### 2023-11-22 globals/metadata.tex
Initial commit

\title{Title of the document}
\author{Author of the document}
\date{\today}
\bibliography{globals/papers}

\newcommand{\acknowledge}{
    Acknowledge
}

\newcommand{\makeabstract}{
    \begin{abstract}
        Abstract of the paper.
    \end{abstract}
}

### 2023-11-22 globals/metadata.tex
Initialize lmcs
Initialize IEEEtran
Initialize lipics

@@
 \title{Title of the document}
-\author{Author of the document}
+
+\author{Jane {Open Access}}%
+       {Dummy University Computing Laboratory, [optional: Address],
+        Country \and
+        My second affiliation, Country \and
+        \url{http://www.myhomepage.edu}
+       }%
+       {johnqpublic@dummyuni.org}%
+       {https://orcid.org/0000-0002-1825-0097}%
+       {(Optional) author-specific funding acknowledgements}
+%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Please provide at least the name of the affiliation and the country.
+The full address is optional.
+Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
+
+\titlerunning{Short title of the document}
 \date{\today}
-\bibliography{globals/papers}
+
+\Copyright{Copyright Holder Name}
+
+%TODO mandatory; please add comma-separated list of keywords
+\keywords{Dummy keyword} 
+\category{} %optional, e.g. invited paper
+\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
 
@@
 }
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-11-24 pnrs.tex
writing the paper

%! TeX program = xelatex
\documentclass[a4paper,11pt]{article}

% INPUTS
\input{globals/packages}
\input{globals/colorscheme}

% CONFIG
\input{globals/macros}
\input{globals/mathdefs}
\input{globals/knowledges.kl}
\input{globals/metadata}

% DOCUMENT
\begin{document}

% FRONT MATTER
\maketitle
\makeabstract
\acknowledge

% MAIN MATTER
\section{Introduction}

There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
This computational model, in the counting case, can be traced back to a model
introduced by \textcite{schutzenberger1965finite}. It has been recently proven
that one can decide the growth rate of such functions, and decide whether they
are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was
conjectured in the PhD thesis of one of the authors, that the membership
problem of $\Nat$-rational series among $\Rel$-rational series was decidable
under the assumption of polynomial growth. This conjecture was based on a
result of \textcite{KARH77}, which we will disprove in \cref{sec:c-example}. We
then provide in \cref{sec:proof} a corrected version of the theorem, together
with a valid proof. As a consequence, it does not invalidate the stated
conjecture.

\AP Before stating the result, let us introduce the necessary vocabulary on
rational series, together with concrete examples so that no doubt is allowed as
to their meaning. A $\Rel$-rational series $f \colon \Sigma \to \Rel$ is
\intro{commutative} whenever $f(uv) = f(vu)$ for all $u,v \in \Sigma^*$. A
polynomial $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational
polynomial} if there exists a \reintro{commutative} $\Rel$-rational series $f$
such that $f(w) = P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, a
\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
series is in fact $\Nat$-rational.

\begin{example}
    With $\Sigma \defined \set{1}$,
    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
    not a \kl{$\Nat$-rational polynomial}.
\end{example}

\begin{example}
    Over a binary alphabet $\Sigma \defined \set{a,b}$,
    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
    but is
    not a \kl{$\Nat$-rational polynomial}.
\end{example}

\AP Let us now give some vocabulary on polynomials with multiple indeterminate
over $\Rel$. 
A \intro{monomial} is a product of variables and integers.
\begin{example}
    $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, 
    but $X + Y$ or $2X^2 + XY$ are not.
\end{example}
A polynomial $P \in \Rel[X_1, \dots, X_k]$ has \intro{non-negative}
if $P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we
do not consider negative values, as the numbers $n_i$ will ultimately count the
number of occurrences of a letter in a word.
\begin{example}
    The polynomial $(X - Y)^2$ is \kl{non-negative},
    and so is the polynomial $X^3$, but the polynomial
    $X^2 - 2X$ is not.
\end{example}
In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
is a \intro{maximal monomial} if it is a maximal element
for the visibility ordering of polynomials \emph{in $\Rat$}
among monomials with non-zero coefficients in $P$.\footnote{
    One could equivalently define maximal elements by removing the (non-zero)
    multiplicative constant of the monomials, and then
    considering the usual divisibility relation of $\Rel$ polynomials.
}
\begin{example}
    In the polynomial 
    $P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
    are $X^2$, $-2 XY$, and $X^2$.
\end{example}
A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$
if its multiplicative constant is positive.
\begin{example}
    In the polynomial $P(X,Y) \defined (X - Y)^2$, 
    the \kl{positive monomials} are
    $X^2$ and $Y^2$.
\end{example}

We are now ready to restate the theorem
of \citeauthor{KARH77}, in terms of its
class $\CoveredPoly$ of polynomials defined as follows:
\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
    Let $\vec{X}$ be a finite tuple of indeterminate.
    The class $\CoveredPoly[\vec{X}]$
    is the class of polynomials $P \in \Rel[\vec{X}]$
    that are \kl{non-negative}
    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
    When the indeterminate are clear from the context, we write
    this class $\CoveredPoly$.
\end{definition}

\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
    Let $\Sigma$ be a finite alphabet.
    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
\end{theorem}

\section{The Counter Example}
\label{sec:c-example}

Let us now introduce the counter example that was found trying to
replicate the original proof. Notice that the counter example
will use three indeterminate, because we will prove
in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds
on polynomial with at most two indeterminate.

\begin{definition}[Counter Example Polynomial]
    \label{def:bad-polynomial}
    We define $P(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
\end{definition}

\begin{theorem}
    \label{thm:counter-example}
    Over $\Sigma \defined \set{a,b,c}$,
    there exists a polynomial $P \in \CoveredPoly$ that is not
    a \kl{$\Nat$-rational polynomial}. Namely,
    $P$ is the polynomial of \cref{def:bad-polynomial}.
\end{theorem}
\begin{proof}
    Let us write $X,Y,Z$ instead of $X_a, X_b, X_c$ to improve
    readability.
    It is clear that $P$ is \kl{non-negative}. We can develop
    the expression of $P$ to 
    obtain
    $P = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
    The \kl{maximal monomials} of $P$
    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
    \kl{positive coefficients}.

    Assume by contradiction that $P$ is a \kl{$\Nat$-rational polynomial}.
    There exists a \kl{commutative}
    $\Nat$-rational series $f \colon \Sigma \to \Nat$
    such that $P(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
    Remark that $P(X,Y,Z) = 0$
    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
    $P(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
    $Z \neq 0$, and $X = Y = 0$.

    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
    Because $L \cap \set{a,b}^*$ is notoriously not a regular language, we
    conclude that $L$ is not a regular language.
    However, $L = f^{-1}(\set{0})$, which is a regular language
    because $f$ is a $\Nat$-rational series (\textbf{TODO: cite some result?}).
\end{proof}

\begin{corollary}
    The result stated in \cite[Theorem 3.3]{KARH77} is false
    for all polynomials with at least $3$ indeterminate.
\end{corollary}

As we will see in the next section, the characterization
of \kl{$\Nat$-rational polynomials} using the set $\CoveredPoly$
holds when the polynomials have at most $2$ indeterminate. This shows
that the examples in \cite{KARH77} are not invalidated, as they
all use at most two indeterminate.

\section{The Corrected Theorem}
\label{sec:proof}

Let us now provide another class of polynomials, $\CorrectPoly$. The definition
will involve fixing the value of indeterminate, for which we use the following
notation: $\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally,
if $\nu$ is a partial function from $\vec{X}$ to $\Nat$, the restriction
$\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminate $\vec{Y}
\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the
domain of $\nu$.

\begin{definition}
    Let $\vec{X}$ be a finite tuple of indeterminate.
    The class $\CorrectPoly[\vec{X}]$ is the collection of
    polynomials $P \in \Rel[\vec{X}]$ such that
    $P$ is 
    such that, for every partial function $\nu \colon \vec{X} \to \Nat$,
    every \kl{maximal monomial} of
    $\restr{P}{\nu}$ has a \kl{positive coefficient}.
\end{definition}

First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
polynomials in $\CorrectPoly$ are \kl{non-negative}.

Before proving that $\CorrectPoly$ captures \kl{$\Nat$-rational polynomials},
let us first check that the counter example provided in
\cref{thm:counter-example} is not in $\CorrectPoly$. For that, notice that for
$P$ defined in \cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which
has a negative coefficient for a \kl{maximal monomial}, namely $-2XY$.

Let us now prove that \kl{$\Nat$-rational polynomials} are
in $\CorrectPoly$.
To simplify the reasoning, we will use results on $\Nat$-polyregular
functions, also known as, polyregular functions with unary output.

\begin{lemma}
    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
\end{lemma}
\begin{proof}[Proof Idea]
    Let $P$ be a \kl{$\Nat$-rational polynomial}.
    Let $\Sigma$  be the finite alphabet such that
    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
    $\Nat$-rational series computing $P$.
    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
    it is a $\Nat$-polyregular function \textbf{CITE}.
    In particular, 
    there exists a tuple $\vec{x}$ of free variables
    $\MSO$ formulas $\psi_1(\vec{x}), \dots, \psi_n(\vec{x})$,
    and positive coefficients $\alpha_1, \dots, \alpha_n \in \Nat$,
    such that
    \begin{equation*}
        f(w) = \left(\sum_{i = 1}^n \alpha_i \vcount{\psi_i}\right)(w) \quad .
    \end{equation*}

    \textbf{TODO: only look at words of the form $\prod_{a \in \Sigma} a^{k_a}$,
    because this is enough}.
    \textbf{TODO: this has positive maximal coefficients}.
    Furthermore, this remains true when restricting the language.
\end{proof}

\begin{fact}
    \label{fact:n-poly-n-poly}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{fact}

The core of the upcoming proof of \cref{lem:correct-to-n-rat} relies on the
following observation: if a polynomial $P$ is written $X^p + Q$, where all
terms of $Q$ have degree less than $p$, then one can find some large enough $K$
so that $X^p - (X - K)^p + Q$ also has a positive maximal coefficient.
It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence
the maximal coefficient can be chosen arbitrarily large by increasing
$K$.
This
trick will work similarly in multivariate polynomials, and the overall
proof scheme works as follows: for a bounded number of possible
values of $X$, we can fix the variable, and have one less item to work on.
For large enough values of $X$ (typically greater than some $K$), we can guarantee that
$X^p - (X - K)^p + Q$ has positive maximal coefficients, hence
can be computed by a $\Nat$-rational series, while
$(X - K)^p$ is clearly computable by a $\Nat$-rational series,
hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
as the sum of two $\Nat$-rational series, and we have
concluded that our original polynomial was a
\kl{$\Nat$-rational polynomial}. There are a number of problems in generalizing
this proof scheme:
\begin{itemize}
    \item We will have to guarantee that every polynomial considered
        is non-negative
    \item We do not only need to check that the \kl{maximal monomials}
        have \kl{positive coefficient}, but that it is the case
        when fixing arbitrarily many variables! (this
        covers the first point too)
\end{itemize}

\begin{lemma}
    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
    $\vec{p} \colon \vec{X}Y \to \Nat$,
    and $M \defined \times \prod_{X \in \vec{X}} X^{p(X)}$.
    Then, for all $K \geq 0$,
    $\Delta_{Y,K} \defined Y^p M - (Y - K)^p M$ has \kl{maximal monomials} with
    \kl{positive coefficients} of weight at least $K$, regardless of
    the variables fixed, as long as the (potential)
    value of $Y$ is greater
    than $K$.
\end{lemma}
\begin{proof}
    Let us first remark that
    \begin{equation*}
        \Delta_{Y,K}
        = KY^{p-1} M + Q M
        \quad .
    \end{equation*}
    In particular, there is a unique \kl{maximal monomial} for $\Delta_{Y,K}$,
    which is precisely $K Y^{p-1} M$, because all monomials
    in $Q M$ divide (strictly) $Y^{p-1} M$.
    Let us fix a variable,
    $X$ to any (non-zero) value $p > 0$. Then
    The \kl{maximal monomial} of $\restr{\Delta_{Y,K}}{X = p}$
    is exactly $\restr{KY^{p-1}M}{X = p}$, which remains
    with a \kl{positive coefficient}, that is proportional to $K$.
    Let us fix $Y$ to some value $K + l$, with $l > 0$,
    then
    $\restr{\Delta_{Y,K}}{Y = K + l}
    = (K + l)^p M - l^p M
    = ((K + l) - l) (\sum_{i = 1}^{p-1} (K + l)^i l^{p-i}) M
    = K \alpha M
    $
    which remains a monomial with a \kl{positive coefficient},
    that is proportional to $K$.
\end{proof}

We can refine the above lemma to use it in the case where
we do not have a single
monomial.

\begin{lemma}
    \label{lem:delta-cool}
    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
    \kl{maximal monomial} of $P$, with $\alpha > 0$.
    Then, there exists $K \geq 0$
    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
    has \kl{positive coefficients} for \kl{maximal monomials}
    in any partial evaluation that satisfies $Y > K$ (if it
    gives a value to $Y$).
\end{lemma}
\begin{proof}
    Let $N$ be greater than the sum of the absolute values of all coefficients
    appearing in $P$, and let $K = 2 N$.

    Let us remark that
    \begin{equation*}
        \Delta_K = \underbrace{(P - \alpha Y^p M)}_{\text{ left hand side }}
        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{\text{ right hand side }}
        \quad .
    \end{equation*}
    an that for every partial evaluation $\nu$ that sends $Y$ to a value
    greater than $K$,
    we have that $\restr{(Y^p M - (Y - K)^p M)}{\nu}$ has
    \kl{maximal monomials} with \kl{positive coefficients}
    proportional to $K$ (note that here ``proportional"
    implies that they are greater than $K$, because we multiply by positive
    integers).

    To conclude, it is just a game of describing how
    \kl{maximal monomials} of $\restr{\Delta_Y}{\nu}$
    can appear. If the \kl{maximal monomial} is obtained
    by summing an element of the left hand side with an element
    of the right hand side, it needs to be a \kl{maximal monomial}
    of the right hand side, and the large value of $K$ guarantees
    that the sum of coefficients is positive.
    If it only appears on the right hand side, it is even better.
    If it only appears on the left hand side, then
    in particular it means that the monomial does not
    divide $\restr{Y^p M}{\nu}$, hence has the same value
    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
    which is guaranteed to be positive because $P \in \CorrectPoly$.

    In particular, we conclude from this that
    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
\end{proof}

Note that we cannot conclude in the above lemma that the result is in
$\CorrectPoly$, because we have a restriction on the range where the variable
$Y$ can be taken.

\begin{lemma}
    \label{lem:correct-to-n-rat}
    Polynomials in $\CorrectPoly$ are
    \kl{$\Nat$-rational polynomials}.
\end{lemma}
\begin{proof}[Proof Sketch]
    We prove by induction on the number $d$ of variables
    and the ``higman ordering of polynomials" (we look at the multiset
    of tuples of things ...) that for every selection
    $K_1, \dots, K_n$ of lower bounds for the different variables
    (that can be $0$),
    functions that are in $\CorrectPoly$ \emph{over tuples of size
    at least nanania} can be computed by a $\Nat$-rational series
    that outputs $0$ outside the domain and the correct value
    inside.

    For $d = 0$,
    the result is true because
    a constant function is a \kl{$\Nat$-rational polynomial}
    if and only if it is \kl{non-negative}.

    Let us now assume that $d > 0$, and in particular
    there is a variable $Y$.
    Consider $P \in \CorrectPoly[\vec{X} Y]$.
    If $Y$ appears at least one in $P$, then it must appear
    in one of the \kl{maximal monomials} of $P$. Otherwise,
    $P \in \CorrectPoly[\vec{X}]$ and we conclude using the
    induction hypothesis.
    Let $\alpha Y^p M$ be a \kl{maximal monomial} in $P$.
    Using \cref{lem:delta-cool},
    we know that there exists $K$ such that
    $P - \alpha (Y - K)^p M$ almost belongs to $\CorrectPoly$,
    and has a strictly smaller "degree multiset".
    By induction hypothesis,
    we conclude that
    the following function is a $\Nat$-rational series
    (beware that we have to consider the lower bounds to apply
    the induction hypothesis):
    \begin{equation*}
        A \defined
        \begin{cases}
            0 & \text{ if some variable is bounded by its } K_i \\
            0 & \text{ otherwise if  } Y \leq K \\
            P - \alpha(Y - K)^pM & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

    Because $\alpha \geq 0$, it is an easy check
    that the following function is also a $\Nat$-rational series
    \begin{equation*}
        B \defined
        \begin{cases}
            0 & \text{ if some variable is bounded by its } K_i \\
            0 & \text{ otherwise if } Y \leq K \\
            \alpha(Y - K)^pM & \text{ otherwise }
        \end{cases}
    \end{equation*}

    Remark that by fixing some of the variables
    to values 
    below the bounds $K_i$ (and potentially $K$ for $Y$)
    we obtain a \emph{finite} number of polynomials with strictly fewer
    indeterminate.
    As a consequence, the following function is computable
    as a $\Nat$-rational series:
    \begin{equation*}
        C \defined
        \begin{cases}
            P & \text{ if some variable is bounded by its } K_i \\
            P & \text{ otherwise if } Y \leq K \\
            0 & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}

    We conclude because
    $P = A + B + C$, all of which are $\Nat$-rational series.
\end{proof}

\begin{theorem}
    The \kl{$\Nat$-rational polynomials} are exactly
    the polynomials in $\CorrectPoly$.
\end{theorem}

\begin{remark}
    In the proof, we actually conclude that
    the \kl{$\Nat$-rational polynomials}
    are exactly those that can be computed 
    by \kl{commutative}
    \emph{star-free} $\Nat$-polyregular functions.
\end{remark}

\begin{lemma}
    \label{lem:correct-covered-2}
    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
\end{lemma}
\begin{proof}
    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
    by considering the empty valuation $\nu \colon \set{X,Y} \to \Nat$.
    For the converse inclusion, let us consider $P(X,Y)$
    that is \kl{non-negative}, such that the \kl{maximal monomials}
    have \kl{positive coefficients}.

    If we fix none of the variables, then the \kl{maximal monomials}
    have \kl{positive coefficients} by assumption. If we fix one of the
    variables, we can assume without loss of generality that we 
    fix $X = k$ for some $k \in \Nat$.
    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
    and therefore must have a positive leading coefficient
    (which is the unique \kl{maximal monomial} in this case)
    or be constant equal to 0. In both cases, the \kl{maximal monomials}
    have \kl{positive coefficients}.
    The same reasoning applies \emph{a fortiori} in the case where
    we fix the two indeterminate, leading to a constant polynomial.
\end{proof}

\section{Conclusion}
\label{sec:ccl}

% BACKMATTER
\printbibliography

\appendix

\end{document}

### 2023-12-22 globals/metadata.tex
Initialize mscs
writing the paper
update

--- 2023-11-22 globals/metadata.tex
@@
-\title{Title of the document}
-\author{Author of the document}
+\title{A corrected version of Karhumäki’s theorem
+       on polynomial $\Nat$-rational series}
+\author{
+    Lopez Aliaume
+}
 \date{\today}
@@
 \newcommand{\acknowledge}{
-    Acknowledge
+    I would like to thank Gaëtan Douéneau-Tabot
+    for mentionning this problem, checking 
+    my counter-example, and supporting me in publishing
+    this paper.
 }
@@
     \begin{abstract}
-        Abstract of the paper.
+        We propose a short counter-example to a
+        previous result on commutative $\Nat$-rational series
+        by Karhumäki, and provide a proof for
+        the corrected statement.
+        This opens the way to the decidability of
+        $\Nat$ polyregular functions 
+        among $\Rel$ porylegular functions, which is
+        an ongoing research.
     \end{abstract}

### 2023-12-22 pnrs.tex
update

@@
 \documentclass[a4paper,11pt]{article}
+
+\usepackage[left=3cm,right=3cm]{geometry}
 
@@
 
+Let us first state the result 
+of \citeauthor{KARH77}, in terms of its
+class $\CoveredPoly$ of polynomials defined as follows:
+\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
+    Let $\vec{X}$ be a finite tuple of indeterminate.
+    The class $\CoveredPoly[\vec{X}]$
+    is the class of polynomials $P \in \Rel[\vec{X}]$
+    that are \kl{non-negative}
+    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
+    When the indeterminate are clear from the context, we write
+    this class $\CoveredPoly$.
+\end{definition}
+
+\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
+    \label{karh:thm}
+    Let $\Sigma$ be a finite alphabet.
+    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
+    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
+\end{theorem}
+
+\paragraph*{Context.}
 There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
@@
 
-\AP Before stating the result, let us introduce the necessary vocabulary on
+\AP
+\paragraph*{Rational series and regular languages.}
+\begin{fact}
+    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
+    is not regular.
+\end{fact}
+
+\begin{fact}
+    regular languages are closed under intersections.
+\end{fact}
+
+TODO.
+
+clasical results from \cite{berstel2011noncommutative}.
+\begin{fact}
+    \label{pre-image-regular:fact}
+    The pre-image of a regular language by a $\Nat$-rational series
+    is a regular language.
+\end{fact}
+
+\begin{fact}
+    \label{polynomial-rational-polyreg:fact}
+    polynomial + rational = unary polyregular.
+\end{fact}
+
+\AP
+\paragraph*{Polynomials.}
+Before stating the result, let us introduce the necessary vocabulary on
 rational series, together with concrete examples so that no doubt is allowed as
@@
 \begin{example}
+    \label{negative-not-nrat:ex}
     With $\Sigma \defined \set{1}$,
@@
 \end{example}
+\begin{proof}
+    TODO:
+    The polynomial $X_1$ is computed by the rational series,
+    the polynomial $X_1^2 + 3$ is computed by,
+    and the polynomial
+    $- X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
+    have non-negative output.
+\end{proof}
+
+We saw in \cref{negative-not-nrat:ex}
+a simple criterion to check whether a polynomial is $\Nat$-rational,
+but being non-negative is not enough, as the following example
+illustrates. Note that the proof scheme
+of this example will be at the core of \cref{thm:counter-example}.
 
@@
 \end{example}
+\begin{proof}
+    Assume by contradiction that
+    $S$ is a $\Nat$-rational series computing $P$.
+    Then, $S^{-1}(\set{0})$ should be a regular language
+    (\cref{pre-image-regular:fact}),
+    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
+    is not a regular languge.
+\end{proof}
 
@@
 \end{example}
-A polynomial $P \in \Rel[X_1, \dots, X_k]$ has \intro{non-negative}
+A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative}
 if $P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we
@@
 \end{example}
-
-We are now ready to restate the theorem
-of \citeauthor{KARH77}, in terms of its
-class $\CoveredPoly$ of polynomials defined as follows:
-\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CoveredPoly[\vec{X}]$
-    is the class of polynomials $P \in \Rel[\vec{X}]$
-    that are \kl{non-negative}
-    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
-    When the indeterminate are clear from the context, we write
-    this class $\CoveredPoly$.
-\end{definition}
-
-\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
-    Let $\Sigma$ be a finite alphabet.
-    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
-    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
-\end{theorem}
 
@@
     However, $L = f^{-1}(\set{0})$, which is a regular language
-    because $f$ is a $\Nat$-rational series (\textbf{TODO: cite some result?}).
+    because $f$ is a $\Nat$-rational series
+    (\cref{pre-image-regular:fact}).
 \end{proof}
@@
 \begin{corollary}
-    The result stated in \cite[Theorem 3.3]{KARH77} is false
+    The result stated in \cite[Theorem 3.3]{KARH77}, restated
+    in \cref{karh:thm}, is false
     for all polynomials with at least $3$ indeterminate.
@@
     Because $f$ has polynomial growth, and is a $\Nat$-rational series,
-    it is a $\Nat$-polyregular function \textbf{CITE}.
+    it is a $\Nat$-polyregular function 
+    (\cref{polynomial-rational-polyreg:fact}).
     In particular, 
@@
 
+\begin{itemize}
+    \item Actually, we have an algorithm that allows
+        us to \emph{decide} if a polynomial is $\Nat$-rational.
+    \item Actually, we obtain a characterisation of
+        commutative 
+        $\Nat$-rational series of polynomial growth
+        among $\Rel$-rational series.
+    \item The question of $\Nat$-polyregular functions
+        among $\Rel$-polyregular functions remains open
+        because of the non-commutativity of the input.
+    \item The question of $\Nat$-rational series (commutative)
+        among $\Rel$-rational series remains open.
+\end{itemize}
+
 % BACKMATTER

### 2023-12-26 globals/metadata.tex
looking again

--- 2023-11-22 globals/metadata.tex
@@
-\title{Title of the document}
-\author{Author of the document}
+\title{A corrected version of Karhumäki’s theorem
+       on commutative polynomial $\Nat$-rational series}
+\author{
+    Lopez Aliaume
+}
 \date{\today}
@@
     \begin{abstract}
-        Abstract of the paper.
+        Deciding whether a $\Rel$-rational series
+        can be computed as a $\Nat$-rational series
+        is an open problem that was solved by Karhumäki in the
+        specific case of $\Rel$-rational series that are commutative
+        and compute a polynomial.
+        We propose a short counter-example to this
+        previous result on commutative $\Nat$-rational series,
+        and prove a corrected version of the theorem, that only
+        requires commutativity and a polynomial growth rate.
+        We then discuss how the syntactic criterion could
+        pave the way towards a decision procedure when the commutativity
+        of the $\Rel$-rational series
+        is not assumed.
     \end{abstract}

### 2023-12-26 pnrs.tex
looking again

@@
 
-Let us first state the result 
-of \citeauthor{KARH77}, in terms of its
-class $\CoveredPoly$ of polynomials defined as follows:
+We will start by formally stating the result of \citeauthor{KARH77}
+in terms of polynomials and commutative $\Nat$-rational series, and
+then introduce the necessary definitions. The characterization of
+commutative $\Nat$-rational series computing polynomial functions
+is described in terms of a class $\CoveredPoly[\vec{X}]$ of
+polynomials with multiple indeterminate.
+
 \begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
@@
 
+We use the term \kl{$\Nat$-rational polynomial}, that will formally be
+stated later on, to denote polynomials that can be computed via a
+$\Nat$-rational series. In the upcoming theorem, this property
+which is \emph{a priori} difficult to decide is reduced to 
+the positivity of the polynomial, and the positivity of \emph{some}
+coefficients.
+
 \begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
@@
 
-\paragraph*{Context.}
-There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
-\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
-This computational model, in the counting case, can be traced back to a model
-introduced by \textcite{schutzenberger1965finite}. It has been recently proven
-that one can decide the growth rate of such functions, and decide whether they
-are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was
-conjectured in the PhD thesis of one of the authors, that the membership
-problem of $\Nat$-rational series among $\Rel$-rational series was decidable
-under the assumption of polynomial growth. This conjecture was based on a
-result of \textcite{KARH77}, which we will disprove in \cref{sec:c-example}. We
-then provide in \cref{sec:proof} a corrected version of the theorem, together
-with a valid proof. As a consequence, it does not invalidate the stated
-conjecture.
+It is not clearly stated in \cite{KARH77}, but \cref{karh:thm} implies the
+decidability of \kl{$\Nat$-rational polynomial} inside polynomials with $\Rel$
+coefficients. We will not formalize this statement yet, as \cref{karh:thm}
+does not hold in general.
 
-\AP
-\paragraph*{Rational series and regular languages.}
-\begin{fact}
+\paragraph*{Context.} There has been recent interest in $\Rel$ and
+$\Nat$-rational series exhibiting \emph{polynomial growth}
+\cite{doueneau2021pebble,bojanczyk2022transducers}. This computational model,
+in the counting case, can be traced back to a model introduced by
+\textcite{schutzenberger1965finite}. It has been recently proven that one can
+decide the growth rate of such functions, and decide whether they are aperiodic
+in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was conjectured in
+the PhD thesis of Gaëtan Douéneau-Tabot, that the membership problem of
+$\Nat$-rational series among $\Rel$-rational series was decidable under the
+assumption of polynomial growth. This conjecture was based on the previously
+introduced result of \textcite{KARH77} (\cref{karh:thm}), which we will
+disprove in \cref{sec:c-example}. We then provide in \cref{sec:proof} a
+corrected version of the theorem, together with a valid proof. As a
+consequence, it does not invalidate the stated conjecture.
+
+\begin{itemize}
+    \item commutative input and output
+    \item commutative output
+    \item no commutativity ?
+\end{itemize}
+
+\AP \paragraph*{Rational series and regular languages.} This paper uses notions
+of regularity and results about \kl{$\Nat$-rational series}, however most of
+the technicalities will stem from the multiple indeterminate of polynomials,
+and all there is to know about regularity in this document will be confined to
+the following facts. For a survey about (non-commutative) rational series, we
+refer the reader to \cite{berstel2011noncommutative}.
+
+\begin{definition}[Rational series]
+    The following are equivalent
+    \begin{itemize}
+        \item weighted automata
+        \item rational expressions
+        \item 
+    \end{itemize}
+\end{definition}
+
+\begin{definition}[Commutative rational series]
+    A $\Rel$-rational series
+    $f \colon \Sigma \to \Rel$ is \intro{commutative} whenever $f(uv) = f(vu)$ for
+    all $u,v \in \Sigma^*$.
+\end{definition}
+
+\begin{fact}[Folklore about regular languages]
+    \label{regular:fact}
     The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular.
+    is not regular, whenever $a,b \in \Sigma$.
+    Regular languages are closed under intersection.
 \end{fact}
@@
 \begin{fact}
-    regular languages are closed under intersections.
+    \label{pre-image-regular:fact}
+    The pre-image of a regular language by a $\Nat$-rational series
+    is a regular language. In particular,
+    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
+    language.
 \end{fact}
 
-TODO.
-
-clasical results from \cite{berstel2011noncommutative}.
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by a $\Nat$-rational series
-    is a regular language.
-\end{fact}
-
-\begin{fact}
-    \label{polynomial-rational-polyreg:fact}
-    polynomial + rational = unary polyregular.
-\end{fact}
-
-\AP
-\paragraph*{Polynomials.}
-Before stating the result, let us introduce the necessary vocabulary on
-rational series, together with concrete examples so that no doubt is allowed as
-to their meaning. A $\Rel$-rational series $f \colon \Sigma \to \Rel$ is
-\intro{commutative} whenever $f(uv) = f(vu)$ for all $u,v \in \Sigma^*$. A
-polynomial $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational
-polynomial} if there exists a \reintro{commutative} $\Rel$-rational series $f$
-such that $f(w) = P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, a
+\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
+coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
+\Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational polynomial} if
+there exists a \reintro{commutative} $\Rel$-rational series $f$ such that $f(w)
+= P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, $P$ is a
 \intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
-series is in fact $\Nat$-rational.
+series is in fact $\Nat$-rational. Beware that we only apply the polynomial
+to \emph{non-negative} values.
 
@@
 \begin{proof}
-    TODO:
-    The polynomial $X_1$ is computed by the rational series,
-    the polynomial $X_1^2 + 3$ is computed by,
-    and the polynomial
-    $- X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
+    The function $w \mapsto |w|$ is a \kl{polyregular function} with unary
+    output of \kl{polynomial growth}, hence  can be computed by a
+    \kl{$\Nat$-rational series}. Hence, $P(X_1) \defined X_1$ is
+    a \kl{$\Nat$-rational polynomial}. Similarly,
+    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-rational series},
+    showing that $Q(X_1) \defined X_1^2 + 3$
+    is a \kl{$\Nat$-rational polynomial}.
+    Finally, 
+    $T(X_1) \defined - X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
     have non-negative output.
@@
     but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not a regular languge.
+    is not a regular language (\cref{regular:fact}).
 \end{proof}
@@
 \AP Let us now give some vocabulary on polynomials with multiple indeterminate
-over $\Rel$. 
-A \intro{monomial} is a product of variables and integers.
-\begin{example}
-    $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, 
-    but $X + Y$ or $2X^2 + XY$ are not.
-\end{example}
-A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative}
-if $P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we
-do not consider negative values, as the numbers $n_i$ will ultimately count the
-number of occurrences of a letter in a word.
-\begin{example}
-    The polynomial $(X - Y)^2$ is \kl{non-negative},
-    and so is the polynomial $X^3$, but the polynomial
-    $X^2 - 2X$ is not.
-\end{example}
+over $\Rel$. A \intro{monomial} is a product of variables and integers. For
+instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, but $X + Y$ or
+$2X^2 + XY$ are not.
+
+\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} if
+$P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we do
+not consider negative values, as the numbers $n_i$ will ultimately count the
+number of occurrences of a letter in a word. As an example, the polynomial $(X
+- Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$, but the
+polynomial $X^2 - 2X$ is not.
+
+\AP
 In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
@@
 }
-\begin{example}
 In the polynomial 
@@
 are $X^2$, $-2 XY$, and $X^2$.
-\end{example}
-A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$
-if its multiplicative constant is positive.
-\begin{example}
-    In the polynomial $P(X,Y) \defined (X - Y)^2$, 
-    the \kl{positive monomials} are
-    $X^2$ and $Y^2$.
-\end{example}
+A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$ if its
+multiplicative constant is positive. In the polynomial $P(X,Y) \defined (X -
+Y)^2$, the \kl{positive monomials} are $X^2$ and $Y^2$.
+
+\paragraph*{Outline of the paper.}
+todo.
 
@@
 
-Let us now introduce the counter example that was found trying to
-replicate the original proof. Notice that the counter example
-will use three indeterminate, because we will prove
-in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds
-on polynomial with at most two indeterminate.
+Let us now introduce the counter example that was found trying to generalise
+the original proof to non-commutative \kl{$\Rel$-rational series} of
+\kl{polynomial growth}. The counter example will use three indeterminate, and
+we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
+polynomial with at most two indeterminate.
 
@@
     \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ is notoriously not a regular language, we
+    Because $L \cap \set{a,b}^*$ is notoriously not a regular language
+    (\cref{regular:fact}), we
     conclude that $L$ is not a regular language.
@@
 
-Let us now provide another class of polynomials, $\CorrectPoly$. The definition
-will involve fixing the value of indeterminate, for which we use the following
-notation: $\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally,
-if $\nu$ is a partial function from $\vec{X}$ to $\Nat$, the restriction
+Let us now provide another class of polynomials, $\CorrectPoly$, that restricts
+$\CoveredPoly$ to a subset that is ``closed under fixing indeterminate", namely,
+polynomials $P(X_1, \dots, X_n) \in \CoveredPoly$ such that partially evaluated
+polynomials remain in $\CoveredPoly$. We use the following notation
+to fix the value of some indeterminate
+$\intro*\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally, if $\nu$ is
+a partial function from $\vec{X}$ to $\Nat$, the restriction
 $\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminate $\vec{Y}
-\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the
-domain of $\nu$.
+\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the domain of
+$\nu$.
 
@@
 First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
-polynomials in $\CorrectPoly$ are \kl{non-negative}.
+polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
+counter example provided in \cref{thm:counter-example} is not in
+$\CorrectPoly$. For that, notice that for $P$ defined in
+\cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
+coefficient for a \kl{maximal monomial}, namely $-2XY$.
 
-Before proving that $\CorrectPoly$ captures \kl{$\Nat$-rational polynomials},
-let us first check that the counter example provided in
-\cref{thm:counter-example} is not in $\CorrectPoly$. For that, notice that for
-$P$ defined in \cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which
-has a negative coefficient for a \kl{maximal monomial}, namely $-2XY$.
+Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
+To simplify the reasoning, we will use results on $\Nat$-polyregular functions,
+also known as, polyregular functions with unary output
+\cite{doueneau2021pebble,bojanczyk2019string}. Note that the connection between
+rational series and polyregular functions is not specific to $\Nat$, as
+$\Rel$-rational series of polynomial growth are computed as polyregular
+functions with output in $\set{+1,-1}$ post-composed with computing the sum of
+outputted letters \cite{LOPEZ23b}. Because we only consider functions with
+output in $\Rel$, and that a commutative output monoid simplifies the definitions,
+let us only define $\Rel$-polyregular functions.
 
-Let us now prove that \kl{$\Nat$-rational polynomials} are
-in $\CorrectPoly$.
-To simplify the reasoning, we will use results on $\Nat$-polyregular
-functions, also known as, polyregular functions with unary output.
+\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
+    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
+    be a morphism, $k \in \Nat$, and 
+    $\pi \colon M^k \to \Rel$ be a production function.
+    The \intro{$\Rel$-polyregular function}
+    $\pi^\dagger \colon \Sigma^* \to \Rel$
+    is computed as follows:
+    \begin{equation*}
+        f(w) \defined
+        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
+        \quad .
+    \end{equation*}
+
+    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
+    is called \intro{$\Nat$-polyregular}.
+\end{definition}
+
+\begin{fact}[see e.g. ...]
+    \label{polynomial-rational-polyreg:fact}
+    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
+    be a function. The following are equivalent:
+    \begin{enumerate}
+        \item $f$ can be computed as a \k{$\Nat$-rational series}
+            of \kl{polynomial growth}.
+        \item $f$ is a \kl{$\Nat$-polyregular function}.
+    \end{enumerate}
+\end{fact}
+
+\begin{fact}
+    \label{fact:n-poly-n-poly}
+    Let $P \in \Nat[\vec{X}]$. Then, $P$
+    is a \kl{$\Nat$-rational polynomial}.
+\end{fact}
 
@@
     In particular, 
-    there exists a tuple $\vec{x}$ of free variables
-    $\MSO$ formulas $\psi_1(\vec{x}), \dots, \psi_n(\vec{x})$,
-    and positive coefficients $\alpha_1, \dots, \alpha_n \in \Nat$,
-    such that
+    there exists a finite monoid $M$,
+    a morphism $\mu \colon \Sigma^* \to M$,
+    an integer $p  \in \Nat$,
+    and a production function $\pi \colon M^p \to \Nat$,
+    such that for all words $w \in \Sigma^*$,
     \begin{equation*}
-        f(w) = \left(\sum_{i = 1}^n \alpha_i \vcount{\psi_i}\right)(w) \quad .
+        f(w) = \sum_{w = u_1 \cdots u_p} \pi(\mu(u_1), \dots, \mu(u_p)) \quad .
     \end{equation*}
+    Because $f$ is \kl{commutative}, we can restrict our attention
+    to words of the form $\prod_{a \in \Sigma} a^{k_a}$ to compute $f$.
+    To that end, let us
+    consider $\omega \in \Nat$ such that
+    $m^\omega$ is an idempotent for all $m \in M$ (this is possible because
+    $M$ is finite).
+    Then, $f( \prod_{a \in \Sigma} a^{ \omega \times k_a} )$
+    can be computed as follows:
+    \begin{align*}
+    f(\prod_{a \in \Sigma} a^{ \omega \times k_a} )
+    &= 
+    \sum_{\prod_{a \in \Sigma} a^{\omega \times k_a}
+    = u_1 \cdots u_p} 
+         \pi(\mu(u_1), \dots, \mu(u_p)) \\
+    &= \pi(a^\omega, \dots, a^\omega) 
+    \end{align*}
+    is a polynomial in $k_a$, whose maximal coefficients are
+    sums of $\alpha_i$.
 
-    \textbf{TODO: only look at words of the form $\prod_{a \in \Sigma} a^{k_a}$,
-    because this is enough}.
+    Furthermore, $f(w) = P(\seqof{\card[a]{w}}{a \in \Sigma})$,
+    hence we can identify coefficients.
     \textbf{TODO: this has positive maximal coefficients}.
@@
 \end{proof}
-
-\begin{fact}
-    \label{fact:n-poly-n-poly}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{fact}
 
@@
 
+\section{Deciding commutative $\Nat$-polyregular functions}
+\label{sec:deciding}
+
+decide nat inside rel in the commutative input case.
+
 \section{Conclusion}
@@
 
+\begin{corollary}
+    The following problem is decidable: given $P \in \Rel[\vec{X}]$,
+    decide whether $P$ is a \kl{$\Nat$-rational polynomial}.
+\end{corollary}
+\begin{proof}
+    Thanks to \cref{karh:thm}, one only has to decide whether
+    $P \in \CoveredPoly$. The procedure is as follows.
+    \begin{enumerate}
+        \item Check if all \kl{maximal monomials} have
+            \kl{positive coefficients} in $P$, if not, answer \texttt{false}.
+        \item Consider $N \in \Nat$ large enough so that
+              $P \geq 0$ when all indeterminate are chosen greater
+              than $N$. This is possible because all \kl{maximal
+              monomials} have \kl{positive coefficients}, and such an $N$
+              is computable.
+        \item Compute $P$
+    \end{enumerate}
+
+    Now, if all \kl{maximal monomials}
+    have \kl{positive coefficients}, then there exists a computable
+    $N \in \Nat$ such that 
+    In particular,
+
+\end{proof}
+
 % BACKMATTER

### 2023-12-26 pnrs.tex
advancing in the results

@@
 
+Let us assert that polynomials that have all coefficients in $\Nat$ are
+\kl{$\Nat$-rational polynomials}, which will be a base case in an upcoming
+induction.
+
 \begin{fact}
@@
 \end{fact}
+
+\begin{lemma}
+    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
+    There exists $\omega \in \Nat$
+    such that for all $p \in \Nat$,
+    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
+    for all $u_1, \dots, u_p \in \Sigma^*$,
+    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
+    whose \kl{maximal monomials} have \kl{positive coefficients},
+    and such that:
+    \begin{equation*}
+        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
+        = P(X_1, \dots, X_p) \quad .
+    \end{equation*}
+\end{lemma}
 
@@
 \begin{theorem}
+    \label{corrected-version:thm}
     The \kl{$\Nat$-rational polynomials} are exactly
@@
 
-decide nat inside rel in the commutative input case.
+The goal of this section is to go beyond polynomials and characterize all
+$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
+that they are \kl{commutative}.
+
+\begin{theorem}
+    \label{decidable-n-poly:thm}
+    Given a \kl{commutative}
+    \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to \Rel$,
+    one can decide if $f$ can be computed by a \kl{$\Nat$-polyregular function},
+    and effectively compute a representation of $f$.
+\end{theorem}
+
+\begin{lemma}
+    \label{decide-correct-poly:lem}
+    One can decide if a polynomial is in $\CorrectPoly$.
+\end{lemma}
+\begin{proof}
+    It is easy to check that maximal monomials have positive coefficients.
+    Assuming that it is the case, remark that 
+    every partial valuation $\nu$
+    such that $\restr{P}{\nu}$ has a negative maximal coefficient
+    must have values that are bounded by the sum of absolute values of
+    coefficients appearing in $P$.
+    In particular, one can check all possible choices for $\nu$
+    within these bounds.
+\end{proof}
+
+To prove the theorem, let us first provide a nice characterization of
+\kl{commutative} \kl{$\Rel$-polyregular functions}
+in terms of polynomials.
+
+\begin{lemma}
+    \label{decompose-polynomial:lem}
+    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
+    \kl{$\Rel$-polyregular function}. There exists $\omega \in \Nat$,
+    and polynomials $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{ 0 \leq n_a < \omega}$,
+    such that for all $w \in \Sigma^*$,
+    \begin{equation*}
+        f(w) = P_{\seqof{\card[a]{w} [\omega]}{a \in \Sigma}}
+        (\seqof{\card[a]{w} / \omega}{a \in \Sigma}) \quad .
+    \end{equation*}
+    Which informally means that for every choice of rest modulo $\omega$
+    for every variable, there exists a polynomial describing the
+    behavior of $f$ when adding multiples of $\omega$ letters at a time.
+
+    Furthermore, the polynomials are computable from $f$.
+\end{lemma}
+\begin{proof}
+\end{proof}
+
+As a consequence, we conclude 
+\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
+    We use \cref{decompose-polynomial:lem}
+    to effectively compute
+    a polynomial representation of $f$. Remark that
+    $f$ is a \kl{$\Nat$-polyregular function} if and only if
+    all these polynomials are \kl{$\Nat$-rational polynomials},
+    which is decidable thanks to
+    \cref{decide-correct-poly:lem,corrected-version:thm}.
+\end{proof}
 
@@
 
+\begin{conjecture}
+    $\CorrectPoly = \setof{P}{ P \geq 0 \wedge \exists K \in \Nat, P \circ (+K) \in \Nat[\vec{X}]}$.
+\end{conjecture}
+\begin{proof}[Proof Sketch]
+    If $S \defined P \circ (+K) \in \Nat[\vec{X}]$, then for every partial application,
+    maximal coefficients of $S$ are positive. As maximal coefficients of 
+    (partial applications of) $S$ are the same as the ones of $P$
+    \emph{except for the constant terms} (full application),
+    where $P$ is already guaranteed to be non-negative, we conclude
+    that $P \in \CorrectPoly$.
+
+    Conversely, we prove by induction on the degree of $P \in \CorrectPoly$,
+    that there exists some $K$ such that $P \circ (+K) \in \Nat[\vec{X}]$.
+    First, notice that $\CorrectPoly$ is closed under pre-composition with
+    $(+K)$. Then, remark that $P(X - K)$ has the same maximal coefficients as
+    $P$ and new maximal coefficients that are multiples of $K$. Hence, for a
+    large enough $K$, we conclude that $P(X - K)$ is the sum of its maximal
+    coefficients, plus a polynomial that lies in $\CorrectPoly$. We conclude by
+    induction hypothesis. 
+\end{proof}
+
+\begin{conjecture}
+    $\CorrectPoly$ is precisely composed of polynomials that
+    have non-negative coefficients in a \emph{binomial basis},
+    i.e., where polynomials are obtained
+    as sums and products of $\binom{X_i}{p_i}$.
+\end{conjecture}
+
 % BACKMATTER

### 2023-12-26 pnrs.tex
yaaay

@@
 \input{globals/metadata}
+
+\usepackage{comment}
+%\includecomment{forshort}
+%\excludecomment{proof}
 
@@
 \cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$.
+coefficient for a \kl{maximal monomial}, namely $-2XY$. Let us also assert that
+polynomials that have all coefficients in $\Nat$ are \kl{$\Nat$-rational
+polynomials}, which will be a base case in an upcoming induction.
+
+\begin{fact}
+    \label{fact:n-poly-n-poly}
+    Let $P \in \Nat[\vec{X}]$. Then, $P$
+    is a \kl{$\Nat$-rational polynomial}.
+\end{fact}
 
@@
 
-Let us assert that polynomials that have all coefficients in $\Nat$ are
-\kl{$\Nat$-rational polynomials}, which will be a base case in an upcoming
-induction.
-
-\begin{fact}
-    \label{fact:n-poly-n-poly}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{fact}
-
 \begin{lemma}
+    \label{n-poly-combinatorics:lem}
     Let $f$ be a  \kl{$\Nat$-polyregular} function. 
@@
 \end{lemma}
+\begin{proof}[Proof Sketch]
+    Let $\omega$ be an idempotent power for the finite monoid $M$.
+    We define an equivalence between partitions of a word 
+    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$
+    as follows: two partitions into $k$ words in $\Sigma^*$ are equivalent
+    when normalizing them with the rules
+    $u_i^{\omega} u_i^{\omega} \to u_i^{\omega}$ for all $1 \leq i \leq p$.
+    This equivalence relation has finite index, and all partitions in the same
+    equivalence class yield the same value through $\pi$, because
+    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid $M$.
+
+    Now, let us remark that the number of partitions
+    in a certain equivalence class is proportional to
+    the number of nanania, that is, is of the form
+    $(X_i - K_i)^{n_i}$ for some $K$. We conclude that
+    maximal monomials indeed have positive coefficients.
+    We even conclude that
+    the polynomial is in $\Nat[\vec{X}]$
+    if we shift the variables with a large enough $K$. 
+\end{proof}
 
@@
     (\cref{polynomial-rational-polyreg:fact}).
-    In particular, 
-    there exists a finite monoid $M$,
-    a morphism $\mu \colon \Sigma^* \to M$,
-    an integer $p  \in \Nat$,
-    and a production function $\pi \colon M^p \to \Nat$,
-    such that for all words $w \in \Sigma^*$,
-    \begin{equation*}
-        f(w) = \sum_{w = u_1 \cdots u_p} \pi(\mu(u_1), \dots, \mu(u_p)) \quad .
-    \end{equation*}
-    Because $f$ is \kl{commutative}, we can restrict our attention
-    to words of the form $\prod_{a \in \Sigma} a^{k_a}$ to compute $f$.
-    To that end, let us
-    consider $\omega \in \Nat$ such that
-    $m^\omega$ is an idempotent for all $m \in M$ (this is possible because
-    $M$ is finite).
-    Then, $f( \prod_{a \in \Sigma} a^{ \omega \times k_a} )$
-    can be computed as follows:
-    \begin{align*}
-    f(\prod_{a \in \Sigma} a^{ \omega \times k_a} )
-    &= 
-    \sum_{\prod_{a \in \Sigma} a^{\omega \times k_a}
-    = u_1 \cdots u_p} 
-         \pi(\mu(u_1), \dots, \mu(u_p)) \\
-    &= \pi(a^\omega, \dots, a^\omega) 
-    \end{align*}
-    is a polynomial in $k_a$, whose maximal coefficients are
-    sums of $\alpha_i$.
 
-    Furthermore, $f(w) = P(\seqof{\card[a]{w}}{a \in \Sigma})$,
-    hence we can identify coefficients.
-    \textbf{TODO: this has positive maximal coefficients}.
-    Furthermore, this remains true when restricting the language.
+    Using \cref{n-poly-combinatorics:lem},
+    we conclude that maximal monomials of $P$ have
+    positive coefficients. Furthermore, when fixing variables
+    in $P$, we are computing a new \kl{$\Nat$-polyregular function},
+    hence we also conclude that the maximal monomials have
+    positive coefficients.
+    We have proven that $P \in \CorrectPoly$.
 \end{proof}
@@
 To prove the theorem, let us first provide a nice characterization of
-\kl{commutative} \kl{$\Rel$-polyregular functions}
-in terms of polynomials.
+\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
+This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
+commutative case.
 
@@
 \begin{proof}
+    equivalent partitions of the word, but commutativity?.q
 \end{proof}
@@
 
+\begin{conjecture}
+    star-free $\Nat$-rational is star-free and $\Nat$-rational?
+\end{conjecture}
+
 % BACKMATTER

### 2024-01-01 pnrs.tex
trying to advance

@@
 \section{Introduction}
+
+There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
+\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
+This computational model, in the counting case, can be traced back to a model
+introduced by \textcite{schutzenberger1965finite}. It has been recently proven
+that one can decide the growth rate of such functions, and decide whether they
+are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. Isolating
+a notion of aperiocidity for functions, and deciding this property is often a
+non-trivial process \cite{droste2019aperiodic}. It was conjectured in the PhD
+thesis of Gaëtan Douéneau-Tabot, that the membership problem of $\Nat$-rational
+series among $\Rel$-rational series was decidable under the assumption of
+polynomial growth. The semantic characterization would imply that
+$\Nat$-rational series of polynomial growth that are aperiodic are precisely
+$\Nat$-rational series of polynomial growth, that can be realised by
+$\Rel$-rational series that are aperiodic. This conjecture was based on the
+previously introduced result of \textcite[Theorem 3.3]{KARH77}, which we will
+disprove in \cref{sec:c-example}. 
+We then provide a corrected version of the theorem, and prove the conjecture
+of Douéneau in the case of \emph{commutative} $\Nat$-rational series of
+polynomial growth.
+
+\section{Preliminaries}
 
@@
 
-\paragraph*{Context.} There has been recent interest in $\Rel$ and
-$\Nat$-rational series exhibiting \emph{polynomial growth}
-\cite{doueneau2021pebble,bojanczyk2022transducers}. This computational model,
-in the counting case, can be traced back to a model introduced by
-\textcite{schutzenberger1965finite}. It has been recently proven that one can
-decide the growth rate of such functions, and decide whether they are aperiodic
-in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was conjectured in
-the PhD thesis of Gaëtan Douéneau-Tabot, that the membership problem of
-$\Nat$-rational series among $\Rel$-rational series was decidable under the
-assumption of polynomial growth. This conjecture was based on the previously
-introduced result of \textcite{KARH77} (\cref{karh:thm}), which we will
-disprove in \cref{sec:c-example}. We then provide in \cref{sec:proof} a
+We then provide in \cref{sec:proof} a
 corrected version of the theorem, together with a valid proof. As a
@@
 
-\section{Deciding commutative $\Nat$-polyregular functions}
+\section{Effective procedures}
 \label{sec:deciding}
@@
 \end{proof}
+
+\subsection{Star-free $\Nat$-polyregular functions}
+
+\begin{definition}[Ultimately polynomial]
+\end{definition}
+
+\begin{fact}
+    Any $\Nat$-rational polynomial is a star-free polyregular function.
+\end{fact}
+
+\begin{fact}
+    One can do anything on small values.
+\end{fact}
+
+Let us now conclude with a nice characterisation.
+\begin{theorem}
+    If
+    $f$ is star-free $\Rel$-polyregular
+    and $f$ is commutative $\Nat$-polyregular,
+    then $f$ is star-free $\Nat$-polyregular.
+\end{theorem}
+\begin{proof}
+    Idea. 
+    Remark that $f$ is ultimately polynomial, and because $f$ 
+    is commutative, this proves that $f$, after some rank,
+    computes a polynomial. But this proves that $f$ is star-free
+    $\Nat$-polyregular.
+    (maybe we need to do an induction on the degree).
+\end{proof}
+
+\begin{corollary}
+    One can decide if a commutative $\Nat$-polyregular function 
+    is star-free.
+\end{corollary}
 
@@
 
+\begin{center}
+    \begin{tabular}{c|cc}
+        \toprule
+        & \textbf{Commutative} & \textbf{General} \\
+        \midrule
+        \textbf{Commutative}
+        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
+        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
+        \textbf{General} & ? & polyregular 
+    \end{tabular}
+\end{center}
+
 % BACKMATTER

### 2024-01-10 globals/metadata.tex
presque un papier

@@
-\title{A corrected version of Karhumäki’s theorem
-       on commutative polynomial $\Nat$-rational series}
+\title{
+    Star-free commutative $\Nat$-polyregular functions:
+    correcting a result of Karhumäki.
+}
 \author{
@@
         previous result on commutative $\Nat$-rational series,
-        and prove a corrected version of the theorem, that only
-        requires commutativity and a polynomial growth rate.
+        and prove a corrected version of the theorem,
+        working of tdd
         We then discuss how the syntactic criterion could
@@
         is not assumed.
+
+        \begin{itemize}
+            \item Deciding $\Nat$ inside $\Rel$.
+            \item Counter example.
+            \item Characterization of $\Nat$-polyregular.
+            \item Characterization of star free $\Nat$-polyregular.
+        \end{itemize}
     \end{abstract}

### 2024-01-10 pnrs.tex
presque un papier

@@
 %! TeX program = xelatex
+%! lang = en-GB
 \documentclass[a4paper,11pt]{article}
@@
 
+\section{TODOs}
+
+\begin{itemize}
+    \item Use $\Nat$-polyregular functions all the time, and just 
+        in the introduction talk about $\Nat$-rational series.
+    \item Prove that $\Nat$-rational series of polynomial 
+        growth are $\Nat$-polyregular functions, maybe using
+        the results about ambiguity in automata?
+        Theorem 5.22
+    \item Define commutativity by a map 
+        $\commute \colon \Sigma^* \to \Nat^{\Sigma}$.
+    \item Define "commutative functions" and its symbol.
+    \item Define "computed by" (like, recognised by) for commutative 
+        functions.
+    \item Change \cref{corrected-version:thm} to use directly
+        star-free!
+    \item Clear up the computability problems.
+    \item Polynomial + Npolyregular => star free?
+    \item Polynomial + N rational => N polyregular?
+    \item cite the paper of Quantitative Monadic Second-Order Logic
+        about why it is interesting class.
+\end{itemize}
+
 % MAIN MATTER
@@
 \section{Preliminaries}
+
+\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
+    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
+    be a morphism, $k \in \Nat$, and 
+    $\pi \colon M^k \to \Rel$ be a production function.
+    The \intro{$\Rel$-polyregular function}
+    $\pi^\dagger \colon \Sigma^* \to \Rel$
+    is computed as follows:
+    \begin{equation*}
+        f(w) \defined
+        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
+        \quad .
+    \end{equation*}
+
+    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
+    is called \intro{$\Nat$-polyregular}.
+\end{definition}
+
+\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
+    \label{polynomial-rational-polyreg:fact}
+    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
+    be a function. The following are equivalent:
+    \begin{enumerate}
+        \item $f$ can be computed as a \k{$\Nat$-rational series}
+            of \kl{polynomial growth}.
+        \item $f$ is a \kl{$\Nat$-polyregular function}.
+    \end{enumerate}
+\end{fact}
+
+\begin{definition}[Letter Counting Function]
+    \label{commutative-map:def}
+    Let $\Sigma$ be a finite alphabet. We define
+    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
+    map that counts occurrences of every letter in the input word, namely:
+    \begin{equation*}
+        \commute[w] \defined \seqof{\card[a]{w}}{a \in \Sigma} \quad .
+    \end{equation*}
+\end{definition}
+
+\begin{definition}[Commutative Function]
+    \label{commutative:def}
+    Let $X$ be a set.
+    A function $f \colon \Sigma^* \to X$ is
+    \intro{commutative} whenever
+    for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
+
+    Equivalently, it is \reintro{commutative} whenever
+    there exists a map $g \colon \Nat^\Sigma \to X$
+    such that $g \circ \commute = f$.
+\end{definition}
+
+\begin{definition}[Represented Function]
+    A function $g \colon \Nat^n \to \Rel$
+    is \intro{represented} by a function $f \colon \Sigma^* \to \Rel$
+    if there exists a map $\mu \colon \set{1, \dots, n} \to \Sigma$
+    such that
+    $g \circ (v \mapsto \seqof{v_{\mu(k)}}{1 \leq k \leq n}) \circ \commute = f$.
+\end{definition}
+
+We need, $\ZRat$, $\NRat$, $\ZPoly$, $\NPoly$, $\ZSF$, and $\NSF$.
 
@@
     \end{itemize}
-\end{definition}
-
-\begin{definition}[Commutative rational series]
-    A $\Rel$-rational series
-    $f \colon \Sigma \to \Rel$ is \intro{commutative} whenever $f(uv) = f(vu)$ for
-    all $u,v \in \Sigma^*$.
 \end{definition}
@@
 let us only define $\Rel$-polyregular functions.
-
-\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a production function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        f(w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-\end{definition}
-
-\begin{fact}[see e.g. ...]
-    \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be a function. The following are equivalent:
-    \begin{enumerate}
-        \item $f$ can be computed as a \k{$\Nat$-rational series}
-            of \kl{polynomial growth}.
-        \item $f$ is a \kl{$\Nat$-polyregular function}.
-    \end{enumerate}
-\end{fact}
 
@@
 
-\begin{remark}
+\begin{theorem}
+    \label{nsf-polynomials:thm}
     In the proof, we actually conclude that
@@
     \emph{star-free} $\Nat$-polyregular functions.
-\end{remark}
+\end{theorem}
 
@@
 \begin{lemma}
+    f is commutative $\Nat$ polyregular, then when iterated we obtain
+    polynomials in $\CorrectPoly$.
+\end{lemma}
+\begin{proof}
+    by induction on the number of letters 
+\end{proof}
+
+\begin{lemma}
     \label{decompose-polynomial:lem}
@@
     equivalent partitions of the word, but commutativity?.q
+    Lemma 5.37 phd gaëtan.
 \end{proof}
@@
 
+Let us now prove that the above characterizations of
+commutative
+$\Nat$-polyregular functions can be combined with the recent
+advances in the study of $\Rel$-polyregular functions 
+\cite{LOPEZ23b}
+to prove the following theorem.
+
+\begin{theorem}
+    \label{zsf-npoly-nsf:thm}
+    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
+    be both a \kl{star-free $\Rel$-polyregular function}
+    and a \kl{commutative} \kl{$\Nat$-polyregular function}.
+    Then, $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}.
+    That is, the following equality holds:
+    \begin{equation*}
+        \ZCommut \cap \ZSF \cap \NPoly
+        = \ZCommut \cap \NSF
+        \quad .
+    \end{equation*}
+\end{theorem}
+
+Note that this proves in the commutative case a conjecture that was 
+formulated by Thesis Gaetan, and that is non trivial, because
+
+\begin{conjecture}
+    \label{nsf-zsf:conj}
+    In the general case, the following equation holds:
+    \begin{equation*}
+        \ZSF \cap \NPoly
+        = 
+        \NSF
+        \quad .
+    \end{equation*}
+\end{conjecture}
+
+As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
+procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
+decidable inside $\NPoly$ under the assumption of commutativity.
+
+\begin{corollary}
+    One can decide if a \kl{commutative} \kl{ $\Rel$-polyregular function}
+    can be realized by a \kl{star-free $\Nat$-polyregular function},
+    in which case the conversion is effective.
+\end{corollary}
+
+The key ingredient of this section is the use of a semantic characterization of
+\kl{star-free $\Rel$-polyregular functions} among \kl{$\Rel$-rational series}
+that generalizes the aperiodicity of languages to functions by the means of
+polynomial behaviors.
+
 \begin{definition}[Ultimately polynomial]
+    \label{ultimately-polynomial:def}
+    Let $\Sigma$ be a finite alphabet. 
+    A function $f \colon \Sigma^* \to \Rel$
+    is \intro{ultimately polynomial}
+    whenever there exists $N_0 \in \Nat$ such that
+    for all $n \in \Nat$
+    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
+    \in \Sigma^*$, there exists a polynomial $P \in \Rel[X_1, \dots, X_n]$
+    such that
+    \begin{equation*}
+        f\left(
+            \alpha_0 \prod_{i = 1}^{n} w_i^{X_i} \alpha_i
+        \right)
+        = 
+        P(X_1, \dots, X_n)
+        \quad 
+        \forall X_1, \dots, X_n \geq N_0
+        \quad .
+    \end{equation*}
 \end{definition}
 
-\begin{fact}
-    Any $\Nat$-rational polynomial is a star-free polyregular function.
-\end{fact}
+Let us recall from \cite{LOPEZ23b} that this definition correctly
+generalizes the indicator functions of regular languages.
 
-\begin{fact}
-    One can do anything on small values.
-\end{fact}
-
-Let us now conclude with a nice characterisation.
-\begin{theorem}
-    If
-    $f$ is star-free $\Rel$-polyregular
-    and $f$ is commutative $\Nat$-polyregular,
-    then $f$ is star-free $\Nat$-polyregular.
-\end{theorem}
+\begin{example}
+    A language $L$ is aperiodic if and only if 
+    the indicator function is ultimately polynomial.
+\end{example}
 \begin{proof}
-    Idea. 
-    Remark that $f$ is ultimately polynomial, and because $f$ 
-    is commutative, this proves that $f$, after some rank,
-    computes a polynomial. But this proves that $f$ is star-free
-    $\Nat$-polyregular.
-    (maybe we need to do an induction on the degree).
+    TODO.
 \end{proof}
 
-\begin{corollary}
-    One can decide if a commutative $\Nat$-polyregular function 
-    is star-free.
-\end{corollary}
+\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
+    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
+
+    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
+    only the empty word $\varepsilon$, and because $f$ is a
+    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
+    Note that $f$ is therefore a constant function, equal to a natural
+    number, hence belongs to $\NSF$.
+
+    \textbf{Induction.}
+    Because $f$ is \kl{ultimately polynomial},
+    there exists $N_0 \in \Nat$, 
+    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
+    such that
+    \begin{equation*}
+        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
+        = 
+        P(\seqof{X_a}{a \in \Sigma})
+        \quad
+        \text{ when }
+        \forall a \in \Sigma, X_a \geq N_0
+        \quad .
+    \end{equation*}
+    Furthermore, leveraging \cref{decompose-polynomial:lem},
+    we obtain $\omega \in \Nat$
+    and a family of polynomials 
+    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
+    such that.
+    Because each of those polynomials coincide with $P$ over an infinite
+    subgrid of $\Nat$,
+    we conclude that all polynomials are in fact equal to a single one.
+    we conclude that the polynomials 
+    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.
+
+    Now, we conclude in particular that the function 
+    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
+    hence is computable by a \kl{star-free $\Nat$-polyregular function},
+    because of \cref{nsf-polynomials:thm}.
+    When fixing values of some indeterminate below $N_0$,
+    we obtain a new function $g$, that remains commutative,
+    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
+    By induction hypothesis, this function $g$ is also 
+    \kl{star-free $\Nat$-polyregular}, and we conclude
+    that 
+    $f = P + \sum g$ which remains star free.
+\end{proof}
 

### 2024-01-10 pnrs.tex
acm style

@@
 %! lang = en-GB
-\documentclass[a4paper,11pt]{article}
-
-\usepackage[left=3cm,right=3cm]{geometry}
+\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
 
@@
 
+\makeabstract
 % DOCUMENT
@@
 \maketitle
-\makeabstract
 \acknowledge
@@
     \label{corrected-version:thm}
-    The \kl{$\Nat$-rational polynomials} are exactly
-    the polynomials in $\CorrectPoly$.
-\end{theorem}
-
-\begin{theorem}
-    \label{nsf-polynomials:thm}
-    In the proof, we actually conclude that
-    the \kl{$\Nat$-rational polynomials}
-    are exactly those that can be computed 
-    by \kl{commutative}
-    \emph{star-free} $\Nat$-polyregular functions.
+    Let $P \in \Rel[\vec{X}]$ be a polynomial.
+    The following are equivalent:
+    \begin{enumerate}
+        \item $P \in \CorrectPoly$,
+        \item $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
+        \item $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
+        \item $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
+    \end{enumerate}
 \end{theorem}

### 2024-01-11 pnrs.tex
cleaning up proofs

@@
 
-\begin{theorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
+\begin{faketheorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
     \label{karh:thm}
@@
     coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
-\end{theorem}
+\end{faketheorem}
 
@@
 decidability of \kl{$\Nat$-rational polynomial} inside polynomials with $\Rel$
-coefficients. We will not formalize this statement yet, as \cref{karh:thm}
-does not hold in general.
+coefficients. We will not formalize this statement yet, as \cref{karh:thm} does
+not hold in general.
 
@@
 
-\begin{fact}
-    \label{fact:n-poly-n-poly}
+\begin{remark}
+    \label{inductive-correct-poly:remark}
+    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
+    The following are equivalent:
+    \begin{enumerate}
+        \item $P \in \CorrectPoly$,
+        \item The \kl{maximal monomials} of 
+            $P$ have \kl{positive coefficients},
+            and 
+            for all $n \in \Nat$,
+            $\restr{P}{X = n} \in \CorrectPoly$.
+    \end{enumerate}
+    Furthermore, if $P \in \Rel[]$ is a polynomial without
+    indeterminate, then $P \in \CorrectPoly$ if and only
+    if its \kl{maximal monomial} (the constant term)
+    has a \kl{positive coefficient}.
+\end{remark}
+\begin{proof}
+    By induction on the number of indeterminate,
+    leveraging the fact that if there exists at least one indeterminate
+    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
+    and partial function $\nu \colon \vec{Y} \to \Nat$,
+    we have 
+    \begin{equation*}
+        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
+        \quad
+        ,
+    \end{equation*}
+    where $\nu[X = n]$ is the partial function from
+    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
+    and maps $X$ to $n$.
+\end{proof}
+
+\begin{example}
+    \label{n-poly-n-poly:example}
     Let $P \in \Nat[\vec{X}]$. Then, $P$
     is a \kl{$\Nat$-rational polynomial}.
-\end{fact}
+\end{example}
+\begin{proof}
+    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
+    be a partial function.
+    Notice that $\restr{P}{\nu}$ continues to be
+    a polynomial with \kl{positive coefficients}, and in particular
+    with 
+    \kl{maximal monomials} having \kl{positive coefficients}.
+\end{proof}
 
@@
 \begin{lemma}
+    \label{n-rat-correct:lem}
     The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
@@
     \kl{maximal monomial} of $P$, with $\alpha > 0$.
-    Then, there exists $K \geq 0$
+    Then, there exists a computable $K \geq 0$
     such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
@@
 $\CorrectPoly$, because we have a restriction on the range where the variable
-$Y$ can be taken.
+$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
+induction on the number $d$ of indeterminate, and the degrees of the
+\kl{maximal monomials}. To make the induction formal, let us define
+$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
+$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
+$\Nat^{\vec{X}}$ with the product ordering, which is a
+\kl{well-quasi-ordering}.
+
+\AP To a polynomial $P$, we assign the set $\MaximalMonomials(P)$ of its
+\kl{maximal monomials}, belonging to $\Pfin(\Monomials)$, which remains a
+\kl{well-quasi-ordering} when endowed with the \intro{Hoare ordering} of
+subsets, that is, $S_1 \hoareleq S_2$ whenever for all \kl{monomial} $M_1 \in
+S_1$, there exists a monomial $M_2 \in S_2$, such that $M_1$ \kl{divides}
+$M_2$.
 
@@
     \label{lem:correct-to-n-rat}
-    Polynomials in $\CorrectPoly$ are
-    \kl{$\Nat$-rational polynomials}.
+    Let $P$ be a polynomial in $\CorrectPoly$.
+    Then, 
+    $P$ is
+    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
 \end{lemma}
-\begin{proof}[Proof Sketch]
-    We prove by induction on the number $d$ of variables
-    and the ``higman ordering of polynomials" (we look at the multiset
-    of tuples of things ...) that for every selection
-    $K_1, \dots, K_n$ of lower bounds for the different variables
-    (that can be $0$),
-    functions that are in $\CorrectPoly$ \emph{over tuples of size
-    at least nanania} can be computed by a $\Nat$-rational series
-    that outputs $0$ outside the domain and the correct value
-    inside.
-
-    For $d = 0$,
-    the result is true because
-    a constant function is a \kl{$\Nat$-rational polynomial}
-    if and only if it is \kl{non-negative}.
-
-    Let us now assume that $d > 0$, and in particular
-    there is a variable $Y$.
-    Consider $P \in \CorrectPoly[\vec{X} Y]$.
-    If $Y$ appears at least one in $P$, then it must appear
-    in one of the \kl{maximal monomials} of $P$. Otherwise,
-    $P \in \CorrectPoly[\vec{X}]$ and we conclude using the
-    induction hypothesis.
-    Let $\alpha Y^p M$ be a \kl{maximal monomial} in $P$.
+\begin{proof}
+    We prove the result by induction on $\MaximalMonomials(P)$.
+
+    \textbf{Base case:} If the \kl{maximal monomial} of $P$ is a constant term,
+    then because $P \in \CorrectPoly$, $P = n \in \Nat$, which is
+    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
+
+    \textbf{Induction:}
+    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
+    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
+    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
     Using \cref{lem:delta-cool},
-    we know that there exists $K$ such that
-    $P - \alpha (Y - K)^p M$ almost belongs to $\CorrectPoly$,
-    and has a strictly smaller "degree multiset".
+    we know that there exists a computable $K$ such that
+    the polynomial
+    $\Diff{K} \defined P - \alpha (Y - K)^p M$
+    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
+    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
+    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
+    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
+    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
+    \hoarele \MaximalMonomials(P)$.
     By induction hypothesis,
     we conclude that
-    the following function is a $\Nat$-rational series
-    (beware that we have to consider the lower bounds to apply
-    the induction hypothesis):
-    \begin{equation*}
-        A \defined
+    $\Diff{K}(Y - K)$ is \kl{represented}
+    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
+    \begin{equation*}
+        f_1(Y, \vec{X}) \defined
         \begin{cases}
-            0 & \text{ if some variable is bounded by its } K_i \\
-            0 & \text{ otherwise if  } Y \leq K \\
-            P - \alpha(Y - K)^pM & \text{ otherwise }
+            0 & \text{ if } Y \leq K \\
+            \Diff{K} & \text{ otherwise }
         \end{cases}
@@
     \begin{equation*}
-        B \defined
+        f_2(Y, \vec{X}) \defined
         \begin{cases}
-            0 & \text{ if some variable is bounded by its } K_i \\
-            0 & \text{ otherwise if } Y \leq K \\
+            0 & \text{ if } Y \leq K \\
             \alpha(Y - K)^pM & \text{ otherwise }
         \end{cases}
-    \end{equation*}
-
-    Remark that by fixing some of the variables
-    to values 
-    below the bounds $K_i$ (and potentially $K$ for $Y$)
-    we obtain a \emph{finite} number of polynomials with strictly fewer
-    indeterminate.
-    As a consequence, the following function is computable
-    as a $\Nat$-rational series:
-    \begin{equation*}
-        C \defined
+        \quad .
+    \end{equation*}
+
+    Because $P \in \CorrectPoly$, for every $n \leq K$, $\restr{P}{Y = n} \in
+    \CorrectPoly$ (\cref{inductive-correct-poly:remark}).
+    Furthermore,
+    $\MaximalMonomials(\restr{P}{Y = n})
+    \hoarele \MaximalMonomials(P)$. Hence, by induction hypothesis,
+    the following function is \kl{represented} by a
+    \kl{star-free $\Nat$-rational series}:
+    \begin{equation*}
+        f_3(Y, \vec{X}) \defined
         \begin{cases}
-            P & \text{ if some variable is bounded by its } K_i \\
-            P & \text{ otherwise if } Y \leq K \\
+            P(Y, \vec{X}) & \text{ if } Y \leq K \\
             0 & \text{ otherwise }
@@
     \end{equation*}
-
-    We conclude because
-    $P = A + B + C$, all of which are $\Nat$-rational series.
-\end{proof}
+    We conclude by remarking that 
+    $P = f_1 + f_2 + f_3$.
+\end{proof}
+
+We are now ready to state the corrected and generalised version of
+\cref{karh:thm}.
 
@@
     \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
-        \item $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-        \item $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
+        \item \label{corrected-1:item} $P \in \CorrectPoly$,
+        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
+        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
+        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
     \end{enumerate}
 \end{theorem}
+\begin{proof}
+    The implications 
+    \cref{corrected-4:item} $\implies$
+    \cref{corrected-3:item} $\implies$
+    \cref{corrected-2:item} are obvious.
+    \cref{lem:correct-to-n-rat} proves
+    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
+    while \cref{n-rat-correct:lem}
+    proves 
+    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
+\end{proof}
 
@@
 
+Before proving \cref{decidable-n-poly:thm}, let us first examine the
+effectiveness of \cref{corrected-version:thm}, namely, provide an algorithm to
+decide whether a polynomial $P$ belongs to $\CorrectPoly$. The idea will be to
+compute a bound to values taken by the valuations $\nu$ appearing in the
+definition of $\CorrectPoly$. As a \emph{mise en jambes}, let us consider
+separately the case of one indeterminate.
+
+\begin{lemma}
+    \label{decide-correct-univariate:lemma}
+    Let $P \in \Rel[X]$, there exists a computable $N \in \Nat$,
+    such that the following are equivalent:
+    \begin{enumerate}
+        \item \label{decide-correct-univariate-1:item}
+            $P \in \CorrectPoly$,
+        \item \label{decide-correct-univariate-2:item}
+            The \kl{maximal monomial}
+            of $P$ has a \kl{positive coefficient},
+            and $P(n) \in \Nat$
+            for all $0 \leq n \leq N_0$.
+    \end{enumerate}
+\end{lemma}
+\begin{proof}
+    Let us first remark that the implication 
+    \cref{decide-correct-univariate-1:item}
+    $\implies$ \cref{decide-correct-univariate-2:item} is trivial for every
+    $N \in \Nat$.
+    For the converse implication,
+    let $P \in \Rel[X]$.
+    If $P$ is a constant polynomial, then $P \in \CorrectPoly$
+    if and only if $P(0) \in \Nat$, and $N \defined 0$
+    suffices.
+    Otherwise, $P$ is non-constant, and let $n \geq 1$ be the \kl{degree} of $P$.
+    Let $P_1$ and $\alpha > 0$ be such that
+    $P \defined P_1 + \alpha X^n$. 
+    To prove that $P \in \CorrectPoly$, it suffices to show that
+    $P(k) \in \Nat$ for all $k \in \Nat$.
+    Let us define $\beta$ as the
+    maximal absolute value of coefficients appearing in $P_1$, which is
+    assumed to be non-zero,
+    and $N_0 \defined \beta + 1 \geq 2$.
+    Remark that for $X \geq N_0$,
+    \begin{align*}
+        P(X) &\geq X^n - \beta \sum_{i = 0}^{n-1} X^i \\
+             &= X^n - \beta \frac{X^n - 1}{X - 1} \\
+             &= \frac{X^{n+1} - X^n - \beta X^n + \beta}{X - 1} \\
+             &= \frac{X^n(X - (\beta + 1)) + \beta}{X - 1} \\
+             &\geq 0
+    \end{align*}
+    This proves that $P(k) \in \Nat$ for all $k \geq N_0$,
+    and because $P$ satisfies \cref{decide-correct-univariate-2:item},
+    $P(k) \in \Nat$ for all $1 \leq k \leq N_0$. We have proven 
+    that $P \in \CorrectPoly$.
+\end{proof}
+
+In the case of multiple indeterminate, we leverage
+\cref{lem:delta-cool} to perform an induction on the degree
+of the polynomials and the number of variables of the polynomial.
+
 \begin{lemma}
     \label{decide-correct-poly:lem}
-    One can decide if a polynomial is in $\CorrectPoly$.
+    One can decide if a polynomial belongs to $\CorrectPoly$.
 \end{lemma}
 \begin{proof}
-    It is easy to check that maximal monomials have positive coefficients.
-    Assuming that it is the case, remark that 
-    every partial valuation $\nu$
-    such that $\restr{P}{\nu}$ has a negative maximal coefficient
-    must have values that are bounded by the sum of absolute values of
-    coefficients appearing in $P$.
-    In particular, one can check all possible choices for $\nu$
-    within these bounds.
+    We prove the result by induction on the number of indeterminate.
+
+    \textbf{Base Case:} If $P$ has zero indeterminate, i.e., is a constant
+    function, then to decide whether $P \in \CorrectPoly$ it suffices to check
+    that $P(0) \in \Nat$.
+
+    \textbf{Induction:} Let $P \in \Rel[Y, \vec{X}]$, we start by checking
+    whether \kl{maximal monomials} of $P$ have \kl{positive coefficients}.
+    Without loss of generality, assume that $Y$ divides some \kl{maximal
+    monomial} $\alpha Y^p M$ with $p \geq 1$. By applying
+    \cref{lem:delta-cool}, we can compute a bound $K \in \Nat$ such that
+    $\Diff{K} \defined P - \alpha (Y - K)^p M$ has \kl{positive coefficients}
+    for its \kl{maximal monomials} for all partial valuations $\nu \colon
+    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
+    particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to
+    $\CorrectPoly$ for all $n > K$. By induction hypothesis, one can
+    decide whether $\restr{P}{X = n}$ belongs to $\CorrectPoly$
+    for all $0 \leq n \leq K$, which is equivalent 
+    to deciding whether
+    $P \in \CorrectPoly$ thanks to \cref{inductive-correct-poly:remark}.
 \end{proof}
@@
 commutative case.
-
-\begin{lemma}
-    f is commutative $\Nat$ polyregular, then when iterated we obtain
-    polynomials in $\CorrectPoly$.
-\end{lemma}
-\begin{proof}
-    by induction on the number of letters 
-\end{proof}
 

### 2024-01-11 pnrs.tex
avancées

@@
 %! TeX program = xelatex
-%! lang = en-GB
+%! lang = en-US
 \documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
@@
     \item Define commutativity by a map 
-        $\commute \colon \Sigma^* \to \Nat^{\Sigma}$.
+        $\commute \colon \Sigma^* \topartial \Nat^{\Sigma}$.
     \item Define "commutative functions" and its symbol.
@@
     \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
+    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \topartial \Nat$
     be a function. The following are equivalent:
@@
     Let $\Sigma$ be a finite alphabet. We define
-    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
+    $\commute \colon \Sigma^* \topartial \Nat^\Sigma$ for the
     map that counts occurrences of every letter in the input word, namely:
@@
     There exists a \kl{commutative}
-    $\Nat$-rational series $f \colon \Sigma \to \Nat$
+    $\Nat$-rational series $f \colon \Sigma \topartial \Nat$
     such that $P(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
@@
     $P$ is 
-    such that, for every partial function $\nu \colon \vec{X} \to \Nat$,
+    such that, for every partial function $\nu \colon \vec{X} \topartial \Nat$,
     every \kl{maximal monomial} of
@@
     $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
-    and partial function $\nu \colon \vec{Y} \to \Nat$,
+    and partial function $\nu \colon \vec{Y} \topartial \Nat$,
     we have 
@@
 \begin{proof}
-    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
+    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \topartial \Nat$
     be a partial function.
@@
 
+The following \cref{n-poly-combinatorics:lem} is a slight refinement over the
+classical combinatorial analysis of \kl{$\Rel$-polyregular functions}
+\cite[Lemma 4.16]{LOPEZ23b} and \kl{$\Nat$-polyregular functions} that takes
+into account \emph{all} \kl{maximal monomials} instead of a specific one
+\cite[Lemma 5.37]{gaetanphd}.
+
 \begin{lemma}
@@
     whose \kl{maximal monomials} have \kl{positive coefficients},
-    and such that:
-    \begin{equation*}
-        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
+    and such that for all $X_1, \dots, X_p \geq \omega$:
+    \begin{equation*}
+        f\left(
+            \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)
+        \right)
         = P(X_1, \dots, X_p) \quad .
@@
 \end{lemma}
-\begin{proof}[Proof Sketch]
-    Let $\omega$ be an idempotent power for the finite monoid $M$.
-    We define an equivalence between partitions of a word 
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$
-    as follows: two partitions into $k$ words in $\Sigma^*$ are equivalent
-    when normalizing them with the rules
-    $u_i^{\omega} u_i^{\omega} \to u_i^{\omega}$ for all $1 \leq i \leq p$.
-    This equivalence relation has finite index, and all partitions in the same
-    equivalence class yield the same value through $\pi$, because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid $M$.
-
-    Now, let us remark that the number of partitions
-    in a certain equivalence class is proportional to
-    the number of nanania, that is, is of the form
-    $(X_i - K_i)^{n_i}$ for some $K$. We conclude that
-    maximal monomials indeed have positive coefficients.
-    We even conclude that
-    the polynomial is in $\Nat[\vec{X}]$
-    if we shift the variables with a large enough $K$. 
+\begin{proof}
+    Let $\omega$ be an idempotent power for the finite monoid $M$,
+    and
+    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
+    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
+    $f = \pi^\dagger$. That is, 
+    $f(w)$ is the sum over all factorizations of $w$
+    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
+    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
+
+    Let us define an equivalence relation $\equiv$ over words of
+    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
+    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
+    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
+    repetitions of the factors $u_i^{\omega}$ in the words.
+    We lift the equivalence relation $\equiv$ to $k$-tuples of
+    words by pointwise application.
+
+    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then
+    $\pi(\mu(v_1), \dots, \mu(v_k)) = \pi(\mu(w_1), \dots, \mu(w_k))$ because
+    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid
+    $M$. In particular, $f(w)$ can be computed as the sum over equivalence
+    classes for $\equiv$, of the value of $\pi$ on one representative,
+    which belongs to $\Nat$.
+
+    Therefore, it suffices to prove that the number of partitions of $w$ in a
+    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
+    with \kl{positive coefficients} for its \kl{maximal monomials}.
+
+    Notice that to produce all elements in a given equivalence class
+    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
+    back to the normalized version
+    of the equivalence class. For all elements in an equivalence class,
+    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
+    and therefore the number of possible choices
+    is either constant, or 
+    is a product of binomials of the form $\binom{X - l}{s}$
+    which has the desired shape.
 \end{proof}
@@
     \label{n-rat-correct:lem}
-    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
+    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
+    by a \kl{$\Nat$-rational series}. Then,
+    $P \in \CorrectPoly$.
 \end{lemma}
-\begin{proof}[Proof Idea]
-    Let $P$ be a \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma$  be the finite alphabet such that
-    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
-    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
-    $\Nat$-rational series computing $P$.
-    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
-    it is a $\Nat$-polyregular function 
+\begin{proof}
+    Let $f$ be the \kl{commutative}
+    \kl{$\Nat$-rational series}
+    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
+    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
+    it is in fact a \kl{$\Nat$-polyregular function}
     (\cref{polynomial-rational-polyreg:fact}).
-
     Using \cref{n-poly-combinatorics:lem},
-    we conclude that maximal monomials of $P$ have
-    positive coefficients. Furthermore, when fixing variables
-    in $P$, we are computing a new \kl{$\Nat$-polyregular function},
-    hence we also conclude that the maximal monomials have
-    positive coefficients.
-    We have proven that $P \in \CorrectPoly$.
-\end{proof}
-
+    there exists a number $\omega \in \Nat$
+    and polynomial $Q$ with \kl{positive coefficients} for its
+    \kl{maximal monomials} such that
+    for all $X_1, \dots, X_p \geq \omega$:
+    \begin{equation*}
+        f\left(
+            \prod_{i = 1}^p (a_i)^{\omega \times X_i}
+        \right)
+        = Q(X_1, \dots, X_p)
+        = P(\omega X_1, \dots, \omega X_p) 
+        \quad .
+    \end{equation*}
+    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
+    as polynomials,
+    and in particular that
+    the \kl{maximal monomials} of 
+    $P$ have \kl{positive coefficients}.
+
+    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
+    the polynomial $\restr{P}{\nu}$ continues to be represented
+    by a \kl{$\Nat$-polyregular function}, namely
+    $f$ partially applied to a word. As a consequence,
+    the \kl{maximal monomials} of
+    $\restr{P}{\nu}$ also have \kl{positive coefficients}, and 
+    we have proven that $P \in \CorrectPoly$.
+\end{proof}
+
+\textbf{TODO: rewrite this}
 The core of the upcoming proof of \cref{lem:correct-to-n-rat} relies on the
 following observation: if a polynomial $P$ is written $X^p + Q$, where all
-terms of $Q$ have degree less than $p$, then one can find some large enough $K$
-so that $X^p - (X - K)^p + Q$ also has a positive maximal coefficient.
-It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence
-the maximal coefficient can be chosen arbitrarily large by increasing
-$K$.
-This
-trick will work similarly in multivariate polynomials, and the overall
-proof scheme works as follows: for a bounded number of possible
-values of $X$, we can fix the variable, and have one less item to work on.
-For large enough values of $X$ (typically greater than some $K$), we can guarantee that
-$X^p - (X - K)^p + Q$ has positive maximal coefficients, hence
-can be computed by a $\Nat$-rational series, while
-$(X - K)^p$ is clearly computable by a $\Nat$-rational series,
-hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
-as the sum of two $\Nat$-rational series, and we have
-concluded that our original polynomial was a
-\kl{$\Nat$-rational polynomial}. There are a number of problems in generalizing
-this proof scheme:
-\begin{itemize}
-    \item We will have to guarantee that every polynomial considered
-        is non-negative
-    \item We do not only need to check that the \kl{maximal monomials}
-        have \kl{positive coefficient}, but that it is the case
-        when fixing arbitrarily many variables! (this
-        covers the first point too)
-\end{itemize}
+terms of $Q$ have \kl{degree} less than $p$, then one can find some large
+enough $K \in \Nat$ so that $X^p - (X - K)^p + Q$ also has a positive maximal
+coefficient. It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence the
+new \kl{maximal monomial} can be made to have arbitrarily large \kl{coefficient}
+by increasing $K$.
+As a consequence, for values $n \leq K$ of $X$, one can compute $\restr{P}{X =
+n}$ which remains in $\CorrectPoly$, and apply an induction hypothesis.
+Furthermore, for values $n > K$ of $X$, one has the guarantee that $X^p - (X -
+K)^p + Q$ has positive maximal coefficients, hence can be computed by a
+$\Nat$-rational series, while $(X - K)^p$ is clearly computable by a
+$\Nat$-rational series, hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
+as the sum of two $\Nat$-rational series, and we have concluded that our
+original polynomial was a \kl{$\Nat$-rational polynomial}.
+
+When multiple indeterminate are involved, a bit of care is needed, and
+this is the technical content of \cref{diff-one-monomial:lem,lem:delta-cool}.
+
+\begin{definition}
+    Let $\vec{X}$ be a tuple of indeterminate, let $Y$
+    be a fresh indeterminate,
+    and 
+    let $P \in \Rel[\vec{X},Y]$ be a polynomial.
+    For every \kl{maximal monomial}
+    $\alpha Y^p M$ of $P$ such that $M \in \Rel[\vec{X}]$, and every $K \in \Nat$,
+    we define
+    \begin{equation*}
+        \Diff{K}{\alpha Y^p M}{P} \defined 
+        P - \alpha (Y - K)^p M \quad .
+    \end{equation*}
+\end{definition}
 
 \begin{lemma}
+    \label{diff-one-monomial:lem}
     Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
-    $\vec{p} \colon \vec{X}Y \to \Nat$,
-    and $M \defined \times \prod_{X \in \vec{X}} X^{p(X)}$.
+    and $P = Y^p M$ where $M$ is a \kl{monomial} over $\vec{X}$.
     Then, for all $K \geq 0$,
-    $\Delta_{Y,K} \defined Y^p M - (Y - K)^p M$ has \kl{maximal monomials} with
-    \kl{positive coefficients} of weight at least $K$, regardless of
-    the variables fixed, as long as the (potential)
-    value of $Y$ is greater
-    than $K$.
+    for all $\nu \colon \vec{X}Y \topartial \Nat$
+    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
+    $\restr{\Diff{K}{Y^pM}{Y^pM}}{\nu}$
+    has \kl{maximal monomials} with
+    coefficients at least $K$, or is the constant polynomial equal to zero.
 \end{lemma}
 \begin{proof}
+    Let us write $\Delta_K$ as a shorthand for
+    $\Diff{K}{Y^p M}{Y^p M}$.
     Let us first remark that
-    \begin{equation*}
-        \Delta_{Y,K}
+    there exists $Q \in \Rel[Y]$ of degree at most $p-2$
+    such that:
+    \begin{equation*}
+        \Delta_K
         = KY^{p-1} M + Q M
@@
     \end{equation*}
-    In particular, there is a unique \kl{maximal monomial} for $\Delta_{Y,K}$,
+    In particular, there is a unique \kl{maximal monomial} for $\Delta_{K}$,
     which is precisely $K Y^{p-1} M$, because all monomials
-    in $Q M$ divide (strictly) $Y^{p-1} M$.
-    Let us fix a variable,
-    $X$ to any (non-zero) value $p > 0$. Then
-    The \kl{maximal monomial} of $\restr{\Delta_{Y,K}}{X = p}$
-    is exactly $\restr{KY^{p-1}M}{X = p}$, which remains
-    with a \kl{positive coefficient}, that is proportional to $K$.
-    Let us fix $Y$ to some value $K + l$, with $l > 0$,
-    then
-    $\restr{\Delta_{Y,K}}{Y = K + l}
-    = (K + l)^p M - l^p M
-    = ((K + l) - l) (\sum_{i = 1}^{p-1} (K + l)^i l^{p-i}) M
-    = K \alpha M
-    $
-    which remains a monomial with a \kl{positive coefficient},
-    that is proportional to $K$.
+    in $Q M$ \kl{strictly divide} $Y^{p-1} M$.
+
+    Let $\nu \colon \vec{X}Y \topartial \Nat$ be a partial valuation such that
+    $\restr{\Delta_K}{\nu} \neq 0$. If $\nu(Y)$ is undefined, then the
+    \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$ is exactly $K Y^{p-1}
+    \restr{M}{\nu}$, and $\restr{M}{\nu} \neq 0$. Therefore, the unique
+    \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$ has a \kl{positive
+    coefficient} that is at least $K$.
+
+    Otherwise, if $\nu(Y) > K$, then $\nu(Y) = K + l$ for some $l \in \Nat$.
+    Let us write $\mu$ for the restriction of $\nu$ to indeterminates of $\vec{X}$.
+    We can compute 
+    \begin{align*}
+        \restr{\Delta_K}{\nu}
+        &= (K + l)^p \restr{M}{\mu} - l^p \restr{M}{\mu}\\
+        &= ((K + l)^p - l^p) \restr{M}{\mu} \\
+        &= (K + l - l)\left(\sum_{i = 0}^{p-1} (K+l)^i l^{p-i}\right) 
+        \restr{M}{\mu} \\
+        &= K \alpha \restr{M}{\mu} 
+        \quad .
+    \end{align*}
+    Which proves that the \kl{maximal monomial} of
+    $\restr{\Delta_K}{\nu}$ has a coefficient
+    that is at least $K$.
 \end{proof}
@@
     Then, there exists a computable $K \geq 0$
-    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
-    has \kl{positive coefficients} for \kl{maximal monomials}
-    in any partial evaluation that satisfies $Y > K$ (if it
-    gives a value to $Y$).
+    such that
+    for all $\nu \colon \vec{X}Y \topartial \Nat$
+    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
+    the \kl{maximal monomials} of 
+    $\restr{\Diff{K}{Y^pM}{P}}{\nu}$
+    have \kl{positive coefficients}.
 \end{lemma}
@@
     Let $N$ be greater than the sum of the absolute values of all coefficients
-    appearing in $P$, and let $K = 2 N$.
-
-    Let us remark that
-    \begin{equation*}
-        \Delta_K = \underbrace{(P - \alpha Y^p M)}_{\text{ left hand side }}
-        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{\text{ right hand side }}
+    appearing in $P$, and let $K \defined N + 1$.
+    Let us write $\Delta_K$
+    as a shorthand for
+    $\Diff{K}{\alpha Y^p M}{P}$, and decompose $\Delta_K$ as follows:
+    \begin{equation*}
+        \Delta_K
+        = \underbrace{(P - \alpha Y^p M)}_{Q_1}
+        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{Q_2}
         \quad .
     \end{equation*}
-    an that for every partial evaluation $\nu$ that sends $Y$ to a value
-    greater than $K$,
-    we have that $\restr{(Y^p M - (Y - K)^p M)}{\nu}$ has
-    \kl{maximal monomials} with \kl{positive coefficients}
-    proportional to $K$ (note that here ``proportional"
-    implies that they are greater than $K$, because we multiply by positive
-    integers).
-
-    To conclude, it is just a game of describing how
-    \kl{maximal monomials} of $\restr{\Delta_Y}{\nu}$
-    can appear. If the \kl{maximal monomial} is obtained
+
+    Applying \cref{diff-one-monomial:lem}, we know that for every partial
+    evaluation $\nu \colon \vec{X}Y \topartial \Nat$ such that $\nu(Y) > K$ or
+    $\nu(Y)$ is undefined, the \kl{maximal monomials} of $\restr{Q_2}{\nu}$
+    have \kl{positive coefficients}, and that these coefficients are either $0$
+    (the polynomial is constant equal to zero) or at least $K$.
+
+    To conclude, it is just a matter of describing how \kl{maximal monomials}
+    of $\restr{\Delta_K}{\nu}$ can appear. Let us proceed by case analysis. If
+    $\restr{Q_2}{\nu} = 0$, then, $\restr{\alpha Y^p M}{\nu} = 0$, and
+    $\restr{\Delta_K}{\nu} = \restr{P}{\nu}$, in which case we conclude because
+    $P \in \CorrectPoly$. Otherwise, we know that $\restr{Q_2}{\nu} \neq 0$,
+    and that its (unique) \kl{maximal monomial} has a \kl{positive coefficient}
+    that is at least $K$.
+    Let us now consider a \kl{maximal monomial} $M_0$
+    of $\restr{P}{\nu}$. This \kl{monomial} is obtained
+    by summing \kl{monomials} from $\restr{Q_1}{\nu}$ and $\restr{Q_2}{\nu}$.
+    Note that, by construction
+
+    Because $\alpha Y^p M$ is a \kl{maximal monomial} of
+    $P$, and since $K$ is greater than the sum of
+    the absolute values of all coefficients appearing in $P$,
+    we conclude that
+
+    If a \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$
+    is obtained
+    by summing a of $\restr{Q_1}{\nu}$
     by summing an element of the left hand side with an element
@@
     $\Diff{K} \defined P - \alpha (Y - K)^p M$
-    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
+    satisfies that for all partial valuations $\nu \colon Y\vec{X} \topartial \Nat$
     such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
@@
     It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \to \Nat$.
+    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
     For the converse inclusion, let us consider $P(X,Y)$
@@
     for its \kl{maximal monomials} for all partial valuations $\nu \colon
-    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
+    Y\vec{X} \topartial \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
     particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to

### 2024-01-12 pnrs.tex
correcting some mistakes

@@
 
-When multiple indeterminate are involved, a bit of care is needed, and
-this is the technical content of \cref{diff-one-monomial:lem,lem:delta-cool}.
+When multiple indeterminate are involved, a bit of care is needed. The key
+ingredient is to notice that polynomials in $\CorrectPoly$ are actually stable
+under a \emph{discrete} analogue of the usual differentiation operator.
+Intuitively, polynomials $P$ in $\CorrectPoly$ have \kl{positive} \kl{maximal
+monomials}, and so does $\partial P$, $\partial^2 P$ etc.
 
 \begin{definition}
-    Let $\vec{X}$ be a tuple of indeterminate, let $Y$
-    be a fresh indeterminate,
+    \label{discrete-derivative:def}
+    Let $\vec{X}$ be a tuple of indeterminate,
+    $K \in \Nat$,
     and 
-    let $P \in \Rel[\vec{X},Y]$ be a polynomial.
-    For every \kl{maximal monomial}
-    $\alpha Y^p M$ of $P$ such that $M \in \Rel[\vec{X}]$, and every $K \in \Nat$,
-    we define
-    \begin{equation*}
-        \Diff{K}{\alpha Y^p M}{P} \defined 
-        P - \alpha (Y - K)^p M \quad .
-    \end{equation*}
+    $P \in \Rel[\vec{X}]$ be a polynomial.
+    \begin{equation*}
+        \Diff{K}{P} \defined 
+        P(\vec{X}) - P(\vec{X} - K) \quad .
+    \end{equation*}
+    Where $P(\vec{X} - K)$ is the polynomial function
+    obtained by pre-composing $P$ with the map 
+    $\vec{X} \mapsto \seqof{X_i - K}{X_i \in \vec{X}}$.
 \end{definition}
 
-\begin{lemma}
-    \label{diff-one-monomial:lem}
-    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
-    and $P = Y^p M$ where $M$ is a \kl{monomial} over $\vec{X}$.
-    Then, for all $K \geq 0$,
-    for all $\nu \colon \vec{X}Y \topartial \Nat$
-    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
-    $\restr{\Diff{K}{Y^pM}{Y^pM}}{\nu}$
-    has \kl{maximal monomials} with
-    coefficients at least $K$, or is the constant polynomial equal to zero.
-\end{lemma}
-\begin{proof}
-    Let us write $\Delta_K$ as a shorthand for
-    $\Diff{K}{Y^p M}{Y^p M}$.
-    Let us first remark that
-    there exists $Q \in \Rel[Y]$ of degree at most $p-2$
-    such that:
-    \begin{equation*}
-        \Delta_K
-        = KY^{p-1} M + Q M
-        \quad .
-    \end{equation*}
-    In particular, there is a unique \kl{maximal monomial} for $\Delta_{K}$,
-    which is precisely $K Y^{p-1} M$, because all monomials
-    in $Q M$ \kl{strictly divide} $Y^{p-1} M$.
-
-    Let $\nu \colon \vec{X}Y \topartial \Nat$ be a partial valuation such that
-    $\restr{\Delta_K}{\nu} \neq 0$. If $\nu(Y)$ is undefined, then the
-    \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$ is exactly $K Y^{p-1}
-    \restr{M}{\nu}$, and $\restr{M}{\nu} \neq 0$. Therefore, the unique
-    \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$ has a \kl{positive
-    coefficient} that is at least $K$.
-
-    Otherwise, if $\nu(Y) > K$, then $\nu(Y) = K + l$ for some $l \in \Nat$.
-    Let us write $\mu$ for the restriction of $\nu$ to indeterminates of $\vec{X}$.
-    We can compute 
-    \begin{align*}
-        \restr{\Delta_K}{\nu}
-        &= (K + l)^p \restr{M}{\mu} - l^p \restr{M}{\mu}\\
-        &= ((K + l)^p - l^p) \restr{M}{\mu} \\
-        &= (K + l - l)\left(\sum_{i = 0}^{p-1} (K+l)^i l^{p-i}\right) 
-        \restr{M}{\mu} \\
-        &= K \alpha \restr{M}{\mu} 
-        \quad .
-    \end{align*}
-    Which proves that the \kl{maximal monomial} of
-    $\restr{\Delta_K}{\nu}$ has a coefficient
-    that is at least $K$.
-\end{proof}
-
-We can refine the above lemma to use it in the case where
-we do not have a single
-monomial.
-
-\begin{lemma}
-    \label{lem:delta-cool}
-    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
-    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
-    \kl{maximal monomial} of $P$, with $\alpha > 0$.
-    Then, there exists a computable $K \geq 0$
-    such that
-    for all $\nu \colon \vec{X}Y \topartial \Nat$
-    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
-    the \kl{maximal monomials} of 
-    $\restr{\Diff{K}{Y^pM}{P}}{\nu}$
-    have \kl{positive coefficients}.
-\end{lemma}
-\begin{proof}
-    Let $N$ be greater than the sum of the absolute values of all coefficients
-    appearing in $P$, and let $K \defined N + 1$.
-    Let us write $\Delta_K$
-    as a shorthand for
-    $\Diff{K}{\alpha Y^p M}{P}$, and decompose $\Delta_K$ as follows:
-    \begin{equation*}
-        \Delta_K
-        = \underbrace{(P - \alpha Y^p M)}_{Q_1}
-        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{Q_2}
-        \quad .
-    \end{equation*}
-
-    Applying \cref{diff-one-monomial:lem}, we know that for every partial
-    evaluation $\nu \colon \vec{X}Y \topartial \Nat$ such that $\nu(Y) > K$ or
-    $\nu(Y)$ is undefined, the \kl{maximal monomials} of $\restr{Q_2}{\nu}$
-    have \kl{positive coefficients}, and that these coefficients are either $0$
-    (the polynomial is constant equal to zero) or at least $K$.
-
-    To conclude, it is just a matter of describing how \kl{maximal monomials}
-    of $\restr{\Delta_K}{\nu}$ can appear. Let us proceed by case analysis. If
-    $\restr{Q_2}{\nu} = 0$, then, $\restr{\alpha Y^p M}{\nu} = 0$, and
-    $\restr{\Delta_K}{\nu} = \restr{P}{\nu}$, in which case we conclude because
-    $P \in \CorrectPoly$. Otherwise, we know that $\restr{Q_2}{\nu} \neq 0$,
-    and that its (unique) \kl{maximal monomial} has a \kl{positive coefficient}
-    that is at least $K$.
-    Let us now consider a \kl{maximal monomial} $M_0$
-    of $\restr{P}{\nu}$. This \kl{monomial} is obtained
-    by summing \kl{monomials} from $\restr{Q_1}{\nu}$ and $\restr{Q_2}{\nu}$.
-    Note that, by construction
-
-    Because $\alpha Y^p M$ is a \kl{maximal monomial} of
-    $P$, and since $K$ is greater than the sum of
-    the absolute values of all coefficients appearing in $P$,
-    we conclude that
-
-    If a \kl{maximal monomial} of $\restr{\Delta_K}{\nu}$
-    is obtained
-    by summing a of $\restr{Q_1}{\nu}$
-    by summing an element of the left hand side with an element
-    of the right hand side, it needs to be a \kl{maximal monomial}
-    of the right hand side, and the large value of $K$ guarantees
-    that the sum of coefficients is positive.
-    If it only appears on the right hand side, it is even better.
-    If it only appears on the left hand side, then
-    in particular it means that the monomial does not
-    divide $\restr{Y^p M}{\nu}$, hence has the same value
-    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
-    which is guaranteed to be positive because $P \in \CorrectPoly$.
-
-    In particular, we conclude from this that
-    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
-\end{proof}
-
-Note that we cannot conclude in the above lemma that the result is in
-$\CorrectPoly$, because we have a restriction on the range where the variable
-$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
-induction on the number $d$ of indeterminate, and the degrees of the
-\kl{maximal monomials}. To make the induction formal, let us define
-$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
-$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
-$\Nat^{\vec{X}}$ with the product ordering, which is a
-\kl{well-quasi-ordering}.
+As one would expect, the discrete derivatives are linear operations on
+polynomials, that commutes with the partial application operators
+$\restr{\cdot}{\nu}$.
+
+\begin{fact}
+    \label{discrete-deriv-linear:fact}
+    For all tuple $\vec{X}$ of indeterminate,
+    for all $K \in \Nat$, for all partial function
+    $\nu \colon \vec{X} \topartial \Nat$:
+    \begin{enumerate}
+        \item $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
+            is a linear operator,
+        \item $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
+            is a linear operator,
+        \item $\Diff{K}{ \cdot } \circ \restr{\cdot}{\nu}
+            = \restr{\cdot}{\nu} \circ \Diff{K}{\cdot}$.
+    \end{enumerate}
+\end{fact}
+
+The upcoming \cref{lem:correct-to-n-rat} is proven using an induction on the
+number $d$ of indeterminate, and the degrees of the \kl{maximal monomials}. To
+make the induction formal, let us define $\Monomials[\vec{X}]$ as the
+(infinite) set of all possible \kl{monomials} over $\vec{X}$, ordered by
+\kl{divisibility}. Remark that this set is isomorphic to $\Nat^{\vec{X}}$ with
+the product ordering, which is a \kl{well-quasi-ordering}.
 
@@
 $M_2$.
+
+Let us now prove that the \kl{discrete differentiation} operator
+can be used to perform an induction.
+
+\begin{fact}
+    \label{derivation-simplifies:fact}
+    For all $P \in \Rel[\vec{X}]$ that are non-constant,
+    for all $K \in \Nat$,
+    $\MaximalMonomials(\Diff{K}{P}) \hoarele
+    \MaximalMonomials(P)$.
+\end{fact}
+
+\begin{fact}
+    Let $M,T$ be two \kl{monomials}, such that
+    $T$ \kl{strictly divides} $M$.
+    Then,
+    $T$ \kl{divides} some \kl{maximal monomial}
+    of $\Diff{K}{M}$.
+\end{fact}
+
+\begin{lemma}
+    \label{derivation-stabilises-correct:lem}
+    Let $P \in \CorrectPoly$,
+    $P_1$ be the sum of \kl{maximal monomials} of $P$,
+    and $P_2 \defined P - P_1$ be the sum of
+    non-maximal monomials of $P$.
+    There exists a computable $K$,
+    such that
+    $(\Diff{K}{P_1} + P_2) \in \CorrectPoly$.
+\end{lemma}
 
@@
 
-    \textbf{Induction:}
+    \textbf{Induction:} Assume that $P$ is not a constant polynomial.
+    Leveraging \cref{derivation-stabilises-correct:lem}, there exists a
+    computable $K \in \Nat$ such that $\Diff{K}{P} \in \CorrectPoly$. Using
+    \cref{derivation-simplifies:fact} (because $P$ is non-constant), we also
+    know that $\MaximalMonomials(\Diff{K}{P}) \hoarele \MaximalMonomials(P)$.
+    By induction hypothesis, we conclude that $\Diff{K}{P}$ is \kl{represented}
+    by a \kl{star-free $\Nat$-polyregular function} $f_1$.
+
     Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
@@
     the polynomial
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$
+    $\Diff{K}{P} \defined P - \alpha (Y - K)^p M$
     satisfies that for all partial valuations $\nu \colon Y\vec{X} \topartial \Nat$
@@
     the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
-    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
-    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
+    In particular, $\Diff{K}{P} (Y - K)$  belongs to $\CorrectPoly$.
+    Furthermore, $\MaximalMonomials(\Diff{K}{P}) = \MaximalMonomials(\Diff{K}{Y - K})
     \hoarele \MaximalMonomials(P)$.
@@
     we conclude that
-    $\Diff{K}(Y - K)$ is \kl{represented}
+    $\Diff{K}{Y - K}$ is \kl{represented}
     by a \kl{star-free $\Nat$-polyregular function} $g_1$,

### 2024-01-16 parts/conclusion.tex
wohoho

--- 2024-01-11 pnrs.tex
@@
 %! TeX program = xelatex
-%! lang = en-GB
-\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
-
-% INPUTS
-\input{globals/packages}
-\input{globals/colorscheme}
-
-% CONFIG
-\input{globals/macros}
-\input{globals/mathdefs}
-\input{globals/knowledges.kl}
-\input{globals/metadata}
-
-\usepackage{comment}
-%\includecomment{forshort}
-%\excludecomment{proof}
-
-\makeabstract
-% DOCUMENT
-\begin{document}
-
-% FRONT MATTER
-\maketitle
-\acknowledge
-
-\section{TODOs}
-
-\begin{itemize}
-    \item Use $\Nat$-polyregular functions all the time, and just 
-        in the introduction talk about $\Nat$-rational series.
-    \item Prove that $\Nat$-rational series of polynomial 
-        growth are $\Nat$-polyregular functions, maybe using
-        the results about ambiguity in automata?
-        Theorem 5.22
-    \item Define commutativity by a map 
-        $\commute \colon \Sigma^* \to \Nat^{\Sigma}$.
-    \item Define "commutative functions" and its symbol.
-    \item Define "computed by" (like, recognised by) for commutative 
-        functions.
-    \item Change \cref{corrected-version:thm} to use directly
-        star-free!
-    \item Clear up the computability problems.
-    \item Polynomial + Npolyregular => star free?
-    \item Polynomial + N rational => N polyregular?
-    \item cite the paper of Quantitative Monadic Second-Order Logic
-        about why it is interesting class.
-\end{itemize}
-
-% MAIN MATTER
-\section{Introduction}
-
-There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
-\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
-This computational model, in the counting case, can be traced back to a model
-introduced by \textcite{schutzenberger1965finite}. It has been recently proven
-that one can decide the growth rate of such functions, and decide whether they
-are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. Isolating
-a notion of aperiocidity for functions, and deciding this property is often a
-non-trivial process \cite{droste2019aperiodic}. It was conjectured in the PhD
-thesis of Gaëtan Douéneau-Tabot, that the membership problem of $\Nat$-rational
-series among $\Rel$-rational series was decidable under the assumption of
-polynomial growth. The semantic characterization would imply that
-$\Nat$-rational series of polynomial growth that are aperiodic are precisely
-$\Nat$-rational series of polynomial growth, that can be realised by
-$\Rel$-rational series that are aperiodic. This conjecture was based on the
-previously introduced result of \textcite[Theorem 3.3]{KARH77}, which we will
-disprove in \cref{sec:c-example}. 
-We then provide a corrected version of the theorem, and prove the conjecture
-of Douéneau in the case of \emph{commutative} $\Nat$-rational series of
-polynomial growth.
-
-\section{Preliminaries}
-
-\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a production function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        f(w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-\end{definition}
-
-\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
-    \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be a function. The following are equivalent:
-    \begin{enumerate}
-        \item $f$ can be computed as a \k{$\Nat$-rational series}
-            of \kl{polynomial growth}.
-        \item $f$ is a \kl{$\Nat$-polyregular function}.
-    \end{enumerate}
-\end{fact}
-
-\begin{definition}[Letter Counting Function]
-    \label{commutative-map:def}
-    Let $\Sigma$ be a finite alphabet. We define
-    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
-    map that counts occurrences of every letter in the input word, namely:
-    \begin{equation*}
-        \commute[w] \defined \seqof{\card[a]{w}}{a \in \Sigma} \quad .
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[Commutative Function]
-    \label{commutative:def}
-    Let $X$ be a set.
-    A function $f \colon \Sigma^* \to X$ is
-    \intro{commutative} whenever
-    for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
-
-    Equivalently, it is \reintro{commutative} whenever
-    there exists a map $g \colon \Nat^\Sigma \to X$
-    such that $g \circ \commute = f$.
-\end{definition}
-
-\begin{definition}[Represented Function]
-    A function $g \colon \Nat^n \to \Rel$
-    is \intro{represented} by a function $f \colon \Sigma^* \to \Rel$
-    if there exists a map $\mu \colon \set{1, \dots, n} \to \Sigma$
-    such that
-    $g \circ (v \mapsto \seqof{v_{\mu(k)}}{1 \leq k \leq n}) \circ \commute = f$.
-\end{definition}
-
-We need, $\ZRat$, $\NRat$, $\ZPoly$, $\NPoly$, $\ZSF$, and $\NSF$.
-
-We will start by formally stating the result of \citeauthor{KARH77}
-in terms of polynomials and commutative $\Nat$-rational series, and
-then introduce the necessary definitions. The characterization of
-commutative $\Nat$-rational series computing polynomial functions
-is described in terms of a class $\CoveredPoly[\vec{X}]$ of
-polynomials with multiple indeterminate.
-
-\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CoveredPoly[\vec{X}]$
-    is the class of polynomials $P \in \Rel[\vec{X}]$
-    that are \kl{non-negative}
-    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
-    When the indeterminate are clear from the context, we write
-    this class $\CoveredPoly$.
-\end{definition}
-
-We use the term \kl{$\Nat$-rational polynomial}, that will formally be
-stated later on, to denote polynomials that can be computed via a
-$\Nat$-rational series. In the upcoming theorem, this property
-which is \emph{a priori} difficult to decide is reduced to 
-the positivity of the polynomial, and the positivity of \emph{some}
-coefficients.
-
-\begin{faketheorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
-    \label{karh:thm}
-    Let $\Sigma$ be a finite alphabet.
-    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
-    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
-\end{faketheorem}
-
-It is not clearly stated in \cite{KARH77}, but \cref{karh:thm} implies the
-decidability of \kl{$\Nat$-rational polynomial} inside polynomials with $\Rel$
-coefficients. We will not formalize this statement yet, as \cref{karh:thm} does
-not hold in general.
-
-We then provide in \cref{sec:proof} a
-corrected version of the theorem, together with a valid proof. As a
-consequence, it does not invalidate the stated conjecture.
-
-\begin{itemize}
-    \item commutative input and output
-    \item commutative output
-    \item no commutativity ?
-\end{itemize}
-
-\AP \paragraph*{Rational series and regular languages.} This paper uses notions
-of regularity and results about \kl{$\Nat$-rational series}, however most of
-the technicalities will stem from the multiple indeterminate of polynomials,
-and all there is to know about regularity in this document will be confined to
-the following facts. For a survey about (non-commutative) rational series, we
-refer the reader to \cite{berstel2011noncommutative}.
-
-\begin{definition}[Rational series]
-    The following are equivalent
-    \begin{itemize}
-        \item weighted automata
-        \item rational expressions
-        \item 
-    \end{itemize}
-\end{definition}
-
-\begin{fact}[Folklore about regular languages]
-    \label{regular:fact}
-    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular, whenever $a,b \in \Sigma$.
-    Regular languages are closed under intersection.
-\end{fact}
-
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by a $\Nat$-rational series
-    is a regular language. In particular,
-    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
-    language.
-\end{fact}
-
-\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational polynomial} if
-there exists a \reintro{commutative} $\Rel$-rational series $f$ such that $f(w)
-= P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, $P$ is a
-\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
-series is in fact $\Nat$-rational. Beware that we only apply the polynomial
-to \emph{non-negative} values.
-
-\begin{example}
-    \label{negative-not-nrat:ex}
-    With $\Sigma \defined \set{1}$,
-    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
-    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    The function $w \mapsto |w|$ is a \kl{polyregular function} with unary
-    output of \kl{polynomial growth}, hence  can be computed by a
-    \kl{$\Nat$-rational series}. Hence, $P(X_1) \defined X_1$ is
-    a \kl{$\Nat$-rational polynomial}. Similarly,
-    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-rational series},
-    showing that $Q(X_1) \defined X_1^2 + 3$
-    is a \kl{$\Nat$-rational polynomial}.
-    Finally, 
-    $T(X_1) \defined - X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
-    have non-negative output.
-\end{proof}
-
-We saw in \cref{negative-not-nrat:ex}
-a simple criterion to check whether a polynomial is $\Nat$-rational,
-but being non-negative is not enough, as the following example
-illustrates. Note that the proof scheme
-of this example will be at the core of \cref{thm:counter-example}.
-
-\begin{example}
-    Over a binary alphabet $\Sigma \defined \set{a,b}$,
-    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
-    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
-    but is
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Assume by contradiction that
-    $S$ is a $\Nat$-rational series computing $P$.
-    Then, $S^{-1}(\set{0})$ should be a regular language
-    (\cref{pre-image-regular:fact}),
-    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not a regular language (\cref{regular:fact}).
-\end{proof}
-
-\AP Let us now give some vocabulary on polynomials with multiple indeterminate
-over $\Rel$. A \intro{monomial} is a product of variables and integers. For
-instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, but $X + Y$ or
-$2X^2 + XY$ are not.
-
-\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} if
-$P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we do
-not consider negative values, as the numbers $n_i$ will ultimately count the
-number of occurrences of a letter in a word. As an example, the polynomial $(X
-- Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$, but the
-polynomial $X^2 - 2X$ is not.
-
-\AP
-In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
-is a \intro{maximal monomial} if it is a maximal element
-for the visibility ordering of polynomials \emph{in $\Rat$}
-among monomials with non-zero coefficients in $P$.\footnote{
-    One could equivalently define maximal elements by removing the (non-zero)
-    multiplicative constant of the monomials, and then
-    considering the usual divisibility relation of $\Rel$ polynomials.
-}
-In the polynomial 
-$P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
-are $X^2$, $-2 XY$, and $X^2$.
-A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$ if its
-multiplicative constant is positive. In the polynomial $P(X,Y) \defined (X -
-Y)^2$, the \kl{positive monomials} are $X^2$ and $Y^2$.
-
-\paragraph*{Outline of the paper.}
-todo.
-
-\section{The Counter Example}
-\label{sec:c-example}
-
-Let us now introduce the counter example that was found trying to generalise
-the original proof to non-commutative \kl{$\Rel$-rational series} of
-\kl{polynomial growth}. The counter example will use three indeterminate, and
-we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
-polynomial with at most two indeterminate.
-
-\begin{definition}[Counter Example Polynomial]
-    \label{def:bad-polynomial}
-    We define $P(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
-\end{definition}
-
-\begin{theorem}
-    \label{thm:counter-example}
-    Over $\Sigma \defined \set{a,b,c}$,
-    there exists a polynomial $P \in \CoveredPoly$ that is not
-    a \kl{$\Nat$-rational polynomial}. Namely,
-    $P$ is the polynomial of \cref{def:bad-polynomial}.
-\end{theorem}
-\begin{proof}
-    Let us write $X,Y,Z$ instead of $X_a, X_b, X_c$ to improve
-    readability.
-    It is clear that $P$ is \kl{non-negative}. We can develop
-    the expression of $P$ to 
-    obtain
-    $P = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
-    The \kl{maximal monomials} of $P$
-    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
-    \kl{positive coefficients}.
-
-    Assume by contradiction that $P$ is a \kl{$\Nat$-rational polynomial}.
-    There exists a \kl{commutative}
-    $\Nat$-rational series $f \colon \Sigma \to \Nat$
-    such that $P(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-    Remark that $P(X,Y,Z) = 0$
-    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-    $P(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
-    $Z \neq 0$, and $X = Y = 0$.
-
-    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
-    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
-    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ is notoriously not a regular language
-    (\cref{regular:fact}), we
-    conclude that $L$ is not a regular language.
-    However, $L = f^{-1}(\set{0})$, which is a regular language
-    because $f$ is a $\Nat$-rational series
-    (\cref{pre-image-regular:fact}).
-\end{proof}
-
-\begin{corollary}
-    The result stated in \cite[Theorem 3.3]{KARH77}, restated
-    in \cref{karh:thm}, is false
-    for all polynomials with at least $3$ indeterminate.
-\end{corollary}
-
-As we will see in the next section, the characterization
-of \kl{$\Nat$-rational polynomials} using the set $\CoveredPoly$
-holds when the polynomials have at most $2$ indeterminate. This shows
-that the examples in \cite{KARH77} are not invalidated, as they
-all use at most two indeterminate.
-
-\section{The Corrected Theorem}
-\label{sec:proof}
-
-Let us now provide another class of polynomials, $\CorrectPoly$, that restricts
-$\CoveredPoly$ to a subset that is ``closed under fixing indeterminate", namely,
-polynomials $P(X_1, \dots, X_n) \in \CoveredPoly$ such that partially evaluated
-polynomials remain in $\CoveredPoly$. We use the following notation
-to fix the value of some indeterminate
-$\intro*\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally, if $\nu$ is
-a partial function from $\vec{X}$ to $\Nat$, the restriction
-$\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminate $\vec{Y}
-\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the domain of
-$\nu$.
-
-\begin{definition}
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CorrectPoly[\vec{X}]$ is the collection of
-    polynomials $P \in \Rel[\vec{X}]$ such that
-    $P$ is 
-    such that, for every partial function $\nu \colon \vec{X} \to \Nat$,
-    every \kl{maximal monomial} of
-    $\restr{P}{\nu}$ has a \kl{positive coefficient}.
-\end{definition}
-
-First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
-polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
-counter example provided in \cref{thm:counter-example} is not in
-$\CorrectPoly$. For that, notice that for $P$ defined in
-\cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$. Let us also assert that
-polynomials that have all coefficients in $\Nat$ are \kl{$\Nat$-rational
-polynomials}, which will be a base case in an upcoming induction.
-
-\begin{remark}
-    \label{inductive-correct-poly:remark}
-    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item The \kl{maximal monomials} of 
-            $P$ have \kl{positive coefficients},
-            and 
-            for all $n \in \Nat$,
-            $\restr{P}{X = n} \in \CorrectPoly$.
-    \end{enumerate}
-    Furthermore, if $P \in \Rel[]$ is a polynomial without
-    indeterminate, then $P \in \CorrectPoly$ if and only
-    if its \kl{maximal monomial} (the constant term)
-    has a \kl{positive coefficient}.
-\end{remark}
-\begin{proof}
-    By induction on the number of indeterminate,
-    leveraging the fact that if there exists at least one indeterminate
-    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
-    and partial function $\nu \colon \vec{Y} \to \Nat$,
-    we have 
-    \begin{equation*}
-        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
-        \quad
-        ,
-    \end{equation*}
-    where $\nu[X = n]$ is the partial function from
-    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
-    and maps $X$ to $n$.
-\end{proof}
-
-\begin{example}
-    \label{n-poly-n-poly:example}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
-    be a partial function.
-    Notice that $\restr{P}{\nu}$ continues to be
-    a polynomial with \kl{positive coefficients}, and in particular
-    with 
-    \kl{maximal monomials} having \kl{positive coefficients}.
-\end{proof}
-
-Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-To simplify the reasoning, we will use results on $\Nat$-polyregular functions,
-also known as, polyregular functions with unary output
-\cite{doueneau2021pebble,bojanczyk2019string}. Note that the connection between
-rational series and polyregular functions is not specific to $\Nat$, as
-$\Rel$-rational series of polynomial growth are computed as polyregular
-functions with output in $\set{+1,-1}$ post-composed with computing the sum of
-outputted letters \cite{LOPEZ23b}. Because we only consider functions with
-output in $\Rel$, and that a commutative output monoid simplifies the definitions,
-let us only define $\Rel$-polyregular functions.
-
-\begin{lemma}
-    \label{n-poly-combinatorics:lem}
-    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
-    There exists $\omega \in \Nat$
-    such that for all $p \in \Nat$,
-    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
-    for all $u_1, \dots, u_p \in \Sigma^*$,
-    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
-    whose \kl{maximal monomials} have \kl{positive coefficients},
-    and such that:
-    \begin{equation*}
-        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
-        = P(X_1, \dots, X_p) \quad .
-    \end{equation*}
-\end{lemma}
-\begin{proof}[Proof Sketch]
-    Let $\omega$ be an idempotent power for the finite monoid $M$.
-    We define an equivalence between partitions of a word 
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$
-    as follows: two partitions into $k$ words in $\Sigma^*$ are equivalent
-    when normalizing them with the rules
-    $u_i^{\omega} u_i^{\omega} \to u_i^{\omega}$ for all $1 \leq i \leq p$.
-    This equivalence relation has finite index, and all partitions in the same
-    equivalence class yield the same value through $\pi$, because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid $M$.
-
-    Now, let us remark that the number of partitions
-    in a certain equivalence class is proportional to
-    the number of nanania, that is, is of the form
-    $(X_i - K_i)^{n_i}$ for some $K$. We conclude that
-    maximal monomials indeed have positive coefficients.
-    We even conclude that
-    the polynomial is in $\Nat[\vec{X}]$
-    if we shift the variables with a large enough $K$. 
-\end{proof}
-
-\begin{lemma}
-    \label{n-rat-correct:lem}
-    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-\end{lemma}
-\begin{proof}[Proof Idea]
-    Let $P$ be a \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma$  be the finite alphabet such that
-    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
-    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
-    $\Nat$-rational series computing $P$.
-    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
-    it is a $\Nat$-polyregular function 
-    (\cref{polynomial-rational-polyreg:fact}).
-
-    Using \cref{n-poly-combinatorics:lem},
-    we conclude that maximal monomials of $P$ have
-    positive coefficients. Furthermore, when fixing variables
-    in $P$, we are computing a new \kl{$\Nat$-polyregular function},
-    hence we also conclude that the maximal monomials have
-    positive coefficients.
-    We have proven that $P \in \CorrectPoly$.
-\end{proof}
-
-The core of the upcoming proof of \cref{lem:correct-to-n-rat} relies on the
-following observation: if a polynomial $P$ is written $X^p + Q$, where all
-terms of $Q$ have degree less than $p$, then one can find some large enough $K$
-so that $X^p - (X - K)^p + Q$ also has a positive maximal coefficient.
-It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence
-the maximal coefficient can be chosen arbitrarily large by increasing
-$K$.
-This
-trick will work similarly in multivariate polynomials, and the overall
-proof scheme works as follows: for a bounded number of possible
-values of $X$, we can fix the variable, and have one less item to work on.
-For large enough values of $X$ (typically greater than some $K$), we can guarantee that
-$X^p - (X - K)^p + Q$ has positive maximal coefficients, hence
-can be computed by a $\Nat$-rational series, while
-$(X - K)^p$ is clearly computable by a $\Nat$-rational series,
-hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
-as the sum of two $\Nat$-rational series, and we have
-concluded that our original polynomial was a
-\kl{$\Nat$-rational polynomial}. There are a number of problems in generalizing
-this proof scheme:
-\begin{itemize}
-    \item We will have to guarantee that every polynomial considered
-        is non-negative
-    \item We do not only need to check that the \kl{maximal monomials}
-        have \kl{positive coefficient}, but that it is the case
-        when fixing arbitrarily many variables! (this
-        covers the first point too)
-\end{itemize}
-
-\begin{lemma}
-    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
-    $\vec{p} \colon \vec{X}Y \to \Nat$,
-    and $M \defined \times \prod_{X \in \vec{X}} X^{p(X)}$.
-    Then, for all $K \geq 0$,
-    $\Delta_{Y,K} \defined Y^p M - (Y - K)^p M$ has \kl{maximal monomials} with
-    \kl{positive coefficients} of weight at least $K$, regardless of
-    the variables fixed, as long as the (potential)
-    value of $Y$ is greater
-    than $K$.
-\end{lemma}
-\begin{proof}
-    Let us first remark that
-    \begin{equation*}
-        \Delta_{Y,K}
-        = KY^{p-1} M + Q M
-        \quad .
-    \end{equation*}
-    In particular, there is a unique \kl{maximal monomial} for $\Delta_{Y,K}$,
-    which is precisely $K Y^{p-1} M$, because all monomials
-    in $Q M$ divide (strictly) $Y^{p-1} M$.
-    Let us fix a variable,
-    $X$ to any (non-zero) value $p > 0$. Then
-    The \kl{maximal monomial} of $\restr{\Delta_{Y,K}}{X = p}$
-    is exactly $\restr{KY^{p-1}M}{X = p}$, which remains
-    with a \kl{positive coefficient}, that is proportional to $K$.
-    Let us fix $Y$ to some value $K + l$, with $l > 0$,
-    then
-    $\restr{\Delta_{Y,K}}{Y = K + l}
-    = (K + l)^p M - l^p M
-    = ((K + l) - l) (\sum_{i = 1}^{p-1} (K + l)^i l^{p-i}) M
-    = K \alpha M
-    $
-    which remains a monomial with a \kl{positive coefficient},
-    that is proportional to $K$.
-\end{proof}
-
-We can refine the above lemma to use it in the case where
-we do not have a single
-monomial.
-
-\begin{lemma}
-    \label{lem:delta-cool}
-    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
-    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
-    \kl{maximal monomial} of $P$, with $\alpha > 0$.
-    Then, there exists a computable $K \geq 0$
-    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
-    has \kl{positive coefficients} for \kl{maximal monomials}
-    in any partial evaluation that satisfies $Y > K$ (if it
-    gives a value to $Y$).
-\end{lemma}
-\begin{proof}
-    Let $N$ be greater than the sum of the absolute values of all coefficients
-    appearing in $P$, and let $K = 2 N$.
-
-    Let us remark that
-    \begin{equation*}
-        \Delta_K = \underbrace{(P - \alpha Y^p M)}_{\text{ left hand side }}
-        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{\text{ right hand side }}
-        \quad .
-    \end{equation*}
-    an that for every partial evaluation $\nu$ that sends $Y$ to a value
-    greater than $K$,
-    we have that $\restr{(Y^p M - (Y - K)^p M)}{\nu}$ has
-    \kl{maximal monomials} with \kl{positive coefficients}
-    proportional to $K$ (note that here ``proportional"
-    implies that they are greater than $K$, because we multiply by positive
-    integers).
-
-    To conclude, it is just a game of describing how
-    \kl{maximal monomials} of $\restr{\Delta_Y}{\nu}$
-    can appear. If the \kl{maximal monomial} is obtained
-    by summing an element of the left hand side with an element
-    of the right hand side, it needs to be a \kl{maximal monomial}
-    of the right hand side, and the large value of $K$ guarantees
-    that the sum of coefficients is positive.
-    If it only appears on the right hand side, it is even better.
-    If it only appears on the left hand side, then
-    in particular it means that the monomial does not
-    divide $\restr{Y^p M}{\nu}$, hence has the same value
-    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
-    which is guaranteed to be positive because $P \in \CorrectPoly$.
-
-    In particular, we conclude from this that
-    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
-\end{proof}
-
-Note that we cannot conclude in the above lemma that the result is in
-$\CorrectPoly$, because we have a restriction on the range where the variable
-$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
-induction on the number $d$ of indeterminate, and the degrees of the
-\kl{maximal monomials}. To make the induction formal, let us define
-$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
-$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
-$\Nat^{\vec{X}}$ with the product ordering, which is a
-\kl{well-quasi-ordering}.
-
-\AP To a polynomial $P$, we assign the set $\MaximalMonomials(P)$ of its
-\kl{maximal monomials}, belonging to $\Pfin(\Monomials)$, which remains a
-\kl{well-quasi-ordering} when endowed with the \intro{Hoare ordering} of
-subsets, that is, $S_1 \hoareleq S_2$ whenever for all \kl{monomial} $M_1 \in
-S_1$, there exists a monomial $M_2 \in S_2$, such that $M_1$ \kl{divides}
-$M_2$.
-
-\begin{lemma}
-    \label{lem:correct-to-n-rat}
-    Let $P$ be a polynomial in $\CorrectPoly$.
-    Then, 
-    $P$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on $\MaximalMonomials(P)$.
-
-    \textbf{Base case:} If the \kl{maximal monomial} of $P$ is a constant term,
-    then because $P \in \CorrectPoly$, $P = n \in \Nat$, which is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-
-    \textbf{Induction:}
-    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
-    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
-    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
-    Using \cref{lem:delta-cool},
-    we know that there exists a computable $K$ such that
-    the polynomial
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$
-    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
-    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
-    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
-    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
-    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
-    \hoarele \MaximalMonomials(P)$.
-    By induction hypothesis,
-    we conclude that
-    $\Diff{K}(Y - K)$ is \kl{represented}
-    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
-    \begin{equation*}
-        f_1(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \Diff{K} & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $\alpha \geq 0$, it is an easy check
-    that the following function is also a $\Nat$-rational series
-    \begin{equation*}
-        f_2(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \alpha(Y - K)^pM & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $P \in \CorrectPoly$, for every $n \leq K$, $\restr{P}{Y = n} \in
-    \CorrectPoly$ (\cref{inductive-correct-poly:remark}).
-    Furthermore,
-    $\MaximalMonomials(\restr{P}{Y = n})
-    \hoarele \MaximalMonomials(P)$. Hence, by induction hypothesis,
-    the following function is \kl{represented} by a
-    \kl{star-free $\Nat$-rational series}:
-    \begin{equation*}
-        f_3(Y, \vec{X}) \defined
-        \begin{cases}
-            P(Y, \vec{X}) & \text{ if } Y \leq K \\
-            0 & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    We conclude by remarking that 
-    $P = f_1 + f_2 + f_3$.
-\end{proof}
-
-We are now ready to state the corrected and generalised version of
-\cref{karh:thm}.
-
-\begin{theorem}
-    \label{corrected-version:thm}
-    Let $P \in \Rel[\vec{X}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item \label{corrected-1:item} $P \in \CorrectPoly$,
-        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
-        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-    \end{enumerate}
-\end{theorem}
-\begin{proof}
-    The implications 
-    \cref{corrected-4:item} $\implies$
-    \cref{corrected-3:item} $\implies$
-    \cref{corrected-2:item} are obvious.
-    \cref{lem:correct-to-n-rat} proves
-    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
-    while \cref{n-rat-correct:lem}
-    proves 
-    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
-\end{proof}
-
-\begin{lemma}
-    \label{lem:correct-covered-2}
-    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
-\end{lemma}
-\begin{proof}
-    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \to \Nat$.
-    For the converse inclusion, let us consider $P(X,Y)$
-    that is \kl{non-negative}, such that the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-
-    If we fix none of the variables, then the \kl{maximal monomials}
-    have \kl{positive coefficients} by assumption. If we fix one of the
-    variables, we can assume without loss of generality that we 
-    fix $X = k$ for some $k \in \Nat$.
-    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
-    and therefore must have a positive leading coefficient
-    (which is the unique \kl{maximal monomial} in this case)
-    or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-    The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminate, leading to a constant polynomial.
-\end{proof}
-
-\section{Effective procedures}
-\label{sec:deciding}
-
-The goal of this section is to go beyond polynomials and characterize all
-$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
-that they are \kl{commutative}.
-
-\begin{theorem}
-    \label{decidable-n-poly:thm}
-    Given a \kl{commutative}
-    \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to \Rel$,
-    one can decide if $f$ can be computed by a \kl{$\Nat$-polyregular function},
-    and effectively compute a representation of $f$.
-\end{theorem}
-
-Before proving \cref{decidable-n-poly:thm}, let us first examine the
-effectiveness of \cref{corrected-version:thm}, namely, provide an algorithm to
-decide whether a polynomial $P$ belongs to $\CorrectPoly$. The idea will be to
-compute a bound to values taken by the valuations $\nu$ appearing in the
-definition of $\CorrectPoly$. As a \emph{mise en jambes}, let us consider
-separately the case of one indeterminate.
-
-\begin{lemma}
-    \label{decide-correct-univariate:lemma}
-    Let $P \in \Rel[X]$, there exists a computable $N \in \Nat$,
-    such that the following are equivalent:
-    \begin{enumerate}
-        \item \label{decide-correct-univariate-1:item}
-            $P \in \CorrectPoly$,
-        \item \label{decide-correct-univariate-2:item}
-            The \kl{maximal monomial}
-            of $P$ has a \kl{positive coefficient},
-            and $P(n) \in \Nat$
-            for all $0 \leq n \leq N_0$.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    Let us first remark that the implication 
-    \cref{decide-correct-univariate-1:item}
-    $\implies$ \cref{decide-correct-univariate-2:item} is trivial for every
-    $N \in \Nat$.
-    For the converse implication,
-    let $P \in \Rel[X]$.
-    If $P$ is a constant polynomial, then $P \in \CorrectPoly$
-    if and only if $P(0) \in \Nat$, and $N \defined 0$
-    suffices.
-    Otherwise, $P$ is non-constant, and let $n \geq 1$ be the \kl{degree} of $P$.
-    Let $P_1$ and $\alpha > 0$ be such that
-    $P \defined P_1 + \alpha X^n$. 
-    To prove that $P \in \CorrectPoly$, it suffices to show that
-    $P(k) \in \Nat$ for all $k \in \Nat$.
-    Let us define $\beta$ as the
-    maximal absolute value of coefficients appearing in $P_1$, which is
-    assumed to be non-zero,
-    and $N_0 \defined \beta + 1 \geq 2$.
-    Remark that for $X \geq N_0$,
-    \begin{align*}
-        P(X) &\geq X^n - \beta \sum_{i = 0}^{n-1} X^i \\
-             &= X^n - \beta \frac{X^n - 1}{X - 1} \\
-             &= \frac{X^{n+1} - X^n - \beta X^n + \beta}{X - 1} \\
-             &= \frac{X^n(X - (\beta + 1)) + \beta}{X - 1} \\
-             &\geq 0
-    \end{align*}
-    This proves that $P(k) \in \Nat$ for all $k \geq N_0$,
-    and because $P$ satisfies \cref{decide-correct-univariate-2:item},
-    $P(k) \in \Nat$ for all $1 \leq k \leq N_0$. We have proven 
-    that $P \in \CorrectPoly$.
-\end{proof}
-
-In the case of multiple indeterminate, we leverage
-\cref{lem:delta-cool} to perform an induction on the degree
-of the polynomials and the number of variables of the polynomial.
-
-\begin{lemma}
-    \label{decide-correct-poly:lem}
-    One can decide if a polynomial belongs to $\CorrectPoly$.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on the number of indeterminate.
-
-    \textbf{Base Case:} If $P$ has zero indeterminate, i.e., is a constant
-    function, then to decide whether $P \in \CorrectPoly$ it suffices to check
-    that $P(0) \in \Nat$.
-
-    \textbf{Induction:} Let $P \in \Rel[Y, \vec{X}]$, we start by checking
-    whether \kl{maximal monomials} of $P$ have \kl{positive coefficients}.
-    Without loss of generality, assume that $Y$ divides some \kl{maximal
-    monomial} $\alpha Y^p M$ with $p \geq 1$. By applying
-    \cref{lem:delta-cool}, we can compute a bound $K \in \Nat$ such that
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$ has \kl{positive coefficients}
-    for its \kl{maximal monomials} for all partial valuations $\nu \colon
-    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
-    particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to
-    $\CorrectPoly$ for all $n > K$. By induction hypothesis, one can
-    decide whether $\restr{P}{X = n}$ belongs to $\CorrectPoly$
-    for all $0 \leq n \leq K$, which is equivalent 
-    to deciding whether
-    $P \in \CorrectPoly$ thanks to \cref{inductive-correct-poly:remark}.
-\end{proof}
-
-To prove the theorem, let us first provide a nice characterization of
-\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
-This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
-commutative case.
-
-\begin{lemma}
-    \label{decompose-polynomial:lem}
-    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
-    \kl{$\Rel$-polyregular function}. There exists $\omega \in \Nat$,
-    and polynomials $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{ 0 \leq n_a < \omega}$,
-    such that for all $w \in \Sigma^*$,
-    \begin{equation*}
-        f(w) = P_{\seqof{\card[a]{w} [\omega]}{a \in \Sigma}}
-        (\seqof{\card[a]{w} / \omega}{a \in \Sigma}) \quad .
-    \end{equation*}
-    Which informally means that for every choice of rest modulo $\omega$
-    for every variable, there exists a polynomial describing the
-    behavior of $f$ when adding multiples of $\omega$ letters at a time.
-
-    Furthermore, the polynomials are computable from $f$.
-\end{lemma}
-\begin{proof}
-    equivalent partitions of the word, but commutativity?.q
-    Lemma 5.37 phd gaëtan.
-\end{proof}
-
-As a consequence, we conclude 
-\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
-    We use \cref{decompose-polynomial:lem}
-    to effectively compute
-    a polynomial representation of $f$. Remark that
-    $f$ is a \kl{$\Nat$-polyregular function} if and only if
-    all these polynomials are \kl{$\Nat$-rational polynomials},
-    which is decidable thanks to
-    \cref{decide-correct-poly:lem,corrected-version:thm}.
-\end{proof}
-
-\subsection{Star-free $\Nat$-polyregular functions}
-
-Let us now prove that the above characterizations of
-commutative
-$\Nat$-polyregular functions can be combined with the recent
-advances in the study of $\Rel$-polyregular functions 
-\cite{LOPEZ23b}
-to prove the following theorem.
-
-\begin{theorem}
-    \label{zsf-npoly-nsf:thm}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be both a \kl{star-free $\Rel$-polyregular function}
-    and a \kl{commutative} \kl{$\Nat$-polyregular function}.
-    Then, $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}.
-    That is, the following equality holds:
-    \begin{equation*}
-        \ZCommut \cap \ZSF \cap \NPoly
-        = \ZCommut \cap \NSF
-        \quad .
-    \end{equation*}
-\end{theorem}
-
-Note that this proves in the commutative case a conjecture that was 
-formulated by Thesis Gaetan, and that is non trivial, because
-
-\begin{conjecture}
-    \label{nsf-zsf:conj}
-    In the general case, the following equation holds:
-    \begin{equation*}
-        \ZSF \cap \NPoly
-        = 
-        \NSF
-        \quad .
-    \end{equation*}
-\end{conjecture}
-
-As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
-procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
-decidable inside $\NPoly$ under the assumption of commutativity.
-
-\begin{corollary}
-    One can decide if a \kl{commutative} \kl{ $\Rel$-polyregular function}
-    can be realized by a \kl{star-free $\Nat$-polyregular function},
-    in which case the conversion is effective.
-\end{corollary}
-
-The key ingredient of this section is the use of a semantic characterization of
-\kl{star-free $\Rel$-polyregular functions} among \kl{$\Rel$-rational series}
-that generalizes the aperiodicity of languages to functions by the means of
-polynomial behaviors.
-
-\begin{definition}[Ultimately polynomial]
-    \label{ultimately-polynomial:def}
-    Let $\Sigma$ be a finite alphabet. 
-    A function $f \colon \Sigma^* \to \Rel$
-    is \intro{ultimately polynomial}
-    whenever there exists $N_0 \in \Nat$ such that
-    for all $n \in \Nat$
-    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
-    \in \Sigma^*$, there exists a polynomial $P \in \Rel[X_1, \dots, X_n]$
-    such that
-    \begin{equation*}
-        f\left(
-            \alpha_0 \prod_{i = 1}^{n} w_i^{X_i} \alpha_i
-        \right)
-        = 
-        P(X_1, \dots, X_n)
-        \quad 
-        \forall X_1, \dots, X_n \geq N_0
-        \quad .
-    \end{equation*}
-\end{definition}
-
-Let us recall from \cite{LOPEZ23b} that this definition correctly
-generalizes the indicator functions of regular languages.
-
-\begin{example}
-    A language $L$ is aperiodic if and only if 
-    the indicator function is ultimately polynomial.
-\end{example}
-\begin{proof}
-    TODO.
-\end{proof}
-
-\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
-    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
-
-    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
-    only the empty word $\varepsilon$, and because $f$ is a
-    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
-    Note that $f$ is therefore a constant function, equal to a natural
-    number, hence belongs to $\NSF$.
-
-    \textbf{Induction.}
-    Because $f$ is \kl{ultimately polynomial},
-    there exists $N_0 \in \Nat$, 
-    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
-    such that
-    \begin{equation*}
-        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
-        = 
-        P(\seqof{X_a}{a \in \Sigma})
-        \quad
-        \text{ when }
-        \forall a \in \Sigma, X_a \geq N_0
-        \quad .
-    \end{equation*}
-    Furthermore, leveraging \cref{decompose-polynomial:lem},
-    we obtain $\omega \in \Nat$
-    and a family of polynomials 
-    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
-    such that.
-    Because each of those polynomials coincide with $P$ over an infinite
-    subgrid of $\Nat$,
-    we conclude that all polynomials are in fact equal to a single one.
-    we conclude that the polynomials 
-    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.
-
-    Now, we conclude in particular that the function 
-    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
-    hence is computable by a \kl{star-free $\Nat$-polyregular function},
-    because of \cref{nsf-polynomials:thm}.
-    When fixing values of some indeterminate below $N_0$,
-    we obtain a new function $g$, that remains commutative,
-    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
-    By induction hypothesis, this function $g$ is also 
-    \kl{star-free $\Nat$-polyregular}, and we conclude
-    that 
-    $f = P + \sum g$ which remains star free.
-\end{proof}
-
+%! lang = en-US
 \section{Conclusion}
@@
 
-\begin{corollary}
-    The following problem is decidable: given $P \in \Rel[\vec{X}]$,
-    decide whether $P$ is a \kl{$\Nat$-rational polynomial}.
-\end{corollary}
-\begin{proof}
-    Thanks to \cref{karh:thm}, one only has to decide whether
-    $P \in \CoveredPoly$. The procedure is as follows.
-    \begin{enumerate}
-        \item Check if all \kl{maximal monomials} have
-            \kl{positive coefficients} in $P$, if not, answer \texttt{false}.
-        \item Consider $N \in \Nat$ large enough so that
-              $P \geq 0$ when all indeterminate are chosen greater
-              than $N$. This is possible because all \kl{maximal
-              monomials} have \kl{positive coefficients}, and such an $N$
-              is computable.
-        \item Compute $P$
-    \end{enumerate}
-
-    Now, if all \kl{maximal monomials}
-    have \kl{positive coefficients}, then there exists a computable
-    $N \in \Nat$ such that 
-    In particular,
-
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly = \setof{P}{ P \geq 0 \wedge \exists K \in \Nat, P \circ (+K) \in \Nat[\vec{X}]}$.
-\end{conjecture}
-\begin{proof}[Proof Sketch]
-    If $S \defined P \circ (+K) \in \Nat[\vec{X}]$, then for every partial application,
-    maximal coefficients of $S$ are positive. As maximal coefficients of 
-    (partial applications of) $S$ are the same as the ones of $P$
-    \emph{except for the constant terms} (full application),
-    where $P$ is already guaranteed to be non-negative, we conclude
-    that $P \in \CorrectPoly$.
-
-    Conversely, we prove by induction on the degree of $P \in \CorrectPoly$,
-    that there exists some $K$ such that $P \circ (+K) \in \Nat[\vec{X}]$.
-    First, notice that $\CorrectPoly$ is closed under pre-composition with
-    $(+K)$. Then, remark that $P(X - K)$ has the same maximal coefficients as
-    $P$ and new maximal coefficients that are multiples of $K$. Hence, for a
-    large enough $K$, we conclude that $P(X - K)$ is the sum of its maximal
-    coefficients, plus a polynomial that lies in $\CorrectPoly$. We conclude by
-    induction hypothesis. 
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly$ is precisely composed of polynomials that
-    have non-negative coefficients in a \emph{binomial basis},
-    i.e., where polynomials are obtained
-    as sums and products of $\binom{X_i}{p_i}$.
-\end{conjecture}
-
-\begin{conjecture}
-    star-free $\Nat$-rational is star-free and $\Nat$-rational?
-\end{conjecture}
-
 \begin{center}
@@
 \end{center}
-
-% BACKMATTER
-\printbibliography
-
-\appendix
-
-\end{document}

### 2024-01-16 parts/correction.tex
wohoho

%! TeX program = xelatex
%! lang = en-US
\section{The Corrected Theorem}
\label{sec:proof}

The counter example provided by \cref{def:bad-polynomial} relies on the fact
that $\CoveredPoly$ is not stable under fixing indeterminates, while
\kl{$\Nat$-polyregular functions} are stable under the residual operation
$u^{-1} \colon f \mapsto f(u \cdot)$. In this section, we prove that closing
$\CorrectPoly$ under assignments of variable is enough to recover from
\cref{karh:thm}.

\AP We use the following notation to fix the value of some indeterminate, if
$P(X,Y)$ is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is
the polynomial $P(1,Y) \in \Rel[Y]$. More generally, if $\nu$ is a partial
function from $\vec{X}$ to $\Nat$, written $\nu \colon \vec{X} \topartial
\Nat$, the restriction $\restr{P(\vec{X})}{\nu}$ is the polynomial with
indeterminates $\vec{Y} \defined \vec{X} - \dom(\nu)$ obtained by fixing the
variables of the domain of $\nu$.

\begin{definition}
    Let $\vec{X}$ be a finite tuple of indeterminates.
    The class $\CorrectPoly[\vec{X}]$ is the collection of
    polynomials $P \in \Rel[\vec{X}]$ such that
    $P$ is 
    such that, for every partial function $\nu \colon \vec{X} \topartial \Nat$,
    every \kl{maximal monomial} of
    $\restr{P}{\nu}$ is \kl{non-negative}.
\end{definition}

First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
counter example provided in \cref{thm:counter-example} is not in
$\CorrectPoly$. For that, notice that for $\BadPoly$ introduced in
\cref{def:bad-polynomial}, $\BadPoly(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
coefficient for a \kl{maximal monomial}, namely $-2XY$.

\begin{remark}
    \label{inductive-correct-poly:remark}
    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item $P \in \CorrectPoly$,
        \item The \kl{maximal monomials} of 
            $P$ are \kl{non-negative}
            and 
            for all $n \in \Nat$,
            $\restr{P}{X = n} \in \CorrectPoly$.
    \end{enumerate}
    Furthermore, if $P \in \Rel[]$ is a polynomial without
    indeterminate, then $P \in \CorrectPoly$ if and only
    if its \kl{maximal monomial} (the constant term)
    is \kl{non-negative}.
\end{remark}
\begin{proof}
    By induction on the number of indeterminate,
    leveraging the fact that if there exists at least one indeterminate
    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
    and partial function $\nu \colon \vec{Y} \topartial \Nat$,
    we have 
    \begin{equation*}
        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
        \quad
        ,
    \end{equation*}
    where $\nu[X = n]$ is the partial function from
    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
    and maps $X$ to $n$.
\end{proof}

Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
The following \cref{n-poly-combinatorics:lem} is a slight refinement over the
classical combinatorial analysis of \kl{$\Rel$-polyregular functions}
\cite[Lemma 4.16]{LOPEZ23b} and \kl{$\Nat$-polyregular functions} that takes
into account \emph{all} \kl{maximal monomials} instead of a specific one
\cite[Lemma 5.37]{gaetanphd}.

\begin{lemma}
    \label{n-poly-combinatorics:lem}
    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
    There exists $\omega \in \Nat$
    such that for all $p \in \Nat$,
    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
    for all $u_1, \dots, u_p \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
    whose \kl{maximal monomials} are \kl{non-negative},
    and such that for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)
        \right)
        = P(X_1, \dots, X_p) \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

    Let us define an equivalence relation $\equiv$ over words of
    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
    repetitions of the factors $u_i^{\omega}$ in the words.
    We lift the equivalence relation $\equiv$ to $k$-tuples of
    words by pointwise application.

    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then
    $\pi(\mu(v_1), \dots, \mu(v_k)) = \pi(\mu(w_1), \dots, \mu(w_k))$ because
    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid
    $M$. In particular, $f(w)$ can be computed as the sum over equivalence
    classes for $\equiv$, of the value of $\pi$ on one representative,
    which belongs to $\Nat$.

    Therefore, it suffices to prove that the number of partitions of $w$ in a
    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
    with \kl{positive coefficients} for its \kl{maximal monomials}.

    Notice that to produce all elements in a given equivalence class
    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
    back to the normalized version
    of the equivalence class. For all elements in an equivalence class,
    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
    and therefore the number of possible choices
    is either constant, or 
    is a product of binomials of the form $\binom{X - l}{s}$
    which has the desired shape.
\end{proof}

\begin{lemma}
    \label{n-rat-correct:lem}
    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
    by a \kl{$\Nat$-rational series}. Then,
    $P \in \CorrectPoly$.
\end{lemma}
\begin{proof}
    Let $f$ be the \kl{commutative}
    \kl{$\Nat$-rational series}
    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
    it is in fact a \kl{$\Nat$-polyregular function}
    (\cref{polynomial-rational-polyreg:fact}).
    Using \cref{n-poly-combinatorics:lem},
    there exists a number $\omega \in \Nat$
    and polynomial $Q$ with \kl{positive coefficients} for its
    \kl{maximal monomials} such that
    for all $X_1, \dots, X_p \geq \omega$:
    \begin{equation*}
        f\left(
            \prod_{i = 1}^p (a_i)^{\omega \times X_i}
        \right)
        = Q(X_1, \dots, X_p)
        = P(\omega X_1, \dots, \omega X_p) 
        \quad .
    \end{equation*}
    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
    as polynomials,
    and in particular that
    the \kl{maximal monomials} of 
    $P$ have \kl{positive coefficients}.

    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
    the polynomial $\restr{P}{\nu}$ continues to be represented
    by a \kl{$\Nat$-polyregular function}, namely
    $f$ partially applied to a word. As a consequence,
    the \kl{maximal monomials} of
    $\restr{P}{\nu}$ also have \kl{positive coefficients}, and 
    we have proven that $P \in \CorrectPoly$.
\end{proof}

Let us also assert that polynomials that have all coefficients in $\Nat$ are
\kl{$\Nat$-rational polynomials}. In particular, \kl{non-negative}
\kl{monomials} are \kl{$\Nat$-rational polynomials}, which will be a base case
in an upcoming induction proving \cref{lem:correct-to-n-rat}.

\begin{lemma}
    \label{n-poly-n-poly:example}
    Let $P \in \Nat[\vec{X}]$. Then, $P$
    is a \kl{$\Nat$-rational polynomial}.
\end{lemma}
\begin{proof}
    We prove it by remarking that $\NPoly$
    is stable under non-negative linear combinations
    and contains the functions ...
\end{proof}

\AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
\kl{discrete derivative} to perform an induction on the \kl{maximal monomials},
and suitable notions of \kl{translations} of the domain of polynomials. Namely,
we write $\translate{K}$ for the \intro{translation function} that maps a
polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 - K, \dots,
X_n - K)$.

\begin{fact}
    \label{translation-invariance:fact}
    The following sets are stable under \kl{translations}
    $\translate{K}$,
    where $K \in \Nat$ is non-negative:
    \begin{itemize}
        \item $\CorrectPoly$,
        \item \kl{$\Nat$-rational polynomials},
        \item Polynomials \kl{represented}
            by \kl{star-free $\Nat$-polyregular functions}.
    \end{itemize}
\end{fact}

The idea behind the \kl{discrete derivative} introduced in
\cref{discrete-derivative:def} is to consider the map $P \mapsto P(X) -
P(X-K)$, which will be formulated using the \kl{translation function} in the
case of multiple variables.

\begin{definition}
    \label{discrete-derivative:def}
    Let $\vec{X}$ be a tuple of indeterminates,
    $K \in \Nat$,
    and 
    $P \in \Rel[\vec{X}]$ be a polynomial.
    \begin{equation*}
        \Diff{K}{P} \defined 
        P - \translate{-K}(P) \quad .
    \end{equation*}
\end{definition}

Let us now prove that for a single indeterminate $X$, polynomials $P \in
\CorrectPoly$ are \kl{$\Nat$-rational polynomials}. This will illustrate how
the different ingredients piece together. Given a polynomial $P \in
\CorrectPoly$, the goal is to build a \kl{$\Nat$-polyregular function} $f$ that
\kl{represents} $P$. Note that in this case, the function has unary input and
unary output. We prove the result on the degree $\deg(P)$ of the polynomial
$P$, where the base case is trivial because when $P$ is a constant function,
this constant must be in $\Nat$. For the induction step, we can assume that $P
= \alpha X^n + Q$, where $\deg(Q) < \deg(P)$, $n \geq 1$, and $\alpha > 0$. On
the one hand, $\alpha X^n$ is a \kl{$\Nat$-rational polynomial}, but on the
other hand, $Q$ may not belong to $\CorrectPoly$. However, there exists $K \in
\Nat$ such that $\Diff{K}{X^n} + Q$ belongs to $\CorrectPoly$, for instance by
considering $K$ to be the sum of absolute values of coefficients of $P$.
Furthermore, $\Diff{K}{X^n} + Q$ has degree smaller than $\deg(P)$, hence is a
\kl{$\Nat$-rational polynomial}. Now, $P = \alpha (X - K)^n + \Diff{K}{X^n} +
Q$, and therefore, $P(X+K) = X^n + (\Diff{K}{X^n} + Q)(X+K)$ is a
\kl{$\Nat$-rational polynomial}, \kl{represented} by a \kl{$\Nat$-polyregular
function} $f_1$. Now, for values $k$ of $X$ that are smaller than $K$, we can
compute $P(k) = \restr{P}{X=k}$, which belongs to $\Nat$. It is possible to
build a \kl{$\Nat$-polyregular function} $f_2 \colon \Nat \to \Nat$ that
outputs $P(k)$ if the input is at most $K$, and $f_1(k)$ otherwise.

In the actual statement of \cref{corrected-version:thm}, the representation of
polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
functions} to \kl{star-free $\Nat$-polyregular functions}. This has to be
compared with the characterization of \kl{star-free $\Rel$-polyregular
functions} in terms of polynomials \cite[Theorem??]{LOPEZ23b},
that we will re-introduce in \cref{star-free:sec}.

When multiple indeterminate are involved, a bit of care is needed. The key
ingredient is that one has a fine control over the \kl{maximal monomials} of
the \kl{discrete derivatives}, even when fixing some indeterminates.
As one would expect, the \kl{discrete derivatives} are linear operations on
polynomials, that commutes with the partial application operators
$\restr{\cdot}{\nu}$, and the \kl{translation operators} $\translate{K}$.

\begin{fact}
    \label{discrete-deriv-linear:fact}
    For all tuple $\vec{X}$ of indeterminates,
    for all $K \in \Nat$, $L \in \Rel$, for all partial function
    $\nu \colon \vec{X} \topartial \Nat$:
    \begin{enumerate}
        \item $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            is a linear operator,
        \item $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            is a linear operator,
        \item $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
            is a linear operator,
        \item $\Diff{K}{ \cdot } \circ \restr{\cdot}{\nu}
            = \restr{\cdot}{\nu} \circ \Diff{K}{\cdot}$.
        \item $\Diff{K}{ \cdot } \circ \translate{L}
            = \translate{K} \circ \Diff{L}{\cdot}$,
        \item $\restr{\cdot}{\nu} \circ \translate{L}
            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
    \end{enumerate}
\end{fact}

\begin{fact}
    \label{translation-maximal:fact}
    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
    $\MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))$.
\end{fact}

\AP Let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[see e.g.][Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$
with the \intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering} \cite[see e.g.][Hoare quasi-ordering]{SCSC12}.

The following two lemmas show how the \kl{discrete differentiation} operator
interacts with \kl{maximal monomials} with respect to the \kl{Hoare ordering}:
it extracts \emph{submaximal} monomials from a given polynomial.

\begin{lemma}
    \label{derivation-simplifies:lemma}
    For all $P \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    $\MaximalMonomials(\Diff{K}{P}) \hoarele
    \MaximalMonomials(P)$.
\end{lemma}
\begin{proof}
    By linearity of $\Diff{K}{P}$ (\cref{discrete-deriv-linear:fact}), 
    it suffices to prove the result
    for \kl{monomials} $S \in \Monomials$.
\end{proof}

\begin{fact}
    Let $M,T$ be two \kl{monomials}, such that
    $T$ \kl{strictly divides} $M$.
    Then,
    $T$ \kl{divides} some \kl{maximal monomial}
    of $\Diff{K}{M}$.
\end{fact}

\begin{lemma}
    \label{derivation-covers:lemma}
    For all $P,Q \in \Rel[\vec{X}]$ that are non-constant,
    for all $K \in \Nat$,
    if $\MaximalMonomials(P) \hoarele \MaximalMonomials(Q)$,
    then
    $\MaximalMonomials(P) \hoareleq \MaximalMonomials(\Diff{K}{Q})$.
\end{lemma}

\begin{lemma}
    \label{derivation-stabilises-correct:lem}
    Let $P \in \CorrectPoly$,
    $P_1$ be the sum of \kl{maximal monomials} of $P$,
    and $P_2 \defined P - P_1$ be the sum of
    non-maximal monomials of $P$.
    There exists a computable $K$,
    such that
    $(\Diff{K}{P_1} + P_2) \in \CorrectPoly$.
\end{lemma}

\begin{lemma}
    \label{derivation-translation:lem}
    Let $P \in \Rel[\vec{X}]$, the following are equivalent
    \begin{enumerate}
        \item $P \in \CorrectPoly$,
        \item $P$ is \kl{non-negative}
            and there exists a computable $K \in \Nat$
            such that $\translate{K}(P) \in \Nat[\vec{X}]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    \textbf{TODO: do the proof}.
    To prove \cref{corrected-5:item} $\implies$ \cref{corrected-1:item}, it
    suffices to remark that for every polynomial $P$, for every partial
    valuation $\nu \colon \vec{X} \topartial \Nat$, for every $K \in \Nat$, the
    \kl{maximal monomials} of $\restr{\translate{K}(P)}{\nu}$ are exactly the
    \kl{maximal monomials} of $\restr{P}{\nu}$, the former being
    \kl{non-negative} if $\translate{K}(P) \in \Nat[\vec{X}]$.

    Conversely, the implication \cref{corrected-1:item} $\implies$
    \cref{corrected-5:item} follows immediately from the fact that polynomials
    in $\CorrectPoly$ are \kl{non-negative}, and on the following induction.
    Let $P \in \CorrectPoly$, if $P$ is constant, then $\translate{0}(P) \in \Nat[\vec{X}]$.
    Otherwise, if $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
    monomials} of $P$, there exists $K \in \Nat$ such that $Q \defined
    \Diff{K}{P_1}+ P_2$ belongs to $\CorrectPoly$,
    and by induction hypothesis there exists $L \in
    \Nat$ such that $\translate{L}{Q} \in \Nat[\vec{X}]$. As a consequence,
    $\translate{K+L}(P) = \translate{L}(P_1) + \translate{K}(\translate{L}(Q))$
    is a sum of two polynomials in $\Nat[\vec{X}]$, and we have concluded.
\end{proof}

\begin{lemma}
    \label{lem:correct-to-n-rat}
    Let $\vec{X}$ be a tuple of indeterminates,
    and let $P \in \Rel[\vec{X}]$.
    If $P \in \CorrectPoly$, then $P$ is \kl{represented}
    by a \kl{star-free $\Nat$-polyregular function},
    which can be explicitly constructed from $P$.
\end{lemma}
\begin{proof}
    We prove the result by induction on $\MaximalMonomials(P)$. 
    In the proof, we write $\vec{X}$ for the indeterminates appearing in $P$,
    that is, we assume without loss of generality that all indeterminates are used.

    \textbf{Base case:} If the (unique) \kl{maximal monomial} of $P$ is a
    constant term. Since $P \in \CorrectPoly$, $P = n \in \Nat$, and therefore
    $P$ is \kl{represented} by a constant \kl{star-free $\Nat$-polyregular
    function}.

    \textbf{Induction:} Assume that $P$ is not a constant polynomial, and let
    us write $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
    monomials} of $P$. We compute a bound $K$ such that $Q \defined
    (\Diff{K}{P_1} + P_2) \in \CorrectPoly$ using
    \cref{derivation-stabilises-correct:lem}. Thanks to
    \cref{derivation-simplifies:lemma}, we also know that $\MaximalMonomials(Q)
    \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
    which is effectively computable.

    Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
    (\cref{n-poly-n-poly:example}).
    Furthermore, $\translate{K}(Q)$ is also (effectively)
    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
    (see \cref{translation-invariance:fact}).
    As a consequence, $\translate{K}(P)$ is (effectively) \kl{represented} 
    by a function $f_\Delta$,
    since it is obtained as the following sum:
    \begin{equation*}
        \translate{K}(P) 
        = P_1 + \translate{K}(Q) \quad .
    \end{equation*}

    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$
    fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
    \hoarele \MaximalMonomials{P}$. Because $\restr{P}{\nu} \in \CorrectPoly$,
    one can use the induction hypothesis to compute a \kl{star-free
    $\Nat$-polyregular function} $f_\nu$ that \kl{represents} $\restr{P}{\nu}$.

    Let us assume that the alphabet over which the (\kl{commutative}) functions
    $f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
    representing the indeterminate $X_i$ of the polynomials. Now, let us define
    by case analysis the following \kl{commutative} \kl{star-free
    $\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
    a_1^{X_1} \cdots a_n^{X_n}$, with $X_1, \dots, X_n \geq 0$.

    \begin{equation*}
        f(w) \defined
        \begin{cases}
            f_\nu(w) & \text{ if } \exists i \in \set{1, \dots, n}, X_i \leq K \\
            f_\Delta(a_1^{X_1 - K} \cdots a_n^{X_n - K}) & \text{ otherwise }
        \end{cases}
        \quad .
    \end{equation*}
    It is clear that
    $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function},
    and that
    $f$ \kl{represents} $P$.
\end{proof}

While \cref{lem:correct-to-n-rat} provides effective conversion, it does not
explicitly state that the nembership is decidable. However, it is not too
difficult to remark that $\CorrectPoly$, the proof scheme can be followed to
obtain decidability. For the sake of completeness, we provide a self-contained
proof hereafter, that furthermore

\begin{lemma}
    \label{decidability-correct:lem}
    Membership in $\CorrectPoly$ is decidable for
    polynomials in $\Rel[\vec{X}]$.
\end{lemma}
\begin{proof}
    Let $P \in \Rel[\vec{X}]$. We prove by induction on
    $\MaximalMonomials(P)$ that whether $P \in \CorrectPoly$
    is decidable.

    If $P$ is a constant polynomial, then
    $P \in \CorrectPoly$ if and only if $P = n \in \Nat$
    which is decidable.

    Otherwise, $P$ is not a constant polynomial, and let us write $P = P_1 +
    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. If $P_1
    \not \in \Nat[\vec{X}]$, then $P \not \in \CorrectPoly$, and this is
    decidable. Hence, we can assume that $P_1 \in \Nat[\vec{X}]$. Using
    \cref{derivation-stabilises-correct:lem}, there exists a computable bound
    $K$, such that if $P \in \CorrectPoly$, then t $Q \defined (\Diff{K}{P_1} +
    P_2) \in \CorrectPoly$. Leveraging, \cref{derivation-simplifies:lemma}, we
    also know that $\MaximalMonomials(Q) \hoarele \MaximalMonomials(P)$, and
    therefore that we can decide whether $Q \in \CorrectPoly$. Notice that if
    $Q \not \in \CorrectPoly$, then $P \not \in \CorrectPoly$, hence we can
    assume that $Q \in \CorrectPoly$.

    Similarly, for partial valuations
    $\nu \colon \vec{X} \topartial \set{0, \dots, K}$
    fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
    \hoarele \MaximalMonomials{P}$. 
    We can check for all these partial valuations that
    $\restr{P}{\nu} \in \CorrectPoly$ by induction hypothesis,
    and if one test is not satisfied, then $P \not \in \CorrectPoly$.

    Let us now conclude that if all the above checks answered correctly,
    then $P \in \CorrectPoly$. Let $\nu \colon \vec{X} \topartial \Nat$
    be a partial valuation. Then, 
    one of the three following cases arise:
    \begin{itemize}
        \item 
            Either,
        $\nu$ fixes \emph{some} indeterminate $X_i$ to a value $n_i < K$.
        In that case, let us define $\delta$ as the partial function that coincides
        with $\nu$ on its domain
        $\dom(\delta) = \dom(\nu) \setminus \set{ X_i }$.
        Then, $\restr{P}{\nu} = \restr{\restr{P}{X_i = n_i}}{\delta}$.
        As we previously checked that $\restr{P}{X_i = n_i} \in \CorrectPoly$,
        we conclude in particular that 
        $\restr{P}{\nu}$ has \kl{non-negative} \kl{maximal monomials}.

        \item Or, for all $X_i$ in the (potentially empty) domain of $\nu$, $\nu(X_i)
            \geq K$. In that case, let us write $\delta \defined
            \translate{-K}(\nu)$, that is the partial map $\delta \colon
            \vec{X} \topartial \Nat$ having the same domain as $\nu$, and such
            that for all $X_i \in \dom(\nu)$, $\delta(X_i) \defined \nu(X_i) -
            K$. Recall that $Q = \Diff{K}{P_1} + P_2$, and therefore that
            $\translate{K}(P) = P_1 + \translate{K}(Q)$. By linearity,
            $\restr{\translate{K}(P)}{\delta} = \restr{P_1}{\delta} +
            \restr{\translate{K}(Q)}{\delta}$, and we conclude that the
            \kl{maximal monomials} of $\restr{\translate{K}(P)}{\delta}$ are
            \kl{non-negative}.
            Since $\restr{\translate{K}(P)}{\delta} = \translate{K}{\restr{P}{\nu}}$,
            and since
            $\translate{K}$ leaves \kl{maximal monomials} invariant,
            we conclude that
            the \kl{maximal monomials} of $\restr{P}{\nu}$
            are the same as those of $\restr{\translate{K}(P)}{\delta}$,
            and are therefore \kl{non-negative}.
            \qedhere
    \end{itemize}
\end{proof}

We are now ready to state the corrected and generalized version of
\cref{karh:thm}, which is the main technical contribution of the paper.

\begin{theorem}
    \label{corrected-version:thm}
    Let $P \in \Rel[\vec{X}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item \label{corrected-1:item} $P \in \CorrectPoly$,
        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item \label{corrected-5:item} $P$ is \kl{non-negative}
            and there exists $K \in \Nat$ such that $\translate{K}(P) \in \Nat[\vec{X}]$.
    \end{enumerate}
    Furthermore, the membership is decidable, and effective conversion
    procedures exist between all the representations.
\end{theorem}
\begin{proof}
    The implications 
    \cref{corrected-4:item} $\implies$
    \cref{corrected-3:item} $\implies$
    \cref{corrected-2:item} are obvious.
    \cref{lem:correct-to-n-rat} proves
    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
    while \cref{n-rat-correct:lem}
    proves 
    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
    Note that the lemmas provide effective conversion procedures,
    and that \cref{lem:correct-to-n-rat} also provides a decision
    procedure.

    The equivalence between
    \cref{corrected-5:item} and \cref{corrected-1:item}
    is precisely \cref{derivation-translation:lem}.
\end{proof}

Let us now state, for the sake of a completeness, the equality of
$\CorrectPoly$ and $\CoveredPoly$ in the case of two indeterminates. This is
relevant because it shows that examples given in \cite{KARH77} are actually
correct, even though based on an invalid result, and it may be the case for
other works based on \cite{KARH77}.

\begin{lemma}
    \label{lem:correct-covered-2}
    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
\end{lemma}
\begin{proof}
    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
    For the converse inclusion, let us consider $P(X,Y)$
    that is \kl{non-negative}, such that the \kl{maximal monomials}
    have \kl{positive coefficients}.

    If we fix none of the variables, then the \kl{maximal monomials}
    have \kl{positive coefficients} by assumption. If we fix one of the
    variables, we can assume without loss of generality that we 
    fix $X = k$ for some $k \in \Nat$.
    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
    and therefore must have a positive leading coefficient
    (which is the unique \kl{maximal monomial} in this case)
    or be constant equal to 0. In both cases, the \kl{maximal monomials}
    have \kl{positive coefficients}.
    The same reasoning applies \emph{a fortiori} in the case where
    we fix the two indeterminate, leading to a constant polynomial.
\end{proof}

### 2024-01-16 parts/counter.tex
wohoho

--- 2024-01-11 pnrs.tex
@@
 %! TeX program = xelatex
-%! lang = en-GB
-\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
+%! lang = en-US
+\section{The Counter Example}
+\label{sec:c-example}
 
-% INPUTS
-\input{globals/packages}
-\input{globals/colorscheme}
-
-% CONFIG
-\input{globals/macros}
-\input{globals/mathdefs}
-\input{globals/knowledges.kl}
-\input{globals/metadata}
-
-\usepackage{comment}
-%\includecomment{forshort}
-%\excludecomment{proof}
-
-\makeabstract
-% DOCUMENT
-\begin{document}
-
-% FRONT MATTER
-\maketitle
-\acknowledge
-
-\section{TODOs}
-
-\begin{itemize}
-    \item Use $\Nat$-polyregular functions all the time, and just 
-        in the introduction talk about $\Nat$-rational series.
-    \item Prove that $\Nat$-rational series of polynomial 
-        growth are $\Nat$-polyregular functions, maybe using
-        the results about ambiguity in automata?
-        Theorem 5.22
-    \item Define commutativity by a map 
-        $\commute \colon \Sigma^* \to \Nat^{\Sigma}$.
-    \item Define "commutative functions" and its symbol.
-    \item Define "computed by" (like, recognised by) for commutative 
-        functions.
-    \item Change \cref{corrected-version:thm} to use directly
-        star-free!
-    \item Clear up the computability problems.
-    \item Polynomial + Npolyregular => star free?
-    \item Polynomial + N rational => N polyregular?
-    \item cite the paper of Quantitative Monadic Second-Order Logic
-        about why it is interesting class.
-\end{itemize}
-
-% MAIN MATTER
-\section{Introduction}
-
-There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
-\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
-This computational model, in the counting case, can be traced back to a model
-introduced by \textcite{schutzenberger1965finite}. It has been recently proven
-that one can decide the growth rate of such functions, and decide whether they
-are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. Isolating
-a notion of aperiocidity for functions, and deciding this property is often a
-non-trivial process \cite{droste2019aperiodic}. It was conjectured in the PhD
-thesis of Gaëtan Douéneau-Tabot, that the membership problem of $\Nat$-rational
-series among $\Rel$-rational series was decidable under the assumption of
-polynomial growth. The semantic characterization would imply that
-$\Nat$-rational series of polynomial growth that are aperiodic are precisely
-$\Nat$-rational series of polynomial growth, that can be realised by
-$\Rel$-rational series that are aperiodic. This conjecture was based on the
-previously introduced result of \textcite[Theorem 3.3]{KARH77}, which we will
-disprove in \cref{sec:c-example}. 
-We then provide a corrected version of the theorem, and prove the conjecture
-of Douéneau in the case of \emph{commutative} $\Nat$-rational series of
-polynomial growth.
-
-\section{Preliminaries}
-
-\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a production function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        f(w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-\end{definition}
-
-\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
-    \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be a function. The following are equivalent:
-    \begin{enumerate}
-        \item $f$ can be computed as a \k{$\Nat$-rational series}
-            of \kl{polynomial growth}.
-        \item $f$ is a \kl{$\Nat$-polyregular function}.
-    \end{enumerate}
-\end{fact}
-
-\begin{definition}[Letter Counting Function]
-    \label{commutative-map:def}
-    Let $\Sigma$ be a finite alphabet. We define
-    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
-    map that counts occurrences of every letter in the input word, namely:
-    \begin{equation*}
-        \commute[w] \defined \seqof{\card[a]{w}}{a \in \Sigma} \quad .
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[Commutative Function]
-    \label{commutative:def}
-    Let $X$ be a set.
-    A function $f \colon \Sigma^* \to X$ is
-    \intro{commutative} whenever
-    for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
-
-    Equivalently, it is \reintro{commutative} whenever
-    there exists a map $g \colon \Nat^\Sigma \to X$
-    such that $g \circ \commute = f$.
-\end{definition}
-
-\begin{definition}[Represented Function]
-    A function $g \colon \Nat^n \to \Rel$
-    is \intro{represented} by a function $f \colon \Sigma^* \to \Rel$
-    if there exists a map $\mu \colon \set{1, \dots, n} \to \Sigma$
-    such that
-    $g \circ (v \mapsto \seqof{v_{\mu(k)}}{1 \leq k \leq n}) \circ \commute = f$.
-\end{definition}
-
-We need, $\ZRat$, $\NRat$, $\ZPoly$, $\NPoly$, $\ZSF$, and $\NSF$.
-
-We will start by formally stating the result of \citeauthor{KARH77}
-in terms of polynomials and commutative $\Nat$-rational series, and
-then introduce the necessary definitions. The characterization of
-commutative $\Nat$-rational series computing polynomial functions
-is described in terms of a class $\CoveredPoly[\vec{X}]$ of
-polynomials with multiple indeterminate.
+We will start by formally stating the result of \citeauthor{KARH77} in terms of
+polynomials and commutative $\Nat$-rational series, and then introduce the
+necessary definitions. The characterization of commutative $\Nat$-rational
+series computing polynomial functions is described in terms of a class
+$\CoveredPoly[\vec{X}]$ of polynomials with multiple indeterminate.
 
@@
     that are \kl{non-negative}
-    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
+    and such that every \kl{maximal monomial} is \kl{non-negative}.
     When the indeterminate are clear from the context, we write
@@
 
-We use the term \kl{$\Nat$-rational polynomial}, that will formally be
-stated later on, to denote polynomials that can be computed via a
-$\Nat$-rational series. In the upcoming theorem, this property
-which is \emph{a priori} difficult to decide is reduced to 
-the positivity of the polynomial, and the positivity of \emph{some}
-coefficients.
-
 \begin{faketheorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
     \label{karh:thm}
-    Let $\Sigma$ be a finite alphabet.
-    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
-    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
+    Let $\Sigma$ be a finite alphabet
+    and $P$ be a polynomial. The following are equivalent:
+    \begin{enumerate}
+        \item $P$ is a \kl{$\Nat$-rational polynomial},
+        \item $P \in \CoveredPoly$.
+    \end{enumerate}
 \end{faketheorem}
 
-It is not clearly stated in \cite{KARH77}, but \cref{karh:thm} implies the
-decidability of \kl{$\Nat$-rational polynomial} inside polynomials with $\Rel$
-coefficients. We will not formalize this statement yet, as \cref{karh:thm} does
-not hold in general.
-
-We then provide in \cref{sec:proof} a
-corrected version of the theorem, together with a valid proof. As a
-consequence, it does not invalidate the stated conjecture.
-
-\begin{itemize}
-    \item commutative input and output
-    \item commutative output
-    \item no commutativity ?
-\end{itemize}
-
-\AP \paragraph*{Rational series and regular languages.} This paper uses notions
-of regularity and results about \kl{$\Nat$-rational series}, however most of
-the technicalities will stem from the multiple indeterminate of polynomials,
-and all there is to know about regularity in this document will be confined to
-the following facts. For a survey about (non-commutative) rational series, we
-refer the reader to \cite{berstel2011noncommutative}.
-
-\begin{definition}[Rational series]
-    The following are equivalent
-    \begin{itemize}
-        \item weighted automata
-        \item rational expressions
-        \item 
-    \end{itemize}
-\end{definition}
-
-\begin{fact}[Folklore about regular languages]
-    \label{regular:fact}
-    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular, whenever $a,b \in \Sigma$.
-    Regular languages are closed under intersection.
-\end{fact}
-
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by a $\Nat$-rational series
-    is a regular language. In particular,
-    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
-    language.
-\end{fact}
-
-\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational polynomial} if
-there exists a \reintro{commutative} $\Rel$-rational series $f$ such that $f(w)
-= P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, $P$ is a
-\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
-series is in fact $\Nat$-rational. Beware that we only apply the polynomial
-to \emph{non-negative} values.
-
-\begin{example}
-    \label{negative-not-nrat:ex}
-    With $\Sigma \defined \set{1}$,
-    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
-    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    The function $w \mapsto |w|$ is a \kl{polyregular function} with unary
-    output of \kl{polynomial growth}, hence  can be computed by a
-    \kl{$\Nat$-rational series}. Hence, $P(X_1) \defined X_1$ is
-    a \kl{$\Nat$-rational polynomial}. Similarly,
-    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-rational series},
-    showing that $Q(X_1) \defined X_1^2 + 3$
-    is a \kl{$\Nat$-rational polynomial}.
-    Finally, 
-    $T(X_1) \defined - X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
-    have non-negative output.
-\end{proof}
-
-We saw in \cref{negative-not-nrat:ex}
-a simple criterion to check whether a polynomial is $\Nat$-rational,
-but being non-negative is not enough, as the following example
-illustrates. Note that the proof scheme
-of this example will be at the core of \cref{thm:counter-example}.
-
-\begin{example}
-    Over a binary alphabet $\Sigma \defined \set{a,b}$,
-    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
-    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
-    but is
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Assume by contradiction that
-    $S$ is a $\Nat$-rational series computing $P$.
-    Then, $S^{-1}(\set{0})$ should be a regular language
-    (\cref{pre-image-regular:fact}),
-    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not a regular language (\cref{regular:fact}).
-\end{proof}
-
-\AP Let us now give some vocabulary on polynomials with multiple indeterminate
-over $\Rel$. A \intro{monomial} is a product of variables and integers. For
-instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, but $X + Y$ or
-$2X^2 + XY$ are not.
-
-\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} if
-$P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we do
-not consider negative values, as the numbers $n_i$ will ultimately count the
-number of occurrences of a letter in a word. As an example, the polynomial $(X
-- Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$, but the
-polynomial $X^2 - 2X$ is not.
-
-\AP
-In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
-is a \intro{maximal monomial} if it is a maximal element
-for the visibility ordering of polynomials \emph{in $\Rat$}
-among monomials with non-zero coefficients in $P$.\footnote{
-    One could equivalently define maximal elements by removing the (non-zero)
-    multiplicative constant of the monomials, and then
-    considering the usual divisibility relation of $\Rel$ polynomials.
-}
-In the polynomial 
-$P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
-are $X^2$, $-2 XY$, and $X^2$.
-A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$ if its
-multiplicative constant is positive. In the polynomial $P(X,Y) \defined (X -
-Y)^2$, the \kl{positive monomials} are $X^2$ and $Y^2$.
-
-\paragraph*{Outline of the paper.}
-todo.
-
-\section{The Counter Example}
-\label{sec:c-example}
-
-Let us now introduce the counter example that was found trying to generalise
+Let us now introduce the counter example that was found trying to generalize
 the original proof to non-commutative \kl{$\Rel$-rational series} of
@@
 
-\begin{definition}[Counter Example Polynomial]
+\begin{definition}
     \label{def:bad-polynomial}
-    We define $P(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
+    We define $\BadPoly(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
 \end{definition}
 
-\begin{theorem}
+\begin{lemma}
     \label{thm:counter-example}
-    Over $\Sigma \defined \set{a,b,c}$,
-    there exists a polynomial $P \in \CoveredPoly$ that is not
-    a \kl{$\Nat$-rational polynomial}. Namely,
-    $P$ is the polynomial of \cref{def:bad-polynomial}.
-\end{theorem}
+    The polynomial $\BadPoly$ belongs to $\CoveredPoly$,
+    but is not a \kl{$\Nat$-rational polynomial}.
+\end{lemma}
 \begin{proof}
-    Let us write $X,Y,Z$ instead of $X_a, X_b, X_c$ to improve
-    readability.
-    It is clear that $P$ is \kl{non-negative}. We can develop
-    the expression of $P$ to 
+    It is clear that $\BadPoly$ is \kl{non-negative} as a sum of two 
+    squares. We can develop
+    the expression of $\BadPoly$ to 
     obtain
-    $P = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
+    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
     The \kl{maximal monomials} of $P$
@@
 
-    Assume by contradiction that $P$ is a \kl{$\Nat$-rational polynomial}.
+    Assume by contradiction that $\BadPoly$ is a \kl{$\Nat$-rational polynomial}.
+    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
     There exists a \kl{commutative}
-    $\Nat$-rational series $f \colon \Sigma \to \Nat$
-    such that $P(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-    Remark that $P(X,Y,Z) = 0$
+    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
+    such that for all $w \in \Sigma^*$,
+    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
+
+    Remark that $\BadPoly(X,Y,Z) = 0$
     if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-    $P(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
+    $\BadPoly(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
     $Z \neq 0$, and $X = Y = 0$.
@@
     However, $L = f^{-1}(\set{0})$, which is a regular language
-    because $f$ is a $\Nat$-rational series
+    because $f$ is a \kl{$\Nat$-rational series}
     (\cref{pre-image-regular:fact}).
@@
 
-As we will see in the next section, the characterization
-of \kl{$\Nat$-rational polynomials} using the set $\CoveredPoly$
-holds when the polynomials have at most $2$ indeterminate. This shows
-that the examples in \cite{KARH77} are not invalidated, as they
-all use at most two indeterminate.
-
-\section{The Corrected Theorem}
-\label{sec:proof}
-
-Let us now provide another class of polynomials, $\CorrectPoly$, that restricts
-$\CoveredPoly$ to a subset that is ``closed under fixing indeterminate", namely,
-polynomials $P(X_1, \dots, X_n) \in \CoveredPoly$ such that partially evaluated
-polynomials remain in $\CoveredPoly$. We use the following notation
-to fix the value of some indeterminate
-$\intro*\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally, if $\nu$ is
-a partial function from $\vec{X}$ to $\Nat$, the restriction
-$\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminate $\vec{Y}
-\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the domain of
-$\nu$.
-
-\begin{definition}
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CorrectPoly[\vec{X}]$ is the collection of
-    polynomials $P \in \Rel[\vec{X}]$ such that
-    $P$ is 
-    such that, for every partial function $\nu \colon \vec{X} \to \Nat$,
-    every \kl{maximal monomial} of
-    $\restr{P}{\nu}$ has a \kl{positive coefficient}.
-\end{definition}
-
-First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
-polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
-counter example provided in \cref{thm:counter-example} is not in
-$\CorrectPoly$. For that, notice that for $P$ defined in
-\cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$. Let us also assert that
-polynomials that have all coefficients in $\Nat$ are \kl{$\Nat$-rational
-polynomials}, which will be a base case in an upcoming induction.
-
-\begin{remark}
-    \label{inductive-correct-poly:remark}
-    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item The \kl{maximal monomials} of 
-            $P$ have \kl{positive coefficients},
-            and 
-            for all $n \in \Nat$,
-            $\restr{P}{X = n} \in \CorrectPoly$.
-    \end{enumerate}
-    Furthermore, if $P \in \Rel[]$ is a polynomial without
-    indeterminate, then $P \in \CorrectPoly$ if and only
-    if its \kl{maximal monomial} (the constant term)
-    has a \kl{positive coefficient}.
-\end{remark}
-\begin{proof}
-    By induction on the number of indeterminate,
-    leveraging the fact that if there exists at least one indeterminate
-    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
-    and partial function $\nu \colon \vec{Y} \to \Nat$,
-    we have 
-    \begin{equation*}
-        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
-        \quad
-        ,
-    \end{equation*}
-    where $\nu[X = n]$ is the partial function from
-    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
-    and maps $X$ to $n$.
-\end{proof}
-
-\begin{example}
-    \label{n-poly-n-poly:example}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
-    be a partial function.
-    Notice that $\restr{P}{\nu}$ continues to be
-    a polynomial with \kl{positive coefficients}, and in particular
-    with 
-    \kl{maximal monomials} having \kl{positive coefficients}.
-\end{proof}
-
-Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-To simplify the reasoning, we will use results on $\Nat$-polyregular functions,
-also known as, polyregular functions with unary output
-\cite{doueneau2021pebble,bojanczyk2019string}. Note that the connection between
-rational series and polyregular functions is not specific to $\Nat$, as
-$\Rel$-rational series of polynomial growth are computed as polyregular
-functions with output in $\set{+1,-1}$ post-composed with computing the sum of
-outputted letters \cite{LOPEZ23b}. Because we only consider functions with
-output in $\Rel$, and that a commutative output monoid simplifies the definitions,
-let us only define $\Rel$-polyregular functions.
-
-\begin{lemma}
-    \label{n-poly-combinatorics:lem}
-    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
-    There exists $\omega \in \Nat$
-    such that for all $p \in \Nat$,
-    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
-    for all $u_1, \dots, u_p \in \Sigma^*$,
-    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
-    whose \kl{maximal monomials} have \kl{positive coefficients},
-    and such that:
-    \begin{equation*}
-        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
-        = P(X_1, \dots, X_p) \quad .
-    \end{equation*}
-\end{lemma}
-\begin{proof}[Proof Sketch]
-    Let $\omega$ be an idempotent power for the finite monoid $M$.
-    We define an equivalence between partitions of a word 
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$
-    as follows: two partitions into $k$ words in $\Sigma^*$ are equivalent
-    when normalizing them with the rules
-    $u_i^{\omega} u_i^{\omega} \to u_i^{\omega}$ for all $1 \leq i \leq p$.
-    This equivalence relation has finite index, and all partitions in the same
-    equivalence class yield the same value through $\pi$, because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid $M$.
-
-    Now, let us remark that the number of partitions
-    in a certain equivalence class is proportional to
-    the number of nanania, that is, is of the form
-    $(X_i - K_i)^{n_i}$ for some $K$. We conclude that
-    maximal monomials indeed have positive coefficients.
-    We even conclude that
-    the polynomial is in $\Nat[\vec{X}]$
-    if we shift the variables with a large enough $K$. 
-\end{proof}
-
-\begin{lemma}
-    \label{n-rat-correct:lem}
-    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-\end{lemma}
-\begin{proof}[Proof Idea]
-    Let $P$ be a \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma$  be the finite alphabet such that
-    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
-    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
-    $\Nat$-rational series computing $P$.
-    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
-    it is a $\Nat$-polyregular function 
-    (\cref{polynomial-rational-polyreg:fact}).
-
-    Using \cref{n-poly-combinatorics:lem},
-    we conclude that maximal monomials of $P$ have
-    positive coefficients. Furthermore, when fixing variables
-    in $P$, we are computing a new \kl{$\Nat$-polyregular function},
-    hence we also conclude that the maximal monomials have
-    positive coefficients.
-    We have proven that $P \in \CorrectPoly$.
-\end{proof}
-
-The core of the upcoming proof of \cref{lem:correct-to-n-rat} relies on the
-following observation: if a polynomial $P$ is written $X^p + Q$, where all
-terms of $Q$ have degree less than $p$, then one can find some large enough $K$
-so that $X^p - (X - K)^p + Q$ also has a positive maximal coefficient.
-It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence
-the maximal coefficient can be chosen arbitrarily large by increasing
-$K$.
-This
-trick will work similarly in multivariate polynomials, and the overall
-proof scheme works as follows: for a bounded number of possible
-values of $X$, we can fix the variable, and have one less item to work on.
-For large enough values of $X$ (typically greater than some $K$), we can guarantee that
-$X^p - (X - K)^p + Q$ has positive maximal coefficients, hence
-can be computed by a $\Nat$-rational series, while
-$(X - K)^p$ is clearly computable by a $\Nat$-rational series,
-hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
-as the sum of two $\Nat$-rational series, and we have
-concluded that our original polynomial was a
-\kl{$\Nat$-rational polynomial}. There are a number of problems in generalizing
-this proof scheme:
-\begin{itemize}
-    \item We will have to guarantee that every polynomial considered
-        is non-negative
-    \item We do not only need to check that the \kl{maximal monomials}
-        have \kl{positive coefficient}, but that it is the case
-        when fixing arbitrarily many variables! (this
-        covers the first point too)
-\end{itemize}
-
-\begin{lemma}
-    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
-    $\vec{p} \colon \vec{X}Y \to \Nat$,
-    and $M \defined \times \prod_{X \in \vec{X}} X^{p(X)}$.
-    Then, for all $K \geq 0$,
-    $\Delta_{Y,K} \defined Y^p M - (Y - K)^p M$ has \kl{maximal monomials} with
-    \kl{positive coefficients} of weight at least $K$, regardless of
-    the variables fixed, as long as the (potential)
-    value of $Y$ is greater
-    than $K$.
-\end{lemma}
-\begin{proof}
-    Let us first remark that
-    \begin{equation*}
-        \Delta_{Y,K}
-        = KY^{p-1} M + Q M
-        \quad .
-    \end{equation*}
-    In particular, there is a unique \kl{maximal monomial} for $\Delta_{Y,K}$,
-    which is precisely $K Y^{p-1} M$, because all monomials
-    in $Q M$ divide (strictly) $Y^{p-1} M$.
-    Let us fix a variable,
-    $X$ to any (non-zero) value $p > 0$. Then
-    The \kl{maximal monomial} of $\restr{\Delta_{Y,K}}{X = p}$
-    is exactly $\restr{KY^{p-1}M}{X = p}$, which remains
-    with a \kl{positive coefficient}, that is proportional to $K$.
-    Let us fix $Y$ to some value $K + l$, with $l > 0$,
-    then
-    $\restr{\Delta_{Y,K}}{Y = K + l}
-    = (K + l)^p M - l^p M
-    = ((K + l) - l) (\sum_{i = 1}^{p-1} (K + l)^i l^{p-i}) M
-    = K \alpha M
-    $
-    which remains a monomial with a \kl{positive coefficient},
-    that is proportional to $K$.
-\end{proof}
-
-We can refine the above lemma to use it in the case where
-we do not have a single
-monomial.
-
-\begin{lemma}
-    \label{lem:delta-cool}
-    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
-    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
-    \kl{maximal monomial} of $P$, with $\alpha > 0$.
-    Then, there exists a computable $K \geq 0$
-    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
-    has \kl{positive coefficients} for \kl{maximal monomials}
-    in any partial evaluation that satisfies $Y > K$ (if it
-    gives a value to $Y$).
-\end{lemma}
-\begin{proof}
-    Let $N$ be greater than the sum of the absolute values of all coefficients
-    appearing in $P$, and let $K = 2 N$.
-
-    Let us remark that
-    \begin{equation*}
-        \Delta_K = \underbrace{(P - \alpha Y^p M)}_{\text{ left hand side }}
-        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{\text{ right hand side }}
-        \quad .
-    \end{equation*}
-    an that for every partial evaluation $\nu$ that sends $Y$ to a value
-    greater than $K$,
-    we have that $\restr{(Y^p M - (Y - K)^p M)}{\nu}$ has
-    \kl{maximal monomials} with \kl{positive coefficients}
-    proportional to $K$ (note that here ``proportional"
-    implies that they are greater than $K$, because we multiply by positive
-    integers).
-
-    To conclude, it is just a game of describing how
-    \kl{maximal monomials} of $\restr{\Delta_Y}{\nu}$
-    can appear. If the \kl{maximal monomial} is obtained
-    by summing an element of the left hand side with an element
-    of the right hand side, it needs to be a \kl{maximal monomial}
-    of the right hand side, and the large value of $K$ guarantees
-    that the sum of coefficients is positive.
-    If it only appears on the right hand side, it is even better.
-    If it only appears on the left hand side, then
-    in particular it means that the monomial does not
-    divide $\restr{Y^p M}{\nu}$, hence has the same value
-    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
-    which is guaranteed to be positive because $P \in \CorrectPoly$.
-
-    In particular, we conclude from this that
-    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
-\end{proof}
-
-Note that we cannot conclude in the above lemma that the result is in
-$\CorrectPoly$, because we have a restriction on the range where the variable
-$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
-induction on the number $d$ of indeterminate, and the degrees of the
-\kl{maximal monomials}. To make the induction formal, let us define
-$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
-$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
-$\Nat^{\vec{X}}$ with the product ordering, which is a
-\kl{well-quasi-ordering}.
-
-\AP To a polynomial $P$, we assign the set $\MaximalMonomials(P)$ of its
-\kl{maximal monomials}, belonging to $\Pfin(\Monomials)$, which remains a
-\kl{well-quasi-ordering} when endowed with the \intro{Hoare ordering} of
-subsets, that is, $S_1 \hoareleq S_2$ whenever for all \kl{monomial} $M_1 \in
-S_1$, there exists a monomial $M_2 \in S_2$, such that $M_1$ \kl{divides}
-$M_2$.
-
-\begin{lemma}
-    \label{lem:correct-to-n-rat}
-    Let $P$ be a polynomial in $\CorrectPoly$.
-    Then, 
-    $P$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on $\MaximalMonomials(P)$.
-
-    \textbf{Base case:} If the \kl{maximal monomial} of $P$ is a constant term,
-    then because $P \in \CorrectPoly$, $P = n \in \Nat$, which is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-
-    \textbf{Induction:}
-    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
-    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
-    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
-    Using \cref{lem:delta-cool},
-    we know that there exists a computable $K$ such that
-    the polynomial
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$
-    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
-    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
-    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
-    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
-    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
-    \hoarele \MaximalMonomials(P)$.
-    By induction hypothesis,
-    we conclude that
-    $\Diff{K}(Y - K)$ is \kl{represented}
-    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
-    \begin{equation*}
-        f_1(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \Diff{K} & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $\alpha \geq 0$, it is an easy check
-    that the following function is also a $\Nat$-rational series
-    \begin{equation*}
-        f_2(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \alpha(Y - K)^pM & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $P \in \CorrectPoly$, for every $n \leq K$, $\restr{P}{Y = n} \in
-    \CorrectPoly$ (\cref{inductive-correct-poly:remark}).
-    Furthermore,
-    $\MaximalMonomials(\restr{P}{Y = n})
-    \hoarele \MaximalMonomials(P)$. Hence, by induction hypothesis,
-    the following function is \kl{represented} by a
-    \kl{star-free $\Nat$-rational series}:
-    \begin{equation*}
-        f_3(Y, \vec{X}) \defined
-        \begin{cases}
-            P(Y, \vec{X}) & \text{ if } Y \leq K \\
-            0 & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    We conclude by remarking that 
-    $P = f_1 + f_2 + f_3$.
-\end{proof}
-
-We are now ready to state the corrected and generalised version of
-\cref{karh:thm}.
-
-\begin{theorem}
-    \label{corrected-version:thm}
-    Let $P \in \Rel[\vec{X}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item \label{corrected-1:item} $P \in \CorrectPoly$,
-        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
-        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-    \end{enumerate}
-\end{theorem}
-\begin{proof}
-    The implications 
-    \cref{corrected-4:item} $\implies$
-    \cref{corrected-3:item} $\implies$
-    \cref{corrected-2:item} are obvious.
-    \cref{lem:correct-to-n-rat} proves
-    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
-    while \cref{n-rat-correct:lem}
-    proves 
-    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
-\end{proof}
-
-\begin{lemma}
-    \label{lem:correct-covered-2}
-    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
-\end{lemma}
-\begin{proof}
-    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \to \Nat$.
-    For the converse inclusion, let us consider $P(X,Y)$
-    that is \kl{non-negative}, such that the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-
-    If we fix none of the variables, then the \kl{maximal monomials}
-    have \kl{positive coefficients} by assumption. If we fix one of the
-    variables, we can assume without loss of generality that we 
-    fix $X = k$ for some $k \in \Nat$.
-    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
-    and therefore must have a positive leading coefficient
-    (which is the unique \kl{maximal monomial} in this case)
-    or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-    The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminate, leading to a constant polynomial.
-\end{proof}
-
-\section{Effective procedures}
-\label{sec:deciding}
-
-The goal of this section is to go beyond polynomials and characterize all
-$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
-that they are \kl{commutative}.
-
-\begin{theorem}
-    \label{decidable-n-poly:thm}
-    Given a \kl{commutative}
-    \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to \Rel$,
-    one can decide if $f$ can be computed by a \kl{$\Nat$-polyregular function},
-    and effectively compute a representation of $f$.
-\end{theorem}
-
-Before proving \cref{decidable-n-poly:thm}, let us first examine the
-effectiveness of \cref{corrected-version:thm}, namely, provide an algorithm to
-decide whether a polynomial $P$ belongs to $\CorrectPoly$. The idea will be to
-compute a bound to values taken by the valuations $\nu$ appearing in the
-definition of $\CorrectPoly$. As a \emph{mise en jambes}, let us consider
-separately the case of one indeterminate.
-
-\begin{lemma}
-    \label{decide-correct-univariate:lemma}
-    Let $P \in \Rel[X]$, there exists a computable $N \in \Nat$,
-    such that the following are equivalent:
-    \begin{enumerate}
-        \item \label{decide-correct-univariate-1:item}
-            $P \in \CorrectPoly$,
-        \item \label{decide-correct-univariate-2:item}
-            The \kl{maximal monomial}
-            of $P$ has a \kl{positive coefficient},
-            and $P(n) \in \Nat$
-            for all $0 \leq n \leq N_0$.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    Let us first remark that the implication 
-    \cref{decide-correct-univariate-1:item}
-    $\implies$ \cref{decide-correct-univariate-2:item} is trivial for every
-    $N \in \Nat$.
-    For the converse implication,
-    let $P \in \Rel[X]$.
-    If $P$ is a constant polynomial, then $P \in \CorrectPoly$
-    if and only if $P(0) \in \Nat$, and $N \defined 0$
-    suffices.
-    Otherwise, $P$ is non-constant, and let $n \geq 1$ be the \kl{degree} of $P$.
-    Let $P_1$ and $\alpha > 0$ be such that
-    $P \defined P_1 + \alpha X^n$. 
-    To prove that $P \in \CorrectPoly$, it suffices to show that
-    $P(k) \in \Nat$ for all $k \in \Nat$.
-    Let us define $\beta$ as the
-    maximal absolute value of coefficients appearing in $P_1$, which is
-    assumed to be non-zero,
-    and $N_0 \defined \beta + 1 \geq 2$.
-    Remark that for $X \geq N_0$,
-    \begin{align*}
-        P(X) &\geq X^n - \beta \sum_{i = 0}^{n-1} X^i \\
-             &= X^n - \beta \frac{X^n - 1}{X - 1} \\
-             &= \frac{X^{n+1} - X^n - \beta X^n + \beta}{X - 1} \\
-             &= \frac{X^n(X - (\beta + 1)) + \beta}{X - 1} \\
-             &\geq 0
-    \end{align*}
-    This proves that $P(k) \in \Nat$ for all $k \geq N_0$,
-    and because $P$ satisfies \cref{decide-correct-univariate-2:item},
-    $P(k) \in \Nat$ for all $1 \leq k \leq N_0$. We have proven 
-    that $P \in \CorrectPoly$.
-\end{proof}
-
-In the case of multiple indeterminate, we leverage
-\cref{lem:delta-cool} to perform an induction on the degree
-of the polynomials and the number of variables of the polynomial.
-
-\begin{lemma}
-    \label{decide-correct-poly:lem}
-    One can decide if a polynomial belongs to $\CorrectPoly$.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on the number of indeterminate.
-
-    \textbf{Base Case:} If $P$ has zero indeterminate, i.e., is a constant
-    function, then to decide whether $P \in \CorrectPoly$ it suffices to check
-    that $P(0) \in \Nat$.
-
-    \textbf{Induction:} Let $P \in \Rel[Y, \vec{X}]$, we start by checking
-    whether \kl{maximal monomials} of $P$ have \kl{positive coefficients}.
-    Without loss of generality, assume that $Y$ divides some \kl{maximal
-    monomial} $\alpha Y^p M$ with $p \geq 1$. By applying
-    \cref{lem:delta-cool}, we can compute a bound $K \in \Nat$ such that
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$ has \kl{positive coefficients}
-    for its \kl{maximal monomials} for all partial valuations $\nu \colon
-    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
-    particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to
-    $\CorrectPoly$ for all $n > K$. By induction hypothesis, one can
-    decide whether $\restr{P}{X = n}$ belongs to $\CorrectPoly$
-    for all $0 \leq n \leq K$, which is equivalent 
-    to deciding whether
-    $P \in \CorrectPoly$ thanks to \cref{inductive-correct-poly:remark}.
-\end{proof}
-
-To prove the theorem, let us first provide a nice characterization of
-\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
-This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
-commutative case.
-
-\begin{lemma}
-    \label{decompose-polynomial:lem}
-    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
-    \kl{$\Rel$-polyregular function}. There exists $\omega \in \Nat$,
-    and polynomials $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{ 0 \leq n_a < \omega}$,
-    such that for all $w \in \Sigma^*$,
-    \begin{equation*}
-        f(w) = P_{\seqof{\card[a]{w} [\omega]}{a \in \Sigma}}
-        (\seqof{\card[a]{w} / \omega}{a \in \Sigma}) \quad .
-    \end{equation*}
-    Which informally means that for every choice of rest modulo $\omega$
-    for every variable, there exists a polynomial describing the
-    behavior of $f$ when adding multiples of $\omega$ letters at a time.
-
-    Furthermore, the polynomials are computable from $f$.
-\end{lemma}
-\begin{proof}
-    equivalent partitions of the word, but commutativity?.q
-    Lemma 5.37 phd gaëtan.
-\end{proof}
-
-As a consequence, we conclude 
-\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
-    We use \cref{decompose-polynomial:lem}
-    to effectively compute
-    a polynomial representation of $f$. Remark that
-    $f$ is a \kl{$\Nat$-polyregular function} if and only if
-    all these polynomials are \kl{$\Nat$-rational polynomials},
-    which is decidable thanks to
-    \cref{decide-correct-poly:lem,corrected-version:thm}.
-\end{proof}
-
-\subsection{Star-free $\Nat$-polyregular functions}
-
-Let us now prove that the above characterizations of
-commutative
-$\Nat$-polyregular functions can be combined with the recent
-advances in the study of $\Rel$-polyregular functions 
-\cite{LOPEZ23b}
-to prove the following theorem.
-
-\begin{theorem}
-    \label{zsf-npoly-nsf:thm}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be both a \kl{star-free $\Rel$-polyregular function}
-    and a \kl{commutative} \kl{$\Nat$-polyregular function}.
-    Then, $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}.
-    That is, the following equality holds:
-    \begin{equation*}
-        \ZCommut \cap \ZSF \cap \NPoly
-        = \ZCommut \cap \NSF
-        \quad .
-    \end{equation*}
-\end{theorem}
-
-Note that this proves in the commutative case a conjecture that was 
-formulated by Thesis Gaetan, and that is non trivial, because
-
-\begin{conjecture}
-    \label{nsf-zsf:conj}
-    In the general case, the following equation holds:
-    \begin{equation*}
-        \ZSF \cap \NPoly
-        = 
-        \NSF
-        \quad .
-    \end{equation*}
-\end{conjecture}
-
-As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
-procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
-decidable inside $\NPoly$ under the assumption of commutativity.
-
-\begin{corollary}
-    One can decide if a \kl{commutative} \kl{ $\Rel$-polyregular function}
-    can be realized by a \kl{star-free $\Nat$-polyregular function},
-    in which case the conversion is effective.
-\end{corollary}
-
-The key ingredient of this section is the use of a semantic characterization of
-\kl{star-free $\Rel$-polyregular functions} among \kl{$\Rel$-rational series}
-that generalizes the aperiodicity of languages to functions by the means of
-polynomial behaviors.
-
-\begin{definition}[Ultimately polynomial]
-    \label{ultimately-polynomial:def}
-    Let $\Sigma$ be a finite alphabet. 
-    A function $f \colon \Sigma^* \to \Rel$
-    is \intro{ultimately polynomial}
-    whenever there exists $N_0 \in \Nat$ such that
-    for all $n \in \Nat$
-    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
-    \in \Sigma^*$, there exists a polynomial $P \in \Rel[X_1, \dots, X_n]$
-    such that
-    \begin{equation*}
-        f\left(
-            \alpha_0 \prod_{i = 1}^{n} w_i^{X_i} \alpha_i
-        \right)
-        = 
-        P(X_1, \dots, X_n)
-        \quad 
-        \forall X_1, \dots, X_n \geq N_0
-        \quad .
-    \end{equation*}
-\end{definition}
-
-Let us recall from \cite{LOPEZ23b} that this definition correctly
-generalizes the indicator functions of regular languages.
-
-\begin{example}
-    A language $L$ is aperiodic if and only if 
-    the indicator function is ultimately polynomial.
-\end{example}
-\begin{proof}
-    TODO.
-\end{proof}
-
-\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
-    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
-
-    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
-    only the empty word $\varepsilon$, and because $f$ is a
-    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
-    Note that $f$ is therefore a constant function, equal to a natural
-    number, hence belongs to $\NSF$.
-
-    \textbf{Induction.}
-    Because $f$ is \kl{ultimately polynomial},
-    there exists $N_0 \in \Nat$, 
-    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
-    such that
-    \begin{equation*}
-        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
-        = 
-        P(\seqof{X_a}{a \in \Sigma})
-        \quad
-        \text{ when }
-        \forall a \in \Sigma, X_a \geq N_0
-        \quad .
-    \end{equation*}
-    Furthermore, leveraging \cref{decompose-polynomial:lem},
-    we obtain $\omega \in \Nat$
-    and a family of polynomials 
-    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
-    such that.
-    Because each of those polynomials coincide with $P$ over an infinite
-    subgrid of $\Nat$,
-    we conclude that all polynomials are in fact equal to a single one.
-    we conclude that the polynomials 
-    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.
-
-    Now, we conclude in particular that the function 
-    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
-    hence is computable by a \kl{star-free $\Nat$-polyregular function},
-    because of \cref{nsf-polynomials:thm}.
-    When fixing values of some indeterminate below $N_0$,
-    we obtain a new function $g$, that remains commutative,
-    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
-    By induction hypothesis, this function $g$ is also 
-    \kl{star-free $\Nat$-polyregular}, and we conclude
-    that 
-    $f = P + \sum g$ which remains star free.
-\end{proof}
-
-\section{Conclusion}
-\label{sec:ccl}
-
-\begin{itemize}
-    \item Actually, we have an algorithm that allows
-        us to \emph{decide} if a polynomial is $\Nat$-rational.
-    \item Actually, we obtain a characterisation of
-        commutative 
-        $\Nat$-rational series of polynomial growth
-        among $\Rel$-rational series.
-    \item The question of $\Nat$-polyregular functions
-        among $\Rel$-polyregular functions remains open
-        because of the non-commutativity of the input.
-    \item The question of $\Nat$-rational series (commutative)
-        among $\Rel$-rational series remains open.
-\end{itemize}
-
-\begin{corollary}
-    The following problem is decidable: given $P \in \Rel[\vec{X}]$,
-    decide whether $P$ is a \kl{$\Nat$-rational polynomial}.
-\end{corollary}
-\begin{proof}
-    Thanks to \cref{karh:thm}, one only has to decide whether
-    $P \in \CoveredPoly$. The procedure is as follows.
-    \begin{enumerate}
-        \item Check if all \kl{maximal monomials} have
-            \kl{positive coefficients} in $P$, if not, answer \texttt{false}.
-        \item Consider $N \in \Nat$ large enough so that
-              $P \geq 0$ when all indeterminate are chosen greater
-              than $N$. This is possible because all \kl{maximal
-              monomials} have \kl{positive coefficients}, and such an $N$
-              is computable.
-        \item Compute $P$
-    \end{enumerate}
-
-    Now, if all \kl{maximal monomials}
-    have \kl{positive coefficients}, then there exists a computable
-    $N \in \Nat$ such that 
-    In particular,
-
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly = \setof{P}{ P \geq 0 \wedge \exists K \in \Nat, P \circ (+K) \in \Nat[\vec{X}]}$.
-\end{conjecture}
-\begin{proof}[Proof Sketch]
-    If $S \defined P \circ (+K) \in \Nat[\vec{X}]$, then for every partial application,
-    maximal coefficients of $S$ are positive. As maximal coefficients of 
-    (partial applications of) $S$ are the same as the ones of $P$
-    \emph{except for the constant terms} (full application),
-    where $P$ is already guaranteed to be non-negative, we conclude
-    that $P \in \CorrectPoly$.
-
-    Conversely, we prove by induction on the degree of $P \in \CorrectPoly$,
-    that there exists some $K$ such that $P \circ (+K) \in \Nat[\vec{X}]$.
-    First, notice that $\CorrectPoly$ is closed under pre-composition with
-    $(+K)$. Then, remark that $P(X - K)$ has the same maximal coefficients as
-    $P$ and new maximal coefficients that are multiples of $K$. Hence, for a
-    large enough $K$, we conclude that $P(X - K)$ is the sum of its maximal
-    coefficients, plus a polynomial that lies in $\CorrectPoly$. We conclude by
-    induction hypothesis. 
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly$ is precisely composed of polynomials that
-    have non-negative coefficients in a \emph{binomial basis},
-    i.e., where polynomials are obtained
-    as sums and products of $\binom{X_i}{p_i}$.
-\end{conjecture}
-
-\begin{conjecture}
-    star-free $\Nat$-rational is star-free and $\Nat$-rational?
-\end{conjecture}
-
-\begin{center}
-    \begin{tabular}{c|cc}
-        \toprule
-        & \textbf{Commutative} & \textbf{General} \\
-        \midrule
-        \textbf{Commutative}
-        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
-        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
-        \textbf{General} & ? & polyregular 
-    \end{tabular}
-\end{center}
-
-% BACKMATTER
-\printbibliography
-
-\appendix
-
-\end{document}
+As we will see in the next section, the characterization of \kl{$\Nat$-rational
+polynomials} using the set $\CoveredPoly$ holds when the polynomials have at
+most $2$ indeterminate. This shows that the examples in \cite{KARH77} are not
+invalidated, as they all use at most two indeterminates. Furthermore, the
+converse implication, that is, that \kl{$\Nat$-rational polynomials} belong to
+$\CoveredPoly$ is correct, and can even be strengthened as we will see in
+\cref{corrected-version:thm}.

### 2024-01-16 parts/intro.tex
wohoho

%! TeX program = xelatex
%! lang = en-US
% MAIN MATTER
\section{Introduction}
\label{introduction:sec}

Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free automata \cite{mcnaughton1971counter}, represented by
star-free regular expressions \cite{schutzenberger1965finite}, definable in
First Order Logic ($\FO$) \cite{perrin1986first}, and recognized by aperiodic
monoids \cite{schutzenberger1965finite}. It is decidable whether a regular
language is aperiodic, thanks to the characterizations in terms of minimal
automaton and/or syntactic monoid.

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}) \cite[see,
e.g.][]{bojanczyk2019string}. In this paper, we will focus on \kl{polyregular
functions}, that shares with regular languages the characterizations in terms
of logic (via $\MSO$ interpretations), automata (via pebble automata or SST
\cite{}), and adds new bridges with programming languages (lambda-terms)
for-transducers. Although the class of "star-free" polyregular functions is
defined and similarly characterized using first-order logic.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, it aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}.

Another way to approach the problem of deciding star-free for polyregular
functions is to not restrict the computational power, nor change the semantics,
but consider restrictions of the output alphabet. In particular, there have
been successful results proven on \emph{unary polyregular functions}, also
known as \kl{$\Nat$-polyregular functions}
\cite{doueneau2021pebble,doueneau2022hiding}. Let us note that results on
\emph{unary output functions} often generalize to the slightly more general
class of functions with \emph{commutative output}, and that algorithms leverage
the absence of combinatorial information in the output. Furthermore,
\kl{$\Nat$-polyregular functions} benefit from theorems proven in the field of
rational series \cite{berstel2011noncommutative}, where they coincide with
weighted automata of polynomial ambiguity, rational series of polynomial
growth, and a notion of quantitative $\MSO$
\cite{berstel2011noncommutative,kreutzer2013,schutzenberger1962}. Although
promising, the restriction to unary output did not yield effective procedure to
decide the aperiodicity of polyregular functions.
Notice that unary outputs in $\Rel$ and $\Nat$
are therefore special kind of weighted automata,
and that questions arising from the weighted automata
theory (good first order logic?) are also connected.
\cite{droste2019aperiodic,reutenauer_series_1980}.

Recently, the introduction of $\Rel$-output polyregular functions, called
\kl{$\Rel$-polyregular functions} (and more generally, commutative invertible
outputs) allowed to solve the general problem \cite{LOPEZ23b}. Note that
$\Rel$-output generalizes $\Nat$-output, but somehow has a combinatorial
description that is farther away from traditional \kl{polyregular functions}
because of the invertible outputs. The proof relies on the construction of a
canonical object essentially based on differences between functions, which is
not applicable to unary output. Note that already for rational series, the
problem is open.

We write $\Poly$ for the class of \kl{polyregular functions},
$\SF$ for the class of \kl{star-free polyregular functions},
$\NRat$ for the class of \kl{$\Nat$-rational series},
and $\ZRat$ for the class of \kl{$\Rel$-rational series}.
Arrows denote strict inclusions, and effectiveness
(both in terms of decidability and of effective representation)
is represented by thick double arrows.
Inclusions that are suspected to be effective have a question mark.
It is also conjectured that $\NPoly \cap \ZSF = \NSF$.

\begin{figure}
    \includestandalone[width=\linewidth]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions. 
    }
\end{figure}

It was conjectured in a thesis that some particular form of polynomials
would characterize star-free functions.This paper follows the line of
research on \kl{polyregular functions} that focuses on semantic
characterizations of the asymptotic behavior, rather than the construction
of canonical objects of computation
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}. The goal was to
decide arrows X,Y and Z, namely, to understand precisely the relationship
between star-free and $\Nat$ output. In the case of \kl{$\Nat$-rational
series}, a result of \citeauthor{KARH77} characterizes \kl{commutative}
\kl{$\Nat$-rational series} among \kl{$\Rel$-rational series}
\cite{KARH77}.
This conjecture was based on the previously introduced result of
\textcite[Theorem 3.3]{KARH77}, which we will disprove in \cref{sec:c-example}.
We then provide a corrected version of the theorem, and prove the conjecture of
Douéneau in the case of \emph{commutative} $\Nat$-rational series of polynomial
growth.

\hrule

\paragraph{Contributions.} 
\begin{itemize}
    \item We provide a candidate semantic characterization
    \item We prove that it is correct for commutative inputs
    \item We provide a counter-example to the theorem of nanania.
    \item We show that effective procedures are possible for all arrows 
        and answer positively to the conjecture of XX
        in the case of commutative inputs.
\end{itemize}

\paragraph*{Outline of the paper.}
\begin{itemize}
    \item In a first section, we recall the preliminary definitions.
    \item In a second section, we provide a counter example to the
        theorem of XX.
    \item In a third section, we provide an alternative result.
    \item In a fourth section, we make the results effective,
        and compute everyting.
\end{itemize}

### 2024-01-16 parts/prelim.tex
wohoho

%! TeX program = xelatex
%! lang = en-US
\section{Preliminaries}
\label{preliminaries:sec}

\subsection{Polyregular Functions}

All functions in this paper have output in $\Nat$ or $\Rel$,
hence, we avoid the definition of \kl{polyregular functions}
in full generality. The following \cref{nat-rel-poly:def}
is one of the equivalent definitions of \cite{LOPEZ23b},
and is similar in shape to the \emph{finite counting automata}
introduced by \textcite{schutzenberger1962}.

\begin{definition}[$\Rel$-polyregular functions {\cite[see, e.g.][]{LOPEZ23b}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a production function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    When the monoid $M$ is \kl{aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

Let us illustrate on an example that these notions actually generalise
blabla.

\begin{example}
    \label{regular-language:ex}
    Let $\Sigma$ be a finite alphabet, and
    $L \subseteq \Sigma^*$. Then,
    $L$ is a regular language if and only if
    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
    \kl{$\Nat$-polyregular function}.
    Furthermore, $L$ is a star-free regular language
    if and only if $\ind{L}$ is a
    \kl{star-free $\Nat$-polyregular function}.
\end{example}

\AP Notice that for a \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to
\Rel$, the \intro{growth rate} of $f$, defined as the map $\growth{f} \colon
\Nat \to \Nat$ via $\growth{f}(n) \defined \sup_{\card{w} = n} |f(w)|$, is
bounded by a polynomial in $n$. This property of having a \intro{polynomial
growth} actually characterizes $\ZPoly$ (resp. $\NPoly$) inside $\ZRat$ (resp.
$\NRat$). We formally state this property in \cref{polynomial-rational-polyreg:fact}.

\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
    \label{polynomial-rational-polyreg:fact}
    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
    be a function. The following are equivalent:
    \begin{enumerate}
        \item $f$ is a \kl{$\Nat$-rational series}
            of \kl{polynomial growth}.
        \item $f$ is a \kl{$\Nat$-polyregular function}.
    \end{enumerate}
\end{fact}

\AP Inside $\ZPoly$, there is an infinite hierarchy of classes, defined using
the \kl{growth rate}. Namely, one defines $\ZPoly[k]$ as the functions $f \in
\ZPoly$ such that $\growth{f}(n) = \bigO(n^k)$. Similarly, we let $\NPoly[k] =
\NPoly \cap \ZPoly[k]$. It follows from the results of \cite{LOPEZ23b} that
this hierarchy is nice and for every $f$, there exists $k$ such that
$\growth{f}(n) = \Theta(n^k)$. This parameter also coincides with the minimal
$k$ such that $f$ is represented by a morphism $\mu \colon M^k \to \Rel$.

\begin{fact}[Folklore about regular languages]
    \label{regular:fact}
    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
    is not regular, whenever $a,b \in \Sigma$.
    Regular languages are closed under intersection.
\end{fact}

\begin{fact}
    \label{pre-image-regular:fact}
    The pre-image of a regular language by a $\Nat$-rational series
    is a regular language. In particular,
    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
    language.
\end{fact}

\subsection{Commutative Input}

\AP 
\paragraph{Commutative Polyregular Functions.}
Because this paper focuses on functions that are commutative both in the
input and the output, let us introduce precisely what it means and how it
relates to other functions from $\Nat^k \to \Nat$. For that, let $\Sigma$ be a
finite alphabet. We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
map that counts occurrences of every letter in the input word, namely: $
\commute[w] \defined (a \mapsto \card[a]{w})$.

\AP Let $X$ be a set. A function $f \colon \Sigma^* \to X$ is
\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
\Nat^\Sigma \to X$ such that $g \circ \commute = f$.

\begin{example}
    \label{commutative-function:ex}
    The map $f \colon w \mapsto \card[a]{w} \times \card[b]{w}$ is \kl{commutative}.
\end{example}

\AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function
$f \colon \Sigma^* \to \Rel$ if there exists a map $\eta \colon \set{1, \dots,
n} \to \Sigma$ such that $g \circ (v \mapsto \seqof{v_{\eta(k)}}{1 \leq k \leq
n}) \circ \commute = f$. A key family of functions will be \kl{represented} by
\kl{$\Rel$-polyregular functions}, namely polynomials with multiple indeterminates.

\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
\Rel[X_1, \dots, X_n]$ is a \intro{$\Rel$-rational polynomial} if it
is \kl{represented} by a \kl{$\Rel$-polyregular function}. Similarly, it is a
\intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
\kl{$\Nat$-polyregular function}.

\begin{example}
    \label{negative-not-nrat:ex}
    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomial},
    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
    not a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function},
    hence, $P(X) \defined X$ is
    a \kl{$\Nat$-rational polynomial}. Similarly,
    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-polyregular function},
    showing that $Q(X) \defined X^2 + 3$
    is a \kl{$\Nat$-rational polynomial}.
    Finally, 
    $T(X) \defined - X$ cannot be 
    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
    have non-negative output.
\end{proof}

\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for
all $n_1, \dots, n_k \geq 0$, $P(n_1, \dots, n_k) \geq 0$. Beware that we do
not consider negative values as input, as the numbers $n_i$ will ultimately
count the number of occurrences of a letter in a word. As an example, the
polynomial $(X - Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$,
but the polynomial $X^2 - 2X$ is not.

We saw in \cref{negative-not-nrat:ex} a simple criterion to check whether a
polynomial is a \kl{$\Nat$-rational polynomial}, but being \kl{non-negative} is
not enough, as the following example illustrates. Note that the proof scheme of
this example will be at the core of \cref{thm:counter-example}.

\begin{example}
    The polynomial $P(X, Y) \defined (X - Y)^2$
    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
    but is
    not a \kl{$\Nat$-rational polynomial}.
\end{example}
\begin{proof}
    Assume by contradiction that
    $S$ is a $\Nat$-rational series computing $P$.
    Then, $S^{-1}(\set{0})$ should be a regular language
    (\cref{pre-image-regular:fact}),
    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
    is not a regular language (\cref{regular:fact}).
\end{proof}

\AP Let us now give some vocabulary on polynomials with multiple indeterminate
over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.
For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
\kl{monomial}, but $X + Y$ or $2X^2 + XY$ are not. We write $\Monomials[X_1,
\dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
Every polynomial $P \in \Rel[X_1, \dots, X_n]$ decomposes uniquely
into a sum of \kl{monomials}.

\AP A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
$T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + X^2
+ X + Y$, the set of \reintro{maximal monomials}, written
$\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  X^2}$. A \kl{monomial} has a
\intro{non-negative coefficient} if its multiplicative constant belongs to
$\Nat$, or equivalently, if it is \kl{non-negative} as a polynomial. In the
polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative monomials} are $X^2$
and $Y^2$.

### 2024-01-16 parts/starfree.tex
wohoho

--- 2024-01-11 pnrs.tex
@@
 %! TeX program = xelatex
-%! lang = en-GB
-\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
-
-% INPUTS
-\input{globals/packages}
-\input{globals/colorscheme}
-
-% CONFIG
-\input{globals/macros}
-\input{globals/mathdefs}
-\input{globals/knowledges.kl}
-\input{globals/metadata}
-
-\usepackage{comment}
-%\includecomment{forshort}
-%\excludecomment{proof}
-
-\makeabstract
-% DOCUMENT
-\begin{document}
-
-% FRONT MATTER
-\maketitle
-\acknowledge
-
-\section{TODOs}
-
-\begin{itemize}
-    \item Use $\Nat$-polyregular functions all the time, and just 
-        in the introduction talk about $\Nat$-rational series.
-    \item Prove that $\Nat$-rational series of polynomial 
-        growth are $\Nat$-polyregular functions, maybe using
-        the results about ambiguity in automata?
-        Theorem 5.22
-    \item Define commutativity by a map 
-        $\commute \colon \Sigma^* \to \Nat^{\Sigma}$.
-    \item Define "commutative functions" and its symbol.
-    \item Define "computed by" (like, recognised by) for commutative 
-        functions.
-    \item Change \cref{corrected-version:thm} to use directly
-        star-free!
-    \item Clear up the computability problems.
-    \item Polynomial + Npolyregular => star free?
-    \item Polynomial + N rational => N polyregular?
-    \item cite the paper of Quantitative Monadic Second-Order Logic
-        about why it is interesting class.
-\end{itemize}
-
-% MAIN MATTER
-\section{Introduction}
-
-There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
-\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
-This computational model, in the counting case, can be traced back to a model
-introduced by \textcite{schutzenberger1965finite}. It has been recently proven
-that one can decide the growth rate of such functions, and decide whether they
-are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. Isolating
-a notion of aperiocidity for functions, and deciding this property is often a
-non-trivial process \cite{droste2019aperiodic}. It was conjectured in the PhD
-thesis of Gaëtan Douéneau-Tabot, that the membership problem of $\Nat$-rational
-series among $\Rel$-rational series was decidable under the assumption of
-polynomial growth. The semantic characterization would imply that
-$\Nat$-rational series of polynomial growth that are aperiodic are precisely
-$\Nat$-rational series of polynomial growth, that can be realised by
-$\Rel$-rational series that are aperiodic. This conjecture was based on the
-previously introduced result of \textcite[Theorem 3.3]{KARH77}, which we will
-disprove in \cref{sec:c-example}. 
-We then provide a corrected version of the theorem, and prove the conjecture
-of Douéneau in the case of \emph{commutative} $\Nat$-rational series of
-polynomial growth.
-
-\section{Preliminaries}
-
-\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a production function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        f(w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-\end{definition}
-
-\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
-    \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be a function. The following are equivalent:
-    \begin{enumerate}
-        \item $f$ can be computed as a \k{$\Nat$-rational series}
-            of \kl{polynomial growth}.
-        \item $f$ is a \kl{$\Nat$-polyregular function}.
-    \end{enumerate}
-\end{fact}
-
-\begin{definition}[Letter Counting Function]
-    \label{commutative-map:def}
-    Let $\Sigma$ be a finite alphabet. We define
-    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
-    map that counts occurrences of every letter in the input word, namely:
-    \begin{equation*}
-        \commute[w] \defined \seqof{\card[a]{w}}{a \in \Sigma} \quad .
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[Commutative Function]
-    \label{commutative:def}
-    Let $X$ be a set.
-    A function $f \colon \Sigma^* \to X$ is
-    \intro{commutative} whenever
-    for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
-
-    Equivalently, it is \reintro{commutative} whenever
-    there exists a map $g \colon \Nat^\Sigma \to X$
-    such that $g \circ \commute = f$.
-\end{definition}
-
-\begin{definition}[Represented Function]
-    A function $g \colon \Nat^n \to \Rel$
-    is \intro{represented} by a function $f \colon \Sigma^* \to \Rel$
-    if there exists a map $\mu \colon \set{1, \dots, n} \to \Sigma$
-    such that
-    $g \circ (v \mapsto \seqof{v_{\mu(k)}}{1 \leq k \leq n}) \circ \commute = f$.
-\end{definition}
-
-We need, $\ZRat$, $\NRat$, $\ZPoly$, $\NPoly$, $\ZSF$, and $\NSF$.
-
-We will start by formally stating the result of \citeauthor{KARH77}
-in terms of polynomials and commutative $\Nat$-rational series, and
-then introduce the necessary definitions. The characterization of
-commutative $\Nat$-rational series computing polynomial functions
-is described in terms of a class $\CoveredPoly[\vec{X}]$ of
-polynomials with multiple indeterminate.
-
-\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CoveredPoly[\vec{X}]$
-    is the class of polynomials $P \in \Rel[\vec{X}]$
-    that are \kl{non-negative}
-    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
-    When the indeterminate are clear from the context, we write
-    this class $\CoveredPoly$.
-\end{definition}
-
-We use the term \kl{$\Nat$-rational polynomial}, that will formally be
-stated later on, to denote polynomials that can be computed via a
-$\Nat$-rational series. In the upcoming theorem, this property
-which is \emph{a priori} difficult to decide is reduced to 
-the positivity of the polynomial, and the positivity of \emph{some}
-coefficients.
-
-\begin{faketheorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
-    \label{karh:thm}
-    Let $\Sigma$ be a finite alphabet.
-    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
-    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
-\end{faketheorem}
-
-It is not clearly stated in \cite{KARH77}, but \cref{karh:thm} implies the
-decidability of \kl{$\Nat$-rational polynomial} inside polynomials with $\Rel$
-coefficients. We will not formalize this statement yet, as \cref{karh:thm} does
-not hold in general.
-
-We then provide in \cref{sec:proof} a
-corrected version of the theorem, together with a valid proof. As a
-consequence, it does not invalidate the stated conjecture.
-
-\begin{itemize}
-    \item commutative input and output
-    \item commutative output
-    \item no commutativity ?
-\end{itemize}
-
-\AP \paragraph*{Rational series and regular languages.} This paper uses notions
-of regularity and results about \kl{$\Nat$-rational series}, however most of
-the technicalities will stem from the multiple indeterminate of polynomials,
-and all there is to know about regularity in this document will be confined to
-the following facts. For a survey about (non-commutative) rational series, we
-refer the reader to \cite{berstel2011noncommutative}.
-
-\begin{definition}[Rational series]
-    The following are equivalent
-    \begin{itemize}
-        \item weighted automata
-        \item rational expressions
-        \item 
-    \end{itemize}
-\end{definition}
-
-\begin{fact}[Folklore about regular languages]
-    \label{regular:fact}
-    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular, whenever $a,b \in \Sigma$.
-    Regular languages are closed under intersection.
-\end{fact}
-
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by a $\Nat$-rational series
-    is a regular language. In particular,
-    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
-    language.
-\end{fact}
-
-\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational polynomial} if
-there exists a \reintro{commutative} $\Rel$-rational series $f$ such that $f(w)
-= P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, $P$ is a
-\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
-series is in fact $\Nat$-rational. Beware that we only apply the polynomial
-to \emph{non-negative} values.
-
-\begin{example}
-    \label{negative-not-nrat:ex}
-    With $\Sigma \defined \set{1}$,
-    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
-    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    The function $w \mapsto |w|$ is a \kl{polyregular function} with unary
-    output of \kl{polynomial growth}, hence  can be computed by a
-    \kl{$\Nat$-rational series}. Hence, $P(X_1) \defined X_1$ is
-    a \kl{$\Nat$-rational polynomial}. Similarly,
-    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-rational series},
-    showing that $Q(X_1) \defined X_1^2 + 3$
-    is a \kl{$\Nat$-rational polynomial}.
-    Finally, 
-    $T(X_1) \defined - X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
-    have non-negative output.
-\end{proof}
-
-We saw in \cref{negative-not-nrat:ex}
-a simple criterion to check whether a polynomial is $\Nat$-rational,
-but being non-negative is not enough, as the following example
-illustrates. Note that the proof scheme
-of this example will be at the core of \cref{thm:counter-example}.
-
-\begin{example}
-    Over a binary alphabet $\Sigma \defined \set{a,b}$,
-    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
-    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
-    but is
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Assume by contradiction that
-    $S$ is a $\Nat$-rational series computing $P$.
-    Then, $S^{-1}(\set{0})$ should be a regular language
-    (\cref{pre-image-regular:fact}),
-    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not a regular language (\cref{regular:fact}).
-\end{proof}
-
-\AP Let us now give some vocabulary on polynomials with multiple indeterminate
-over $\Rel$. A \intro{monomial} is a product of variables and integers. For
-instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, but $X + Y$ or
-$2X^2 + XY$ are not.
-
-\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} if
-$P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we do
-not consider negative values, as the numbers $n_i$ will ultimately count the
-number of occurrences of a letter in a word. As an example, the polynomial $(X
-- Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$, but the
-polynomial $X^2 - 2X$ is not.
-
-\AP
-In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
-is a \intro{maximal monomial} if it is a maximal element
-for the visibility ordering of polynomials \emph{in $\Rat$}
-among monomials with non-zero coefficients in $P$.\footnote{
-    One could equivalently define maximal elements by removing the (non-zero)
-    multiplicative constant of the monomials, and then
-    considering the usual divisibility relation of $\Rel$ polynomials.
-}
-In the polynomial 
-$P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
-are $X^2$, $-2 XY$, and $X^2$.
-A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$ if its
-multiplicative constant is positive. In the polynomial $P(X,Y) \defined (X -
-Y)^2$, the \kl{positive monomials} are $X^2$ and $Y^2$.
-
-\paragraph*{Outline of the paper.}
-todo.
-
-\section{The Counter Example}
-\label{sec:c-example}
-
-Let us now introduce the counter example that was found trying to generalise
-the original proof to non-commutative \kl{$\Rel$-rational series} of
-\kl{polynomial growth}. The counter example will use three indeterminate, and
-we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
-polynomial with at most two indeterminate.
-
-\begin{definition}[Counter Example Polynomial]
-    \label{def:bad-polynomial}
-    We define $P(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
-\end{definition}
-
-\begin{theorem}
-    \label{thm:counter-example}
-    Over $\Sigma \defined \set{a,b,c}$,
-    there exists a polynomial $P \in \CoveredPoly$ that is not
-    a \kl{$\Nat$-rational polynomial}. Namely,
-    $P$ is the polynomial of \cref{def:bad-polynomial}.
-\end{theorem}
-\begin{proof}
-    Let us write $X,Y,Z$ instead of $X_a, X_b, X_c$ to improve
-    readability.
-    It is clear that $P$ is \kl{non-negative}. We can develop
-    the expression of $P$ to 
-    obtain
-    $P = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
-    The \kl{maximal monomials} of $P$
-    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
-    \kl{positive coefficients}.
-
-    Assume by contradiction that $P$ is a \kl{$\Nat$-rational polynomial}.
-    There exists a \kl{commutative}
-    $\Nat$-rational series $f \colon \Sigma \to \Nat$
-    such that $P(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-    Remark that $P(X,Y,Z) = 0$
-    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-    $P(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
-    $Z \neq 0$, and $X = Y = 0$.
-
-    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
-    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
-    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ is notoriously not a regular language
-    (\cref{regular:fact}), we
-    conclude that $L$ is not a regular language.
-    However, $L = f^{-1}(\set{0})$, which is a regular language
-    because $f$ is a $\Nat$-rational series
-    (\cref{pre-image-regular:fact}).
-\end{proof}
-
-\begin{corollary}
-    The result stated in \cite[Theorem 3.3]{KARH77}, restated
-    in \cref{karh:thm}, is false
-    for all polynomials with at least $3$ indeterminate.
-\end{corollary}
-
-As we will see in the next section, the characterization
-of \kl{$\Nat$-rational polynomials} using the set $\CoveredPoly$
-holds when the polynomials have at most $2$ indeterminate. This shows
-that the examples in \cite{KARH77} are not invalidated, as they
-all use at most two indeterminate.
-
-\section{The Corrected Theorem}
-\label{sec:proof}
-
-Let us now provide another class of polynomials, $\CorrectPoly$, that restricts
-$\CoveredPoly$ to a subset that is ``closed under fixing indeterminate", namely,
-polynomials $P(X_1, \dots, X_n) \in \CoveredPoly$ such that partially evaluated
-polynomials remain in $\CoveredPoly$. We use the following notation
-to fix the value of some indeterminate
-$\intro*\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally, if $\nu$ is
-a partial function from $\vec{X}$ to $\Nat$, the restriction
-$\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminate $\vec{Y}
-\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the domain of
-$\nu$.
-
-\begin{definition}
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CorrectPoly[\vec{X}]$ is the collection of
-    polynomials $P \in \Rel[\vec{X}]$ such that
-    $P$ is 
-    such that, for every partial function $\nu \colon \vec{X} \to \Nat$,
-    every \kl{maximal monomial} of
-    $\restr{P}{\nu}$ has a \kl{positive coefficient}.
-\end{definition}
-
-First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
-polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
-counter example provided in \cref{thm:counter-example} is not in
-$\CorrectPoly$. For that, notice that for $P$ defined in
-\cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$. Let us also assert that
-polynomials that have all coefficients in $\Nat$ are \kl{$\Nat$-rational
-polynomials}, which will be a base case in an upcoming induction.
-
-\begin{remark}
-    \label{inductive-correct-poly:remark}
-    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item The \kl{maximal monomials} of 
-            $P$ have \kl{positive coefficients},
-            and 
-            for all $n \in \Nat$,
-            $\restr{P}{X = n} \in \CorrectPoly$.
-    \end{enumerate}
-    Furthermore, if $P \in \Rel[]$ is a polynomial without
-    indeterminate, then $P \in \CorrectPoly$ if and only
-    if its \kl{maximal monomial} (the constant term)
-    has a \kl{positive coefficient}.
-\end{remark}
-\begin{proof}
-    By induction on the number of indeterminate,
-    leveraging the fact that if there exists at least one indeterminate
-    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
-    and partial function $\nu \colon \vec{Y} \to \Nat$,
-    we have 
-    \begin{equation*}
-        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
-        \quad
-        ,
-    \end{equation*}
-    where $\nu[X = n]$ is the partial function from
-    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
-    and maps $X$ to $n$.
-\end{proof}
-
-\begin{example}
-    \label{n-poly-n-poly:example}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
-    be a partial function.
-    Notice that $\restr{P}{\nu}$ continues to be
-    a polynomial with \kl{positive coefficients}, and in particular
-    with 
-    \kl{maximal monomials} having \kl{positive coefficients}.
-\end{proof}
-
-Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-To simplify the reasoning, we will use results on $\Nat$-polyregular functions,
-also known as, polyregular functions with unary output
-\cite{doueneau2021pebble,bojanczyk2019string}. Note that the connection between
-rational series and polyregular functions is not specific to $\Nat$, as
-$\Rel$-rational series of polynomial growth are computed as polyregular
-functions with output in $\set{+1,-1}$ post-composed with computing the sum of
-outputted letters \cite{LOPEZ23b}. Because we only consider functions with
-output in $\Rel$, and that a commutative output monoid simplifies the definitions,
-let us only define $\Rel$-polyregular functions.
-
-\begin{lemma}
-    \label{n-poly-combinatorics:lem}
-    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
-    There exists $\omega \in \Nat$
-    such that for all $p \in \Nat$,
-    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
-    for all $u_1, \dots, u_p \in \Sigma^*$,
-    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
-    whose \kl{maximal monomials} have \kl{positive coefficients},
-    and such that:
-    \begin{equation*}
-        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
-        = P(X_1, \dots, X_p) \quad .
-    \end{equation*}
-\end{lemma}
-\begin{proof}[Proof Sketch]
-    Let $\omega$ be an idempotent power for the finite monoid $M$.
-    We define an equivalence between partitions of a word 
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$
-    as follows: two partitions into $k$ words in $\Sigma^*$ are equivalent
-    when normalizing them with the rules
-    $u_i^{\omega} u_i^{\omega} \to u_i^{\omega}$ for all $1 \leq i \leq p$.
-    This equivalence relation has finite index, and all partitions in the same
-    equivalence class yield the same value through $\pi$, because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid $M$.
-
-    Now, let us remark that the number of partitions
-    in a certain equivalence class is proportional to
-    the number of nanania, that is, is of the form
-    $(X_i - K_i)^{n_i}$ for some $K$. We conclude that
-    maximal monomials indeed have positive coefficients.
-    We even conclude that
-    the polynomial is in $\Nat[\vec{X}]$
-    if we shift the variables with a large enough $K$. 
-\end{proof}
-
-\begin{lemma}
-    \label{n-rat-correct:lem}
-    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-\end{lemma}
-\begin{proof}[Proof Idea]
-    Let $P$ be a \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma$  be the finite alphabet such that
-    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
-    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
-    $\Nat$-rational series computing $P$.
-    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
-    it is a $\Nat$-polyregular function 
-    (\cref{polynomial-rational-polyreg:fact}).
-
-    Using \cref{n-poly-combinatorics:lem},
-    we conclude that maximal monomials of $P$ have
-    positive coefficients. Furthermore, when fixing variables
-    in $P$, we are computing a new \kl{$\Nat$-polyregular function},
-    hence we also conclude that the maximal monomials have
-    positive coefficients.
-    We have proven that $P \in \CorrectPoly$.
-\end{proof}
-
-The core of the upcoming proof of \cref{lem:correct-to-n-rat} relies on the
-following observation: if a polynomial $P$ is written $X^p + Q$, where all
-terms of $Q$ have degree less than $p$, then one can find some large enough $K$
-so that $X^p - (X - K)^p + Q$ also has a positive maximal coefficient.
-It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence
-the maximal coefficient can be chosen arbitrarily large by increasing
-$K$.
-This
-trick will work similarly in multivariate polynomials, and the overall
-proof scheme works as follows: for a bounded number of possible
-values of $X$, we can fix the variable, and have one less item to work on.
-For large enough values of $X$ (typically greater than some $K$), we can guarantee that
-$X^p - (X - K)^p + Q$ has positive maximal coefficients, hence
-can be computed by a $\Nat$-rational series, while
-$(X - K)^p$ is clearly computable by a $\Nat$-rational series,
-hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
-as the sum of two $\Nat$-rational series, and we have
-concluded that our original polynomial was a
-\kl{$\Nat$-rational polynomial}. There are a number of problems in generalizing
-this proof scheme:
-\begin{itemize}
-    \item We will have to guarantee that every polynomial considered
-        is non-negative
-    \item We do not only need to check that the \kl{maximal monomials}
-        have \kl{positive coefficient}, but that it is the case
-        when fixing arbitrarily many variables! (this
-        covers the first point too)
-\end{itemize}
-
-\begin{lemma}
-    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
-    $\vec{p} \colon \vec{X}Y \to \Nat$,
-    and $M \defined \times \prod_{X \in \vec{X}} X^{p(X)}$.
-    Then, for all $K \geq 0$,
-    $\Delta_{Y,K} \defined Y^p M - (Y - K)^p M$ has \kl{maximal monomials} with
-    \kl{positive coefficients} of weight at least $K$, regardless of
-    the variables fixed, as long as the (potential)
-    value of $Y$ is greater
-    than $K$.
-\end{lemma}
-\begin{proof}
-    Let us first remark that
-    \begin{equation*}
-        \Delta_{Y,K}
-        = KY^{p-1} M + Q M
-        \quad .
-    \end{equation*}
-    In particular, there is a unique \kl{maximal monomial} for $\Delta_{Y,K}$,
-    which is precisely $K Y^{p-1} M$, because all monomials
-    in $Q M$ divide (strictly) $Y^{p-1} M$.
-    Let us fix a variable,
-    $X$ to any (non-zero) value $p > 0$. Then
-    The \kl{maximal monomial} of $\restr{\Delta_{Y,K}}{X = p}$
-    is exactly $\restr{KY^{p-1}M}{X = p}$, which remains
-    with a \kl{positive coefficient}, that is proportional to $K$.
-    Let us fix $Y$ to some value $K + l$, with $l > 0$,
-    then
-    $\restr{\Delta_{Y,K}}{Y = K + l}
-    = (K + l)^p M - l^p M
-    = ((K + l) - l) (\sum_{i = 1}^{p-1} (K + l)^i l^{p-i}) M
-    = K \alpha M
-    $
-    which remains a monomial with a \kl{positive coefficient},
-    that is proportional to $K$.
-\end{proof}
-
-We can refine the above lemma to use it in the case where
-we do not have a single
-monomial.
-
-\begin{lemma}
-    \label{lem:delta-cool}
-    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
-    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
-    \kl{maximal monomial} of $P$, with $\alpha > 0$.
-    Then, there exists a computable $K \geq 0$
-    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
-    has \kl{positive coefficients} for \kl{maximal monomials}
-    in any partial evaluation that satisfies $Y > K$ (if it
-    gives a value to $Y$).
-\end{lemma}
-\begin{proof}
-    Let $N$ be greater than the sum of the absolute values of all coefficients
-    appearing in $P$, and let $K = 2 N$.
-
-    Let us remark that
-    \begin{equation*}
-        \Delta_K = \underbrace{(P - \alpha Y^p M)}_{\text{ left hand side }}
-        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{\text{ right hand side }}
-        \quad .
-    \end{equation*}
-    an that for every partial evaluation $\nu$ that sends $Y$ to a value
-    greater than $K$,
-    we have that $\restr{(Y^p M - (Y - K)^p M)}{\nu}$ has
-    \kl{maximal monomials} with \kl{positive coefficients}
-    proportional to $K$ (note that here ``proportional"
-    implies that they are greater than $K$, because we multiply by positive
-    integers).
-
-    To conclude, it is just a game of describing how
-    \kl{maximal monomials} of $\restr{\Delta_Y}{\nu}$
-    can appear. If the \kl{maximal monomial} is obtained
-    by summing an element of the left hand side with an element
-    of the right hand side, it needs to be a \kl{maximal monomial}
-    of the right hand side, and the large value of $K$ guarantees
-    that the sum of coefficients is positive.
-    If it only appears on the right hand side, it is even better.
-    If it only appears on the left hand side, then
-    in particular it means that the monomial does not
-    divide $\restr{Y^p M}{\nu}$, hence has the same value
-    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
-    which is guaranteed to be positive because $P \in \CorrectPoly$.
-
-    In particular, we conclude from this that
-    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
-\end{proof}
-
-Note that we cannot conclude in the above lemma that the result is in
-$\CorrectPoly$, because we have a restriction on the range where the variable
-$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
-induction on the number $d$ of indeterminate, and the degrees of the
-\kl{maximal monomials}. To make the induction formal, let us define
-$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
-$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
-$\Nat^{\vec{X}}$ with the product ordering, which is a
-\kl{well-quasi-ordering}.
-
-\AP To a polynomial $P$, we assign the set $\MaximalMonomials(P)$ of its
-\kl{maximal monomials}, belonging to $\Pfin(\Monomials)$, which remains a
-\kl{well-quasi-ordering} when endowed with the \intro{Hoare ordering} of
-subsets, that is, $S_1 \hoareleq S_2$ whenever for all \kl{monomial} $M_1 \in
-S_1$, there exists a monomial $M_2 \in S_2$, such that $M_1$ \kl{divides}
-$M_2$.
-
-\begin{lemma}
-    \label{lem:correct-to-n-rat}
-    Let $P$ be a polynomial in $\CorrectPoly$.
-    Then, 
-    $P$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on $\MaximalMonomials(P)$.
-
-    \textbf{Base case:} If the \kl{maximal monomial} of $P$ is a constant term,
-    then because $P \in \CorrectPoly$, $P = n \in \Nat$, which is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-
-    \textbf{Induction:}
-    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
-    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
-    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
-    Using \cref{lem:delta-cool},
-    we know that there exists a computable $K$ such that
-    the polynomial
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$
-    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
-    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
-    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
-    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
-    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
-    \hoarele \MaximalMonomials(P)$.
-    By induction hypothesis,
-    we conclude that
-    $\Diff{K}(Y - K)$ is \kl{represented}
-    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
-    \begin{equation*}
-        f_1(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \Diff{K} & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $\alpha \geq 0$, it is an easy check
-    that the following function is also a $\Nat$-rational series
-    \begin{equation*}
-        f_2(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \alpha(Y - K)^pM & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $P \in \CorrectPoly$, for every $n \leq K$, $\restr{P}{Y = n} \in
-    \CorrectPoly$ (\cref{inductive-correct-poly:remark}).
-    Furthermore,
-    $\MaximalMonomials(\restr{P}{Y = n})
-    \hoarele \MaximalMonomials(P)$. Hence, by induction hypothesis,
-    the following function is \kl{represented} by a
-    \kl{star-free $\Nat$-rational series}:
-    \begin{equation*}
-        f_3(Y, \vec{X}) \defined
-        \begin{cases}
-            P(Y, \vec{X}) & \text{ if } Y \leq K \\
-            0 & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    We conclude by remarking that 
-    $P = f_1 + f_2 + f_3$.
-\end{proof}
-
-We are now ready to state the corrected and generalised version of
-\cref{karh:thm}.
-
-\begin{theorem}
-    \label{corrected-version:thm}
-    Let $P \in \Rel[\vec{X}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item \label{corrected-1:item} $P \in \CorrectPoly$,
-        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
-        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-    \end{enumerate}
-\end{theorem}
-\begin{proof}
-    The implications 
-    \cref{corrected-4:item} $\implies$
-    \cref{corrected-3:item} $\implies$
-    \cref{corrected-2:item} are obvious.
-    \cref{lem:correct-to-n-rat} proves
-    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
-    while \cref{n-rat-correct:lem}
-    proves 
-    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
-\end{proof}
-
-\begin{lemma}
-    \label{lem:correct-covered-2}
-    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
-\end{lemma}
-\begin{proof}
-    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \to \Nat$.
-    For the converse inclusion, let us consider $P(X,Y)$
-    that is \kl{non-negative}, such that the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-
-    If we fix none of the variables, then the \kl{maximal monomials}
-    have \kl{positive coefficients} by assumption. If we fix one of the
-    variables, we can assume without loss of generality that we 
-    fix $X = k$ for some $k \in \Nat$.
-    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
-    and therefore must have a positive leading coefficient
-    (which is the unique \kl{maximal monomial} in this case)
-    or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-    The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminate, leading to a constant polynomial.
-\end{proof}
-
-\section{Effective procedures}
+%! lang = en-US
+\section{Beyond Polynomials}
 \label{sec:deciding}
@@
 \end{theorem}
-
-Before proving \cref{decidable-n-poly:thm}, let us first examine the
-effectiveness of \cref{corrected-version:thm}, namely, provide an algorithm to
-decide whether a polynomial $P$ belongs to $\CorrectPoly$. The idea will be to
-compute a bound to values taken by the valuations $\nu$ appearing in the
-definition of $\CorrectPoly$. As a \emph{mise en jambes}, let us consider
-separately the case of one indeterminate.
-
-\begin{lemma}
-    \label{decide-correct-univariate:lemma}
-    Let $P \in \Rel[X]$, there exists a computable $N \in \Nat$,
-    such that the following are equivalent:
-    \begin{enumerate}
-        \item \label{decide-correct-univariate-1:item}
-            $P \in \CorrectPoly$,
-        \item \label{decide-correct-univariate-2:item}
-            The \kl{maximal monomial}
-            of $P$ has a \kl{positive coefficient},
-            and $P(n) \in \Nat$
-            for all $0 \leq n \leq N_0$.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    Let us first remark that the implication 
-    \cref{decide-correct-univariate-1:item}
-    $\implies$ \cref{decide-correct-univariate-2:item} is trivial for every
-    $N \in \Nat$.
-    For the converse implication,
-    let $P \in \Rel[X]$.
-    If $P$ is a constant polynomial, then $P \in \CorrectPoly$
-    if and only if $P(0) \in \Nat$, and $N \defined 0$
-    suffices.
-    Otherwise, $P$ is non-constant, and let $n \geq 1$ be the \kl{degree} of $P$.
-    Let $P_1$ and $\alpha > 0$ be such that
-    $P \defined P_1 + \alpha X^n$. 
-    To prove that $P \in \CorrectPoly$, it suffices to show that
-    $P(k) \in \Nat$ for all $k \in \Nat$.
-    Let us define $\beta$ as the
-    maximal absolute value of coefficients appearing in $P_1$, which is
-    assumed to be non-zero,
-    and $N_0 \defined \beta + 1 \geq 2$.
-    Remark that for $X \geq N_0$,
-    \begin{align*}
-        P(X) &\geq X^n - \beta \sum_{i = 0}^{n-1} X^i \\
-             &= X^n - \beta \frac{X^n - 1}{X - 1} \\
-             &= \frac{X^{n+1} - X^n - \beta X^n + \beta}{X - 1} \\
-             &= \frac{X^n(X - (\beta + 1)) + \beta}{X - 1} \\
-             &\geq 0
-    \end{align*}
-    This proves that $P(k) \in \Nat$ for all $k \geq N_0$,
-    and because $P$ satisfies \cref{decide-correct-univariate-2:item},
-    $P(k) \in \Nat$ for all $1 \leq k \leq N_0$. We have proven 
-    that $P \in \CorrectPoly$.
-\end{proof}
-
-In the case of multiple indeterminate, we leverage
-\cref{lem:delta-cool} to perform an induction on the degree
-of the polynomials and the number of variables of the polynomial.
-
-\begin{lemma}
-    \label{decide-correct-poly:lem}
-    One can decide if a polynomial belongs to $\CorrectPoly$.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on the number of indeterminate.
-
-    \textbf{Base Case:} If $P$ has zero indeterminate, i.e., is a constant
-    function, then to decide whether $P \in \CorrectPoly$ it suffices to check
-    that $P(0) \in \Nat$.
-
-    \textbf{Induction:} Let $P \in \Rel[Y, \vec{X}]$, we start by checking
-    whether \kl{maximal monomials} of $P$ have \kl{positive coefficients}.
-    Without loss of generality, assume that $Y$ divides some \kl{maximal
-    monomial} $\alpha Y^p M$ with $p \geq 1$. By applying
-    \cref{lem:delta-cool}, we can compute a bound $K \in \Nat$ such that
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$ has \kl{positive coefficients}
-    for its \kl{maximal monomials} for all partial valuations $\nu \colon
-    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
-    particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to
-    $\CorrectPoly$ for all $n > K$. By induction hypothesis, one can
-    decide whether $\restr{P}{X = n}$ belongs to $\CorrectPoly$
-    for all $0 \leq n \leq K$, which is equivalent 
-    to deciding whether
-    $P \in \CorrectPoly$ thanks to \cref{inductive-correct-poly:remark}.
-\end{proof}
 
@@
 
+\begin{conjecture}
+    $\CorrectPoly$ is precisely composed of polynomials that
+    have non-negative coefficients in a \emph{binomial basis},
+    i.e., where polynomials are obtained
+    as sums and products of $\binom{X_i}{p_i}$.
+\end{conjecture}
+
 \subsection{Star-free $\Nat$-polyregular functions}
+\label{star-free:sec}
 
@@
     \begin{equation*}
-        \ZCommut \cap \ZSF \cap \NPoly
+        \ZCommut \cap \ZSF[k] \cap \NPoly
+        = \ZCommut \cap \ZSF \cap \NPoly[k]
         = \ZCommut \cap \NSF
@@
 \end{proof}
-
-\section{Conclusion}
-\label{sec:ccl}
-
-\begin{itemize}
-    \item Actually, we have an algorithm that allows
-        us to \emph{decide} if a polynomial is $\Nat$-rational.
-    \item Actually, we obtain a characterisation of
-        commutative 
-        $\Nat$-rational series of polynomial growth
-        among $\Rel$-rational series.
-    \item The question of $\Nat$-polyregular functions
-        among $\Rel$-polyregular functions remains open
-        because of the non-commutativity of the input.
-    \item The question of $\Nat$-rational series (commutative)
-        among $\Rel$-rational series remains open.
-\end{itemize}
-
-\begin{corollary}
-    The following problem is decidable: given $P \in \Rel[\vec{X}]$,
-    decide whether $P$ is a \kl{$\Nat$-rational polynomial}.
-\end{corollary}
-\begin{proof}
-    Thanks to \cref{karh:thm}, one only has to decide whether
-    $P \in \CoveredPoly$. The procedure is as follows.
-    \begin{enumerate}
-        \item Check if all \kl{maximal monomials} have
-            \kl{positive coefficients} in $P$, if not, answer \texttt{false}.
-        \item Consider $N \in \Nat$ large enough so that
-              $P \geq 0$ when all indeterminate are chosen greater
-              than $N$. This is possible because all \kl{maximal
-              monomials} have \kl{positive coefficients}, and such an $N$
-              is computable.
-        \item Compute $P$
-    \end{enumerate}
-
-    Now, if all \kl{maximal monomials}
-    have \kl{positive coefficients}, then there exists a computable
-    $N \in \Nat$ such that 
-    In particular,
-
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly = \setof{P}{ P \geq 0 \wedge \exists K \in \Nat, P \circ (+K) \in \Nat[\vec{X}]}$.
-\end{conjecture}
-\begin{proof}[Proof Sketch]
-    If $S \defined P \circ (+K) \in \Nat[\vec{X}]$, then for every partial application,
-    maximal coefficients of $S$ are positive. As maximal coefficients of 
-    (partial applications of) $S$ are the same as the ones of $P$
-    \emph{except for the constant terms} (full application),
-    where $P$ is already guaranteed to be non-negative, we conclude
-    that $P \in \CorrectPoly$.
-
-    Conversely, we prove by induction on the degree of $P \in \CorrectPoly$,
-    that there exists some $K$ such that $P \circ (+K) \in \Nat[\vec{X}]$.
-    First, notice that $\CorrectPoly$ is closed under pre-composition with
-    $(+K)$. Then, remark that $P(X - K)$ has the same maximal coefficients as
-    $P$ and new maximal coefficients that are multiples of $K$. Hence, for a
-    large enough $K$, we conclude that $P(X - K)$ is the sum of its maximal
-    coefficients, plus a polynomial that lies in $\CorrectPoly$. We conclude by
-    induction hypothesis. 
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly$ is precisely composed of polynomials that
-    have non-negative coefficients in a \emph{binomial basis},
-    i.e., where polynomials are obtained
-    as sums and products of $\binom{X_i}{p_i}$.
-\end{conjecture}
-
-\begin{conjecture}
-    star-free $\Nat$-rational is star-free and $\Nat$-rational?
-\end{conjecture}
-
-\begin{center}
-    \begin{tabular}{c|cc}
-        \toprule
-        & \textbf{Commutative} & \textbf{General} \\
-        \midrule
-        \textbf{Commutative}
-        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
-        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
-        \textbf{General} & ? & polyregular 
-    \end{tabular}
-\end{center}
-
-% BACKMATTER
-\printbibliography
-
-\appendix
-
-\end{document}

### 2024-01-16 pnrs.tex
adding todos
wohoho

--- 2024-01-11 pnrs.tex
@@
         about why it is interesting class.
+    \item Example of polyregular functions
+    \item Example of non-polyregular functions
+    \item Expand a bit on the notion of "derivative"
+        and divisibility of monomials.
+    \item What about matrices ? They are not minimisable... so we do not know
+        about their eigenvalues ?
+    \item What about the residual transducer ?
+    \item What about weigthed automata? Does it mean something?
+        -> we have unary input, so the automata really is an automata.
+        -> we output numbers in $\Nat$
+        -> residuals are what? totally ordered?
+        -> recursive calls?
+        -> if the function computes polynomials in $\Nat$, we can build
+        the usual thing? and then we add "a few steps before"?
+
+    \item Is the proof simpler by assuming
+        $P(X + K) \in \Nat[X]$? This is way better.
+
+    \item Prove decidability of the property.
+
+    \item Show the equivalence with binomial
+
+    \item What about 
 \end{itemize}
 
-% MAIN MATTER
-\section{Introduction}
-
-There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
-\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
-This computational model, in the counting case, can be traced back to a model
-introduced by \textcite{schutzenberger1965finite}. It has been recently proven
-that one can decide the growth rate of such functions, and decide whether they
-are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. Isolating
-a notion of aperiocidity for functions, and deciding this property is often a
-non-trivial process \cite{droste2019aperiodic}. It was conjectured in the PhD
-thesis of Gaëtan Douéneau-Tabot, that the membership problem of $\Nat$-rational
-series among $\Rel$-rational series was decidable under the assumption of
-polynomial growth. The semantic characterization would imply that
-$\Nat$-rational series of polynomial growth that are aperiodic are precisely
-$\Nat$-rational series of polynomial growth, that can be realised by
-$\Rel$-rational series that are aperiodic. This conjecture was based on the
-previously introduced result of \textcite[Theorem 3.3]{KARH77}, which we will
-disprove in \cref{sec:c-example}. 
-We then provide a corrected version of the theorem, and prove the conjecture
-of Douéneau in the case of \emph{commutative} $\Nat$-rational series of
-polynomial growth.
-
-\section{Preliminaries}
-
-\begin{definition}[$\Rel$-polyregular functions, {\cite[see, e.g.]{LOPEZ23b}}]
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a production function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        f(w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-\end{definition}
-
-\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
-    \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be a function. The following are equivalent:
-    \begin{enumerate}
-        \item $f$ can be computed as a \k{$\Nat$-rational series}
-            of \kl{polynomial growth}.
-        \item $f$ is a \kl{$\Nat$-polyregular function}.
-    \end{enumerate}
-\end{fact}
-
-\begin{definition}[Letter Counting Function]
-    \label{commutative-map:def}
-    Let $\Sigma$ be a finite alphabet. We define
-    $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
-    map that counts occurrences of every letter in the input word, namely:
-    \begin{equation*}
-        \commute[w] \defined \seqof{\card[a]{w}}{a \in \Sigma} \quad .
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[Commutative Function]
-    \label{commutative:def}
-    Let $X$ be a set.
-    A function $f \colon \Sigma^* \to X$ is
-    \intro{commutative} whenever
-    for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
-
-    Equivalently, it is \reintro{commutative} whenever
-    there exists a map $g \colon \Nat^\Sigma \to X$
-    such that $g \circ \commute = f$.
-\end{definition}
-
-\begin{definition}[Represented Function]
-    A function $g \colon \Nat^n \to \Rel$
-    is \intro{represented} by a function $f \colon \Sigma^* \to \Rel$
-    if there exists a map $\mu \colon \set{1, \dots, n} \to \Sigma$
-    such that
-    $g \circ (v \mapsto \seqof{v_{\mu(k)}}{1 \leq k \leq n}) \circ \commute = f$.
-\end{definition}
-
-We need, $\ZRat$, $\NRat$, $\ZPoly$, $\NPoly$, $\ZSF$, and $\NSF$.
-
-We will start by formally stating the result of \citeauthor{KARH77}
-in terms of polynomials and commutative $\Nat$-rational series, and
-then introduce the necessary definitions. The characterization of
-commutative $\Nat$-rational series computing polynomial functions
-is described in terms of a class $\CoveredPoly[\vec{X}]$ of
-polynomials with multiple indeterminate.
-
-\begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CoveredPoly[\vec{X}]$
-    is the class of polynomials $P \in \Rel[\vec{X}]$
-    that are \kl{non-negative}
-    and such that every \kl{maximal monomial} has a \kl{positive coefficient}.
-    When the indeterminate are clear from the context, we write
-    this class $\CoveredPoly$.
-\end{definition}
-
-We use the term \kl{$\Nat$-rational polynomial}, that will formally be
-stated later on, to denote polynomials that can be computed via a
-$\Nat$-rational series. In the upcoming theorem, this property
-which is \emph{a priori} difficult to decide is reduced to 
-the positivity of the polynomial, and the positivity of \emph{some}
-coefficients.
-
-\begin{faketheorem}[{\cite[Theorem 3.3, page 4]{KARH77}}] 
-    \label{karh:thm}
-    Let $\Sigma$ be a finite alphabet.
-    The set of \kl{$\Nat$-rational polynomials} over $\Sigma$
-    coincides with $\CoveredPoly[\seqof{X_a}{a \in \Sigma}]$.
-\end{faketheorem}
-
-It is not clearly stated in \cite{KARH77}, but \cref{karh:thm} implies the
-decidability of \kl{$\Nat$-rational polynomial} inside polynomials with $\Rel$
-coefficients. We will not formalize this statement yet, as \cref{karh:thm} does
-not hold in general.
-
-We then provide in \cref{sec:proof} a
-corrected version of the theorem, together with a valid proof. As a
-consequence, it does not invalidate the stated conjecture.
-
-\begin{itemize}
-    \item commutative input and output
-    \item commutative output
-    \item no commutativity ?
-\end{itemize}
-
-\AP \paragraph*{Rational series and regular languages.} This paper uses notions
-of regularity and results about \kl{$\Nat$-rational series}, however most of
-the technicalities will stem from the multiple indeterminate of polynomials,
-and all there is to know about regularity in this document will be confined to
-the following facts. For a survey about (non-commutative) rational series, we
-refer the reader to \cite{berstel2011noncommutative}.
-
-\begin{definition}[Rational series]
-    The following are equivalent
-    \begin{itemize}
-        \item weighted automata
-        \item rational expressions
-        \item 
-    \end{itemize}
-\end{definition}
-
-\begin{fact}[Folklore about regular languages]
-    \label{regular:fact}
-    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular, whenever $a,b \in \Sigma$.
-    Regular languages are closed under intersection.
-\end{fact}
-
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by a $\Nat$-rational series
-    is a regular language. In particular,
-    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
-    language.
-\end{fact}
-
-\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[\seqof{X_a}{a \in \Sigma}]$ is a \intro{$\Rel$-rational polynomial} if
-there exists a \reintro{commutative} $\Rel$-rational series $f$ such that $f(w)
-= P(\seqof{ \card[a]{w} }{a \in \Sigma})$. Similarly, $P$ is a
-\intro{$\Nat$-rational polynomial} if the \reintro{commutative} $\Rel$-rational
-series is in fact $\Nat$-rational. Beware that we only apply the polynomial
-to \emph{non-negative} values.
-
-\begin{example}
-    \label{negative-not-nrat:ex}
-    With $\Sigma \defined \set{1}$,
-    the polynomials $X_1$, and $X_1^2 + 3$ are \kl{$\Nat$-rational polynomial},
-    but $- X_1$ is a \kl{$\Rel$-rational polynomial} that is 
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    The function $w \mapsto |w|$ is a \kl{polyregular function} with unary
-    output of \kl{polynomial growth}, hence  can be computed by a
-    \kl{$\Nat$-rational series}. Hence, $P(X_1) \defined X_1$ is
-    a \kl{$\Nat$-rational polynomial}. Similarly,
-    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-rational series},
-    showing that $Q(X_1) \defined X_1^2 + 3$
-    is a \kl{$\Nat$-rational polynomial}.
-    Finally, 
-    $T(X_1) \defined - X_1$ cannot be $\Nat$-rational as $\Nat$-rational series
-    have non-negative output.
-\end{proof}
-
-We saw in \cref{negative-not-nrat:ex}
-a simple criterion to check whether a polynomial is $\Nat$-rational,
-but being non-negative is not enough, as the following example
-illustrates. Note that the proof scheme
-of this example will be at the core of \cref{thm:counter-example}.
-
-\begin{example}
-    Over a binary alphabet $\Sigma \defined \set{a,b}$,
-    the polynomial $P(X_a, X_b) \defined (X_a - X_b)^2$
-    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
-    but is
-    not a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Assume by contradiction that
-    $S$ is a $\Nat$-rational series computing $P$.
-    Then, $S^{-1}(\set{0})$ should be a regular language
-    (\cref{pre-image-regular:fact}),
-    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not a regular language (\cref{regular:fact}).
-\end{proof}
-
-\AP Let us now give some vocabulary on polynomials with multiple indeterminate
-over $\Rel$. A \intro{monomial} is a product of variables and integers. For
-instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, but $X + Y$ or
-$2X^2 + XY$ are not.
-
-\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} if
-$P(n_1, \dots, n_k) \geq 0$ for all $n_1, \dots, n_k \geq 0$. Beware that we do
-not consider negative values, as the numbers $n_i$ will ultimately count the
-number of occurrences of a letter in a word. As an example, the polynomial $(X
-- Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$, but the
-polynomial $X^2 - 2X$ is not.
-
-\AP
-In a polynomial $P \in \Rel[X_1, \dots, X_k]$, a monomial 
-is a \intro{maximal monomial} if it is a maximal element
-for the visibility ordering of polynomials \emph{in $\Rat$}
-among monomials with non-zero coefficients in $P$.\footnote{
-    One could equivalently define maximal elements by removing the (non-zero)
-    multiplicative constant of the monomials, and then
-    considering the usual divisibility relation of $\Rel$ polynomials.
-}
-In the polynomial 
-$P(X,Y) \defined X^2 - 2XY + X^2 + X + Y$, the \reintro{maximal monomials}
-are $X^2$, $-2 XY$, and $X^2$.
-A \kl{monomial} has \intro{positive coefficient} in a polynomial $P$ if its
-multiplicative constant is positive. In the polynomial $P(X,Y) \defined (X -
-Y)^2$, the \kl{positive monomials} are $X^2$ and $Y^2$.
-
-\paragraph*{Outline of the paper.}
-todo.
-
-\section{The Counter Example}
-\label{sec:c-example}
-
-Let us now introduce the counter example that was found trying to generalise
-the original proof to non-commutative \kl{$\Rel$-rational series} of
-\kl{polynomial growth}. The counter example will use three indeterminate, and
-we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
-polynomial with at most two indeterminate.
-
-\begin{definition}[Counter Example Polynomial]
-    \label{def:bad-polynomial}
-    We define $P(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
-\end{definition}
-
-\begin{theorem}
-    \label{thm:counter-example}
-    Over $\Sigma \defined \set{a,b,c}$,
-    there exists a polynomial $P \in \CoveredPoly$ that is not
-    a \kl{$\Nat$-rational polynomial}. Namely,
-    $P$ is the polynomial of \cref{def:bad-polynomial}.
-\end{theorem}
-\begin{proof}
-    Let us write $X,Y,Z$ instead of $X_a, X_b, X_c$ to improve
-    readability.
-    It is clear that $P$ is \kl{non-negative}. We can develop
-    the expression of $P$ to 
-    obtain
-    $P = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
-    The \kl{maximal monomials} of $P$
-    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
-    \kl{positive coefficients}.
-
-    Assume by contradiction that $P$ is a \kl{$\Nat$-rational polynomial}.
-    There exists a \kl{commutative}
-    $\Nat$-rational series $f \colon \Sigma \to \Nat$
-    such that $P(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-    Remark that $P(X,Y,Z) = 0$
-    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-    $P(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
-    $Z \neq 0$, and $X = Y = 0$.
-
-    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
-    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
-    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ is notoriously not a regular language
-    (\cref{regular:fact}), we
-    conclude that $L$ is not a regular language.
-    However, $L = f^{-1}(\set{0})$, which is a regular language
-    because $f$ is a $\Nat$-rational series
-    (\cref{pre-image-regular:fact}).
-\end{proof}
-
-\begin{corollary}
-    The result stated in \cite[Theorem 3.3]{KARH77}, restated
-    in \cref{karh:thm}, is false
-    for all polynomials with at least $3$ indeterminate.
-\end{corollary}
-
-As we will see in the next section, the characterization
-of \kl{$\Nat$-rational polynomials} using the set $\CoveredPoly$
-holds when the polynomials have at most $2$ indeterminate. This shows
-that the examples in \cite{KARH77} are not invalidated, as they
-all use at most two indeterminate.
-
-\section{The Corrected Theorem}
-\label{sec:proof}
-
-Let us now provide another class of polynomials, $\CorrectPoly$, that restricts
-$\CoveredPoly$ to a subset that is ``closed under fixing indeterminate", namely,
-polynomials $P(X_1, \dots, X_n) \in \CoveredPoly$ such that partially evaluated
-polynomials remain in $\CoveredPoly$. We use the following notation
-to fix the value of some indeterminate
-$\intro*\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y)$. More generally, if $\nu$ is
-a partial function from $\vec{X}$ to $\Nat$, the restriction
-$\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminate $\vec{Y}
-\defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the domain of
-$\nu$.
-
-\begin{definition}
-    Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CorrectPoly[\vec{X}]$ is the collection of
-    polynomials $P \in \Rel[\vec{X}]$ such that
-    $P$ is 
-    such that, for every partial function $\nu \colon \vec{X} \to \Nat$,
-    every \kl{maximal monomial} of
-    $\restr{P}{\nu}$ has a \kl{positive coefficient}.
-\end{definition}
-
-First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
-polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
-counter example provided in \cref{thm:counter-example} is not in
-$\CorrectPoly$. For that, notice that for $P$ defined in
-\cref{def:bad-polynomial}, $P(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$. Let us also assert that
-polynomials that have all coefficients in $\Nat$ are \kl{$\Nat$-rational
-polynomials}, which will be a base case in an upcoming induction.
-
-\begin{remark}
-    \label{inductive-correct-poly:remark}
-    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item The \kl{maximal monomials} of 
-            $P$ have \kl{positive coefficients},
-            and 
-            for all $n \in \Nat$,
-            $\restr{P}{X = n} \in \CorrectPoly$.
-    \end{enumerate}
-    Furthermore, if $P \in \Rel[]$ is a polynomial without
-    indeterminate, then $P \in \CorrectPoly$ if and only
-    if its \kl{maximal monomial} (the constant term)
-    has a \kl{positive coefficient}.
-\end{remark}
-\begin{proof}
-    By induction on the number of indeterminate,
-    leveraging the fact that if there exists at least one indeterminate
-    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
-    and partial function $\nu \colon \vec{Y} \to \Nat$,
-    we have 
-    \begin{equation*}
-        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
-        \quad
-        ,
-    \end{equation*}
-    where $\nu[X = n]$ is the partial function from
-    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
-    and maps $X$ to $n$.
-\end{proof}
-
-\begin{example}
-    \label{n-poly-n-poly:example}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Let $P \in \Nat[\vec{X}]$, and $\nu \colon \vec{X} \to \Nat$
-    be a partial function.
-    Notice that $\restr{P}{\nu}$ continues to be
-    a polynomial with \kl{positive coefficients}, and in particular
-    with 
-    \kl{maximal monomials} having \kl{positive coefficients}.
-\end{proof}
-
-Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-To simplify the reasoning, we will use results on $\Nat$-polyregular functions,
-also known as, polyregular functions with unary output
-\cite{doueneau2021pebble,bojanczyk2019string}. Note that the connection between
-rational series and polyregular functions is not specific to $\Nat$, as
-$\Rel$-rational series of polynomial growth are computed as polyregular
-functions with output in $\set{+1,-1}$ post-composed with computing the sum of
-outputted letters \cite{LOPEZ23b}. Because we only consider functions with
-output in $\Rel$, and that a commutative output monoid simplifies the definitions,
-let us only define $\Rel$-polyregular functions.
-
-\begin{lemma}
-    \label{n-poly-combinatorics:lem}
-    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
-    There exists $\omega \in \Nat$
-    such that for all $p \in \Nat$,
-    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
-    for all $u_1, \dots, u_p \in \Sigma^*$,
-    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
-    whose \kl{maximal monomials} have \kl{positive coefficients},
-    and such that:
-    \begin{equation*}
-        f(\alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i))
-        = P(X_1, \dots, X_p) \quad .
-    \end{equation*}
-\end{lemma}
-\begin{proof}[Proof Sketch]
-    Let $\omega$ be an idempotent power for the finite monoid $M$.
-    We define an equivalence between partitions of a word 
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$
-    as follows: two partitions into $k$ words in $\Sigma^*$ are equivalent
-    when normalizing them with the rules
-    $u_i^{\omega} u_i^{\omega} \to u_i^{\omega}$ for all $1 \leq i \leq p$.
-    This equivalence relation has finite index, and all partitions in the same
-    equivalence class yield the same value through $\pi$, because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid $M$.
-
-    Now, let us remark that the number of partitions
-    in a certain equivalence class is proportional to
-    the number of nanania, that is, is of the form
-    $(X_i - K_i)^{n_i}$ for some $K$. We conclude that
-    maximal monomials indeed have positive coefficients.
-    We even conclude that
-    the polynomial is in $\Nat[\vec{X}]$
-    if we shift the variables with a large enough $K$. 
-\end{proof}
-
-\begin{lemma}
-    \label{n-rat-correct:lem}
-    The \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-\end{lemma}
-\begin{proof}[Proof Idea]
-    Let $P$ be a \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma$  be the finite alphabet such that
-    $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$,
-    and $f \colon \Sigma^* \to \Nat$ be the \kl{commutative}
-    $\Nat$-rational series computing $P$.
-    Because $f$ has polynomial growth, and is a $\Nat$-rational series,
-    it is a $\Nat$-polyregular function 
-    (\cref{polynomial-rational-polyreg:fact}).
-
-    Using \cref{n-poly-combinatorics:lem},
-    we conclude that maximal monomials of $P$ have
-    positive coefficients. Furthermore, when fixing variables
-    in $P$, we are computing a new \kl{$\Nat$-polyregular function},
-    hence we also conclude that the maximal monomials have
-    positive coefficients.
-    We have proven that $P \in \CorrectPoly$.
-\end{proof}
-
-The core of the upcoming proof of \cref{lem:correct-to-n-rat} relies on the
-following observation: if a polynomial $P$ is written $X^p + Q$, where all
-terms of $Q$ have degree less than $p$, then one can find some large enough $K$
-so that $X^p - (X - K)^p + Q$ also has a positive maximal coefficient.
-It works because $X^p - (X - K)^p = K X^{p-1} + \cdots$, hence
-the maximal coefficient can be chosen arbitrarily large by increasing
-$K$.
-This
-trick will work similarly in multivariate polynomials, and the overall
-proof scheme works as follows: for a bounded number of possible
-values of $X$, we can fix the variable, and have one less item to work on.
-For large enough values of $X$ (typically greater than some $K$), we can guarantee that
-$X^p - (X - K)^p + Q$ has positive maximal coefficients, hence
-can be computed by a $\Nat$-rational series, while
-$(X - K)^p$ is clearly computable by a $\Nat$-rational series,
-hence $(X^p - (X - K)^p + Q) + (X - K)^p$ is computable
-as the sum of two $\Nat$-rational series, and we have
-concluded that our original polynomial was a
-\kl{$\Nat$-rational polynomial}. There are a number of problems in generalizing
-this proof scheme:
-\begin{itemize}
-    \item We will have to guarantee that every polynomial considered
-        is non-negative
-    \item We do not only need to check that the \kl{maximal monomials}
-        have \kl{positive coefficient}, but that it is the case
-        when fixing arbitrarily many variables! (this
-        covers the first point too)
-\end{itemize}
-
-\begin{lemma}
-    Let $\vec{X}$ be a tuple of indeterminate, $Y$ be a fresh indeterminate,
-    $\vec{p} \colon \vec{X}Y \to \Nat$,
-    and $M \defined \times \prod_{X \in \vec{X}} X^{p(X)}$.
-    Then, for all $K \geq 0$,
-    $\Delta_{Y,K} \defined Y^p M - (Y - K)^p M$ has \kl{maximal monomials} with
-    \kl{positive coefficients} of weight at least $K$, regardless of
-    the variables fixed, as long as the (potential)
-    value of $Y$ is greater
-    than $K$.
-\end{lemma}
-\begin{proof}
-    Let us first remark that
-    \begin{equation*}
-        \Delta_{Y,K}
-        = KY^{p-1} M + Q M
-        \quad .
-    \end{equation*}
-    In particular, there is a unique \kl{maximal monomial} for $\Delta_{Y,K}$,
-    which is precisely $K Y^{p-1} M$, because all monomials
-    in $Q M$ divide (strictly) $Y^{p-1} M$.
-    Let us fix a variable,
-    $X$ to any (non-zero) value $p > 0$. Then
-    The \kl{maximal monomial} of $\restr{\Delta_{Y,K}}{X = p}$
-    is exactly $\restr{KY^{p-1}M}{X = p}$, which remains
-    with a \kl{positive coefficient}, that is proportional to $K$.
-    Let us fix $Y$ to some value $K + l$, with $l > 0$,
-    then
-    $\restr{\Delta_{Y,K}}{Y = K + l}
-    = (K + l)^p M - l^p M
-    = ((K + l) - l) (\sum_{i = 1}^{p-1} (K + l)^i l^{p-i}) M
-    = K \alpha M
-    $
-    which remains a monomial with a \kl{positive coefficient},
-    that is proportional to $K$.
-\end{proof}
-
-We can refine the above lemma to use it in the case where
-we do not have a single
-monomial.
-
-\begin{lemma}
-    \label{lem:delta-cool}
-    Let $\vec{X}$ be a tuple of indeterminate, and $Y$ be a fresh indeterminate.
-    Let $P \in \Rel[\vec{X}Y]$ be in $\CorrectPoly$, and $\alpha Y^p M$ be a
-    \kl{maximal monomial} of $P$, with $\alpha > 0$.
-    Then, there exists a computable $K \geq 0$
-    such that $\Delta_{K} \defined P - \alpha (Y - K)^p M$
-    has \kl{positive coefficients} for \kl{maximal monomials}
-    in any partial evaluation that satisfies $Y > K$ (if it
-    gives a value to $Y$).
-\end{lemma}
-\begin{proof}
-    Let $N$ be greater than the sum of the absolute values of all coefficients
-    appearing in $P$, and let $K = 2 N$.
-
-    Let us remark that
-    \begin{equation*}
-        \Delta_K = \underbrace{(P - \alpha Y^p M)}_{\text{ left hand side }}
-        + \underbrace{\alpha (Y^p M - (Y - K)^p M)}_{\text{ right hand side }}
-        \quad .
-    \end{equation*}
-    an that for every partial evaluation $\nu$ that sends $Y$ to a value
-    greater than $K$,
-    we have that $\restr{(Y^p M - (Y - K)^p M)}{\nu}$ has
-    \kl{maximal monomials} with \kl{positive coefficients}
-    proportional to $K$ (note that here ``proportional"
-    implies that they are greater than $K$, because we multiply by positive
-    integers).
-
-    To conclude, it is just a game of describing how
-    \kl{maximal monomials} of $\restr{\Delta_Y}{\nu}$
-    can appear. If the \kl{maximal monomial} is obtained
-    by summing an element of the left hand side with an element
-    of the right hand side, it needs to be a \kl{maximal monomial}
-    of the right hand side, and the large value of $K$ guarantees
-    that the sum of coefficients is positive.
-    If it only appears on the right hand side, it is even better.
-    If it only appears on the left hand side, then
-    in particular it means that the monomial does not
-    divide $\restr{Y^p M}{\nu}$, hence has the same value
-    in $\restr{\Delta_K}{\nu}$ or in $\restr{P}{\nu}$,
-    which is guaranteed to be positive because $P \in \CorrectPoly$.
-
-    In particular, we conclude from this that
-    $\Delta_Y \geq 0$, on all tuples such that $Y > K$.
-\end{proof}
-
-Note that we cannot conclude in the above lemma that the result is in
-$\CorrectPoly$, because we have a restriction on the range where the variable
-$Y$ can be taken. The upcoming \cref{lem:correct-to-n-rat} is proven using an
-induction on the number $d$ of indeterminate, and the degrees of the
-\kl{maximal monomials}. To make the induction formal, let us define
-$\Monomials[\vec{X}]$ as the (infinite) set of all possible \kl{monomials} over
-$\vec{X}$, ordered by \kl{divisibility}. Remark that this set is isomorphic to
-$\Nat^{\vec{X}}$ with the product ordering, which is a
-\kl{well-quasi-ordering}.
-
-\AP To a polynomial $P$, we assign the set $\MaximalMonomials(P)$ of its
-\kl{maximal monomials}, belonging to $\Pfin(\Monomials)$, which remains a
-\kl{well-quasi-ordering} when endowed with the \intro{Hoare ordering} of
-subsets, that is, $S_1 \hoareleq S_2$ whenever for all \kl{monomial} $M_1 \in
-S_1$, there exists a monomial $M_2 \in S_2$, such that $M_1$ \kl{divides}
-$M_2$.
-
-\begin{lemma}
-    \label{lem:correct-to-n-rat}
-    Let $P$ be a polynomial in $\CorrectPoly$.
-    Then, 
-    $P$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on $\MaximalMonomials(P)$.
-
-    \textbf{Base case:} If the \kl{maximal monomial} of $P$ is a constant term,
-    then because $P \in \CorrectPoly$, $P = n \in \Nat$, which is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-
-    \textbf{Induction:}
-    Assume that $\alpha Y^p M \in \MaximalMonomials(P)$
-    for some indeterminate $Y$, $\alpha \neq 0$, and $p \geq 1$.
-    Because $P \in \CorrectPoly$, we know that $\alpha > 0$.
-    Using \cref{lem:delta-cool},
-    we know that there exists a computable $K$ such that
-    the polynomial
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$
-    satisfies that for all partial valuations $\nu \colon Y\vec{X} \to \Nat$
-    such that $\nu(Y) > K$ or $\nu(Y)$ is undefined,
-    the \kl{maximal monomials} of $\Diff{K}$ have \kl{positive coefficients}.
-    In particular, $\Diff{K}(Y - K)$  belongs to $\CorrectPoly$.
-    Furthermore, $\MaximalMonomials(\Diff{K}) = \MaximalMonomials(\Diff{K}(Y - K))
-    \hoarele \MaximalMonomials(P)$.
-    By induction hypothesis,
-    we conclude that
-    $\Diff{K}(Y - K)$ is \kl{represented}
-    by a \kl{star-free $\Nat$-polyregular function} $g_1$,
-    \begin{equation*}
-        f_1(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \Diff{K} & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $\alpha \geq 0$, it is an easy check
-    that the following function is also a $\Nat$-rational series
-    \begin{equation*}
-        f_2(Y, \vec{X}) \defined
-        \begin{cases}
-            0 & \text{ if } Y \leq K \\
-            \alpha(Y - K)^pM & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-
-    Because $P \in \CorrectPoly$, for every $n \leq K$, $\restr{P}{Y = n} \in
-    \CorrectPoly$ (\cref{inductive-correct-poly:remark}).
-    Furthermore,
-    $\MaximalMonomials(\restr{P}{Y = n})
-    \hoarele \MaximalMonomials(P)$. Hence, by induction hypothesis,
-    the following function is \kl{represented} by a
-    \kl{star-free $\Nat$-rational series}:
-    \begin{equation*}
-        f_3(Y, \vec{X}) \defined
-        \begin{cases}
-            P(Y, \vec{X}) & \text{ if } Y \leq K \\
-            0 & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    We conclude by remarking that 
-    $P = f_1 + f_2 + f_3$.
-\end{proof}
-
-We are now ready to state the corrected and generalised version of
-\cref{karh:thm}.
-
-\begin{theorem}
-    \label{corrected-version:thm}
-    Let $P \in \Rel[\vec{X}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item \label{corrected-1:item} $P \in \CorrectPoly$,
-        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
-        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
-    \end{enumerate}
-\end{theorem}
-\begin{proof}
-    The implications 
-    \cref{corrected-4:item} $\implies$
-    \cref{corrected-3:item} $\implies$
-    \cref{corrected-2:item} are obvious.
-    \cref{lem:correct-to-n-rat} proves
-    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
-    while \cref{n-rat-correct:lem}
-    proves 
-    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
-\end{proof}
-
-\begin{lemma}
-    \label{lem:correct-covered-2}
-    $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
-\end{lemma}
-\begin{proof}
-    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \to \Nat$.
-    For the converse inclusion, let us consider $P(X,Y)$
-    that is \kl{non-negative}, such that the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-
-    If we fix none of the variables, then the \kl{maximal monomials}
-    have \kl{positive coefficients} by assumption. If we fix one of the
-    variables, we can assume without loss of generality that we 
-    fix $X = k$ for some $k \in \Nat$.
-    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
-    and therefore must have a positive leading coefficient
-    (which is the unique \kl{maximal monomial} in this case)
-    or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have \kl{positive coefficients}.
-    The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminate, leading to a constant polynomial.
-\end{proof}
-
-\section{Effective procedures}
-\label{sec:deciding}
-
-The goal of this section is to go beyond polynomials and characterize all
-$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
-that they are \kl{commutative}.
-
-\begin{theorem}
-    \label{decidable-n-poly:thm}
-    Given a \kl{commutative}
-    \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to \Rel$,
-    one can decide if $f$ can be computed by a \kl{$\Nat$-polyregular function},
-    and effectively compute a representation of $f$.
-\end{theorem}
-
-Before proving \cref{decidable-n-poly:thm}, let us first examine the
-effectiveness of \cref{corrected-version:thm}, namely, provide an algorithm to
-decide whether a polynomial $P$ belongs to $\CorrectPoly$. The idea will be to
-compute a bound to values taken by the valuations $\nu$ appearing in the
-definition of $\CorrectPoly$. As a \emph{mise en jambes}, let us consider
-separately the case of one indeterminate.
-
-\begin{lemma}
-    \label{decide-correct-univariate:lemma}
-    Let $P \in \Rel[X]$, there exists a computable $N \in \Nat$,
-    such that the following are equivalent:
-    \begin{enumerate}
-        \item \label{decide-correct-univariate-1:item}
-            $P \in \CorrectPoly$,
-        \item \label{decide-correct-univariate-2:item}
-            The \kl{maximal monomial}
-            of $P$ has a \kl{positive coefficient},
-            and $P(n) \in \Nat$
-            for all $0 \leq n \leq N_0$.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    Let us first remark that the implication 
-    \cref{decide-correct-univariate-1:item}
-    $\implies$ \cref{decide-correct-univariate-2:item} is trivial for every
-    $N \in \Nat$.
-    For the converse implication,
-    let $P \in \Rel[X]$.
-    If $P$ is a constant polynomial, then $P \in \CorrectPoly$
-    if and only if $P(0) \in \Nat$, and $N \defined 0$
-    suffices.
-    Otherwise, $P$ is non-constant, and let $n \geq 1$ be the \kl{degree} of $P$.
-    Let $P_1$ and $\alpha > 0$ be such that
-    $P \defined P_1 + \alpha X^n$. 
-    To prove that $P \in \CorrectPoly$, it suffices to show that
-    $P(k) \in \Nat$ for all $k \in \Nat$.
-    Let us define $\beta$ as the
-    maximal absolute value of coefficients appearing in $P_1$, which is
-    assumed to be non-zero,
-    and $N_0 \defined \beta + 1 \geq 2$.
-    Remark that for $X \geq N_0$,
-    \begin{align*}
-        P(X) &\geq X^n - \beta \sum_{i = 0}^{n-1} X^i \\
-             &= X^n - \beta \frac{X^n - 1}{X - 1} \\
-             &= \frac{X^{n+1} - X^n - \beta X^n + \beta}{X - 1} \\
-             &= \frac{X^n(X - (\beta + 1)) + \beta}{X - 1} \\
-             &\geq 0
-    \end{align*}
-    This proves that $P(k) \in \Nat$ for all $k \geq N_0$,
-    and because $P$ satisfies \cref{decide-correct-univariate-2:item},
-    $P(k) \in \Nat$ for all $1 \leq k \leq N_0$. We have proven 
-    that $P \in \CorrectPoly$.
-\end{proof}
-
-In the case of multiple indeterminate, we leverage
-\cref{lem:delta-cool} to perform an induction on the degree
-of the polynomials and the number of variables of the polynomial.
-
-\begin{lemma}
-    \label{decide-correct-poly:lem}
-    One can decide if a polynomial belongs to $\CorrectPoly$.
-\end{lemma}
-\begin{proof}
-    We prove the result by induction on the number of indeterminate.
-
-    \textbf{Base Case:} If $P$ has zero indeterminate, i.e., is a constant
-    function, then to decide whether $P \in \CorrectPoly$ it suffices to check
-    that $P(0) \in \Nat$.
-
-    \textbf{Induction:} Let $P \in \Rel[Y, \vec{X}]$, we start by checking
-    whether \kl{maximal monomials} of $P$ have \kl{positive coefficients}.
-    Without loss of generality, assume that $Y$ divides some \kl{maximal
-    monomial} $\alpha Y^p M$ with $p \geq 1$. By applying
-    \cref{lem:delta-cool}, we can compute a bound $K \in \Nat$ such that
-    $\Diff{K} \defined P - \alpha (Y - K)^p M$ has \kl{positive coefficients}
-    for its \kl{maximal monomials} for all partial valuations $\nu \colon
-    Y\vec{X} \to \Nat$ such that $Y > K$ (if $Y$ is assigned a value). In
-    particular, we conclude that, for all $\restr{P}{Y = n}$ belongs to
-    $\CorrectPoly$ for all $n > K$. By induction hypothesis, one can
-    decide whether $\restr{P}{X = n}$ belongs to $\CorrectPoly$
-    for all $0 \leq n \leq K$, which is equivalent 
-    to deciding whether
-    $P \in \CorrectPoly$ thanks to \cref{inductive-correct-poly:remark}.
-\end{proof}
-
-To prove the theorem, let us first provide a nice characterization of
-\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
-This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
-commutative case.
-
-\begin{lemma}
-    \label{decompose-polynomial:lem}
-    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
-    \kl{$\Rel$-polyregular function}. There exists $\omega \in \Nat$,
-    and polynomials $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{ 0 \leq n_a < \omega}$,
-    such that for all $w \in \Sigma^*$,
-    \begin{equation*}
-        f(w) = P_{\seqof{\card[a]{w} [\omega]}{a \in \Sigma}}
-        (\seqof{\card[a]{w} / \omega}{a \in \Sigma}) \quad .
-    \end{equation*}
-    Which informally means that for every choice of rest modulo $\omega$
-    for every variable, there exists a polynomial describing the
-    behavior of $f$ when adding multiples of $\omega$ letters at a time.
-
-    Furthermore, the polynomials are computable from $f$.
-\end{lemma}
-\begin{proof}
-    equivalent partitions of the word, but commutativity?.q
-    Lemma 5.37 phd gaëtan.
-\end{proof}
-
-As a consequence, we conclude 
-\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
-    We use \cref{decompose-polynomial:lem}
-    to effectively compute
-    a polynomial representation of $f$. Remark that
-    $f$ is a \kl{$\Nat$-polyregular function} if and only if
-    all these polynomials are \kl{$\Nat$-rational polynomials},
-    which is decidable thanks to
-    \cref{decide-correct-poly:lem,corrected-version:thm}.
-\end{proof}
-
-\subsection{Star-free $\Nat$-polyregular functions}
-
-Let us now prove that the above characterizations of
-commutative
-$\Nat$-polyregular functions can be combined with the recent
-advances in the study of $\Rel$-polyregular functions 
-\cite{LOPEZ23b}
-to prove the following theorem.
-
-\begin{theorem}
-    \label{zsf-npoly-nsf:thm}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be both a \kl{star-free $\Rel$-polyregular function}
-    and a \kl{commutative} \kl{$\Nat$-polyregular function}.
-    Then, $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}.
-    That is, the following equality holds:
-    \begin{equation*}
-        \ZCommut \cap \ZSF \cap \NPoly
-        = \ZCommut \cap \NSF
-        \quad .
-    \end{equation*}
-\end{theorem}
-
-Note that this proves in the commutative case a conjecture that was 
-formulated by Thesis Gaetan, and that is non trivial, because
-
-\begin{conjecture}
-    \label{nsf-zsf:conj}
-    In the general case, the following equation holds:
-    \begin{equation*}
-        \ZSF \cap \NPoly
-        = 
-        \NSF
-        \quad .
-    \end{equation*}
-\end{conjecture}
-
-As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
-procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
-decidable inside $\NPoly$ under the assumption of commutativity.
-
-\begin{corollary}
-    One can decide if a \kl{commutative} \kl{ $\Rel$-polyregular function}
-    can be realized by a \kl{star-free $\Nat$-polyregular function},
-    in which case the conversion is effective.
-\end{corollary}
-
-The key ingredient of this section is the use of a semantic characterization of
-\kl{star-free $\Rel$-polyregular functions} among \kl{$\Rel$-rational series}
-that generalizes the aperiodicity of languages to functions by the means of
-polynomial behaviors.
-
-\begin{definition}[Ultimately polynomial]
-    \label{ultimately-polynomial:def}
-    Let $\Sigma$ be a finite alphabet. 
-    A function $f \colon \Sigma^* \to \Rel$
-    is \intro{ultimately polynomial}
-    whenever there exists $N_0 \in \Nat$ such that
-    for all $n \in \Nat$
-    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
-    \in \Sigma^*$, there exists a polynomial $P \in \Rel[X_1, \dots, X_n]$
-    such that
-    \begin{equation*}
-        f\left(
-            \alpha_0 \prod_{i = 1}^{n} w_i^{X_i} \alpha_i
-        \right)
-        = 
-        P(X_1, \dots, X_n)
-        \quad 
-        \forall X_1, \dots, X_n \geq N_0
-        \quad .
-    \end{equation*}
-\end{definition}
-
-Let us recall from \cite{LOPEZ23b} that this definition correctly
-generalizes the indicator functions of regular languages.
-
-\begin{example}
-    A language $L$ is aperiodic if and only if 
-    the indicator function is ultimately polynomial.
-\end{example}
-\begin{proof}
-    TODO.
-\end{proof}
-
-\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
-    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
-
-    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
-    only the empty word $\varepsilon$, and because $f$ is a
-    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
-    Note that $f$ is therefore a constant function, equal to a natural
-    number, hence belongs to $\NSF$.
-
-    \textbf{Induction.}
-    Because $f$ is \kl{ultimately polynomial},
-    there exists $N_0 \in \Nat$, 
-    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
-    such that
-    \begin{equation*}
-        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
-        = 
-        P(\seqof{X_a}{a \in \Sigma})
-        \quad
-        \text{ when }
-        \forall a \in \Sigma, X_a \geq N_0
-        \quad .
-    \end{equation*}
-    Furthermore, leveraging \cref{decompose-polynomial:lem},
-    we obtain $\omega \in \Nat$
-    and a family of polynomials 
-    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
-    such that.
-    Because each of those polynomials coincide with $P$ over an infinite
-    subgrid of $\Nat$,
-    we conclude that all polynomials are in fact equal to a single one.
-    we conclude that the polynomials 
-    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.
-
-    Now, we conclude in particular that the function 
-    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
-    hence is computable by a \kl{star-free $\Nat$-polyregular function},
-    because of \cref{nsf-polynomials:thm}.
-    When fixing values of some indeterminate below $N_0$,
-    we obtain a new function $g$, that remains commutative,
-    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
-    By induction hypothesis, this function $g$ is also 
-    \kl{star-free $\Nat$-polyregular}, and we conclude
-    that 
-    $f = P + \sum g$ which remains star free.
-\end{proof}
-
-\section{Conclusion}
-\label{sec:ccl}
-
-\begin{itemize}
-    \item Actually, we have an algorithm that allows
-        us to \emph{decide} if a polynomial is $\Nat$-rational.
-    \item Actually, we obtain a characterisation of
-        commutative 
-        $\Nat$-rational series of polynomial growth
-        among $\Rel$-rational series.
-    \item The question of $\Nat$-polyregular functions
-        among $\Rel$-polyregular functions remains open
-        because of the non-commutativity of the input.
-    \item The question of $\Nat$-rational series (commutative)
-        among $\Rel$-rational series remains open.
-\end{itemize}
-
-\begin{corollary}
-    The following problem is decidable: given $P \in \Rel[\vec{X}]$,
-    decide whether $P$ is a \kl{$\Nat$-rational polynomial}.
-\end{corollary}
-\begin{proof}
-    Thanks to \cref{karh:thm}, one only has to decide whether
-    $P \in \CoveredPoly$. The procedure is as follows.
-    \begin{enumerate}
-        \item Check if all \kl{maximal monomials} have
-            \kl{positive coefficients} in $P$, if not, answer \texttt{false}.
-        \item Consider $N \in \Nat$ large enough so that
-              $P \geq 0$ when all indeterminate are chosen greater
-              than $N$. This is possible because all \kl{maximal
-              monomials} have \kl{positive coefficients}, and such an $N$
-              is computable.
-        \item Compute $P$
-    \end{enumerate}
-
-    Now, if all \kl{maximal monomials}
-    have \kl{positive coefficients}, then there exists a computable
-    $N \in \Nat$ such that 
-    In particular,
-
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly = \setof{P}{ P \geq 0 \wedge \exists K \in \Nat, P \circ (+K) \in \Nat[\vec{X}]}$.
-\end{conjecture}
-\begin{proof}[Proof Sketch]
-    If $S \defined P \circ (+K) \in \Nat[\vec{X}]$, then for every partial application,
-    maximal coefficients of $S$ are positive. As maximal coefficients of 
-    (partial applications of) $S$ are the same as the ones of $P$
-    \emph{except for the constant terms} (full application),
-    where $P$ is already guaranteed to be non-negative, we conclude
-    that $P \in \CorrectPoly$.
-
-    Conversely, we prove by induction on the degree of $P \in \CorrectPoly$,
-    that there exists some $K$ such that $P \circ (+K) \in \Nat[\vec{X}]$.
-    First, notice that $\CorrectPoly$ is closed under pre-composition with
-    $(+K)$. Then, remark that $P(X - K)$ has the same maximal coefficients as
-    $P$ and new maximal coefficients that are multiples of $K$. Hence, for a
-    large enough $K$, we conclude that $P(X - K)$ is the sum of its maximal
-    coefficients, plus a polynomial that lies in $\CorrectPoly$. We conclude by
-    induction hypothesis. 
-\end{proof}
-
-\begin{conjecture}
-    $\CorrectPoly$ is precisely composed of polynomials that
-    have non-negative coefficients in a \emph{binomial basis},
-    i.e., where polynomials are obtained
-    as sums and products of $\binom{X_i}{p_i}$.
-\end{conjecture}
-
-\begin{conjecture}
-    star-free $\Nat$-rational is star-free and $\Nat$-rational?
-\end{conjecture}
-
-\begin{center}
-    \begin{tabular}{c|cc}
-        \toprule
-        & \textbf{Commutative} & \textbf{General} \\
-        \midrule
-        \textbf{Commutative}
-        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
-        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
-        \textbf{General} & ? & polyregular 
-    \end{tabular}
-\end{center}
+\input{parts/intro.tex}
+\input{parts/prelim.tex}
+\input{parts/counter.tex}
+\input{parts/correction.tex}
+\input{parts/starfree.tex}
+\input{parts/conclusion.tex}
 

### 2024-01-17 parts/correction.tex
residual transducer??

@@
 coefficient for a \kl{maximal monomial}, namely $-2XY$.
-
-\begin{remark}
-    \label{inductive-correct-poly:remark}
-    Let $P \in \Rel[X,\vec{Y}]$ be a polynomial.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item The \kl{maximal monomials} of 
-            $P$ are \kl{non-negative}
-            and 
-            for all $n \in \Nat$,
-            $\restr{P}{X = n} \in \CorrectPoly$.
-    \end{enumerate}
-    Furthermore, if $P \in \Rel[]$ is a polynomial without
-    indeterminate, then $P \in \CorrectPoly$ if and only
-    if its \kl{maximal monomial} (the constant term)
-    is \kl{non-negative}.
-\end{remark}
-\begin{proof}
-    By induction on the number of indeterminate,
-    leveraging the fact that if there exists at least one indeterminate
-    $X$, i.e., that $\vec{X} = X \vec{Y}$, then for all $n \in \Nat$
-    and partial function $\nu \colon \vec{Y} \topartial \Nat$,
-    we have 
-    \begin{equation*}
-        \restr{\left(\restr{P}{X = n}\right)}{\nu} = \restr{P}{ \nu[X = n]}
-        \quad
-        ,
-    \end{equation*}
-    where $\nu[X = n]$ is the partial function from
-    $\vec{X}$ to $\Nat$ that coincides with $\nu$ on $\vec{Y}$
-    and maps $X$ to $n$.
-\end{proof}
 
@@
 \begin{lemma}
-    \label{derivation-translation:lem}
-    Let $P \in \Rel[\vec{X}]$, the following are equivalent
-    \begin{enumerate}
-        \item $P \in \CorrectPoly$,
-        \item $P$ is \kl{non-negative}
-            and there exists a computable $K \in \Nat$
-            such that $\translate{K}(P) \in \Nat[\vec{X}]$.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    \textbf{TODO: do the proof}.
-    To prove \cref{corrected-5:item} $\implies$ \cref{corrected-1:item}, it
-    suffices to remark that for every polynomial $P$, for every partial
-    valuation $\nu \colon \vec{X} \topartial \Nat$, for every $K \in \Nat$, the
-    \kl{maximal monomials} of $\restr{\translate{K}(P)}{\nu}$ are exactly the
-    \kl{maximal monomials} of $\restr{P}{\nu}$, the former being
-    \kl{non-negative} if $\translate{K}(P) \in \Nat[\vec{X}]$.
-
-    Conversely, the implication \cref{corrected-1:item} $\implies$
-    \cref{corrected-5:item} follows immediately from the fact that polynomials
-    in $\CorrectPoly$ are \kl{non-negative}, and on the following induction.
-    Let $P \in \CorrectPoly$, if $P$ is constant, then $\translate{0}(P) \in \Nat[\vec{X}]$.
-    Otherwise, if $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
-    monomials} of $P$, there exists $K \in \Nat$ such that $Q \defined
-    \Diff{K}{P_1}+ P_2$ belongs to $\CorrectPoly$,
-    and by induction hypothesis there exists $L \in
-    \Nat$ such that $\translate{L}{Q} \in \Nat[\vec{X}]$. As a consequence,
-    $\translate{K+L}(P) = \translate{L}(P_1) + \translate{K}(\translate{L}(Q))$
-    is a sum of two polynomials in $\Nat[\vec{X}]$, and we have concluded.
-\end{proof}
-
-\begin{lemma}
     \label{lem:correct-to-n-rat}
@@
 While \cref{lem:correct-to-n-rat} provides effective conversion, it does not
-explicitly state that the nembership is decidable. However, it is not too
+explicitly state that the membership is decidable. However, it is not too
 difficult to remark that $\CorrectPoly$, the proof scheme can be followed to
-obtain decidability. For the sake of completeness, we provide a self-contained
-proof hereafter, that furthermore
+obtain decidability. For the sake of completeness and readability,
+we extract this fact in the following lemma.
 
@@
         \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
-        \item \label{corrected-5:item} $P$ is \kl{non-negative}
-            and there exists $K \in \Nat$ such that $\translate{K}(P) \in \Nat[\vec{X}]$.
     \end{enumerate}
@@
     procedure.
-
-    The equivalence between
-    \cref{corrected-5:item} and \cref{corrected-1:item}
-    is precisely \cref{derivation-translation:lem}.
 \end{proof}
@@
 \end{proof}
+
+It was conjectured in the 
+\cite[Open question 5.55]{gaetanphd}.
+\cite[Conjecture 7.61]{gaetanphd}
+Let us remark that a previous conjecture was not true, and provide the
+recursively defined variant. This can be understood in the light of
+\cref{corrected-version:thm}, as the fact that \kl{$\Nat$-rational polynomials}
+are essentially $\Nat$-polynomials.
+
+\begin{example}
+    The polynomial $\BadPoly$ is not a 
+    \kl{$\Nat$-rational polynomial},
+    but is \kl{non-negative} and satisfies
+    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
+\end{example}
+
+\begin{lemma}
+    \label{derivation-translation:lem}
+    Let $P \in \Rel[\vec{X}]$, the following are equivalent
+    \begin{enumerate}
+        \item \label{d-t-correct:item} $P \in \CorrectPoly$,
+        \item \label{d-t-transl:item}
+            There exists $K \in \Nat$,
+            such that for every partial
+            function $\nu \colon \vec{X} \topartial \Nat$,
+            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
+    \end{enumerate}
+\end{lemma}
+\begin{proof}
+    To prove \cref{d-t-transl:item} $\implies$
+    \cref{d-t-correct:item}, let us remark that
+    if $P$ is \kl{non-negative} and $\translate{K}(P) \in \Nat[\vec{X}]$,
+    then $\translate{K}(P) \in \CorrectPoly$ thanks to
+    \cref{n-poly-n-poly:example}.
+    In particular, 
+    then for every partial valuation 
+    $\nu \colon \vec{X} \topartial \Nat$, 
+    $\restr{\translate{K}(P)}{\nu} \in \Nat[\vec{X}]$,
+    and in particular conclude that
+    \kl{maximal monomials} of $\restr{\translate{K}(P)}{\nu}$ are exactly the
+    \kl{maximal monomials} of $\restr{P}{\nu}$, the former being
+    \kl{non-negative} if $\translate{K}(P) \in \Nat[\vec{X}]$.
+
+    Conversely, the implication \cref{d-t-correct:item} $\implies$
+    \cref{d-t-transl:item} is proven by induction on $\MaximalMonomials(P)$.
+    If $P$ is a constant polynomial, then $P = n$ for some $n \in \Nat$,
+    and we conclude that $\translate{0}(P) \in \Nat[\vec{X}]$.
+    Otherwise, $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
+    monomials} of $P$.
+    Using \cref{derivation-stabilises-correct:lem},
+    there exists a computable $K \in \Nat$
+    such that 
+    $Q \defined
+    \Diff{K}{P_1} + P_2$ belongs to $\CorrectPoly$.
+    Since, $\MaximalMonomials(Q) \hoarele \MaximalMonomials(P)$,
+    we conclude by induction hypothesis that
+    there exists $L \in \Nat$ such that 
+    for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
+    $\translate{L}(\restr{Q}{\nu}) \in \Nat[\vec{X}]$. 
+    Similarly,
+    for all partial functions $\nu \colon \vec{X} \topartial \Nat$
+    fixing at least one indeterminate (appearing in $P$),
+    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials(P)$,
+    hence there exists $L_\nu \in \Nat$ such that
+    $\translate{L_\nu}{\restr{P}{\nu}} \in \Nat[\vec{X}]$.
+    Let us define $L_m$ to be the maximum of $L$, and $L_\nu$
+    where $\nu$ ranges over partial functions described above.
+
+    Let us prove that for all partial functions
+    $\nu \colon \vec{X} \topartial \Nat$,
+    $\translate{L_m + K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
+    If $\nu$ is the empty valuation, then $\restr{P}{\nu} = P$,
+    and
+    $\translate{L_m + K}(P) = \translate{L_m}(P_1) + \translate{K+L_m}(Q)$
+    is a sum of two polynomials in $\Nat[\vec{X}]$.
+    Otherwise,
+    $\translate{L_m + K}(\restr{P}{\nu}) = \translate{K}(\translate{L_m}(\restr{P}{\nu}))$
+    belongs to $\Nat[\vec{X}]$ since the latter is stable under
+    positive translations.
+\end{proof}

### 2024-01-17 parts/prelim.tex
residual transducer??

@@
 \end{definition}
+
+Remark II.21. It follows from Remark II.6, [11, Proposition 6.1], and Theorem
+II.20, that Z-rational series of polynomial growth are exactly those computable
+by weigthed automata with coefficients in {0,1,−1} of polynomial ambiguity. We
+are not aware of a direct proof of this correspondence.
 
@@
 \end{example}
+
+Let us briefly state that one can decide whether a \kl{$\Rel$-polyregular
+function} $f$ is \kl{commutative}, ensuring that we are working in a decidable
+subclass of functions.
+
+\begin{lemma}
+    \label{decide-commutative:lemma}
+    One can decide if a function $f \in \ZPoly$
+    is \kl{commutative}.
+\end{lemma}
+\begin{proof}
+    Let $\Sigma$ be a finite alphabet, endowed with a
+    total ordering over its letters,
+    and let $f \colon \Sigma^* \to \Rel$ be a
+    \kl{$\Rel$-polyregular function}.
+    The map $\polysort \colon \Sigma^* \to \Sigma^*$
+    that sorts the letters in the input word $w$ according
+    to the chosen ordering is a \kl{polyregular function}
+    \cite[Proposition II.12]{LOPEZ23b}.
+    As a consequence,
+    $(f \circ \polysort) \in \ZPoly$.
+    Finally, $f$ is \kl{commutative} if and only
+    if $f = f \circ \polysort$, which is decidable
+    in the case of $\ZPoly$
+    \cite[Corollary II.24]{LOPEZ23b}, and more generally in the case 
+    of $\ZRat$ \cite[Corollary 3.6 p 38]{berstel2011noncommutative}.
+\end{proof}
+
+\cref{decide-commutative:lemma} does not state decidability in the case of
+\kl{$\Rel$-rational series}, and its proof relies on the stability of $\ZPoly$
+under the pre-composition by \kl{polyregular functions}, which does not hold
+for $\ZRat$. However, and for the sake of completeness, let us sketch a (more
+involved) proof that \kl{commutativity} is decidable for this larger class,
+using non-trivial results coming from topological things.
+
+\begin{lemma}
+    \label{decidable-commutative-rat:lemma}
+    Let $f \in \ZRat$, it is decidable whether
+    $f$ is \kl{commutative}.
+\end{lemma}
+\begin{proof}
+    Because $f$ is a \kl{$\Rel$-rational series},
+    there exists a $n \in \Nat$,
+    vectors $v_1$ and $v_2$,
+    and $n$ by $n$ integer matrices $\seqof{M_a}{a \in \Sigma}$,
+    such that
+    \begin{equation*}
+        \forall w \in \Sigma^*,
+        f(w) = {}^t v_1 \left(\prod_{i = 1}^{|w|} M_{w_i}\right) v_2 \quad .
+    \end{equation*}
+    To simplify notations, 
+    let us write $M_w \defined \prod_{i = 1}^{|w|} M_{w_i}$,
+    and 
+    $S \defined \setof{M_w}{w \in \Sigma^+}$ be the
+    semigroup generated by $\seqof{M_a}{a \in \Sigma}$.
+
+    The function $f$ is \kl{commutative} if and only if for all $u,v \in
+    \Sigma^+$, ${}^t v_1 (M_u M_v  - M_v M_u) v_2 = 0$. The above is a
+    polynomial equation involving the coefficients of $M_u$ and $M_v$.
+    Let us compute $\overline{S \times S}$ the Zariski closure of $S \times S$.
+    The above polynomial vanishes over $S \times S$ if and only
+    if it belongs to the Zariski closure, which is decidable
+    \cite{HROUPOWO18}.
+\end{proof}
 

### 2024-01-17 parts/starfree.tex
residual transducer??

@@
 
+\subsection{$\Nat$-polyregular functions}
+
 The goal of this section is to go beyond polynomials and characterize all
@@
 
+\begin{definition}
+    \label{k-combinatorial:def}
+    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
+    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
+    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
+    such that
+    for all
+    $\alpha_0, \dots, \alpha_k \in \Sigma^*$,
+    for all $u_1, \dots, u_k \in \Sigma^*$,
+    there exists a \kl{$\Nat$-rational polynomial} $P$
+    satisfying for all $X_1, \dots, X_k \geq \omega$:
+    \begin{equation*}
+        f
+        \left(
+            \alpha_0 \prod_{i = 1}^k u_i^{\omega X_i} \alpha_i
+        \right)
+        = 
+        P(X_1, \dots, X_k) \quad .
+    \end{equation*}
+\end{definition}
+
 \begin{theorem}
     \label{decidable-n-poly:thm}
-    Given a \kl{commutative}
-    \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to \Rel$,
-    one can decide if $f$ can be computed by a \kl{$\Nat$-polyregular function},
-    and effectively compute a representation of $f$.
+    Let $k \in \Nat$, and 
+    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
+    The following are equivalent:
+    \begin{enumerate}
+        \item $f$ is \kl{$(k,\Nat)$-combinatorial},
+        \item $f \in \NPoly$,
+    \end{enumerate}
+    Furthermore, the properties are decidable,
+    and one can effectively compute a representation of $f$.
 \end{theorem}
@@
 This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
-commutative case.
+commutative case. Informally, we state that \kl{commutative}
+\kl{$\Rel$-polyregular functions} are obtained by applying different
+\kl{$\Rel$-rational polynomials} based on the modulos...
+
+\AP Given $\omega \in \Nat$, let us write $\phi \colon \Nat \to \set{0, \dots,
+\omega} \times \set{0, \dots, \omega - 1}$ defined as $\phi(n) = (t, r)$ where
+$t = \min(n, \omega)$, and $r$ is the rest of the Euclidian division of $n$ by
+$\omega$.
+We let $\phi^\dagger \colon \Nat^k \to$.
 
@@
     Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
-    \kl{$\Rel$-polyregular function}. There exists $\omega \in \Nat$,
-    and polynomials $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{ 0 \leq n_a < \omega}$,
+    \kl{$\Rel$-polyregular function}. There exists a computable
+    $\omega \in \Nat$,
+    and computable 
+    polynomials $\seqof{P_{\rho}}{\rho \in \Delta_\omega}$,
     such that for all $w \in \Sigma^*$,
     \begin{equation*}
-        f(w) = P_{\seqof{\card[a]{w} [\omega]}{a \in \Sigma}}
-        (\seqof{\card[a]{w} / \omega}{a \in \Sigma}) \quad .
+        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) = P_{\phi(X_1, \dots, X_n)} (X_1 / \omega, \dots, X_n / \omega)
+        \quad .
     \end{equation*}
-    Which informally means that for every choice of rest modulo $\omega$
-    for every variable, there exists a polynomial describing the
-    behavior of $f$ when adding multiples of $\omega$ letters at a time.
-
-    Furthermore, the polynomials are computable from $f$.
 \end{lemma}
@@
 
-\begin{conjecture}
-    $\CorrectPoly$ is precisely composed of polynomials that
-    have non-negative coefficients in a \emph{binomial basis},
-    i.e., where polynomials are obtained
-    as sums and products of $\binom{X_i}{p_i}$.
-\end{conjecture}
-
 \subsection{Star-free $\Nat$-polyregular functions}
@@
 
-Let us now prove that the above characterizations of
-commutative
-$\Nat$-polyregular functions can be combined with the recent
-advances in the study of $\Rel$-polyregular functions 
-\cite{LOPEZ23b}
-to prove the following theorem.
+Let us now prove that the above characterizations of \kl{commutative}
+\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
+the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to prove the
+following theorem.
 
@@
     \label{zsf-npoly-nsf:thm}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be both a \kl{star-free $\Rel$-polyregular function}
-    and a \kl{commutative} \kl{$\Nat$-polyregular function}.
-    Then, $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}.
-    That is, the following equality holds:
+    Let $\Sigma$ be a finite alphabet and $k \in \Nat$.
+    Then,
     \begin{equation*}
         \ZCommut \cap \ZSF[k] \cap \NPoly
-        = \ZCommut \cap \ZSF \cap \NPoly[k]
-        = \ZCommut \cap \NSF
+        = \ZCommut \cap \NSF[k]
         \quad .
@@
     \label{nsf-zsf:conj}
-    In the general case, the following equation holds:
+    In the general case, the following equation holds for all $k \in \Nat$:
     \begin{equation*}
-        \ZSF \cap \NPoly
+        \ZSF[k] \cap \NPoly
         = 
-        \NSF
+        \NSF[k]
         \quad .
@@
 
+\textbf{TODO: add the growth rate here, we are actually minimising it}.
 \begin{corollary}
-    One can decide if a \kl{commutative} \kl{ $\Rel$-polyregular function}
+    One can decide if a \kl{commutative} \kl{$\Rel$-polyregular function}
     can be realized by a \kl{star-free $\Nat$-polyregular function},
@@
 \end{corollary}
+\begin{proof}
+    From \cite[Theorem V.13]{LOPEZ23b}, it is decidable whether
+    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
+    it is decidable whether $f \in \NPoly$.
+    Furthermore, one can enumerate functions in $\NSF[k]$,
+    and check equivalence to effectively represent $f$ as a
+    \kl{star-free $\Nat$-polyregular function}.
+\end{proof}
 
@@
 \end{proof}
+
+\subsection{Non-negative Residual Transducers}
+
+The goal of the upcoming section is to provide a notion of residual transducer
+for \kl{commutative} \kl{$\Nat$-polyregular functions}.
+
+\begin{fact}
+    For every $K \in \Nat$, and every $n \in \Nat$,
+    $\Nat^n$ is a \kl{well-quasi-ordering}
+    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
+    if and only if $\vec{u} = \vec{v}$ or
+    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
+    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
+\end{fact}
+\begin{proof}
+    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
+    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
+    one can extract an infinite monotone subsequence for
+    the product ordering.
+    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
+    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
+    and the original sequence was good. Or for all $i < j$,
+    we have $\vec{u}_i$ that is strictly smaller than 
+    $\vec{u}_j$ for the product ordering. As a consequence,
+    $\vec{u}_0 \leq \vec{u}_K$.
+\end{proof}
+
+\begin{fact}
+    Let $P \in \Nat[X,\vec{Y}]$, and $K,L \in \Nat$,
+    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \Nat[X, \vec{Y}]$
+    whenever $L \geq K$.
+\end{fact}
+\begin{proof}
+    We can write
+    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
+    Now,
+    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
+    is precisely
+    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$,
+    remaining with positive coefficients.
+\end{proof}
+
+\begin{fact}
+    Let $P \in \CorrectPoly[X,\vec{Y}]$, and $K,L \in \Nat$,
+    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \CorrectPoly$
+    whenever $L \geq K$.
+\end{fact}
+\begin{proof}
+    We can write
+    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
+    Now,
+    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
+    is precisely
+    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$.
+    Now, $P_i(\vec{Y})$
+    remaining with positive coefficients.
+\end{proof}
+
+\begin{fact}
+    Let $P \in \CorrectPoly[X_1, \dots, X_n]$. 
+    For all
+    $\vec{u} \leq \vec{v} \in \Nat^n$,
+    $Q(X_1, \dots, X_n)
+    \defined 
+    P(X_1 + v_1, \dots, X_n + v_n)
+    -
+    P(X_1 + u_1, \dots, X_n + u_n)$
+    belongs to $\CorrectPoly[X_1, \dots, X_n]$.
+\end{fact}
+\begin{proof}
+    Let us consider $K \in \Nat$ such that for all partial valuations
+    $\nu \colon \vec{X} \topartial \Nat$,
+    $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
+
+    Remark that
+    \begin{align*}
+        Q
+        &= 
+        P(X_1 + v_1, \dots, X_n + v_n)
+        -
+        P(X_1 + u_1, \dots, X_n + u_n) \\
+        &=
+        P(X_1 + v_1, X_2 + v_2 , \dots, X_n + v_n) - P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) \\
+        &+
+        P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) - P(X_1 + u_1, X_2 + u_2, \dots, X_n + v_n) \\
+        &+ \dots
+    \end{align*}
+
+    Let $\nu \colon \vec{X} \topartial \Nat$
+    be a partial valuation, let us show that
+    $\translate{K}(\restr{Q}{\nu}) \in \Nat[\vec{X}]$.
+
+    \begin{align*}
+        \translate{K}(\restr{Q}{\nu})
+        &= 
+        \translate{K}(\restr{P(X_1 + v_1, \dots, X_n + v_n)}{\nu})
+        -
+        \translate{K}(\restr{P(X_1 + u_1, \dots, X_n + u_n)}{\nu})
+    \end{align*}
+\end{proof}
+
+\begin{lemma}
+    Let $k \in \Nat$,
+    $f \in \NPoly[k]$,
+    and $\seqof{u_i}{i \in \Nat}$ be a sequence of
+    elements in $\Nat^n$. There exists $i < j$
+    such that
+    $f(u_j \cdot) - f(u_i \cdot) \in \NPoly[k-1]$.
+\end{lemma}
+
+\begin{defintion}
+    Residual $\Nat$-transducer.
+\end{definition}
+
+\begin{lemma}
+    Computable 
+\end{lemma}
+
+\begin{theorem}
+    If and only if this one is counter-free.
+\end{theorem}
+
+Note that it is not needed for decidability, as we are able to use directly the
+results on $\ZSF$, but this provides an independent proof, and highlights the
+versatility of this notion of "residual transducer".

### 2024-01-17 pnrs.tex
residual transducer??

@@
 
-\section{TODOs}
-
-\begin{itemize}
-    \item Use $\Nat$-polyregular functions all the time, and just 
-        in the introduction talk about $\Nat$-rational series.
-    \item Prove that $\Nat$-rational series of polynomial 
-        growth are $\Nat$-polyregular functions, maybe using
-        the results about ambiguity in automata?
-        Theorem 5.22
-    \item Define commutativity by a map 
-        $\commute \colon \Sigma^* \topartial \Nat^{\Sigma}$.
-    \item Define "commutative functions" and its symbol.
-    \item Define "computed by" (like, recognised by) for commutative 
-        functions.
-    \item Change \cref{corrected-version:thm} to use directly
-        star-free!
-    \item Clear up the computability problems.
-    \item Polynomial + Npolyregular => star free?
-    \item Polynomial + N rational => N polyregular?
-    \item cite the paper of Quantitative Monadic Second-Order Logic
-        about why it is interesting class.
-    \item Example of polyregular functions
-    \item Example of non-polyregular functions
-    \item Expand a bit on the notion of "derivative"
-        and divisibility of monomials.
-    \item What about matrices ? They are not minimisable... so we do not know
-        about their eigenvalues ?
-    \item What about the residual transducer ?
-    \item What about weigthed automata? Does it mean something?
-        -> we have unary input, so the automata really is an automata.
-        -> we output numbers in $\Nat$
-        -> residuals are what? totally ordered?
-        -> recursive calls?
-        -> if the function computes polynomials in $\Nat$, we can build
-        the usual thing? and then we add "a few steps before"?
-
-    \item Is the proof simpler by assuming
-        $P(X + K) \in \Nat[X]$? This is way better.
-
-    \item Prove decidability of the property.
-
-    \item Show the equivalence with binomial
-
-    \item What about 
-\end{itemize}
-
 \input{parts/intro.tex}

### 2024-01-17 parts/starfree.tex
ultimate polynomiality is a lie

@@
 for \kl{commutative} \kl{$\Nat$-polyregular functions}.
+This will be based on another characterization of $\CorrectPoly$
+using partial derivatives. Namely, we define
+
+\begin{definition}
+    For every polynomial $P \in \Rel[X \vec{Y}]$, we define 
+    the partial derivative with respect to $X$ as follows:
+    \begin{equation*}
+        \partial_X P \defined P(X + 1, \vec{Y}) - P(X, \vec{Y}) \quad .
+    \end{equation*}
+\end{definition}
 
 \begin{fact}
-    For every $K \in \Nat$, and every $n \in \Nat$,
-    $\Nat^n$ is a \kl{well-quasi-ordering}
-    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
-    if and only if $\vec{u} = \vec{v}$ or
-    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
-    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
+    Let $P \in \CorrectPoly[X_1, \dots, X_n]$, then
+    for all $1 \leq i \leq n$,
+    $\partial_{X_i} P \in \CorrectPoly$.
 \end{fact}
 \begin{proof}
-    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
-    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
-    one can extract an infinite monotone subsequence for
-    the product ordering.
-    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
-    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
-    and the original sequence was good. Or for all $i < j$,
-    we have $\vec{u}_i$ that is strictly smaller than 
-    $\vec{u}_j$ for the product ordering. As a consequence,
-    $\vec{u}_0 \leq \vec{u}_K$.
+    We prove the result by induction on the number of variables.
+    Let $\nu \colon \vec{X} \topartial \Nat$.
+    Two cases, either $X_i \in \dom(\nu)$,
+    in which case, we remark that
+    we can split $\nu$ into two valuations,
+    and that coefficients are good.
+
+    Otherwise, evaluation commutes with partial,
+    and we conclude by induction hypothesis.
 \end{proof}
+
+\begin{remark}
+    The converse does not hold,
+    for instance $P(X) \defined X - 1$ has all its derivatives
+    that are in $\CorrectPoly$, but is not itself in $\CorrectPoly$.
+\end{remark}
 
@@
     remaining with positive coefficients.
+\end{proof}
+\begin{fact}
+    For every $K \in \Nat$, and every $n \in \Nat$,
+    $\Nat^n$ is a \kl{well-quasi-ordering}
+    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
+    if and only if $\vec{u} = \vec{v}$ or
+    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
+    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
+\end{fact}
+\begin{proof}
+    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
+    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
+    one can extract an infinite monotone subsequence for
+    the product ordering.
+    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
+    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
+    and the original sequence was good. Or for all $i < j$,
+    we have $\vec{u}_i$ that is strictly smaller than 
+    $\vec{u}_j$ for the product ordering. As a consequence,
+    $\vec{u}_0 \leq \vec{u}_K$.
 \end{proof}
@@
 
+\begin{definition}
+    Residual $\Nat$-transducer.
+\end{definition}
+
+\begin{lemma}
+    If $P \in \CorrectPoly$, then it is computable
+    by a $\Nat$-residual transducer.
+\end{lemma}
+
+\begin{theorem}
+    The following are equivalent for \kl{commutative}.
+    \begin{itemize}
+        \item $f$ is in $\NSF$,
+        \item The residual automata of $f$ has no counters
+    \end{itemize}
+\end{theorem}
+
+\section{Beyond commutativity}
+
+\begin{lemma}
+    Let $f \in \NPoly[k]$,
+    and $h \colon \Nat^p \to \Sigma^*$ be a polyregular function.
+    Then,
+    $f \circ h$ is a \kl{commutative} \kl{polyregular function},
+    and
+    the set $\setof{ f(u \cdot) \circ h }{u \in \Sigma^* }$
+    is a \kl{well-quasi-ordering}
+    for the order
+    $g_1 \leq g_2$ iff 
+    $g_2 - g_1 \in \NPoly[k-1]$.
+\end{lemma}
+
+\begin{claim}
+    Let $k \in \Nat$,
+    there exists a finite family $\mathcal{F}_k$
+    of \kl{commutative} \kl{polyregular functions}
+    such that the following are equivalent
+    for all $f \in \ZPoly[k]$:
+    \begin{itemize}
+        \item $f \in \NPoly[k]$,
+        \item For all $h \in \mathcal{F}_k$,
+            $(f \circ h)$ is a \kl{commutative}
+            \kl{$\Nat$-polyregular function}.
+    \end{itemize}
+\end{claim}
+
 \begin{lemma}
@@
 
-\begin{defintion}
-    Residual $\Nat$-transducer.
-\end{definition}
-
 \begin{lemma}
-    Computable 
+    If $f \in \NPoly[k]$, then the algorithm computes
+    a residual transducer of $f$.
 \end{lemma}
@@
 \begin{theorem}
-    If and only if this one is counter-free.
+    The following are equivalent for $f \in \NPoly$.
+    \begin{itemize}
+        \item The function $f \in \NSF$,
+        \item The residual transduucer of $f$ has no counters,
+        \item For all commutative polyregular functions $h$,
+            $f \circ h$ is a \kl{commutative}
+            \kl{star-free $\Nat$-polyregular function},
+        \item $f$ is \kl{ultimately polynomial}.
+    \end{itemize}
 \end{theorem}

### 2024-01-18 globals/metadata.tex
acm style
re-organising transducers and such

--- 2023-12-22 globals/metadata.tex
@@
-\title{A corrected version of Karhumäki’s theorem
-       on polynomial $\Nat$-rational series}
+\title{
+    Star-free $\Nat$-polyregular functions:
+    the commutative case.
+}
 \author{
-    Lopez Aliaume
+    Aliaume Lopez
 }
+\affiliation{%
+  \institution{Univeristy of Warsaw}
+  \city{Warsaw}
+  \country{Poland}}
+\email{ad.lopez@uw.edu.pl}
+
 \date{\today}
@@
 \newcommand{\acknowledge}{
-    I would like to thank Gaëtan Douéneau-Tabot
-    for mentionning this problem, checking 
+    I would like to thank 
+    Anonymous Author(s)
+    %Gaëtan Douéneau-Tabot
+    for mentioning this problem, checking 
     my counter-example, and supporting me in publishing
@@
     \begin{abstract}
-        We propose a short counter-example to a
-        previous result on commutative $\Nat$-rational series
-        by Karhumäki, and provide a proof for
-        the corrected statement.
-        This opens the way to the decidability of
-        $\Nat$ polyregular functions 
-        among $\Rel$ porylegular functions, which is
-        an ongoing research.
+        This paper answers two questions about 
+        commutative unary output polyregular functions, called
+        $\Nat$-polyregular functions, that forms a proper subset
+        of $\Nat$-rational series.
+        We show that given a $\Rel$-rational series,
+        it is decidable whether it is computable via a commutative
+        $\Nat$-polyregular function, and provide a counter-example
+        to the theorem of Karhumäki that studied the same question
+        in the case of polynomials.
+
+        We also prove that it is decidable whether a commutative $\Nat$-polyregular
+        function is star-free, by proving the stronger statement
+        that star-free $\Rel$-polyregular functions that are
+        $\Nat$-polyregular are computable using a star-free $\Nat$-polyregular function.
+        This answers positively to previous conjectures on $\Nat$-polyregular functions,
+        and the study of the commutative case seems to be useful
+        towards a proof in the non-commutative case.
     \end{abstract}
 }
+
+\setcopyright{acmlicensed}
+\copyrightyear{2018}
+\acmYear{2018}
+\acmDOI{XXXXXXX.XXXXXXX}

### 2024-01-18 parts/correction.tex
re-organising transducers and such

@@
 %! lang = en-US
-\section{The Corrected Theorem}
+\subsection{The Corrected Theorem}
 \label{sec:proof}
@@
 that $\CoveredPoly$ is not stable under fixing indeterminates, while
-\kl{$\Nat$-polyregular functions} are stable under the residual operation
-$u^{-1} \colon f \mapsto f(u \cdot)$. In this section, we prove that closing
+\kl{$\Nat$-polyregular functions} are. In this section, we prove that closing
 $\CorrectPoly$ under assignments of variable is enough to recover from
-\cref{karh:thm}.
+\cref{karh:thm}. Because the previous attempt to characterize this class of
+polynomials did not hold, this part of the paper will try to be extremely
+precise.
 
@@
         f\left(
-            \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)
+            \alpha_0 \prod_{i = 1}^p u_i^{\omega \times X_i} \alpha_i
         \right)
@@
 
-\begin{lemma}
+\begin{corollary}
     \label{n-rat-correct:lem}
@@
     $P \in \CorrectPoly$.
-\end{lemma}
+\end{corollary}
 \begin{proof}
@@
 
-Let us also assert that polynomials that have all coefficients in $\Nat$ are
-\kl{$\Nat$-rational polynomials}. In particular, \kl{non-negative}
-\kl{monomials} are \kl{$\Nat$-rational polynomials}, which will be a base case
-in an upcoming induction proving \cref{lem:correct-to-n-rat}.
-
-\begin{lemma}
-    \label{n-poly-n-poly:example}
-    Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
-\end{lemma}
-\begin{proof}
-    We prove it by remarking that $\NPoly$
-    is stable under non-negative linear combinations
-    and contains the functions ...
-\end{proof}
-
 \AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
@@
 we write $\translate{K}$ for the \intro{translation function} that maps a
-polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 - K, \dots,
-X_n - K)$.
+polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 + K, \dots,
+X_n + K)$.
 
@@
 polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
-functions} to \kl{star-free $\Nat$-polyregular functions}. This has to be
-compared with the characterization of \kl{star-free $\Rel$-polyregular
-functions} in terms of polynomials \cite[Theorem??]{LOPEZ23b},
-that we will re-introduce in \cref{star-free:sec}.
+functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
+surprising, given the conjecture that \kl{ultimate polynomial}
+\kl{$\Nat$-polyregular functions} (re-introduced in \cref{star-free:sec}) were
+conjectured to be star free in \cite{LOPEZ23b}.
 
@@
 
-The following two lemmas show how the \kl{discrete differentiation} operator
+The following \cref{derivation-simplifies:lemma,derivation-covers:lemma} 
+show how the \kl{discrete differentiation} operator
 interacts with \kl{maximal monomials} with respect to the \kl{Hoare ordering}:
@@
 \begin{fact}
-    Let $M,T$ be two \kl{monomials}, such that
+    Let $K > 0$,
+    and let $M,T$ be two \kl{monomials}, such that
     $T$ \kl{strictly divides} $M$.
@@
     For all $P,Q \in \Rel[\vec{X}]$ that are non-constant,
-    for all $K \in \Nat$,
+    for all $K > 0$,
     if $\MaximalMonomials(P) \hoarele \MaximalMonomials(Q)$,
@@
 While \cref{lem:correct-to-n-rat} provides effective conversion, it does not
-explicitly state that the membership is decidable. However, it is not too
-difficult to remark that $\CorrectPoly$, the proof scheme can be followed to
-obtain decidability. For the sake of completeness and readability,
-we extract this fact in the following lemma.
+explicitly state that the membership is decidable to keep the proof clearer. A
+similar proof scheme can be followed to conclude that membership is decidable.
 
@@
     \label{decidability-correct:lem}
-    Membership in $\CorrectPoly$ is decidable for
-    polynomials in $\Rel[\vec{X}]$.
+    It is decidable whether a polynomial 
+    $P \in \Rel[\vec{X}]$
+    belongs to $\CorrectPoly$.
 \end{lemma}

### 2024-01-18 parts/counter.tex
re-organising transducers and such

@@
 %! lang = en-US
-\section{The Counter Example}
-\label{sec:c-example}
+\section{Polynomials}
 
-We will start by formally stating the result of \citeauthor{KARH77} in terms of
-polynomials and commutative $\Nat$-rational series, and then introduce the
-necessary definitions. The characterization of commutative $\Nat$-rational
-series computing polynomial functions is described in terms of a class
+The goal of this section is to provide semantic characterizations of
+\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is a
+\kl{$\Nat$-rational polynomial}, and if so, explicitly compute a
+\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$.
+
+It is an easy check that polynomials with coefficients in $\Nat$ are
+\kl{$\Nat$-rational polynomials} (\cref{n-poly-n-poly:example}). However,
+\cref{negative-but-npoly:ex} provides an example of a polynomial with negative
+coefficients that is a \kl{$\Nat$-rational polynomial}. As a consequence, the
+class of interest strictly contains $\Nat[\vec{X}]$.
+
+\begin{lemma}
+    \label{n-poly-n-poly:example}
+    Let $P \in \Nat[\vec{X}]$. Then, $P$
+    is a \kl{$\Nat$-rational polynomial}.
+\end{lemma}
+\begin{proof}
+    We prove it by remarking that $\NPoly$
+    is stable under non-negative linear combinations
+    and contains the functions ...
+\end{proof}
+
+\begin{example}
+    \label{negative-but-npoly:ex}
+    The polynomial $P(X) \defined X^2 -2X + 2$ is
+    a \kl{$\Nat$-rational polynomial}.
+\end{example}
+\begin{proof}
+    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
+    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
+    It is clear that $f \in \NPoly$, and that 
+    $P$ is \kl{represented} by
+    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
+\end{proof}
+
+Let us now introduce the characterization of \kl{$\Nat$-rational polynomials}
+given by \citeauthor{KARH77}, which is stated using a class
 $\CoveredPoly[\vec{X}]$ of polynomials with multiple indeterminate.
@@
 
+\subsection{The counter example}
+\label{sec:c-example}
+
 Let us now introduce the counter example that was found trying to generalize
@@
 we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
-polynomial with at most two indeterminate.
+polynomial with at most two indeterminate. In particular, the examples
+appearing in \cite{KARH77} are not invalidated, as they all use at most two
+indeterminates. Furthermore, the converse implication, that is, that
+\kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is correct, and can
+even be strengthened as we will see in \cref{corrected-version:thm}.
 
@@
 \end{corollary}
-
-As we will see in the next section, the characterization of \kl{$\Nat$-rational
-polynomials} using the set $\CoveredPoly$ holds when the polynomials have at
-most $2$ indeterminate. This shows that the examples in \cite{KARH77} are not
-invalidated, as they all use at most two indeterminates. Furthermore, the
-converse implication, that is, that \kl{$\Nat$-rational polynomials} belong to
-$\CoveredPoly$ is correct, and can even be strengthened as we will see in
-\cref{corrected-version:thm}.

### 2024-01-18 parts/intro.tex
re-organising transducers and such

@@
 e.g.][]{bojanczyk2019string}. In this paper, we will focus on \kl{polyregular
-functions}, that shares with regular languages the characterizations in terms
-of logic (via $\MSO$ interpretations), automata (via pebble automata or SST
-\cite{}), and adds new bridges with programming languages (lambda-terms)
-for-transducers. Although the class of "star-free" polyregular functions is
-defined and similarly characterized using first-order logic.
+functions}, whose class is denoted $\Poly$, and that shares with regular
+languages the characterizations in terms of logic (via $\MSO$ interpretations),
+automata (via pebble automata), and adds new bridges with programming languages
+via its relationships with lambda-terms and so-called for-transducers. As for
+regular languages, the class $\SF$ of \intro{star-free polyregular function}
+can equivalently be introduced in terms of first-order definability,
+counter-free pebble transducers, and suitable restrictions of for-tranducers.
 
@@
 Indeed, for sequential functions that admit a canonical object, aperiodicity is
-decidable \cite{choffrut03}, and more recently, it aperiodicity has been proven
+decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
 decidable for rational functions
@@
 decide the aperiodicity of polyregular functions.
-Notice that unary outputs in $\Rel$ and $\Nat$
-are therefore special kind of weighted automata,
-and that questions arising from the weighted automata
-theory (good first order logic?) are also connected.
-\cite{droste2019aperiodic,reutenauer_series_1980}.
 
-Recently, the introduction of $\Rel$-output polyregular functions, called
-\kl{$\Rel$-polyregular functions} (and more generally, commutative invertible
-outputs) allowed to solve the general problem \cite{LOPEZ23b}. Note that
-$\Rel$-output generalizes $\Nat$-output, but somehow has a combinatorial
-description that is farther away from traditional \kl{polyregular functions}
-because of the invertible outputs. The proof relies on the construction of a
-canonical object essentially based on differences between functions, which is
-not applicable to unary output. Note that already for rational series, the
-problem is open.
+Recently, aperiodicity has been shown decidable over a restricted class of
+\kl{polyregular functions} called \kl{$\Rel$-polyregular functions} and denoted
+by $\ZPoly$ \cite{LOPEZ23b}.  These functions generalise \kl{$\Nat$-polyregular
+functions} by allowing negative outputs, and therefore enabling a limited form
+of backtracking. The decidability of aperiodicity for \kl{$\Rel$-polyregular
+functions} relies on the construction of a canonical object called the
+\intro{residual transducer}, the latter being essentially based on differences
+between functions, crucially leveraging negative outputs.
+If the proof method does not carry from $\Rel$-output to $\Nat$-output
+functions, it was conjectured that the semantic characterization of \kl{star
+free $\Rel$-polyregular functions} ($\ZSF$) would continue to characterize
+\kl{star-free $\Nat$-polyregular functions} ($\NSF$), and in particular that
+$\NPoly \cap \ZSF = \NSF$.
 
-We write $\Poly$ for the class of \kl{polyregular functions},
-$\SF$ for the class of \kl{star-free polyregular functions},
-$\NRat$ for the class of \kl{$\Nat$-rational series},
-and $\ZRat$ for the class of \kl{$\Rel$-rational series}.
-Arrows denote strict inclusions, and effectiveness
-(both in terms of decidability and of effective representation)
-is represented by thick double arrows.
-Inclusions that are suspected to be effective have a question mark.
-It is also conjectured that $\NPoly \cap \ZSF = \NSF$.
+Let us remark that, \kl{$\Nat$-polyregular functions} and
+\kl{$\Rel$-polyregular functions} are proper subsets respectively of
+\kl{$\Nat$-rational series} ($\ZRat$) and of \kl{$\Rel$-rational series}
+($\NRat$). For these larger classes of functions, several competing notions of
+aperiodicity exist \cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b},
+and decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
+Furthermore, the decidability of $\NRat$ among $\ZRat$ is also a longstanding
+open problem, that was to the knowledge of the authors, only solved in the case
+of \emph{polynomials} by \textcite{KARH77}.
+
+We refer the reader to \cref{previously-known-inclusions:fig} to get an
+overview of the different classes of functions appearing in the paper. Arrows
+denote strict inclusions, and effectiveness (both in terms of decidability and
+of effective representation) is represented by thick double arrows. Inclusions
+that are suspected to be effective are represented using a dashed arrow
+together with a question mark.
 
@@
     \caption{
-        Decidability and inclusions of classes of functions. 
+        Decidability and inclusions of classes of functions,
+        arranged along two axes. The first one is the complexity
+        of the output alphabet ($\Nat$, $\Rat$, $\Sigma$). The second
+        one is the allowed computational power
+        (star-free polyregular functions, polyregular functions, 
+        rational series).
     }
+    \label{previously-known-inclusions:fig}
 \end{figure}
 
+\paragraph{Contributions.} We prove that for \kl{commutative} functions, all
+the membership problems involving the classes $\ZSF$, $\ZPoly$, $\NSF$, and
+$\NPoly$ are all decidable (this is the lower left square of inclusions in
+\cref{previously-known-inclusions:fig}), and effective conversion algorithms
+are provided. Furthermore, we show that in the commutative case, the conjecture
+stating that $\ZSF \cap \NPoly = \NSF$ holds.
+
+These results are obtained by completely understanding which \emph{polynomials}
+are computable using \kl{$\Nat$-polyregular functions}, and lifting these
+results from polynomials to \kl{$\Rel$-polyregular functions}. This study
+invalidates the main theorem of \cite{KARH77}, for which we provide an explicit
+counter-example.
+
+\begin{itemize}
+    \item We show that commutativity is decidable.
+    \item We provide a candidate semantic characterization
+    \item We prove that it is correct for commutative inputs
+    \item We provide a counter-example to the theorem of nanania.
+    \item We show that effective procedures are possible for all arrows 
+        and answer positively to the conjecture of XX
+        in the case of commutative inputs.
+\end{itemize}
 It was conjectured in a thesis that some particular form of polynomials
-would characterize star-free functions.This paper follows the line of
+would characterize star-free functions.
+
+This paper follows the line of
 research on \kl{polyregular functions} that focuses on semantic
@@
 
+\cref{thm:counter-example}
+\cref{corrected-version:thm}
+
+\cref{decidable-n-poly:thm}
+\cref{zsf-npoly-nsf:thm}
+
 \hrule
 
-\paragraph{Contributions.} 
-\begin{itemize}
-    \item We provide a candidate semantic characterization
-    \item We prove that it is correct for commutative inputs
-    \item We provide a counter-example to the theorem of nanania.
-    \item We show that effective procedures are possible for all arrows 
-        and answer positively to the conjecture of XX
-        in the case of commutative inputs.
-\end{itemize}
-
 \paragraph*{Outline of the paper.}
+In \cref{preliminaries:sec}, we introduce the computational models
+and prove that 
 \begin{itemize}

### 2024-01-18 parts/prelim.tex
re-organising transducers and such

@@
 
+Let us briefly remark that the characterizations of subclasses of
+\kl{polyregular functions} that are obtained in practice can be obtained by
+pre-composition with a \kl{commutative} \kl{regular function}.
+
+\begin{remark}
+    Let $f \in \ZRat$, and $k \in \Nat$,
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \ZPoly[k]$,
+        \item For every \kl{commutative} \kl{polyregular function} $h$
+            of \kl{growth rate} $l \in \Nat$,
+            $(f \circ h) \in \ZPoly[k+l-1]$.
+    \end{enumerate}
+\end{remark}
+
+\begin{remark}
+    Let $f \in \ZPoly$.
+    The following are equivalent:
+    \begin{enumerate}
+        \item $f \in \ZSF$,
+        \item For every \kl{commutative} \kl{star-free polyregular function} $h$,
+            $(f \circ h) \in \ZSF$.
+    \end{enumerate}
+\end{remark}
+
 \AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function

### 2024-01-18 parts/starfree.tex
re-organising transducers and such

@@
 \label{sec:deciding}
+
+Let us now move on from polynomials to general \kl{$\Rel$-polyregular
+functions}. As we will prove, \kl{commutativity} reduces $\ZPoly$ to a
+combination of a finite number of polynomials, arranged using modulo counting.
+We will first decide $\NPoly$ inside $\ZPoly$, and then leverage this to decide
+$\NSF$ inside $\NPoly$, incidentally proving that $\ZSF \cap \NPoly = \NSF$.
 
@@
         \item $f$ is \kl{$(k,\Nat)$-combinatorial},
-        \item $f \in \NPoly$,
+        \item $f \in \NPoly[k]$,
     \end{enumerate}
@@
 This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
-commutative case. Informally, we state that \kl{commutative}
-\kl{$\Rel$-polyregular functions} are obtained by applying different
-\kl{$\Rel$-rational polynomials} based on the modulos...
+commutative case.
 
@@
 $t = \min(n, \omega)$, and $r$ is the rest of the Euclidian division of $n$ by
-$\omega$.
-We let $\phi^\dagger \colon \Nat^k \to$.
+$\omega$. We let $\phi^\dagger \colon \Nat^k \to$.
 
@@
 \subsection{Non-negative Residual Transducers}
+
+\begin{definition}
+    Partial applications $\app{f}{u} \defined w \mapsto f(uw)$.
+\end{definition}
+
+\begin{definition}
+    Residuals $\Res(f) \defined \setof{ \app{f}{u} }{u \in \Sigma^*}$.
+\end{definition}
+\begin{definition}
+    We say that $f \natleq[k] g$ if and only if $g - f \in \NPoly[k]$.
+\end{definition}
 

### 2024-01-22 parts/correction.tex
update proofs and paper

@@
         f\left(
-            \alpha_0 \prod_{i = 1}^p u_i^{\omega \times X_i} \alpha_i
+            \alpha_0 \prod_{i = 1}^p u_i^{\omega X_i} \alpha_i
         \right)
@@
 
+\begin{fact}
+    \label{maxi-monomials-submodular:fact}
+    Let $P, Q$ be two polynomials.
+    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
+    \cup \MaximalMonomials(Q)$.
+\end{fact}
+
 The following \cref{derivation-simplifies:lemma,derivation-covers:lemma} 
@@
 it extracts \emph{submaximal} monomials from a given polynomial.
+
+\begin{fact}
+    \label{derivation-monomials:fact}
+    Let $K > 0$,
+    and let $M,T$ be two \kl{monomials}, such that
+    $T$ \kl{strictly divides} $M$.
+    Then,
+    all \kl{monomials} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
+    and 
+    $T$ \kl{divides} some \kl{maximal monomial}
+    of $\Diff{K}{M}$, which has a coefficient
+    that is a multiple of $K$.
+\end{fact}
+\begin{proof}
+    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
+    where $k_1, \dots, k_n > 0$ and $n > 0$.
+    Let us compute
+    \begin{align*}
+        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
+                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
+    \end{align*}
+    Therefore, 
+    the \kl{maximal monomials} of $\Diff{K}{M}$
+    are precisely
+    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
+    which \kl{strictly divide} $M$.
+    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
+    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
+    for all $1 \leq i \leq n$, $t_i \leq k_i$,
+    and there exists $1 \leq p \leq n$ such that$ t_j < k_j$.
+    In particular, $T$ \kl{divides}
+    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
+\end{proof}
 
@@
 \begin{proof}
-    By linearity of $\Diff{K}{P}$ (\cref{discrete-deriv-linear:fact}), 
-    it suffices to prove the result
-    for \kl{monomials} $S \in \Monomials$.
+    The result holds for \kl{monomials}
+    thanks to \cref{derivation-monomials:fact}.
+    It extends by linearity to polynomials 
+    thanks to \cref{maxi-monomials-submodular:fact,discrete-deriv-linear:fact}.
 \end{proof}
-
-\begin{fact}
-    Let $K > 0$,
-    and let $M,T$ be two \kl{monomials}, such that
-    $T$ \kl{strictly divides} $M$.
-    Then,
-    $T$ \kl{divides} some \kl{maximal monomial}
-    of $\Diff{K}{M}$.
-\end{fact}
-
-\begin{lemma}
-    \label{derivation-covers:lemma}
-    For all $P,Q \in \Rel[\vec{X}]$ that are non-constant,
-    for all $K > 0$,
-    if $\MaximalMonomials(P) \hoarele \MaximalMonomials(Q)$,
-    then
-    $\MaximalMonomials(P) \hoareleq \MaximalMonomials(\Diff{K}{Q})$.
-\end{lemma}
 
@@
     such that
-    $(\Diff{K}{P_1} + P_2) \in \CorrectPoly$.
+    $Q \defined (\Diff{K}{P_1} + P_2) \in \CorrectPoly$.
 \end{lemma}
+\begin{proof}
+    If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
+    hence for $K = 0$, $Q \in \CorrectPoly$.
+    Otherwise, let $\nu \colon \vec{X} \topartial \Nat$,
+    and let us analyze the \kl{maximal monomials} $T$
+    of $\restr{Q}{\nu}$.
+    Because $T$ is a \kl{monomial} of $\restr{Q}{\nu}$,
+    it \kl{divides} some \kl{monomial} in $Q$, and
+    using \cref{derivation-monomials:fact},
+    it \kl{divides} some \kl{monomial} $M$ in 
+    $\Diff{K}{P_1}$.
+    Now, the monomial $T$ is computed as follows:
+    \begin{align*}
+        T &= 
+        \sum_{S \in \Monomials(Q), T \divides S}
+        \restr{S}{\nu} \\
+          &= 
+          \sum_{S \in \Monomials(\Diff{K}{P_1}), T \divides S}
+            \restr{S}{\nu}
+        + 
+          \sum_{S \in \Monomials(P_2), T \divides S}
+            \restr{S}{\nu}
+        \\
+          &\geq
+          \sum_{S \in \Monomials(\Diff{K}{P_1}), T \divides S}
+            \restr{S}{\nu}
+          - 
+            \left|
+          \sum_{S \in \Monomials(P_2), T \divides S}
+            \restr{S}{\nu}
+            \right|
+    \end{align*}
+
+\end{proof}
 

### 2024-01-22 parts/starfree.tex
update proofs and paper

--- 2024-01-17 parts/starfree.tex
@@
 \end{proof}
-
-\subsection{Non-negative Residual Transducers}
-
-The goal of the upcoming section is to provide a notion of residual transducer
-for \kl{commutative} \kl{$\Nat$-polyregular functions}.
-
-\begin{fact}
-    For every $K \in \Nat$, and every $n \in \Nat$,
-    $\Nat^n$ is a \kl{well-quasi-ordering}
-    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
-    if and only if $\vec{u} = \vec{v}$ or
-    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
-    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
-\end{fact}
-\begin{proof}
-    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
-    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
-    one can extract an infinite monotone subsequence for
-    the product ordering.
-    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
-    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
-    and the original sequence was good. Or for all $i < j$,
-    we have $\vec{u}_i$ that is strictly smaller than 
-    $\vec{u}_j$ for the product ordering. As a consequence,
-    $\vec{u}_0 \leq \vec{u}_K$.
-\end{proof}
-
-\begin{fact}
-    Let $P \in \Nat[X,\vec{Y}]$, and $K,L \in \Nat$,
-    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \Nat[X, \vec{Y}]$
-    whenever $L \geq K$.
-\end{fact}
-\begin{proof}
-    We can write
-    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
-    Now,
-    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
-    is precisely
-    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$,
-    remaining with positive coefficients.
-\end{proof}
-
-\begin{fact}
-    Let $P \in \CorrectPoly[X,\vec{Y}]$, and $K,L \in \Nat$,
-    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \CorrectPoly$
-    whenever $L \geq K$.
-\end{fact}
-\begin{proof}
-    We can write
-    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
-    Now,
-    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
-    is precisely
-    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$.
-    Now, $P_i(\vec{Y})$
-    remaining with positive coefficients.
-\end{proof}
-
-\begin{fact}
-    Let $P \in \CorrectPoly[X_1, \dots, X_n]$. 
-    For all
-    $\vec{u} \leq \vec{v} \in \Nat^n$,
-    $Q(X_1, \dots, X_n)
-    \defined 
-    P(X_1 + v_1, \dots, X_n + v_n)
-    -
-    P(X_1 + u_1, \dots, X_n + u_n)$
-    belongs to $\CorrectPoly[X_1, \dots, X_n]$.
-\end{fact}
-\begin{proof}
-    Let us consider $K \in \Nat$ such that for all partial valuations
-    $\nu \colon \vec{X} \topartial \Nat$,
-    $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
-
-    Remark that
-    \begin{align*}
-        Q
-        &= 
-        P(X_1 + v_1, \dots, X_n + v_n)
-        -
-        P(X_1 + u_1, \dots, X_n + u_n) \\
-        &=
-        P(X_1 + v_1, X_2 + v_2 , \dots, X_n + v_n) - P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) \\
-        &+
-        P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) - P(X_1 + u_1, X_2 + u_2, \dots, X_n + v_n) \\
-        &+ \dots
-    \end{align*}
-
-    Let $\nu \colon \vec{X} \topartial \Nat$
-    be a partial valuation, let us show that
-    $\translate{K}(\restr{Q}{\nu}) \in \Nat[\vec{X}]$.
-
-    \begin{align*}
-        \translate{K}(\restr{Q}{\nu})
-        &= 
-        \translate{K}(\restr{P(X_1 + v_1, \dots, X_n + v_n)}{\nu})
-        -
-        \translate{K}(\restr{P(X_1 + u_1, \dots, X_n + u_n)}{\nu})
-    \end{align*}
-\end{proof}
-
-\begin{lemma}
-    Let $k \in \Nat$,
-    $f \in \NPoly[k]$,
-    and $\seqof{u_i}{i \in \Nat}$ be a sequence of
-    elements in $\Nat^n$. There exists $i < j$
-    such that
-    $f(u_j \cdot) - f(u_i \cdot) \in \NPoly[k-1]$.
-\end{lemma}
-
-\begin{defintion}
-    Residual $\Nat$-transducer.
-\end{definition}
-
-\begin{lemma}
-    Computable 
-\end{lemma}
-
-\begin{theorem}
-    If and only if this one is counter-free.
-\end{theorem}
-
-Note that it is not needed for decidability, as we are able to use directly the
-results on $\ZSF$, but this provides an independent proof, and highlights the
-versatility of this notion of "residual transducer".

### 2024-01-22 parts/transducers.tex
update proofs and paper

%! TeX program = xelatex
%! TeX root = ../pnrs.tex
%! lang = en-US
\section{Beyond commutativity}
\label{beyond-commutative:sec}

The goal of this section is to provide tools that do not require
\kl{commutative} \kl{$\Nat$-polyregular functions}. To that end, we introduce
the notion of \kl{$k$-residual transducer}, that generalizes the notion of
residual transducer introduced for \kl{$\Rel$-polyregular functions} in
\cite{LOPEZ23b}. We will first introduce the object, provide an algorithm (with
an oracle function) that computes said object, and finally, check that in
simple cases, the aperiodicity of a function can be read on its
\kl{$k$-residual transducer}. The main drawbacks of the generalisation are the
following: one needs to decide whether a \kl{$\Rel$-polyregular function} is in
$\NPoly$, which is only known in the \kl{commutative} case, and the
characterization in terms of counter free also assumes \kl{commutativity}.
However, we believe that the introduction of this object is relevant.

\AP
Let us first introduce the notion of \intro{residual} of a function
$f \colon \Sigma^* \to \Rel$, by defining
$\app{f}{u} \defined w \mapsto f(uw)$.
The collection of \reintro{residuals} of a function $f$
is denoted $\Res(f)$ and is defined as
the set of $\app{f}{u}$ where $u$ ranges over words in $\Sigma^*$.

\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
cannot be symmetric, and we are going to introduce a partial ordering $f
\natleq[k] g$ if and only if $g - f \in \NPoly[k]$.

The goal of the upcoming section is to provide a notion of residual transducer
for \kl{commutative} \kl{$\Nat$-polyregular functions}. The notion of
transducer is based on the \kl{marble transducers} \cite{EHB99}.

\begin{definition}
    Let $\mathcal{H}$ be a family of functions
    from $\Sigma^*$ to $\Nat$.
    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
    a tuple $(Q, q_0, \delta, \lambda, F)$ where
    \begin{itemize}
        \item $Q$ is a finite set of states,
        \item $q_0 \in Q$ is an initial state,
        \item $\delta \colon Q \times \Sigma \to Q$
            is a deterministic transition function,
        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
            is a correction function,
        \item $F \colon Q \to \Nat$ is an output function.
    \end{itemize}
\end{definition}

Given the syntax of an \kl{$\mathcal{H}$ transducer}, let us now 
introduce its semantics, which is given by a simple recursion
scheme.
\begin{definition}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Nat$
    is defined inductively  as follows:
    \begin{itemize}
        \item $\aTransd(q, \varepsilon) \defined F(q)$;
        \item $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    \end{itemize}
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

\AP As in the case of regular automata, we generalize the function $\delta
\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
$\lambda(q,\varepsilon) \defined 0$, and $\lambda(q, au) \defined \app{\lambda(q,a)}{u}
+ \lambda(\delta(q,a), u)$. Following these equations, we obtain that
$\aTransd(q_0, uv) = \aTransd(\delta(q_0, u), v) + \lambda(q_0, u)(v)$.
Furthermore, let us notice that if $\mathcal{H}$ is closed under finite sums
and shifts, then $\lambda(q, u) \in \mathcal{H}$ for all $(q,u) \in Q \times
\Sigma^*$.

\AP Let us recall that a counter in an automaton is a pair $u,w$ of words
together with an integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0,
u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.

\begin{theorem}[Cite gaëtan]
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    The following are equivalent:
    \begin{itemize}
        \item $f \in \NPoly[k]$,
        \item $f$ is computed by a \kl{$\NPoly[k-1]$-transducer}.
    \end{itemize}
    \begin{itemize}
        \item $f \in \NSF[k]$,
        \item $f$ is computed by a \kl{$\NSF[k-1]$-transducer}
            that is counter-free.
    \end{itemize}
\end{theorem}

Let us now introduce the notion of residual transducer, based on the
notions of ordering introduced prior to this paragraph.

\begin{definition}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{k-residual transducer}
    of $f$ 
    when
    \begin{itemize}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is \kl{downwards closed}
            for $\prefleq$;
        \item For all $u, v \in Q$, and for all $a \in \Sigma$,
            if
            $\delta(v,a) = u$
            then
            $u \prefleq va$,
            $\lambda(v,a) = \app{f}{va} - \app{f}{u}$,
            and $\lambda(v,a) \in \NPoly[k-1]$.
    \end{itemize}
\end{definition}

\begin{fact}
    The residual transducer is unique.
\end{fact}

We introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
given a function $f$. Notice that this algorithm requires the ability to test
if a function belongs to $\NPoly[k]$, which is only known to be feasible for
\kl{commutative} polyregular functions. However, the termination of this
algorithm also proves the existence of a \kl{$k$-residual transducer}.

\begin{algorithm}
    $Q \defined \{ \varepsilon \}$;
    $O \defined \setof{ a }{ a \in \Sigma}$;
    $\delta \defined \emptyset$;
    $\lambda \defined \emptyset$;
    $F \defined \emptyset$;

    \While{$O \neq \emptyset$}{
        choose $ua \in O$;

        $O \defined O \setminus \set{ ua }$;

        \eIf{$\exists v \in Q, v = \max_{\prefleq} \setof{w \in Q}{w \resleq{f}{k} ua}$}{
            $\delta(u, a) \defined v$;

            $\lambda(u, a) \defined \app{f}{ua} - \app{f}{v}$;
        }{
            $Q \defined Q \uplus \set{ ua }$;

            $\delta(u,a) \defined ua$;

            $\lambda(u,a) \defined 0$;

            $O \defined O \cup \setof{ uab }{b \in \Sigma}$;
        }
    }
    \For{$u \in Q$}{
        $F(u) \defined f(u)$;
    }
    \Return{$(Q, \varepsilon, \delta, \lambda, F)$};
    \caption{Computing a $(N,k)$-residual transducer.}
    \label{residual:algo}
\end{algorithm}

\begin{lemma}
    \label{correct-residual:lemma}
    If \cref{residual:algo} terminates on 
    an input $f \in \NPoly$, $k \in \Nat$,
    then it computes a \kl{$k$-residual transducer} of $f$.
\end{lemma}
\begin{proof}
    TODO.
\end{proof}

\begin{fact}
    \label{q-o-prefix-cool:fact}
    Let $f \in \NPoly$ and $k \in \Nat$.
    At each step of the \texttt{while loop}
    of \cref{residual:algo}, the sets
    $Q$ and $O$ are such that
    \begin{enumerate}
        \item $Q \cup O$ is a \kl{downwards closed} subset of 
            $\Sigma^*$ for $\prefleq$;
        \item elements in $O$ are pairwise incomparable
            for $\prefleq$, and are maximal
            for $\prefleq$ inside $Q \cup O$.
    \end{enumerate}
\end{fact}
\begin{proof}
    Let us write $Q_i$ and $O_i$ for the value of the variables
    $Q$ and $O$ at step $i$ of the \texttt{while loop}.
    We prove the desired property by induction on $i$.

    For $i=0$, the property is true because
    $Q_0 = \set{\varepsilon}$ and $O_0 = \setof{a}{a \in \Sigma}$.

    For $i+1$. Either the \texttt{if} branch was taken, in which case $Q_{i+1}
    \cup O_{i+1} = (Q_i \cup O_i) \setminus \set{u}$ for some $u \in O_i$. This
    set remains \kl{downwards closed}, and elements in $O_{i+1}$ remain maximal
    elements.

    If the \texttt{else} branch was taken, then there exists $u \in O_i$ such
    that $Q_{i+1} = Q_i \cup \set{u}$ and $O_{i+1} = O_i \setminus \set{ u }
    \cup \setof{ ua }{ a \in \Sigma}$. We conclude that $Q_{i+1} \cup O_{i+1} =
    Q_i \cup O_i \cup \setof{ ua }{a \in \Sigma}$ continues to be \kl{downwards
    closed} for $\prefleq$. Let $v \in Q_{i+1} \cup O_{i+1}$ be such that $ua
    \prefleq v$ for some $a \in \Sigma$. Then $u \prefleq v$, and $u = v$ since
    $u$ was a maximal element. As a consequence, $ua$ is a maximal element for
    all $a \in \Sigma$. Assume by contradiction that $ua$ is comparable with
    some $v \in O_{i+1}$ with $ua \neq v$, it cannot be that $ua \prefleq v$ by
    the above argument, and if $v \prefleq ua$ with $v \neq ua$, then $v
    \prefleq u$ and $u = v$, which is absurd since $v \not \in O{i+1}$.
    We have concluded that $O_{i+1}$ continues to have pairwise incomparable
    elements.
\end{proof}

\begin{lemma}
    \label{wqo-implies-termination:lemma}
    Let $f \in \NPoly$, and $k \in \Nat$ be such that
    $(\Sigma^*, \resleq{f}{k} \cup \preforth)$ is a \kl{well-quasi-ordering}.
    Then, \cref{residual:algo} terminates on the input $(f,k)$.
\end{lemma}
\begin{proof}
    Assume towards a contradiction that
    \cref{residual:algo} does not terminate.
    Then, the \texttt{else} branch in the \texttt{while loop}
    must be taken infinitely often.
    This means that the set $Q$ of states grows arbitrarily large.

    Let us write $\seqof{Q_i}{i \in \Nat}$ for the set of states $Q$ at step
    $i$ of the execution of \cref{residual:algo}. Applying
    \cref{q-o-prefix-cool:fact}, we know that for all $i \in \Nat$, $Q_i$ is
    \kl{downwards closed} for $\prefleq$. Let us write $Q_\infty \defined
    \bigcup_{i \in \Nat} Q_i$. The set $Q_\infty$ is infinite, and is
    \kl{downwards closed} for $\prefleq$. As a consequence, it is an infinite
    tree with a finite branching (at most $\card{\Sigma}$), and has an infinite
    branch $\seqof{u_j}{j \in \Nat}$ thanks to König’s lemma.

    Let us prove that this infinite branch is a \kl{bad sequence} for the
    ordering $\preforth \cup \resleq{f}{k}$. Because all elements in the
    sequence are pairwise comparable for $\prefleq$, it suffices to prove that
    the sequence is a \kl{bad sequence} for $\resleq{f}{k}$.

    Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
    know that $u_j \in Q_j$ and $u_p \in Q_p$. Then, at step $p-1$ of the
    algorithm, $u_j \in Q_{p-1}$, since $u_j \in Q_j \subseteq Q_{p-1}$.
    Because $u_j \prefleq u_p$ and $u_j \resleq{f}{k} u_p$,
    \cref{residual:algo} must take the \texttt{if} branch at step $p-1$. As a
    consequence, $u_p \not\in Q_{p}$, which is absurd.

    We have proven that the infinite branch is a \kl{bad sequence}
    for $\resleq{f}{k}$, but this contradicts the assumption
    that $\seqof{u_j}{j \in \Nat}$ is a \kl{good sequence}.

    Hence, \cref{residual:algo} must terminate.
\end{proof}

\begin{lemma}
    \label{n-poly-k-implies-wqo:lemma}
    Let $k \in \Nat$, 
    let $\aTransd$ be a \kl{$\NPoly[k-1]$-transducer},
    and let $f$ be the function $f$ \kl{computed} by $\aTransd$.
    Then,
    $(\Sigma^*, \resleq{f}{k-1} \cup \preforth)$ is
    \kl{well-quasi-ordering}.
\end{lemma}
\begin{proof}
    Assume towards a contradiction that there exists
    a \kl{bad sequence} $\seqof{u_i}{i \in \Nat}$
    of words in $\Sigma^*$.
    In particular, for all $i < j$, 
    $u_i$ is comparable to $u_j$ for the \kl{prefix ordering}.
    Without loss of generality, because $\prefleq$ is well-founded,
    we can assume that for all $i < j$,
    $u_i \prefleq u_j$.

    Let $\aTransd = (Q, q_0, \delta, \lambda, F)$.
    Because $Q$ is a finite set,
    there exists $i < j$
    such that 
    $\delta(q_0, u_i) = \delta(q_0, u_j)$. Since 
    $u_i \prefleq u_j$, there exists $v \in \Sigma^*$
    such that $u_i v = u_j$.
    Let us now remark that for all $w \in \Sigma^*$,
    \begin{align*}
        f(u_j w) &= f(u_i v w) \\
                 &= \aTransd(q_0, u_i v w) \\
                 &= \aTransd(\delta(q_0, u_i v), w)
                   + \lambda(q_0, u_i v) (w) \\
                 &= \aTransd(\delta(q_0, u_i), w)
                   + \lambda(q_0, u_i v) (w) \\
                 &= f(u_i w) 
                   + \lambda(q_0, u_i v) (w)
    \end{align*}

    Because $\NPoly[k-1]$ is closed under shifts and finite sums,
    we conclude that
    $\app{f}{u_j} - \app{f}{u_i}$ belongs to $\NPoly[k-1]$,
    proving that $u_i \resleq{f}{k-1} u_j$.
\end{proof}

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item $f \in \NPoly[k]$;
        \item every $\prefleq$ monotone sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k-1}$.
        \item $f$ is \kl{computed} by a
            \kl{$k$-residual transducer};
    \end{enumerate}
\end{theorem}
\begin{proof}
    The first two items are known to be equivalent since 
    the phd of gaëtan.
    We have proven that transducers imply
    wqo, and that wqo implies $k$-residual transducer.
\end{proof}

\begin{corollary}
    In the \kl{commutative} case, one can effectively compute
    a \kl{$k$-residual transducer} of a function $f \in \NPoly$.
\end{corollary}

\begin{example}
    Let us consider $f(w) \defined (\card{w} - 3)^2$, 
    which is a \kl{star-free $\Nat$-polyregular function}.
    Then,
    $\app{f}{a} - f$ is not in $\NPoly$,
    but
    $\app{f}{a^10} - f$ is in $\NPoly$,
    hence there is a non-trivial counter.
    However,
    for all $n > 1$ such that
    $\app{f}{a^n} - f$ belongs to $\NPoly$,
    $\app{f}{a^n} - \app{f}{a^{n-1}}$ belongs to $\NPoly$ too.
\end{example}
\begin{proof}
    Remark that $\app{f}{a} = (\card{w} - 2)^2$,
    hence that $\app{f}{a}- f = 2\card{w} - 5$.
    Similarly, $\app{f}{a^10} - f = 20\card{w} + k$ where $k \geq 0$.

    Finally,
    let $n > 1$ be such that
    $\app{f}{a^n} - f$ belong to $\NPoly$.
    In this specific example,
    $\app{f}{a^n} = (\card{w} + n -3)^2$,
    and $\app{f}{a^n} - f = (2 \card{w} + n - 6)n$.
    The latter belongs to $\NPoly$ if and only if
    $n \geq 6$. Let $n \geq 6$, 
    remark that
    $\app{f}{a^n} - \app{f}{a^{n-1}}
    = 2 \card{w} + 2n - 7$. Because $n = n_0 + 6$ with $n_0 \geq 0$,
    we can rewrite the latter 
    as $2 \card{w} + 2n_0 + 5$, which belongs to $\NPoly$.
\end{proof}

\begin{lemma}
    Let $f \colon \Sigma^* \to \Nat$ be a \kl{star-free $\Nat$-polyregular
    function}. Then, for all $u \in \Sigma^*$, for all $n > 1$, for all $w \in
    \Sigma^+$, if $\app{f}{u w^n} - \app{f}{u}$ belongs to $\NPoly[k-1]$, then
    $\app{f}{uw^n} - \app{f}{u w^{n-1}}$ belongs to $\NPoly[k-1]$.
\end{lemma}
\begin{proof}
    By induction on $n$, 
\end{proof}

Even though computability is not guaranteed, we have the following proof
scheme to decide aperiodicity of \kl{$\Nat$-polyregular functions}.
\begin{lemma}
    Let $f \in \NPoly[k]$ be \kl{ultimately polynomial},
    then every \kl{$k$-residual transducer} computing 
    $f$ has no counters, and has labels
    that are themselves \kl{ultimately polynomial}
    \kl{$\Nat$-polyregular functions}.
\end{lemma}
\begin{proof}

    We prove the result by induction on $k$. By definition
    of a residual transducer, the labels are \kl{ultimately polynomial}.
    Now, assume that there exists a counter
    in some \kl{$k$-residual transducer}
    $\aTransd \defined (Q, q_0, \delta, \lambda, F)$ of $f$. That is, there exists $\alpha \in \Sigma^*$,
    $w \in \Sigma^+$, and $n > 1$, such that
    $\delta(q_0,\alpha w^n) = \delta(q_0, \alpha)$.
    We want to prove that
    $\delta(q_0, \alpha w) = \delta(q_0, \alpha)$.
    Because $f$ is \kl{ultimately polynomial},
    it is in particular in $\ZSF[k]$, and therefore
    $g \defined \app{f}{\alpha w} - \app{f}{\alpha}$
    belongs to $\ZSF[k-1]$.
    It remains to be proven that this difference is a function in
    $\NPoly$.

    To that end, notice that the map $f(\alpha w^X \mathcal{M})$ is a
    \kl{commutative} \kl{star-free $\Nat$-polyregular function}. For a large
    enough $N_0$, we conclude that $f(\alpha w^{N_0+1} \mathcal{M}) - f(\alpha
    w^{N_0} \mathcal{M})$ belongs to $\NPoly$, hence that $\alpha w^{N_0}
    \resleq{f}{k-1} \alpha w^{N_0 + 1}$. Does that prove that $\alpha
    \resleq{f}{k-1} \alpha w$? Not at all and we cannot conclude.
\end{proof}

\begin{theorem}
    The following are equivalent for $f \in \NPoly[k]$.
    \begin{itemize}
        \item The function $f \in \NSF[k]$,
        \item $f$ is \kl{ultimately polynomial}.
        \item Every \kl{$k$-residual transducer} of $f$ has no counters,
        \item $f$ is \kl{computed} by some
            \kl{$\NSF[k-1]$-transducer}.
    \end{itemize}
\end{theorem}
\begin{proof}
    The only implication to prove is the first implies the last.
    We know that the automata is counter free, so the languages of the states
    are star-free. Now, we just write 
    \begin{equation*}
        TODO
    \end{equation*}
    and we have finished the proof.
\end{proof}

Remark that the theorem is effective because being \kl{ultimately polynomial}
is an effective property! However, effective conversions are not possible as of
now, because the $k$-residual transducer algorithm uses an oracle to decide if
a function is in $\NPoly$.

\begin{definition}
    For every polynomial $P \in \Rel[X \vec{Y}]$, we define 
    the partial derivative with respect to $X$ as follows:
    \begin{equation*}
        \partial_X P \defined P(X + 1, \vec{Y}) - P(X, \vec{Y}) \quad .
    \end{equation*}
\end{definition}

\begin{fact}
    Let $P \in \CorrectPoly[X_1, \dots, X_n]$, then
    for all $1 \leq i \leq n$,
    $\partial_{X_i} P \in \CorrectPoly$.
\end{fact}
\begin{proof}
    We prove the result by induction on the number of variables.
    Let $\nu \colon \vec{X} \topartial \Nat$.
    Two cases, either $X_i \in \dom(\nu)$,
    in which case, we remark that
    we can split $\nu$ into two valuations,
    and that coefficients are good.

    Otherwise, evaluation commutes with partial,
    and we conclude by induction hypothesis.
\end{proof}

\begin{remark}
    The converse does not hold,
    for instance $P(X) \defined X - 1$ has all its derivatives
    that are in $\CorrectPoly$, but is not itself in $\CorrectPoly$.
\end{remark}

\begin{fact}
    Let $P \in \Nat[X,\vec{Y}]$, and $K,L \in \Nat$,
    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \Nat[X, \vec{Y}]$
    whenever $L \geq K$.
\end{fact}
\begin{proof}
    We can write
    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
    Now,
    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
    is precisely
    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$,
    remaining with positive coefficients.
\end{proof}

\begin{fact}
    Let $P \in \CorrectPoly[X,\vec{Y}]$, and $K,L \in \Nat$,
    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \CorrectPoly$
    whenever $L \geq K$.
\end{fact}
\begin{proof}
    We can write
    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
    Now,
    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
    is precisely
    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$.
    Now, $P_i(\vec{Y})$
    remaining with positive coefficients.
\end{proof}
\begin{fact}
    For every $K \in \Nat$, and every $n \in \Nat$,
    $\Nat^n$ is a \kl{well-quasi-ordering}
    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
    if and only if $\vec{u} = \vec{v}$ or
    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
\end{fact}
\begin{proof}
    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
    one can extract an infinite monotone subsequence for
    the product ordering.
    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
    and the original sequence was good. Or for all $i < j$,
    we have $\vec{u}_i$ that is strictly smaller than 
    $\vec{u}_j$ for the product ordering. As a consequence,
    $\vec{u}_0 \leq \vec{u}_K$.
\end{proof}

\begin{fact}
    Let $P \in \CorrectPoly[X_1, \dots, X_n]$. 
    For all
    $\vec{u} \leq \vec{v} \in \Nat^n$,
    $Q(X_1, \dots, X_n)
    \defined 
    P(X_1 + v_1, \dots, X_n + v_n)
    -
    P(X_1 + u_1, \dots, X_n + u_n)$
    belongs to $\CorrectPoly[X_1, \dots, X_n]$.
\end{fact}
\begin{proof}
    Let us consider $K \in \Nat$ such that for all partial valuations
    $\nu \colon \vec{X} \topartial \Nat$,
    $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.

    Remark that
    \begin{align*}
        Q
        &= 
        P(X_1 + v_1, \dots, X_n + v_n)
        -
        P(X_1 + u_1, \dots, X_n + u_n) \\
        &=
        P(X_1 + v_1, X_2 + v_2 , \dots, X_n + v_n) - P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) \\
        &+
        P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) - P(X_1 + u_1, X_2 + u_2, \dots, X_n + v_n) \\
        &+ \dots
    \end{align*}

    Let $\nu \colon \vec{X} \topartial \Nat$
    be a partial valuation, let us show that
    $\translate{K}(\restr{Q}{\nu}) \in \Nat[\vec{X}]$.

    \begin{align*}
        \translate{K}(\restr{Q}{\nu})
        &= 
        \translate{K}(\restr{P(X_1 + v_1, \dots, X_n + v_n)}{\nu})
        -
        \translate{K}(\restr{P(X_1 + u_1, \dots, X_n + u_n)}{\nu})
    \end{align*}
\end{proof}

\begin{definition}
    Residual $\Nat$-transducer.
\end{definition}

\begin{lemma}
    If $P \in \CorrectPoly$, then it is computable
    by a $\Nat$-residual transducer.
\end{lemma}

\begin{theorem}
    The following are equivalent for \kl{commutative}.
    \begin{itemize}
        \item $f$ is in $\NSF$,
        \item The residual automata of $f$ has no counters
    \end{itemize}
\end{theorem}

### 2024-01-23 parts/transducers.tex
avancées

@@
 simple cases, the aperiodicity of a function can be read on its
-\kl{$k$-residual transducer}. The main drawbacks of the generalisation are the
+\kl{$k$-residual transducer}. The main drawbacks of the generalization are the
 following: one needs to decide whether a \kl{$\Rel$-polyregular function} is in
@@
 However, we believe that the introduction of this object is relevant.
-
-\AP
-Let us first introduce the notion of \intro{residual} of a function
-$f \colon \Sigma^* \to \Rel$, by defining
-$\app{f}{u} \defined w \mapsto f(uw)$.
-The collection of \reintro{residuals} of a function $f$
-is denoted $\Res(f)$ and is defined as
-the set of $\app{f}{u}$ where $u$ ranges over words in $\Sigma^*$.
-
-\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
-built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
-- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
-cannot be symmetric, and we are going to introduce a partial ordering $f
-\natleq[k] g$ if and only if $g - f \in \NPoly[k]$.
 
@@
 
-\AP Let us recall that a counter in an automaton is a pair $u,w$ of words
-together with an integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0,
-u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.
+\begin{example}
+    \label{non-canonical-transd:ex}
+    Let $f \colon \set{a}^* \to \Nat$
+    be defined as follows
+    \begin{equation*}
+        f(a^n) \defined
+        \begin{cases}
+            1 & \text{ if } n = 0 \\
+            0 & \text{ if } n = 1 \\
+            n & \text{ otherwise }
+        \end{cases}
+        \quad .
+    \end{equation*}
+    The function is computed by the two automata
+    in \cref{non-canonical-transd:fig}.
+\end{example}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{0.9\linewidth}
+        \begin{tikzpicture}[
+            etat/.style={minimum size=2em}
+            ]
+            \node[etat,state,initial,
+                accepting by arrow,
+                accepting text={$1$},
+                accepting where=below,
+                ] (A) at (0,0) {$\varepsilon$};
+            \node[etat,state,
+                accepting by arrow,
+                accepting text={$0$},
+                accepting where=below,
+                ] (B) at (2,0) {$a$};
+
+            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
+
+            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
+        \end{tikzpicture}
+        \caption{This automaton has a \kl{counter}.}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.9\linewidth}
+        \begin{tikzpicture}[
+            etat/.style={minimum size=2em}
+            ]
+            \node[etat,state,initial,
+                accepting by arrow,
+                accepting text={$1$},
+                accepting where=below,
+                ] (A) at (0,0) {$\varepsilon$};
+            \node[etat,state,
+                accepting by arrow,
+                accepting text={$0$},
+                accepting where=below,
+                ] (B) at (2,0) {$a$};
+
+            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
+
+            \draw[->] (B) edge[loop right] node[midway, right]
+                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
+        \end{tikzpicture}
+        \caption{This automaton is \kl{counter-free}.}
+    \end{subfigure}
+    \caption{Two \kl{$\NPoly[0]$-transducers}
+    computing \cref{non-canonical-transd:ex}.}
+    \label{non-canonical-transd:fig}
+\end{figure}
+
+\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
+words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
+\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.
+An automaton is \reintro{counter-free} if it contains no \reintro{counters}.
+This notion can be applied to \kl{$\mathcal{H}$-transducers} by
+considering counters of their underlying automata.
 
@@
 
-Let us now introduce the notion of residual transducer, based on the
-notions of ordering introduced prior to this paragraph.
+\AP Let us now introduce the notion of \intro{residual} of a function $f
+\colon \Sigma^* \to \Rel$, by defining $\app{f}{u} \defined w \mapsto f(uw)$.
+The collection of \reintro{residuals} of a function $f$ is denoted $\Res(f)$
+and is defined as the set of $\app{f}{u}$ where $u$ ranges over words in
+$\Sigma^*$.
+
+\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
+built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
+- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
+cannot be symmetric, and furthermore, we are going to order words rather than
+residuals. Namely, given $k \in \Nat$, and $f \colon \Sigma^* \to \Nat$, we
+define $u \resleq{f}{k} v$ if and only if $\app{f}{v} - \app{f}{u} \in
+\NPoly[k-1]$.
+
+\begin{fact}
+    \label{good-residual-ordering:fact}
+    Let $k \in \Nat$, and let $f \colon \Sigma^* \to \Nat$. Then,
+    $(\resleq{f}{k})$ is a quasi-ordering, satisfying the following
+    extra properties:
+    \begin{enumerate}
+        \item For all $u,v,w \in \Sigma^*$, $u \resleq{f}{k} v$
+            implies $uw \resleq{f}{k} vw$,
+        \item If $u \resleq{f}{k} v$ and $v \resleq{f}{k} u$,
+            then $\app{f}{u} = \app{f}{v}$.
+    \end{enumerate}
+\end{fact}
+
+There is one major issue arising when defining the notion of residual
+transducer based on the quasi-ordering $\resleq{f}{k}$. The naïve approach is
+to consider as states some representatives for the $\resleq{f}{k}$ minimal
+elements of $\Sigma^*$, all computing the same function
+(\cref{good-residual-ordering:fact}), and define transitions by letting
+$\delta(u, a)$ be \emph{one} state $v$ such that $v \resleq{f}{k} ua$. This
+naïve approach does not yield a canonical model.
+For instance, the function
+of \cref{non-canonical-transd:ex} can be computed in two ways if we only
+consider $\resleq{f}{k}$ minimal elements, as witnessed in
+\cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
+$\NSF[0]$, but one of the two presented transducers has a \kl{counter}.
+
+This is why in \cref{residual-transducer:def}, we introduce the prefix ordering
+$\prefleq$ to further constrain how the model computes.
 
 \begin{definition}
+    \label{residual-transducer:def}
     Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
@@
     when
-    \begin{itemize}
+    it is a \kl{$\NPoly[k-1]$-transducer}
+    satisfying the following properties:
+    \begin{enumerate}
         \item $\aTransd$ \kl{computes} $f$;
-        \item $Q \subseteq \Sigma^*$ is \kl{downwards closed}
+        \item $Q \subseteq \Sigma^*$ is a \kl{downwards closed}
             for $\prefleq$;
-        \item For all $u, v \in Q$, and for all $a \in \Sigma$,
-            if
-            $\delta(v,a) = u$
-            then
-            $u \prefleq va$,
-            $\lambda(v,a) = \app{f}{va} - \app{f}{u}$,
-            and $\lambda(v,a) \in \NPoly[k-1]$.
-    \end{itemize}
+        \item $q_0 = \varepsilon$;
+        \item every state $q \in Q$ is accessible from $q_0$;
+        \item For all $u, a \in Q$,
+            $\delta(u,a)$ is the $\prefleq$-maximal $v \in Q$
+            such that $v \prefleq ua$, and $v \resleq{f}{k} ua$.
+        \item For all $u,a \in Q$,
+            $\lambda(u,a) = \app{f}{ua} - \app{f}{\delta(u,a)}$.
+    \end{enumerate}
 \end{definition}
 
+Because of the restriction on prefixes, the \kl{$k$-residual transducer}
+is actually unique when it exists.
+
 \begin{fact}
-    The residual transducer is unique.
+    \label{unique-res-transducer:fact}
+    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
+    Then $f$ has at most one \kl{$k$-residual transducer}.
 \end{fact}
+\begin{proof}
+    Let $\aTransd_1$ and $\aTransd_2$ be two
+    \kl{$k$-residual transducers} for $f$.
+    The two initial states must be $\varepsilon$.
+    Let us prove by induction on $u \in \Sigma^*$ that
+    $\delta_1(\varepsilon, u) = \delta_2(\varepsilon, u)$
+    and that $Q_1$ equals $Q_2$ over prefixes of $u$.
+    This will prove that 
+    $Q_1 = Q_2$, hence that $\aTransd_1 = \aTransd_2$.
 
-We introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
+    Let $u \in \Sigma^* \cap Q_1 \cap Q_2$ and $a \in \Sigma$, $v_1 \in Q_1$ be
+    defined as $v \defined \delta_1(u,a)$, and $v_2 \defined \delta_2(u,a)$.
+    Remark that by induction hypothesis, for all $v \prefle u$, $v \in Q_1 \cap
+    Q_2$. If $\delta_1(u,a) = Q_1$, it means that for all $v \in Q_2$ such that
+    $v \prefle ua$, we have $\neg( v \resleq{f}{k} ua )$. The only possible
+    transition in $\aTransd_2$ is therefore $\delta_2(u,a) = ua$, and $ua \in
+    Q_2$. Similarly, if $\delta_1(u,a) \prefleq u$, then $\delta_2(u,a) =
+    \delta_1(u,a)$ by definition of $\delta_2$ as a maximum.
+\end{proof}
+
+We now introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
 given a function $f$. Notice that this algorithm requires the ability to test
@@
 \kl{commutative} polyregular functions. However, the termination of this
-algorithm also proves the existence of a \kl{$k$-residual transducer}.
+algorithm also proves the existence of the \kl{$k$-residual transducer}.
 
@@
 
-        \eIf{$\exists v \in Q, v = \max_{\prefleq} \setof{w \in Q}{w \resleq{f}{k} ua}$}{
+        \eIf{$\exists v \in Q, v = \max_{\prefleq} \setof{w \in Q}{w \prefleq ua \wedge w \resleq{f}{k} ua}$}{
             $\delta(u, a) \defined v$;
@@
     \Return{$(Q, \varepsilon, \delta, \lambda, F)$};
-    \caption{Computing a $(N,k)$-residual transducer.}
+    \caption{Computing a $k$-residual transducer given a function $f$.}
     \label{residual:algo}
@@
 
-\begin{lemma}
-    \label{correct-residual:lemma}
-    If \cref{residual:algo} terminates on 
-    an input $f \in \NPoly$, $k \in \Nat$,
-    then it computes a \kl{$k$-residual transducer} of $f$.
-\end{lemma}
-\begin{proof}
-    TODO.
-\end{proof}
-
 \begin{fact}
     \label{q-o-prefix-cool:fact}
-    Let $f \in \NPoly$ and $k \in \Nat$.
+    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
     At each step of the \texttt{while loop}
@@
 \begin{lemma}
+    \label{correct-residual:lemma}
+    If \cref{residual:algo} terminates on 
+    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
+    then it computes the \kl{$k$-residual transducer} of $f$.
+\end{lemma}
+\begin{proof}
+    Because of \cref{q-o-prefix-cool:fact},
+    we already know that $q_0 = \varepsilon$,
+    $Q$ is a \kl{downwards closed} subset of $\Sigma^*$
+    for $\prefleq$, 
+    that every state of $Q$ is accessible from $q_0$.
+    Notice that at every step,
+    $\lambda(u,a)$ is defined as
+    $\app{f}{ua} - \app{f}{\delta(u,a)}$.
+    Finally, since $Q \cup O$ is a \kl{downwards closed} subset of $\Sigma^*$
+    at every step,
+    we have that at step $i$,
+    for all $ua \in O_i$,
+    $\setof{w \in Q}{w \prefleq ua} = \setof{w \in Q_i}{ w \prefleq ua}$,
+    which proves that the maximum considered in the algorithm
+    is indeed computing correctly.
+\end{proof}
+
+\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
+quasi-ordered set $(X, \leq)$ is \kl{good} whenever there exist $i < j$ such
+that $u_i \leq u_j$. The set $X$ is a \kl{well-quasi-ordering} when
+every infinite sequence is \kl{good}. A sequence is \kl{bad}
+when it is not \kl{good}.
+
+\begin{lemma}
     \label{wqo-implies-termination:lemma}
-    Let $f \in \NPoly$, and $k \in \Nat$ be such that
-    $(\Sigma^*, \resleq{f}{k} \cup \preforth)$ is a \kl{well-quasi-ordering}.
+    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
+    every infinite, $\prefleq$-increasing sequence is \kl{good}
+    in $(\Sigma^*, \resleq{f}{k})$.
     Then, \cref{residual:algo} terminates on the input $(f,k)$.
@@
     Let us prove that this infinite branch is a \kl{bad sequence} for the
-    ordering $\preforth \cup \resleq{f}{k}$. Because all elements in the
-    sequence are pairwise comparable for $\prefleq$, it suffices to prove that
-    the sequence is a \kl{bad sequence} for $\resleq{f}{k}$.
-
+    ordering $\resleq{f}{k}$.
     Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
@@
     We have proven that the infinite branch is a \kl{bad sequence}
-    for $\resleq{f}{k}$, but this contradicts the assumption
-    that $\seqof{u_j}{j \in \Nat}$ is a \kl{good sequence}.
-
+    for $\resleq{f}{k}$, which contradicts the assumption.
     Hence, \cref{residual:algo} must terminate.
@@
     \label{n-poly-k-implies-wqo:lemma}
-    Let $k \in \Nat$, 
-    let $\aTransd$ be a \kl{$\NPoly[k-1]$-transducer},
-    and let $f$ be the function $f$ \kl{computed} by $\aTransd$.
-    Then,
-    $(\Sigma^*, \resleq{f}{k-1} \cup \preforth)$ is
-    \kl{well-quasi-ordering}.
+    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
+    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
 \end{lemma}
 \begin{proof}
-    Assume towards a contradiction that there exists
-    a \kl{bad sequence} $\seqof{u_i}{i \in \Nat}$
-    of words in $\Sigma^*$.
-    In particular, for all $i < j$, 
-    $u_i$ is comparable to $u_j$ for the \kl{prefix ordering}.
-    Without loss of generality, because $\prefleq$ is well-founded,
-    we can assume that for all $i < j$,
-    $u_i \prefleq u_j$.
+    Because $f \in \NPoly[k]$, there exists
+    a tuple $\vec{x}$ of first order free variables,
+    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
+    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
+    such that
+    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.
 
-    Let $\aTransd = (Q, q_0, \delta, \lambda, F)$.
-    Because $Q$ is a finite set,
-    there exists $i < j$
-    such that 
-    $\delta(q_0, u_i) = \delta(q_0, u_j)$. Since 
-    $u_i \prefleq u_j$, there exists $v \in \Sigma^*$
-    such that $u_i v = u_j$.
-    Let us now remark that for all $w \in \Sigma^*$,
+    Let $q$ be the maximal quantifier rank of formulas $\seqof{\psi_i}{1 \leq i
+    \leq n}$. To a word $u \in \Sigma^*$, we associate the vector $\MSO^q(u)$
+    that maps an $\MSO$-type with $\ell \leq |\vec{x}|$ free variables to the
+    number of realizations of this type in $u$.
+
+    Let $u, v \in \Sigma^*$ such that $\MSO^q(u) \leq \MSO^q(v)$, which means
+    that every $\MSO$ type (of quantifier rank $q$ and with at most $n$ free
+    variables) has at least as many realizations in $v$ than it has in $u$.
+    Remark that by the compositionality of $\MSO$ over words (for instance, see
+    the Feferman-Vaught theorem \cite{FEVAU59,MAKOW04}), for all $\MSO^q$ types
+    $t(\vec{x})$, there are finitely many $\MSO^q$ types $t_l^j(\vec{y_i}),
+    t_r^j(\vec{z_i})$ with $\vec{x} = \vec{y_i} \uplus \vec{z_i}$
+    for $1 \leq j \leq N_0$, such that for every
+    tuple $\vec{a}$ of elements in a word $uv$, $\MSO^q(\vec{a} / uv) =
+    t(\vec{x})$ if and only if there exists $1 \leq j \leq N_0$,
+    such that $\vec{a} = \vec{b} \uplus \vec{c}$,
+    $\MSO^q(\vec{b} / u) =
+    t_l^j(\vec{y_i})$, and $\MSO^q(\vec{c} / v) = t_r^j(\vec{z_i})$.
+    We write $t = t_l \odot t_r$ to signify
+    that $\MSO^q(\vec{bc} / uv) = t$
+    if and only if $\MSO^q(\vec{b}/u) = t_l$
+    and $\MSO^q(\vec{c}/v) = t_r$.
+
+    As a consequence, if $\MSO^q(u) \leq \MSO^q(v)$, then 
+    for all $w \in \Sigma^*$: 
     \begin{align*}
-        f(u_j w) &= f(u_i v w) \\
-                 &= \aTransd(q_0, u_i v w) \\
-                 &= \aTransd(\delta(q_0, u_i v), w)
-                   + \lambda(q_0, u_i v) (w) \\
-                 &= \aTransd(\delta(q_0, u_i), w)
-                   + \lambda(q_0, u_i v) (w) \\
-                 &= f(u_i w) 
-                   + \lambda(q_0, u_i v) (w)
+        & f(vw) - f(uw) \\
+        &= 
+        \sum_{i = 1}^n m_i
+        \left[
+            \vcount{\phi_i(\vec{x})} (vw) -
+            \vcount{\phi_i(\vec{x})} (uw)
+        \right] \\
+        &= 
+        \sum_{i = 1}^n
+        m_i
+            \sum_{\phi_i \in t(\vec{x})}
+        \left[
+            \vcount{t(\vec{x})}(vw)
+            -
+            \vcount{t(\vec{x})}(uw)
+        \right] \\
+        &= 
+        \sum_{i = 1}^n
+        m_i
+        \sum_{1 \leq j \leq N_0}
+        \sum_{\phi_i \in t_l^j(\vec{y}) \odot t_r^j(\vec{z})}
+        \\
+        &\quad\quad
+        \underbrace{
+        \left[
+            \vcount{t_r^j(\vec{y})}(v)
+            -
+            \vcount{t_r^j(\vec{y})}(u)
+        \right] 
+    }_{ \in \Nat }
+            \times 
+            \vcount{t_l^j(\vec{z})}(w)
     \end{align*}
 
-    Because $\NPoly[k-1]$ is closed under shifts and finite sums,
-    we conclude that
-    $\app{f}{u_j} - \app{f}{u_i}$ belongs to $\NPoly[k-1]$,
-    proving that $u_i \resleq{f}{k-1} u_j$.
+    We have proven that if $\MSO^q(u) \leq \MSO^q(v)$, then $u \resleq{f}{k}
+    v$. Recall that $\Nat^p$ is a \kl{well-quasi-ordering} when endowed with
+    the product ordering, and therefore that $\setof{\MSO^q(u)}{u \in
+    \Sigma^*}$ is a \kl{well-quasi-ordering}.
+
+    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
+    Without loss of generality, one can assume that for all $i \neq j$, $u_i
+    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
+    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
+    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
+    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
+    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
+    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
 \end{proof}
@@
         \item $f \in \NPoly[k]$;
-        \item every $\prefleq$ monotone sequence
+        \item $(\Sigma^*, \resleq{f}{k})$ is a
+            \kl{well-quasi-ordering};
+        \item every $\prefleq$-increasing sequence
             of $\Sigma^*$  is a \kl{good sequence}
-            for $\resleq{f}{k-1}$.
+            for $\resleq{f}{k}$;
         \item $f$ is \kl{computed} by a
-            \kl{$k$-residual transducer};
+            \kl{$k$-residual transducer}.
     \end{enumerate}
+    If $f$ is asumed to be \kl{commutative}, then all of the above
+    properties are decidable, and the conversions effective.
 \end{theorem}
@@
 
-\begin{corollary}
-    In the \kl{commutative} case, one can effectively compute
-    a \kl{$k$-residual transducer} of a function $f \in \NPoly$.
-\end{corollary}
+\begin{lemma}
+    \label{aperiodic-iff-residual:lem}
+    Let $L$ be a regular language.
+    The following are equivalent:
+    \begin{enumerate}
+        \item $L$ is aperiodic;
+        \item The \kl{$0$-residual transducer} of $\ind{L}$
+            is \kl{counter-free}.
+    \end{enumerate}
+\end{lemma}
+\begin{proof}
+    If the $0$-residual transducer is counter free,
+    then $L$ is aperiodic because the minimal automata is also counter-free.
+
+    % TODO
+    \textbf{TODO Claim:}
+    on every branch of the automata, the words have distinct residuals.
+
+    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
+    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
+    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
+    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
+    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
+    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
+    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
+    w_2) = t$.
+
+    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
+    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
+    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
+    strict prefix of $s_0$. It must exist, because otherwise the two are always
+    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
+    this contradicts the minimality of $u$.
+
+    We have proven that $s_0$ and $s_1$ are comparable, hence they
+    are equal.
+    Finally, we have proven that $\delta(q, w_1) = s_0$,
+    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
+
+    As a consequence $t$ and $q$ were comparable for the prefix relation,
+    hence equal, and therefore $\delta(q, w) = q$.
+\end{proof}
+
+\begin{conjecture}
+    Let $f \in \NPoly$. The following are equivalent,
+    \begin{enumerate}
+        \item $f \in \ZSF$,
+        \item the \kl{$k$-residual transducer} of $f$
+            is \kl{counter-free},
+            and its labels belong to $\ZSF$.
+    \end{enumerate}
+\end{conjecture}
+\begin{proof}
+    If it has a counter free residual transducer recursively, then
+    it clearly belongs to $\ZSF$.
+
+    % TODO: mix orderings and trees and maximal elements :(
+
+    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
+    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
+    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
+    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
+    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
+    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
+    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
+    w_2) = t$.
+
+    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
+    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
+    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
+    strict prefix of $s_0$. It must exist, because otherwise the two are always
+    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
+    this contradicts the minimality of $u$.
+
+    We have proven that $s_0$ and $s_1$ are comparable, hence they
+    are equal.
+    Finally, we have proven that $\delta(q, w_1) = s_0$,
+    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
+
+    As a consequence $t$ and $q$ were comparable for the prefix relation,
+    hence equal, and therefore $\delta(q, w) = q$.
+\end{proof}
 
@@
     The only implication to prove is the first implies the last.
-    We know that the automata is counter free, so the languages of the states
+    We know that the automaton is counter free, so the languages of the states
     are star-free. Now, we just write 

### 2024-01-25 parts/appendix.tex
writing to publish

%! TeX program = xelatex
%! lang = en-US
\section{Proofs of
section \ref{preliminaries:sec}}

\begin{proof}[Proof of \cref{decidable-commutative-rat:lemma}]
    Because $f$ is a \kl{$\Rel$-rational series},
    there exists a $n \in \Nat$,
    vectors $v_1$ and $v_2$,
    and $n$ by $n$ integer matrices $\seqof{M_a}{a \in \Sigma}$,
    such that
    \begin{equation*}
        \forall w \in \Sigma^*,
        f(w) = {}^t v_1 \left(\prod_{i = 1}^{|w|} M_{w_i}\right) v_2 \quad .
    \end{equation*}
    To simplify notations, 
    let us write $M_w \defined \prod_{i = 1}^{|w|} M_{w_i}$,
    and 
    $S \defined \setof{M_w}{w \in \Sigma^+}$ be the
    semigroup generated by $\seqof{M_a}{a \in \Sigma}$.

    The function $f$ is \kl{commutative} if and only if for all $u,v \in
    \Sigma^+$, ${}^t v_1 (M_u M_v  - M_v M_u) v_2 = 0$. The above is a
    polynomial equation involving the coefficients of $M_u$ and $M_v$.
    Let us compute $\overline{S \times S}$ the Zariski closure of $S \times S$.
    The above polynomial vanishes over $S \times S$ if and only
    if it belongs to the Zariski closure, which is decidable
    \cite{HROUPOWO18}.
\end{proof}

\begin{proof}[Proof of
    \cref{pre-compose-growth-commut:lemma}]
    If $f \in \ZPoly[k]$, then 
    for every \kl{polyregular function} $g \in \Poly[\ell]$,
    $f \circ g \in \ZPoly[k+ \ell - 1]$
    \cite[]{LOPEZ23b}.

    Conversely, it was proven in \cite[Theorem III.3]{LOPEZ23b}
    that $f \in \ZPoly[k]$ if and only if
    for all $\alpha_0, \cdots, \alpha_k$,
    for all $w_1, \dots, w_k$,
    we have 
    \begin{equation}
        \label{k-pumpable:eqn}
        f\left(
            \alpha_0 \prod_{i = 1}^k w_i^{X_i} \alpha_i
        \right)
        = \bigO( (X_1 + \cdots X_k)^{k} )
        \quad .
    \end{equation}
    Or, the \kl{commutative} \kl{star-free polyregular function} 
    $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
    a word $u$
    to $\alpha \prod_{i = 1}^k w_i^{\card[i]{u}}$
    has \kl{growth rate} $1$.
    Hence, 
    $f \circ h \in \ZPoly[k]$, 
    and we indeed conclude that 
    \cref{k-pumpable:eqn}
    holds, i.e. that $f \in \ZPoly[k]$.
\end{proof}

\begin{proof}[Proof of
    \cref{pre-compose-sf-commut:lemma}]
    If $f \in \ZSF$, then for all $h \in \SF$,
    $f \circ h \in \ZSF$, and \emph{a fortiori}
    for \kl{commutative} functions $h$.

    Conversely, assume that $f \circ h \in \ZSF$
    for all \kl{commutative} functions $h \in \SF$.
    Using \cite[Theorem V.13]{LOPEZ23b},
    to conclude that $f \in \ZSF$,
    it suffices to prove that
    for all $\alpha_0, \cdots, \alpha_k \in \Sigma^*$,
    for all $w_1, \dots, w_k \in \Sigma^*$,
    there exists a polynomial $P \in \Rel[X_1, \dots,X_k]$
    and a constant $N_0 \in \Nat$,
    such that if $X_1, \dots, X_k \geq N_0$:
    \begin{equation}
        \label{ultimate-polynomial:eqn}
        f\left(
            \alpha_0 \prod_{i = 1}^k w_i^{X_i} \alpha_i
        \right)
        = P(X_1, \dots, X_k)
        \quad .
    \end{equation}
    Let us consider
    the \kl{commutative} \kl{star-free polyregular function}
    $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
    a word $u$
    to $\alpha \prod_{i = 1}^k w_i^{\card[i]{u}}$.
    We know that
    $f \circ h \in \ZSF$, hence, that 
    there exists $Q \in \Rel[X_1, \dots, X_n]$
    and $M_0 \in \Nat$,
    such that for all $X_1, \dots, X_n \geq N_0$,
    $f \circ h( \prod_{i = 1}^k \underline{i}^{X_i}) = Q(X_1, \dots, X_n)$.
    In particular,
    we can take $N_0 = M_0$, and $P = Q$ to conclude that
    \cref{ultimate-polynomial:eqn} holds, hence, that
    $f \in \ZSF$.
\end{proof}

\section{Proofs of
section \ref{polynomials:sec}}

\begin{proof}[Proof of
    \cref{n-poly-n-poly:example}]
    The polynomials of $\Nat[\vec{X}]$
    are obtained from basic functions (constant in $\Nat$,
    $X_i$ for some $1 \leq i \leq n$)
    by products and sums. Because the basic functions are
    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
    that are closed under these operations, we conclude.
\end{proof}

\begin{proof}[Proof of
    \cref{n-poly-combinatorics:lem}
    ]

    Let $\omega$ be an idempotent power for the finite monoid $M$,
    and
    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
    $f = \pi^\dagger$. That is, 
    $f(w)$ is the sum over all factorizations of $w$
    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.

    Let us define an equivalence relation $\equiv$ over words of
    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
    repetitions of the factors $u_i^{\omega}$ in the words.
    We lift the equivalence relation $\equiv$ to $k$-tuples of
    words by pointwise application.

    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then
    $\pi(\mu(v_1), \dots, \mu(v_k)) = \pi(\mu(w_1), \dots, \mu(w_k))$ because
    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid
    $M$. In particular, $f(w)$ can be computed as the sum over equivalence
    classes for $\equiv$, of the value of $\pi$ on one representative,
    which belongs to $\Nat$.

    Therefore, it suffices to prove that the number of partitions of $w$ in a
    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
    with \kl{positive coefficients} for its \kl{maximal monomials}.

    Notice that to produce all elements in a given equivalence class
    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
    back to the normalized version
    of the equivalence class. For all elements in an equivalence class,
    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
    and therefore the number of possible choices
    is either constant, or 
    is a product of binomials of the form $\binom{X - l}{s}$
    which has the desired shape.
\end{proof}

\begin{proof}[Proof of 
    \cref{derivation-monomials:fact}
    ]
    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
    where $k_1, \dots, k_n > 0$ and $n > 0$.
    Let us compute
    \begin{align*}
        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
    \end{align*}
    Therefore, 
    the \kl{maximal monomials} of $\Diff{K}{M}$
    are precisely
    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
    which \kl{strictly divide} $M$.
    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
    for all $1 \leq i \leq n$, $t_i \leq k_i$,
    and there exists $1 \leq p \leq n$ such that$ t_j < k_j$.
    In particular, $T$ \kl{divides}
    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
\end{proof}

\begin{proof}[Proof of \cref{discrete-deriv-linear:fact}]
    TODO or to skip.
\end{proof}

\begin{proof}[Proof of \cref{maxi-monomials-submodular:fact}]
\end{proof}

\begin{proof}[Proof of \cref{derivation-monomials:fact}]
\end{proof}

\begin{proof}[Proof of \cref{derivation-simplifies:lemma}]
    Let $P \in \Rel[\vec{X}]$ and
    $K \in \Nat$.
    Leveraging 
    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
    \begin{align*}
        \MaximalMonomials\left(\Diff{K}{P}\right)
        &=
        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
        &\hoareleq
        \bigcup_{M \in \Monomials(P)} 
        \MaximalMonomials(\Diff{K}{M}) \\
        &\hoarele
        \MaximalMonomials(P)
        \quad .
        \qedhere
    \end{align*}
\end{proof}
\begin{proof}[Proof of \cref{derivation-stabilises-correct:lem}]
    If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
    hence for $K = 0$, $Q \in \CorrectPoly$.
    Otherwise, let $\nu \colon \vec{X} \topartial \Nat$,
    and let us analyze the \kl{maximal monomials} $T$
    of $\restr{Q}{\nu}$.
    Because $T$ is a \kl{monomial} of $\restr{Q}{\nu}$,
    it \kl{divides} some \kl{monomial} in $Q$, and
    using \cref{derivation-monomials:fact},
    it \kl{divides} some \kl{monomial} $M$ in 
    $\Diff{K}{P_1}$.
    Now, the monomial $T$ is computed as follows:
    \begin{align*}
        T &= 
        \sum_{S \in \Monomials(Q), T \divides S}
        \restr{S}{\nu} \\
          &= 
          \sum_{S \in \Monomials(\Diff{K}{P_1}), T \divides S}
            \restr{S}{\nu}
        + 
          \sum_{S \in \Monomials(P_2), T \divides S}
            \restr{S}{\nu}
        \\
          &\geq
          \sum_{S \in \Monomials(\Diff{K}{P_1}), T \divides S}
            \restr{S}{\nu}
          - 
            \left|
          \sum_{S \in \Monomials(P_2), T \divides S}
            \restr{S}{\nu}
            \right|
    \end{align*}

\end{proof}

\begin{proof}[Proof of
    \cref{decidability-correct:lem}]
    Let $P \in \Rel[\vec{X}]$. We prove by induction on
    $\MaximalMonomials(P)$ that whether $P \in \CorrectPoly$
    is decidable.

    If $P$ is a constant polynomial, then
    $P \in \CorrectPoly$ if and only if $P = n \in \Nat$
    which is decidable.

    Otherwise, $P$ is not a constant polynomial, and let us write $P = P_1 +
    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. If $P_1
    \not \in \Nat[\vec{X}]$, then $P \not \in \CorrectPoly$, and this is
    decidable. Hence, we can assume that $P_1 \in \Nat[\vec{X}]$. Using
    \cref{derivation-stabilises-correct:lem}, there exists a computable bound
    $K$, such that if $P \in \CorrectPoly$, then t $Q \defined (\Diff{K}{P_1} +
    P_2) \in \CorrectPoly$. Leveraging, \cref{derivation-simplifies:lemma}, we
    also know that $\MaximalMonomials(Q) \hoarele \MaximalMonomials(P)$, and
    therefore that we can decide whether $Q \in \CorrectPoly$. Notice that if
    $Q \not \in \CorrectPoly$, then $P \not \in \CorrectPoly$, hence we can
    assume that $Q \in \CorrectPoly$.

    Similarly, for partial valuations
    $\nu \colon \vec{X} \topartial \set{0, \dots, K}$
    fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
    \hoarele \MaximalMonomials{P}$. 
    We can check for all these partial valuations that
    $\restr{P}{\nu} \in \CorrectPoly$ by induction hypothesis,
    and if one test is not satisfied, then $P \not \in \CorrectPoly$.

    Let us now conclude that if all the above checks answered correctly,
    then $P \in \CorrectPoly$. Let $\nu \colon \vec{X} \topartial \Nat$
    be a partial valuation. Then, 
    one of the three following cases arise:
    \begin{itemize}
        \item 
            Either,
        $\nu$ fixes \emph{some} indeterminate $X_i$ to a value $n_i < K$.
        In that case, let us define $\delta$ as the partial function that coincides
        with $\nu$ on its domain
        $\dom(\delta) = \dom(\nu) \setminus \set{ X_i }$.
        Then, $\restr{P}{\nu} = \restr{\restr{P}{X_i = n_i}}{\delta}$.
        As we previously checked that $\restr{P}{X_i = n_i} \in \CorrectPoly$,
        we conclude in particular that 
        $\restr{P}{\nu}$ has \kl{non-negative} \kl{maximal monomials}.

        \item Or, for all $X_i$ in the (potentially empty) domain of $\nu$, $\nu(X_i)
            \geq K$. In that case, let us write $\delta \defined
            \translate{-K}(\nu)$, that is the partial map $\delta \colon
            \vec{X} \topartial \Nat$ having the same domain as $\nu$, and such
            that for all $X_i \in \dom(\nu)$, $\delta(X_i) \defined \nu(X_i) -
            K$. Recall that $Q = \Diff{K}{P_1} + P_2$, and therefore that
            $\translate{K}(P) = P_1 + \translate{K}(Q)$. By linearity,
            $\restr{\translate{K}(P)}{\delta} = \restr{P_1}{\delta} +
            \restr{\translate{K}(Q)}{\delta}$, and we conclude that the
            \kl{maximal monomials} of $\restr{\translate{K}(P)}{\delta}$ are
            \kl{non-negative}.
            Since $\restr{\translate{K}(P)}{\delta} = \translate{K}{\restr{P}{\nu}}$,
            and since
            $\translate{K}$ leaves \kl{maximal monomials} invariant,
            we conclude that
            the \kl{maximal monomials} of $\restr{P}{\nu}$
            are the same as those of $\restr{\translate{K}(P)}{\delta}$,
            and are therefore \kl{non-negative}.
            \qedhere
    \end{itemize}
\end{proof}

\begin{proof}[Proof of \cref{derivation-translation:lem}]

    To prove \cref{d-t-transl:item} $\implies$ \cref{d-t-correct:item}, let us
    remark that if $P$ is \kl{non-negative} and $\translate{K}(P) \in
    \Nat[\vec{X}]$, then $\translate{K}(P) \in \CorrectPoly$ thanks to
    \cref{n-poly-n-poly:example}. Then, for every partial valuation $\nu \colon
    \vec{X} \topartial \Nat$, $\translate{K}(\restr{P}{\nu}) \in
    \Nat[\vec{X}]$, and in particular conclude that \kl{maximal monomials} of
    $\translate{K}(\restr{P}{\nu})$ are exactly the \kl{maximal monomials} of
    $\restr{P}{\nu}$ thanks to \cref{translation-maximal:fact}, and are therefore
    \kl{non-negative}.

    Conversely, the implication \cref{d-t-correct:item} $\implies$
    \cref{d-t-transl:item} is proven by induction on $\MaximalMonomials(P)$. If
    $P$ is a constant polynomial, then $P = n$ for some $n \in \Nat$, and we
    conclude that $\translate{0}(P) \in \Nat[\vec{X}]$. Otherwise, $P = P_1 +
    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. Using
    \cref{derivation-stabilises-correct:lem}, there exists a computable $K \in
    \Nat$ such that $Q \defined \Diff{K}{P_1} + \translate{K}(P_2)$ belongs to
    $\CorrectPoly$. Since, $\MaximalMonomials(Q) \hoarele
    \MaximalMonomials(P)$, we conclude by induction hypothesis that there
    exists $L \in \Nat$ such that for every partial valuation $\nu \colon
    \vec{X} \topartial \Nat$, $\translate{L}(\restr{Q}{\nu}) \in
    \Nat[\vec{X}]$. Similarly, for all partial functions $\nu \colon \vec{X}
    \topartial \set{0, \dots, K}$ fixing at least one indeterminate of $P$,
    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials(P)$, hence
    there exists $L_\nu \in \Nat$ such that $\translate{L_\nu}(\restr{P}{\nu})
    \in \Nat[\vec{X}]$. Let us define $L_m$ to be the maximum of $L$, and
    $L_\nu$ where $\nu$ ranges over partial functions described above.

    Let us prove that for all partial functions $\nu \colon \vec{X} \topartial
    \Nat$, the translation $\translate{L_m + K}(\restr{P}{\nu})$ belongs to
    $\Nat[\vec{X}]$. Without loss of generality, $\nu$ fixes only
    indeterminates that appear in $P$. If $\nu$ fixes at least one
    indeterminate $X_i$ to a value $\nu(X_i) = k \leq K$, then we can write
    $\nu = \mu [X_i = k]$, where $\mu$ does not fix the value of $X_i$. By
    construction, we know that $\translate{L_i}(\restr{\restr{P}{X_i =
    k}}{\mu})$ belongs to $\Nat[\vec{X}]$, and therefore that $\translate{K +
    L_m}(\restr{P}{\nu})$ does so too. The only case left is when all the
    variables appearing in $\nu$ are greater than $K$.
    In that case, $\nu = \translate{K}(\mu)$ for some partial 
    valuation $\mu$.
    In this case,
    $\translate{K}(\restr{P}{\translate{K}(\mu)}
    = \restr{\translate{K}(P)}{\mu}$
    thanks to 
    \cref{discrete-deriv-linear:fact}.
    Remark that by definition,
    $\translate{K}(P) = P_1 + Q$,
    hence that
    $\translate{K + L_m}(P) = \translate{L_m}(P_1) + \translate{L_m}(Q)$
    belongs to $\Nat[\vec{X}]$.
    Finally,
    we conclude that $\restr{\translate{K+L_m}(P)}{\mu}$
    belongs to $\Nat[\vec{X}]$
    because
    the latter is closed under partial applications.
\end{proof}

\section{Proofs of section \ref{beyond-polynomials:sec}}

\section{Proofs of section \ref{beyond-commutative:sec}}

\begin{proof}[Proof of
    \cref{q-o-prefix-cool:fact}]
    Let us write $Q_i$ and $O_i$ for the value of the variables
    $Q$ and $O$ at step $i$ of the \texttt{while loop}.
    We prove the desired property by induction on $i$.

    For $i=0$, the property is true because
    $Q_0 = \set{\varepsilon}$ and $O_0 = \setof{a}{a \in \Sigma}$.

    For $i+1$. Either the \texttt{if} branch was taken, in which case $Q_{i+1}
    \cup O_{i+1} = (Q_i \cup O_i) \setminus \set{u}$ for some $u \in O_i$. This
    set remains \kl{downwards closed}, and elements in $O_{i+1}$ remain maximal
    elements.

    If the \texttt{else} branch was taken, then there exists $u \in O_i$ such
    that $Q_{i+1} = Q_i \cup \set{u}$ and $O_{i+1} = O_i \setminus \set{ u }
    \cup \setof{ ua }{ a \in \Sigma}$. We conclude that $Q_{i+1} \cup O_{i+1} =
    Q_i \cup O_i \cup \setof{ ua }{a \in \Sigma}$ continues to be \kl{downwards
    closed} for $\prefleq$. Let $v \in Q_{i+1} \cup O_{i+1}$ be such that $ua
    \prefleq v$ for some $a \in \Sigma$. Then $u \prefleq v$, and $u = v$ since
    $u$ was a maximal element. As a consequence, $ua$ is a maximal element for
    all $a \in \Sigma$. Assume by contradiction that $ua$ is comparable with
    some $v \in O_{i+1}$ with $ua \neq v$, it cannot be that $ua \prefleq v$ by
    the above argument, and if $v \prefleq ua$ with $v \neq ua$, then $v
    \prefleq u$ and $u = v$, which is absurd since $v \not \in O{i+1}$.
    We have concluded that $O_{i+1}$ continues to have pairwise incomparable
    elements.
\end{proof}
\begin{proof}[Proof of \cref{wqo-implies-termination:lemma}]
    Assume towards a contradiction that
    \cref{residual:algo} does not terminate.
    Then, the \texttt{else} branch in the \texttt{while loop}
    must be taken infinitely often.
    This means that the set $Q$ of states grows arbitrarily large.

    Let us write $\seqof{Q_i}{i \in \Nat}$ for the set of states $Q$ at step
    $i$ of the execution of \cref{residual:algo}. Applying
    \cref{q-o-prefix-cool:fact}, we know that for all $i \in \Nat$, $Q_i$ is
    \kl{downwards closed} for $\prefleq$. Let us write $Q_\infty \defined
    \bigcup_{i \in \Nat} Q_i$. The set $Q_\infty$ is infinite, and is
    \kl{downwards closed} for $\prefleq$. As a consequence, it is an infinite
    tree with a finite branching (at most $\card{\Sigma}$), and has an infinite
    branch $\seqof{u_j}{j \in \Nat}$ thanks to König’s lemma.

    Let us prove that this infinite branch is a \kl{bad sequence} for the
    ordering $\resleq{f}{k}$.
    Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
    know that $u_j \in Q_j$ and $u_p \in Q_p$. Then, at step $p-1$ of the
    algorithm, $u_j \in Q_{p-1}$, since $u_j \in Q_j \subseteq Q_{p-1}$.
    Because $u_j \prefleq u_p$ and $u_j \resleq{f}{k} u_p$,
    \cref{residual:algo} must take the \texttt{if} branch at step $p-1$. As a
    consequence, $u_p \not\in Q_{p}$, which is absurd.

    We have proven that the infinite branch is a \kl{bad sequence}
    for $\resleq{f}{k}$, which contradicts the assumption.
    Hence, \cref{residual:algo} must terminate.
\end{proof}

\begin{proof}[Proof of \cref{n-poly-k-implies-wqo:lemma}]
    Because $f \in \NPoly[k]$, there exists
    a tuple $\vec{x}$ of first order free variables,
    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
    such that
    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.

    Let $q$ be the maximal quantifier rank of formulas $\seqof{\psi_i}{1 \leq i
    \leq n}$. To a word $u \in \Sigma^*$, we associate the vector $\MSO^q(u)$
    that maps an $\MSO$-type with $\ell \leq |\vec{x}|$ free variables to the
    number of realizations of this type in $u$.

    Let $u, v \in \Sigma^*$ such that $\MSO^q(u) \leq \MSO^q(v)$, which means
    that every $\MSO$ type (of quantifier rank $q$ and with at most $n$ free
    variables) has at least as many realizations in $v$ than it has in $u$.
    Remark that by the compositionality of $\MSO$ over words (for instance, see
    the Feferman-Vaught theorem \cite{FEVAU59,MAKOW04}), for all $\MSO^q$ types
    $t(\vec{x})$, there are finitely many $\MSO^q$ types $t_l^j(\vec{y_i}),
    t_r^j(\vec{z_i})$ with $\vec{x} = \vec{y_i} \uplus \vec{z_i}$
    for $1 \leq j \leq N_0$, such that for every
    tuple $\vec{a}$ of elements in a word $uv$, $\MSO^q(\vec{a} / uv) =
    t(\vec{x})$ if and only if there exists $1 \leq j \leq N_0$,
    such that $\vec{a} = \vec{b} \uplus \vec{c}$,
    $\MSO^q(\vec{b} / u) =
    t_l^j(\vec{y_i})$, and $\MSO^q(\vec{c} / v) = t_r^j(\vec{z_i})$.
    We write $t = t_l \odot t_r$ to signify
    that $\MSO^q(\vec{bc} / uv) = t$
    if and only if $\MSO^q(\vec{b}/u) = t_l$
    and $\MSO^q(\vec{c}/v) = t_r$.

    As a consequence, if $\MSO^q(u) \leq \MSO^q(v)$, then 
    for all $w \in \Sigma^*$: 
    \begin{align*}
        & f(vw) - f(uw) \\
        &= 
        \sum_{i = 1}^n m_i
        \left[
            \vcount{\phi_i(\vec{x})} (vw) -
            \vcount{\phi_i(\vec{x})} (uw)
        \right] \\
        &= 
        \sum_{i = 1}^n
        m_i
            \sum_{\phi_i \in t(\vec{x})}
        \left[
            \vcount{t(\vec{x})}(vw)
            -
            \vcount{t(\vec{x})}(uw)
        \right] \\
        &= 
        \sum_{i = 1}^n
        m_i
        \sum_{1 \leq j \leq N_0}
        \sum_{\phi_i \in t_l^j(\vec{y}) \odot t_r^j(\vec{z})}
        \\
        &\quad\quad
        \underbrace{
        \left[
            \vcount{t_r^j(\vec{y})}(v)
            -
            \vcount{t_r^j(\vec{y})}(u)
        \right] 
    }_{ \in \Nat }
            \times 
            \vcount{t_l^j(\vec{z})}(w)
    \end{align*}

    We have proven that if $\MSO^q(u) \leq \MSO^q(v)$, then $u \resleq{f}{k}
    v$. Recall that $\Nat^p$ is a \kl{well-quasi-ordering} when endowed with
    the product ordering, and therefore that $\setof{\MSO^q(u)}{u \in
    \Sigma^*}$ is a \kl{well-quasi-ordering}.

    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
    Without loss of generality, one can assume that for all $i \neq j$, $u_i
    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
\end{proof}

### 2024-01-25 parts/conclusion.tex
writing to publish

@@
 
+ This suggests the
+following conjecture in the non-commutative case:
+
+\begin{conjecture}
+    Let $k \in \Nat$,
+    and $f \in \ZPoly[k]$. Then, $f \in \NPoly[k]$
+    if and only if $f$ is \kl{combinatorial}, which is decidable,
+    and an effective conversion procedure exists.
+\end{conjecture}
+
 \begin{center}
@@
 \end{center}
+
+\begin{conjecture}
+    For commutative functions, the residual transducer 
+    is helpful.
+\end{conjecture}
+
+\begin{fact}
+    \label{translation-invariance:fact}
+    The following sets are stable under \kl{translations}
+    $\translate{K}$,
+    where $K \in \Nat$ is non-negative:
+    \begin{itemize}
+        \item $\CorrectPoly$,
+        \item \kl{$\Nat$-rational polynomials},
+        \item Polynomials \kl{represented}
+            by \kl{star-free $\Nat$-polyregular functions}.
+    \end{itemize}
+\end{fact}
+
+\begin{conjecture}
+    Let $f \in \NPoly$. The following are equivalent,
+    \begin{enumerate}
+        \item $f \in \ZSF$,
+        \item the \kl{$k$-residual transducer} of $f$
+            is \kl{counter-free},
+            and its labels belong to $\ZSF$.
+    \end{enumerate}
+\end{conjecture}
+\begin{proof}
+    If it has a counter free residual transducer recursively, then
+    it clearly belongs to $\ZSF$.
+
+    % TODO: mix orderings and trees and maximal elements :(
+
+    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
+    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
+    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
+    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
+    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
+    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
+    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
+    w_2) = t$.
+
+    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
+    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
+    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
+    strict prefix of $s_0$. It must exist, because otherwise the two are always
+    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
+    this contradicts the minimality of $u$.
+
+    We have proven that $s_0$ and $s_1$ are comparable, hence they
+    are equal.
+    Finally, we have proven that $\delta(q, w_1) = s_0$,
+    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
+
+    As a consequence $t$ and $q$ were comparable for the prefix relation,
+    hence equal, and therefore $\delta(q, w) = q$.
+\end{proof}
+
+We do not know if \kl{commutativity} is decidable for more powerful models of
+computation, and conjecture that it is decidable for \kl{polyregular functions}
+in general.
+
+\begin{conjecture}
+    Let $f \in \Poly$. One can decide if $f$ is \kl{commutative}.
+\end{conjecture}

### 2024-01-25 parts/correction.tex
writing to publish

--- 2024-01-18 parts/correction.tex
@@
 $\CorrectPoly$ under assignments of variable is enough to recover from
-\cref{karh:thm}. Because the previous attempt to characterize this class of
-polynomials did not hold, this part of the paper will try to be extremely
-precise.
+\cref{karh:thm}.
 
@@
 Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-The following \cref{n-poly-combinatorics:lem} is a slight refinement over the
-classical combinatorial analysis of \kl{$\Rel$-polyregular functions}
-\cite[Lemma 4.16]{LOPEZ23b} and \kl{$\Nat$-polyregular functions} that takes
-into account \emph{all} \kl{maximal monomials} instead of a specific one
-\cite[Lemma 5.37]{gaetanphd}.
+This follows from the correct implication in the statement of \cref{karh:thm},
+but we provide a self-contained proof using a refinement of the
+classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{LOPEZ23b} and
+$\NPoly$ \cite[Lemma 5.37]{gaetanphd}.
 
@@
 \end{lemma}
-\begin{proof}
-    Let $\omega$ be an idempotent power for the finite monoid $M$,
-    and
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
-    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
-    $f = \pi^\dagger$. That is, 
-    $f(w)$ is the sum over all factorizations of $w$
-    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
-    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
-
-    Let us define an equivalence relation $\equiv$ over words of
-    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
-    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
-    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
-    repetitions of the factors $u_i^{\omega}$ in the words.
-    We lift the equivalence relation $\equiv$ to $k$-tuples of
-    words by pointwise application.
-
-    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then
-    $\pi(\mu(v_1), \dots, \mu(v_k)) = \pi(\mu(w_1), \dots, \mu(w_k))$ because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid
-    $M$. In particular, $f(w)$ can be computed as the sum over equivalence
-    classes for $\equiv$, of the value of $\pi$ on one representative,
-    which belongs to $\Nat$.
-
-    Therefore, it suffices to prove that the number of partitions of $w$ in a
-    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
-    with \kl{positive coefficients} for its \kl{maximal monomials}.
-
-    Notice that to produce all elements in a given equivalence class
-    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
-    back to the normalized version
-    of the equivalence class. For all elements in an equivalence class,
-    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
-    and therefore the number of possible choices
-    is either constant, or 
-    is a product of binomials of the form $\binom{X - l}{s}$
-    which has the desired shape.
-\end{proof}
 
@@
     there exists a number $\omega \in \Nat$
-    and polynomial $Q$ with \kl{positive coefficients} for its
+    and polynomial $Q$ with \kl{non-negative}
     \kl{maximal monomials} such that
@@
         f\left(
-            \prod_{i = 1}^p (a_i)^{\omega \times X_i}
+            \prod_{i = 1}^p a_i^{\omega X_i}
         \right)
@@
     the \kl{maximal monomials} of 
-    $P$ have \kl{positive coefficients}.
+    $P$ are \kl{non-negative}.
 
@@
     the \kl{maximal monomials} of
-    $\restr{P}{\nu}$ also have \kl{positive coefficients}, and 
+    $\restr{P}{\nu}$ are also \kl{non-negative}, 
+    and
     we have proven that $P \in \CorrectPoly$.
@@
 \AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
-\kl{discrete derivative} to perform an induction on the \kl{maximal monomials},
-and suitable notions of \kl{translations} of the domain of polynomials. Namely,
-we write $\translate{K}$ for the \intro{translation function} that maps a
-polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 + K, \dots,
-X_n + K)$.
-
-\begin{fact}
-    \label{translation-invariance:fact}
-    The following sets are stable under \kl{translations}
-    $\translate{K}$,
-    where $K \in \Nat$ is non-negative:
-    \begin{itemize}
-        \item $\CorrectPoly$,
-        \item \kl{$\Nat$-rational polynomials},
-        \item Polynomials \kl{represented}
-            by \kl{star-free $\Nat$-polyregular functions}.
-    \end{itemize}
-\end{fact}
-
-The idea behind the \kl{discrete derivative} introduced in
-\cref{discrete-derivative:def} is to consider the map $P \mapsto P(X) -
-P(X-K)$, which will be formulated using the \kl{translation function} in the
-case of multiple variables.
+\kl{discrete derivative} to perform an induction on the \kl{maximal monomials}.
+This notion of derivation is built by translating the domain of the polynomial.
+To that end, let us write $\translate{K}$ for the \intro{translation function}
+that maps a polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 +
+K, \dots, X_n + K)$.
 
@@
         \Diff{K}{P} \defined 
-        P - \translate{-K}(P) \quad .
+        \translate{K}(P) - P
+        \quad .
     \end{equation*}
@@
 
-Let us now prove that for a single indeterminate $X$, polynomials $P \in
-\CorrectPoly$ are \kl{$\Nat$-rational polynomials}. This will illustrate how
-the different ingredients piece together. Given a polynomial $P \in
-\CorrectPoly$, the goal is to build a \kl{$\Nat$-polyregular function} $f$ that
-\kl{represents} $P$. Note that in this case, the function has unary input and
-unary output. We prove the result on the degree $\deg(P)$ of the polynomial
-$P$, where the base case is trivial because when $P$ is a constant function,
-this constant must be in $\Nat$. For the induction step, we can assume that $P
-= \alpha X^n + Q$, where $\deg(Q) < \deg(P)$, $n \geq 1$, and $\alpha > 0$. On
-the one hand, $\alpha X^n$ is a \kl{$\Nat$-rational polynomial}, but on the
-other hand, $Q$ may not belong to $\CorrectPoly$. However, there exists $K \in
-\Nat$ such that $\Diff{K}{X^n} + Q$ belongs to $\CorrectPoly$, for instance by
-considering $K$ to be the sum of absolute values of coefficients of $P$.
-Furthermore, $\Diff{K}{X^n} + Q$ has degree smaller than $\deg(P)$, hence is a
-\kl{$\Nat$-rational polynomial}. Now, $P = \alpha (X - K)^n + \Diff{K}{X^n} +
-Q$, and therefore, $P(X+K) = X^n + (\Diff{K}{X^n} + Q)(X+K)$ is a
-\kl{$\Nat$-rational polynomial}, \kl{represented} by a \kl{$\Nat$-polyregular
-function} $f_1$. Now, for values $k$ of $X$ that are smaller than $K$, we can
-compute $P(k) = \restr{P}{X=k}$, which belongs to $\Nat$. It is possible to
-build a \kl{$\Nat$-polyregular function} $f_2 \colon \Nat \to \Nat$ that
-outputs $P(k)$ if the input is at most $K$, and $f_1(k)$ otherwise.
-
-In the actual statement of \cref{corrected-version:thm}, the representation of
-polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
-functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
-surprising, given the conjecture that \kl{ultimate polynomial}
-\kl{$\Nat$-polyregular functions} (re-introduced in \cref{star-free:sec}) were
-conjectured to be star free in \cite{LOPEZ23b}.
-
-When multiple indeterminate are involved, a bit of care is needed. The key
-ingredient is that one has a fine control over the \kl{maximal monomials} of
-the \kl{discrete derivatives}, even when fixing some indeterminates.
 As one would expect, the \kl{discrete derivatives} are linear operations on
-polynomials, that commutes with the partial application operators
-$\restr{\cdot}{\nu}$, and the \kl{translation operators} $\translate{K}$.
+polynomials, that commutes with \kl{translation operators}. However, the
+\kl{translations operators} do not perfectly commute with partial applications
+$\restr{\cdot}{\nu}$.
 
@@
     \label{discrete-deriv-linear:fact}
-    For all tuple $\vec{X}$ of indeterminates,
+    For all tuples $\vec{X}$ of indeterminates,
     for all $K \in \Nat$, $L \in \Rel$, for all partial function
@@
     \begin{enumerate}
-        \item $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
-            is a linear operator,
-        \item $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
-            is a linear operator,
-        \item $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
-            is a linear operator,
-        \item $\Diff{K}{ \cdot } \circ \restr{\cdot}{\nu}
-            = \restr{\cdot}{\nu} \circ \Diff{K}{\cdot}$.
+        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
+        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
+        and
+        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
+            are linear operators,
         \item $\Diff{K}{ \cdot } \circ \translate{L}
@@
 
-\begin{fact}
-    \label{translation-maximal:fact}
-    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
-    $\MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))$.
-\end{fact}
+To build some intuition on the proof to come, let us sketch the argument in the
+case of a single indeterminate $X$. Given a polynomial $P \in \CorrectPoly$,
+the goal is to build a \kl{$\Nat$-polyregular function} $f$ that
+\kl{represents} $P$. Note that in this case, the function has unary input and
+unary output. We prove the result by induction on the degree $\deg(P)$ of the
+polynomial $P$, where the base case is trivial because when $P$ is a constant
+function, this constant must be in $\Nat$. For the induction step, we can
+assume that $P = \alpha X^n + Q$, where $\deg(Q) < \deg(P)$, $n \geq 1$, and
+$\alpha > 0$. On the one hand, $\alpha X^n$ is a \kl{$\Nat$-rational
+polynomial}, but on the other hand, $Q$ may not belong to $\CorrectPoly$.
+However, there exists $K \in \Nat$ such that $\Diff{K}{X^n} + \translate{K}(Q)$
+belongs to $\CorrectPoly$, for instance by considering $K$ to be the sum of
+absolute values of coefficients of $P$. Furthermore, $\Diff{K}{X^n} +
+\translate{K}(Q)$ has degree smaller than $\deg(P)$, hence is a
+\kl{$\Nat$-rational polynomial}. Now, $\translate{K}(P) = \alpha X^n +
+\Diff{K}{X^n} + \translate{K}(Q)$ is a \kl{$\Nat$-rational polynomial},
+represented by a function $f_1$. Now, for values $k$ of $X$ that are smaller
+than $K$, we can compute $P(k) = \restr{P}{X=k}$, which belongs to $\Nat$. It
+is possible to build a \kl{$\Nat$-polyregular function} $f_2 \colon \Nat \to
+\Nat$ that outputs $P(k)$ if the input is at most $K$, and $f_1(k - K)$
+otherwise.
 
-\AP Let us now introduce the ordering over which the induction of
+In the actual statement of \cref{corrected-version:thm}, the representation of
+polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
+functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
+surprising, given the conjecture that \kl{ultimately polynomial}
+\kl{$\Nat$-polyregular functions} (re-introduced in \cref{star-free:sec}) were
+conjectured to be star free (\cref{zsf-nsf:conjecture}), and that polynomials
+are always \kl{ultimately polynomial}.
+
+When multiple indeterminate are involved, the notion of 
+degree must be adapted, and to ensure that the decision procedure
+is effective, one must obtain explicit bounds on the
+translations that are needed.
+
+\AP To that end, let us now introduce the ordering over which the induction of
 \cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
@@
 
-The following \cref{derivation-simplifies:lemma,derivation-covers:lemma} 
-show how the \kl{discrete differentiation} operator
-interacts with \kl{maximal monomials} with respect to the \kl{Hoare ordering}:
-it extracts \emph{submaximal} monomials from a given polynomial.
+\begin{fact}
+    \label{maxi-monomials-submodular:fact}
+    Let $P, Q$ be two polynomials.
+    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
+    \cup \MaximalMonomials(Q)$.
+\end{fact}
+
+Let us now illustrate how the \kl{discrete differentiation} operator interacts
+with \kl{maximal monomials} with respect to the \kl{Hoare ordering}: it
+extracts \emph{submaximal} monomials from a given polynomial.
+
+\begin{fact}
+    \label{translation-maximal:fact}
+    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
+    \begin{equation*}
+        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
+        \quad .
+    \end{equation*}
+\end{fact}
+
+\begin{fact}
+    \label{derivation-monomials:fact}
+    Let $K > 0$,
+    and let $M,T$ be two \kl{monomials}, such that
+    $T$ \kl{strictly divides} $M$.
+    Then,
+    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
+    and 
+    $T$ \kl{divides} some \kl{maximal monomial}
+    of $\Diff{K}{M}$, which has a coefficient
+    that is a multiple of $K$.
+\end{fact}
 
@@
 \end{lemma}
-\begin{proof}
-    By linearity of $\Diff{K}{P}$ (\cref{discrete-deriv-linear:fact}), 
-    it suffices to prove the result
-    for \kl{monomials} $S \in \Monomials$.
-\end{proof}
 
-\begin{fact}
-    Let $K > 0$,
-    and let $M,T$ be two \kl{monomials}, such that
-    $T$ \kl{strictly divides} $M$.
-    Then,
-    $T$ \kl{divides} some \kl{maximal monomial}
-    of $\Diff{K}{M}$.
-\end{fact}
-
-\begin{lemma}
-    \label{derivation-covers:lemma}
-    For all $P,Q \in \Rel[\vec{X}]$ that are non-constant,
-    for all $K > 0$,
-    if $\MaximalMonomials(P) \hoarele \MaximalMonomials(Q)$,
-    then
-    $\MaximalMonomials(P) \hoareleq \MaximalMonomials(\Diff{K}{Q})$.
-\end{lemma}
+The following \cref{derivation-stabilises-correct:lem} is the main
+combinatorial argument of this section. Note that we must be careful
+about it. \textbf{TODO: write a nice thing here!}.
 
@@
     such that
-    $(\Diff{K}{P_1} + P_2) \in \CorrectPoly$.
+    $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$.
 \end{lemma}
@@
 \end{lemma}
-\begin{proof}
-    Let $P \in \Rel[\vec{X}]$. We prove by induction on
-    $\MaximalMonomials(P)$ that whether $P \in \CorrectPoly$
-    is decidable.
-
-    If $P$ is a constant polynomial, then
-    $P \in \CorrectPoly$ if and only if $P = n \in \Nat$
-    which is decidable.
-
-    Otherwise, $P$ is not a constant polynomial, and let us write $P = P_1 +
-    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. If $P_1
-    \not \in \Nat[\vec{X}]$, then $P \not \in \CorrectPoly$, and this is
-    decidable. Hence, we can assume that $P_1 \in \Nat[\vec{X}]$. Using
-    \cref{derivation-stabilises-correct:lem}, there exists a computable bound
-    $K$, such that if $P \in \CorrectPoly$, then t $Q \defined (\Diff{K}{P_1} +
-    P_2) \in \CorrectPoly$. Leveraging, \cref{derivation-simplifies:lemma}, we
-    also know that $\MaximalMonomials(Q) \hoarele \MaximalMonomials(P)$, and
-    therefore that we can decide whether $Q \in \CorrectPoly$. Notice that if
-    $Q \not \in \CorrectPoly$, then $P \not \in \CorrectPoly$, hence we can
-    assume that $Q \in \CorrectPoly$.
-
-    Similarly, for partial valuations
-    $\nu \colon \vec{X} \topartial \set{0, \dots, K}$
-    fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
-    \hoarele \MaximalMonomials{P}$. 
-    We can check for all these partial valuations that
-    $\restr{P}{\nu} \in \CorrectPoly$ by induction hypothesis,
-    and if one test is not satisfied, then $P \not \in \CorrectPoly$.
-
-    Let us now conclude that if all the above checks answered correctly,
-    then $P \in \CorrectPoly$. Let $\nu \colon \vec{X} \topartial \Nat$
-    be a partial valuation. Then, 
-    one of the three following cases arise:
-    \begin{itemize}
-        \item 
-            Either,
-        $\nu$ fixes \emph{some} indeterminate $X_i$ to a value $n_i < K$.
-        In that case, let us define $\delta$ as the partial function that coincides
-        with $\nu$ on its domain
-        $\dom(\delta) = \dom(\nu) \setminus \set{ X_i }$.
-        Then, $\restr{P}{\nu} = \restr{\restr{P}{X_i = n_i}}{\delta}$.
-        As we previously checked that $\restr{P}{X_i = n_i} \in \CorrectPoly$,
-        we conclude in particular that 
-        $\restr{P}{\nu}$ has \kl{non-negative} \kl{maximal monomials}.
-
-        \item Or, for all $X_i$ in the (potentially empty) domain of $\nu$, $\nu(X_i)
-            \geq K$. In that case, let us write $\delta \defined
-            \translate{-K}(\nu)$, that is the partial map $\delta \colon
-            \vec{X} \topartial \Nat$ having the same domain as $\nu$, and such
-            that for all $X_i \in \dom(\nu)$, $\delta(X_i) \defined \nu(X_i) -
-            K$. Recall that $Q = \Diff{K}{P_1} + P_2$, and therefore that
-            $\translate{K}(P) = P_1 + \translate{K}(Q)$. By linearity,
-            $\restr{\translate{K}(P)}{\delta} = \restr{P_1}{\delta} +
-            \restr{\translate{K}(Q)}{\delta}$, and we conclude that the
-            \kl{maximal monomials} of $\restr{\translate{K}(P)}{\delta}$ are
-            \kl{non-negative}.
-            Since $\restr{\translate{K}(P)}{\delta} = \translate{K}{\restr{P}{\nu}}$,
-            and since
-            $\translate{K}$ leaves \kl{maximal monomials} invariant,
-            we conclude that
-            the \kl{maximal monomials} of $\restr{P}{\nu}$
-            are the same as those of $\restr{\translate{K}(P)}{\delta}$,
-            and are therefore \kl{non-negative}.
-            \qedhere
-    \end{itemize}
-\end{proof}
 
@@
     that is \kl{non-negative}, such that the \kl{maximal monomials}
-    have \kl{positive coefficients}.
+    are  \kl{non-negative}.
 
     If we fix none of the variables, then the \kl{maximal monomials}
-    have \kl{positive coefficients} by assumption. If we fix one of the
+    are \kl{non-negative} by assumption. If we fix one of the
     variables, we can assume without loss of generality that we 
@@
     or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have \kl{positive coefficients}.
+    have positive coefficient.
     The same reasoning applies \emph{a fortiori} in the case where
@@
 
-It was conjectured in the 
-\cite[Open question 5.55]{gaetanphd}.
-\cite[Conjecture 7.61]{gaetanphd}
-Let us remark that a previous conjecture was not true, and provide the
-recursively defined variant. This can be understood in the light of
-\cref{corrected-version:thm}, as the fact that \kl{$\Nat$-rational polynomials}
-are essentially $\Nat$-polynomials.
-
-\begin{example}
-    The polynomial $\BadPoly$ is not a 
-    \kl{$\Nat$-rational polynomial},
-    but is \kl{non-negative} and satisfies
-    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
-\end{example}
+Let us also state that polynomials in $\CorrectPoly$ are, up to suitable
+translations, polynomials in $\Nat[\vec{X}]$. Beware that partial applications
+are still needed in this characterization, as \cref{bad-poly-translate:ex}
+illustrates.
 
@@
     \label{derivation-translation:lem}
-    Let $P \in \Rel[\vec{X}]$, the following are equivalent
+    Let $P \in \Rel[\vec{X}]$, the following are equivalent:
     \begin{enumerate}
@@
 \end{lemma}
-\begin{proof}
-    To prove \cref{d-t-transl:item} $\implies$
-    \cref{d-t-correct:item}, let us remark that
-    if $P$ is \kl{non-negative} and $\translate{K}(P) \in \Nat[\vec{X}]$,
-    then $\translate{K}(P) \in \CorrectPoly$ thanks to
-    \cref{n-poly-n-poly:example}.
-    In particular, 
-    then for every partial valuation 
-    $\nu \colon \vec{X} \topartial \Nat$, 
-    $\restr{\translate{K}(P)}{\nu} \in \Nat[\vec{X}]$,
-    and in particular conclude that
-    \kl{maximal monomials} of $\restr{\translate{K}(P)}{\nu}$ are exactly the
-    \kl{maximal monomials} of $\restr{P}{\nu}$, the former being
-    \kl{non-negative} if $\translate{K}(P) \in \Nat[\vec{X}]$.
 
-    Conversely, the implication \cref{d-t-correct:item} $\implies$
-    \cref{d-t-transl:item} is proven by induction on $\MaximalMonomials(P)$.
-    If $P$ is a constant polynomial, then $P = n$ for some $n \in \Nat$,
-    and we conclude that $\translate{0}(P) \in \Nat[\vec{X}]$.
-    Otherwise, $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
-    monomials} of $P$.
-    Using \cref{derivation-stabilises-correct:lem},
-    there exists a computable $K \in \Nat$
-    such that 
-    $Q \defined
-    \Diff{K}{P_1} + P_2$ belongs to $\CorrectPoly$.
-    Since, $\MaximalMonomials(Q) \hoarele \MaximalMonomials(P)$,
-    we conclude by induction hypothesis that
-    there exists $L \in \Nat$ such that 
-    for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
-    $\translate{L}(\restr{Q}{\nu}) \in \Nat[\vec{X}]$. 
-    Similarly,
-    for all partial functions $\nu \colon \vec{X} \topartial \Nat$
-    fixing at least one indeterminate (appearing in $P$),
-    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials(P)$,
-    hence there exists $L_\nu \in \Nat$ such that
-    $\translate{L_\nu}{\restr{P}{\nu}} \in \Nat[\vec{X}]$.
-    Let us define $L_m$ to be the maximum of $L$, and $L_\nu$
-    where $\nu$ ranges over partial functions described above.
-
-    Let us prove that for all partial functions
-    $\nu \colon \vec{X} \topartial \Nat$,
-    $\translate{L_m + K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
-    If $\nu$ is the empty valuation, then $\restr{P}{\nu} = P$,
-    and
-    $\translate{L_m + K}(P) = \translate{L_m}(P_1) + \translate{K+L_m}(Q)$
-    is a sum of two polynomials in $\Nat[\vec{X}]$.
-    Otherwise,
-    $\translate{L_m + K}(\restr{P}{\nu}) = \translate{K}(\translate{L_m}(\restr{P}{\nu}))$
-    belongs to $\Nat[\vec{X}]$ since the latter is stable under
-    positive translations.
-\end{proof}
+\begin{example}
+    \label{bad-poly-translate:ex}
+    The polynomial $\BadPoly$ is not a 
+    \kl{$\Nat$-rational polynomial},
+    but is \kl{non-negative} and satisfies
+    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
+\end{example}

### 2024-01-25 parts/counter.tex
writing to publish

@@
 \section{Polynomials}
+\label{polynomials:sec}
 
@@
 
-\begin{lemma}
+\begin{fact}
     \label{n-poly-n-poly:example}
@@
     is a \kl{$\Nat$-rational polynomial}.
-\end{lemma}
-\begin{proof}
-    We prove it by remarking that $\NPoly$
-    is stable under non-negative linear combinations
-    and contains the functions ...
-\end{proof}
+\end{fact}
 
@@
     \label{negative-but-npoly:ex}
-    The polynomial $P(X) \defined X^2 -2X + 2$ is
-    a \kl{$\Nat$-rational polynomial}.
+    Let $P \defined X^2 -2X + 2$. Then $P$
+    is a \kl{$\Nat$-rational polynomial}.
 \end{example}
@@
 Let us now introduce the counter example that was found trying to generalize
-the original proof to non-commutative \kl{$\Rel$-rational series} of
-\kl{polynomial growth}. The counter example will use three indeterminate, and
+the original proof $\ZPoly$.
+The counter example will use three indeterminates, and
 we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
@@
 \begin{proof}
-    It is clear that $\BadPoly$ is \kl{non-negative} as a sum of two 
-    squares. We can develop
+    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
     the expression of $\BadPoly$ to 
@@
     \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ is notoriously not a regular language
-    (\cref{regular:fact}), we
+    Because $L \cap \set{a,b}^*$ not a regular language,
+    we
     conclude that $L$ is not a regular language.
-    However, $L = f^{-1}(\set{0})$, which is a regular language
-    because $f$ is a \kl{$\Nat$-rational series}
+    However, $L = f^{-1}(\set{0})$ is a regular language
     (\cref{pre-image-regular:fact}). 
@@
     in \cref{karh:thm}, is false
-    for all polynomials with at least $3$ indeterminate.
+    when allowing at least $3$ indeterminates.
 \end{corollary}

### 2024-01-25 parts/intro.tex
writing to publish

@@
 characterizations, namely the \emph{star-free} regular languages. These are
-obtained via counter-free automata \cite{mcnaughton1971counter}, represented by
-star-free regular expressions \cite{schutzenberger1965finite}, definable in
-First Order Logic ($\FO$) \cite{perrin1986first}, and recognized by aperiodic
-monoids \cite{schutzenberger1965finite}. It is decidable whether a regular
-language is aperiodic, thanks to the characterizations in terms of minimal
-automaton and/or syntactic monoid.
+obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
+represented by star-free regular expressions \cite{schutzenberger1965finite},
+definable in First Order Logic ($\FO$) \cite{perrin1986first}, and recognized
+by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
+regular language is aperiodic, thanks to the characterizations in terms of
+minimal automaton and/or syntactic monoid.
 
@@
 languages to functions, leading to classes of various expressiveness
-(sequential, rational, regular, and \kl{polyregular functions}) \cite[see,
-e.g.][]{bojanczyk2019string}. In this paper, we will focus on \kl{polyregular
-functions}, whose class is denoted $\Poly$, and that shares with regular
-languages the characterizations in terms of logic (via $\MSO$ interpretations),
+(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
+we will focus on \kl{polyregular functions} ($\Poly$), that shares with regular
+languages its characterizations in terms of logic (via $\MSO$ interpretations),
 automata (via pebble automata), and adds new bridges with programming languages
-via its relationships with lambda-terms and so-called for-transducers. As for
-regular languages, the class $\SF$ of \intro{star-free polyregular function}
-can equivalently be introduced in terms of first-order definability,
-counter-free pebble transducers, and suitable restrictions of for-tranducers.
+via its relationships with lambda-terms and so-called for-transducers
+\cite[see, e.g.][]{bojanczyk2019string}. As for regular languages, the class
+$\SF$ of \intro{star-free polyregular function} can equivalently be introduced
+in terms of first-order definability, counter-free pebble transducers, and
+suitable restrictions of for-tranducers.
 
@@
 semantics of function equality to account for the \emph{origins} in the output
-\cite{bojanczyk14}.
+\cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
+interesting parameter is the \kl{growth rate} of a function $f$, which is given
+by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
+$\Poly$. We write $\Poly[k]$ for the \kl{polyregular functions} of \kl{growth
+rate} $k$. It has been shown that the parameter $k$ controls the maximal
+complexity of the $\MSO$ interpretation needed to represent the function
+\cite{bojanczyk2022transducers}.
 
-Another way to approach the problem of deciding star-free for polyregular
-functions is to not restrict the computational power, nor change the semantics,
-but consider restrictions of the output alphabet. In particular, there have
-been successful results proven on \emph{unary polyregular functions}, also
-known as \kl{$\Nat$-polyregular functions}
-\cite{doueneau2021pebble,doueneau2022hiding}. Let us note that results on
-\emph{unary output functions} often generalize to the slightly more general
-class of functions with \emph{commutative output}, and that algorithms leverage
-the absence of combinatorial information in the output. Furthermore,
-\kl{$\Nat$-polyregular functions} benefit from theorems proven in the field of
-rational series \cite{berstel2011noncommutative}, where they coincide with
-weighted automata of polynomial ambiguity, rational series of polynomial
-growth, and a notion of quantitative $\MSO$
-\cite{berstel2011noncommutative,kreutzer2013,schutzenberger1962}. Although
-promising, the restriction to unary output did not yield effective procedure to
-decide the aperiodicity of polyregular functions.
+Another way to approach the problem of deciding if a \kl{polyregular function}
+is \kl{star-free}, is to not restrict the computational power (via the
+\kl{growth rate}), nor change the semantics (by adding origins), but consider
+restrictions of the output alphabet. In particular, there have been successful
+results proven on \emph{unary polyregular functions}, also known as
+\kl{$\Nat$-polyregular functions} ($\NPoly$)
+\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
+rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
+needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
+\cite{bojanczyk2022transducers}. Let us remark that the restriction to
+\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
+outputs in the free commuative semigroup generated by a finite alphabet
+$\Gamma$. In this sense, the shift to \emph{unary output} is really about
+enforcing commutativity on the output.
 
 Recently, aperiodicity has been shown decidable over a restricted class of
-\kl{polyregular functions} called \kl{$\Rel$-polyregular functions} and denoted
-by $\ZPoly$ \cite{LOPEZ23b}.  These functions generalise \kl{$\Nat$-polyregular
-functions} by allowing negative outputs, and therefore enabling a limited form
-of backtracking. The decidability of aperiodicity for \kl{$\Rel$-polyregular
-functions} relies on the construction of a canonical object called the
-\intro{residual transducer}, the latter being essentially based on differences
-between functions, crucially leveraging negative outputs.
-If the proof method does not carry from $\Rel$-output to $\Nat$-output
-functions, it was conjectured that the semantic characterization of \kl{star
-free $\Rel$-polyregular functions} ($\ZSF$) would continue to characterize
-\kl{star-free $\Nat$-polyregular functions} ($\NSF$), and in particular that
-$\NPoly \cap \ZSF = \NSF$.
+\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
+$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\ZPoly$)
+\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
+by allowing negative outputs, and therefore enabling a limited form of
+backtracking. As for the unary outputs, these results generalize to
+\kl{polyregular functions} having outputs in the free commutative \emph{group}
+generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
+\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
+object called the \emph{residual transducer}, the latter being essentially
+based on differences between functions, crucially leveraging negative outputs
+(i.e., the presence of inverse elements). If the proof method does not carry
+from $\Rel$-output to $\Nat$-output functions, it was conjectured that the
+semantic characterization of \kl{star-free $\Rel$-polyregular functions}
+($\ZSF$) would continue to characterize \kl{star-free $\Nat$-polyregular
+functions} ($\NSF$).
 
-Let us remark that, \kl{$\Nat$-polyregular functions} and
-\kl{$\Rel$-polyregular functions} are proper subsets respectively of
-\kl{$\Nat$-rational series} ($\ZRat$) and of \kl{$\Rel$-rational series}
-($\NRat$). For these larger classes of functions, several competing notions of
-aperiodicity exist \cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b},
-and decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
-Furthermore, the decidability of $\NRat$ among $\ZRat$ is also a longstanding
-open problem, that was to the knowledge of the authors, only solved in the case
-of \emph{polynomials} by \textcite{KARH77}.
+\begin{conjecture}[{\cite[Conjecture 7.61]{gaetanphd}}]
+    \label{zsf-nsf:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \NPoly[k]$
+    belongs to $\NSF[k]$ if and only if
+    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
+    In particular,
+    $\NSF[k] = \ZSF[k] \cap \NPoly$.
+\end{conjecture}
+
+It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
+of \kl{$\Nat$-rational series} ($\NRat$) (resp. of \kl{$\Rel$-rational series}
+$\ZRat$), that have been characterized in terms of \kl{growth rate}, ambiguity,
+and a notion of \emph{quantitative} $\MSO$
+\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
+theorems obtained on these \cite{berstel2011noncommutative}. However, for these
+larger classes of functions, several competing notions of aperiodicity exist
+\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
+decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
+Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
+open problem, that was to the knowledge of the authors, only solved by
+\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
+\emph{polynomials}. The decidability of $\NPoly$ inside $\ZPoly$ (resp. $\NSF$
+inside $\ZSF$) remains an open problem
+\cite[see e.g.][Open question 5.55]{gaetanphd}.
 
@@
         arranged along two axes. The first one is the complexity
-        of the output alphabet ($\Nat$, $\Rat$, $\Sigma$). The second
+        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
         one is the allowed computational power
@@
 
-\paragraph{Contributions.} We prove that for \kl{commutative} functions, all
-the membership problems involving the classes $\ZSF$, $\ZPoly$, $\NSF$, and
-$\NPoly$ are all decidable (this is the lower left square of inclusions in
-\cref{previously-known-inclusions:fig}), and effective conversion algorithms
-are provided. Furthermore, we show that in the commutative case, the conjecture
-stating that $\ZSF \cap \NPoly = \NSF$ holds.
+\paragraph{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
+functions} that are furthermore \kl{commutative}. Remark that this enforces a
+commutativity property both on the input alphabet and on the output alphabet.
+In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
+and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
+$\ZPoly$ are all decidable, with effective conversion algorithms. This
+effectively shows that all arrows in the lower left square of
+\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
+setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
+function} as the class of \kl{$\Rel$-polyregular functions} that are
+\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
+property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
+As a byproduct of our analysis, we provide a counter example to the
+characterization of polynomials computed by \kl{$\Nat$-rational series} given
+by \textcite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
 
-These results are obtained by completely understanding which \emph{polynomials}
-are computable using \kl{$\Nat$-polyregular functions}, and lifting these
-results from polynomials to \kl{$\Rel$-polyregular functions}. This study
-invalidates the main theorem of \cite{KARH77}, for which we provide an explicit
-counter-example.
+Finally, we make a first step towards understanding the non-commutative by
+introducing a canonical model of computation for \kl{$\Nat$-polyregular
+functions} built upon residuals, called the \kl{residual transducer}. This
+allows the characterization of $\NPoly[k]$ as the class of functions $f$ such
+that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
+\kl{well-quasi-ordering} with respect to the partial ordering $g \npolyleq[k]
+h$ if and only if $h - g \in \NPoly[k-1]$. In this sense, it generalizes to
+orderings the characterization of $\ZPoly[k]$ as the class of functions $f$
+such that $\Res(f)$ has finite index for the equivalence relation $g
+\zpolyequiv[k] h$ if and only if $g - h \in \ZPoly[k-1]$. The \kl{residual
+transducer} is effectively computable if $f$ is a \kl{commutative}
+\kl{$\Nat$-polyregular function}, and it is conjectured to be computable for
+all functions in $\NPoly$. Finally, we prove that for $f \in \NPoly[0]$, $f \in
+\NSF$ if and only if the \kl{residual transducer} of $f$ has no \kl{counters},
+and conjecture that in full generality, $f \in \NSF$ if and only if the
+\kl{residual transducer} of $f$ is \kl{counter-free}.
 
-\begin{itemize}
-    \item We show that commutativity is decidable.
-    \item We provide a candidate semantic characterization
-    \item We prove that it is correct for commutative inputs
-    \item We provide a counter-example to the theorem of nanania.
-    \item We show that effective procedures are possible for all arrows 
-        and answer positively to the conjecture of XX
-        in the case of commutative inputs.
-\end{itemize}
-It was conjectured in a thesis that some particular form of polynomials
-would characterize star-free functions.
-
-This paper follows the line of
-research on \kl{polyregular functions} that focuses on semantic
-characterizations of the asymptotic behavior, rather than the construction
-of canonical objects of computation
-\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}. The goal was to
-decide arrows X,Y and Z, namely, to understand precisely the relationship
-between star-free and $\Nat$ output. In the case of \kl{$\Nat$-rational
-series}, a result of \citeauthor{KARH77} characterizes \kl{commutative}
-\kl{$\Nat$-rational series} among \kl{$\Rel$-rational series}
-\cite{KARH77}.
-This conjecture was based on the previously introduced result of
-\textcite[Theorem 3.3]{KARH77}, which we will disprove in \cref{sec:c-example}.
-We then provide a corrected version of the theorem, and prove the conjecture of
-Douéneau in the case of \emph{commutative} $\Nat$-rational series of polynomial
-growth.
-
-\cref{thm:counter-example}
-\cref{corrected-version:thm}
-
-\cref{decidable-n-poly:thm}
-\cref{zsf-npoly-nsf:thm}
-
-\hrule
-
-\paragraph*{Outline of the paper.}
-In \cref{preliminaries:sec}, we introduce the computational models
-and prove that 
-\begin{itemize}
-    \item In a first section, we recall the preliminary definitions.
-    \item In a second section, we provide a counter example to the
-        theorem of XX.
-    \item In a third section, we provide an alternative result.
-    \item In a fourth section, we make the results effective,
-        and compute everyting.
-\end{itemize}
+\paragraph{Outline of the paper.} In \cref{preliminaries:sec}, we introduce the
+multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
+$\Rel$-polyregular functions), show that one can decide if a function $f \in
+\ZPoly$ is \kl{commutative} (\cref{decide-commutative:lemma}), and give the
+basic definitions used on multivariate polynomials. In \cref{polynomials:sec},
+we focus on \emph{polynomials}, by introducing the \cref{karh:thm} of
+\textcite{KARH77}, providing a counter-example in \cref{thm:counter-example},
+and an alternative characterization in \cref{corrected-version:thm}. In
+\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
+decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
+(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture} holds
+for \kl{commutative} functions (\cref{zsf-npoly-nsf:thm}). Finally, in
+\cref{beyond-commutative:sec}, we introduce the notion of \kl{residual
+transducer} (\cref{residual-transducer:def}), and prove in
+\cref{non-commutative-npoly:thm} that it characterizes $\NPoly$ inside
+$\ZPoly$.

### 2024-01-25 parts/prelim.tex
update proofs and paper
writing to publish

--- 2024-01-16 parts/prelim.tex
@@
 
+\AP
+In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of integers
+(resp. nonnegative integers). The capital letters $\Sigma,\Gamma$ denotes fixed
+alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$ (resp.
+$\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
+$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
+\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
+$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
+that the reader is familiar with the basics of automata theory, in particular
+the notions of monoid morphisms, idempotents in monoids, monadic second-order
+($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
+\cite{thomas1997languages}).
+
+\AP
+For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^*
+\to \set{0,1}$ as the indicator function of $L$. If $f,g \colon \Sigma^* \to
+\Rel$ are two functions, we define their Cauchy product as $f \cauchy g (w)
+\defined \sum_{uv = w} f(u)g(w)$. Given a set $X$ of functions from $\Sigma^*$
+to $\Rel$, and a subset $\mathcal{S}$ of $\Rel$, we define
+$\Span{\mathbb{S}}(X)$ as the smallest subset of functions from $\Sigma^* \to
+\Rel$ containing $X$ and closed under (finite) sums and multiplication with $s
+\in \mathbb{S}$.
+
+\AP
+Given an $\MSO$ formula with first-order free variables $\varphi(\vec{x})$, we
+define $\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$
+that maps a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$,
+such that $w, \nu \models \varphi(\vec{x})$.
+
+\AP Let $f \colon \Sigma^* \to \Rel$ be a function. We define
+the \intro{growth rate} of
+$f$, $\growth{f} \colon \Nat \to \Nat$, as follows $\growth{f}(n) \defined
+\sup_{\card{w} = n} |f(w)|$. A function $f$ has \kl{growth rate} $k \in \Nat$
+if $\growth{f} = \bigO(n^k)$, and if such a $k$ exists,
+we say that $f$ has \kl{polynomial growth}.
+
 \subsection{Polyregular Functions}
 
-All functions in this paper have output in $\Nat$ or $\Rel$,
-hence, we avoid the definition of \kl{polyregular functions}
-in full generality. The following \cref{nat-rel-poly:def}
-is one of the equivalent definitions of \cite{LOPEZ23b},
-and is similar in shape to the \emph{finite counting automata}
-introduced by \textcite{schutzenberger1962}.
+\AP All functions in this paper have output in $\Nat$ or $\Rel$, hence, we
+avoid the definition of \kl{polyregular functions} in full generality. The
+following \cref{nat-rel-poly:def} is one of the equivalent definitions of
+\cite{LOPEZ23b}, and is similar in shape to the \emph{finite counting automata}
+introduced by \textcite{schutzenberger1962}. Let us recall that a monoid $M$ is
+\intro(monoid){aperiodic} whenever there exists $n \in \Nat$ such that for all
+$x \in M$, $x^{n+1} = x^n$.
 
@@
     is called \intro{$\Nat$-polyregular}.
-    When the monoid $M$ is \kl{aperiodic}
+    When the monoid $M$ is \kl(monoid){aperiodic}
     then
@@
 
-Let us illustrate on an example that these notions actually generalise
-blabla.
+\begin{example}
+    The map $f \colon w \mapsto \card{w} + 1$
+    belongs to $\NSF$.
+\end{example}
+\begin{proof}
+    Let us define $M \defined (\set{1}, \times)$ which is 
+    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
+    defined by $\mu(w) \defined 1$, and
+    $\pi \colon M^2 \to \Nat$
+    that is the constant function equal to $1$.
+    We check that for all $w \in \Sigma^*$:
+    \begin{equation*}
+        \pi^\dagger(w)
+        =
+        \sum_{uv = w} 1
+        =
+        \card{w} + 1
+        = f(w)
+        \quad . 
+        \qedhere
+    \end{equation*}
+\end{proof}
+
+\AP Let us briefly recall in the following lemma alternative characterizations
+of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
+analysis of their \kl{commutative} counterpart, and others that provide
+a nicer syntax to construct examples.
+
+\begin{lemma}[TODO]
+    \label{polynomial-rational-polyreg:fact}
+    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NPoly$ (resp. $\ZPoly$);
+        \item $f$ is a \kl{polyregular function} with output
+            in $\set{1}^*$,
+            postcomposed with $\polysum$
+            (resp. with output in $\set{-1,+1}^*$);
+        \item $f$ has \kl{polynomial growth};
+        \item $f$ can be constructed
+            using $+$, $\cauchy$, and multiplication by
+            an integer $m \in \Nat$ (resp. $m \in \Rel$)
+            starting from
+            indicator functions $\ind{L}$ of regular languages;
+        \item $f$ belongs to
+            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
+            with $\mathbb{S} = \Nat$
+            (resp. $\mathbb{S} = \Rel$).
+    \end{enumerate}
+\end{lemma}
+
+Similarly, we provide several known characterizations of
+$\NSF$ and $\ZSF$ in the following lemma.
+
+\begin{lemma}[TODO]
+    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NSF$ (resp. $\ZSF$);
+        \item $f$ is a \kl{star-free polyregular function} with output
+            in $\set{1}^*$,
+            postcomposed with $\polysum$
+            (resp. with output in $\set{-1,+1}^*$);
+        \item $f$ can be constructed
+            using $+$, $\cauchy$, and multiplication by
+            an integer $m \in \Nat$ (resp. $m \in \Rel$)
+            starting from
+            indicator functions $\ind{L}$ of star-free regular languages;
+        \item $f$ belongs to
+            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
+            with $\mathbb{S} = \Nat$
+            (resp. $\mathbb{S} = \Rel$).
+    \end{enumerate}
+\end{lemma}
 
@@
 
-\AP Notice that for a \kl{$\Rel$-polyregular function} $f \colon \Sigma^* \to
-\Rel$, the \intro{growth rate} of $f$, defined as the map $\growth{f} \colon
-\Nat \to \Nat$ via $\growth{f}(n) \defined \sup_{\card{w} = n} |f(w)|$, is
-bounded by a polynomial in $n$. This property of having a \intro{polynomial
-growth} actually characterizes $\ZPoly$ (resp. $\NPoly$) inside $\ZRat$ (resp.
-$\NRat$). We formally state this property in \cref{polynomial-rational-polyreg:fact}.
-
-\begin{fact}[{\cite[Exercise 1.2 p 169]{berstel2011noncommutative}}]
-    \label{polynomial-rational-polyreg:fact}
-    Let $\Sigma$ be a finite alphabet, and $f \colon \Sigma^* \to \Nat$
-    be a function. The following are equivalent:
-    \begin{enumerate}
-        \item $f$ is a \kl{$\Nat$-rational series}
-            of \kl{polynomial growth}.
-        \item $f$ is a \kl{$\Nat$-polyregular function}.
-    \end{enumerate}
-\end{fact}
-
 \AP Inside $\ZPoly$, there is an infinite hierarchy of classes, defined using
 the \kl{growth rate}. Namely, one defines $\ZPoly[k]$ as the functions $f \in
-\ZPoly$ such that $\growth{f}(n) = \bigO(n^k)$. Similarly, we let $\NPoly[k] =
-\NPoly \cap \ZPoly[k]$. It follows from the results of \cite{LOPEZ23b} that
-this hierarchy is nice and for every $f$, there exists $k$ such that
-$\growth{f}(n) = \Theta(n^k)$. This parameter also coincides with the minimal
-$k$ such that $f$ is represented by a morphism $\mu \colon M^k \to \Rel$.
-
-\begin{fact}[Folklore about regular languages]
-    \label{regular:fact}
-    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular, whenever $a,b \in \Sigma$.
-    Regular languages are closed under intersection.
-\end{fact}
-
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by a $\Nat$-rational series
-    is a regular language. In particular,
-    for all $\Nat$-rational series $S$, $S^{-1}(\set{0})$ is a regular
-    language.
-\end{fact}
+\ZPoly$ such that $\growth{f}(n) = \bigO(n^k)$, i.e., have \kl{growth rate}
+$k$. Similarly, we let $\NPoly[k] = \NPoly \cap \ZPoly[k]$.
 
@@
 
+\paragraph{Commutative Polyregular Functions.}
 \AP 
-\paragraph{Commutative Polyregular Functions.}
 Because this paper focuses on functions that are commutative both in the
@@
 
+Let us briefly state that one can decide whether a \kl{$\Rel$-polyregular
+function} $f$ is \kl{commutative}, ensuring that we are working in a decidable
+subclass of functions.
+
+\begin{lemma}
+    \label{decide-commutative:lemma}
+    Let $f \in \ZPoly$. One can decide if 
+    $f$
+    is \kl{commutative}.
+\end{lemma}
+\begin{proof}
+    Let $\Sigma$ be a finite alphabet, endowed with a
+    total ordering over its letters,
+    and let $f \colon \Sigma^* \to \Rel$ be a
+    \kl{$\Rel$-polyregular function}.
+    The map $\polysort \colon \Sigma^* \to \Sigma^*$
+    that sorts the letters in the input word $w$ according
+    to the chosen ordering is a \kl{polyregular function}
+    \cite[Proposition II.12]{LOPEZ23b}.
+    As a consequence,
+    $(f \circ \polysort) \in \ZPoly$.
+    Finally, $f$ is \kl{commutative} if and only
+    if $f = f \circ \polysort$, which is decidable
+    in the case of $\ZPoly$
+    \cite[Corollary II.24]{LOPEZ23b}, and more generally in the case 
+    of $\ZRat$ \cite[Corollary 3.6 p 38]{berstel2011noncommutative}.
+\end{proof}
+
+For the sake of completeness, let us also state that \kl{commutativity} is
+decidable for \kl{$\Rel$-rational series}, although using more involved
+arguments, since the latter are not closed under pre-composition by a
+\kl{polyregular function}.
+
+\begin{lemma}
+    \label{decidable-commutative-rat:lemma}
+    Let $f \in \ZRat$. One can decide if
+    $f$ is \kl{commutative}.
+\end{lemma}
+
+Let us briefly remark that the characterizations of subclasses of
+\kl{polyregular functions} that are obtained in practice can be obtained by
+pre-composition with a \kl{commutative} \kl{polyregular function}. While the
+two following
+\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma} do not
+provide simpler proofs, they suggest the central position of \kl{commutative}
+\kl{star-free polyregular functions} as \emph{test functions} for various
+properties (growth rate, aperiodicity).
+
+\begin{lemma}
+    \label{pre-compose-growth-commut:lemma}
+    Let $f \in \ZRat$, and $k \in \Nat$,
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \ZPoly[k]$,
+        \item For every \kl{commutative} \kl{star-free polyregular function} $h$
+            of \kl{growth rate} $l \in \Nat$,
+            $(f \circ h) \in \ZPoly[k+l-1]$.
+    \end{enumerate}
+\end{lemma}
+
+\begin{lemma}
+    \label{pre-compose-sf-commut:lemma}
+    Let $f \in \ZPoly$.
+    The following are equivalent:
+    \begin{enumerate}
+        \item $f \in \ZSF$,
+        \item For every \kl{commutative} \kl{star-free polyregular function} $h$,
+            $(f \circ h) \in \ZSF$.
+    \end{enumerate}
+\end{lemma}
+
 \AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function
@@
 
-\AP \paragraph*{Polynomials.} All polynomials considered in this paper have
+\paragraph*{Polynomials.} \AP All polynomials considered in this paper have
 coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[X_1, \dots, X_n]$ is a \intro{$\Rel$-rational polynomial} if it
-is \kl{represented} by a \kl{$\Rel$-polyregular function}. Similarly, it is a
-\intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
-\kl{$\Nat$-polyregular function}.
+\Rel[X_1, \dots, X_n]$ is a \intro{$\Nat$-rational polynomial} if it is
+\kl{represented} by a \kl{$\Nat$-polyregular function}. The analogue notion of
+\intro{$\Rel$-rational polynomials} is not of particular interest, since every
+polynomial $P$ is \kl{represented} by a \kl{$\Rel$-polyregular function}.
 
@@
     \label{negative-not-nrat:ex}
-    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomial},
+    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomials},
     but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
@@
 \begin{proof}
-    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function},
-    hence, $P(X) \defined X$ is
+    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
+    Thus, 
+    $P(X) \defined X$ is
     a \kl{$\Nat$-rational polynomial}. Similarly,
@@
     a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
-    have non-negative output.
+    are non-negative.
 \end{proof}
@@
 
-We saw in \cref{negative-not-nrat:ex} a simple criterion to check whether a
-polynomial is a \kl{$\Nat$-rational polynomial}, but being \kl{non-negative} is
-not enough, as the following example illustrates. Note that the proof scheme of
-this example will be at the core of \cref{thm:counter-example}.
+All \kl{$\Nat$-rational polynomials} are \kl{non-negative}, but the converse does
+not hold,
+as the following example illustrates. Note that the proof scheme of
+this example will be at the core of \cref{thm:counter-example}, and leverages
+a key property of $\NRat$ recalled hereafter.
+
+\begin{fact}
+    \label{pre-image-regular:fact}
+    The pre-image of a regular language by a \kl{$\Nat$-rational series}
+    is a regular language. 
+\end{fact}
 
 \begin{example}
-    The polynomial $P(X, Y) \defined (X - Y)^2$
-    is a \kl{$\Rel$-rational polynomial} that is \kl{non-negative},
-    but is
+    Let $P(X, Y) \defined (X - Y)^2$.
+    Then $P$ is
+    is \kl{non-negative}, but is
     not a \kl{$\Nat$-rational polynomial}.
@@
     Assume by contradiction that
-    $S$ is a $\Nat$-rational series computing $P$.
-    Then, $S^{-1}(\set{0})$ should be a regular language
+    $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
+    Then, $f^{-1}(\set{0})$ is a regular language
     (\cref{pre-image-regular:fact}),
     but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not a regular language (\cref{regular:fact}).
+    is not.
 \end{proof}
@@
 $\Nat$, or equivalently, if it is \kl{non-negative} as a polynomial. In the
-polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative monomials} are $X^2$
-and $Y^2$.
+polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials}
+are $X^2$ and $Y^2$.

### 2024-01-25 parts/transducers.tex
writing to publish

@@
 \end{fact}
-\begin{proof}
-    Let us write $Q_i$ and $O_i$ for the value of the variables
-    $Q$ and $O$ at step $i$ of the \texttt{while loop}.
-    We prove the desired property by induction on $i$.
-
-    For $i=0$, the property is true because
-    $Q_0 = \set{\varepsilon}$ and $O_0 = \setof{a}{a \in \Sigma}$.
-
-    For $i+1$. Either the \texttt{if} branch was taken, in which case $Q_{i+1}
-    \cup O_{i+1} = (Q_i \cup O_i) \setminus \set{u}$ for some $u \in O_i$. This
-    set remains \kl{downwards closed}, and elements in $O_{i+1}$ remain maximal
-    elements.
-
-    If the \texttt{else} branch was taken, then there exists $u \in O_i$ such
-    that $Q_{i+1} = Q_i \cup \set{u}$ and $O_{i+1} = O_i \setminus \set{ u }
-    \cup \setof{ ua }{ a \in \Sigma}$. We conclude that $Q_{i+1} \cup O_{i+1} =
-    Q_i \cup O_i \cup \setof{ ua }{a \in \Sigma}$ continues to be \kl{downwards
-    closed} for $\prefleq$. Let $v \in Q_{i+1} \cup O_{i+1}$ be such that $ua
-    \prefleq v$ for some $a \in \Sigma$. Then $u \prefleq v$, and $u = v$ since
-    $u$ was a maximal element. As a consequence, $ua$ is a maximal element for
-    all $a \in \Sigma$. Assume by contradiction that $ua$ is comparable with
-    some $v \in O_{i+1}$ with $ua \neq v$, it cannot be that $ua \prefleq v$ by
-    the above argument, and if $v \prefleq ua$ with $v \neq ua$, then $v
-    \prefleq u$ and $u = v$, which is absurd since $v \not \in O{i+1}$.
-    We have concluded that $O_{i+1}$ continues to have pairwise incomparable
-    elements.
-\end{proof}
 
@@
 \end{lemma}
-\begin{proof}
-    Assume towards a contradiction that
-    \cref{residual:algo} does not terminate.
-    Then, the \texttt{else} branch in the \texttt{while loop}
-    must be taken infinitely often.
-    This means that the set $Q$ of states grows arbitrarily large.
-
-    Let us write $\seqof{Q_i}{i \in \Nat}$ for the set of states $Q$ at step
-    $i$ of the execution of \cref{residual:algo}. Applying
-    \cref{q-o-prefix-cool:fact}, we know that for all $i \in \Nat$, $Q_i$ is
-    \kl{downwards closed} for $\prefleq$. Let us write $Q_\infty \defined
-    \bigcup_{i \in \Nat} Q_i$. The set $Q_\infty$ is infinite, and is
-    \kl{downwards closed} for $\prefleq$. As a consequence, it is an infinite
-    tree with a finite branching (at most $\card{\Sigma}$), and has an infinite
-    branch $\seqof{u_j}{j \in \Nat}$ thanks to König’s lemma.
-
-    Let us prove that this infinite branch is a \kl{bad sequence} for the
-    ordering $\resleq{f}{k}$.
-    Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
-    know that $u_j \in Q_j$ and $u_p \in Q_p$. Then, at step $p-1$ of the
-    algorithm, $u_j \in Q_{p-1}$, since $u_j \in Q_j \subseteq Q_{p-1}$.
-    Because $u_j \prefleq u_p$ and $u_j \resleq{f}{k} u_p$,
-    \cref{residual:algo} must take the \texttt{if} branch at step $p-1$. As a
-    consequence, $u_p \not\in Q_{p}$, which is absurd.
-
-    We have proven that the infinite branch is a \kl{bad sequence}
-    for $\resleq{f}{k}$, which contradicts the assumption.
-    Hence, \cref{residual:algo} must terminate.
-\end{proof}
 
@@
 \end{lemma}
-\begin{proof}
-    Because $f \in \NPoly[k]$, there exists
-    a tuple $\vec{x}$ of first order free variables,
-    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
-    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
-    such that
-    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.
-
-    Let $q$ be the maximal quantifier rank of formulas $\seqof{\psi_i}{1 \leq i
-    \leq n}$. To a word $u \in \Sigma^*$, we associate the vector $\MSO^q(u)$
-    that maps an $\MSO$-type with $\ell \leq |\vec{x}|$ free variables to the
-    number of realizations of this type in $u$.
-
-    Let $u, v \in \Sigma^*$ such that $\MSO^q(u) \leq \MSO^q(v)$, which means
-    that every $\MSO$ type (of quantifier rank $q$ and with at most $n$ free
-    variables) has at least as many realizations in $v$ than it has in $u$.
-    Remark that by the compositionality of $\MSO$ over words (for instance, see
-    the Feferman-Vaught theorem \cite{FEVAU59,MAKOW04}), for all $\MSO^q$ types
-    $t(\vec{x})$, there are finitely many $\MSO^q$ types $t_l^j(\vec{y_i}),
-    t_r^j(\vec{z_i})$ with $\vec{x} = \vec{y_i} \uplus \vec{z_i}$
-    for $1 \leq j \leq N_0$, such that for every
-    tuple $\vec{a}$ of elements in a word $uv$, $\MSO^q(\vec{a} / uv) =
-    t(\vec{x})$ if and only if there exists $1 \leq j \leq N_0$,
-    such that $\vec{a} = \vec{b} \uplus \vec{c}$,
-    $\MSO^q(\vec{b} / u) =
-    t_l^j(\vec{y_i})$, and $\MSO^q(\vec{c} / v) = t_r^j(\vec{z_i})$.
-    We write $t = t_l \odot t_r$ to signify
-    that $\MSO^q(\vec{bc} / uv) = t$
-    if and only if $\MSO^q(\vec{b}/u) = t_l$
-    and $\MSO^q(\vec{c}/v) = t_r$.
-
-    As a consequence, if $\MSO^q(u) \leq \MSO^q(v)$, then 
-    for all $w \in \Sigma^*$: 
-    \begin{align*}
-        & f(vw) - f(uw) \\
-        &= 
-        \sum_{i = 1}^n m_i
-        \left[
-            \vcount{\phi_i(\vec{x})} (vw) -
-            \vcount{\phi_i(\vec{x})} (uw)
-        \right] \\
-        &= 
-        \sum_{i = 1}^n
-        m_i
-            \sum_{\phi_i \in t(\vec{x})}
-        \left[
-            \vcount{t(\vec{x})}(vw)
-            -
-            \vcount{t(\vec{x})}(uw)
-        \right] \\
-        &= 
-        \sum_{i = 1}^n
-        m_i
-        \sum_{1 \leq j \leq N_0}
-        \sum_{\phi_i \in t_l^j(\vec{y}) \odot t_r^j(\vec{z})}
-        \\
-        &\quad\quad
-        \underbrace{
-        \left[
-            \vcount{t_r^j(\vec{y})}(v)
-            -
-            \vcount{t_r^j(\vec{y})}(u)
-        \right] 
-    }_{ \in \Nat }
-            \times 
-            \vcount{t_l^j(\vec{z})}(w)
-    \end{align*}
-
-    We have proven that if $\MSO^q(u) \leq \MSO^q(v)$, then $u \resleq{f}{k}
-    v$. Recall that $\Nat^p$ is a \kl{well-quasi-ordering} when endowed with
-    the product ordering, and therefore that $\setof{\MSO^q(u)}{u \in
-    \Sigma^*}$ is a \kl{well-quasi-ordering}.
-
-    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
-    Without loss of generality, one can assume that for all $i \neq j$, $u_i
-    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
-    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
-    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
-    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
-    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
-    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
-\end{proof}
 
@@
 
-\begin{conjecture}
-    Let $f \in \NPoly$. The following are equivalent,
-    \begin{enumerate}
-        \item $f \in \ZSF$,
-        \item the \kl{$k$-residual transducer} of $f$
-            is \kl{counter-free},
-            and its labels belong to $\ZSF$.
-    \end{enumerate}
-\end{conjecture}
+\begin{lemma}
+    Let $\NSF[0] = \NPoly \cap \NSF[0]$,
+    and elements of $\NSF[0]$
+    have a \kl{counter-free} \kl{$0$-residual transducer}.
+\end{lemma}
 \begin{proof}
-    If it has a counter free residual transducer recursively, then
-    it clearly belongs to $\ZSF$.
+    TODO.
+\end{proof}
 
-    % TODO: mix orderings and trees and maximal elements :(
-
-    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
-    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
-    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
-    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
-    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
-    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
-    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
-    w_2) = t$.
-
-    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
-    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
-    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
-    strict prefix of $s_0$. It must exist, because otherwise the two are always
-    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
-    this contradicts the minimality of $u$.
-
-    We have proven that $s_0$ and $s_1$ are comparable, hence they
-    are equal.
-    Finally, we have proven that $\delta(q, w_1) = s_0$,
-    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
-
-    As a consequence $t$ and $q$ were comparable for the prefix relation,
-    hence equal, and therefore $\delta(q, w) = q$.
-\end{proof}
+\textbf{TODO:} one may think that we can ask that $u \resleq{f}{k} u w^n$
+implies $u \resleq{f}{k} uw$, but it is false, as the following example will
+demonstrate. This is why the definition of counter is more involved in our
+case.
 
@@
     hence there is a non-trivial counter.
-    However,
-    for all $n > 1$ such that
-    $\app{f}{a^n} - f$ belongs to $\NPoly$,
-    $\app{f}{a^n} - \app{f}{a^{n-1}}$ belongs to $\NPoly$ too.
 \end{example}
@@
 \begin{lemma}
+    \label{sf-no-periods-on-sequences:lemma}
     Let $f \colon \Sigma^* \to \Nat$ be a \kl{star-free $\Nat$-polyregular
-    function}. Then, for all $u \in \Sigma^*$, for all $n > 1$, for all $w \in
-    \Sigma^+$, if $\app{f}{u w^n} - \app{f}{u}$ belongs to $\NPoly[k-1]$, then
-    $\app{f}{uw^n} - \app{f}{u w^{n-1}}$ belongs to $\NPoly[k-1]$.
+    function}. Then,
+    for all $u, w \in \Sigma^*$,
+    there exists $n \in \Nat$ such that
+    $u w^n \resleq{f}{k} u w^{n+1}$.
 \end{lemma}
-\begin{proof}
-    By induction on $n$, 
+\begin{proof}[Proof sketch.]
+    Let $t$ be a first order type,
+    the map $g_t \colon X \mapsto \vcount{t}(uw^X)$ is a
+    \kl{commutative} \kl{star-free $\Nat$-polyregular function},
+    hence ultimately a polynomial, and therefore,
+    translated enough, it is a $\Nat[X]$ polynomial.
+    But this means that $g_t(X+1) - g_t(X) \in \Nat[X]$ for a large
+    enough $X$, which is precisely what suffices.
+    Now, we can do this for every first order type $t$
+    with fewer variables and lower quantifier rank than those
+    needed to define $f$, and conclude.
 \end{proof}
@@
 a function is in $\NPoly$.
+
+\section{Trash polynomials}
 
@@
 \end{theorem}
+
+\begin{fact}[Folklore about regular languages]
+    \label{regular:fact}
+    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
+    is not regular, whenever $a,b \in \Sigma$.
+    Regular languages are closed under intersection.
+\end{fact}

### 2024-01-25 parts/transducers.tex
adding conjectures

@@
 
+\begin{lemma}
+    \label{trivial-for-unary:lemma}
+    Let $f \in \NPoly$ having unary input.
+    In particular, $f$ is commutative.
+    Then, $f \in \NSF$ if and only if 
+    the \kl{residual transducer} of $f$
+    is \kl{counter-free}.
+\end{lemma}
+\begin{proof}
+    If there is a counter... well we will find it ^^
+\end{proof}
+
+\begin{conjecture}
+    If $f$ is commutative, and the \kl{residual transducer}
+    has a counter, then it has a single letter counter.
+\end{conjecture}
+
+\begin{conjecture}
+    If $f$ is commutative, then 
+    $f \in \NSF$ if and only if the \kl{residual transducer}
+    of $f$ is \kl{counter-free}.
+\end{conjecture}
+
 \section{Trash polynomials}

### 2024-01-26 parts/appendix.tex
preparing the witdrawal

@@
 
-    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then
-    $\pi(\mu(v_1), \dots, \mu(v_k)) = \pi(\mu(w_1), \dots, \mu(w_k))$ because
-    $\mu(u_i^\omega) = \mu(u_i)^\omega$ is an idempotent element of the monoid
-    $M$. In particular, $f(w)$ can be computed as the sum over equivalence
-    classes for $\equiv$, of the value of $\pi$ on one representative,
-    which belongs to $\Nat$.
+    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then their
+    production are the same, i.e., $\pi(\mu(v_1), \dots, \mu(v_k)) =
+    \pi(\mu(w_1), \dots, \mu(w_k))$. This holds because $\mu(u_i^\omega) =
+    \mu(u_i)^\omega$ is an idempotent element of the monoid $M$. In particular,
+    $f(w)$ can be computed as the sum over equivalence classes for $\equiv$, of
+    the value of $\pi$ on one representative, which belongs to $\Nat$.
 
@@
     single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
-    with \kl{positive coefficients} for its \kl{maximal monomials}.
+    with \kl{non-negative} \kl{maximal monomials}.
 
@@
     of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
-    which \kl{strictly divide} $M$.
+    which \kl{strictly divides} $M$.
     Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
@@
     $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
-\end{proof}
-
-\begin{proof}[Proof of \cref{discrete-deriv-linear:fact}]
-    TODO or to skip.
-\end{proof}
-
-\begin{proof}[Proof of \cref{maxi-monomials-submodular:fact}]
-\end{proof}
-
-\begin{proof}[Proof of \cref{derivation-monomials:fact}]
 \end{proof}
@@
 \end{proof}
+
 \begin{proof}[Proof of \cref{derivation-stabilises-correct:lem}]
@@
     hence for $K = 0$, $Q \in \CorrectPoly$.
-    Otherwise, let $\nu \colon \vec{X} \topartial \Nat$,
-    and let us analyze the \kl{maximal monomials} $T$
-    of $\restr{Q}{\nu}$.
-    Because $T$ is a \kl{monomial} of $\restr{Q}{\nu}$,
-    it \kl{divides} some \kl{monomial} in $Q$, and
-    using \cref{derivation-monomials:fact},
-    it \kl{divides} some \kl{monomial} $M$ in 
-    $\Diff{K}{P_1}$.
-    Now, the monomial $T$ is computed as follows:
+
+    Otherwise, let $\alpha$ be the maximal absolute value 
+    of coefficients appearing in $P$,
+    and $D$ be the maximal number of divisors for
+    a monomial in $P$.
+    We define $K \defined \alpha \times D + 1$.
+
+    Let $\nu \colon \vec{X} \topartial \Nat$, we are going to prove that
+    $\restr{Q}{\nu}$ has \kl{non-negative} \kl{maximal monomials}. To that end,
+    let us consider $T$ such a maximal monomial, and let us write $T = \beta
+    T_1$ so that $T_1$ has coefficient $1$ and $\beta \in \Rel$. Because we
+    will need to normalize other monomials in this way, we define for the rest
+    of the proof the function $\mathsf{unit}$ that maps a monomial to the
+    monomial obtained by replacing its constant term by $1$ if the monomial is
+    non-zero, and leaving the monomial unchanged otherwise.
+
+    Let us write $\mathcal{D}_1$ to be the set of \kl{monomials} of
+    $\Diff{K}{P_1}$ that are \kl{divided} by $T$. Similarly, let us write
+    $\mathcal{D}_2$ for the set of monomials of $\translate{K}(P_2)$ that are
+    \kl{divided} by $T$. Thanks to \cref{derivation-monomials:fact}, we know
+    that every element of $\mathcal{D}_2$ divides some element in
+    $\mathcal{D}_1$. We write $\mathcal{D}_2^M$ for the set of monomials $S \in
+    \mathcal{D}_2$ such that $S$ \kl{divides} $M$.
+
+    Now, remark that the coefficient $\beta$ is obtained as follows:
     \begin{align*}
-        T &= 
-        \sum_{S \in \Monomials(Q), T \divides S}
-        \restr{S}{\nu} \\
-          &= 
-          \sum_{S \in \Monomials(\Diff{K}{P_1}), T \divides S}
-            \restr{S}{\nu}
-        + 
-          \sum_{S \in \Monomials(P_2), T \divides S}
-            \restr{S}{\nu}
+        \beta &= \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
+               + \sum_{M \in \mathcal{D}_2} \restr{M / T_1}{\nu} \\
+              &\geq \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
+              - \sum_{M \in \mathcal{D}_2} \left|\restr{M / T_1}{\nu}\right|
               \\
               &\geq
-          \sum_{S \in \Monomials(\Diff{K}{P_1}), T \divides S}
-            \restr{S}{\nu}
+              \sum_{M \in \mathcal{D}_1}
+              \left[
+              \restr{M / T_1}{\nu}
               -
-            \left|
-          \sum_{S \in \Monomials(P_2), T \divides S}
-            \restr{S}{\nu}
-            \right|
+              \sum_{S \in \mathcal{D}_2^M}
+              \left|\restr{S/T_1}{\nu}\right|
+              \right]
+              \\
+              &\geq
+              \sum_{M \in \mathcal{D}_1}
+              \restr{\mathsf{unit}(M)/T_1}{\nu}
+              \left[
+              K
+              - \alpha D
+              \right]
+              \\
+              &\geq 0 \quad . \qedhere
     \end{align*}
-
-\end{proof}
-
-\begin{proof}[Proof of
-    \cref{decidability-correct:lem}]
-    Let $P \in \Rel[\vec{X}]$. We prove by induction on
-    $\MaximalMonomials(P)$ that whether $P \in \CorrectPoly$
-    is decidable.
-
-    If $P$ is a constant polynomial, then
-    $P \in \CorrectPoly$ if and only if $P = n \in \Nat$
-    which is decidable.
-
-    Otherwise, $P$ is not a constant polynomial, and let us write $P = P_1 +
-    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. If $P_1
-    \not \in \Nat[\vec{X}]$, then $P \not \in \CorrectPoly$, and this is
-    decidable. Hence, we can assume that $P_1 \in \Nat[\vec{X}]$. Using
-    \cref{derivation-stabilises-correct:lem}, there exists a computable bound
-    $K$, such that if $P \in \CorrectPoly$, then t $Q \defined (\Diff{K}{P_1} +
-    P_2) \in \CorrectPoly$. Leveraging, \cref{derivation-simplifies:lemma}, we
-    also know that $\MaximalMonomials(Q) \hoarele \MaximalMonomials(P)$, and
-    therefore that we can decide whether $Q \in \CorrectPoly$. Notice that if
-    $Q \not \in \CorrectPoly$, then $P \not \in \CorrectPoly$, hence we can
-    assume that $Q \in \CorrectPoly$.
-
-    Similarly, for partial valuations
-    $\nu \colon \vec{X} \topartial \set{0, \dots, K}$
-    fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
-    \hoarele \MaximalMonomials{P}$. 
-    We can check for all these partial valuations that
-    $\restr{P}{\nu} \in \CorrectPoly$ by induction hypothesis,
-    and if one test is not satisfied, then $P \not \in \CorrectPoly$.
-
-    Let us now conclude that if all the above checks answered correctly,
-    then $P \in \CorrectPoly$. Let $\nu \colon \vec{X} \topartial \Nat$
-    be a partial valuation. Then, 
-    one of the three following cases arise:
-    \begin{itemize}
-        \item 
-            Either,
-        $\nu$ fixes \emph{some} indeterminate $X_i$ to a value $n_i < K$.
-        In that case, let us define $\delta$ as the partial function that coincides
-        with $\nu$ on its domain
-        $\dom(\delta) = \dom(\nu) \setminus \set{ X_i }$.
-        Then, $\restr{P}{\nu} = \restr{\restr{P}{X_i = n_i}}{\delta}$.
-        As we previously checked that $\restr{P}{X_i = n_i} \in \CorrectPoly$,
-        we conclude in particular that 
-        $\restr{P}{\nu}$ has \kl{non-negative} \kl{maximal monomials}.
-
-        \item Or, for all $X_i$ in the (potentially empty) domain of $\nu$, $\nu(X_i)
-            \geq K$. In that case, let us write $\delta \defined
-            \translate{-K}(\nu)$, that is the partial map $\delta \colon
-            \vec{X} \topartial \Nat$ having the same domain as $\nu$, and such
-            that for all $X_i \in \dom(\nu)$, $\delta(X_i) \defined \nu(X_i) -
-            K$. Recall that $Q = \Diff{K}{P_1} + P_2$, and therefore that
-            $\translate{K}(P) = P_1 + \translate{K}(Q)$. By linearity,
-            $\restr{\translate{K}(P)}{\delta} = \restr{P_1}{\delta} +
-            \restr{\translate{K}(Q)}{\delta}$, and we conclude that the
-            \kl{maximal monomials} of $\restr{\translate{K}(P)}{\delta}$ are
-            \kl{non-negative}.
-            Since $\restr{\translate{K}(P)}{\delta} = \translate{K}{\restr{P}{\nu}}$,
-            and since
-            $\translate{K}$ leaves \kl{maximal monomials} invariant,
-            we conclude that
-            the \kl{maximal monomials} of $\restr{P}{\nu}$
-            are the same as those of $\restr{\translate{K}(P)}{\delta}$,
-            and are therefore \kl{non-negative}.
-            \qedhere
-    \end{itemize}
 \end{proof}
@@
 
-    To prove \cref{d-t-transl:item} $\implies$ \cref{d-t-correct:item}, let us
-    remark that if $P$ is \kl{non-negative} and $\translate{K}(P) \in
-    \Nat[\vec{X}]$, then $\translate{K}(P) \in \CorrectPoly$ thanks to
-    \cref{n-poly-n-poly:example}. Then, for every partial valuation $\nu \colon
-    \vec{X} \topartial \Nat$, $\translate{K}(\restr{P}{\nu}) \in
-    \Nat[\vec{X}]$, and in particular conclude that \kl{maximal monomials} of
+    We will first prove the equivalence between \cref{d-t-transl:item} and
+    \cref{d-t-correct:item} without keeping track of the constant $K$, and then
+    notice that this $K$ is actually computable. This will simplify the reading
+    process.
+
+    We will first prove that \cref{d-t-transl:item} $\implies$
+    \cref{d-t-correct:item}, for any choice of $K \in \Nat$. Indeed, let us
+    consider some partial valuation $\nu \colon \vec{X} \topartial \Nat$. By
+    assumption, $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$, and in
+    particular conclude that \kl{maximal monomials} of
     $\translate{K}(\restr{P}{\nu})$ are exactly the \kl{maximal monomials} of
-    $\restr{P}{\nu}$ thanks to \cref{translation-maximal:fact}, and are therefore
-    \kl{non-negative}.
+    $\restr{P}{\nu}$ thanks to \cref{translation-maximal:fact}, and are
+    therefore \kl{non-negative}.
 
@@
     the latter is closed under partial applications.
+
+    Now, remark that in the proof, we have computed $K$ recursively from $P$,
+    and only applied partial functions using values below $K$. This not only
+    shows that $K$ is computable from $P$, but also that
+    \cref{d-t-transl-fin:item} and \cref{d-t-transl:item} are equivalent.
 \end{proof}

### 2024-01-26 parts/correction.tex
preparing the witdrawal

@@
     Let $f$ be a  \kl{$\Nat$-polyregular} function. 
-    There exists $\omega \in \Nat$
+    There exists a computable $\omega \in \Nat$
     such that for all $p \in \Nat$,
@@
 The following \cref{derivation-stabilises-correct:lem} is the main
-combinatorial argument of this section. Note that we must be careful
-about it. \textbf{TODO: write a nice thing here!}.
+combinatorial argument of this section. It leverages the positivity of the
+\kl{maximal monomials} to compute a shift large enough to make lower degree
+coefficients positive. This lemma will be lifted to a given polynomial $P$ by
+noticing that if $K \in \Nat$, and $P = P_1 + P_2$, then $\translate{K}(P) =
+P_1 + \Diff{K}{P_1} + \translate{K}(P_2)$.
 
@@
     monomials} of $P$. We compute a bound $K$ such that $Q \defined
-    (\Diff{K}{P_1} + P_2) \in \CorrectPoly$ using
+    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$ using
     \cref{derivation-stabilises-correct:lem}. Thanks to
@@
     \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
-    which is effectively computable.
+    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} which is
+    effectively computable.
 
     Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
-    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
-    (\cref{n-poly-n-poly:example}).
-    Furthermore, $\translate{K}(Q)$ is also (effectively)
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
-    (see \cref{translation-invariance:fact}).
-    As a consequence, $\translate{K}(P)$ is (effectively) \kl{represented} 
-    by a function $f_\Delta$,
-    since it is obtained as the following sum:
-    \begin{equation*}
-        \translate{K}(P) 
-        = P_1 + \translate{K}(Q) \quad .
-    \end{equation*}
+    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular
+    function} (\cref{n-poly-n-poly:example}). As a consequence,
+    $\translate{K}(P) = P_1 + Q$ is (effectively) \kl{represented} by a
+    function $f_\Delta$.
 
-    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$
-    fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
+    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
+    K}$ fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
     \hoarele \MaximalMonomials{P}$. Because $\restr{P}{\nu} \in \CorrectPoly$,
@@
 explicitly state that the membership is decidable to keep the proof clearer. A
-similar proof scheme can be followed to conclude that membership is decidable.
+similar proof scheme can be followed to conclude that membership is decidable,
+and even show that elements in $\CorrectPoly$ are, up to suitable translations,
+polynomials in $\Nat[\vec{X}]$. Beware that partial applications are still
+needed in this characterization, as \cref{bad-poly-translate:ex} illustrates.
 
 \begin{lemma}
+    \label{derivation-translation:lem}
+    Let $P \in \Rel[\vec{X}]$, 
+    there exists a computable $K \in \Nat$
+    such that the following are equivalent:
+    \begin{enumerate}
+        \item \label{d-t-correct:item} $P \in \CorrectPoly$,
+        \item \label{d-t-transl:item}
+            for 
+            all $\nu \colon \vec{X} \topartial \Nat$,
+            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$,
+        \item \label{d-t-transl-fin:item}
+            for all
+            $\nu \colon \vec{X} \topartial \set{0, \dots, K}$,
+            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
+    \end{enumerate}
+\end{lemma}
+
+\begin{corollary}
     \label{decidability-correct:lem}
-    It is decidable whether a polynomial 
-    $P \in \Rel[\vec{X}]$
-    belongs to $\CorrectPoly$.
-\end{lemma}
+    One can decide membership in $\CorrectPoly$.
+\end{corollary}
+\begin{proof}
+    Given $P \in \Rel[\vec{X}]$,
+    one applies \cref{derivation-translation:lem}
+    to obtain a bound $K$, and then 
+    computes $\translate{K}(\restr{P}{\nu})$ for all 
+    partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
+    Then $P \in \CorrectPoly$ if and only if all those polynomials
+    belong to $\Nat[\vec{X}]$, which is decidable.
+\end{proof}
+
+\begin{example}
+    \label{bad-poly-translate:ex}
+    The polynomial $\BadPoly$ is not a 
+    \kl{$\Nat$-rational polynomial},
+    but is \kl{non-negative} and satisfies
+    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
+\end{example}
 
@@
 \end{proof}
-
-Let us also state that polynomials in $\CorrectPoly$ are, up to suitable
-translations, polynomials in $\Nat[\vec{X}]$. Beware that partial applications
-are still needed in this characterization, as \cref{bad-poly-translate:ex}
-illustrates.
-
-\begin{lemma}
-    \label{derivation-translation:lem}
-    Let $P \in \Rel[\vec{X}]$, the following are equivalent:
-    \begin{enumerate}
-        \item \label{d-t-correct:item} $P \in \CorrectPoly$,
-        \item \label{d-t-transl:item}
-            There exists $K \in \Nat$,
-            such that for every partial
-            function $\nu \colon \vec{X} \topartial \Nat$,
-            $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
-    \end{enumerate}
-\end{lemma}
-
-\begin{example}
-    \label{bad-poly-translate:ex}
-    The polynomial $\BadPoly$ is not a 
-    \kl{$\Nat$-rational polynomial},
-    but is \kl{non-negative} and satisfies
-    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
-\end{example}

### 2024-01-26 parts/prelim.tex
preparing the witdrawal

@@
             indicator functions $\ind{L}$ of regular languages;
-        \item $f$ belongs to
+        \item \label{npoly-counting-mso:item} $f$ belongs to
             $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
@@
 
+\textbf{This is wrong, we should talk about the number of variables,
+    and say that we do not know in the case of $\NSF$ if it corresponds
+    to the minimal number of variables needed!}
 \AP Inside $\ZPoly$, there is an infinite hierarchy of classes, defined using

### 2024-01-26 parts/starfree.tex
writing to publish
preparing the witdrawal

@@
 \section{Beyond Polynomials}
-\label{sec:deciding}
+\label{beyond-polynomials:sec}
 
-Let us now move on from polynomials to general \kl{$\Rel$-polyregular
-functions}. As we will prove, \kl{commutativity} reduces $\ZPoly$ to a
-combination of a finite number of polynomials, arranged using modulo counting.
-We will first decide $\NPoly$ inside $\ZPoly$, and then leverage this to decide
-$\NSF$ inside $\NPoly$, incidentally proving that $\ZSF \cap \NPoly = \NSF$.
+Let us now go beyond polynomials and consider the case of \kl{commutative}
+\kl{$\Rel$-polyregular functions}. The hypothesis of \kl{commutativity} reduces
+$\ZPoly$ to a combination of a finite number of polynomials, arranged by
+counting modulo some numbers. This characterization will be proven in
+\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
+procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
+\cref{decompose-polynomial:lem} is actually a refinement of
+\cref{n-poly-combinatorics:lem} under the extra assumption of
+\kl{commutativity}.
+
+\AP To formally state the lemma, we will need some notations that will
+formalize the intuitive idea of counting tuples of integers modulo some
+constant. Let $\omega \in \Nat$, we write the set of \intro{types modulo
+$\omega$} $\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
+    \dots, \omega - 1}$. To a number $n \in \Nat$, we associate its
+    \intro{$\omega$-type} written $\moduloType[\omega](n)$ which is defined as the pair
+    $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest of the Euclidian
+    division of $n$ by $\omega$. This notion of type is lifted to vectors in
+    $\Nat^p$ for any $p \in \Nat$ by pointwise application.
+
+\begin{lemma}
+    \label{decompose-polynomial:lem}
+    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
+    \kl{$\Rel$-polyregular function},
+    where $\Sigma = \set{a_1, \dots, a_n}$.
+    There exists a computable
+    $\omega \in \Nat$,
+    and computable 
+    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
+    such that for all $w \in \Sigma^*$,
+    \begin{equation*}
+        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
+        = P_{\moduloType[\omega](X_1, \dots, X_n)}
+        \left(
+            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
+        \right)
+        \quad .
+    \end{equation*}
+\end{lemma}
+\begin{proof}
+    We apply
+    \cref{n-poly-combinatorics:lem} to compute the $\omega$.
+    Using this $\omega$,
+    for any choice of
+    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
+    there exists a polynomial $P_r \in \Rel[X_1, \dots, X_p]$
+    such that $X_1, \dots, X_p \geq \omega$:
+    \begin{equation*}
+        f\left(
+        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
+        = P_r(X_1, \dots, X_p) \quad .
+    \end{equation*}
+    To get the full family of polynomials needed requires
+    just to only iterate subsets of letters, and fix the other ones
+    to be below the threshold $\omega^2$.
+\end{proof}
 
@@
 The goal of this section is to go beyond polynomials and characterize all
-$\Rel$-polyregular functions that are $\Nat$-polyregular, under the assumption
-that they are \kl{commutative}.
+\kl{$\Rel$-polyregular functions} that are \kl{$\Nat$-polyregular}, under the
+assumption that they are \kl{commutative}. We will provide a semantic property
+in the form of \cref{k-combinatorial:def}, that is conjecture to characterize
+$\NPoly$ inside $\ZPoly$, even in the non-commutative setting.
 
@@
     for all $u_1, \dots, u_k \in \Sigma^*$,
-    there exists a \kl{$\Nat$-rational polynomial} $P$
+    there exists a polynomial $P$
+    such that $\translate{K}(P) \in \CorrectPoly$, and
     satisfying for all $X_1, \dots, X_k \geq \omega$:
@@
     \end{equation*}
+    We say that $f$ is \reintro{combinatorial}
+    if it is \reintro{$(k,\Nat)$-combinatorial} for all $k \in \Nat$.
 \end{definition}
@@
     \begin{enumerate}
-        \item $f$ is \kl{$(k,\Nat)$-combinatorial},
-        \item $f \in \NPoly[k]$,
+        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
+        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
     \end{enumerate}
@@
 \end{theorem}
+\begin{proof}
+    \textbf{TODO: we need to talk about the number of variables hehe}
+    Let us prove that
+    \cref{f-npoly-combi:item}
+    implies \cref{f-combinatorial:item}. 
+    We compute the $\omega$ by applying 
+    \cref{n-poly-combinatorics:lem}, and then
+    notice that the polynomial $P$ obtained is
+    by definition a \kl{$\Nat$-rational polynomial}.
 
-To prove the theorem, let us first provide a nice characterization of
-\kl{commutative} \kl{$\Rel$-polyregular functions} in terms of polynomials.
-This is actually a refinement of \cref{n-poly-combinatorics:lem} in the
-commutative case.
-
-\AP Given $\omega \in \Nat$, let us write $\phi \colon \Nat \to \set{0, \dots,
-\omega} \times \set{0, \dots, \omega - 1}$ defined as $\phi(n) = (t, r)$ where
-$t = \min(n, \omega)$, and $r$ is the rest of the Euclidian division of $n$ by
-$\omega$. We let $\phi^\dagger \colon \Nat^k \to$.
-
-\begin{lemma}
-    \label{decompose-polynomial:lem}
-    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
-    \kl{$\Rel$-polyregular function}. There exists a computable
-    $\omega \in \Nat$,
-    and computable 
-    polynomials $\seqof{P_{\rho}}{\rho \in \Delta_\omega}$,
-    such that for all $w \in \Sigma^*$,
-    \begin{equation*}
-        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) = P_{\phi(X_1, \dots, X_n)} (X_1 / \omega, \dots, X_n / \omega)
-        \quad .
-    \end{equation*}
-\end{lemma}
-\begin{proof}
-    equivalent partitions of the word, but commutativity?.q
-    Lemma 5.37 phd gaëtan.
-\end{proof}
-
-As a consequence, we conclude 
-\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
-    We use \cref{decompose-polynomial:lem}
-    to effectively compute
-    a polynomial representation of $f$. Remark that
-    $f$ is a \kl{$\Nat$-polyregular function} if and only if
-    all these polynomials are \kl{$\Nat$-rational polynomials},
-    which is decidable thanks to
-    \cref{decide-correct-poly:lem,corrected-version:thm}.
+    For the converse implication, we use \cref{decompose-polynomial:lem} to
+    effectively compute a polynomial representation of $f$. Using the
+    assumption, all of these polynomials are actually \kl{$\Nat$-rational
+    polynomials}. Let us now construct a new way to compute $f$, by first
+    computing the \kl{$\omega$-type} of the input (which is possible in
+    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
+    which is possible in $\NPoly$.
 \end{proof}
@@
 
-Note that this proves in the commutative case a conjecture that was 
-formulated by Thesis Gaetan, and that is non trivial, because
-
-\begin{conjecture}
-    \label{nsf-zsf:conj}
-    In the general case, the following equation holds for all $k \in \Nat$:
-    \begin{equation*}
-        \ZSF[k] \cap \NPoly
-        = 
-        \NSF[k]
-        \quad .
-    \end{equation*}
-\end{conjecture}
-
 As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
 procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
-decidable inside $\NPoly$ under the assumption of commutativity.
+decidable inside $\NPoly$ under the assumption of commutativity,
+and that the effective procedure can be used
 
-\textbf{TODO: add the growth rate here, we are actually minimising it}.
 \begin{corollary}
@@
     A language $L$ is aperiodic if and only if 
-    the indicator function is ultimately polynomial.
+    $\ind{L}$ is \kl{ultimately polynomial}.
 \end{example}
-\begin{proof}
-    TODO.
-\end{proof}
 
@@
     and a family of polynomials 
-    $\seqof{P_{\seqof{n_a}{a \in \Sigma}}}{0 \leq n_a < \omega}$
-    such that.
-    Because each of those polynomials coincide with $P$ over an infinite
-    subgrid of $\Nat$,
-    we conclude that all polynomials are in fact equal to a single one.
-    we conclude that the polynomials 
-    $P_{\seqof{n_a}{a \in \Sigma}}$ are all equal.
+    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^\Sigma}$
+    that represent $f$.
+    For all $t \in \ModuloTypes[\omega]^\Sigma$
+    that have threshold greater than $\omega^2$,
+    we conclude that $P_t$ and $P$ coincide (up to translation) over an infinite grid,
+    hence we conclude that
+    if all letters appear at least $N_1 \defined \max (\omega^2, N_0)$ times,
+    then $f$ computes $P$. But this also shows that
+    $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
+    \kl{represented} by a function $f_1 \in \NSF[k]$.
 
-    Now, we conclude in particular that the function 
-    computing $P$ is \kl{$\Nat$-polyregular}, because $f$ is,
-    hence is computable by a \kl{star-free $\Nat$-polyregular function},
-    because of \cref{nsf-polynomials:thm}.
-    When fixing values of some indeterminate below $N_0$,
-    we obtain a new function $g$, that remains commutative,
-    remains ultimately polynomial, and remains \kl{$\Nat$-polyregular}.
-    By induction hypothesis, this function $g$ is also 
-    \kl{star-free $\Nat$-polyregular}, and we conclude
-    that 
-    $f = P + \sum g$ which remains star free.
+    To compute $f$ on words where some letter appears fewer than
+    $N_1$ times, we partially apply $f$ to obtain a function $g$
+    over a restricted input alphabet. Using the induction hypothesis,
+    $g$ belongs to $\NSF[k]$, and we can therefore (effectively) compute
+    a representation of $f$ in $\NSF[k]$.
 \end{proof}

### 2024-01-26 parts/transducers.tex
preparing the witdrawal

@@
 
-The goal of this section is to provide tools that do not require
-\kl{commutative} \kl{$\Nat$-polyregular functions}. To that end, we introduce
-the notion of \kl{$k$-residual transducer}, that generalizes the notion of
-residual transducer introduced for \kl{$\Rel$-polyregular functions} in
-\cite{LOPEZ23b}. We will first introduce the object, provide an algorithm (with
-an oracle function) that computes said object, and finally, check that in
-simple cases, the aperiodicity of a function can be read on its
-\kl{$k$-residual transducer}. The main drawbacks of the generalization are the
-following: one needs to decide whether a \kl{$\Rel$-polyregular function} is in
-$\NPoly$, which is only known in the \kl{commutative} case, and the
-characterization in terms of counter free also assumes \kl{commutativity}.
-However, we believe that the introduction of this object is relevant.
+The goal of this section is to provide tools that do not require the assumption
+of \kl{commutativity}. To that end, we introduce the notion of \kl{$k$-residual
+transducer}, that generalizes the notion of residual transducer introduced for
+\kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}. This notion of residual
+transducer has its own interest, as it shifts the attention from (compatible)
+equivalence relations of finite index that are the classical tool in automata
+theory, towards (compatible) order relations that are \kl{well-quasi-ordered},
+which is the order counterpart of having finite index.
+
+While \kl{$k$-residual transducers} will provide canonical objects associated
+to functions in $\NPoly$, their effective construction requires the ability to
+decide $\NPoly$ inside $\ZPoly$, which is not currently known (except in the
+commutative case). Furthermore, the correspondence between \kl{counters} in
+this automaton and \kl{star-free $\Nat$-polyregular functions} is non-trivial,
+and cannot be used \emph{as is} to decide membership in $\NSF$
+based on this analysis of the \kl{$k$-residual transducer}.
+
+\AP The upcoming sections will heavily refer to the notion of \intro{residual} of
+a function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
+\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
+denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
+over words in $\Sigma^*$.
+
+\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
+quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
+that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
+infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
+\reintro{good}. The notion of \reintro{good sequences} can also be applied to
+binary relations that are not orderings, and a binary relation $R$ for which
+every infinite sequence is \reintro{good} is said to be \intro{well}
+\cite{MELL98}.
+
+\subsection{Residual Transducers and WQOs}
 
@@
 for \kl{commutative} \kl{$\Nat$-polyregular functions}. The notion of
-transducer is based on the \kl{marble transducers} \cite{EHB99}.
+transducer is based on the marble transducers \cite{EHB99}, and follows the
+general pattern of transducers introduced in \cite{LOPEZ23b}.
 
@@
 
-Given the syntax of an \kl{$\mathcal{H}$ transducer}, let us now 
+Given the syntax of an \kl{$\mathcal{H}$-transducer}, let us now 
 introduce its semantics, which is given by a simple recursion
@@
 \end{definition}
-
-\AP As in the case of regular automata, we generalize the function $\delta
-\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
-q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
-$\lambda(q,\varepsilon) \defined 0$, and $\lambda(q, au) \defined \app{\lambda(q,a)}{u}
-+ \lambda(\delta(q,a), u)$. Following these equations, we obtain that
-$\aTransd(q_0, uv) = \aTransd(\delta(q_0, u), v) + \lambda(q_0, u)(v)$.
-Furthermore, let us notice that if $\mathcal{H}$ is closed under finite sums
-and shifts, then $\lambda(q, u) \in \mathcal{H}$ for all $(q,u) \in Q \times
-\Sigma^*$.
 
@@
 
+\AP As in the case of regular automata, we generalize the function $\delta
+\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
+q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
+$\lambda(q,\varepsilon) \defined 0$, and recursively define $\lambda(q, au)
+\defined \app{\lambda(q,a)}{u} + \lambda(\delta(q,a), u)$. Following these
+equations, we can write for all $q \in Q$, $u,v \in \Sigma^*$, $\aTransd(q, uv)
+= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice
+that if $\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q,
+u) \in \mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.
+
 \AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
@@
 
-\begin{theorem}[Cite gaëtan]
+It is already known in the literature that this notion of transducers correctly
+captures both $\NSF$ and $\NPoly$ in the sense that the following fact will
+clarify.
+
+\begin{fact}[Cite gaëtan]
     Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
@@
     \begin{itemize}
-        \item $f \in \NPoly[k]$,
-        \item $f$ is computed by a \kl{$\NPoly[k-1]$-transducer}.
+        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
+        \item $f$ is computed by an \kl{$\NPoly[k-1]$-transducer}
+            (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
+            \kl{counter-free}).
     \end{itemize}
-    \begin{itemize}
-        \item $f \in \NSF[k]$,
-        \item $f$ is computed by a \kl{$\NSF[k-1]$-transducer}
-            that is counter-free.
-    \end{itemize}
-\end{theorem}
+\end{fact}
 
-\AP Let us now introduce the notion of \intro{residual} of a function $f
-\colon \Sigma^* \to \Rel$, by defining $\app{f}{u} \defined w \mapsto f(uw)$.
-The collection of \reintro{residuals} of a function $f$ is denoted $\Res(f)$
-and is defined as the set of $\app{f}{u}$ where $u$ ranges over words in
-$\Sigma^*$.
-
-\AP Let $k \in \Nat$. The residual transducer in the case of $\Rel$ output was
-built using an equivalence relation defined as $f \equiv_k g$ if and only if $g
-- f \in \ZPoly[k-1]$. However, in the case of $\Nat$ output, the relation
-cannot be symmetric, and furthermore, we are going to order words rather than
-residuals. Namely, given $k \in \Nat$, and $f \colon \Sigma^* \to \Nat$, we
-define $u \resleq{f}{k} v$ if and only if $\app{f}{v} - \app{f}{u} \in
-\NPoly[k-1]$.
+\AP Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if and only if $(g - f)
+\in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if $(g - f) \in
+\NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we are going to
+study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f), \npolyleq[k])$. However,
+to simplify notations, instead of writing $\app{f}{u} \npolyleq[k] \app{f}{v}$
+when $u,v \in \Sigma^*$, we will use the convenient notation $u \resleq{f}{k}
+v$, and directly consider the space $(\Sigma^*, \resleq{f}{k})$.
 
@@
 
-There is one major issue arising when defining the notion of residual
-transducer based on the quasi-ordering $\resleq{f}{k}$. The naïve approach is
-to consider as states some representatives for the $\resleq{f}{k}$ minimal
-elements of $\Sigma^*$, all computing the same function
-(\cref{good-residual-ordering:fact}), and define transitions by letting
-$\delta(u, a)$ be \emph{one} state $v$ such that $v \resleq{f}{k} ua$. This
-naïve approach does not yield a canonical model.
-For instance, the function
-of \cref{non-canonical-transd:ex} can be computed in two ways if we only
-consider $\resleq{f}{k}$ minimal elements, as witnessed in
+Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
+canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
+approach is to consider as states some representatives for the $\resleq{f}{k}$
+minimal elements of $\Sigma^*$ (assuming there are finitely many of them), all
+computing the same function (\cref{good-residual-ordering:fact}), and define
+transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
+\resleq{f}{k} ua$. This naïve approach does not yield a canonical model. For
+instance, the function of \cref{non-canonical-transd:ex} can be computed in two
+ways if we only consider $\resleq{f}{k}$ minimal elements, as witnessed in
 \cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
-$\NSF[0]$, but one of the two presented transducers has a \kl{counter}.
-
-This is why in \cref{residual-transducer:def}, we introduce the prefix ordering
+$\NSF[0]$, but one of the two presented transducers has a \kl{counter}. This is
+why in \cref{residual-transducer:def}, we introduce the prefix ordering
 $\prefleq$ to further constrain how the model computes.
@@
 
-\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
-quasi-ordered set $(X, \leq)$ is \kl{good} whenever there exist $i < j$ such
-that $u_i \leq u_j$. The set $X$ is a \kl{well-quasi-ordering} when
-every infinite sequence is \kl{good}. A sequence is \kl{bad}
-when it is not \kl{good}.
-
 \begin{lemma}
@@
     every infinite, $\prefleq$-increasing sequence is \kl{good}
-    in $(\Sigma^*, \resleq{f}{k})$.
+    in $(\Sigma^*, \resleq{f}{k})$
+    (or equivalently, such that the relation $(\prefleq \Rightarrow \resleq{f}{k})$
+    is \kl{well}).
     Then, \cref{residual:algo} terminates on the input $(f,k)$.
 \end{lemma}
+
+The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
+section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
+\Rightarrow \resleq{f}{k})$ is \kl{well}, but $\resleq{f}{k}$ is a
+\kl{well-quasi-ordering}. This has to be compared with the property of being
+\kl{$(k,\Nat)$-combinatorial}, which states that ultimately, the function is
+increasing in every direction.
 
@@
 
+\begin{conjecture}
+    Let $f \in \ZPoly[k]$
+    be \kl{combinatorial}.
+    Then, $f$ is non-negative and for every pair
+    $(u,w) \in \Sigma^*$,
+    the sequence $\seqof{u w^n}{n \in \Nat}$
+    is \kl{good}.
+\end{conjecture}
+\begin{proof}
+    By induction on $k$.
+\end{proof}
+
+\begin{conjecture}
+    Let $f \in \ZPoly[k]$
+    be a non-negative function such that for every
+    $(u,w) \in \Sigma^*$,
+    the sequence $\seqof{u w^n}{n \in \Nat}$
+    is \kl{good}.
+    Then, $f$ belongs to $\NPoly[k]$.
+\end{conjecture}
+\begin{proof}
+    By induction on $k$.
+    For $k = 0$, non-negative is enough.
+    To go beyond, we
+
+\end{proof}
+
 \begin{theorem}
     \label{non-commutative-npoly:thm}
-    Let $f \in \ZPoly$ and $k \in \Nat$,
+    Let $f \in \ZPoly$ be a non-negative function, 
+    and $k \in \Nat$,
     the following are equivalent:
@@
 \end{proof}
+
+\subsection{Aperiodicity and Counters}
 

### 2024-01-31 parts/transducers.tex
adding a small cojnecture
adding polyblind?

@@
 \cite{MELL98}.
+
+\subsection{Polyblind functions}
+
+We are going to prove that polyblind functions are essentially
+commutative.
+
+\AP Given $q,k \in \Nat$, we define $\MSO^q_k$ to be the collection of types of
+tuples of at most $k$ variables using quantifier rank at most $q$. We extend
+the notation to a function $\MSO^q_k (u)$ that assigns to a word $u$ the number
+of representative for each type.
+
+\begin{lemma}
+    Let $q \in \Nat$. The image
+    $\MSO^q_1(\Sigma^*)$ is a semilinear set.
+\end{lemma}
+\begin{proof}
+    Here is a context-free grammar that enumerates words
+    in $\Sigma^*$ with letters colored by their $\MSO^q_1$ type.
+    The initial non-terminal decides the global $\MSO^q$ type of the word,
+    and given a non-terminal with an $\MSO^q$-type, we can split it in three
+    by choosing a letter $a \in \Sigma$, an $\MSO^q$-type for the left,
+    and an $\MSO^q$-type for the right, so that the types are coherent.
+    We just have to remember the $\MSO^q$ type of the left and right
+    of the current non-terminal to do that properly.
+
+    By the Parikh’s theorem, the image of this is a semilinear set, that
+    precisely computes $\MSO^q_1(\Sigma^*)$.
+\end{proof}
+
+\AP
+Recall that a \intro{polyblind function} is a function such that.
+
+\begin{lemma}
+    Let $f \colon \Sigma^* \to \Rel$ be a \kl{polyblind function}.
+    Then, there exists a computable function $g \colon \Nat^p \to \Rel$,
+    that is \kl{commutative} and such that.
+\end{lemma}
+
+\AP
+Look at the image of $\MSO^q_k$, is it semi-algebraic instead of semi-linear?
+\begin{lemma}
+    $\MSO^q_k(\Sigma^*)$ is a set of tuples that is obtained as a finite union 
+    images of polynomials?
+\end{lemma}
+
+\subsection{}
 
@@
 \end{lemma}
+
+\begin{conjecture}
+    Let $q,k \in \Nat$.
+    Associate to a word $w$ the number of representatives
+    of every $\MSO(q,k)$ type. 
+    The image of this map is a semilinear set.
+    In particular, it has a simple basis.
+\end{conjecture}
 

### 2024-01-31 globals/metadata.tex
avancées
writing to publish
lipics compiles

@@
-\title{
-    Star-free $\Nat$-polyregular functions:
-    the commutative case.
-}
-\author{
-    Aliaume Lopez
-}
-\affiliation{%
-  \institution{Univeristy of Warsaw}
-  \city{Warsaw}
-  \country{Poland}}
-\email{ad.lopez@uw.edu.pl}
+\date{\today}
 
+\title{Commutative $\Nat$-polyregular functions}
+
+\author{Aliaume Lopez}%
+       {University of Warsaw}%
+       {ad.lopez@uw.edu.pl}%
+       {https://orcid.org/0000-0002-1825-0097}%
+       {Gaëtan Douéneau-Tabot}
+
+\authorrunning{A. Lopez}
+
+%\titlerunning{Commutative $\Nat$-polyregular functions}
 \date{\today}
-\bibliography{globals/papers}
+
+\Copyright{Copyright Holder Name}
+
+%TODO mandatory; please add comma-separated list of keywords
+\keywords{Dummy keyword} 
+\category{} %optional, e.g. invited paper
+\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
 
 \newcommand{\acknowledge}{
-    I would like to thank 
-    Anonymous Author(s)
-    %Gaëtan Douéneau-Tabot
-    for mentioning this problem, checking 
-    my counter-example, and supporting me in publishing
-    this paper.
+    Acknowledge Gaëtan
 }
@@
 
-\setcopyright{acmlicensed}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Access}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{CVIT 2016}
+\EventAcronym{CVIT}
+\EventYear{2016}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-02-12 globals/metadata.tex
update

@@
        {https://orcid.org/0000-0002-1825-0097}%
-       {Gaëtan Douéneau-Tabot}
+       {}
 
@@
 
-\Copyright{Copyright Holder Name}
+\Copyright{Aliaume Lopez}
 
 %TODO mandatory; please add comma-separated list of keywords
-\keywords{Dummy keyword} 
+\keywords{Rational series, weighted automata, polyregular function, commutative input} 
 \category{} %optional, e.g. invited paper
 \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+
+\ccsdesc[500]{Theory of Computation~Models of Computation}
+\ccsdesc[500]{Theory of computation~Automata extensions}
 
@@
 \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
+\EventShortTitle{ICALP 2023}
+\EventAcronym{ICALP}
+\EventYear{2023}
 \EventDate{December 24--27, 2016}

### 2024-02-12 parts/transducers.tex
update

@@
 
-\subsection{Polyblind functions}
+\subsection{Deciding $\Nat$-polyregular functions}
 
-We are going to prove that polyblind functions are essentially
-commutative.
+The goal is to prove \cref{decide-n-poly-general:thm}.
+To that end, we will use factorisation forests.
+\begin{definition}
+    Forests of depth $d$ by induction.
+\end{definition}
 
-\AP Given $q,k \in \Nat$, we define $\MSO^q_k$ to be the collection of types of
-tuples of at most $k$ variables using quantifier rank at most $q$. We extend
-the notation to a function $\MSO^q_k (u)$ that assigns to a word $u$ the number
-of representative for each type.
+\begin{remark}
+    Place the nested higman embedding on the forests, it is a well-quasi-ordering.
+\end{remark}
+
+\begin{definition}
+    Counting functions. $\vcount{q}$ where $q$ is a pattern.
+\end{definition}
+
+\begin{definition}
+    Divisibility of patterns.
+\end{definition}
+
+\begin{definition}
+    Forest iteration, where $q$ is a pattern.
+\end{definition}
+
+\begin{theorem}[Simon, restated in ]
+    For every $f \in \ZPoly[k]$,
+    there exists a computable function $h \colon \Sigma^* \to F_d(M)$,
+    and a function $g \colon F_d(M) \to \Rel$ that is a counting function
+    such that.
+\end{theorem}
+
+\textbf{TODO: domains should have a good shape, i.e.,
+    intersections of sentences of the shape
+    this pattern has at most "X" matches, and should be
+    downwards closed}.
+
+\textbf{TODO: problem, there is a problem, it is not because
+    the number of matches is unbounded, that we can
+    iterate the shape itself no?}
 
 \begin{lemma}
-    Let $q \in \Nat$. The image
-    $\MSO^q_1(\Sigma^*)$ is a semilinear set.
+    A function $g \colon F_d(M) \to \Rel$ is $\Nat$-polyregular
+    over a domain $D$ if and only if for every 
+    pattern iteration, the function obtained is
+    $\Nat$-polyregular (commutative).
 \end{lemma}
 \begin{proof}
-    Here is a context-free grammar that enumerates words
-    in $\Sigma^*$ with letters colored by their $\MSO^q_1$ type.
-    The initial non-terminal decides the global $\MSO^q$ type of the word,
-    and given a non-terminal with an $\MSO^q$-type, we can split it in three
-    by choosing a letter $a \in \Sigma$, an $\MSO^q$-type for the left,
-    and an $\MSO^q$-type for the right, so that the types are coherent.
-    We just have to remember the $\MSO^q$ type of the left and right
-    of the current non-terminal to do that properly.
+    By induction on the domain $D$, and on the divisibility ordering
+    of patterns.
+    Base case, for all pattern iteration, the value of $g$ is bounded.
+    Then, $g$ has finitely many outputs, and they are all in $\Nat$,
+    we have concluded.
 
-    By the Parikh’s theorem, the image of this is a semilinear set, that
-    precisely computes $\MSO^q_1(\Sigma^*)$.
+    Let $q_i$ be a maximal pattern of $g$. If $q_i$ cannot be iterated in $D$,
+    then the value of $q_i$ is bounded by some constant, and we can rewrite it
+    as counting "full" patterns?
+
+    Otherwise, $q_i$ can be iterated, and therefore we can
+    pump to prove that its coefficient must be strictly positive.
+    Then, using some derivation magic, we can recurse.
 \end{proof}
 
-\AP
-Recall that a \intro{polyblind function} is a function such that.
-
-\begin{lemma}
-    Let $f \colon \Sigma^* \to \Rel$ be a \kl{polyblind function}.
-    Then, there exists a computable function $g \colon \Nat^p \to \Rel$,
-    that is \kl{commutative} and such that.
-\end{lemma}
-
-\AP
-Look at the image of $\MSO^q_k$, is it semi-algebraic instead of semi-linear?
-\begin{lemma}
-    $\MSO^q_k(\Sigma^*)$ is a set of tuples that is obtained as a finite union 
-    images of polynomials?
-\end{lemma}
-
-\subsection{}
+\begin{theorem}
+    \label{decide-n-poly-general:thm}
+    Let $f \in \ZPoly[k]$, the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NPoly$;
+        \item $f \in \NPoly[k]$;
+        \item for all $h$ commutative polyregular function,
+            $f \circ h$ in $\NPoly$
+        \item for all pattern iteration,
+            $f(pattern)$ is a \kl{$\Nat$-rational polynomial}.
+    \end{enumerate}
+    Furthermore, the properties are decidable.
+\end{theorem}
+\begin{proof}
+    (1) implies (2) is okay, because of generic results.
+    (2) implies (3) implies (4) is completely fine.
+    Now, let us prove that (4) implies $1$.
+\end{proof}
 
@@
 \end{fact}
+
+\subsection{Polyblind functions}
+
+We are going to prove that polyblind functions are essentially
+commutative.
+
+\AP Given $q,k \in \Nat$, we define $\MSO^q_k$ to be the collection of types of
+tuples of at most $k$ variables using quantifier rank at most $q$. We extend
+the notation to a function $\MSO^q_k (u)$ that assigns to a word $u$ the number
+of representative for each type.
+
+\begin{lemma}
+    Let $q \in \Nat$. The image
+    $\MSO^q_1(\Sigma^*)$ is a semilinear set.
+\end{lemma}
+\begin{proof}
+    Here is a context-free grammar that enumerates words
+    in $\Sigma^*$ with letters colored by their $\MSO^q_1$ type.
+    The initial non-terminal decides the global $\MSO^q$ type of the word,
+    and given a non-terminal with an $\MSO^q$-type, we can split it in three
+    by choosing a letter $a \in \Sigma$, an $\MSO^q$-type for the left,
+    and an $\MSO^q$-type for the right, so that the types are coherent.
+    We just have to remember the $\MSO^q$ type of the left and right
+    of the current non-terminal to do that properly.
+
+    By the Parikh’s theorem, the image of this is a semilinear set, that
+    precisely computes $\MSO^q_1(\Sigma^*)$.
+\end{proof}
+
+\AP
+Recall that a \intro{polyblind function} is a function such that.
+
+\begin{lemma}
+    Let $f \colon \Sigma^* \to \Rel$ be a \kl{polyblind function}.
+    Then, there exists a computable function $g \colon \Nat^p \to \Rel$,
+    that is \kl{commutative} and such that the following are equivalent:
+    \begin{itemize}
+        \item $g$ is $\Nat$-polyregular
+        \item $f$ is $\Nat$-polyregular
+    \end{itemize}
+\end{lemma}
+
+\AP
+Look at the image of $\MSO^q_k$, is it semi-algebraic instead of semi-linear?
+\begin{lemma}
+    $\MSO^q_k(\Sigma^*)$ is a set of tuples that is obtained as a finite union 
+    images of polynomials?
+\end{lemma}

### 2024-02-12 parts/appendix.tex
big big rewrite

@@
 
-\begin{proof}[Proof of \cref{decidable-commutative-rat:lemma}]
+\begin{proof}[Proof of \cref{size-of-word-nsf:ex}]
+    Let us define $M \defined (\set{1}, \times)$ which is 
+    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
+    defined by $\mu(w) \defined 1$, and
+    $\pi \colon M^2 \to \Nat$
+    that is the constant function equal to $1$.
+    We check that for all $w \in \Sigma^*$:
+    \begin{equation*}
+        \pi^\dagger(w)
+        =
+        \sum_{uv = w} 1
+        =
+        \card{w} + 1
+        = f(w)
+        \quad . 
+        \qedhere
+    \end{equation*}
+\end{proof}
+
+\begin{proof}[Proof of \cref{decidable-commutative-poly:lemma}, in the polyregular case]
+    Let $\Sigma$ be a finite alphabet, endowed with a
+    total ordering over its letters,
+    and let $f \colon \Sigma^* \to \Rel$ be a
+    \kl{$\Rel$-polyregular function}.
+    The map $\polysort \colon \Sigma^* \to \Sigma^*$
+    that sorts the letters in the input word $w$ according
+    to the chosen ordering is a \kl{polyregular function}
+    \cite[Proposition II.12]{LOPEZ23b}.
+    As a consequence,
+    $(f \circ \polysort) \in \ZPoly$.
+    Finally, $f$ is \kl{commutative} if and only
+    if $f = f \circ \polysort$, which is decidable
+    in the case of $\ZPoly$
+    \cite[Corollary II.24]{LOPEZ23b}, and more generally in the case 
+    of $\ZRat$ \cite[Corollary 3.6 p 38]{berstel2011noncommutative}.
+\end{proof}
+
+\begin{proof}[Proof of \cref{decidable-commutative-rat:lemma} in the general case]
     Because $f$ is a \kl{$\Rel$-rational series},
@@
     $f \circ g \in \ZPoly[k+ \ell - 1]$
-    \cite[]{LOPEZ23b}.
+    \cite{LOPEZ23b}.
 
@@
 
+\begin{proof}[Proof of \cref{negative-not-nrat:ex}]
+    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
+    Thus, 
+    $P(X) \defined X$ is
+    a \kl{$\Nat$-rational polynomial}. Similarly,
+    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-polyregular function},
+    showing that $Q(X) \defined X^2 + 3$
+    is a \kl{$\Nat$-rational polynomial}.
+    Finally, 
+    $T(X) \defined - X$ cannot be 
+    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
+    are non-negative.
+\end{proof}
+
 \section{Proofs of
 section \ref{polynomials:sec}}
+
+\begin{proof}[Proof of \cref{negative-but-npoly:ex}]
+    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
+    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
+    It is clear that $f \in \NPoly$, and that 
+    $P$ is \kl{represented} by
+    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
+\end{proof}
 
@@
 \end{proof}
+
+\begin{proof}[Proof of \cref{n-rat-correct:lem}]
+    Let $f$ be the \kl{commutative}
+    \kl{$\Nat$-rational series}
+    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
+    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
+    it is in fact a \kl{$\Nat$-polyregular function}
+    (\cref{polynomial-rational-polyreg:fact}).
+    Using \cref{n-poly-combinatorics:lem},
+    there exists a number $\omega \in \Nat$
+    and polynomial $Q$ with \kl{non-negative}
+    \kl{maximal monomials} such that
+    for all $X_1, \dots, X_p \geq \omega$:
+    \begin{equation*}
+        f\left(
+            \prod_{i = 1}^p a_i^{\omega X_i}
+        \right)
+        = Q(X_1, \dots, X_p)
+        = P(\omega X_1, \dots, \omega X_p) 
+        \quad .
+    \end{equation*}
+    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
+    as polynomials,
+    and in particular that
+    the \kl{maximal monomials} of 
+    $P$ are \kl{non-negative}.
+
+    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
+    the polynomial $\restr{P}{\nu}$ continues to be represented
+    by a \kl{$\Nat$-polyregular function}, namely
+    $f$ partially applied to a word. As a consequence,
+    the \kl{maximal monomials} of
+    $\restr{P}{\nu}$ are also \kl{non-negative}, 
+    and
+    we have proven that $P \in \CorrectPoly$.
+\end{proof}
+
+\begin{fact}
+    \label{maxi-monomials-submodular:fact}
+    Let $P, Q$ be two polynomials.
+    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
+    \cup \MaximalMonomials(Q)$.
+\end{fact}
+
+\begin{fact}
+    \label{translation-maximal:fact}
+    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
+    \begin{equation*}
+        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
+        \quad .
+    \end{equation*}
+\end{fact}
 
@@
 
+\begin{proof}[Proof of \cref{decidability-correct:lem}]
+    Given $P \in \Rel[\vec{X}]$,
+    one applies \cref{derivation-translation:lem}
+    to obtain a bound $K$, and then 
+    computes $\translate{K}(\restr{P}{\nu})$ for all 
+    partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
+    Then $P \in \CorrectPoly$ if and only if all those polynomials
+    belong to $\Nat[\vec{X}]$, which is decidable.
+\end{proof}
+
 \section{Proofs of section \ref{beyond-polynomials:sec}}
+
+\begin{proof}[Proof of \cref{decompose-polynomial:lem}]
+    We apply
+    \cref{n-poly-combinatorics:lem} to compute the $\omega$.
+    Using this $\omega$,
+    for any choice of
+    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
+    there exists a polynomial $P_r \in \Rel[X_1, \dots, X_p]$
+    such that $X_1, \dots, X_p \geq \omega$:
+    \begin{equation*}
+        f\left(
+        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
+        = P_r(X_1, \dots, X_p) \quad .
+    \end{equation*}
+    To get the full family of polynomials needed requires
+    just to only iterate subsets of letters, and fix the other ones
+    to be below the threshold $\omega^2$.
+\end{proof}
+
+\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
+    \textbf{TODO: we need to talk about the number of variables hehe}
+    Let us prove that
+    \cref{f-npoly-combi:item}
+    implies \cref{f-combinatorial:item}. 
+    We compute the $\omega$ by applying 
+    \cref{n-poly-combinatorics:lem}, and then
+    notice that the polynomial $P$ obtained is
+    by definition a \kl{$\Nat$-rational polynomial}.
+
+    For the converse implication, we use \cref{decompose-polynomial:lem} to
+    effectively compute a polynomial representation of $f$. Using the
+    assumption, all of these polynomials are actually \kl{$\Nat$-rational
+    polynomials}. Let us now construct a new way to compute $f$, by first
+    computing the \kl{$\omega$-type} of the input (which is possible in
+    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
+    which is possible in $\NPoly$.
+\end{proof}
+
+\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
+    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
+
+    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
+    only the empty word $\varepsilon$, and because $f$ is a
+    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
+    Note that $f$ is therefore a constant function, equal to a natural
+    number, hence belongs to $\NSF$.
+
+    \textbf{Induction.}
+    Because $f$ is \kl{ultimately polynomial},
+    there exists $N_0 \in \Nat$, 
+    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
+    such that
+    \begin{equation*}
+        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
+        = 
+        P(\seqof{X_a}{a \in \Sigma})
+        \quad
+        \text{ when }
+        \forall a \in \Sigma, X_a \geq N_0
+        \quad .
+    \end{equation*}
+    Furthermore, leveraging \cref{decompose-polynomial:lem},
+    we obtain $\omega \in \Nat$
+    and a family of polynomials 
+    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^\Sigma}$
+    that represent $f$.
+    For all $t \in \ModuloTypes[\omega]^\Sigma$
+    that have threshold greater than $\omega^2$,
+    we conclude that $P_t$ and $P$ coincide (up to translation) over an infinite grid,
+    hence we conclude that
+    if all letters appear at least $N_1 \defined \max (\omega^2, N_0)$ times,
+    then $f$ computes $P$. But this also shows that
+    $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
+    \kl{represented} by a function $f_1 \in \NSF[k]$.
+
+    To compute $f$ on words where some letter appears fewer than $N_1$ times,
+    we partially apply $f$ to obtain a function $g$ over a restricted input
+    alphabet. Using the induction hypothesis, $g$ belongs to $\NSF[k]$, and we
+    can therefore (effectively) compute a representation of $f$ in $\NSF[k]$.
+
+    For the effective part of the theorem, 
+    we know from \cite[Theorem V.13]{LOPEZ23b}, it is decidable whether
+    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
+    it is decidable whether $f \in \NPoly$.
+    Furthermore, one can enumerate functions in $\NSF[k]$,
+    and check equivalence to effectively represent $f$ as a
+    \kl{star-free $\Nat$-polyregular function}.
+\end{proof}
 
@@
 
+\begin{fact}
+    \label{unique-res-transducer:fact}
+    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
+    Then $f$ has at most one \kl{$k$-residual transducer}.
+\end{fact}
+\begin{proof}[Proof of \cref{unique-res-transducer:fact}]
+    Let $\aTransd_1$ and $\aTransd_2$ be two
+    \kl{$k$-residual transducers} for $f$.
+    The two initial states must be $\varepsilon$.
+    Let us prove by induction on $u \in \Sigma^*$ that
+    $\delta_1(\varepsilon, u) = \delta_2(\varepsilon, u)$
+    and that $Q_1$ equals $Q_2$ over prefixes of $u$.
+    This will prove that 
+    $Q_1 = Q_2$, hence that $\aTransd_1 = \aTransd_2$.
+
+    Let $u \in \Sigma^* \cap Q_1 \cap Q_2$ and $a \in \Sigma$, $v_1 \in Q_1$ be
+    defined as $v \defined \delta_1(u,a)$, and $v_2 \defined \delta_2(u,a)$.
+    Remark that by induction hypothesis, for all $v \prefle u$, $v \in Q_1 \cap
+    Q_2$. If $\delta_1(u,a) = Q_1$, it means that for all $v \in Q_2$ such that
+    $v \prefle ua$, we have $\neg( v \resleq{f}{k} ua )$. The only possible
+    transition in $\aTransd_2$ is therefore $\delta_2(u,a) = ua$, and $ua \in
+    Q_2$. Similarly, if $\delta_1(u,a) \prefleq u$, then $\delta_2(u,a) =
+    \delta_1(u,a)$ by definition of $\delta_2$ as a maximum.
+\end{proof}
+
+\begin{fact}
+    \label{q-o-prefix-cool:fact}
+    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
+    At each step of the \texttt{while loop}
+    of \cref{residual:algo}, the sets
+    $Q$ and $O$ are such that
+    \begin{enumerate}
+        \item $Q \cup O$ is a \kl{downwards closed} subset of 
+            $\Sigma^*$ for $\prefleq$;
+        \item elements in $O$ are pairwise incomparable
+            for $\prefleq$, and are maximal
+            for $\prefleq$ inside $Q \cup O$.
+    \end{enumerate}
+\end{fact}
 \begin{proof}[Proof of
@@
 \end{proof}
+
+\begin{proof}[Proof of \cref{correct-residual:lemma}]
+    Because of \cref{q-o-prefix-cool:fact},
+    we already know that $q_0 = \varepsilon$,
+    $Q$ is a \kl{downwards closed} subset of $\Sigma^*$
+    for $\prefleq$, 
+    that every state of $Q$ is accessible from $q_0$.
+    Notice that at every step,
+    $\lambda(u,a)$ is defined as
+    $\app{f}{ua} - \app{f}{\delta(u,a)}$.
+    Finally, since $Q \cup O$ is a \kl{downwards closed} subset of $\Sigma^*$
+    at every step,
+    we have that at step $i$,
+    for all $ua \in O_i$,
+    $\setof{w \in Q}{w \prefleq ua} = \setof{w \in Q_i}{ w \prefleq ua}$,
+    which proves that the maximum considered in the algorithm
+    is indeed computing correctly.
+\end{proof}
+
 \begin{proof}[Proof of \cref{wqo-implies-termination:lemma}]
@@
     tree with a finite branching (at most $\card{\Sigma}$), and has an infinite
-    branch $\seqof{u_j}{j \in \Nat}$ thanks to König’s lemma.
+    branch $\seqof{u_j}{j \in \Nat}$ thanks to König' s lemma.
 
@@
         \sum_{\phi_i \in t_l^j(\vec{y}) \odot t_r^j(\vec{z})}
-        \\
-        &\quad\quad
         \underbrace{
@@
 \end{proof}
+
+\begin{proof}[Proof of \cref{aperiodic-iff-residual:lem}]
+    It is clear that $\NSF \subseteq \ZSF$, and known that if $f \in \ZSF$
+    then it is \kl{ultimately polynomial}. Furthermore, if the \kl{$0$-residual
+    transducer} of $f$ is \kl{counter-free}, then $f \in \NSF$
+    using \cref{transducer-nsf-npoly:lemma}.
+
+    Assume that $f$ is \kl{ultimately polynomial}, let us prove that the
+    \kl{$0$-residual transducer} of $f$ is \kl{counter-free}. 
+    Note that because $f \in \NPoly[0]$,
+    $u \resleq{f}{0} v$ if and only if $\app{f}{u} = \app{f}{v}$.
+    In particular, in a \kl{$0$-residual transducer} of $f$,  two states that
+    represent the same residual must be incomparable for the prefix relation.
+
+    Let $(q,w^n)$ be
+    a counter with $n \geq 1$. This means that $\delta(q, w^n) = q$ in the
+    automaton, and implies that $q \resleq{f}{0} qw^n$, hence that $\app{f}{q}
+    = \app{f}{qw^n}$. Because $f$ is \kl{ultimately polynomial},
+    we know that $\app{f}{qw^n} = \app{f}{qw^{n+1}}$,
+    hence that $\app{f}{qw} = \app{f}{q}$.
+
+    Let us write $t \defined \delta(q,w) = \delta(q,w^{n+1})$. We know that
+    $\app{f}{q} = \app{f}{t}$. Assume by contradiction that $t$ and $q$ are
+    incomparable for the prefix relation. Let us split $w = w_1 w_2$ where
+    $w_1$ is the shortest prefix of $w$ such that $s_0 \defined \delta(q,w_1)$
+    is an ancestor of $q$ and of $t$ for the prefix relation, it must exist
+    because $\delta(q,w_1 w_2) = t$.
+
+    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
+    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
+    the smallest prefix $v$ of $w_1$ such that $\delta(t, v)$ is a strict
+    prefix of $s_0$. It must exist, otherwise $s_0$ is always a prefix of
+    $s_1$. Because $\app{f}{t} = \app{f}{q}$, we conclude that $\app{f}{tv} =
+    \app{f}{qv}$. However, this contradicts the minimality of $w_1$, since
+    $\delta(t,v)$ is an ancestor of $q$ and $t$.
+
+    We have proven that $s_0$ and $s_1$ are comparable, hence they are equal,
+    since $\app{f}{s_1} = \app{f}{t w_1} = \app{f}{q w_1} = \app{f}{s_0}$.
+    Finally, we have proven that $\delta(q, w_1) = s_0$, $\delta(s_0, w_2) =
+    t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
+    As a consequence $t$ and $q$ were comparable for the prefix relation,
+    hence equal, and therefore $\delta(q, w) = q$.
+\end{proof}
+
+\begin{proof}[Proof sketch of \cref{sf-no-periods-on-sequences:lemma}]
+    Let $t$ be a first order type,
+    the map $g_t \colon X \mapsto \vcount{t}(uw^X)$ is a
+    \kl{commutative} \kl{star-free $\Nat$-polyregular function},
+    hence ultimately a polynomial, and therefore,
+    translated enough, it is a $\Nat[X]$ polynomial.
+    But this means that $g_t(X+1) - g_t(X) \in \Nat[X]$ for a large
+    enough $X$, which is precisely what suffices.
+    Now, we can do this for every first order type $t$
+    with fewer variables and lower quantifier rank than those
+    needed to define $f$, and conclude.
+\end{proof}

### 2024-02-12 parts/conclusion.tex
big big rewrite

--- 2024-01-16 parts/conclusion.tex
@@
 %! lang = en-US
-\section{Conclusion}
+\section{Outlook}
 \label{sec:ccl}
+
+While \kl{$k$-residual transducers} will provide canonical objects associated
+to functions in $\NPoly$, their effective construction requires the ability to
+decide $\NPoly$ inside $\ZPoly$, which is not currently known (except in the
+commutative case). Furthermore, the correspondence between \kl{counters} in
+this automaton and \kl{star-free $\Nat$-polyregular functions} is non-trivial,
+and cannot be used \emph{as is} to decide membership in $\NSF$
+based on this analysis of the \kl{$k$-residual transducer}.
+
+Let us prove that the \kl{residual transducer} is enough to decide aperiodicity
+in the case of functions in $\NPoly[0]$, that is, with finite image. In
+particular, this generalizes the result of \cite{LOPEZ23b}, by proving that
+$\NSF[0] = \NPoly[0] \cap \ZSF$. Even though the proposed model is canonical,
+and allows to effectively decide aperiodicity for \kl{$\Nat$-polyregular
+functions} of growth rate $0$ (that is, with bounded image), it is not clear
+yet if it can be leveraged to decide aperiodicity for \kl{$\Nat$-polyregular
+functions} of higher growth rates.
+
+\begin{lemma}
+    \label{aperiodic-iff-residual:lem}
+    Let $f \in \NPoly[0]$, the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NSF$,
+        \item $f \in \ZSF$.
+        \item $f$ is \kl{ultimately polynomial},
+        \item The \kl{$0$-residual transducer} of $f$ is \kl{counter-free},
+    \end{enumerate}
+\end{lemma}
+
+\textbf{TODO:} one may think that we can ask that $u \resleq{f}{k} u w^n$
+implies $u \resleq{f}{k} uw$, but it is false, as the following example will
+demonstrate. This is why the definition of counter is more involved in our
+case.
+
+\begin{lemma}
+    \label{sf-no-periods-on-sequences:lemma}
+    Let $f \colon \Sigma^* \to \Nat$ be a \kl{star-free $\Nat$-polyregular
+    function}. Then,
+    for all $u, w \in \Sigma^*$,
+    there exists $n \in \Nat$ such that
+    $u w^n \resleq{f}{k} u w^{n+1}$.
+\end{lemma}
+
+\begin{theorem}
+    The following are equivalent for \kl{commutative}.
+    \begin{itemize}
+        \item $f$ is in $\NSF$,
+        \item The residual automata of $f$ has no counters
+    \end{itemize}
+\end{theorem}
 
@@
 
-\begin{center}
-    \begin{tabular}{c|cc}
-        \toprule
-        & \textbf{Commutative} & \textbf{General} \\
-        \midrule
-        \textbf{Commutative}
-        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
-        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
-        \textbf{General} & ? & polyregular 
-    \end{tabular}
-\end{center}
+ This suggests the
+following conjecture in the non-commutative case:
+
+\begin{conjecture}
+    Let $k \in \Nat$,
+    and $f \in \ZPoly[k]$. Then, $f \in \NPoly[k]$
+    if and only if $f$ is \kl{combinatorial}, which is decidable,
+    and an effective conversion procedure exists.
+\end{conjecture}

### 2024-02-12 parts/correction.tex
big big rewrite

@@
 \end{corollary}
-\begin{proof}
-    Let $f$ be the \kl{commutative}
-    \kl{$\Nat$-rational series}
-    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
-    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
-    it is in fact a \kl{$\Nat$-polyregular function}
-    (\cref{polynomial-rational-polyreg:fact}).
-    Using \cref{n-poly-combinatorics:lem},
-    there exists a number $\omega \in \Nat$
-    and polynomial $Q$ with \kl{non-negative}
-    \kl{maximal monomials} such that
-    for all $X_1, \dots, X_p \geq \omega$:
-    \begin{equation*}
-        f\left(
-            \prod_{i = 1}^p a_i^{\omega X_i}
-        \right)
-        = Q(X_1, \dots, X_p)
-        = P(\omega X_1, \dots, \omega X_p) 
-        \quad .
-    \end{equation*}
-    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
-    as polynomials,
-    and in particular that
-    the \kl{maximal monomials} of 
-    $P$ are \kl{non-negative}.
-
-    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
-    the polynomial $\restr{P}{\nu}$ continues to be represented
-    by a \kl{$\Nat$-polyregular function}, namely
-    $f$ partially applied to a word. As a consequence,
-    the \kl{maximal monomials} of
-    $\restr{P}{\nu}$ are also \kl{non-negative}, 
-    and
-    we have proven that $P \in \CorrectPoly$.
-\end{proof}
 
@@
 polynomials, that commutes with \kl{translation operators}. However, the
-\kl{translations operators} do not perfectly commute with partial applications
+\kl{translation operators} do not perfectly commute with partial applications
 $\restr{\cdot}{\nu}$.
@@
         \item $\Diff{K}{ \cdot } \circ \translate{L}
-            = \translate{K} \circ \Diff{L}{\cdot}$,
+            = \translate{L} \circ \Diff{K}{\cdot}$,
         \item $\restr{\cdot}{\nu} \circ \translate{L}
@@
 \end{fact}
-
-To build some intuition on the proof to come, let us sketch the argument in the
-case of a single indeterminate $X$. Given a polynomial $P \in \CorrectPoly$,
-the goal is to build a \kl{$\Nat$-polyregular function} $f$ that
-\kl{represents} $P$. Note that in this case, the function has unary input and
-unary output. We prove the result by induction on the degree $\deg(P)$ of the
-polynomial $P$, where the base case is trivial because when $P$ is a constant
-function, this constant must be in $\Nat$. For the induction step, we can
-assume that $P = \alpha X^n + Q$, where $\deg(Q) < \deg(P)$, $n \geq 1$, and
-$\alpha > 0$. On the one hand, $\alpha X^n$ is a \kl{$\Nat$-rational
-polynomial}, but on the other hand, $Q$ may not belong to $\CorrectPoly$.
-However, there exists $K \in \Nat$ such that $\Diff{K}{X^n} + \translate{K}(Q)$
-belongs to $\CorrectPoly$, for instance by considering $K$ to be the sum of
-absolute values of coefficients of $P$. Furthermore, $\Diff{K}{X^n} +
-\translate{K}(Q)$ has degree smaller than $\deg(P)$, hence is a
-\kl{$\Nat$-rational polynomial}. Now, $\translate{K}(P) = \alpha X^n +
-\Diff{K}{X^n} + \translate{K}(Q)$ is a \kl{$\Nat$-rational polynomial},
-represented by a function $f_1$. Now, for values $k$ of $X$ that are smaller
-than $K$, we can compute $P(k) = \restr{P}{X=k}$, which belongs to $\Nat$. It
-is possible to build a \kl{$\Nat$-polyregular function} $f_2 \colon \Nat \to
-\Nat$ that outputs $P(k)$ if the input is at most $K$, and $f_1(k - K)$
-otherwise.
 
@@
 
-When multiple indeterminate are involved, the notion of 
-degree must be adapted, and to ensure that the decision procedure
-is effective, one must obtain explicit bounds on the
-translations that are needed.
-
-\AP To that end, let us now introduce the ordering over which the induction of
+\AP When multiple indeterminate are involved, the notion of degree must be
+adapted, and to ensure that the decision procedure is effective, one must
+obtain explicit bounds on the translations that are needed. To that end, let us
+now introduce the ordering over which the induction of
 \cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
@@
 \kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
-ordering \cite[see e.g.][Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$
-with the \intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
+ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
+\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
 \kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
 $M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
-\kl{well-quasi-ordering} \cite[see e.g.][Hoare quasi-ordering]{SCSC12}.
-
-\begin{fact}
-    \label{maxi-monomials-submodular:fact}
-    Let $P, Q$ be two polynomials.
-    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
-    \cup \MaximalMonomials(Q)$.
-\end{fact}
+\kl{well-quasi-ordering} \cite[Hoare quasi-ordering]{SCSC12}.
 
@@
 extracts \emph{submaximal} monomials from a given polynomial.
-
-\begin{fact}
-    \label{translation-maximal:fact}
-    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
-    \begin{equation*}
-        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
-        \quad .
-    \end{equation*}
-\end{fact}
 
@@
     For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
-    K}$ fixing at least one indeterminate, $\MaximalMonomials(\restr{P}{\nu})
-    \hoarele \MaximalMonomials{P}$. Because $\restr{P}{\nu} \in \CorrectPoly$,
-    one can use the induction hypothesis to compute a \kl{star-free
-    $\Nat$-polyregular function} $f_\nu$ that \kl{represents} $\restr{P}{\nu}$.
+    K}$ fixing at least one indeterminate,  we know that
+    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
+    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
+    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
+    \kl{represents} $\restr{P}{\nu}$.
 
@@
 \end{corollary}
-\begin{proof}
-    Given $P \in \Rel[\vec{X}]$,
-    one applies \cref{derivation-translation:lem}
-    to obtain a bound $K$, and then 
-    computes $\translate{K}(\restr{P}{\nu})$ for all 
-    partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
-    Then $P \in \CorrectPoly$ if and only if all those polynomials
-    belong to $\Nat[\vec{X}]$, which is decidable.
-\end{proof}
 

### 2024-02-12 parts/counter.tex
big big rewrite

@@
 %! lang = en-US
-\section{Polynomials}
+\section{$\Nat$-rational Polynomials}
 \label{polynomials:sec}
@@
 \kl{$\Nat$-rational polynomial}, and if so, explicitly compute a
-\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$.
-
-It is an easy check that polynomials with coefficients in $\Nat$ are
-\kl{$\Nat$-rational polynomials} (\cref{n-poly-n-poly:example}). However,
-\cref{negative-but-npoly:ex} provides an example of a polynomial with negative
-coefficients that is a \kl{$\Nat$-rational polynomial}. As a consequence, the
-class of interest strictly contains $\Nat[\vec{X}]$.
+\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$. It is an easy
+check that polynomials with coefficients in $\Nat$ are \kl{$\Nat$-rational
+polynomials} (\cref{n-poly-n-poly:example}). However,
+\cref{negative-but-npoly:ex} provides a polynomial with negative coefficients
+that is a \kl{$\Nat$-rational polynomial}. As a consequence, the class of
+interest strictly contains $\Nat[\vec{X}]$.
 
@@
 \end{example}
-\begin{proof}
-    Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
-    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
-    It is clear that $f \in \NPoly$, and that 
-    $P$ is \kl{represented} by
-    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
-\end{proof}
 
@@
 
-Let us now introduce the counter example that was found trying to generalize
-the original proof $\ZPoly$.
-The counter example will use three indeterminates, and
-we will prove in \cref{sec:proof} that \citeauthor{KARH77}’s theorem holds on
-polynomial with at most two indeterminate. In particular, the examples
-appearing in \cite{KARH77} are not invalidated, as they all use at most two
-indeterminates. Furthermore, the converse implication, that is, that
-\kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is correct, and can
-even be strengthened as we will see in \cref{corrected-version:thm}.
+Let us now provide a counter example to \cref{karh:thm}. The counter example
+will use three indeterminates, and we will prove in \cref{sec:proof} that
+\cref{karh:thm} holds on polynomial with at most two indeterminate. In
+particular, the examples appearing in \cite{KARH77} are not invalidated, as
+they all use at most two indeterminates. Furthermore, the converse implication,
+that is, that \kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is
+correct, and can even be strengthened as we will see in
+\cref{corrected-version:thm}.
 
@@
     The \kl{maximal monomials} of $P$
-    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which have
-    \kl{positive coefficients}.
+    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
+    \kl{non-negative}.
 

### 2024-02-12 parts/intro.tex
adding polyblind?
test
big big rewrite

@@
 via its relationships with lambda-terms and so-called for-transducers
-\cite[see, e.g.][]{bojanczyk2019string}. As for regular languages, the class
+\cite{bojanczyk2019string}. As for regular languages, the class
 $\SF$ of \intro{star-free polyregular function} can equivalently be introduced
@@
 \cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
-interesting parameter is the \kl{growth rate} of a function $f$, which is given
+interesting parameter is the \intro{growth rate} of a function $f$, which is given
 by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
@@
 (i.e., the presence of inverse elements). If the proof method does not carry
-from $\Rel$-output to $\Nat$-output functions, it was conjectured that the
-semantic characterization of \kl{star-free $\Rel$-polyregular functions}
-($\ZSF$) would continue to characterize \kl{star-free $\Nat$-polyregular
-functions} ($\NSF$).
+from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
+that the semantic characterization of \kl{star-free $\Rel$-polyregular
+functions} ($\ZSF$) would continue to characterize \kl{star-free
+$\Nat$-polyregular functions} ($\NSF$).
 
@@
 \textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
-\emph{polynomials}. The decidability of $\NPoly$ inside $\ZPoly$ (resp. $\NSF$
-inside $\ZSF$) remains an open problem
-\cite[see e.g.][Open question 5.55]{gaetanphd}.
+\emph{polynomials}, with an incorrect proof. The decidability of $\NPoly$
+inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem \cite[Open
+question 5.55]{gaetanphd}.
 
@@
 \begin{figure}
-    \includestandalone[width=\linewidth]{tikz/class-inclusions}
+    \includestandalone[width=5cm]{tikz/class-inclusions}
     \caption{
@@
 
-\paragraph{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
+\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
 functions} that are furthermore \kl{commutative}. Remark that this enforces a
@@
 that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
-\kl{well-quasi-ordering} with respect to the partial ordering $g \npolyleq[k]
-h$ if and only if $h - g \in \NPoly[k-1]$. In this sense, it generalizes to
-orderings the characterization of $\ZPoly[k]$ as the class of functions $f$
-such that $\Res(f)$ has finite index for the equivalence relation $g
-\zpolyequiv[k] h$ if and only if $g - h \in \ZPoly[k-1]$. The \kl{residual
-transducer} is effectively computable if $f$ is a \kl{commutative}
-\kl{$\Nat$-polyregular function}, and it is conjectured to be computable for
-all functions in $\NPoly$. Finally, we prove that for $f \in \NPoly[0]$, $f \in
-\NSF$ if and only if the \kl{residual transducer} of $f$ has no \kl{counters},
-and conjecture that in full generality, $f \in \NSF$ if and only if the
-\kl{residual transducer} of $f$ is \kl{counter-free}.
+\kl{well-quasi-ordering}, generalizing the characterization of $\ZPoly[k]$ as
+the class of functions $f$ such that $\Res(f)$ has finite index (for some
+equivalence relation). The \kl{residual transducer} is effectively computable
+if $f$ is a \kl{commutative} \kl{$\Nat$-polyregular function}, and it is
+conjectured to be computable for all functions in $\NPoly$.
 
-\paragraph{Outline of the paper.} In \cref{preliminaries:sec}, we introduce the
-multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
+\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
+the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
 $\Rel$-polyregular functions), show that one can decide if a function $f \in
-\ZPoly$ is \kl{commutative} (\cref{decide-commutative:lemma}), and give the
+\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give the
 basic definitions used on multivariate polynomials. In \cref{polynomials:sec},

### 2024-02-12 parts/prelim.tex
big big rewrite

@@
 
-\AP
-In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of integers
-(resp. nonnegative integers). The capital letters $\Sigma,\Gamma$ denotes fixed
-alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$ (resp.
-$\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
+\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
+integers (resp. nonnegative integers). The capital letters $\Sigma,\Gamma$
+denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
+(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
 $\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
@@
 
-\AP
-For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^*
-\to \set{0,1}$ as the indicator function of $L$. If $f,g \colon \Sigma^* \to
-\Rel$ are two functions, we define their Cauchy product as $f \cauchy g (w)
-\defined \sum_{uv = w} f(u)g(w)$. Given a set $X$ of functions from $\Sigma^*$
-to $\Rel$, and a subset $\mathcal{S}$ of $\Rel$, we define
+\AP In order to build functions in $\NPoly$, the following constructions will
+be useful. For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon
+\Sigma^* \to \set{0,1}$ as the indicator function of $L$. Given an $\MSO$
+formula with first-order free variables $\varphi(\vec{x})$, we define
+$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
+a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
+$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
+to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
 $\Span{\mathbb{S}}(X)$ as the smallest subset of functions from $\Sigma^* \to
@@
 \in \mathbb{S}$.
-
-\AP
-Given an $\MSO$ formula with first-order free variables $\varphi(\vec{x})$, we
-define $\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$
-that maps a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$,
-such that $w, \nu \models \varphi(\vec{x})$.
-
-\AP Let $f \colon \Sigma^* \to \Rel$ be a function. We define
-the \intro{growth rate} of
-$f$, $\growth{f} \colon \Nat \to \Nat$, as follows $\growth{f}(n) \defined
-\sup_{\card{w} = n} |f(w)|$. A function $f$ has \kl{growth rate} $k \in \Nat$
-if $\growth{f} = \bigO(n^k)$, and if such a $k$ exists,
-we say that $f$ has \kl{polynomial growth}.
 
@@
 
-\begin{definition}[$\Rel$-polyregular functions {\cite[see, e.g.][]{LOPEZ23b}}]
+\begin{definition}[$\Rel$-polyregular functions {\cite{LOPEZ23b}}]
     \label{nat-rel-poly:def}
@@
 \begin{example}
+    \label{size-of-word-nsf:ex}
     The map $f \colon w \mapsto \card{w} + 1$
@@
 \end{example}
-\begin{proof}
-    Let us define $M \defined (\set{1}, \times)$ which is 
-    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
-    defined by $\mu(w) \defined 1$, and
-    $\pi \colon M^2 \to \Nat$
-    that is the constant function equal to $1$.
-    We check that for all $w \in \Sigma^*$:
-    \begin{equation*}
-        \pi^\dagger(w)
-        =
-        \sum_{uv = w} 1
-        =
-        \card{w} + 1
-        = f(w)
-        \quad . 
-        \qedhere
-    \end{equation*}
-\end{proof}
 
@@
 analysis of their \kl{commutative} counterpart, and others that provide
-a nicer syntax to construct examples.
+a nicer syntax to construct examples. These characterizations are known, and
+can be found in \cite{gaetanphd,LOPEZ23b}.
 
-\begin{lemma}[TODO]
+\begin{lemma}
     \label{polynomial-rational-polyreg:fact}
@@
         \item $f$ has \kl{polynomial growth};
-        \item $f$ can be constructed
-            using $+$, $\cauchy$, and multiplication by
-            an integer $m \in \Nat$ (resp. $m \in \Rel$)
-            starting from
-            indicator functions $\ind{L}$ of regular languages;
         \item \label{npoly-counting-mso:item} $f$ belongs to
@@
 
-Similarly, we provide several known characterizations of
-$\NSF$ and $\ZSF$ in the following lemma.
+Similar characterizations can be obtained for $\NSF$ and $\ZSF$.
 
-\begin{lemma}[TODO]
+\begin{lemma}
     Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
@@
             (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ can be constructed
-            using $+$, $\cauchy$, and multiplication by
-            an integer $m \in \Nat$ (resp. $m \in \Rel$)
-            starting from
-            indicator functions $\ind{L}$ of star-free regular languages;
         \item $f$ belongs to
@@
 
-\textbf{This is wrong, we should talk about the number of variables,
-    and say that we do not know in the case of $\NSF$ if it corresponds
-    to the minimal number of variables needed!}
-\AP Inside $\ZPoly$, there is an infinite hierarchy of classes, defined using
-the \kl{growth rate}. Namely, one defines $\ZPoly[k]$ as the functions $f \in
-\ZPoly$ such that $\growth{f}(n) = \bigO(n^k)$, i.e., have \kl{growth rate}
-$k$. Similarly, we let $\NPoly[k] = \NPoly \cap \ZPoly[k]$.
+\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
+limiting the maximal number of free variables used to construct the function as
+a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
+$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
+$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
+$k$ free variables. It has been proven in
+\cite{LOPEZ23b,bojanczyk2022transducers} that the classes $\ZPoly[k]$ (resp.
+$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
+that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\growth{f}(n) =
+\bigO(n^k)$, and similarly for the other classes.
 
@@
 
-\paragraph{Commutative Polyregular Functions.}
-\AP 
-Because this paper focuses on functions that are commutative both in the
+\AP Because this paper focuses on functions that are commutative both in the
 input and the output, let us introduce precisely what it means and how it
@@
 map that counts occurrences of every letter in the input word, namely: $
-\commute[w] \defined (a \mapsto \card[a]{w})$.
+\commute[w] \defined (a \mapsto \card[a]{w})$. Let $X$ be a set. A function $f
+\colon \Sigma^* \to X$ is \intro{commutative} whenever for all $u,v \in
+\Sigma^*$, $f(uv) = f(vu)$. Equivalently, it is \reintro{commutative} whenever
+there exists a map $g \colon \Nat^\Sigma \to X$ such that $g \circ \commute =
+f$.
 
-\AP Let $X$ be a set. A function $f \colon \Sigma^* \to X$ is
-\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
-Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
-\Nat^\Sigma \to X$ such that $g \circ \commute = f$.
-
-\begin{example}
-    \label{commutative-function:ex}
-    The map $f \colon w \mapsto \card[a]{w} \times \card[b]{w}$ is \kl{commutative}.
-\end{example}
-
-Let us briefly state that one can decide whether a \kl{$\Rel$-polyregular
-function} $f$ is \kl{commutative}, ensuring that we are working in a decidable
-subclass of functions.
+To ensure that we are working inside a decidable subclass of \kl{$\Rel$-rational series}
+let us state that commutativity is decidable. The proof is straightforward in the case
+of \kl{$\Rel$-polyregular functions}, and more involved for general \kl{$\Rel$-rational series}.
 
 \begin{lemma}
-    \label{decide-commutative:lemma}
-    Let $f \in \ZPoly$. One can decide if 
+    \label{decidable-commutative-poly:lemma}
+    \label{decidable-commutative-rat:lemma}
+    Let $f \in \ZRat$. One can decide if 
     $f$
     is \kl{commutative}.
-\end{lemma}
-\begin{proof}
-    Let $\Sigma$ be a finite alphabet, endowed with a
-    total ordering over its letters,
-    and let $f \colon \Sigma^* \to \Rel$ be a
-    \kl{$\Rel$-polyregular function}.
-    The map $\polysort \colon \Sigma^* \to \Sigma^*$
-    that sorts the letters in the input word $w$ according
-    to the chosen ordering is a \kl{polyregular function}
-    \cite[Proposition II.12]{LOPEZ23b}.
-    As a consequence,
-    $(f \circ \polysort) \in \ZPoly$.
-    Finally, $f$ is \kl{commutative} if and only
-    if $f = f \circ \polysort$, which is decidable
-    in the case of $\ZPoly$
-    \cite[Corollary II.24]{LOPEZ23b}, and more generally in the case 
-    of $\ZRat$ \cite[Corollary 3.6 p 38]{berstel2011noncommutative}.
-\end{proof}
-
-For the sake of completeness, let us also state that \kl{commutativity} is
-decidable for \kl{$\Rel$-rational series}, although using more involved
-arguments, since the latter are not closed under pre-composition by a
-\kl{polyregular function}.
-
-\begin{lemma}
-    \label{decidable-commutative-rat:lemma}
-    Let $f \in \ZRat$. One can decide if
-    $f$ is \kl{commutative}.
 \end{lemma}
@@
 
-\paragraph*{Polynomials.} \AP All polynomials considered in this paper have
+\subsection{Polynomials} \AP All polynomials considered in this paper have
 coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
@@
 \end{example}
-\begin{proof}
-    The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
-    Thus, 
-    $P(X) \defined X$ is
-    a \kl{$\Nat$-rational polynomial}. Similarly,
-    $w \mapsto |w|^2 + 3$ is a \kl{$\Nat$-polyregular function},
-    showing that $Q(X) \defined X^2 + 3$
-    is a \kl{$\Nat$-rational polynomial}.
-    Finally, 
-    $T(X) \defined - X$ cannot be 
-    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
-    are non-negative.
-\end{proof}
 
 \AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for
-all $n_1, \dots, n_k \geq 0$, $P(n_1, \dots, n_k) \geq 0$. Beware that we do
-not consider negative values as input, as the numbers $n_i$ will ultimately
-count the number of occurrences of a letter in a word. As an example, the
-polynomial $(X - Y)^2$ is \kl{non-negative}, and so is the polynomial $X^3$,
-but the polynomial $X^2 - 2X$ is not.
+all non-negative integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1,
+\dots, n_k)$ of the polynomial is non-negative. Beware that we do not consider
+negative values as input, as the numbers $n_i$ will ultimately count the number
+of occurrences of a letter in a word. As an example, the polynomial $(X - Y)^2$
+is \kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
+2X$ is not.
 

### 2024-02-12 parts/starfree.tex
big big rewrite

@@
 \label{beyond-polynomials:sec}
+\label{star-free:sec}
 
@@
     \kl{$\Rel$-polyregular function},
-    where $\Sigma = \set{a_1, \dots, a_n}$.
+    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
     There exists a computable
@@
 \end{lemma}
-\begin{proof}
-    We apply
-    \cref{n-poly-combinatorics:lem} to compute the $\omega$.
-    Using this $\omega$,
-    for any choice of
-    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
-    there exists a polynomial $P_r \in \Rel[X_1, \dots, X_p]$
-    such that $X_1, \dots, X_p \geq \omega$:
-    \begin{equation*}
-        f\left(
-        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
-        = P_r(X_1, \dots, X_p) \quad .
-    \end{equation*}
-    To get the full family of polynomials needed requires
-    just to only iterate subsets of letters, and fix the other ones
-    to be below the threshold $\omega^2$.
-\end{proof}
-
-\subsection{$\Nat$-polyregular functions}
 
@@
     for all $u_1, \dots, u_k \in \Sigma^*$,
-    there exists a polynomial $P$
-    such that $\translate{K}(P) \in \CorrectPoly$, and
+    there exists a polynomial $P \in \CorrectPoly$,  
     satisfying for all $X_1, \dots, X_k \geq \omega$:
@@
 \end{theorem}
-\begin{proof}
-    \textbf{TODO: we need to talk about the number of variables hehe}
-    Let us prove that
-    \cref{f-npoly-combi:item}
-    implies \cref{f-combinatorial:item}. 
-    We compute the $\omega$ by applying 
-    \cref{n-poly-combinatorics:lem}, and then
-    notice that the polynomial $P$ obtained is
-    by definition a \kl{$\Nat$-rational polynomial}.
-
-    For the converse implication, we use \cref{decompose-polynomial:lem} to
-    effectively compute a polynomial representation of $f$. Using the
-    assumption, all of these polynomials are actually \kl{$\Nat$-rational
-    polynomials}. Let us now construct a new way to compute $f$, by first
-    computing the \kl{$\omega$-type} of the input (which is possible in
-    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
-    which is possible in $\NPoly$.
-\end{proof}
-
-\subsection{Star-free $\Nat$-polyregular functions}
-\label{star-free:sec}
 
@@
 the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to prove the
-following theorem.
-
-\begin{theorem}
-    \label{zsf-npoly-nsf:thm}
-    Let $\Sigma$ be a finite alphabet and $k \in \Nat$.
-    Then,
-    \begin{equation*}
-        \ZCommut \cap \ZSF[k] \cap \NPoly
-        = \ZCommut \cap \NSF[k]
-        \quad .
-    \end{equation*}
-\end{theorem}
-
-As a corollary of \cref{zsf-npoly-nsf:thm} and the effective decision
-procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that $\NSF$ is
-decidable inside $\NPoly$ under the assumption of commutativity,
-and that the effective procedure can be used
-
-\begin{corollary}
-    One can decide if a \kl{commutative} \kl{$\Rel$-polyregular function}
-    can be realized by a \kl{star-free $\Nat$-polyregular function},
-    in which case the conversion is effective.
-\end{corollary}
-\begin{proof}
-    From \cite[Theorem V.13]{LOPEZ23b}, it is decidable whether
-    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
-    it is decidable whether $f \in \NPoly$.
-    Furthermore, one can enumerate functions in $\NSF[k]$,
-    and check equivalence to effectively represent $f$ as a
-    \kl{star-free $\Nat$-polyregular function}.
-\end{proof}
-
-The key ingredient of this section is the use of a semantic characterization of
-\kl{star-free $\Rel$-polyregular functions} among \kl{$\Rel$-rational series}
-that generalizes the aperiodicity of languages to functions by the means of
-polynomial behaviors.
+following theorem. As a corollary of \cref{zsf-npoly-nsf:thm} and the effective
+decision procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that
+$\NSF$ is decidable inside $\NPoly$ under the assumption of commutativity, and
+that the effective procedure can be used The key ingredient of this section is
+the use of a semantic characterization of \kl{star-free $\Rel$-polyregular
+functions} among \kl{$\Rel$-rational series} that generalizes the aperiodicity
+of languages to functions by the means of polynomial behaviors.
 
@@
 
-\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
-    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
-
-    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
-    only the empty word $\varepsilon$, and because $f$ is a
-    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
-    Note that $f$ is therefore a constant function, equal to a natural
-    number, hence belongs to $\NSF$.
-
-    \textbf{Induction.}
-    Because $f$ is \kl{ultimately polynomial},
-    there exists $N_0 \in \Nat$, 
-    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
-    such that
-    \begin{equation*}
-        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
-        = 
-        P(\seqof{X_a}{a \in \Sigma})
-        \quad
-        \text{ when }
-        \forall a \in \Sigma, X_a \geq N_0
-        \quad .
-    \end{equation*}
-    Furthermore, leveraging \cref{decompose-polynomial:lem},
-    we obtain $\omega \in \Nat$
-    and a family of polynomials 
-    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^\Sigma}$
-    that represent $f$.
-    For all $t \in \ModuloTypes[\omega]^\Sigma$
-    that have threshold greater than $\omega^2$,
-    we conclude that $P_t$ and $P$ coincide (up to translation) over an infinite grid,
-    hence we conclude that
-    if all letters appear at least $N_1 \defined \max (\omega^2, N_0)$ times,
-    then $f$ computes $P$. But this also shows that
-    $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
-    \kl{represented} by a function $f_1 \in \NSF[k]$.
-
-    To compute $f$ on words where some letter appears fewer than
-    $N_1$ times, we partially apply $f$ to obtain a function $g$
-    over a restricted input alphabet. Using the induction hypothesis,
-    $g$ belongs to $\NSF[k]$, and we can therefore (effectively) compute
-    a representation of $f$ in $\NSF[k]$.
-\end{proof}
+\begin{theorem}
+    \label{zsf-npoly-nsf:thm}
+    Let $\Sigma$ be a finite alphabet, $k \in \Nat$,
+    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
+    \kl{$\Nat$-polyregular function}.
+    Then, the following are equivalent:
+    \begin{itemize}
+        \item $f$ is \kl{ultimately polynomial},
+        \item $f \in \ZSF$,
+        \item $f \in \NSF[k]$.
+    \end{itemize}
+    Furthermore, membership is decidable and effective conversions
+    are possible.
+\end{theorem}

### 2024-02-12 parts/transducers.tex
big big rewrite

--- 2024-01-26 parts/transducers.tex
@@
 
-The goal of this section is to provide tools that do not require the assumption
-of \kl{commutativity}. To that end, we introduce the notion of \kl{$k$-residual
-transducer}, that generalizes the notion of residual transducer introduced for
-\kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}. This notion of residual
-transducer has its own interest, as it shifts the attention from (compatible)
-equivalence relations of finite index that are the classical tool in automata
-theory, towards (compatible) order relations that are \kl{well-quasi-ordered},
-which is the order counterpart of having finite index.
-
-While \kl{$k$-residual transducers} will provide canonical objects associated
-to functions in $\NPoly$, their effective construction requires the ability to
-decide $\NPoly$ inside $\ZPoly$, which is not currently known (except in the
-commutative case). Furthermore, the correspondence between \kl{counters} in
-this automaton and \kl{star-free $\Nat$-polyregular functions} is non-trivial,
-and cannot be used \emph{as is} to decide membership in $\NSF$
-based on this analysis of the \kl{$k$-residual transducer}.
-
-\AP The upcoming sections will heavily refer to the notion of \intro{residual} of
-a function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
-\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
-denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
-over words in $\Sigma^*$.
-
-\AP Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
+\AP The goal of this section is to provide tools that do not require the
+assumption of \kl{commutativity}. To that end, we introduce the notion of
+\kl{$k$-residual transducer}, that generalizes the notion of residual
+transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}.
+This notion of residual transducer has its own interest, as it shifts the
+attention from (compatible) equivalence relations of finite index that are the
+classical tool in automata theory, towards (compatible) order relations that
+are \kl{well-quasi-ordered}, which is the order counterpart of having finite
+index. Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
 quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
@@
 
-\subsection{Residual Transducers and WQOs}
-
-The goal of the upcoming section is to provide a notion of residual transducer
-for \kl{commutative} \kl{$\Nat$-polyregular functions}. The notion of
-transducer is based on the marble transducers \cite{EHB99}, and follows the
-general pattern of transducers introduced in \cite{LOPEZ23b}.
+The following definition of transducers is based on
+the marble transducers \cite{EHB99}, and follows the general pattern of
+transducers introduced in \cite{LOPEZ23b}.
 
@@
     Let $f \colon \set{a}^* \to \Nat$
-    be defined as follows
-    \begin{equation*}
-        f(a^n) \defined
-        \begin{cases}
-            1 & \text{ if } n = 0 \\
-            0 & \text{ if } n = 1 \\
-            n & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    The function is computed by the two automata
+    be defined as
+    $f(\varepsilon) = 1$,
+    $f(a) = 0$,
+    and $f(a^n) = n - 1$ for all $n \geq 2$.
+    \textbf{TODO: split the example}
+    $f(\varepsilon) = 1$,
+    $f(a) = 0$,
+    $f(a^2) = 1$,
+    and $f(a^n) = n - 3$ for all $n \geq 3$.
+    The function is \kl{computed} by the two automata
     in \cref{non-canonical-transd:fig}.
@@
     \centering
-    \begin{subfigure}[b]{0.9\linewidth}
+    \begin{subfigure}[b]{0.49\linewidth}
         \begin{tikzpicture}[
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.9\linewidth}
+    \begin{subfigure}[b]{0.49\linewidth}
         \begin{tikzpicture}[
@@
     \caption{Two \kl{$\NPoly[0]$-transducers}
-    computing \cref{non-canonical-transd:ex}.}
+    \kl{computing} the function of \cref{non-canonical-transd:ex}.}
     \label{non-canonical-transd:fig}
@@
 
-\AP As in the case of regular automata, we generalize the function $\delta
+\AP As in the case of finite automata, we generalize the function $\delta
 \colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
@@
 equations, we can write for all $q \in Q$, $u,v \in \Sigma^*$, $\aTransd(q, uv)
-= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice
-that if $\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q,
-u) \in \mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.
+= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice that if
+$\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q, u) \in
+\mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.
 
@@
 An automaton is \reintro{counter-free} if it contains no \reintro{counters}.
-This notion can be applied to \kl{$\mathcal{H}$-transducers} by
-considering counters of their underlying automata.
+This notion can be applied to \kl{$\mathcal{H}$-transducers} by considering
+counters of their underlying automata. It is already known in the literature
+that this notion of transducers correctly captures both $\NSF$ and $\NPoly$ in
+the sense that the following fact will clarify.
 
-It is already known in the literature that this notion of transducers correctly
-captures both $\NSF$ and $\NPoly$ in the sense that the following fact will
-clarify.
-
-\begin{fact}[Cite gaëtan]
+\begin{lemma}[{\cite{gaetanphd}}]
+    \label{transducer-nsf-npoly:lemma}
     Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
@@
         \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
-        \item $f$ is computed by an \kl{$\NPoly[k-1]$-transducer}
+        \item $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
             (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
@@
     \end{itemize}
-\end{fact}
+\end{lemma}
 
-\AP Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if and only if $(g - f)
-\in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if $(g - f) \in
-\NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we are going to
-study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f), \npolyleq[k])$. However,
-to simplify notations, instead of writing $\app{f}{u} \npolyleq[k] \app{f}{v}$
-when $u,v \in \Sigma^*$, we will use the convenient notation $u \resleq{f}{k}
-v$, and directly consider the space $(\Sigma^*, \resleq{f}{k})$.
+\AP The core concept of this section is that of a \intro{residual} of a
+function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
+\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
+denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
+over words in $\Sigma^*$. Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if
+and only if $(g - f) \in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if
+$(g - f) \in \NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we
+are going to study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f),
+\npolyleq[k])$. However, to simplify notations, instead of writing $\app{f}{u}
+\npolyleq[k] \app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient
+notation $u \resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
+\resleq{f}{k})$.
 
@@
 transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
-\resleq{f}{k} ua$. This naïve approach does not yield a canonical model. For
-instance, the function of \cref{non-canonical-transd:ex} can be computed in two
-ways if we only consider $\resleq{f}{k}$ minimal elements, as witnessed in
-\cref{non-canonical-transd:fig}. Even worse, the function $f$ belongs to
-$\NSF[0]$, but one of the two presented transducers has a \kl{counter}. This is
-why in \cref{residual-transducer:def}, we introduce the prefix ordering
-$\prefleq$ to further constrain how the model computes.
+\resleq{f}{k} ua$. This naïve approach does not yield a canonical model which
+is corrected by the following definition.
 
@@
     A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
-    is a \intro{k-residual transducer}
+    is a \intro{$k$-residual transducer}
     of $f$ 
@@
 is actually unique when it exists.
-
-\begin{fact}
-    \label{unique-res-transducer:fact}
-    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
-    Then $f$ has at most one \kl{$k$-residual transducer}.
-\end{fact}
-\begin{proof}
-    Let $\aTransd_1$ and $\aTransd_2$ be two
-    \kl{$k$-residual transducers} for $f$.
-    The two initial states must be $\varepsilon$.
-    Let us prove by induction on $u \in \Sigma^*$ that
-    $\delta_1(\varepsilon, u) = \delta_2(\varepsilon, u)$
-    and that $Q_1$ equals $Q_2$ over prefixes of $u$.
-    This will prove that 
-    $Q_1 = Q_2$, hence that $\aTransd_1 = \aTransd_2$.
-
-    Let $u \in \Sigma^* \cap Q_1 \cap Q_2$ and $a \in \Sigma$, $v_1 \in Q_1$ be
-    defined as $v \defined \delta_1(u,a)$, and $v_2 \defined \delta_2(u,a)$.
-    Remark that by induction hypothesis, for all $v \prefle u$, $v \in Q_1 \cap
-    Q_2$. If $\delta_1(u,a) = Q_1$, it means that for all $v \in Q_2$ such that
-    $v \prefle ua$, we have $\neg( v \resleq{f}{k} ua )$. The only possible
-    transition in $\aTransd_2$ is therefore $\delta_2(u,a) = ua$, and $ua \in
-    Q_2$. Similarly, if $\delta_1(u,a) \prefleq u$, then $\delta_2(u,a) =
-    \delta_1(u,a)$ by definition of $\delta_2$ as a maximum.
-\end{proof}
-
 We now introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
@@
 
-\begin{fact}
-    \label{q-o-prefix-cool:fact}
-    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
-    At each step of the \texttt{while loop}
-    of \cref{residual:algo}, the sets
-    $Q$ and $O$ are such that
-    \begin{enumerate}
-        \item $Q \cup O$ is a \kl{downwards closed} subset of 
-            $\Sigma^*$ for $\prefleq$;
-        \item elements in $O$ are pairwise incomparable
-            for $\prefleq$, and are maximal
-            for $\prefleq$ inside $Q \cup O$.
-    \end{enumerate}
-\end{fact}
-
 \begin{lemma}
@@
 \end{lemma}
-\begin{proof}
-    Because of \cref{q-o-prefix-cool:fact},
-    we already know that $q_0 = \varepsilon$,
-    $Q$ is a \kl{downwards closed} subset of $\Sigma^*$
-    for $\prefleq$, 
-    that every state of $Q$ is accessible from $q_0$.
-    Notice that at every step,
-    $\lambda(u,a)$ is defined as
-    $\app{f}{ua} - \app{f}{\delta(u,a)}$.
-    Finally, since $Q \cup O$ is a \kl{downwards closed} subset of $\Sigma^*$
-    at every step,
-    we have that at step $i$,
-    for all $ua \in O_i$,
-    $\setof{w \in Q}{w \prefleq ua} = \setof{w \in Q_i}{ w \prefleq ua}$,
-    which proves that the maximum considered in the algorithm
-    is indeed computing correctly.
-\end{proof}
 
@@
 section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
-\Rightarrow \resleq{f}{k})$ is \kl{well}, but $\resleq{f}{k}$ is a
+\Rightarrow \resleq{f}{k})$ is \kl{well}, but the quas-ordering $(\resleq{f}{k})$ is a
 \kl{well-quasi-ordering}. This has to be compared with the property of being
@@
 
-\begin{conjecture}
-    Let $f \in \ZPoly[k]$
-    be \kl{combinatorial}.
-    Then, $f$ is non-negative and for every pair
-    $(u,w) \in \Sigma^*$,
-    the sequence $\seqof{u w^n}{n \in \Nat}$
-    is \kl{good}.
-\end{conjecture}
-\begin{proof}
-    By induction on $k$.
-\end{proof}
-
-\begin{conjecture}
-    Let $f \in \ZPoly[k]$
-    be a non-negative function such that for every
-    $(u,w) \in \Sigma^*$,
-    the sequence $\seqof{u w^n}{n \in \Nat}$
-    is \kl{good}.
-    Then, $f$ belongs to $\NPoly[k]$.
-\end{conjecture}
-\begin{proof}
-    By induction on $k$.
-    For $k = 0$, non-negative is enough.
-    To go beyond, we
-
-\end{proof}
-
 \begin{theorem}
@@
     \begin{enumerate}
-        \item $f$ is \kl{computed}
+        \item \label{n-poly-1-transd:item} $f$ is \kl{computed}
             by a \kl{$\NPoly[k-1]$-transducer};
-        \item $f \in \NPoly[k]$;
-        \item $(\Sigma^*, \resleq{f}{k})$ is a
+        \item \label{n-poly-k:item} $f \in \NPoly[k]$;
+        \item \label{n-poly-wqo:item} $(\Sigma^*, \resleq{f}{k})$ is a
             \kl{well-quasi-ordering};
-        \item every $\prefleq$-increasing sequence
+        \item \label{n-poly-well:item} every $\prefleq$-increasing sequence
             of $\Sigma^*$  is a \kl{good sequence}
             for $\resleq{f}{k}$;
-        \item $f$ is \kl{computed} by a
+        \item \label{n-poly-residual:item} $f$ is \kl{computed} by a
             \kl{$k$-residual transducer}.
     \end{enumerate}
-    If $f$ is asumed to be \kl{commutative}, then all of the above
+    If $f$ is assumed to be \kl{commutative}, then all the above
     properties are decidable, and the conversions effective.
@@
 \begin{proof}
-    The first two items are known to be equivalent since 
-    the phd of gaëtan.
-    We have proven that transducers imply
-    wqo, and that wqo implies $k$-residual transducer.
+    \cref{n-poly-1-transd:item} implies \cref{n-poly-k:item} by
+    \cref{transducer-nsf-npoly:lemma}. Then,
+    \cref{n-poly-k:item} implies \cref{n-poly-wqo:item} by
+    \cref{n-poly-k-implies-wqo:lemma}.
+    The implication \cref{n-poly-wqo:item} $\Rightarrow$ \cref{n-poly-well:item}
+    is obvious.
+    Then, \cref{wqo-implies-termination:lemma} proves
+    that \cref{n-poly-well:item} implies \cref{n-poly-residual:item}.
+    Finally, because a \kl{$k$-residual transducer} is a \kl{$\NPoly[k-1]$-transducer},
+    \cref{n-poly-residual:item} implies \cref{n-poly-1-transd:item}.
 \end{proof}
-
-\subsection{Aperiodicity and Counters}
-
-\begin{lemma}
-    \label{aperiodic-iff-residual:lem}
-    Let $L$ be a regular language.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $L$ is aperiodic;
-        \item The \kl{$0$-residual transducer} of $\ind{L}$
-            is \kl{counter-free}.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    If the $0$-residual transducer is counter free,
-    then $L$ is aperiodic because the minimal automata is also counter-free.
-
-    % TODO
-    \textbf{TODO Claim:}
-    on every branch of the automata, the words have distinct residuals.
-
-    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
-    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
-    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
-    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
-    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
-    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
-    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
-    w_2) = t$.
-
-    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
-    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
-    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
-    strict prefix of $s_0$. It must exist, because otherwise the two are always
-    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
-    this contradicts the minimality of $u$.
-
-    We have proven that $s_0$ and $s_1$ are comparable, hence they
-    are equal.
-    Finally, we have proven that $\delta(q, w_1) = s_0$,
-    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
-
-    As a consequence $t$ and $q$ were comparable for the prefix relation,
-    hence equal, and therefore $\delta(q, w) = q$.
-\end{proof}
-
-\begin{lemma}
-    Let $\NSF[0] = \NPoly \cap \NSF[0]$,
-    and elements of $\NSF[0]$
-    have a \kl{counter-free} \kl{$0$-residual transducer}.
-\end{lemma}
-\begin{proof}
-    TODO.
-\end{proof}
-
-\textbf{TODO:} one may think that we can ask that $u \resleq{f}{k} u w^n$
-implies $u \resleq{f}{k} uw$, but it is false, as the following example will
-demonstrate. This is why the definition of counter is more involved in our
-case.
-
-\begin{example}
-    Let us consider $f(w) \defined (\card{w} - 3)^2$, 
-    which is a \kl{star-free $\Nat$-polyregular function}.
-    Then,
-    $\app{f}{a} - f$ is not in $\NPoly$,
-    but
-    $\app{f}{a^10} - f$ is in $\NPoly$,
-    hence there is a non-trivial counter.
-\end{example}
-\begin{proof}
-    Remark that $\app{f}{a} = (\card{w} - 2)^2$,
-    hence that $\app{f}{a}- f = 2\card{w} - 5$.
-    Similarly, $\app{f}{a^10} - f = 20\card{w} + k$ where $k \geq 0$.
-
-    Finally,
-    let $n > 1$ be such that
-    $\app{f}{a^n} - f$ belong to $\NPoly$.
-    In this specific example,
-    $\app{f}{a^n} = (\card{w} + n -3)^2$,
-    and $\app{f}{a^n} - f = (2 \card{w} + n - 6)n$.
-    The latter belongs to $\NPoly$ if and only if
-    $n \geq 6$. Let $n \geq 6$, 
-    remark that
-    $\app{f}{a^n} - \app{f}{a^{n-1}}
-    = 2 \card{w} + 2n - 7$. Because $n = n_0 + 6$ with $n_0 \geq 0$,
-    we can rewrite the latter 
-    as $2 \card{w} + 2n_0 + 5$, which belongs to $\NPoly$.
-\end{proof}
-
-\begin{lemma}
-    \label{sf-no-periods-on-sequences:lemma}
-    Let $f \colon \Sigma^* \to \Nat$ be a \kl{star-free $\Nat$-polyregular
-    function}. Then,
-    for all $u, w \in \Sigma^*$,
-    there exists $n \in \Nat$ such that
-    $u w^n \resleq{f}{k} u w^{n+1}$.
-\end{lemma}
-\begin{proof}[Proof sketch.]
-    Let $t$ be a first order type,
-    the map $g_t \colon X \mapsto \vcount{t}(uw^X)$ is a
-    \kl{commutative} \kl{star-free $\Nat$-polyregular function},
-    hence ultimately a polynomial, and therefore,
-    translated enough, it is a $\Nat[X]$ polynomial.
-    But this means that $g_t(X+1) - g_t(X) \in \Nat[X]$ for a large
-    enough $X$, which is precisely what suffices.
-    Now, we can do this for every first order type $t$
-    with fewer variables and lower quantifier rank than those
-    needed to define $f$, and conclude.
-\end{proof}
-
-Even though computability is not guaranteed, we have the following proof
-scheme to decide aperiodicity of \kl{$\Nat$-polyregular functions}.
-\begin{lemma}
-    Let $f \in \NPoly[k]$ be \kl{ultimately polynomial},
-    then every \kl{$k$-residual transducer} computing 
-    $f$ has no counters, and has labels
-    that are themselves \kl{ultimately polynomial}
-    \kl{$\Nat$-polyregular functions}.
-\end{lemma}
-\begin{proof}
-
-    We prove the result by induction on $k$. By definition
-    of a residual transducer, the labels are \kl{ultimately polynomial}.
-    Now, assume that there exists a counter
-    in some \kl{$k$-residual transducer}
-    $\aTransd \defined (Q, q_0, \delta, \lambda, F)$ of $f$. That is, there exists $\alpha \in \Sigma^*$,
-    $w \in \Sigma^+$, and $n > 1$, such that
-    $\delta(q_0,\alpha w^n) = \delta(q_0, \alpha)$.
-    We want to prove that
-    $\delta(q_0, \alpha w) = \delta(q_0, \alpha)$.
-    Because $f$ is \kl{ultimately polynomial},
-    it is in particular in $\ZSF[k]$, and therefore
-    $g \defined \app{f}{\alpha w} - \app{f}{\alpha}$
-    belongs to $\ZSF[k-1]$.
-    It remains to be proven that this difference is a function in
-    $\NPoly$.
-
-    To that end, notice that the map $f(\alpha w^X \mathcal{M})$ is a
-    \kl{commutative} \kl{star-free $\Nat$-polyregular function}. For a large
-    enough $N_0$, we conclude that $f(\alpha w^{N_0+1} \mathcal{M}) - f(\alpha
-    w^{N_0} \mathcal{M})$ belongs to $\NPoly$, hence that $\alpha w^{N_0}
-    \resleq{f}{k-1} \alpha w^{N_0 + 1}$. Does that prove that $\alpha
-    \resleq{f}{k-1} \alpha w$? Not at all and we cannot conclude.
-\end{proof}
-
-\begin{theorem}
-    The following are equivalent for $f \in \NPoly[k]$.
-    \begin{itemize}
-        \item The function $f \in \NSF[k]$,
-        \item $f$ is \kl{ultimately polynomial}.
-        \item Every \kl{$k$-residual transducer} of $f$ has no counters,
-        \item $f$ is \kl{computed} by some
-            \kl{$\NSF[k-1]$-transducer}.
-    \end{itemize}
-\end{theorem}
-\begin{proof}
-    The only implication to prove is the first implies the last.
-    We know that the automaton is counter free, so the languages of the states
-    are star-free. Now, we just write 
-    \begin{equation*}
-        TODO
-    \end{equation*}
-    and we have finished the proof.
-\end{proof}
-
-Remark that the theorem is effective because being \kl{ultimately polynomial}
-is an effective property! However, effective conversions are not possible as of
-now, because the $k$-residual transducer algorithm uses an oracle to decide if
-a function is in $\NPoly$.
-
-\begin{lemma}
-    \label{trivial-for-unary:lemma}
-    Let $f \in \NPoly$ having unary input.
-    In particular, $f$ is commutative.
-    Then, $f \in \NSF$ if and only if 
-    the \kl{residual transducer} of $f$
-    is \kl{counter-free}.
-\end{lemma}
-\begin{proof}
-    If there is a counter... well we will find it ^^
-\end{proof}
-
-\begin{conjecture}
-    If $f$ is commutative, and the \kl{residual transducer}
-    has a counter, then it has a single letter counter.
-\end{conjecture}
-
-\begin{conjecture}
-    If $f$ is commutative, then 
-    $f \in \NSF$ if and only if the \kl{residual transducer}
-    of $f$ is \kl{counter-free}.
-\end{conjecture}
-
-\section{Trash polynomials}
-
-\begin{definition}
-    For every polynomial $P \in \Rel[X \vec{Y}]$, we define 
-    the partial derivative with respect to $X$ as follows:
-    \begin{equation*}
-        \partial_X P \defined P(X + 1, \vec{Y}) - P(X, \vec{Y}) \quad .
-    \end{equation*}
-\end{definition}
-
-\begin{fact}
-    Let $P \in \CorrectPoly[X_1, \dots, X_n]$, then
-    for all $1 \leq i \leq n$,
-    $\partial_{X_i} P \in \CorrectPoly$.
-\end{fact}
-\begin{proof}
-    We prove the result by induction on the number of variables.
-    Let $\nu \colon \vec{X} \topartial \Nat$.
-    Two cases, either $X_i \in \dom(\nu)$,
-    in which case, we remark that
-    we can split $\nu$ into two valuations,
-    and that coefficients are good.
-
-    Otherwise, evaluation commutes with partial,
-    and we conclude by induction hypothesis.
-\end{proof}
-
-\begin{remark}
-    The converse does not hold,
-    for instance $P(X) \defined X - 1$ has all its derivatives
-    that are in $\CorrectPoly$, but is not itself in $\CorrectPoly$.
-\end{remark}
-
-\begin{fact}
-    Let $P \in \Nat[X,\vec{Y}]$, and $K,L \in \Nat$,
-    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \Nat[X, \vec{Y}]$
-    whenever $L \geq K$.
-\end{fact}
-\begin{proof}
-    We can write
-    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
-    Now,
-    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
-    is precisely
-    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$,
-    remaining with positive coefficients.
-\end{proof}
-
-\begin{fact}
-    Let $P \in \CorrectPoly[X,\vec{Y}]$, and $K,L \in \Nat$,
-    then $P(X+L, \vec{Y}) - P(X+K,\vec{Y}) \in \CorrectPoly$
-    whenever $L \geq K$.
-\end{fact}
-\begin{proof}
-    We can write
-    $P = \sum_{i = 0}^n P_i(\vec{Y}) X^i$, with $P_i \in \Nat[\vec{Y}]$.
-    Now,
-    $P(X+L,\vec{Y}) - P(X+K,\vec{Y})$
-    is precisely
-    $\sum_{i = 1}^n P_i(\vec{Y}) (L-K) \left[ \sum_{j + l + 1= i} (X+L)^j (X+K)^l \right]$.
-    Now, $P_i(\vec{Y})$
-    remaining with positive coefficients.
-\end{proof}
-\begin{fact}
-    For every $K \in \Nat$, and every $n \in \Nat$,
-    $\Nat^n$ is a \kl{well-quasi-ordering}
-    when endowed with the quasi-ordering $\vec{u} \leq_K \vec{v}$
-    if and only if $\vec{u} = \vec{v}$ or
-    $\vec{u}$ is smaller than $\vec{v}$ for the product ordering and
-    $\sum_{i = 1}^n  (\vec{v} - \vec{u})_i \geq K$.
-\end{fact}
-\begin{proof}
-    First, $\Nat^n$ with the product ordering is a \kl{well-quasi-ordering}.
-    Therefore, given a sequence $\seqof{\vec{u}_i}{i \in \Nat}$,
-    one can extract an infinite monotone subsequence for
-    the product ordering.
-    Either there exists $i < j$ such that $\vec{u}_i = \vec{u}_j$
-    in the subsequence, in which case $\vec{u}_i \leq \vec{u}_j$
-    and the original sequence was good. Or for all $i < j$,
-    we have $\vec{u}_i$ that is strictly smaller than 
-    $\vec{u}_j$ for the product ordering. As a consequence,
-    $\vec{u}_0 \leq \vec{u}_K$.
-\end{proof}
-
-\begin{fact}
-    Let $P \in \CorrectPoly[X_1, \dots, X_n]$. 
-    For all
-    $\vec{u} \leq \vec{v} \in \Nat^n$,
-    $Q(X_1, \dots, X_n)
-    \defined 
-    P(X_1 + v_1, \dots, X_n + v_n)
-    -
-    P(X_1 + u_1, \dots, X_n + u_n)$
-    belongs to $\CorrectPoly[X_1, \dots, X_n]$.
-\end{fact}
-\begin{proof}
-    Let us consider $K \in \Nat$ such that for all partial valuations
-    $\nu \colon \vec{X} \topartial \Nat$,
-    $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$.
-
-    Remark that
-    \begin{align*}
-        Q
-        &= 
-        P(X_1 + v_1, \dots, X_n + v_n)
-        -
-        P(X_1 + u_1, \dots, X_n + u_n) \\
-        &=
-        P(X_1 + v_1, X_2 + v_2 , \dots, X_n + v_n) - P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) \\
-        &+
-        P(X_1 + u_1, X_2 + v_2, \dots, X_n + v_n) - P(X_1 + u_1, X_2 + u_2, \dots, X_n + v_n) \\
-        &+ \dots
-    \end{align*}
-
-    Let $\nu \colon \vec{X} \topartial \Nat$
-    be a partial valuation, let us show that
-    $\translate{K}(\restr{Q}{\nu}) \in \Nat[\vec{X}]$.
-
-    \begin{align*}
-        \translate{K}(\restr{Q}{\nu})
-        &= 
-        \translate{K}(\restr{P(X_1 + v_1, \dots, X_n + v_n)}{\nu})
-        -
-        \translate{K}(\restr{P(X_1 + u_1, \dots, X_n + u_n)}{\nu})
-    \end{align*}
-\end{proof}
-
-\begin{definition}
-    Residual $\Nat$-transducer.
-\end{definition}
-
-\begin{lemma}
-    If $P \in \CorrectPoly$, then it is computable
-    by a $\Nat$-residual transducer.
-\end{lemma}
-
-\begin{theorem}
-    The following are equivalent for \kl{commutative}.
-    \begin{itemize}
-        \item $f$ is in $\NSF$,
-        \item The residual automata of $f$ has no counters
-    \end{itemize}
-\end{theorem}
-
-\begin{fact}[Folklore about regular languages]
-    \label{regular:fact}
-    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
-    is not regular, whenever $a,b \in \Sigma$.
-    Regular languages are closed under intersection.
-\end{fact}

### 2024-02-12 parts/trash.tex
keeping trash

\subsection{Polyblind functions}

We are going to prove that polyblind functions are essentially
commutative.

\AP Given $q,k \in \Nat$, we define $\MSO^q_k$ to be the collection of types of
tuples of at most $k$ variables using quantifier rank at most $q$. We extend
the notation to a function $\MSO^q_k (u)$ that assigns to a word $u$ the number
of representative for each type.

\begin{lemma}
    Let $q \in \Nat$. The image
    $\MSO^q_1(\Sigma^*)$ is a semilinear set.
\end{lemma}
\begin{proof}
    Here is a context-free grammar that enumerates words
    in $\Sigma^*$ with letters colored by their $\MSO^q_1$ type.
    The initial non-terminal decides the global $\MSO^q$ type of the word,
    and given a non-terminal with an $\MSO^q$-type, we can split it in three
    by choosing a letter $a \in \Sigma$, an $\MSO^q$-type for the left,
    and an $\MSO^q$-type for the right, so that the types are coherent.
    We just have to remember the $\MSO^q$ type of the left and right
    of the current non-terminal to do that properly.

    By the Parikh’s theorem, the image of this is a semilinear set, that
    precisely computes $\MSO^q_1(\Sigma^*)$.
\end{proof}

\AP
Recall that a \intro{polyblind function} is a function such that.

\begin{lemma}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{polyblind function}.
    Then, there exists a computable function $g \colon \Nat^p \to \Rel$,
    that is \kl{commutative} and such that the following are equivalent:
    \begin{itemize}
        \item $g$ is $\Nat$-polyregular
        \item $f$ is $\Nat$-polyregular
    \end{itemize}
\end{lemma}

\AP
Look at the image of $\MSO^q_k$, is it semi-algebraic instead of semi-linear?
\begin{lemma}
    $\MSO^q_k(\Sigma^*)$ is a set of tuples that is obtained as a finite union 
    images of polynomials?
\end{lemma}

\begin{center}
    \begin{tabular}{c|cc}
        \toprule
        & \textbf{Commutative} & \textbf{General} \\
        \midrule
        \textbf{Commutative}
        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
        \textbf{General} & ? & polyregular 
    \end{tabular}
\end{center}

\begin{conjecture}
    For commutative functions, the residual transducer 
    is helpful.
\end{conjecture}

\begin{fact}
    \label{translation-invariance:fact}
    The following sets are stable under \kl{translations}
    $\translate{K}$,
    where $K \in \Nat$ is non-negative:
    \begin{itemize}
        \item $\CorrectPoly$,
        \item \kl{$\Nat$-rational polynomials},
        \item Polynomials \kl{represented}
            by \kl{star-free $\Nat$-polyregular functions}.
    \end{itemize}
\end{fact}

\begin{conjecture}
    Let $f \in \NPoly$. The following are equivalent,
    \begin{enumerate}
        \item $f \in \ZSF$,
        \item the \kl{$k$-residual transducer} of $f$
            is \kl{counter-free},
            and its labels belong to $\ZSF$.
    \end{enumerate}
\end{conjecture}
\begin{proof}
    If it has a counter free residual transducer recursively, then
    it clearly belongs to $\ZSF$.

    % TODO: mix orderings and trees and maximal elements :(

    Now, assume that $L$ is aperiodic, and let $(q,w^n)$ be a counter with $n
    \geq 1$. We know that $qw \equiv_L q$ because the language $L$ is
    aperiodic. Let us write $t \defined \delta(q,w)$, we know that $t \equiv_L
    q$. Assume by contradiction that $t$ and $q$ are incomparable for the
    prefix relation. Let us split $w = w_1 w_2$ where $w_1$ is the shortest
    prefix of $w$ such that $s_0 \defined \delta(q,w_1)$ is an ancestor of $q$
    and of $t$ for the prefix relation, it must exist because $\delta(q,w_1
    w_2) = t$.

    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
    the smallest $v$ for the prefix relation such that $\delta(t, v)$ is a
    strict prefix of $s_0$. It must exist, because otherwise the two are always
    comparable. Since $t \equiv_L q$, we know that $tv \equiv_L qv$, but then,
    this contradicts the minimality of $u$.

    We have proven that $s_0$ and $s_1$ are comparable, hence they
    are equal.
    Finally, we have proven that $\delta(q, w_1) = s_0$,
    $\delta(s_0, w_2) = t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.

    As a consequence $t$ and $q$ were comparable for the prefix relation,
    hence equal, and therefore $\delta(q, w) = q$.
\end{proof}

We do not know if \kl{commutativity} is decidable for more powerful models of
computation, and conjecture that it is decidable for \kl{polyregular functions}
in general.

\begin{conjecture}
    Let $f \in \Poly$. One can decide if $f$ is \kl{commutative}.
\end{conjecture}

The goal is to prove \cref{decide-n-poly-general:thm}.
To that end, we will use factorisation forests.
\begin{definition}
    Forests of depth $d$ by induction.
\end{definition}

\begin{remark}
    Place the nested higman embedding on the forests, it is a well-quasi-ordering.
\end{remark}

\begin{definition}
    Counting functions. $\vcount{q}$ where $q$ is a pattern.
\end{definition}

\begin{definition}
    Divisibility of patterns.
\end{definition}

\begin{definition}
    Forest iteration, where $q$ is a pattern.
\end{definition}

\begin{theorem}[Simon, restated in ]
    For every $f \in \ZPoly[k]$,
    there exists a computable function $h \colon \Sigma^* \to F_d(M)$,
    and a function $g \colon F_d(M) \to \Rel$ that is a counting function
    such that.
\end{theorem}

\textbf{TODO: domains should have a good shape, i.e.,
    intersections of sentences of the shape
    this pattern has at most "X" matches, and should be
    downwards closed}.

\textbf{TODO: problem, there is a problem, it is not because
    the number of matches is unbounded, that we can
    iterate the shape itself no?}

\begin{lemma}
    A function $g \colon F_d(M) \to \Rel$ is $\Nat$-polyregular
    over a domain $D$ if and only if for every 
    pattern iteration, the function obtained is
    $\Nat$-polyregular (commutative).
\end{lemma}
\begin{proof}
    By induction on the domain $D$, and on the divisibility ordering
    of patterns.
    Base case, for all pattern iteration, the value of $g$ is bounded.
    Then, $g$ has finitely many outputs, and they are all in $\Nat$,
    we have concluded.

    Let $q_i$ be a maximal pattern of $g$. If $q_i$ cannot be iterated in $D$,
    then the value of $q_i$ is bounded by some constant, and we can rewrite it
    as counting "full" patterns?

    Otherwise, $q_i$ can be iterated, and therefore we can
    pump to prove that its coefficient must be strictly positive.
    Then, using some derivation magic, we can recurse.
\end{proof}

\begin{theorem}
    \label{decide-n-poly-general:thm}
    Let $f \in \ZPoly[k]$, the following are equivalent:
    \begin{enumerate}
        \item $f \in \NPoly$;
        \item $f \in \NPoly[k]$;
        \item for all $h$ commutative polyregular function,
            $f \circ h$ in $\NPoly$
        \item for all pattern iteration,
            $f(pattern)$ is a \kl{$\Nat$-rational polynomial}.
    \end{enumerate}
    Furthermore, the properties are decidable.
\end{theorem}
\begin{proof}
    (1) implies (2) is okay, because of generic results.
    (2) implies (3) implies (4) is completely fine.
    Now, let us prove that (4) implies $1$.
\end{proof}

\begin{conjecture}
    Let $q,k \in \Nat$.
    Associate to a word $w$ the number of representatives
    of every $\MSO(q,k)$ type. 
    The image of this map is a semilinear set.
    In particular, it has a simple basis.
\end{conjecture}

\begin{conjecture}
    Let $f \in \ZPoly[k]$
    be \kl{combinatorial}.
    Then, $f$ is non-negative and for every pair
    $(u,w) \in \Sigma^*$,
    the sequence $\seqof{u w^n}{n \in \Nat}$
    is \kl{good}.
\end{conjecture}
\begin{proof}
    By induction on $k$.
\end{proof}

\begin{conjecture}
    Let $f \in \ZPoly[k]$
    be a non-negative function such that for every
    $(u,w) \in \Sigma^*$,
    the sequence $\seqof{u w^n}{n \in \Nat}$
    is \kl{good}.
    Then, $f$ belongs to $\NPoly[k]$.
\end{conjecture}
\begin{proof}
    By induction on $k$.
    For $k = 0$, non-negative is enough.
    To go beyond, we

\end{proof}

\begin{fact}[Folklore about regular languages]
    \label{regular:fact}
    The language $\setof{ w \in \Sigma}{ \card[a]{w} = \card[b]{w}}$
    is not regular, whenever $a,b \in \Sigma$.
    Regular languages are closed under intersection.
\end{fact}

### 2024-02-14 globals/metadata.tex
uploading to ICALP

@@
     \begin{abstract}
-        This paper answers two questions about 
+        This paper addresses two questions about 
         commutative unary output polyregular functions, called
@@
         in the case of polynomials.
-
         We also prove that it is decidable whether a commutative $\Nat$-polyregular

### 2024-02-14 parts/appendix.tex
uploading to ICALP

@@
     \cite{HROUPOWO18}.
-\end{proof}
-
-\begin{proof}[Proof of
-    \cref{pre-compose-growth-commut:lemma}]
-    If $f \in \ZPoly[k]$, then 
-    for every \kl{polyregular function} $g \in \Poly[\ell]$,
-    $f \circ g \in \ZPoly[k+ \ell - 1]$
-    \cite{LOPEZ23b}.
-
-    Conversely, it was proven in \cite[Theorem III.3]{LOPEZ23b}
-    that $f \in \ZPoly[k]$ if and only if
-    for all $\alpha_0, \cdots, \alpha_k$,
-    for all $w_1, \dots, w_k$,
-    we have 
-    \begin{equation}
-        \label{k-pumpable:eqn}
-        f\left(
-            \alpha_0 \prod_{i = 1}^k w_i^{X_i} \alpha_i
-        \right)
-        = \bigO( (X_1 + \cdots X_k)^{k} )
-        \quad .
-    \end{equation}
-    Or, the \kl{commutative} \kl{star-free polyregular function} 
-    $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
-    a word $u$
-    to $\alpha \prod_{i = 1}^k w_i^{\card[i]{u}}$
-    has \kl{growth rate} $1$.
-    Hence, 
-    $f \circ h \in \ZPoly[k]$, 
-    and we indeed conclude that 
-    \cref{k-pumpable:eqn}
-    holds, i.e. that $f \in \ZPoly[k]$.
-\end{proof}
-
-\begin{proof}[Proof of
-    \cref{pre-compose-sf-commut:lemma}]
-    If $f \in \ZSF$, then for all $h \in \SF$,
-    $f \circ h \in \ZSF$, and \emph{a fortiori}
-    for \kl{commutative} functions $h$.
-
-    Conversely, assume that $f \circ h \in \ZSF$
-    for all \kl{commutative} functions $h \in \SF$.
-    Using \cite[Theorem V.13]{LOPEZ23b},
-    to conclude that $f \in \ZSF$,
-    it suffices to prove that
-    for all $\alpha_0, \cdots, \alpha_k \in \Sigma^*$,
-    for all $w_1, \dots, w_k \in \Sigma^*$,
-    there exists a polynomial $P \in \Rel[X_1, \dots,X_k]$
-    and a constant $N_0 \in \Nat$,
-    such that if $X_1, \dots, X_k \geq N_0$:
-    \begin{equation}
-        \label{ultimate-polynomial:eqn}
-        f\left(
-            \alpha_0 \prod_{i = 1}^k w_i^{X_i} \alpha_i
-        \right)
-        = P(X_1, \dots, X_k)
-        \quad .
-    \end{equation}
-    Let us consider
-    the \kl{commutative} \kl{star-free polyregular function}
-    $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
-    a word $u$
-    to $\alpha \prod_{i = 1}^k w_i^{\card[i]{u}}$.
-    We know that
-    $f \circ h \in \ZSF$, hence, that 
-    there exists $Q \in \Rel[X_1, \dots, X_n]$
-    and $M_0 \in \Nat$,
-    such that for all $X_1, \dots, X_n \geq N_0$,
-    $f \circ h( \prod_{i = 1}^k \underline{i}^{X_i}) = Q(X_1, \dots, X_n)$.
-    In particular,
-    we can take $N_0 = M_0$, and $P = Q$ to conclude that
-    \cref{ultimate-polynomial:eqn} holds, hence, that
-    $f \in \ZSF$.
 \end{proof}
@@
     \cref{d-t-transl-fin:item} and \cref{d-t-transl:item} are equivalent.
-\end{proof}
 
-\begin{proof}[Proof of \cref{decidability-correct:lem}]
-    Given $P \in \Rel[\vec{X}]$,
+    For the effective decision procedure,
+    given $P \in \Rel[\vec{X}]$,
     one applies \cref{derivation-translation:lem}
@@
 \begin{proof}[Proof of \cref{decidable-n-poly:thm}]
-    \textbf{TODO: we need to talk about the number of variables hehe}
     Let us prove that
@@
 \begin{proof}[Proof sketch of \cref{sf-no-periods-on-sequences:lemma}]
-    Let $t$ be a first order type,
-    the map $g_t \colon X \mapsto \vcount{t}(uw^X)$ is a
-    \kl{commutative} \kl{star-free $\Nat$-polyregular function},
-    hence ultimately a polynomial, and therefore,
-    translated enough, it is a $\Nat[X]$ polynomial.
-    But this means that $g_t(X+1) - g_t(X) \in \Nat[X]$ for a large
-    enough $X$, which is precisely what suffices.
-    Now, we can do this for every first order type $t$
-    with fewer variables and lower quantifier rank than those
-    needed to define $f$, and conclude.
+    Let $f \in \NSF[k]$, $q \in \Nat$, and 
+    write $f = \sum_{i=1}^n m_i \times \vcount{\phi_i(\vec{x})}$, where
+    $\phi_i \in \FO$ has quantifier rank at most $q$
+    and $\card{\vec{x}} = k$.
+
+    As in the proof of \cref{n-poly-k-implies-wqo:lemma}, we are going to
+    assign a tuple of integers to a word $u \in \Sigma^*$ by counting the
+    number of realizations of each $\FO^q$ type of at most $k$ variables in
+    $u$. To that end, let us write $\FO^q(u)$ this vector.
+
+    First, let us notice that it suffices to prove that for some $n \in \Nat$,
+    $\FO^q(uw^n) \leq \FO^q(uw^{n+1})$ to conclude, since $\app{f}{uw^{n+1}} -
+    \app{f}{uw^n}$ will be obtained as a positive combination of counting
+    first-order types in the argument.
+
+    Let $t$ be a first order type with at most $k$ free variables and
+    quantifier rank at most $q$. The map $g_t \colon X \mapsto
+    \vcount{t}(uw^X)$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular
+    function}. As a consequence, there exists $N_0$ and a polynomial $P \in
+    \CorrectPoly$ such that for all $X \geq N_0$, $g_t(X) = P(X)$. Now, because
+    $P \in \CorrectPoly$, there exists a $K \in \Nat$ such that
+    $\translate{K}(\Diff{1}{P}) \in \CorrectPoly$, and in particular $g_t(X +
+    N_0 + K+1) - g_t(X + N_0 + K) \in \Nat$, for all $X \geq 0$.
+
+    Because there are finitely many non-equivalent $\FO^q$ types with at most 
+    $k$ free variables, we can take the maximum of the $K$'s obtained for each 
+    of those, and conclude.
 \end{proof}
+
+\begin{proof}[Proof of
+    \cref{pre-compose-growth-commut:lemma}]
+    If $f \in \ZPoly[k]$, then 
+    for every \kl{polyregular function} $g \in \Poly[\ell]$,
+    $f \circ g \in \ZPoly[k+ \ell - 1]$
+    \cite{LOPEZ23b}.
+
+    Conversely, it was proven in \cite[Theorem III.3]{LOPEZ23b}
+    that $f \in \ZPoly[k]$ if and only if
+    for all $\alpha_0, \cdots, \alpha_k$,
+    for all $w_1, \dots, w_k$,
+    we have 
+    \begin{equation}
+        \label{k-pumpable:eqn}
+        f\left(
+            \alpha_0 \prod_{i = 1}^k w_i^{X_i} \alpha_i
+        \right)
+        = \bigO( (X_1 + \cdots X_k)^{k} )
+        \quad .
+    \end{equation}
+    Or, the \kl{commutative} \kl{star-free polyregular function} 
+    $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
+    a word $u$
+    to $\alpha \prod_{i = 1}^k w_i^{\card[i]{u}}$
+    has \kl{growth rate} $1$.
+    Hence, 
+    $f \circ h \in \ZPoly[k]$, 
+    and we indeed conclude that 
+    \cref{k-pumpable:eqn}
+    holds, i.e. that $f \in \ZPoly[k]$.
+\end{proof}
+
+\begin{proof}[Proof of
+    \cref{pre-compose-sf-commut:lemma}]
+    If $f \in \ZSF$, then for all $h \in \SF$,
+    $f \circ h \in \ZSF$, and \emph{a fortiori}
+    for \kl{commutative} functions $h$.
+
+    Conversely, assume that $f \circ h \in \ZSF$
+    for all \kl{commutative} functions $h \in \SF$.
+    Using \cite[Theorem V.13]{LOPEZ23b},
+    to conclude that $f \in \ZSF$,
+    it suffices to prove that
+    for all $\alpha_0, \cdots, \alpha_k \in \Sigma^*$,
+    for all $w_1, \dots, w_k \in \Sigma^*$,
+    there exists a polynomial $P \in \Rel[X_1, \dots,X_k]$
+    and a constant $N_0 \in \Nat$,
+    such that if $X_1, \dots, X_k \geq N_0$:
+    \begin{equation}
+        \label{ultimate-polynomial:eqn}
+        f\left(
+            \alpha_0 \prod_{i = 1}^k w_i^{X_i} \alpha_i
+        \right)
+        = P(X_1, \dots, X_k)
+        \quad .
+    \end{equation}
+    Let us consider
+    the \kl{commutative} \kl{star-free polyregular function}
+    $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
+    a word $u$
+    to $\alpha \prod_{i = 1}^k w_i^{\card[i]{u}}$.
+    We know that
+    $f \circ h \in \ZSF$, hence, that 
+    there exists $Q \in \Rel[X_1, \dots, X_n]$
+    and $M_0 \in \Nat$,
+    such that for all $X_1, \dots, X_n \geq N_0$,
+    $f \circ h( \prod_{i = 1}^k \underline{i}^{X_i}) = Q(X_1, \dots, X_n)$.
+    In particular,
+    we can take $N_0 = M_0$, and $P = Q$ to conclude that
+    \cref{ultimate-polynomial:eqn} holds, hence, that
+    $f \in \ZSF$.
+\end{proof}
+
+\begin{proof}[Proof of \cref{z-poly-commutative-encoding:remark}]
+    We are going to prove that if $f \in \ZPoly[k]$ is polyblind, 
+    then the conjecture \cref{z-poly-commutative-encoding:conj} holds when allowing 
+    for non polyregular functions $\mathsf{enc}$ and $\mathsf{dec}$.
+
+    Let us consider $f \in \ZPoly[k]$ that is polyblind.
+    Using \cite[Theorem 6.12]{gaetanphd},
+    we know that there exists a family of $\MSO$ formulas
+    $\seqof{\psi_{i,j}(x)}{1 \leq i,j \leq n}$,
+    and coefficients $\seqof{z_i}{1 \leq i \leq n}$,
+    such that:
+    \begin{equation*}
+        f = \sum_{i = 1}^n z_i \times \prod_{j = 1}^n \vcount{\psi_{ij}(x)}
+        \quad .
+    \end{equation*}
+
+    Let us now consider a monoid $M$ that recognizes all formulas $\psi_{ij}$
+    for $1 \leq i,j \leq n$ with a morphism $\mu \colon \Sigma^* \to M$. Let us
+    consider the map $\mathsf{t} \colon \Sigma^* \to (M^3)^*$, that associates
+    to a word $w \in \Sigma^*$ the word $\mathsf{t}(w) \defined \prod_{i =
+    1}^{|w|} ( \mu(w_{< i}), \mu(w_i), \mu(w_{> i}))$. That is, to every letter
+    one associates its left and right context in the monoid.
+
+    Let us define $\mathsf{s} \colon \Sigma^* \to (\Nat)^{M^3}$
+    by letting $\mathsf{s}(w)$ be the Parikh image of $\mathsf{t}$, that is,
+    counting the number of occurrences of every letter.
+    It is clear how to define the polyregular function $g \colon (\Nat)^{M^3} \to \Rel$
+    that makes the following diagram commute, at is it even a polynomial function
+    with $M^3$ indeterminates:
+    \begin{center}
+        \begin{tikzcd}
+            (M^3)^* 
+            \arrow{dr}[below,anchor=center,rotate=-39,yshift=-3pt]{\mathsf{count}}
+            & \Sigma^* \arrow{l}[above]{\mathsf{t}} \arrow{r}{f} \arrow{d}[left]{\mathsf{s}} & \Rel \\
+            & \Nat^{M^3} \arrow{ur}[below]{g} & \\
+        \end{tikzcd}
+    \end{center}
+    One problem is that the map $\mathsf{s}$ is not surjective, because of
+    potential dependencies between the types encountered in $w$.
+    Let us however notice that one can build a context-free grammar
+    recognizing exactly the image of $\mathsf{t}$. Namely,
+    \begin{align*}
+        S &\to S^{1,m,1} &  m \in M \\
+        S^{m_r, m, m_l} &\to (m_l, m, m_r) & \exists a \in \Sigma, \mu(a) = m \\
+        S^{m_r, m, m_l} &\to 
+        S^{m_r, m_x, m_y m_l}
+        S^{m_r m_x, m_y, m_l}
+                        & m_x m_y = m
+    \end{align*}
+
+    Using Parikh's Theorem \cite{parikh66context},
+    we conclude that the image of $\mathsf{s}$ is a semilinear set,
+    i.e., that there exists a finite set $Q$,
+    numbers $\seqof{d_q}{q \in Q}$,
+    vectors $\seqof{v_{i,q}}{0 \leq i \leq d_q, q \in Q}$ in $\Nat^{M^3}$,
+    such that
+    \begin{equation*}
+        \mathsf{s}(\Sigma^*) =
+        \bigcup_{q \in Q}
+        \setof{
+            v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
+        }{
+            (\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}
+        } \quad .
+    \end{equation*}
+
+    We can therefore define a function $\mathsf{decode} \colon \Sigma^* \to
+    \sum_{q \in Q} \Nat^{d_q}$ that maps to a word $w$ to \emph{some} $q \in Q$
+    and parameters $(\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}$ such that
+    $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$. By taming the
+    semilinear set prior to this step, we can ensure that the function is in
+    fact uniquely defined, this is done by computing an equivalent semilinear
+    set that is \emph{unambiguous} and \emph{proper} \cite{chistikov16}.
+    Similarly, let us define $\mathsf{encode} \colon \sum_{q \in Q} \Nat^{d_q}
+    \to \Sigma^*$ that maps an element $q \in Q$ with parameters $(\lambda_1,
+    \dots, \lambda_{d_q}) \in \Nat^{d_q}$ to \emph{some} word $w \in \Sigma^*$
+    such that $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$.
+    Notice that $\mathsf{decode} \circ \mathsf{encode} = \mathsf{id}$, by the
+    unicity of the representation in $\sum_{q \in Q} \Nat^{d_q}$.
+    We can define the map $\mathsf{repr} \colon \sum_{q \in Q} \Nat^{d_q} \to \Nat^{M^3}$
+    as follows
+    \begin{equation*}
+        \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \defined
+        v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
+        \quad .
+    \end{equation*}
+    Finally, let us define for $q \in Q$ the map $g_q \colon \Nat^{d_q} \to \Rel$ as the polynomial
+    expression:
+    \begin{equation*}
+        g_q(\lambda_1, \dots, \lambda_{d_q}) \defined \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \quad .
+    \end{equation*}
+
+    We have proven that the following diagram commutes:
+    \begin{center}
+        \begin{tikzcd}
+            \Sigma^* \arrow{rr}{f} 
+                     \arrow{dr}{\mathsf{s}}
+                     \arrow[bend right=20]{dd}[rotate=90,anchor=center,yshift=4pt]{\mathsf{decode}}
+                     &        & \Rel \\
+                     & \Nat^{M^3} \arrow{ur}[below]{g} & \\
+            \sum_{q \in Q} \Nat^{d_q}
+            \arrow[bend right=20]{uu}[right,rotate=90,anchor=center,yshift=-4pt]{\mathsf{encode}}
+            \arrow{ur}[left]{\mathsf{repr}}
+            \arrow[bend right=45]{uurr}[right]{\seqof{g_q}{q \in Q}}
+                     & & \\
+        \end{tikzcd}
+    \end{center}
+
+    The only missing part of this proof is the fact that $\mathsf{encode}$,
+    $\mathsf{decode}$ can be computed as \kl{polyregular functions}. Notice
+    that \emph{mutatis mutandis}, the same proof scheme applies to $f \in
+    \ZSF[k]$, by assuming that the monoid is aperiodic, since the functions
+    constructed (apart from $\mathsf{encode}$ and $\mathsf{decode}$) will be
+    \kl{star-free}.
+\end{proof}

### 2024-02-14 parts/conclusion.tex
uploading to ICALP

--- 2024-01-16 parts/conclusion.tex
@@
 %! lang = en-US
-\section{Conclusion}
+\section{Outlook}
 \label{sec:ccl}
 
-\begin{itemize}
-    \item Actually, we have an algorithm that allows
-        us to \emph{decide} if a polynomial is $\Nat$-rational.
-    \item Actually, we obtain a characterisation of
-        commutative 
-        $\Nat$-rational series of polynomial growth
-        among $\Rel$-rational series.
-    \item The question of $\Nat$-polyregular functions
-        among $\Rel$-polyregular functions remains open
-        because of the non-commutativity of the input.
-    \item The question of $\Nat$-rational series (commutative)
-        among $\Rel$-rational series remains open.
-\end{itemize}
+In the commutative case, the conjectured semantic characterizations of $\NSF$
+and $\NPoly$ have been proven to hold. Furthermore, the existence of a
+canonical transducer computing functions in $\NPoly$ opens the door to several
+questions with respect to aperiodicity. We can prove that for functions in
+$\NPoly[0]$, aperiodicity is equivalent to the absence of counters in the
+\kl{residual transducer} of the function (see
+\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
+higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
+In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number
+of states computing the function is \kl{counter-free}, which suggests that a
+more sophisticated object is needed to observe aperiodicity.
 
-\begin{center}
-    \begin{tabular}{c|cc}
-        \toprule
-        & \textbf{Commutative} & \textbf{General} \\
-        \midrule
-        \textbf{Commutative}
-        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
-        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
-        \textbf{General} & ? & polyregular 
-    \end{tabular}
-\end{center}
+\begin{lemma}
+    \label{aperiodic-iff-residual:lem}
+    Let $f \in \NPoly[0]$, the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NSF$,
+        \item $f \in \ZSF$,
+        \item $f$ is \kl{ultimately polynomial},
+        \item The \kl{$0$-residual transducer} of $f$ is \kl{counter-free}.
+    \end{enumerate}
+\end{lemma}
+
+\begin{example}
+    \label{non-aperiodic-residual-transd:ex}
+    Let us define
+    $\BadExKo(\varepsilon) = 1$,
+    $\BadExKo(a) = 0$,
+    $\BadExKo(a^2) = 1$,
+    and $\BadExKo(a^n) = n - 3$ for all $n \geq 3$.
+    The \kl{$0$-residual transducer} of $\BadExKo$ has a \kl{counter} and two states.
+    Furthermore,
+    every \kl{$\NPoly[0]$-transducer} with two states contains a \kl{counter}.
+\end{example}
+
+\AP In the spirit of the characterization of $\NPoly$ based the
+$(\resleq{f}{k})$ quasi-ordering (\cref{non-commutative-npoly:thm}), we
+introduce the notion of \intro{aperiodic ordering} of $\Sigma^*$ as follows:
+$(\Sigma^*, \leq)$ is \reintro[aperiodic ordering]{aperiodic} whenever for all
+$u, w \in \Sigma^*$, there exists $N_0 \in \Nat$, such that the sequence
+$\seqof{uw^n}{n \geq N_0}$ is increasing. This leads to the following
+\cref{sf-no-periods-on-sequences:conj}, of which the easy implication is
+already known to hold (\cref{sf-no-periods-on-sequences:lemma}), and requires
+the introduction of the \emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of
+$(\resleq{f}{k})$, defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(v - u) \in
+\NSF[k-1]$.
+
+\begin{lemma}
+    \label{sf-no-periods-on-sequences:lemma}
+    Let $k \in \Nat$, and $f \in \SF[k]$. Then,
+    $(\Sigma^*, \resleqsf{f}{k})$ is an
+    \kl[aperiodic ordering]{aperiodic} \kl{well-quasi-ordering}.
+\end{lemma}
+
+\begin{conjecture}
+    \label{sf-no-periods-on-sequences:conj}
+    For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
+    $f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
+    \kl{well-quasi-ordering}.
+\end{conjecture}
+
+Let us conclude with a more general remark on the status of commutative input
+in the study of polyregular functions. Notice that semantic characterizations
+of subclasses of \kl{polyregular functions} (\kl{combinatorial}, \kl{ultimately
+polynomial}, pumpable) arise from pre-composition with a \kl{commutative}
+\kl{star-free polyregular function}, as formally stated in
+\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
+Generalizing this idea, we conjecture that an effective encoding of $\ZPoly$ in
+terms of commutative functions could be obtained, as stated in
+\cref{z-poly-commutative-encoding:conj}, playing the same role as the
+decomposition obtained in \cref{decompose-polynomial:lem} in the
+\kl{commutative} case.
+
+\begin{remark}
+    \label{pre-compose-growth-commut:lemma}
+    Let $f \in \ZRat$, and $k \in \Nat$. Then,
+    $f \in \ZPoly[k]$ if and only if 
+    for every \kl{commutative} \kl{star-free polyregular function} $h$
+            of \kl{growth rate} $l \in \Nat$,
+            $(f \circ h) \in \ZPoly[k+l-1]$.
+\end{remark}
+
+\begin{remark}
+    \label{pre-compose-sf-commut:lemma}
+    Let $f \in \ZPoly$. Then, $f \in \ZSF$,
+    if and only if for every \kl{commutative} \kl{star-free polyregular function} $h$,
+            $(f \circ h) \in \ZSF$.
+\end{remark}
+
+\begin{conjecture}
+    \label{z-poly-commutative-encoding:conj}
+    Let $k \in \Nat$, and let $f \in \ZPoly[k]$ (resp. $\ZSF[k]$).
+    There exists a finite set $Q$, 
+    parameters $l_q \in \Nat$ for all $q \in Q$,
+    \kl{commutative} polyregular functions (resp. \kl{star-free polyregular})
+    $(g_q)_{q \in Q}$
+    from $\Nat^{l_q}$ to $\Rel$,
+    a surjective polyregular function (resp. \kl{star-free polyregular})
+    $\mathsf{dec} \colon \Sigma^* \tosurj \sum_{q \in Q} \Nat^{l_q}$,
+    and a polyregular function (resp. \kl{star-free polyregular})
+    $\mathsf{enc} \colon \sum_{q \in Q} \Nat^{l_q} \to \Sigma^*$,
+    such that
+    $f = (\sum_{q \in Q} g_q) \circ \mathsf{dec}$
+    and $\mathsf{dec} \circ \mathsf{enc} = \mathsf{id}$.
+\end{conjecture}
+
+In the specific subclass of \emph{polyblind functions}
+\cite{nguyn21comparison,doueneau2022hiding}, parts of the conjecture holds,
+because they essentially behave as commutative functions \cite[Theorem
+6.12]{gaetanphd}. Indeed, a suitable application of Parikh's Theorem
+\cite{parikh66context} guarantees the existence of the functions $\mathsf{enc}$
+and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
+We consider this partial result as a strong indicator in favor of the
+conjecture.
+
+\begin{remark}
+    \label{z-poly-commutative-encoding:remark}
+    The conjecture \cref{z-poly-commutative-encoding:conj} holds for the
+    subclass of polyblind functions when dropping the assumption that
+    $\mathsf{enc}$ and $\mathsf{dec}$ are \kl{polyregular functions}
+    (resp. \kl{star-free polyregular functions}).
+\end{remark}

### 2024-02-14 parts/correction.tex
uploading to ICALP

@@
     Let $\vec{X}$ be a finite tuple of indeterminates.
-    The class $\CorrectPoly[\vec{X}]$ is the collection of
+    The class $\intro*\CorrectPoly[\vec{X}]$ is the collection of
     polynomials $P \in \Rel[\vec{X}]$ such that
@@
 \AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
-\kl{discrete derivative} to perform an induction on the \kl{maximal monomials}.
+\intro{discrete derivative} to perform an induction on the \kl{maximal monomials}.
 This notion of derivation is built by translating the domain of the polynomial.
-To that end, let us write $\translate{K}$ for the \intro{translation function}
+To that end, let us write $\intro*\translate{K}$ for the \intro{translation function}
 that maps a polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 +
@@
     \begin{equation*}
-        \Diff{K}{P} \defined 
+        \intro*\Diff{K}{P} \defined 
         \translate{K}(P) - P
@@
 
-\begin{fact}
+\begin{remark}
     \label{discrete-deriv-linear:fact}
     For all tuples $\vec{X}$ of indeterminates,
-    for all $K \in \Nat$, $L \in \Rel$, for all partial function
+    for all $K \in \Nat$, $L \in \Rel$, for all partial functions
     $\nu \colon \vec{X} \topartial \Nat$:
@@
     \end{enumerate}
-\end{fact}
+\end{remark}
 
@@
             for 
-            all $\nu \colon \vec{X} \topartial \Nat$,
+            all partial functions $\nu \colon \vec{X} \topartial \Nat$,
             $\translate{K}(\restr{P}{\nu}) \in \Nat[\vec{X}]$,
         \item \label{d-t-transl-fin:item}
-            for all
+            for all partial functions
             $\nu \colon \vec{X} \topartial \set{0, \dots, K}$,
@@
     \end{enumerate}
+    Furthermore, the membership is decidable.
 \end{lemma}
-
-\begin{corollary}
-    \label{decidability-correct:lem}
-    One can decide membership in $\CorrectPoly$.
-\end{corollary}
 

### 2024-02-14 parts/counter.tex
uploading to ICALP

@@
 
-\begin{fact}
+\begin{remark}
     \label{n-poly-n-poly:example}
@@
     is a \kl{$\Nat$-rational polynomial}.
-\end{fact}
+\end{remark}
 
@@
     Let $\vec{X}$ be a finite tuple of indeterminate.
-    The class $\CoveredPoly[\vec{X}]$
+    The class $\intro*\CoveredPoly[\vec{X}]$
     is the class of polynomials $P \in \Rel[\vec{X}]$
@@
     When the indeterminate are clear from the context, we write
-    this class $\CoveredPoly$.
+    this class $\reintro*\CoveredPoly$.
 \end{definition}
@@
     \label{def:bad-polynomial}
-    We define $\BadPoly(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
+    We define $\intro*\BadPoly(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
 \end{definition}

### 2024-02-14 parts/intro.tex
uploading to ICALP

@@
 defined in terms of deterministic finite automata and regular expressions
-\cite{kleene1956representation}, Monadic Second Order Logic ($\MSO$)
+\cite{kleene1956representation}, Monadic Second Order Logic ($\intro*\MSO$)
 \cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
@@
 represented by star-free regular expressions \cite{schutzenberger1965finite},
-definable in First Order Logic ($\FO$) \cite{perrin1986first}, and recognized
+definable in First Order Logic ($\intro*\FO$) \cite{perrin1986first}, and recognized
 by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
@@
 (sequential, rational, regular, and \kl{polyregular functions}). In this paper,
-we will focus on \kl{polyregular functions} ($\Poly$), that shares with regular
+we will focus on \intro{polyregular functions} ($\intro*\Poly$), that shares with regular
 languages its characterizations in terms of logic (via $\MSO$ interpretations),
@@
 \cite{bojanczyk2019string}. As for regular languages, the class
-$\SF$ of \intro{star-free polyregular function} can equivalently be introduced
+$\intro*\SF$ of \intro{star-free polyregular function} can equivalently be introduced
 in terms of first-order definability, counter-free pebble transducers, and
@@
 difficult in the case of \kl{polyregular functions}, due to the lack of
-canonical objects akin to the syntactic monoid or the minimal automaton.
-Indeed, for sequential functions that admit a canonical object, aperiodicity is
-decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
-decidable for rational functions
-\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
-regular functions, although it has been proven decidable when changing the
-semantics of function equality to account for the \emph{origins} in the output
-\cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
-interesting parameter is the \intro{growth rate} of a function $f$, which is given
-by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
-$\Poly$. We write $\Poly[k]$ for the \kl{polyregular functions} of \kl{growth
-rate} $k$. It has been shown that the parameter $k$ controls the maximal
-complexity of the $\MSO$ interpretation needed to represent the function
-\cite{bojanczyk2022transducers}.
+canonical objects akin to the syntactic monoid or the minimal automaton. For
+sequential functions (that admit a canonical object), aperiodicity is decidable
+\cite{choffrut03}, and more recently, aperiodicity has been proven decidable
+for rational functions \cite{filiot2016aperiodicity,filiot2018canonical}. The
+problem remains open for regular functions, although it has been proven
+decidable when changing the semantics of function equality to account for the
+\emph{origins} in the output \cite{bojanczyk14}. To construct subclasses of
+\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
+of a function $f$, which is given by a number $k \in \Nat$, which can be used
+to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
+\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
+functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
+controls the maximal complexity of the $\MSO$ interpretation needed to
+represent the function \cite{bojanczyk2022transducers}.
 
@@
 results proven on \emph{unary polyregular functions}, also known as
-\kl{$\Nat$-polyregular functions} ($\NPoly$)
+\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
 \cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
@@
 \kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
-$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\ZPoly$)
+$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
 \cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
@@
 based on differences between functions, crucially leveraging negative outputs
-(i.e., the presence of inverse elements). If the proof method does not carry
+(i.e., the presence of inverse elements). Although the proof method does not carry
 from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
 that the semantic characterization of \kl{star-free $\Rel$-polyregular
-functions} ($\ZSF$) would continue to characterize \kl{star-free
-$\Nat$-polyregular functions} ($\NSF$).
+functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
+$\Nat$-polyregular functions} ($\intro*\NSF$).
 
@@
 It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
-of \kl{$\Nat$-rational series} ($\NRat$) (resp. of \kl{$\Rel$-rational series}
-$\ZRat$), that have been characterized in terms of \kl{growth rate}, ambiguity,
-and a notion of \emph{quantitative} $\MSO$
+of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
+series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
+ambiguity, and a notion of \emph{quantitative} $\MSO$
 \cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
@@
 \textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
-\emph{polynomials}, with an incorrect proof. The decidability of $\NPoly$
-inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem \cite[Open
-question 5.55]{gaetanphd}.
+\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
+$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
+\cite[Open question 5.55]{gaetanphd}.
 
@@
 \begin{figure}
-    \includestandalone[width=5cm]{tikz/class-inclusions}
+    \centering
+    \includestandalone[width=7cm]{tikz/class-inclusions}
     \caption{
@@
 
-Finally, we make a first step towards understanding the non-commutative by
+Finally, we make a first step towards understanding the non-commutative case by
 introducing a canonical model of computation for \kl{$\Nat$-polyregular
 functions} built upon residuals, called the \kl{residual transducer}. This
-allows the characterization of $\NPoly[k]$ as the class of functions $f$ such
-that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
-\kl{well-quasi-ordering}, generalizing the characterization of $\ZPoly[k]$ as
-the class of functions $f$ such that $\Res(f)$ has finite index (for some
-equivalence relation). The \kl{residual transducer} is effectively computable
-if $f$ is a \kl{commutative} \kl{$\Nat$-polyregular function}, and it is
-conjectured to be computable for all functions in $\NPoly$.
+residual transducer is obtained by shifting attention from the notion of
+equivalence relations of finite index over $\Sigma^*$, used in traditional
+automata theory and in the study of $\ZPoly$, to the notion of
+\kl{well-quasi-orderings}. The residual transducer is effectively computable if
+$f$ is \kl{commutative} and \kl{$\Nat$-polyregular}, and it is conjectured to
+be computable for all functions in $\NPoly$.
 
@@
 $\Rel$-polyregular functions), show that one can decide if a function $f \in
-\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give the
-basic definitions used on multivariate polynomials. In \cref{polynomials:sec},
-we focus on \emph{polynomials}, by introducing the \cref{karh:thm} of
-\textcite{KARH77}, providing a counter-example in \cref{thm:counter-example},
-and an alternative characterization in \cref{corrected-version:thm}. In
-\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
-decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
-(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture} holds
-for \kl{commutative} functions (\cref{zsf-npoly-nsf:thm}). Finally, in
-\cref{beyond-commutative:sec}, we introduce the notion of \kl{residual
-transducer} (\cref{residual-transducer:def}), and prove in
-\cref{non-commutative-npoly:thm} that it characterizes $\NPoly$ inside
-$\ZPoly$.
+\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
+the basic definitions used on multivariate polynomials. In
+\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
+\cref{karh:thm} of \textcite{KARH77}, providing a counter-example in
+\cref{thm:counter-example}, and an alternative characterization in
+\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
+results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
+\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
+\cref{zsf-nsf:conjecture} holds for \kl{commutative} functions
+(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
+introduce the notion of \kl{residual transducer}
+(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
+that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
+\kl{well-quasi-ordered} collection of residuals.

### 2024-02-14 parts/prelim.tex
uploading to ICALP

@@
 analysis of their \kl{commutative} counterpart, and others that provide
-a nicer syntax to construct examples. These characterizations are known, and
+a nicer syntax to construct examples. These characterizations were previously known, and
 can be found in \cite{gaetanphd,LOPEZ23b}.
@@
 
-Similar characterizations can be obtained for $\NSF$ and $\ZSF$.
+Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
+generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
 
@@
 $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
-that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\growth{f}(n) =
-\bigO(n^k)$, and similarly for the other classes.
+that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
+\bigO\left(\card{w}^k\right)$. and similarly for the other classes.
 
@@
 finite alphabet. We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
-map that counts occurrences of every letter in the input word, namely: $
-\commute[w] \defined (a \mapsto \card[a]{w})$. Let $X$ be a set. A function $f
-\colon \Sigma^* \to X$ is \intro{commutative} whenever for all $u,v \in
-\Sigma^*$, $f(uv) = f(vu)$. Equivalently, it is \reintro{commutative} whenever
-there exists a map $g \colon \Nat^\Sigma \to X$ such that $g \circ \commute =
-f$.
+map that counts occurrences of every letter in the input word (that is,
+computes the Parikh vector) namely: $ \commute[w] \defined (a \mapsto
+\card[a]{w})$. Let $X$ be a set, a function $f \colon \Sigma^* \to X$ is
+\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
+Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
+\Nat^\Sigma \to X$ such that $g \circ \commute = f$.
 
@@
     is \kl{commutative}.
-\end{lemma}
-
-Let us briefly remark that the characterizations of subclasses of
-\kl{polyregular functions} that are obtained in practice can be obtained by
-pre-composition with a \kl{commutative} \kl{polyregular function}. While the
-two following
-\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma} do not
-provide simpler proofs, they suggest the central position of \kl{commutative}
-\kl{star-free polyregular functions} as \emph{test functions} for various
-properties (growth rate, aperiodicity).
-
-\begin{lemma}
-    \label{pre-compose-growth-commut:lemma}
-    Let $f \in \ZRat$, and $k \in \Nat$,
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \ZPoly[k]$,
-        \item For every \kl{commutative} \kl{star-free polyregular function} $h$
-            of \kl{growth rate} $l \in \Nat$,
-            $(f \circ h) \in \ZPoly[k+l-1]$.
-    \end{enumerate}
-\end{lemma}
-
-\begin{lemma}
-    \label{pre-compose-sf-commut:lemma}
-    Let $f \in \ZPoly$.
-    The following are equivalent:
-    \begin{enumerate}
-        \item $f \in \ZSF$,
-        \item For every \kl{commutative} \kl{star-free polyregular function} $h$,
-            $(f \circ h) \in \ZSF$.
-    \end{enumerate}
 \end{lemma}
@@
 For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
-\kl{monomial}, but $X + Y$ or $2X^2 + XY$ are not. We write $\Monomials[X_1,
+\kl{monomial}, but $X + Y$ or $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
 \dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
@@
 + X + Y$, the set of \reintro{maximal monomials}, written
-$\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  X^2}$. A \kl{monomial} has a
-\intro{non-negative coefficient} if its multiplicative constant belongs to
-$\Nat$, or equivalently, if it is \kl{non-negative} as a polynomial. In the
-polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials}
-are $X^2$ and $Y^2$.
+$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  X^2}$.  In the polynomial
+$P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials} are $X^2$ and
+$Y^2$.

### 2024-02-14 parts/starfree.tex
uploading to ICALP

@@
 
-Let us now go beyond polynomials and consider the case of \kl{commutative}
-\kl{$\Rel$-polyregular functions}. The hypothesis of \kl{commutativity} reduces
-$\ZPoly$ to a combination of a finite number of polynomials, arranged by
-counting modulo some numbers. This characterization will be proven in
-\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
-procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
-\cref{decompose-polynomial:lem} is actually a refinement of
-\cref{n-poly-combinatorics:lem} under the extra assumption of
+The goal of this section is to go beyond polynomials and consider the case of
+\kl{commutative} \kl{$\Rel$-polyregular functions}. The hypothesis of
+\kl{commutativity} reduces $\ZPoly$ to a combination of a finite number of
+polynomials, arranged by counting modulo some numbers. This characterization
+will be proven in \cref{decompose-polynomial:lem}, and will be the key
+ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$
+inside $\NPoly$. Note that \cref{decompose-polynomial:lem} is actually a
+refinement of \cref{n-poly-combinatorics:lem} under the extra assumption of
 \kl{commutativity}.
 
-\AP To formally state the lemma, we will need some notations that will
+\AP To actually state the lemma, we will need some notations that will
 formalize the intuitive idea of counting tuples of integers modulo some
-constant. Let $\omega \in \Nat$, we write the set of \intro{types modulo
-$\omega$} $\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
-    \dots, \omega - 1}$. To a number $n \in \Nat$, we associate its
-    \intro{$\omega$-type} written $\moduloType[\omega](n)$ which is defined as the pair
-    $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest of the Euclidian
-    division of $n$ by $\omega$. This notion of type is lifted to vectors in
-    $\Nat^p$ for any $p \in \Nat$ by pointwise application.
+constant. Let $\omega \in \Nat$, we define the set $\intro*\ModuloTypes[\omega]
+\defined \set{0, \dots, \omega^2} \times \set{0, \dots, \omega - 1}$ of
+\intro{types modulo $\omega$}. To a number $n \in \Nat$, we associate its
+\reintro{$\omega$-type} written $\intro*\moduloType[\omega](n)$ which is
+defined as the pair $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest
+of the Euclidian division of $n$ by $\omega$. This notion of type is lifted to
+vectors in $\Nat^p$ for any $p \in \Nat$ by pointwise application.
 
@@
 
-The goal of this section is to go beyond polynomials and characterize all
-\kl{$\Rel$-polyregular functions} that are \kl{$\Nat$-polyregular}, under the
-assumption that they are \kl{commutative}. We will provide a semantic property
-in the form of \cref{k-combinatorial:def}, that is conjecture to characterize
-$\NPoly$ inside $\ZPoly$, even in the non-commutative setting.
+In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
+of \emph{polynomial pumping arguments}
+\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
+semantic property in the form of \cref{k-combinatorial:def}, that is conjecture
+to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.
 
@@
 \kl{$\Nat$-polyregular functions} can be combined with the recent advances in
-the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to prove the
-following theorem. As a corollary of \cref{zsf-npoly-nsf:thm} and the effective
-decision procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that
-$\NSF$ is decidable inside $\NPoly$ under the assumption of commutativity, and
-that the effective procedure can be used The key ingredient of this section is
-the use of a semantic characterization of \kl{star-free $\Rel$-polyregular
-functions} among \kl{$\Rel$-rational series} that generalizes the aperiodicity
-of languages to functions by the means of polynomial behaviors.
+the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to decide
+aperiodicity. The key ingredient of this study is the use of a semantic
+characterization of \kl{star-free $\Rel$-polyregular functions} among
+\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
+functions by the means of polynomial behaviors.
 
@@
     \label{zsf-npoly-nsf:thm}
-    Let $\Sigma$ be a finite alphabet, $k \in \Nat$,
+    Let $\Sigma$ be a finite alphabet, 
     and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
@@
     Then, the following are equivalent:
-    \begin{itemize}
+    \begin{enumerate}
         \item $f$ is \kl{ultimately polynomial},
         \item $f \in \ZSF$,
-        \item $f \in \NSF[k]$.
-    \end{itemize}
+        \item $f \in \NSF$.
+    \end{enumerate}
     Furthermore, membership is decidable and effective conversions

### 2024-02-14 parts/transducers.tex
uploading to ICALP

@@
 
-The following definition of transducers is based on
-the marble transducers \cite{EHB99}, and follows the general pattern of
-transducers introduced in \cite{LOPEZ23b}.
+As a model of computation, we will use the following definition of transducers,
+that is based on the marble transducers \cite{EHB99}, and follows the general
+pattern of transducers introduced in \cite{LOPEZ23b}. Note that we only
+consider $\Nat$-valued functions, but could consider other commutative
+semigroups.
 
@@
 introduce its semantics, which is given by a simple recursion
-scheme.
+scheme. Intuitively, the function computed by a transducer is
+
 \begin{definition}
@@
     \label{non-canonical-transd:ex}
-    Let $f \colon \set{a}^* \to \Nat$
+    Let $\BadExOk \colon \set{a}^* \to \Nat$
     be defined as
-    $f(\varepsilon) = 1$,
-    $f(a) = 0$,
-    and $f(a^n) = n - 1$ for all $n \geq 2$.
-    \textbf{TODO: split the example}
-    $f(\varepsilon) = 1$,
-    $f(a) = 0$,
-    $f(a^2) = 1$,
-    and $f(a^n) = n - 3$ for all $n \geq 3$.
-    The function is \kl{computed} by the two automata
-    in \cref{non-canonical-transd:fig}.
+    $\BadExOk(\varepsilon) = 1$,
+    $\BadExOk(a) = 0$,
+    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
+    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
+    depicted in \cref{non-canonical-transd:fig}.
 \end{example}
@@
     \caption{Two \kl{$\NPoly[0]$-transducers}
-    \kl{computing} the function of \cref{non-canonical-transd:ex}.}
+    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
     \label{non-canonical-transd:fig}
@@
 
-\AP As in the case of finite automata, we generalize the function $\delta
-\colon Q \times \Sigma \to Q$ to words by defining $\delta(q, \varepsilon) =
-q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. Similarly, we write
-$\lambda(q,\varepsilon) \defined 0$, and recursively define $\lambda(q, au)
-\defined \app{\lambda(q,a)}{u} + \lambda(\delta(q,a), u)$. Following these
-equations, we can write for all $q \in Q$, $u,v \in \Sigma^*$, $\aTransd(q, uv)
-= \aTransd(\delta(q, u), v) + \lambda(q, u)(v)$. Finally, let us notice that if
-$\mathcal{H}$ is closed under finite sums and shifts, then $\lambda(q, u) \in
-\mathcal{H}$ for all $(q,u) \in Q \times \Sigma^*$.
+\AP It is already known that this notion of transducers exactly captures $\NSF$
+and $\NPoly$, and generalizes to outputs into other commutative semigroups than
+$\Nat$. To make the above statement formal, let us recall that a
+\intro{counter} in an automaton is a pair $u,w$ of words together with an
+integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0, u)$ and
+$\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$; where we used the
+extension of $\delta \colon Q \times \Sigma \to Q$ to words by defining
+$\delta(q, \varepsilon) = q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. An
+automaton is \reintro{counter-free} if it contains no \reintro{counters}. This
+notion can be lifted to \kl{$\mathcal{H}$-transducers} by considering counters
+of their underlying automata.
 
-\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
-words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
-\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$.
-An automaton is \reintro{counter-free} if it contains no \reintro{counters}.
-This notion can be applied to \kl{$\mathcal{H}$-transducers} by considering
-counters of their underlying automata. It is already known in the literature
-that this notion of transducers correctly captures both $\NSF$ and $\NPoly$ in
-the sense that the following fact will clarify.
-
-\begin{lemma}[{\cite{gaetanphd}}]
+\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{gaetanphd}}]
     \label{transducer-nsf-npoly:lemma}
@@
 \AP The core concept of this section is that of a \intro{residual} of a
-function $f \colon \Sigma^* \to \Rel$,  defined by $\app{f}{u} \defined w
-\mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
-denoted $\Res(f)$ and is defined as the set of $\app{f}{u}$ where $u$ ranges
-over words in $\Sigma^*$. Let $k \in \Nat$. We define $f \zpolyequiv[k] g$ if
-and only if $(g - f) \in \ZPoly[k-1]$, and $f \npolyleq[k] g$ if and only if
-$(g - f) \in \NPoly[k-1]$. Given a function $f \colon \Sigma^* \to \Rel$, we
-are going to study both $(\Res(f), \zpolyequiv[k])$, and $(\Res(f),
-\npolyleq[k])$. However, to simplify notations, instead of writing $\app{f}{u}
-\npolyleq[k] \app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient
-notation $u \resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
-\resleq{f}{k})$.
+function $f \colon \Sigma^* \to \Rel$,  defined by $\intro*\app{f}{u} \defined
+w \mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
+denoted $\intro*\Res(f)$ and is defined as the set of $\reintro*\app{f}{u}$
+where $u$ ranges over words in $\Sigma^*$. Let $k \in \Nat$. We define $f
+\intro*\zpolyequiv[k] g$ if and only if $(g - f) \in \ZPoly[k-1]$, and $f
+\intro*\npolyleq[k] g$ if and only if $(g - f) \in \NPoly[k-1]$. Given a
+function $f \colon \Sigma^* \to \Rel$, we are going to study both $(\Res(f),
+\reintro*\zpolyequiv[k])$, and $(\Res(f), \reintro*\npolyleq[k])$. However, to
+simplify notations, instead of writing $\app{f}{u} \reintro*\npolyleq[k]
+\app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient notation $u
+\intro*\resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
+\reintro*\resleq{f}{k})$.
 
-\begin{fact}
+\begin{remark}
     \label{good-residual-ordering:fact}
@@
     \end{enumerate}
-\end{fact}
+\end{remark}
 
+\AP
 Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
@@
 transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
-\resleq{f}{k} ua$. This naïve approach does not yield a canonical model which
-is corrected by the following definition.
+\resleq{f}{k} ua$. This naïve approach does not yield a canonical model, as
+illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
+\cref{non-canonical-transd:fig} computing the function $\BadExOk$ of
+\cref{non-canonical-transd:ex}, and having as states minimal elements for $\resleq{\BadExOk}{0}$.
+To obtain a
+canonical notion, we enforce that the set of states is a \intro{downwards
+closed} subset of $\Sigma^*$ for the prefix ordering $(\prefleq)$, that is, if
+$u \in \Sigma^*$ is a state, then every prefix of $u$ must also be a state.
 
@@
 
-Because of the restriction on prefixes, the \kl{$k$-residual transducer}
-is actually unique when it exists.
-We now introduce \cref{residual:algo} to compute a \kl{$k$-residual transducer}
-given a function $f$. Notice that this algorithm requires the ability to test
-if a function belongs to $\NPoly[k]$, which is only known to be feasible for
-\kl{commutative} polyregular functions. However, the termination of this
-algorithm also proves the existence of the \kl{$k$-residual transducer}.
+Because of the restriction on prefixes, the \kl{$k$-residual transducer} is
+actually unique when it exists. We now introduce \cref{residual:algo} to
+compute a \kl{$k$-residual transducer} given a function $f$. Notice that this
+algorithm requires the ability to test if a function belongs to $\NPoly[k]$,
+which is only known to be feasible for \kl{commutative} \kl{polyregular
+functions}. However, the termination of this algorithm also proves the
+existence of the \kl{$k$-residual transducer}.
 
@@
             $\delta(u, a) \defined v$;
-
             $\lambda(u, a) \defined \app{f}{ua} - \app{f}{v}$;
@@
             $Q \defined Q \uplus \set{ ua }$;
-
             $\delta(u,a) \defined ua$;
-
             $\lambda(u,a) \defined 0$;
-
             $O \defined O \cup \setof{ uab }{b \in \Sigma}$;
@@
 section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
-\Rightarrow \resleq{f}{k})$ is \kl{well}, but the quas-ordering $(\resleq{f}{k})$ is a
+\Rightarrow \resleq{f}{k})$ is \kl{well}, but the quasi-ordering $(\resleq{f}{k})$ is a
 \kl{well-quasi-ordering}. This has to be compared with the property of being

### 2024-02-14 globals/metadata.tex
update ICALP

--- 2024-02-12 globals/metadata.tex
@@
     \begin{abstract}
-        This paper answers two questions about 
-        commutative unary output polyregular functions, called
-        $\Nat$-polyregular functions, that forms a proper subset
-        of $\Nat$-rational series.
-        We show that given a $\Rel$-rational series,
-        it is decidable whether it is computable via a commutative
-        $\Nat$-polyregular function, and provide a counter-example
-        to the theorem of Karhumäki that studied the same question
-        in the case of polynomials.
-
-        We also prove that it is decidable whether a commutative $\Nat$-polyregular
-        function is star-free, by proving the stronger statement
-        that star-free $\Rel$-polyregular functions that are
-        $\Nat$-polyregular are computable using a star-free $\Nat$-polyregular function.
-        This answers positively to previous conjectures on $\Nat$-polyregular functions,
-        and the study of the commutative case seems to be useful
-        towards a proof in the non-commutative case.
+This paper addresses two questions regarding $\Nat$-polyregular functions, that
+forms a proper subset of $\Nat$-rational series. We show that given a $\Rel$-rational
+series, it is decidable whether it is computable via a commutative
+$\Nat$-polyregular function, and provide a counter-example to the theorem of
+Karhumäki that studied the same question in the case of polynomials. We also
+prove that it is decidable whether a commutative $\Nat$-polyregular function is
+star-free, by proving the stronger statement that star-free $\Rel$-polyregular
+functions that are $\Nat$-polyregular are in fact computable using a star-free
+$\Nat$-polyregular function. Building towards answering the same questions in the
+non-commutative case, we present a canonical model of computation of
+$\Nat$-polyregular functions by generalizing the notion of residual transducers
+previously introduced in $\Rel$-polyregular functions.
     \end{abstract}
@@
 \SeriesVolume{42}
-\ArticleNo{23}
+\ArticleNo{285}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2024-02-14 parts/appendix.tex
update ICALP

@@
     \kl{$\Nat$-rational series}
-    with domain $\Sigma \defined \set{a_1, \dots, a_p}$
+    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
     that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
@@
     $P \in \CorrectPoly$, there exists a $K \in \Nat$ such that
-    $\translate{K}(\Diff{1}{P}) \in \CorrectPoly$, and in particular $g_t(X +
-    N_0 + K+1) - g_t(X + N_0 + K) \in \Nat$, for all $X \geq 0$.
+    $\translate{K}(\Diff{1}{P}) \in \Nat[X]$
+    (\cref{derivation-translation:lem}), and in particular $g_t(X + N_0 + K+1)
+    - g_t(X + N_0 + K) \in \Nat$, for all $X \geq 0$.
 
@@
     We can define the map $\mathsf{repr} \colon \sum_{q \in Q} \Nat^{d_q} \to \Nat^{M^3}$
-    as follows
+    as follows:
     \begin{equation*}
@@
 
-    The only missing part of this proof is the fact that $\mathsf{encode}$,
+    The only missing part of this proof is the fact that the maps $\mathsf{encode}$ and
     $\mathsf{decode}$ can be computed as \kl{polyregular functions}. Notice

### 2024-02-14 parts/correction.tex
update ICALP

@@
 surprising, given the conjecture that \kl{ultimately polynomial}
-\kl{$\Nat$-polyregular functions} (re-introduced in \cref{star-free:sec}) were
-conjectured to be star free (\cref{zsf-nsf:conjecture}), and that polynomials
-are always \kl{ultimately polynomial}.
+\kl{$\Nat$-polyregular functions} (the formal definition will be given in
+\cref{star-free:sec}) were conjectured to be star free
+(\cref{zsf-nsf:conjecture}), and that polynomials are always \kl{ultimately
+polynomial}.
 
@@
 adapted, and to ensure that the decision procedure is effective, one must
-obtain explicit bounds on the translations that are needed. To that end, let us
-now introduce the ordering over which the induction of
+obtain explicit bounds on the \kl{translations} that are needed. To that end,
+let us now introduce the ordering over which the induction of
 \cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
@@
 $M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
-\kl{well-quasi-ordering} \cite[Hoare quasi-ordering]{SCSC12}.
+\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
+quasi-ordering]{SCSC12}.
 

### 2024-02-14 parts/intro.tex
update ICALP

@@
 
+\AP
 Traditional results from automata theory state the equivalence of languages
@@
 
-There have been recurrent attempts to generalize the notions of regularity from
-languages to functions, leading to classes of various expressiveness
-(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
-we will focus on \intro{polyregular functions} ($\intro*\Poly$), that shares with regular
+\AP There have been recurrent attempts to generalize the notions of regularity
+from languages to functions, leading to classes of various expressiveness (for
+instance, sequential, rational, regular, and \kl{polyregular functions})
+\cite{EM65,Aho1969,eilenberg1974automata,EM02}. In this paper, we will focus on
+\intro{polyregular functions} ($\intro*\Poly$), that shares with regular
 languages its characterizations in terms of logic (via $\MSO$ interpretations),
@@
 via its relationships with lambda-terms and so-called for-transducers
-\cite{bojanczyk2019string}. As for regular languages, the class
-$\intro*\SF$ of \intro{star-free polyregular function} can equivalently be introduced
-in terms of first-order definability, counter-free pebble transducers, and
-suitable restrictions of for-tranducers.
+\cite{bojanczyk2019string}. As for regular languages, the class $\intro*\SF$ of
+\intro{star-free polyregular function} can equivalently be introduced in terms
+of first-order definability, counter-free pebble transducers, and suitable
+restrictions of for-tranducers.
 
+\AP
 Classical problems, including the decidability of aperiodicity, become
@@
 
+\AP
 Another way to approach the problem of deciding if a \kl{polyregular function}
@@
 
+\AP
 Recently, aperiodicity has been shown decidable over a restricted class of
@@
 
+\AP
 It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
@@
 
+\AP
 We refer the reader to \cref{previously-known-inclusions:fig} to get an
@@
 
+The approach taken in this paper avoids leveraging tools from the theory of
+factorizations forests in finite monoids \cite{simon90}, which has been the
+traditional method used to understand (commutative output) \kl{polyregular
+functions} \cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, and instead
+focus on intrinsic description of the functions themselves (via canonical
+computational models such as the \kl{residual transducer} of
+\cref{residual-transducer:def}, or explicit encoding/decoding into polynomials
+via \cref{decompose-polynomial:lem}).
+
 \paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce

### 2024-02-14 parts/prelim.tex
update ICALP

@@
 \Sigma^* \to \set{0,1}$ as the indicator function of $L$. Given an $\MSO$
-formula with first-order free variables $\varphi(\vec{x})$, we define
-$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
-a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
-$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
-to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
-$\Span{\mathbb{S}}(X)$ as the smallest subset of functions from $\Sigma^* \to
-\Rel$ containing $X$ and closed under (finite) sums and multiplication with $s
-\in \mathbb{S}$.
+(resp. $\FO$) formula with first-order free variables $\varphi(\vec{x})$, we
+define $\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$
+that maps a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$,
+such that $w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from
+$\Sigma^*$ to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
+$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
+$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
+multiplication with $s \in \mathbb{S}$.
 
@@
 
-\AP All functions in this paper have output in $\Nat$ or $\Rel$, hence, we
-avoid the definition of \kl{polyregular functions} in full generality. The
-following \cref{nat-rel-poly:def} is one of the equivalent definitions of
-\cite{LOPEZ23b}, and is similar in shape to the \emph{finite counting automata}
-introduced by \textcite{schutzenberger1962}. Let us recall that a monoid $M$ is
-\intro(monoid){aperiodic} whenever there exists $n \in \Nat$ such that for all
-$x \in M$, $x^{n+1} = x^n$.
+\AP Because the functions of interest in this paper have output in $\Nat$ or
+$\Rel$, we will only provide the definiton of \kl{polyregular functions} for
+these two output semgroups, and we refer the reader to
+\cite{bojanczyk2019string} for the general definition of \kl{polyregular
+functions}. The following \cref{nat-rel-poly:def} is one of the equivalent
+definitions of \cite{LOPEZ23b}, and is similar in shape to the \emph{finite
+counting automata} introduced by \textcite{schutzenberger1962}. Let us recall
+that a monoid $M$ is \intro(monoid){aperiodic} whenever there exists $n \in
+\Nat$ such that for all $x \in M$, $x^{n+1} = x^n$.
 
@@
 of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
-analysis of their \kl{commutative} counterpart, and others that provide
-a nicer syntax to construct examples. These characterizations were previously known, and
-can be found in \cite{gaetanphd,LOPEZ23b}.
+analysis of their \kl{commutative} counterpart, and others that provide a nicer
+syntax to construct examples. These characterizations were previously known,
+and can be found for instance in \cite{gaetanphd,LOPEZ23b}.
 
@@
 
-\AP Because this paper focuses on functions that are commutative both in the
-input and the output, let us introduce precisely what it means and how it
-relates to other functions from $\Nat^k \to \Nat$. For that, let $\Sigma$ be a
-finite alphabet. We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
+\AP 
+Let $\Sigma$ be a finite alphabet. 
+We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
 map that counts occurrences of every letter in the input word (that is,
@@
 n} \to \Sigma$ such that $g \circ (v \mapsto \seqof{v_{\eta(k)}}{1 \leq k \leq
-n}) \circ \commute = f$. A key family of functions will be \kl{represented} by
-\kl{$\Rel$-polyregular functions}, namely polynomials with multiple indeterminates.
+n}) \circ \commute = f$. This notion will be useful to formally state that a
+polynomial ``is'' a \kl{polyregular function}.
 
@@
 
-All \kl{$\Nat$-rational polynomials} are \kl{non-negative}, but the converse does
-not hold,
-as the following example illustrates. Note that the proof scheme of
-this example will be at the core of \cref{thm:counter-example}, and leverages
-a key property of $\NRat$ recalled hereafter.
+All \kl{$\Nat$-rational polynomials} are \kl{non-negative}, but the converse
+does not hold, as the following example illustrates. Note that the proof scheme
+of this example will be at the core of the counter-example to \cite[Theoerm
+3.3]{KARH77} in \cref{thm:counter-example}, and leverages a key property of
+$\NRat$ recalled hereafter.
 

### 2024-02-14 parts/starfree.tex
update ICALP

@@
     Furthermore, the properties are decidable,
-    and one can effectively compute a representation of $f$.
+    and conversions effective.
 \end{theorem}
@@
 \kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
-functions by the means of polynomial behaviors.
+functions by the means of polynomial behaviors (see
+\cref{aperidic-ultimately-polynomial:ex}).
 
@@
 
-Let us recall from \cite{LOPEZ23b} that this definition correctly
-generalizes the indicator functions of regular languages.
-
 \begin{example}
+    \label{aperidic-ultimately-polynomial:ex}
     A language $L$ is aperiodic if and only if 
@@
     \end{enumerate}
-    Furthermore, membership is decidable and effective conversions
-    are possible.
+    Furthermore, membership is decidable and conversions are effective.
 \end{theorem}

### 2024-02-14 parts/intro.tex
typo

@@
 
-The approach taken in this paper avoids leveraging tools from the theory of
+The approach taken in this paper avoids the use of tools from the theory of
 factorizations forests in finite monoids \cite{simon90}, which has been the
@@
 functions} \cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, and instead
-focus on intrinsic description of the functions themselves (via canonical
+focuses on intrinsic description of the functions themselves (via canonical
 computational models such as the \kl{residual transducer} of

### 2024-02-14 parts/starfree.tex
typo

@@
 \cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
-semantic property in the form of \cref{k-combinatorial:def}, that is conjecture
+semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
 to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

### 2024-03-27 reviews/rebuttal.md
review

---
title: Rebuttal
author: Aliaume LOPEZ
date: 2024-03-27
lang: en-GB
---

We would like to thank the reviewers for their comments and suggestions, which
are all constructive and will definitely help to improve the quality of the
paper. In particular, we will add the reference to the ArXiv paper
`arXiv:2301.09234` in the introduction which is definitely relevant.

Before answering to the questions of the reviewers, we would like to point out
that the summary of Reviewer 1 is slightly misleading as the paper does not
manage to decide whether a function is "$\mathbb{N}$-polyregular" in general,
but only does so in the case where the function is assumed to be *commutative*
(which is a strong restriction). Without the assumption of commutativity, the
problem remains open. We will make sure to put more emphasis on this point.

---

It is our understanding that a significant part of the reviewers' concerns
relates to the high number of typos and grammatical errors in the manuscript.
We sincerely apologize for these mistakes that made reading this paper
difficult, and we acknowledge the lack of consistency in the bibliographic
section and in particular the mix of unicode and non unicode bibtex entries
which butchered the name of Lê Thành Dũng Tito Nguyên. We will definitely
correct these issues in the revised version of the manuscript. We decided to
not respond to specific typos and grammatical errors in this rebuttal and focus
on more global issues.

---

A second point raised by the reviewers is the lack of clarity for non-expert
readers, with a technical introduction which does not take enough time to
explain the state of the art, and for instance clearly state results obtained
in [9] to compare them with what is done in this paper. To address this issue,
we will take more time to introduce historical results and explain the various
classes of functions that have been studied before, expanding Figure 1 with
concrete examples. We will take time to introduce the various notions of
string-to-string functions (mealy machines, sequential transducers, rational
functions, regular functions, polyregular functions) with a clear table
comparing them and stating the status of "deciding aperiodicity" for each of
these classes.

However, it seems reasonable for us to assume that the reader is familiar at
least with finite monoids, monadic second order logic on words, and weighted
automata (hence, with rational series) which are well established theories:
describing them in a few sentences with a reference to a standard book should
already greatly increase the readability of the paper.

The informal uses of notions such as "growth rate" and "commutativity" in the
introduction were intended to provide intuitions to the reader, but we
understand that it was counterproductive and will be more careful in the
revised version. This is not an isolated issue, and we will make sure to avoid
such informal statements.

We also understand that whether a proof belongs to the appendix or the main
body of the paper felt arbitrary, and that the key ingredients of the proofs
where not clearly identified. While it is difficult to answer to such a comment
without rewriting the paper, we will try to illustrate on a few examples.

- Lemma 7 is a simple proof in the case of polyregular functions, and has
  a sligthly more technical proof[^1] in the case of rational series, but the
  main interest of the paper is in polyregular functions, and therefore the
  proof is in appendix.
- Lemma 19 is a technical result which while technically new, follows proof
  schemes similar to those of [9, 12], and provides another proof of the
  correct part of the theorem of [23]. This is why it is in the appendix.
- Lemma 26 is the actual core of Theorem 29, which is a corrected version of
  a theorem from [23]. In the (incorrect) proof of [23], the error lies in the
  inductive step of the analogue of Lemma 26, which is why we decided to put
  and this is why this particular proof was kept in the main body of the paper,
  while the other technical details about multivariate polynomials were moved
  to the appendix. 
- Lemma 31 has a proof that is a bit technical, but the important part of the
  lemma is actually the statement itself, which decomposes a commutative
  polyregular function into actual *polynomials*. Hence, the proof is in
  appendix.

However, we agree that all the proofs of sections 5 and 6 are in appendix
except for the ones that combine technical lemmas, and a better trade-off
between the sections will be found in the revised version of the paper. We
believe that the interesting proofs are those of Lemmas 44, and 45, which
really leverage the notion of well-quasi-ordering.

We will therefore revise the "Outlook" section to shape it more as a discussion
of the results and their implications, which will leave space for a more
detailed introduction, explanations of the results, and interesting proofs of
section 5.

---

To conclude on the relationship between this paper and [9], which should be
made clearer, let us try to replace both papers in context.

The concern of [9] is about $\mathbb{Z}$-polyregular functions (which are
a subclass of $\mathbb{Z}$-rational series), defining a notion of
"aperiodicity" for such functions, and proving that aperiodicity is decidable.
The work of [9] could be put into context as follows: since the introduction of
polyregular functions [4,5], there has been a corresponding notion of
"aperiodic" polyregular functions. However, decidability of aperiodicity
remains an open question, even in the specific case of unary output (the
function output words over a single letter alphabet). Notice that polyregular
functions with unary output are in fact functions from $\Sigma^*$ to
$\mathbb{N}$ (we write a number in unary).

While the work of [9] could be interpreted as studying $\mathbb{Z}$-rational
series, it is really about deciding aperiodicity of a "variation" of
polyregular functions with unary output that allows "negative outputs".
Unfortunately this "variation" is substantial, and the relationship between
$\mathbb{Z}$-polyregular functions (of [9]) and $\mathbb{N}$-polyregular
functions (unary output) remained unclear up to this paper. In particular the
decidability of aperiodicity for $\mathbb{N}$-polyregular functions could not
be deduced directly from [9].

This paper provides the combinatorial tools to understand the differences
between $\mathbb{Z}$-polyregular functions and $\mathbb{N}$-polyregular
functions (for instance, by giving decision procedures), and leverages these
tools to prove that aperiodicity is decidable for the latter **in the
commutative case**. This is a significant step towards the decidability of
aperiodicity for polyregular functions. The next step would be to drop the
assumption of commutativity, and then to allow the outputs to non-unary
alphabets. Both steps are beyond the scope of the current paper.

We hope that the above clarifications and proposed changes will address the
concerns of the reviewers, and would like to thank them again for the
constructive feedback.

[^1]: There is actually a simpler proof of Lemma 7 in the case of rational
series that avoids the use of Zariski closures and is more direct.

### 2024-03-28 parts/conclusion.tex
update ICALP
update bibliography

@@
 In the specific subclass of \emph{polyblind functions}
-\cite{nguyn21comparison,doueneau2022hiding}, parts of the conjecture holds,
+\cite{LENP21,DOUE22}, parts of the conjecture holds,
 because they essentially behave as commutative functions \cite[Theorem
-6.12]{gaetanphd}. Indeed, a suitable application of Parikh's Theorem
-\cite{parikh66context} guarantees the existence of the functions $\mathsf{enc}$
+6.12]{DOUE23}. Indeed, a suitable application of Parikh's Theorem
+\cite{PARI66} guarantees the existence of the functions $\mathsf{enc}$
 and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
-We consider this partial result as a strong indicator in favor of the
+Although \cref{z-poly-commutative-encoding:remark} is not yet applicable in
+practice, we consider this partial result as a strong indicator in favor of the
 conjecture.

### 2024-03-28 parts/intro.tex
update bibliography

--- 2024-02-14 parts/intro.tex
@@
 
-\AP
-Traditional results from automata theory state the equivalence of languages
+\AP Traditional results from automata theory state the equivalence of languages
 defined in terms of deterministic finite automata and regular expressions
-\cite{kleene1956representation}, Monadic Second Order Logic ($\intro*\MSO$)
-\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
-Among regular languages, there exists a robust subclass sharing similar
-characterizations, namely the \emph{star-free} regular languages. These are
-obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
-represented by star-free regular expressions \cite{schutzenberger1965finite},
-definable in First Order Logic ($\intro*\FO$) \cite{perrin1986first}, and recognized
-by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
-regular language is aperiodic, thanks to the characterizations in terms of
-minimal automaton and/or syntactic monoid.
+\cite{KLEE56}, Monadic Second Order Logic ($\intro*\MSO$) \cite{BUCH60}, or
+finite monoids \cite{SCHU61}. Among regular languages, there exists a robust
+subclass sharing similar characterizations, namely the \emph{star-free} regular
+languages. These are obtained via counter-free (minimal) automata
+\cite{MNPA71}, represented by star-free regular expressions \cite{SCHU65},
+definable in First Order Logic ($\intro*\FO$) \cite{PEPI86}, and recognized by
+aperiodic monoids \cite{SCHU65}. It is decidable whether a regular language is
+aperiodic, thanks to the characterizations in terms of minimal automaton and/or
+syntactic monoid.
 
@@
 instance, sequential, rational, regular, and \kl{polyregular functions})
-\cite{EM65,Aho1969,eilenberg1974automata,EM02}. In this paper, we will focus on
+\cite{ELME65,AHUL69,EILE74,ENMA02}. In this paper, we will focus on
 \intro{polyregular functions} ($\intro*\Poly$), that shares with regular
@@
 via its relationships with lambda-terms and so-called for-transducers
-\cite{bojanczyk2019string}. As for regular languages, the class $\intro*\SF$ of
+\cite{BOKL19}. As for regular languages, the class $\intro*\SF$ of
 \intro{star-free polyregular function} can equivalently be introduced in terms
@@
 sequential functions (that admit a canonical object), aperiodicity is decidable
-\cite{choffrut03}, and more recently, aperiodicity has been proven decidable
-for rational functions \cite{filiot2016aperiodicity,filiot2018canonical}. The
+\cite{CHOF03}, and more recently, aperiodicity has been proven decidable
+for rational functions \cite{FGL16,FGLM18}. The
 problem remains open for regular functions, although it has been proven
 decidable when changing the semantics of function equality to account for the
-\emph{origins} in the output \cite{bojanczyk14}. To construct subclasses of
+\emph{origins} in the output \cite{BOJA14}. To construct subclasses of
 \kl{polyregular function}, an interesting parameter is the \intro{growth rate}
@@
 controls the maximal complexity of the $\MSO$ interpretation needed to
-represent the function \cite{bojanczyk2022transducers}.
+represent the function \cite{BOJA22}.
 
@@
 \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
-\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
+\cite{DOUE21,DOUE22}. For instance, the \kl{growth
 rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
-needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
-\cite{bojanczyk2022transducers}. Let us remark that the restriction to
+needed to compute $f$ \cite{DOUE21}, which does not hold on $\Poly$
+\cite{BOJA22}. Let us remark that the restriction to
 \emph{unary output} in the case of \kl{polyregular functions} can be lifted to
@@
 $\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
-\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
+\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
 by allowing negative outputs, and therefore enabling a limited form of
@@
 
-\begin{conjecture}[{\cite[Conjecture 7.61]{gaetanphd}}]
+\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
     \label{zsf-nsf:conjecture}
@@
 ambiguity, and a notion of \emph{quantitative} $\MSO$
-\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
-theorems obtained on these \cite{berstel2011noncommutative}. However, for these
+\cite{SCHU62,KRRC13}, and therefore can benefit from generic
+theorems obtained on these \cite{BERE10}. However, for these
 larger classes of functions, several competing notions of aperiodicity exist
-\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
+\cite{DRGA19,REUT80,CDTL23}, and the
 decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
@@
 $\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
-\cite[Open question 5.55]{gaetanphd}.
+\cite[Open question 5.55]{DOUE23}.
 
@@
 
-The approach taken in this paper avoids leveraging tools from the theory of
-factorizations forests in finite monoids \cite{simon90}, which has been the
+The approach taken in this paper avoids the use of tools from the theory of
+factorizations forests in finite monoids \cite{SIMO90}, which has been the
 traditional method used to understand (commutative output) \kl{polyregular
-functions} \cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, and instead
-focus on intrinsic description of the functions themselves (via canonical
+functions} \cite{DOUE21,DOUE22,CDTL23}, and instead
+focuses on intrinsic description of the functions themselves (via canonical
 computational models such as the \kl{residual transducer} of

### 2024-03-28 pnrs.tex
update proofs and paper
writing to publish
lipics compiles
update
big big rewrite
update bibliography

@@
 %! TeX program = xelatex
+%! TeX root = pnrs.tex
 %! lang = en-US
-\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
+\documentclass[anonymous,a4paper,UKenglish,cleveref,autoref,thm-restate]{lipics-v2021}
 
@@
 \input{globals/knowledges.kl}
-\input{globals/metadata}
 
@@
 
-\makeabstract
+\crefname{fact}{Fact}{Facts}
+\Crefname{fact}{Fact}{Facts}
+\crefname{algocf}{Algorithm}{Algorithms}
+\Crefname{algocf}{Algorithm}{Algorithms}
+
 % DOCUMENT
 \begin{document}
+\input{globals/metadata}
 
@@
 \maketitle
-\acknowledge
+\makeabstract
+%\acknowledge
 
@@
 \input{parts/starfree.tex}
+\input{parts/transducers.tex}
 \input{parts/conclusion.tex}
@@
 % BACKMATTER
-\printbibliography
+\bibliographystyle{plainurl}
+\bibliography{transducer-bib/automata,%
+transducer-bib/model-theory,%
+transducer-bib/sublinear,%
+transducer-bib/rational-series,%
+transducer-bib/polyregular,%
+transducer-bib/wqos}
 
@@
 
+\include{parts/appendix.tex}
+
 \end{document}

### 2024-03-29 parts/appendix.tex
adding section for proofs in appendix
update bibliography
rewrote intro

@@
 %! lang = en-US
-\section{Proofs of
-section \ref{preliminaries:sec}}
+\section{Proofs of section \ref{preliminaries:sec}}
 
-\begin{proof}[Proof of \cref{size-of-word-nsf:ex}]
+\begin{proofof}[size-of-word-nsf:ex]
     Let us define $M \defined (\set{1}, \times)$ which is 
@@
     \end{equation*}
-\end{proof}
+\end{proofof}
 
@@
     to the chosen ordering is a \kl{polyregular function}
-    \cite[Proposition II.12]{LOPEZ23b}.
+    \cite[Proposition II.12]{CDTL23}.
     As a consequence,
@@
     in the case of $\ZPoly$
-    \cite[Corollary II.24]{LOPEZ23b}, and more generally in the case 
-    of $\ZRat$ \cite[Corollary 3.6 p 38]{berstel2011noncommutative}.
+    \cite[Corollary II.24]{CDTL23}, and more generally in the case 
+    of $\ZRat$ \cite[Corollary 3.6 p 38]{BERE10}.
 \end{proof}
@@
 
-\section{Proofs of
-section \ref{polynomials:sec}}
+\section{Proofs of section \ref{polynomials:sec}}
 
@@
     For the effective part of the theorem, 
-    we know from \cite[Theorem V.13]{LOPEZ23b}, it is decidable whether
+    we know from \cite[Theorem V.13]{CDTL23}, it is decidable whether
     $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
@@
 
+\section{Proofs of section \ref{sec:ccl}}
+
 \begin{proof}[Proof of \cref{aperiodic-iff-residual:lem}]
@@
     $f \circ g \in \ZPoly[k+ \ell - 1]$
-    \cite{LOPEZ23b}.
+    \cite{CDTL23}.
 
-    Conversely, it was proven in \cite[Theorem III.3]{LOPEZ23b}
+    Conversely, it was proven in \cite[Theorem III.3]{CDTL23}
     that $f \in \ZPoly[k]$ if and only if
@@
     for all \kl{commutative} functions $h \in \SF$.
-    Using \cite[Theorem V.13]{LOPEZ23b},
+    Using \cite[Theorem V.13]{CDTL23},
     to conclude that $f \in \ZSF$,
@@
     Let us consider $f \in \ZPoly[k]$ that is polyblind.
-    Using \cite[Theorem 6.12]{gaetanphd},
+    Using \cite[Theorem 6.12]{DOUE23},
     we know that there exists a family of $\MSO$ formulas
@@
 
-    Using Parikh's Theorem \cite{parikh66context},
+    Using Parikh's Theorem \cite{PARI66},
     we conclude that the image of $\mathsf{s}$ is a semilinear set,
@@
     fact uniquely defined, this is done by computing an equivalent semilinear
-    set that is \emph{unambiguous} and \emph{proper} \cite{chistikov16}.
+    set that is \emph{unambiguous} and \emph{proper} \cite{CHHA16}.
     Similarly, let us define $\mathsf{encode} \colon \sum_{q \in Q} \Nat^{d_q}

### 2024-03-29 parts/conclusion.tex
rewrote intro

@@
 the introduction of the \emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of
-$(\resleq{f}{k})$, defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(v - u) \in
+$(\resleq{f}{k})$, defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u}) \in
 \NSF[k-1]$.
@@
     \label{z-poly-commutative-encoding:remark}
-    The conjecture \cref{z-poly-commutative-encoding:conj} holds for the
+    The \cref{z-poly-commutative-encoding:conj} holds for the
     subclass of polyblind functions when dropping the assumption that

### 2024-03-29 parts/correction.tex
rewrote intro

@@
 but we provide a self-contained proof using a refinement of the
-classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{LOPEZ23b} and
-$\NPoly$ \cite[Lemma 5.37]{gaetanphd}.
+classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and
+$\NPoly$ \cite[Lemma 5.37]{DOUE23}.
 
@@
 with \kl{maximal monomials} with respect to the \kl{Hoare ordering}: it
-extracts \emph{submaximal} monomials from a given polynomial.
+extracts \emph{sub-maximal} monomials from a given polynomial.
 

### 2024-03-29 parts/counter.tex
rewrote intro

@@
 The goal of this section is to provide semantic characterizations of
-\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is a
-\kl{$\Nat$-rational polynomial}, and if so, explicitly compute a
+\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is an
+\kl{$\Nat$-rational polynomial}, and if so, explicitly compute an
 \kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$. It is an easy
@@
 \cref{negative-but-npoly:ex} provides a polynomial with negative coefficients
-that is a \kl{$\Nat$-rational polynomial}. As a consequence, the class of
+that is an \kl{$\Nat$-rational polynomial}. As a consequence, the class of
 interest strictly contains $\Nat[\vec{X}]$.
@@
     Let $P \in \Nat[\vec{X}]$. Then, $P$
-    is a \kl{$\Nat$-rational polynomial}.
+    is an \kl{$\Nat$-rational polynomial}.
 \end{remark}
@@
     Let $P \defined X^2 -2X + 2$. Then $P$
-    is a \kl{$\Nat$-rational polynomial}.
+    is an \kl{$\Nat$-rational polynomial}.
 \end{example}
@@
 Let us now introduce the characterization of \kl{$\Nat$-rational polynomials}
-given by \citeauthor{KARH77}, which is stated using a class
+given by \cite{KARH77}, which is stated using a class
 $\CoveredPoly[\vec{X}]$ of polynomials with multiple indeterminate.
@@
     \begin{enumerate}
-        \item $P$ is a \kl{$\Nat$-rational polynomial},
+        \item $P$ is an \kl{$\Nat$-rational polynomial},
         \item $P \in \CoveredPoly$.
@@
     The polynomial $\BadPoly$ belongs to $\CoveredPoly$,
-    but is not a \kl{$\Nat$-rational polynomial}.
+    but is not an \kl{$\Nat$-rational polynomial}.
 \end{lemma}
@@
 
-    Assume by contradiction that $\BadPoly$ is a \kl{$\Nat$-rational polynomial}.
+    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
     Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.

### 2024-03-29 parts/intro.tex
correct citations
rewrote intro

@@
 \AP There have been recurrent attempts to generalize the notions of regularity
-from languages to functions, leading to classes of various expressiveness (for
-instance, sequential, rational, regular, and \kl{polyregular functions})
-\cite{ELME65,AHUL69,EILE74,ENMA02}. In this paper, we will focus on
-\intro{polyregular functions} ($\intro*\Poly$), that shares with regular
-languages its characterizations in terms of logic (via $\MSO$ interpretations),
-automata (via pebble automata), and adds new bridges with programming languages
-via its relationships with lambda-terms and so-called for-transducers
-\cite{BOKL19}. As for regular languages, the class $\intro*\SF$ of
-\intro{star-free polyregular function} can equivalently be introduced in terms
-of first-order definability, counter-free pebble transducers, and suitable
-restrictions of for-tranducers.
+from languages to functions, leading to classes of various expressiveness:
+Mealy Machines ($\intro*\Mealy$) \cite{MEAL55}, sequential functions
+($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
+\cite{EILE74}, regular functions ($\intro*\Regular$) \cite{ENMA02},
+and polyregular functions ($\intro*\Poly$) \cite{BOKL19}. For all of these
+models, an \emph{aperiodic} counterpart can be defined (respectively,
+$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
+$\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
+counter-free automata, star-free languages and aperiodic monoids to the
+functional setting \cite{FKT14,BOJA14,CADA15,BOKL19}. The inclusions between
+these classes of functions are all known to be strict, and we depicted in
+\cref{computational-models:fig} the status of the \emph{membership problem}
+associated to these strict inclusions (that is, given a function $f$, decide if
+$f$ can be computed by a function of a proper subclass). Note that in
+\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
+$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
+the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
+larger class $\mathsf{C}$. As an example, to conclude that a function can be
+computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
+enough to prove that it is sequential ($\Sequential$), and computed by an
+aperiodic polyregular function ($\SF$).
 
-\AP
-Classical problems, including the decidability of aperiodicity, become
-difficult in the case of \kl{polyregular functions}, due to the lack of
-canonical objects akin to the syntactic monoid or the minimal automaton. For
-sequential functions (that admit a canonical object), aperiodicity is decidable
-\cite{CHOF03}, and more recently, aperiodicity has been proven decidable
-for rational functions \cite{FGL16,FGLM18}. The
-problem remains open for regular functions, although it has been proven
-decidable when changing the semantics of function equality to account for the
-\emph{origins} in the output \cite{BOJA14}. To construct subclasses of
-\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
-of a function $f$, which is given by a number $k \in \Nat$, which can be used
-to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
-\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
-functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
-controls the maximal complexity of the $\MSO$ interpretation needed to
-represent the function \cite{BOJA22}.
+\AP To approach the \emph{aperiodicity} membership problems the traditional
+approach has been to develop so-called canonical objects, akin to the minimal
+automaton or the syntactic monoid of regular languages, over which aperiodicity
+is equivalent to a simple syntactic criterion (e.g. having no counters). This
+is the proof scheme that has been implemented for sequential functions
+\cite{CHOF03}, and more recently for rational functions \cite{FGL18}.
+Unfortunately, such a canonical object remains to be defined for regular and
+polyregular functions.
 
-\AP
-Another way to approach the problem of deciding if a \kl{polyregular function}
-is \kl{star-free}, is to not restrict the computational power (via the
-\kl{growth rate}), nor change the semantics (by adding origins), but consider
-restrictions of the output alphabet. In particular, there have been successful
-results proven on \emph{unary polyregular functions}, also known as
-\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
-\cite{DOUE21,DOUE22}. For instance, the \kl{growth
-rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
-needed to compute $f$ \cite{DOUE21}, which does not hold on $\Poly$
-\cite{BOJA22}. Let us remark that the restriction to
-\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
-outputs in the free commuative semigroup generated by a finite alphabet
-$\Gamma$. In this sense, the shift to \emph{unary output} is really about
-enforcing commutativity on the output.
+\begin{figure}
+    \centering
+    \begin{tikzpicture}[
+        xscale=1.2,
+        decidable/.style={
+            ->, thick, color=black
+        },
+        unknown/.style={
+            ->, dashed, color=gray
+        }
+        ]
+        % Create nodes for the classes and their aperiodic variants
+        % Mealy / sequential / rational / regular / polyregular
+        \node (mealy)  at (0, 2) {$\Mealy$};
+        \node (seq)    at (2, 2) {$\Sequential$};
+        \node (rat)    at (4, 2) {$\Rational$};
+        \node (reg)    at (6, 2) {$\Regular$};
+        \node (poly)   at (8, 2) {$\Poly$};
+        % Aperiodic variants
+        \node (amealy) at (0,0) {$\AMealy$};
+        \node (aseq)   at (2,0) {$\ASequential$};
+        \node (arat)   at (4,0) {$\ARational$};
+        \node (areg)   at (6,0) {$\ARegular$};
+        \node (apoly)  at (8,0) {$\SF$};
+        % Arrows
+        % The decidability is known for inclusions
+        % mealy -> sequential -> rational -> regular -> polyregular
+        % similarly for the aperiodic variants
+        \draw[decidable] (mealy) --
+            node[midway, above] {folklore}
+            (seq);
+        \draw[decidable] (seq)   -- 
+            node[midway, above] {\cite{CHOF03}}
+            (rat);
+        \draw[decidable] (rat)   -- 
+            node[midway, above] {\cite{FGRS13,BGMP15}}
+            (reg);
+        \draw[decidable] (reg)   -- 
+            node[midway, above] {\cite{BOKL19}}
+            (poly);
+        \draw[decidable] (amealy) --
+            (aseq);
+        \draw[decidable] (aseq) --
+            (arat);
+        \draw[decidable] (arat) --
+            (areg);
+        \draw[decidable] (areg) --
+            (apoly);
+        % Decidability of the aperidoic variants are known
+        % for mealy, sequential, rational but not for regular nor polyregular.
+        % we draw the arrows
+        \draw[decidable] (amealy) --
+            node[midway, left] {\cite{SCHU65,MNPA71}}
+            (mealy);
+        \draw[decidable] (aseq) -- 
+            node[midway, left] {\cite{CHOF03}}
+            (seq);
+        \draw[decidable] (arat) -- 
+            node[midway, left] {\cite{FGL16,FGLM18}}
+            (rat);
+        \draw[unknown] (areg) --
+            node[midway, left] {$\simeq$\cite{BOJA14}}
+            (reg);
+        \draw[unknown] (apoly) --
+            node[midway, left] {?}
+            (poly);
+    \end{tikzpicture}
+    \caption{
+        In this picture, arrows denote strict inclusions. Plain
+        arrows signify that the corresponding membership problem is decidable,
+        while dashed arrows signify that the corresponding membership problem
+        is conjectured to be decidable. In the case of \kl{regular functions}
+        ($\Regular$),
+        the decidability has only been obtained under a different notion
+        of equivalence of functions \cite{BOJA14}.
+    }
+    \label{computational-models:fig}
+\end{figure}
 
-\AP
-Recently, aperiodicity has been shown decidable over a restricted class of
-\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
-$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
-\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
-by allowing negative outputs, and therefore enabling a limited form of
-backtracking. As for the unary outputs, these results generalize to
-\kl{polyregular functions} having outputs in the free commutative \emph{group}
-generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
-\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
-object called the \emph{residual transducer}, the latter being essentially
-based on differences between functions, crucially leveraging negative outputs
-(i.e., the presence of inverse elements). Although the proof method does not carry
-from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
-that the semantic characterization of \kl{star-free $\Rel$-polyregular
-functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
-$\Nat$-polyregular functions} ($\intro*\NSF$).
+\AP In order to approach the decidability of decision problems on
+\kl{polyregular functions}, restricting the output alphabet to a single letter
+has proven to be a fruitful approach \cite{DOUE21,DOUE22}. The key advantage of
+single-letter output alphabets is that the particular order in which letters
+are produced by the machine is irrelevant. This allows to obtain results that
+would otherwise fail because of this ordering, such as the so-called
+\emph{pebble minimization theorem} of \cite{DOUE21}, which fails in the general
+case \cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary
+output} in the case of \kl{polyregular functions} can often be lifted to
+outputs in the free commutative semigroup generated by a finite alphabet
+$\Gamma$. In this sense, the shift to \emph{single letter output alphabet} is
+really about enforcing commutativity on the output. However, even with the
+restriction to single letter output alphabets, the decidability of
+\kl{star-free polyregular functions} among \kl{polyregular functions} remains
+an open problem \cite[Conjecture 7.61]{DOUE23}.
 
-\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
-    \label{zsf-nsf:conjecture}
-    Let $k \in \Nat$.
-    A function $f \in \NPoly[k]$
-    belongs to $\NSF[k]$ if and only if
-    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
-    In particular,
-    $\NSF[k] = \ZSF[k] \cap \NPoly$.
-\end{conjecture}
+\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
+as having outputs in $\Nat$, and this is why they were introduced respectively
+as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
+$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
+point of view, $\NPoly$ is a subclass of a larger and well-studied class of
+functions called (non-commutative) \intro{rational series}
+\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
+\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
+\kl{$\Nat$-rational series} that was frequently rediscovered in the literature
+\cite{SCHU62,KRRC13}. In \cite{CDTL23}, the authors introduced the class of
+\kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$) that generalize
+\kl{$\Nat$-polyregular functions} by allowing negative outputs. This remains a
+subclass of \kl{rational series}, more precisely, of the class of
+\kl{$\Rel$-rational series} ($\intro*\ZRat$). For this class of functions the
+aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
+Note however that $\ZPoly$ does not correspond to a restriction of
+\kl{polyregular functions}, because of the extra expressive power introduced by
+the negative outputs \cite{CDTL23}. Furthermore, deciding the membership of
+$\NRat$ inside $\ZRat$ is an open problem
+(see e.g. \cite{KARH77}),
+that remains open even in the
+restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}.
 
-\AP
-It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
-of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
-series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
-ambiguity, and a notion of \emph{quantitative} $\MSO$
-\cite{SCHU62,KRRC13}, and therefore can benefit from generic
-theorems obtained on these \cite{BERE10}. However, for these
-larger classes of functions, several competing notions of aperiodicity exist
-\cite{DRGA19,REUT80,CDTL23}, and the
-decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
-Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
-open problem, that was to the knowledge of the authors, only solved by
-\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
-\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
-$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
-\cite[Open question 5.55]{DOUE23}.
-
-\AP
-We refer the reader to \cref{previously-known-inclusions:fig} to get an
-overview of the different classes of functions appearing in the paper. Arrows
-denote strict inclusions, and effectiveness (both in terms of decidability and
-of effective representation) is represented by thick double arrows. Inclusions
-that are suspected to be effective are represented using a dashed arrow
-together with a question mark.
+\AP It was conjectured that, analogously to classes appearing in
+\cref{computational-models:fig}, the following equation would hold $\NSF =
+\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
+decidability of aperiodicity for single letter output polyregular functions
+since aperiodicity is decidable for $\ZSF$. We refer the reader to
+\cref{previously-known-inclusions:fig} to get an overview of the different
+classes of functions appearing in the paper. Arrows denote strict inclusions,
+and effectiveness (both in terms of decidability and of effective
+representation) is represented by thick double arrows. Inclusions that are
+suspected to be effective are represented using a dashed arrow together with a
+question mark.
 
@@
 characterization of polynomials computed by \kl{$\Nat$-rational series} given
-by \textcite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
+by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
 
@@
 \cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
-\cref{karh:thm} of \textcite{KARH77}, providing a counter-example in
+\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
 \cref{thm:counter-example}, and an alternative characterization in

### 2024-03-29 parts/prelim.tex
rewrote intro

@@
 \AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
-integers (resp. nonnegative integers). The capital letters $\Sigma,\Gamma$
+integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
 denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
@@
 ($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
-\cite{thomas1997languages}).
+\cite{THOM97}).
 
@@
 
+\AP
+% How the problem has been approached so far
+To construct subclasses of
+\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
+of a function $f$, which is given by a number $k \in \Nat$, which can be used
+to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
+\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
+functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
+controls the maximal complexity of the $\MSO$ interpretation needed to
+represent the function \cite{BOJA22}.
+
 \AP Because the functions of interest in this paper have output in $\Nat$ or
-$\Rel$, we will only provide the definiton of \kl{polyregular functions} for
-these two output semgroups, and we refer the reader to
-\cite{bojanczyk2019string} for the general definition of \kl{polyregular
-functions}. The following \cref{nat-rel-poly:def} is one of the equivalent
-definitions of \cite{LOPEZ23b}, and is similar in shape to the \emph{finite
-counting automata} introduced by \textcite{schutzenberger1962}. Let us recall
-that a monoid $M$ is \intro(monoid){aperiodic} whenever there exists $n \in
-\Nat$ such that for all $x \in M$, $x^{n+1} = x^n$.
+$\Rel$, we will only provide the definition of \kl{polyregular functions} for
+these two output semigroups, and we refer the reader to \cite{BOKL19} for the
+general definition of \kl{polyregular functions}. The following
+\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
+and is similar in shape to the \emph{finite counting automata} introduced by
+\cite{SCHU62}. Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
+whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
+x^n$.
 
-\begin{definition}[$\Rel$-polyregular functions {\cite{LOPEZ23b}}]
+\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
     \label{nat-rel-poly:def}
@@
     belongs to $\NSF$.
+    \proofref{size-of-word-nsf:ex}
 \end{example}
@@
 syntax to construct examples. These characterizations were previously known,
-and can be found for instance in \cite{gaetanphd,LOPEZ23b}.
+and can be found for instance in \cite{DOUE23,CDTL23}.
 
@@
             in $\set{1}^*$,
-            postcomposed with $\polysum$
+            post-composed with $\polysum$
             (resp. with output in $\set{-1,+1}^*$);
@@
 $k$ free variables. It has been proven in
-\cite{LOPEZ23b,bojanczyk2022transducers} that the classes $\ZPoly[k]$ (resp.
+\cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
 $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
 that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
-\bigO\left(\card{w}^k\right)$. and similarly for the other classes.
+\bigO\left(\card{w}^k\right)$, and similarly for the other classes.
 
@@
 coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[X_1, \dots, X_n]$ is a \intro{$\Nat$-rational polynomial} if it is
+\Rel[X_1, \dots, X_n]$ is an \intro{$\Nat$-rational polynomial} if it is
 \kl{represented} by a \kl{$\Nat$-polyregular function}. The analogue notion of
@@
     but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
-    not a \kl{$\Nat$-rational polynomial}.
+    not an \kl{$\Nat$-rational polynomial}.
 \end{example}
@@
     \label{pre-image-regular:fact}
-    The pre-image of a regular language by a \kl{$\Nat$-rational series}
+    The pre-image of a regular language by an \kl{$\Nat$-rational series}
     is a regular language. 
@@
     is \kl{non-negative}, but is
-    not a \kl{$\Nat$-rational polynomial}.
+    not an \kl{$\Nat$-rational polynomial}.
 \end{example}

### 2024-03-29 parts/starfree.tex
rewrote intro

--- 2024-02-14 parts/starfree.tex
@@
 \kl{commutativity}.
+
+\AP
+% Connection with Weighted Automata
+% introduce weighted automata and rational series here
+% but whether it can be leveraged remains unclear.
+Recently, aperiodicity has been shown decidable over a restricted class of
+\kl{polyregular functions} where the output belongs to $\Rel$ instead of
+$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
+\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
+by allowing negative outputs, and therefore enabling a limited form of
+backtracking. As for the unary outputs, these results generalize to
+\kl{polyregular functions} having outputs in the free commutative \emph{group}
+generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
+\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
+object called the \emph{residual transducer}, the latter being essentially
+based on differences between functions, crucially leveraging negative outputs
+(i.e., the presence of inverse elements). Although the proof method does not carry
+from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
+that the semantic characterization of \kl{star-free $\Rel$-polyregular
+functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
+$\Nat$-polyregular functions} ($\intro*\NSF$). 
+% UNCLEAR
+
+\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
+    \label{zsf-nsf:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \NPoly[k]$
+    belongs to $\NSF[k]$ if and only if
+    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
+    In particular,
+    $\NSF[k] = \ZSF[k] \cap \NPoly$.
+\end{conjecture}
+
+\AP
+It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
+of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
+series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
+ambiguity, and a notion of \emph{quantitative} $\MSO$
+\cite{SCHU62,KRRC13}, and therefore can benefit from generic
+theorems obtained on these \cite{BERE10}. However, for these
+larger classes of functions, several competing notions of aperiodicity exist
+\cite{DRGA19,REUT80,CDTL23}, and the
+decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
+Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
+open problem, that was to the knowledge of the authors, only solved by
+\cite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
+\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
+$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
+\cite[Open question 5.55]{DOUE23}.
 
@@
 of \emph{polynomial pumping arguments}
-\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
-semantic property in the form of \cref{k-combinatorial:def}, that is conjecture
+\cite{DOUE21,DOUE22,CDTL23}, we will provide a
+semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
 to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.
@@
 \kl{$\Nat$-polyregular functions} can be combined with the recent advances in
-the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to decide
+the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
 aperiodicity. The key ingredient of this study is the use of a semantic

### 2024-03-29 parts/transducers.tex
update ICALP
rewrote intro

@@
 \kl{$k$-residual transducer}, that generalizes the notion of residual
-transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{LOPEZ23b}.
+transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{CDTL23}.
 This notion of residual transducer has its own interest, as it shifts the
@@
 that is based on the marble transducers \cite{EHB99}, and follows the general
-pattern of transducers introduced in \cite{LOPEZ23b}. Note that we only
+pattern of transducers introduced in \cite{CDTL23}. Note that we only
 consider $\Nat$-valued functions, but could consider other commutative
@@
 
-\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{gaetanphd}}]
+\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
     \label{transducer-nsf-npoly:lemma}
@@
 denoted $\intro*\Res(f)$ and is defined as the set of $\reintro*\app{f}{u}$
-where $u$ ranges over words in $\Sigma^*$. Let $k \in \Nat$. We define $f
+where $u$ ranges over words in $\Sigma^*$. Given $k \in \Nat$, we define $f
 \intro*\zpolyequiv[k] g$ if and only if $(g - f) \in \ZPoly[k-1]$, and $f
-\intro*\npolyleq[k] g$ if and only if $(g - f) \in \NPoly[k-1]$. Given a
-function $f \colon \Sigma^* \to \Rel$, we are going to study both $(\Res(f),
+\intro*\npolyleq[k] g$ if and only if $(g - f) \in \NPoly[k-1]$. The spaces of
+interest for a function $f \in \ZPoly[k]$ will be $(\Res(f),
 \reintro*\zpolyequiv[k])$, and $(\Res(f), \reintro*\npolyleq[k])$. However, to

### 2024-03-29 parts/intro.tex
rewrite intro

@@
 is the proof scheme that has been implemented for sequential functions
-\cite{CHOF03}, and more recently for rational functions \cite{FGL18}.
+\cite{CHOF03}, and more recently for rational functions \cite{FGL16,FGLM18}.
 Unfortunately, such a canonical object remains to be defined for regular and
@@
 \AP It was conjectured that, analogously to classes appearing in
-\cref{computational-models:fig}, the following equation would hold $\NSF =
+\cref{computational-models:fig}, the following equation would hold: $\NSF =
 \NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
@@
 commutativity property both on the input alphabet and on the output alphabet.
-In this restricted setting, we answer positively to \cref{zsf-nsf:conjecture},
-and prove that the membership problems between $\NSF$, $\ZSF$, $\NPoly$, and
-$\ZPoly$ are all decidable, with effective conversion algorithms. This
-effectively shows that all arrows in the lower left square of
-\cref{previously-known-inclusions:fig} are decidable in the \kl{commutative}
-setting. Furthermore, we characterize \kl{commutative} \kl{$\Nat$-polyregular
-function} as the class of \kl{$\Rel$-polyregular functions} that are
-\kl{combinatorial} (defined page \kpageref{combinatorial}), which is a semantic
-property akin to the one \kl{ultimate polynomiality} that characterizes $\ZSF$.
-As a byproduct of our analysis, we provide a counter example to the
-characterization of polynomials computed by \kl{$\Nat$-rational series} given
-by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
+In this restricted setting, we answer positively to \cite[Conjecture
+7.61]{DOUE23} (restated in \cref{zsf-nsf:conjecture}), and prove that the
+membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
+decidable, with effective conversion algorithms. This effectively shows that
+all arrows in the upper left square of \cref{previously-known-inclusions:fig}
+are decidable in the \kl{commutative} setting. Furthermore, we characterize
+\kl{commutative} \kl{$\Nat$-polyregular function} as the class of
+\kl{$\Rel$-polyregular functions} that are \kl{combinatorial} (defined page
+\kpageref{combinatorial}), which is a semantic property akin to the one
+\kl{ultimate polynomiality} that characterizes $\ZSF$ inside $\ZPoly$
+\cite{CDTL23}. As a byproduct of our analysis, we provide a counter example to
+the characterization of polynomials computed by \kl{$\Nat$-rational series}
+given by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
 
@@
 \kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
-\cref{zsf-nsf:conjecture} holds for \kl{commutative} functions
+\cref{zsf-nsf:conjecture}
+\cite[Conjecture 7.61]{DOUE23} 
+holds for \kl{commutative} functions
 (\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we

### 2024-03-29 parts/prelim.tex
rewrite intro

@@
 ($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
-\cite{THOM97}).
+\cite{THOM97}). Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
+whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
+x^n$.
 
-\AP In order to build functions in $\NPoly$, the following constructions will
-be useful. For every language $L \subseteq \Sigma^*$ we define $\ind{L} \colon
-\Sigma^* \to \set{0,1}$ as the indicator function of $L$. Given an $\MSO$
-(resp. $\FO$) formula with first-order free variables $\varphi(\vec{x})$, we
-define $\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$
-that maps a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$,
-such that $w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from
-$\Sigma^*$ to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
-$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
-$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
-multiplication with $s \in \mathbb{S}$.
+\AP We also assume the reader to be familiar with rational series and weighted
+automata, and refer to \cite{BERE10} for a comprehensive introduction. Let us
+recall that one of the equivalent definitions of a $K$-rational series (where
+$K$ is a ring) is given by weighted automata, which are non-deterministic
+finite automata with transitions labelled by elements of $K$, and whose output
+on a given word is the sum over all accepting paths of the products of the
+weights along the path.
+
+\AP We use the notation $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the map
+that counts occurrences of every letter in the input word (that is, computes
+the Parikh vector) namely: $ \commute[w] \defined (a \mapsto \card[a]{w})$.
+Given a set $X$, a function $f \colon \Sigma^* \to X$ is \intro{commutative}
+whenever for all $u \in \Sigma^*$, for all permutations $\sigma$ of $\set{1,
+\dots, \card{w}}$, $f(\sigma(u)) = f(u)$. Equivalently, it is
+\reintro{commutative} whenever there exists a map $g \colon \Nat^\Sigma \to X$
+such that $g \circ \commute = f$. A function $g \colon \Nat^n \to X$ is
+\intro{represented} by a function $f \colon \Sigma^* \to X$ if there exists a
+map $\eta \colon \set{1, \dots, n} \to \Sigma$ such that $g \circ (v \colon
+\Nat^\Sigma \mapsto \seqof{v(\eta(k))}{1 \leq k \leq n} \colon \Nat^n) \circ
+\commute = f$. This notion will be useful to formally state that a polynomial
+``is'' a \kl{polyregular function}.
 
 \subsection{Polyregular Functions}
-
-\AP
-% How the problem has been approached so far
-To construct subclasses of
-\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
-of a function $f$, which is given by a number $k \in \Nat$, which can be used
-to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
-\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
-functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
-controls the maximal complexity of the $\MSO$ interpretation needed to
-represent the function \cite{BOJA22}.
 
@@
 and is similar in shape to the \emph{finite counting automata} introduced by
-\cite{SCHU62}. Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
-whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
-x^n$.
+\cite{SCHU62}.
 
@@
     be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a production function.
+    $\pi \colon M^k \to \Rel$ be a function.
     The \intro{$\Rel$-polyregular function}
@@
     is called \intro{$\Nat$-polyregular}.
+    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
     When the monoid $M$ is \kl(monoid){aperiodic}
@@
     belongs to $\NSF$.
-    \proofref{size-of-word-nsf:ex}
 \end{example}
+\begin{proof}
+    Let us define $M \defined (\set{1}, \times)$ which is 
+    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
+    defined by $\mu(w) \defined 1$, and
+    $\pi \colon M^2 \to \Nat$
+    that is the constant function equal to $1$.
+    We check that for all $w \in \Sigma^*$:
+    \begin{equation*}
+        \pi^\dagger(w)
+        =
+        \sum_{uv = w} 1
+        =
+        \card{w} + 1
+        = f(w)
+        \quad . 
+        \qedhere
+    \end{equation*}
+\end{proof}
 
-\AP Let us briefly recall in the following lemma alternative characterizations
+\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
+defined as the minimal $k$ such that $\card{f(w)} =
+\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
+$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp. $\reintro*\NPoly[k]$) for
+the functions in $\ZPoly$ (resp. $\NPoly$) that have growth rate at most $k$.
+It is clear that a function $f \in \ZPoly$ defined using a production function
+$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
+generally it is known that \kl{polyregular functions} have \kl{polynomial
+growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
+growth}, a typical example being the function $a^n \mapsto 2^n$.
+
+\AP Let us briefly recall in the following lemmas alternative characterizations
 of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
@@
 syntax to construct examples. These characterizations were previously known,
-and can be found for instance in \cite{DOUE23,CDTL23}.
+and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
+\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
+indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
+first-order free variables $\varphi(\vec{x})$, we define
+$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
+a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
+$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
+to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
+$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
+$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
+multiplication with elements $s \in \mathbb{S}$.
 
@@
 
-\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
-limiting the maximal number of free variables used to construct the function as
-a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
-$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
-$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
-$k$ free variables. It has been proven in
-\cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
-$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
-that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
-\bigO\left(\card{w}^k\right)$, and similarly for the other classes.
+It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
+$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
+of first order free variables used to construct the function as in
+\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.
 
-\subsection{Commutative Input}
-
-\AP 
-Let $\Sigma$ be a finite alphabet. 
-We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
-map that counts occurrences of every letter in the input word (that is,
-computes the Parikh vector) namely: $ \commute[w] \defined (a \mapsto
-\card[a]{w})$. Let $X$ be a set, a function $f \colon \Sigma^* \to X$ is
-\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
-Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
-\Nat^\Sigma \to X$ such that $g \circ \commute = f$.
-
-To ensure that we are working inside a decidable subclass of \kl{$\Rel$-rational series}
-let us state that commutativity is decidable. The proof is straightforward in the case
-of \kl{$\Rel$-polyregular functions}, and more involved for general \kl{$\Rel$-rational series}.
+Let us briefly state that \kl{commutativity} is a decidable property of
+\kl{$\Rel$-rational series}. The proof is straightforward in the case of
+\kl{$\Rel$-polyregular functions}, and slightly more involved for general
+\kl{$\Rel$-rational series}.
 
@@
 \end{lemma}
-
-\AP A function $g \colon \Nat^n \to \Rel$ is \intro{represented} by a function
-$f \colon \Sigma^* \to \Rel$ if there exists a map $\eta \colon \set{1, \dots,
-n} \to \Sigma$ such that $g \circ (v \mapsto \seqof{v_{\eta(k)}}{1 \leq k \leq
-n}) \circ \commute = f$. This notion will be useful to formally state that a
-polynomial ``is'' a \kl{polyregular function}.
 

### 2024-03-29 parts/correction.tex
big edits

@@
 
-\AP When multiple indeterminate are involved, the notion of degree must be
+\AP When multiple indeterminates are involved, the notion of degree must be
 adapted, and to ensure that the decision procedure is effective, one must
@@
 
-Let us now state, for the sake of a completeness, the equality of
-$\CorrectPoly$ and $\CoveredPoly$ in the case of two indeterminates. This is
-relevant because it shows that examples given in \cite{KARH77} are actually
-correct, even though based on an invalid result, and it may be the case for
-other works based on \cite{KARH77}.
+Let us now state, for the sake of completeness, the equality of $\CorrectPoly$
+and $\CoveredPoly$ in the case of two indeterminates. This is relevant because
+it shows that examples given in \cite{KARH77} are actually correct, even though
+based on an invalid result, and it may be the case for other works based on
+\cite{KARH77}.
 
@@
     The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminate, leading to a constant polynomial.
+    we fix the two indeterminates, leading to a constant polynomial.
 \end{proof}

### 2024-03-29 parts/intro.tex
big edits

@@
 \AP There have been recurrent attempts to generalize the notions of regularity
-from languages to functions, leading to classes of various expressiveness:
-Mealy Machines ($\intro*\Mealy$) \cite{MEAL55}, sequential functions
-($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
-\cite{EILE74}, regular functions ($\intro*\Regular$) \cite{ENMA02},
-and polyregular functions ($\intro*\Poly$) \cite{BOKL19}. For all of these
-models, an \emph{aperiodic} counterpart can be defined (respectively,
-$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
-$\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
-counter-free automata, star-free languages and aperiodic monoids to the
-functional setting \cite{FKT14,BOJA14,CADA15,BOKL19}. The inclusions between
+from languages to functions, leading to classes of various expressiveness, such
+as\footnote{This list is far from exhaustive.} Mealy Machines ($\intro*\Mealy$)
+\cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
+rational functions ($\intro*\Rational$) \cite{EILE74}, regular functions
+($\intro*\Regular$) \cite{ENMA02}, and polyregular functions ($\intro*\Poly$)
+\cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
+defined (respectively, $\intro*\AMealy$, $\intro*\ASequential$,
+$\intro*\ARational$, $\intro*\ARegular$, $\intro*\SF$), lifting the
+correspondence between counter-free automata, star-free languages and aperiodic
+monoids to the functional setting
+\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
 these classes of functions are all known to be strict, and we depicted in
@@
 \cref{previously-known-inclusions:fig} to get an overview of the different
-classes of functions appearing in the paper. Arrows denote strict inclusions,
-and effectiveness (both in terms of decidability and of effective
-representation) is represented by thick double arrows. Inclusions that are
-suspected to be effective are represented using a dashed arrow together with a
-question mark.
+classes of functions appearing in the paper.
 
@@
         rational series).
+        Arrows denote strict inclusions,
+        and effectiveness (both in terms of decidability and of effective
+        representation) is represented by thick double arrows. Inclusions that are
+        suspected to be effective are represented using a dashed arrow together with a
+        question mark.
     }

### 2024-03-29 parts/starfree.tex
big edits

@@
 \kl{commutativity}.
-
-\AP
-% Connection with Weighted Automata
-% introduce weighted automata and rational series here
-% but whether it can be leveraged remains unclear.
-Recently, aperiodicity has been shown decidable over a restricted class of
-\kl{polyregular functions} where the output belongs to $\Rel$ instead of
-$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
-\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
-by allowing negative outputs, and therefore enabling a limited form of
-backtracking. As for the unary outputs, these results generalize to
-\kl{polyregular functions} having outputs in the free commutative \emph{group}
-generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
-\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
-object called the \emph{residual transducer}, the latter being essentially
-based on differences between functions, crucially leveraging negative outputs
-(i.e., the presence of inverse elements). Although the proof method does not carry
-from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
-that the semantic characterization of \kl{star-free $\Rel$-polyregular
-functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
-$\Nat$-polyregular functions} ($\intro*\NSF$). 
-% UNCLEAR
-
-\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
-    \label{zsf-nsf:conjecture}
-    Let $k \in \Nat$.
-    A function $f \in \NPoly[k]$
-    belongs to $\NSF[k]$ if and only if
-    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
-    In particular,
-    $\NSF[k] = \ZSF[k] \cap \NPoly$.
-\end{conjecture}
-
-\AP
-It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
-of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
-series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
-ambiguity, and a notion of \emph{quantitative} $\MSO$
-\cite{SCHU62,KRRC13}, and therefore can benefit from generic
-theorems obtained on these \cite{BERE10}. However, for these
-larger classes of functions, several competing notions of aperiodicity exist
-\cite{DRGA19,REUT80,CDTL23}, and the
-decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
-Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
-open problem, that was to the knowledge of the authors, only solved by
-\cite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
-\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
-$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
-\cite[Open question 5.55]{DOUE23}.
 
@@
 
+\AP
+The decidability of aperiodicity for
+\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
+object called the \emph{residual transducer}, the latter being essentially
+based on differences between functions, crucially leveraging negative outputs
+(i.e., the presence of inverse elements). Although the proof method does not carry
+from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
+that the semantic property of being \kl{ultimately polynomial}
+would also characterize $\NSF$ inside $\NPoly$, a conjecture that
+is restated hereafter.
+
+\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
+    \label{zsf-nsf:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \NPoly[k]$
+    belongs to $\NSF[k]$ if and only if
+    it is \kl{ultimately polynomial} (defined page \kpageref{ultimately polynomial}).
+    In particular,
+    $\NSF[k] = \ZSF[k] \cap \NPoly$.
+\end{conjecture}
+
+\AP
+It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
+of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
+series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
+ambiguity, and a notion of \emph{quantitative} $\MSO$
+\cite{SCHU62,KRRC13}, and therefore can benefit from generic
+theorems obtained on these \cite{BERE10}. However, for these
+larger classes of functions, several competing notions of aperiodicity exist
+\cite{DRGA19,REUT80,CDTL23}, and the
+decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
+Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
+open problem, that was to the knowledge of the authors, only solved by
+\cite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
+\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
+$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
+\cite[Open question 5.55]{DOUE23}.
+
 \begin{theorem}

### 2024-03-29 parts/appendix.tex
rewrite intro
appendix + proofrefs

@@
     if it belongs to the Zariski closure, which is decidable
-    \cite{HROUPOWO18}.
+    \cite{HOPW18}.
 \end{proof}
 
-\begin{proof}[Proof of \cref{negative-not-nrat:ex}]
+\begin{proof}[A simpler proof of \cref{decidable-commutative-rat:lemma} in the general case]
+    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
+    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
+    As a consequence, a function $f$ is commutative if and only if
+    $f \circ c = f = f \circ t$.
+    When $f$ is a rational series,
+    $f \circ c$ and $f \circ t$ are both rational series that can be
+    effectively computed from $f$, and since equivalence
+    of rational series is decidable 
+    \cite{BERE10}
+    we have obtained a decision procedure.
+\end{proof}
+
+\begin{proofof}[negative-not-nrat:ex]
     The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
@@
     are non-negative.
-\end{proof}
+\end{proofof}
 
@@
 
-\begin{proof}[Proof of
-    \cref{n-poly-n-poly:example}]
+\begin{proofof}[n-poly-n-poly:example]
     The polynomials of $\Nat[\vec{X}]$
@@
     that are closed under these operations, we conclude.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of
-    \cref{n-poly-combinatorics:lem}
-    ]
-
+\begin{proofof}[n-poly-combinatorics:lem]
     Let $\omega$ be an idempotent power for the finite monoid $M$,
@@
     which has the desired shape.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{n-rat-correct:lem}]
+\begin{proofof}[n-rat-correct:lem]
     Let $f$ be the \kl{commutative}
@@
     we have proven that $P \in \CorrectPoly$.
-\end{proof}
+\end{proofof}
 
@@
 
-\begin{proof}[Proof of 
-    \cref{derivation-monomials:fact}
-    ]
+\begin{proofof}[derivation-monomials:fact]
     Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
@@
     $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{derivation-simplifies:lemma}]
+\begin{proofof}[derivation-simplifies:lemma]
     Let $P \in \Rel[\vec{X}]$ and
@@
     \end{align*}
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{derivation-stabilises-correct:lem}]
+\begin{proofof}[derivation-stabilises-correct:lem]
     If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
@@
     \end{align*}
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{derivation-translation:lem}]
+\begin{proofof}[derivation-translation:lem]
 
@@
     belong to $\Nat[\vec{X}]$, which is decidable.
-\end{proof}
+\end{proofof}
+
+\begin{proofof}[lem:correct-covered-2]
+    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
+    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
+    For the converse inclusion, let us consider $P(X,Y)$
+    that is \kl{non-negative}, such that the \kl{maximal monomials}
+    are  \kl{non-negative}.
+
+    If we fix none of the variables, then the \kl{maximal monomials}
+    are \kl{non-negative} by assumption. If we fix one of the
+    variables, we can assume without loss of generality that we 
+    fix $X = k$ for some $k \in \Nat$.
+    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
+    and therefore must have a positive leading coefficient
+    (which is the unique \kl{maximal monomial} in this case)
+    or be constant equal to 0. In both cases, the \kl{maximal monomials}
+    have positive coefficient.
+    The same reasoning applies \emph{a fortiori} in the case where
+    we fix the two indeterminates, leading to a constant polynomial.
+\end{proofof}
 

### 2024-03-29 parts/conclusion.tex
appendix + proofrefs

@@
 
-In the commutative case, the conjectured semantic characterizations of $\NSF$
-and $\NPoly$ have been proven to hold. Furthermore, the existence of a
-canonical transducer computing functions in $\NPoly$ opens the door to several
-questions with respect to aperiodicity. We can prove that for functions in
-$\NPoly[0]$, aperiodicity is equivalent to the absence of counters in the
-\kl{residual transducer} of the function (see
-\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
-higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
-In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number
-of states computing the function is \kl{counter-free}, which suggests that a
-more sophisticated object is needed to observe aperiodicity.
+Let us conclude with a more general remark on the status of commutative input
+in the study of unary output polyregular functions. Notice that all semantic
+characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
+polynomial}, pumpable) arise from pre-composition with a \kl{commutative}
+\kl{star-free polyregular function}, as formally stated in
+\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}. Following
+this idea, we conjecture that $\ZPoly$ and $\NPoly$ can be described as
+combinations of commutative functions, leading to
+\cref{z-poly-commutative-encoding:conj}. This conjecture would play the role of
+\cref{decompose-polynomial:lem} in the study of functions in $\ZPoly$.
 
 \begin{lemma}
-    \label{aperiodic-iff-residual:lem}
-    Let $f \in \NPoly[0]$, the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NSF$,
-        \item $f \in \ZSF$,
-        \item $f$ is \kl{ultimately polynomial},
-        \item The \kl{$0$-residual transducer} of $f$ is \kl{counter-free}.
-    \end{enumerate}
-\end{lemma}
-
-\begin{example}
-    \label{non-aperiodic-residual-transd:ex}
-    Let us define
-    $\BadExKo(\varepsilon) = 1$,
-    $\BadExKo(a) = 0$,
-    $\BadExKo(a^2) = 1$,
-    and $\BadExKo(a^n) = n - 3$ for all $n \geq 3$.
-    The \kl{$0$-residual transducer} of $\BadExKo$ has a \kl{counter} and two states.
-    Furthermore,
-    every \kl{$\NPoly[0]$-transducer} with two states contains a \kl{counter}.
-\end{example}
-
-\AP In the spirit of the characterization of $\NPoly$ based the
-$(\resleq{f}{k})$ quasi-ordering (\cref{non-commutative-npoly:thm}), we
-introduce the notion of \intro{aperiodic ordering} of $\Sigma^*$ as follows:
-$(\Sigma^*, \leq)$ is \reintro[aperiodic ordering]{aperiodic} whenever for all
-$u, w \in \Sigma^*$, there exists $N_0 \in \Nat$, such that the sequence
-$\seqof{uw^n}{n \geq N_0}$ is increasing. This leads to the following
-\cref{sf-no-periods-on-sequences:conj}, of which the easy implication is
-already known to hold (\cref{sf-no-periods-on-sequences:lemma}), and requires
-the introduction of the \emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of
-$(\resleq{f}{k})$, defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u}) \in
-\NSF[k-1]$.
-
-\begin{lemma}
-    \label{sf-no-periods-on-sequences:lemma}
-    Let $k \in \Nat$, and $f \in \SF[k]$. Then,
-    $(\Sigma^*, \resleqsf{f}{k})$ is an
-    \kl[aperiodic ordering]{aperiodic} \kl{well-quasi-ordering}.
-\end{lemma}
-
-\begin{conjecture}
-    \label{sf-no-periods-on-sequences:conj}
-    For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
-    $f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
-    \kl{well-quasi-ordering}.
-\end{conjecture}
-
-Let us conclude with a more general remark on the status of commutative input
-in the study of polyregular functions. Notice that semantic characterizations
-of subclasses of \kl{polyregular functions} (\kl{combinatorial}, \kl{ultimately
-polynomial}, pumpable) arise from pre-composition with a \kl{commutative}
-\kl{star-free polyregular function}, as formally stated in
-\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
-Generalizing this idea, we conjecture that an effective encoding of $\ZPoly$ in
-terms of commutative functions could be obtained, as stated in
-\cref{z-poly-commutative-encoding:conj}, playing the same role as the
-decomposition obtained in \cref{decompose-polynomial:lem} in the
-\kl{commutative} case.
-
-\begin{remark}
     \label{pre-compose-growth-commut:lemma}
@@
             $(f \circ h) \in \ZPoly[k+l-1]$.
-\end{remark}
+\end{lemma}
 
-\begin{remark}
+\begin{lemma}
     \label{pre-compose-sf-commut:lemma}
@@
             $(f \circ h) \in \ZSF$.
-\end{remark}
+\end{lemma}
 
@@
 and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
-Although \cref{z-poly-commutative-encoding:remark} is not yet applicable in
-practice, we consider this partial result as a strong indicator in favor of the
-conjecture.
 
-\begin{remark}
+\begin{lemma}
     \label{z-poly-commutative-encoding:remark}
@@
     (resp. \kl{star-free polyregular functions}).
-\end{remark}
+\end{lemma}

### 2024-03-29 parts/correction.tex
appendix + proofrefs

@@
 
+\AP
 The counter example provided by \cref{def:bad-polynomial} relies on the fact
@@
 \cref{karh:thm}.
-
-\AP We use the following notation to fix the value of some indeterminate, if
+We use the following notation to fix the value of some indeterminate, if
 $P(X,Y)$ is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is
@@
     \label{n-poly-combinatorics:lem}
-    Let $f$ be a  \kl{$\Nat$-polyregular} function. 
+    Let $f$ be an \kl{$\Nat$-polyregular} function. 
     There exists a computable $\omega \in \Nat$
@@
     \end{equation*}
+    \proofref{n-poly-combinatorics:lem}
 \end{lemma}
@@
     \label{n-rat-correct:lem}
-    Let $P \in \Rel[X_1, \dots, X_p]$ be \kl{represented}
-    by a \kl{$\Nat$-rational series}. Then,
+    Let $P \in \Rel[X_1, \dots, X_p]$ be an \kl{$\Nat$-rational polynomial}.
+    Then,
     $P \in \CorrectPoly$.
+    \proofref{n-rat-correct:lem}
 \end{corollary}
@@
     that is a multiple of $K$.
+    \proofref{derivation-monomials:fact}
 \end{fact}
@@
     \MaximalMonomials(P)$.
+    \proofref{derivation-simplifies:lemma}
 \end{lemma}
@@
     $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$.
+    \proofref{derivation-stabilises-correct:lem}
 \end{lemma}
@@
     Furthermore, the membership is decidable.
+    \proofref{derivation-translation:lem}
 \end{lemma}
@@
 based on an invalid result, and it may be the case for other works based on
-\cite{KARH77}.
+\cite{KARH77}. The proof is based the fact that for (non-negative) univariate
+it is enough that the dominant coefficient is non-negative to be an
+\kl{$\Nat$-rational polynomial}.
 
@@
     $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
+    \proofref{lem:correct-covered-2}
 \end{lemma}
-\begin{proof}
-    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
-    For the converse inclusion, let us consider $P(X,Y)$
-    that is \kl{non-negative}, such that the \kl{maximal monomials}
-    are  \kl{non-negative}.
-
-    If we fix none of the variables, then the \kl{maximal monomials}
-    are \kl{non-negative} by assumption. If we fix one of the
-    variables, we can assume without loss of generality that we 
-    fix $X = k$ for some $k \in \Nat$.
-    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
-    and therefore must have a positive leading coefficient
-    (which is the unique \kl{maximal monomial} in this case)
-    or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have positive coefficient.
-    The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminates, leading to a constant polynomial.
-\end{proof}

### 2024-03-29 parts/intro.tex
appendix + proofrefs

@@
 aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
-Note however that $\ZPoly$ does not correspond to a restriction of
-\kl{polyregular functions}, because of the extra expressive power introduced by
-the negative outputs \cite{CDTL23}. Furthermore, deciding the membership of
-$\NRat$ inside $\ZRat$ is an open problem
-(see e.g. \cite{KARH77}),
-that remains open even in the
-restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}.
+Beware that concurrent (non-equivalent) notions of aperiodicity exists in the
+research area of rational series \cite{REUT80,DRGA19,CDTL23}, and that we only
+refer to the one applying to rational series of polynomial growths introduced
+in \cite{CDTL23}.
 
-\AP It was conjectured that, analogously to classes appearing in
+\AP This approach through rational series has several limitations. First,
+$\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
+because of the extra expressive power introduced by the negative outputs
+\cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
+is an open problem (see e.g. \cite{KARH77}), that remains open even in the
+restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
+was conjectured that, analogously to classes appearing in
 \cref{computational-models:fig}, the following equation would hold: $\NSF =

### 2024-03-29 parts/prelim.tex
appendix + proofrefs

@@
     belongs to $\NSF$.
+    \proofref{size-of-word-nsf:ex}
 \end{example}
-\begin{proof}
-    Let us define $M \defined (\set{1}, \times)$ which is 
-    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
-    defined by $\mu(w) \defined 1$, and
-    $\pi \colon M^2 \to \Nat$
-    that is the constant function equal to $1$.
-    We check that for all $w \in \Sigma^*$:
-    \begin{equation*}
-        \pi^\dagger(w)
-        =
-        \sum_{uv = w} 1
-        =
-        \card{w} + 1
-        = f(w)
-        \quad . 
-        \qedhere
-    \end{equation*}
-\end{proof}
 
@@
 
-Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
-generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
-
 \begin{lemma}
@@
 
-\begin{example}
-    \label{regular-language:ex}
-    Let $\Sigma$ be a finite alphabet, and
-    $L \subseteq \Sigma^*$. Then,
-    $L$ is a regular language if and only if
-    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
-    \kl{$\Nat$-polyregular function}.
-    Furthermore, $L$ is a star-free regular language
-    if and only if $\ind{L}$ is a
-    \kl{star-free $\Nat$-polyregular function}.
-\end{example}
-
-It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
-$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
-of first order free variables used to construct the function as in
-\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.
+%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
+%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
+%\begin{example}
+%    \label{regular-language:ex}
+%    Let $\Sigma$ be a finite alphabet, and
+%    $L \subseteq \Sigma^*$. Then,
+%    $L$ is a regular language if and only if
+%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
+%    \kl{$\Nat$-polyregular function}.
+%    Furthermore, $L$ is a star-free regular language
+%    if and only if $\ind{L}$ is a
+%    \kl{star-free $\Nat$-polyregular function}.
+%\end{example}
+%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
+%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
+%of first order free variables used to construct the function as in
+%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.
 
@@
     not an \kl{$\Nat$-rational polynomial}.
+    \proofref{negative-not-nrat:ex}
 \end{example}

### 2024-03-29 parts/starfree.tex
appendix + proofrefs

--- 2024-02-14 parts/starfree.tex
@@
 of \emph{polynomial pumping arguments}
-\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
+\cite{DOUE21,DOUE22,CDTL23}, we will provide a
 semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
@@
 
+\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}
+relies on the construction of a canonical object called the \emph{residual
+transducer}, the latter being essentially based on differences between
+functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
+proof method does not carry from $\Rel$-output functions to $\Nat$-output
+functions, it was conjectured that the semantic property of being
+\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
+conjecture that is restated hereafter.
+
+\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
+    \label{zsf-nsf:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \NPoly[k]$
+    belongs to $\NSF[k]$ if and only if
+    it is \kl{ultimately polynomial}.
+    In particular,
+    $\NSF[k] = \ZSF[k] \cap \NPoly$.
+\end{conjecture}
+
+We answer positively to \cref{zsf-nsf:conjecture} in the commutative case, by
+leveraging the semantic characterizations respectively of $\ZSF$ inside
+$\ZPoly$ (\kl{ultimately polynomial}) and $\NPoly$ inside $\ZPoly$
+(\kl{$(k,\Nat)$-combinatorial}), which is possible thanks to the decomposition
+obtained in \cref{decompose-polynomial:lem}. Decidability then follows
+immediately from the effective conversions, and the previous decidability
+result for $\ZSF$.
+
 \begin{theorem}

### 2024-03-29 parts/transducers.tex
appendix + proofrefs

@@
 that is based on the marble transducers \cite{EHB99}, and follows the general
-pattern of transducers introduced in \cite{CDTL23}. Note that we only
-consider $\Nat$-valued functions, but could consider other commutative
-semigroups.
+pattern of transducers introduced in \cite{CDTL23}. Note that we only consider
+$\Nat$-valued functions, but could consider other commutative semigroups.
 
@@
 \end{definition}
-
-Given the syntax of an \kl{$\mathcal{H}$-transducer}, let us now 
-introduce its semantics, which is given by a simple recursion
-scheme. Intuitively, the function computed by a transducer is
 
@@
     Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
-    The following are equivalent:
-    \begin{itemize}
-        \item $f \in \NPoly[k]$ (respectively $\NSF[k]$),
-        \item $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
+    Then, 
+    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
+    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
         (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
         \kl{counter-free}).
-    \end{itemize}
 \end{lemma}
@@
 \end{proof}
+
+\AP
+We can prove that for functions in $\NPoly[0]$, aperiodicity is equivalent to
+the absence of counters in the \kl{residual transducer} of the function (see
+\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
+higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
+In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
+states computing the function is \kl{counter-free}, which suggests that a more
+sophisticated object is needed to observe aperiodicity. In the spirit of the
+characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
+(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
+ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
+ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
+\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
+This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
+the easy implication is already known to hold
+(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
+\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
+defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
+\in \NSF[k-1]$.
+
+\begin{lemma}
+    \label{aperiodic-iff-residual:lem}
+    Let $f \in \NPoly[0]$. Then,
+    $f \in \NSF$ if and only if 
+    $f \in \ZSF$, if and only if
+    $f$ is \kl{ultimately polynomial}, if and only if 
+    the \kl{$0$-residual transducer} of $f$ is \kl{counter-free}.
+\end{lemma}
+
+\begin{example}
+    \label{non-aperiodic-residual-transd:ex}
+    Let us define
+    $\BadExKo(\varepsilon) = 1$,
+    $\BadExKo(a) = 0$,
+    $\BadExKo(a^2) = 1$,
+    and $\BadExKo(a^n) = n - 3$ for all $n \geq 3$.
+    The \kl{$0$-residual transducer} of $\BadExKo$ has a \kl{counter} and two states.
+    Furthermore,
+    every \kl{$\NPoly[0]$-transducer} with two states contains a \kl{counter}.
+\end{example}
+
+\begin{lemma}
+    \label{sf-no-periods-on-sequences:lemma}
+    Let $k \in \Nat$, and $f \in \SF[k]$. Then,
+    $(\Sigma^*, \resleqsf{f}{k})$ is an
+    \kl[aperiodic ordering]{aperiodic} \kl{well-quasi-ordering}.
+\end{lemma}
+
+\begin{conjecture}
+    \label{sf-no-periods-on-sequences:conj}
+    For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
+    $f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
+    \kl{well-quasi-ordering}.
+\end{conjecture}

### 2024-03-29 parts/appendix.tex
mise en forme

@@
 
-\begin{proof}[Proof of \cref{decompose-polynomial:lem}]
+\begin{proofof}[decompose-polynomial:lem]
     We apply
@@
     to be below the threshold $\omega^2$.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{decidable-n-poly:thm}]
+\begin{proofof}[decidable-n-poly:thm]
     Let us prove that
@@
     which is possible in $\NPoly$.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{zsf-npoly-nsf:thm}]
+\begin{proofof}[zsf-npoly-nsf:thm]
     We proceed by induction on the size of the alphabet $\card{\Sigma}$.
@@
     \kl{star-free $\Nat$-polyregular function}.
-\end{proof}
+\end{proofof}
 
@@
 
-\begin{proof}[Proof of \cref{correct-residual:lemma}]
+\begin{proofof}[correct-residual:lemma]
     Because of \cref{q-o-prefix-cool:fact},
@@
     is indeed computing correctly.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{wqo-implies-termination:lemma}]
+\begin{proofof}[wqo-implies-termination:lemma]
     Assume towards a contradiction that
@@
     Hence, \cref{residual:algo} must terminate.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{n-poly-k-implies-wqo:lemma}]
+\begin{proofof}[n-poly-k-implies-wqo:lemma]
     Because $f \in \NPoly[k]$, there exists
@@
     exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
-\end{proof}
+\end{proofof}
 
@@
 
-\begin{proof}[Proof of \cref{aperiodic-iff-residual:lem}]
+\begin{proofof}[aperiodic-iff-residual:lem]
     It is clear that $\NSF \subseteq \ZSF$, and known that if $f \in \ZSF$
@@
     hence equal, and therefore $\delta(q, w) = q$.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof sketch of \cref{sf-no-periods-on-sequences:lemma}]
+\begin{proofof}[sf-no-periods-on-sequences:lemma]
     Let $f \in \NSF[k]$, $q \in \Nat$, and 
@@
     of those, and conclude.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of
-    \cref{pre-compose-growth-commut:lemma}]
+\begin{proofof}[pre-compose-growth-commut:lemma]
     If $f \in \ZPoly[k]$, then 
@@
     holds, i.e. that $f \in \ZPoly[k]$.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of
-    \cref{pre-compose-sf-commut:lemma}]
+\begin{proofof}[pre-compose-sf-commut:lemma]
     If $f \in \ZSF$, then for all $h \in \SF$,
@@
     $f \in \ZSF$.
-\end{proof}
+\end{proofof}
 
-\begin{proof}[Proof of \cref{z-poly-commutative-encoding:remark}]
+\begin{proofof}[z-poly-commutative-encoding:remark]
     We are going to prove that if $f \in \ZPoly[k]$ is polyblind, 
@@
     \kl{star-free}.
-\end{proof}
+\end{proofof}

### 2024-03-29 parts/prelim.tex
mise en forme

@@
 \intro{maximal monomial} if it is a maximal element for the \kl{divisibility
-ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + X^2
+ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
 + X + Y$, the set of \reintro{maximal monomials}, written
-$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  X^2}$.  In the polynomial
+$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  Y^2}$.  In the polynomial
 $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials} are $X^2$ and

### 2024-04-02 parts/intro.tex
typos

@@
 \cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
-rational functions ($\intro*\Rational$) \cite{EILE74}, regular functions
-($\intro*\Regular$) \cite{ENMA02}, and polyregular functions ($\intro*\Poly$)
+rational functions ($\intro*\Rational$) \cite{EILE74}, \intro{regular functions}
+($\intro*\Regular$) \cite{ENMA02}, and \intro{polyregular functions} ($\intro*\Poly$)
 \cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
@@
 enough to prove that it is sequential ($\Sequential$), and computed by an
-aperiodic polyregular function ($\SF$).
+\intro{aperiodic polyregular function} ($\SF$).
 
@@
 \kl{$\Nat$-polyregular functions} correspond to a natural subclass of
-\kl{$\Nat$-rational series} that was frequently rediscovered in the literature
-\cite{SCHU62,KRRC13}. In \cite{CDTL23}, the authors introduced the class of
-\kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$) that generalize
-\kl{$\Nat$-polyregular functions} by allowing negative outputs. This remains a
-subclass of \kl{rational series}, more precisely, of the class of
-\kl{$\Rel$-rational series} ($\intro*\ZRat$). For this class of functions the
-aperiodic variant $\intro*\ZSF$ has been shown to be decidable \cite{CDTL23}.
-Beware that concurrent (non-equivalent) notions of aperiodicity exists in the
-research area of rational series \cite{REUT80,DRGA19,CDTL23}, and that we only
-refer to the one applying to rational series of polynomial growths introduced
-in \cite{CDTL23}.
+\intro{$\Nat$-rational series} ($\intro*\NRat$) that was frequently
+rediscovered in the literature \cite{SCHU62,KRRC13}. In \cite{CDTL23}, the
+authors introduced the class of \kl{$\Rel$-polyregular functions}
+($\intro*\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
+negative outputs. This remains a subclass of \kl{rational series}, more
+precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
+this class of functions the aperiodic variant $\intro*\ZSF$ has been shown to
+be decidable \cite{CDTL23}. Beware that several non-equivalent notions of
+aperiodicity coexist for rational series \cite{REUT80,DRGA19,CDTL23}, and that
+we only refer to the one applying to rational series of polynomial growths
+introduced in \cite{CDTL23}.
 
@@
         arranged along two axes. The first one is the complexity
-        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
+        of the output alphabet ($\Rel$, $\Nat$, $\Sigma$). The second
         one is the allowed computational power
-        (star-free polyregular functions, polyregular functions, 
-        rational series).
+        (\kl{star-free polyregular functions}, \kl{polyregular functions}, 
+        \kl{rational series}).
         Arrows denote strict inclusions,

### 2024-04-02 parts/prelim.tex
typos

@@
 ($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
-\cite{THOM97}). Let us recall that a monoid $M$ is \intro(monoid){aperiodic}
-whenever for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} =
-x^n$.
+\cite{THOM97}). As aperiodicity will be a central notion of this paper, let us
+recall that a monoid $M$ is \intro(monoid){aperiodic} whenever for all $x \in
+M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.
 
-\AP We also assume the reader to be familiar with rational series and weighted
-automata, and refer to \cite{BERE10} for a comprehensive introduction. Let us
-recall that one of the equivalent definitions of a $K$-rational series (where
-$K$ is a ring) is given by weighted automata, which are non-deterministic
-finite automata with transitions labelled by elements of $K$, and whose output
-on a given word is the sum over all accepting paths of the products of the
-weights along the path.
+\AP We also assume the reader to be familiar with \kl{rational series} and
+weighted automata, and refer to \cite{BERE10} for a comprehensive introduction.
+Let us recall that one of the equivalent definitions of a $K$-rational series
+(where $K$ is a ring) is given by the functions from $\Sigma^*$ to $K$ computed
+by weighted automata, which are non-deterministic finite automata with
+transitions labelled by elements of $K$, and whose output on a given word is
+the sum over all accepting paths of the products of the weights along the path.
 

### 2024-04-22 parts/correction.tex
moving two variables paragraphs

@@
 
-Let us now state, for the sake of completeness, the equality of $\CorrectPoly$
-and $\CoveredPoly$ in the case of two indeterminates. This is relevant because
-it shows that examples given in \cite{KARH77} are actually correct, even though
-based on an invalid result, and it may be the case for other works based on
-\cite{KARH77}. The proof is based the fact that for (non-negative) univariate
-it is enough that the dominant coefficient is non-negative to be an
-\kl{$\Nat$-rational polynomial}.
+The counter example of \cref{thm:counter-example} used three indeterminates,
+and we will prove in \cref{sec:proof} that \cref{karh:thm} holds on polynomial
+with at most two indeterminate. In particular, the examples appearing in
+\cite{KARH77} are not invalidated, as they all use at most two indeterminates.
+In particular, we derive the equality of $\CorrectPoly$ and $\CoveredPoly$ in
+the case of two indeterminates.  The proof is based the fact that for
+(non-negative) univariate it is enough that the dominant coefficient is
+non-negative to be an \kl{$\Nat$-rational polynomial}.
 

### 2024-04-22 parts/counter.tex
appendix + proofrefs
typos
moving two variables paragraphs

@@
 
-\begin{remark}
+\begin{lemma}
     \label{n-poly-n-poly:example}
@@
     is an \kl{$\Nat$-rational polynomial}.
-\end{remark}
+    \proofref{n-poly-n-poly:example}
+\end{lemma}
 
@@
     is an \kl{$\Nat$-rational polynomial}.
+    \proofref{negative-but-npoly:ex}
 \end{example}
@@
     Let $\Sigma$ be a finite alphabet
-    and $P$ be a polynomial. The following are equivalent:
-    \begin{enumerate}
-        \item $P$ is an \kl{$\Nat$-rational polynomial},
-        \item $P \in \CoveredPoly$.
-    \end{enumerate}
+    and $P$ be a polynomial. Then,
+    $P$ is an \kl{$\Nat$-rational polynomial}
+    if and only if 
+    $P \in \CoveredPoly$.
 \end{faketheorem}
@@
 
-Let us now provide a counter example to \cref{karh:thm}. The counter example
-will use three indeterminates, and we will prove in \cref{sec:proof} that
-\cref{karh:thm} holds on polynomial with at most two indeterminate. In
-particular, the examples appearing in \cite{KARH77} are not invalidated, as
-they all use at most two indeterminates. Furthermore, the converse implication,
-that is, that \kl{$\Nat$-rational polynomials} belong to $\CoveredPoly$ is
-correct, and can even be strengthened as we will see in
-\cref{corrected-version:thm}.
+Let us now provide a counter example to \cref{karh:thm}.
 

### 2024-04-22 parts/appendix.tex
typos
moving remark to appendix

@@
 
-\begin{proof}[Proof of \cref{negative-but-npoly:ex}]
+\begin{proofof}[negative-but-npoly:ex]
     Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
@@
     $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
-\end{proof}
+\end{proofof}
 
@@
 \end{proofof}
+
+As one would expect, the \kl{discrete derivatives} are linear operations on
+polynomials, that commutes with \kl{translation operators}. However, the
+\kl{translation operators} do not perfectly commute with partial applications
+$\restr{\cdot}{\nu}$.
+
+\begin{remark}
+    \label{discrete-deriv-linear:fact}
+    For all tuples $\vec{X}$ of indeterminates,
+    for all $K \in \Nat$, $L \in \Rel$, for all partial functions
+    $\nu \colon \vec{X} \topartial \Nat$:
+    \begin{enumerate}
+        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
+        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
+        and
+        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
+            are linear operators,
+        \item $\Diff{K}{ \cdot } \circ \translate{L}
+            = \translate{L} \circ \Diff{K}{\cdot}$,
+        \item $\restr{\cdot}{\nu} \circ \translate{L}
+            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
+    \end{enumerate}
+\end{remark}
 

### 2024-04-22 parts/correction.tex
moving remark to appendix

@@
 \end{definition}
-
-As one would expect, the \kl{discrete derivatives} are linear operations on
-polynomials, that commutes with \kl{translation operators}. However, the
-\kl{translation operators} do not perfectly commute with partial applications
-$\restr{\cdot}{\nu}$.
-
-\begin{remark}
-    \label{discrete-deriv-linear:fact}
-    For all tuples $\vec{X}$ of indeterminates,
-    for all $K \in \Nat$, $L \in \Rel$, for all partial functions
-    $\nu \colon \vec{X} \topartial \Nat$:
-    \begin{enumerate}
-        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
-        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
-        and
-        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
-            are linear operators,
-        \item $\Diff{K}{ \cdot } \circ \translate{L}
-            = \translate{L} \circ \Diff{K}{\cdot}$,
-        \item $\restr{\cdot}{\nu} \circ \translate{L}
-            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
-    \end{enumerate}
-\end{remark}
 

### 2024-04-22 parts/appendix.tex
moving fact to appendix

@@
 
+\begin{fact}
+    \label{derivation-monomials:fact}
+    Let $K > 0$,
+    and let $M,T$ be two \kl{monomials}, such that
+    $T$ \kl{strictly divides} $M$.
+    Then,
+    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
+    and 
+    $T$ \kl{divides} some \kl{maximal monomial}
+    of $\Diff{K}{M}$, which has a coefficient
+    that is a multiple of $K$.
+    \proofref{derivation-monomials:fact}
+\end{fact}
 \begin{proofof}[derivation-monomials:fact]

### 2024-04-22 parts/correction.tex
moving fact to appendix

@@
 extracts \emph{sub-maximal} monomials from a given polynomial.
-
-\begin{fact}
-    \label{derivation-monomials:fact}
-    Let $K > 0$,
-    and let $M,T$ be two \kl{monomials}, such that
-    $T$ \kl{strictly divides} $M$.
-    Then,
-    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
-    and 
-    $T$ \kl{divides} some \kl{maximal monomial}
-    of $\Diff{K}{M}$, which has a coefficient
-    that is a multiple of $K$.
-    \proofref{derivation-monomials:fact}
-\end{fact}
 

### 2024-04-22 parts/appendix.tex
factoring + commenting proofs

@@
 
+\begin{lemma}
+    \label{correct-residual:lemma}
+    If \cref{residual:algo} terminates on 
+    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
+    then it computes the \kl{$k$-residual transducer} of $f$.
+    \proofref{correct-residual:lemma}
+\end{lemma}
+
 \begin{proofof}[correct-residual:lemma]
@@
 
+\begin{lemma}
+    \label{wqo-implies-termination:lemma}
+    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
+    every infinite, $\prefleq$-increasing sequence is \kl{good}
+    in $(\Sigma^*, \resleq{f}{k})$
+    (or equivalently, such that the relation $(\prefleq \Rightarrow \resleq{f}{k})$
+    is \kl{well}).
+    Then, \cref{residual:algo} terminates on the input $(f,k)$.
+    \proofref{wqo-implies-termination:lemma}
+\end{lemma}
 \begin{proofof}[wqo-implies-termination:lemma]
@@
 
+\begin{lemma}
+    \label{n-poly-k-implies-wqo:lemma}
+    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
+    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
+    \proofref{n-poly-k-implies-wqo:lemma}
+\end{lemma}
 \begin{proofof}[n-poly-k-implies-wqo:lemma]

### 2024-04-22 parts/counter.tex
factoring + commenting proofs

@@
 
-\subsection{The counter example}
+%\subsection{The counter example}
 \label{sec:c-example}
@@
 Let us now provide a counter example to \cref{karh:thm}. 
-
 \begin{definition}
@@
 \end{lemma}
-\begin{proof}
-    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
-    the expression of $\BadPoly$ to 
-    obtain
-    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
-    The \kl{maximal monomials} of $P$
-    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
-    \kl{non-negative}.
-
-    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
-    There exists a \kl{commutative}
-    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
-    such that for all $w \in \Sigma^*$,
-    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-
-    Remark that $\BadPoly(X,Y,Z) = 0$
-    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-    $\BadPoly(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
-    $Z \neq 0$, and $X = Y = 0$.
-
-    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
-    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
-    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ not a regular language,
-    we
-    conclude that $L$ is not a regular language.
-    However, $L = f^{-1}(\set{0})$ is a regular language
-    (\cref{pre-image-regular:fact}). 
-\end{proof}
+%
+%\begin{proof}
+%    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
+%    the expression of $\BadPoly$ to 
+%    obtain
+%    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
+%    The \kl{maximal monomials} of $P$
+%    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
+%    \kl{non-negative}.
+%
+%    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
+%    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
+%    There exists a \kl{commutative}
+%    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
+%    such that for all $w \in \Sigma^*$,
+%    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
+%
+%    Remark that $\BadPoly(X,Y,Z) = 0$
+%    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
+%    $\BadPoly(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
+%    $Z \neq 0$, and $X = Y = 0$.
+%
+%    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
+%    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
+%    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
+%    Because $L \cap \set{a,b}^*$ not a regular language,
+%    we
+%    conclude that $L$ is not a regular language.
+%    However, $L = f^{-1}(\set{0})$ is a regular language
+%    (\cref{pre-image-regular:fact}). 
+%\end{proof}
 

### 2024-04-22 parts/transducers.tex
mise en forme
factoring + commenting proofs

@@
 
-\begin{lemma}
-    \label{correct-residual:lemma}
-    If \cref{residual:algo} terminates on 
-    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
-    then it computes the \kl{$k$-residual transducer} of $f$.
-\end{lemma}
-
-\begin{lemma}
-    \label{wqo-implies-termination:lemma}
-    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
-    every infinite, $\prefleq$-increasing sequence is \kl{good}
-    in $(\Sigma^*, \resleq{f}{k})$
-    (or equivalently, such that the relation $(\prefleq \Rightarrow \resleq{f}{k})$
-    is \kl{well}).
-    Then, \cref{residual:algo} terminates on the input $(f,k)$.
-\end{lemma}
-
 The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
@@
 increasing in every direction.
-
-\begin{lemma}
-    \label{n-poly-k-implies-wqo:lemma}
-    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
-    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
-\end{lemma}
 
@@
     the \kl{$0$-residual transducer} of $f$ is \kl{counter-free}.
+    \proofref{aperiodic-iff-residual:lem}
 \end{lemma}
@@
     \kl[aperiodic ordering]{aperiodic} \kl{well-quasi-ordering}.
+    \proofref{sf-no-periods-on-sequences:lemma}
 \end{lemma}

### 2024-04-23 parts/appendix.tex
move to appendix a lot of stuff

@@
 
+\section{Proofs of section \ref{polynomials:sec}}
+
+\begin{proofof}[n-poly-n-poly:example]
+    The polynomials of $\Nat[\vec{X}]$
+    are obtained from basic functions (constant in $\Nat$,
+    $X_i$ for some $1 \leq i \leq n$)
+    by products and sums. Because the basic functions are
+    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
+    that are closed under these operations, we conclude.
+\end{proofof}
+
 \begin{proofof}[negative-not-nrat:ex]
@@
     is a \kl{$\Nat$-rational polynomial}.
-    Finally, 
-    $T(X) \defined - X$ cannot be 
-    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
-    are non-negative.
-\end{proofof}
-
-\section{Proofs of section \ref{polynomials:sec}}
-
-\begin{proofof}[negative-but-npoly:ex]
+
     Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
@@
     $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
-\end{proofof}
-
-\begin{proofof}[n-poly-n-poly:example]
-    The polynomials of $\Nat[\vec{X}]$
-    are obtained from basic functions (constant in $\Nat$,
-    $X_i$ for some $1 \leq i \leq n$)
-    by products and sums. Because the basic functions are
-    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
-    that are closed under these operations, we conclude.
+
+    Finally, 
+    $T(X) \defined - X$ cannot be 
+    a \kl{$\Nat$-rational polynomial} as \kl{$\Nat$-polyregular functions}
+    are non-negative.
+\end{proofof}
+
+\begin{proofof}[thm:counter-example]
+    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
+    the expression of $\BadPoly$ to 
+    obtain
+    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
+    The \kl{maximal monomials} of $P$
+    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
+    \kl{non-negative}.
+
+    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
+    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
+    There exists a \kl{commutative}
+    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
+    such that for all $w \in \Sigma^*$,
+    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
+
+    Remark that $\BadPoly(X,Y,Z) = 0$
+    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
+    $\BadPoly(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
+    $Z \neq 0$, and $X = Y = 0$.
+
+    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
+    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
+    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
+    Because $L \cap \set{a,b}^*$ not a regular language,
+    we
+    conclude that $L$ is not a regular language.
+    However, $L = f^{-1}(\set{0})$ is a regular language
+    (\cref{pre-image-regular:fact}). 
 \end{proofof}
@@
     \end{align*}
+\end{proofof}
+
+\begin{proofof}[lem:correct-to-n-rat]
+    We prove the result by induction on $\MaximalMonomials(P)$. 
+    In the proof, we write $\vec{X}$ for the indeterminates appearing in $P$,
+    that is, we assume without loss of generality that all indeterminates are used.
+
+    \textbf{Base case:} If the (unique) \kl{maximal monomial} of $P$ is a
+    constant term. Since $P \in \CorrectPoly$, $P = n \in \Nat$, and therefore
+    $P$ is \kl{represented} by a constant \kl{star-free $\Nat$-polyregular
+    function}.
+
+    \textbf{Induction:} Assume that $P$ is not a constant polynomial, and let
+    us write $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
+    monomials} of $P$. We compute a bound $K$ such that $Q \defined
+    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$ using
+    \cref{derivation-stabilises-correct:lem}. Thanks to
+    \cref{derivation-simplifies:lemma}, we also know that $\MaximalMonomials(Q)
+    \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
+    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} which is
+    effectively computable.
+
+    Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
+    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular
+    function} (\cref{n-poly-n-poly:example}). As a consequence,
+    $\translate{K}(P) = P_1 + Q$ is (effectively) \kl{represented} by a
+    function $f_\Delta$.
+
+    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
+    K}$ fixing at least one indeterminate,  we know that
+    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
+    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
+    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
+    \kl{represents} $\restr{P}{\nu}$.
+
+    Let us assume that the alphabet over which the (\kl{commutative}) functions
+    $f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
+    representing the indeterminate $X_i$ of the polynomials. Now, let us define
+    by case analysis the following \kl{commutative} \kl{star-free
+    $\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
+    a_1^{X_1} \cdots a_n^{X_n}$, with $X_1, \dots, X_n \geq 0$.
+
+    \begin{equation*}
+        f(w) \defined
+        \begin{cases}
+            f_\nu(w) & \text{ if } \exists i \in \set{1, \dots, n}, X_i \leq K \\
+            f_\Delta(a_1^{X_1 - K} \cdots a_n^{X_n - K}) & \text{ otherwise }
+        \end{cases}
+        \quad .
+    \end{equation*}
+    Remark that
+    $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}
+    that
+    \kl{represents} $P$.
 \end{proofof}
@@
 
-\section{Proofs of section \ref{sec:ccl}}
+\begin{proofof}[non-commutative-npoly:thm]
+    \cref{n-poly-1-transd:item} implies \cref{n-poly-k:item} by
+    \cref{transducer-nsf-npoly:lemma}. Then,
+    \cref{n-poly-k:item} implies \cref{n-poly-wqo:item} by
+    \cref{n-poly-k-implies-wqo:lemma}.
+    The implication \cref{n-poly-wqo:item} $\Rightarrow$ \cref{n-poly-well:item}
+    is obvious.
+    Then, \cref{wqo-implies-termination:lemma} proves
+    that \cref{n-poly-well:item} implies \cref{n-poly-residual:item}.
+    Finally, because a \kl{$k$-residual transducer} is a \kl{$\NPoly[k-1]$-transducer},
+    \cref{n-poly-residual:item} implies \cref{n-poly-1-transd:item}.
+\end{proofof}
 
@@
 \end{proofof}
+
+\section{Proofs of section \ref{sec:ccl}}
 

### 2024-04-23 parts/correction.tex
move to appendix a lot of stuff

@@
     \label{discrete-derivative:def}
-    Let $\vec{X}$ be a tuple of indeterminates,
+    Let
     $K \in \Nat$,
     and 
-    $P \in \Rel[\vec{X}]$ be a polynomial.
-    \begin{equation*}
+    $P \in \Rel[\vec{X}]$ be a polynomial,
+    then 
+    %\begin{equation*}
+    $
         \intro*\Diff{K}{P} \defined 
         \translate{K}(P) - P
-        \quad .
-    \end{equation*}
+    $.
+    %    \quad .
+    %\end{equation*}
 \end{definition}
-
-In the actual statement of \cref{corrected-version:thm}, the representation of
-polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
-functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
-surprising, given the conjecture that \kl{ultimately polynomial}
-\kl{$\Nat$-polyregular functions} (the formal definition will be given in
-\cref{star-free:sec}) were conjectured to be star free
-(\cref{zsf-nsf:conjecture}), and that polynomials are always \kl{ultimately
-polynomial}.
 
@@
 
+%In the actual statement of \cref{corrected-version:thm}, the representation of
+%polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
+%functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
+%surprising, given the conjecture that \kl{ultimately polynomial}
+%\kl{$\Nat$-polyregular functions} (the formal definition will be given in
+%\cref{star-free:sec}) were conjectured to be star free
+%(\cref{zsf-nsf:conjecture}), and that polynomials are always \kl{ultimately
+%polynomial}.
+
 \begin{lemma}
@@
 \end{lemma}
-\begin{proof}
-    We prove the result by induction on $\MaximalMonomials(P)$. 
-    In the proof, we write $\vec{X}$ for the indeterminates appearing in $P$,
-    that is, we assume without loss of generality that all indeterminates are used.
-
-    \textbf{Base case:} If the (unique) \kl{maximal monomial} of $P$ is a
-    constant term. Since $P \in \CorrectPoly$, $P = n \in \Nat$, and therefore
-    $P$ is \kl{represented} by a constant \kl{star-free $\Nat$-polyregular
-    function}.
-
-    \textbf{Induction:} Assume that $P$ is not a constant polynomial, and let
-    us write $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
-    monomials} of $P$. We compute a bound $K$ such that $Q \defined
-    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$ using
-    \cref{derivation-stabilises-correct:lem}. Thanks to
-    \cref{derivation-simplifies:lemma}, we also know that $\MaximalMonomials(Q)
-    \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} which is
-    effectively computable.
-
-    Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
-    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular
-    function} (\cref{n-poly-n-poly:example}). As a consequence,
-    $\translate{K}(P) = P_1 + Q$ is (effectively) \kl{represented} by a
-    function $f_\Delta$.
-
-    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
-    K}$ fixing at least one indeterminate,  we know that
-    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
-    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
-    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
-    \kl{represents} $\restr{P}{\nu}$.
-
-    Let us assume that the alphabet over which the (\kl{commutative}) functions
-    $f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
-    representing the indeterminate $X_i$ of the polynomials. Now, let us define
-    by case analysis the following \kl{commutative} \kl{star-free
-    $\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
-    a_1^{X_1} \cdots a_n^{X_n}$, with $X_1, \dots, X_n \geq 0$.
-
-    \begin{equation*}
-        f(w) \defined
-        \begin{cases}
-            f_\nu(w) & \text{ if } \exists i \in \set{1, \dots, n}, X_i \leq K \\
-            f_\Delta(a_1^{X_1 - K} \cdots a_n^{X_n - K}) & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    It is clear that
-    $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function},
-    and that
-    $f$ \kl{represents} $P$.
-\end{proof}
 
@@
 The counter example of \cref{thm:counter-example} used three indeterminates,
-and we will prove in \cref{sec:proof} that \cref{karh:thm} holds on polynomial
-with at most two indeterminate. In particular, the examples appearing in
-\cite{KARH77} are not invalidated, as they all use at most two indeterminates.
-In particular, we derive the equality of $\CorrectPoly$ and $\CoveredPoly$ in
-the case of two indeterminates.  The proof is based the fact that for
-(non-negative) univariate it is enough that the dominant coefficient is
-non-negative to be an \kl{$\Nat$-rational polynomial}.
+and this is not a coincidence: in the particular case of at most two
+indeterminates, the classes $\CorrectPoly$ and $\CoveredPoly$ coincide. In
+particular, the examples appearing in \cite{KARH77} are not invalidated, as
+they all use at most two indeterminates. This is because, in the
+univariate case, being non-negative and having non-negative maximal coefficient
+implies being an \kl{$\Nat$-rational polynomial}.
 

### 2024-04-23 parts/counter.tex
move to appendix a lot of stuff

@@
 \label{polynomials:sec}
+%\subsection{The counter example}
+\label{sec:c-example}
 
-The goal of this section is to provide semantic characterizations of
-\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is an
-\kl{$\Nat$-rational polynomial}, and if so, explicitly compute an
-\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$. It is an easy
-check that polynomials with coefficients in $\Nat$ are \kl{$\Nat$-rational
-polynomials} (\cref{n-poly-n-poly:example}). However,
-\cref{negative-but-npoly:ex} provides a polynomial with negative coefficients
-that is an \kl{$\Nat$-rational polynomial}. As a consequence, the class of
-interest strictly contains $\Nat[\vec{X}]$.
+\AP A polynomial $P \in \Rel[X_1, \dots, X_n]$ is an \intro{$\Nat$-rational
+polynomial} if it is \kl{represented} by a \kl{$\Nat$-polyregular function}.
+The analogue notion of \intro{$\Rel$-rational polynomials} is not of particular
+interest, since every polynomial $P$ is \kl{represented} by a
+\kl{$\Rel$-polyregular function}. It is an easy check that polynomials with
+coefficients in $\Nat$ are \kl{$\Nat$-rational polynomials}
+(\cref{n-poly-n-poly:example}). However, \cref{negative-but-npoly:ex} provides
+a polynomial with negative coefficients that is an \kl{$\Nat$-rational
+polynomial}. As a consequence, the class of interest strictly contains
+$\Nat[\vec{X}]$.
 
@@
 \begin{example}
+    \label{negative-not-nrat:ex}
     \label{negative-but-npoly:ex}
-    Let $P \defined X^2 -2X + 2$. Then $P$
-    is an \kl{$\Nat$-rational polynomial}.
-    \proofref{negative-but-npoly:ex}
+    The polynomials $X$, $X^2 + 3$,
+    and $X^2 - 2X + 2$
+    are \kl{$\Nat$-rational polynomials},
+    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
+    not an \kl{$\Nat$-rational polynomial}.
+    \proofref{negative-not-nrat:ex}
 \end{example}
 
-Let us now introduce the characterization of \kl{$\Nat$-rational polynomials}
-given by \cite{KARH77}, which is stated using a class
-$\CoveredPoly[\vec{X}]$ of polynomials with multiple indeterminate.
+The goal of this section is to provide semantic characterizations of
+\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is an
+\kl{$\Nat$-rational polynomial}, and if so, explicitly compute an
+\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$. To that end, let
+us now introduce the characterization of \kl{$\Nat$-rational polynomials} given
+by \cite{KARH77}, which is stated using a class $\CoveredPoly[\vec{X}]$ of
+polynomials with multiple indeterminates.
 
 \begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
-    Let $\vec{X}$ be a finite tuple of indeterminate.
+    Let $\vec{X}$ be a finite tuple of indeterminates.
     The class $\intro*\CoveredPoly[\vec{X}]$
@@
     and such that every \kl{maximal monomial} is \kl{non-negative}.
-    When the indeterminate are clear from the context, we write
+    When the indeterminates are clear from the context, we write
     this class $\reintro*\CoveredPoly$.
@@
 
-%\subsection{The counter example}
-\label{sec:c-example}
+To understand why \cref{karh:thm} is false, we need to understand why some
+non-negative polynomials are not \kl{$\Nat$-rational polynomials}, which is
+illustrated by \cref{non-neg-not-nrat:ex}. Note that the proof scheme of this
+example is almost a counter-example to \cite[Theoerm 3.3]{KARH77} in
+\cref{thm:counter-example}, and leverages a key property of $\NRat$ recalled
+hereafter.
 
-Let us now provide a counter example to \cref{karh:thm}. 
+\begin{fact}
+    \label{pre-image-regular:fact}
+    The pre-image of a regular language by an \kl{$\Nat$-rational series}
+    is a regular language. 
+\end{fact}
+
+\begin{example}
+    \label{non-neg-not-nrat:ex}
+    Let $P(X, Y) \defined (X - Y)^2$.
+    Then $P$ is
+    is \kl{non-negative}, but is
+    not an \kl{$\Nat$-rational polynomial}.
+\end{example}
+\begin{proof}
+    Assume by contradiction that
+    $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
+    Then, $f^{-1}(\set{0})$ is a regular language
+    (\cref{pre-image-regular:fact}),
+    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
+    is not.
+\end{proof}
+
+Let us now design a counter example to \cref{karh:thm} by suitably
+tweaking \cref{non-neg-not-nrat:ex} to ensure that the polynomial not only is
+\kl{non-negative}, but also belongs to
+$\CoveredPoly$.
+
 \begin{definition}
@@
     but is not an \kl{$\Nat$-rational polynomial}.
-\end{lemma}
-%
-%\begin{proof}
-%    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
-%    the expression of $\BadPoly$ to 
-%    obtain
-%    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
-%    The \kl{maximal monomials} of $P$
-%    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
-%    \kl{non-negative}.
-%
-%    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
-%    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
-%    There exists a \kl{commutative}
-%    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
-%    such that for all $w \in \Sigma^*$,
-%    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-%
-%    Remark that $\BadPoly(X,Y,Z) = 0$
-%    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-%    $\BadPoly(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
-%    $Z \neq 0$, and $X = Y = 0$.
-%
-%    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
-%    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
-%    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-%    Because $L \cap \set{a,b}^*$ not a regular language,
-%    we
-%    conclude that $L$ is not a regular language.
-%    However, $L = f^{-1}(\set{0})$ is a regular language
-%    (\cref{pre-image-regular:fact}). 
-%\end{proof}
-
-\begin{corollary}
-    The result stated in \cite[Theorem 3.3]{KARH77}, restated
+    As a corollary, 
+    the result stated in \cite[Theorem 3.3]{KARH77}, restated
     in \cref{karh:thm}, is false
     when allowing at least $3$ indeterminates.
-\end{corollary}
+\end{lemma}

### 2024-04-23 parts/prelim.tex
move to appendix a lot of stuff

@@
 \subsection{Polynomials} \AP All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[X_1, \dots, X_n]$ is an \intro{$\Nat$-rational polynomial} if it is
-\kl{represented} by a \kl{$\Nat$-polyregular function}. The analogue notion of
-\intro{$\Rel$-rational polynomials} is not of particular interest, since every
-polynomial $P$ is \kl{represented} by a \kl{$\Rel$-polyregular function}.
-
-\begin{example}
-    \label{negative-not-nrat:ex}
-    The polynomials $X$, and $X^2 + 3$ are \kl{$\Nat$-rational polynomials},
-    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
-    not an \kl{$\Nat$-rational polynomial}.
-    \proofref{negative-not-nrat:ex}
-\end{example}
-
-\AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for
+coefficients in $\Rel$ unless explicitly stated otherwise. 
+A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for
 all non-negative integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1,
@@
 
-All \kl{$\Nat$-rational polynomials} are \kl{non-negative}, but the converse
-does not hold, as the following example illustrates. Note that the proof scheme
-of this example will be at the core of the counter-example to \cite[Theoerm
-3.3]{KARH77} in \cref{thm:counter-example}, and leverages a key property of
-$\NRat$ recalled hereafter.
-
-\begin{fact}
-    \label{pre-image-regular:fact}
-    The pre-image of a regular language by an \kl{$\Nat$-rational series}
-    is a regular language. 
-\end{fact}
-
-\begin{example}
-    Let $P(X, Y) \defined (X - Y)^2$.
-    Then $P$ is
-    is \kl{non-negative}, but is
-    not an \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Assume by contradiction that
-    $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
-    Then, $f^{-1}(\set{0})$ is a regular language
-    (\cref{pre-image-regular:fact}),
-    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
-    is not.
-\end{proof}
-
-\AP Let us now give some vocabulary on polynomials with multiple indeterminate
+\AP Let us now give some vocabulary on polynomials with multiple indeterminates
 over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.

### 2024-04-23 parts/transducers.tex
move to appendix a lot of stuff

@@
     is defined inductively as follows:
-    \begin{itemize}
-        \item $\aTransd(q, \varepsilon) \defined F(q)$;
-        \item $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
+        $\aTransd(q, \varepsilon) \defined F(q)$, and
+        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
             + \lambda(q,a)(u)$.
-    \end{itemize}
     The function \intro{computed} by a transducer $\aTransd$
@@
 \end{theorem}
-\begin{proof}
-    \cref{n-poly-1-transd:item} implies \cref{n-poly-k:item} by
-    \cref{transducer-nsf-npoly:lemma}. Then,
-    \cref{n-poly-k:item} implies \cref{n-poly-wqo:item} by
-    \cref{n-poly-k-implies-wqo:lemma}.
-    The implication \cref{n-poly-wqo:item} $\Rightarrow$ \cref{n-poly-well:item}
-    is obvious.
-    Then, \cref{wqo-implies-termination:lemma} proves
-    that \cref{n-poly-well:item} implies \cref{n-poly-residual:item}.
-    Finally, because a \kl{$k$-residual transducer} is a \kl{$\NPoly[k-1]$-transducer},
-    \cref{n-poly-residual:item} implies \cref{n-poly-1-transd:item}.
-\end{proof}
 

### 2024-04-23 parts/prelim.tex
move to appendix

@@
 
+\AP
+As a model of computation, we will use the following definition of transducers,
+that is based on the marble transducers \cite{EHB99}, and follows the general
+pattern of transducers introduced in \cite{CDTL23}. Note that we only consider
+$\Nat$-valued functions, but could consider other commutative semigroups.
+
+\begin{definition}
+    Let $\mathcal{H}$ be a family of functions
+    from $\Sigma^*$ to $\Nat$.
+    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
+    a tuple $(Q, q_0, \delta, \lambda, F)$ where
+    \begin{itemize}
+        \item $Q$ is a finite set of states,
+        \item $q_0 \in Q$ is an initial state,
+        \item $\delta \colon Q \times \Sigma \to Q$
+            is a deterministic transition function,
+        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
+            is a correction function,
+        \item $F \colon Q \to \Nat$ is an output function.
+    \end{itemize}
+\end{definition}
+
+\begin{definition}
+    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
+    The function
+    $\aTransd \colon Q \times \Sigma^* \to \Nat$
+    is defined inductively as follows:
+        $\aTransd(q, \varepsilon) \defined F(q)$, and
+        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
+            + \lambda(q,a)(u)$.
+    The function \intro{computed} by a transducer $\aTransd$
+    is $w \mapsto \aTransd(q_0, w)$.
+\end{definition}
+
 \subsection{Polynomials} \AP All polynomials considered in this paper have

### 2024-04-23 parts/transducers.tex
move to appendix

@@
 \cite{MELL98}.
-
-As a model of computation, we will use the following definition of transducers,
-that is based on the marble transducers \cite{EHB99}, and follows the general
-pattern of transducers introduced in \cite{CDTL23}. Note that we only consider
-$\Nat$-valued functions, but could consider other commutative semigroups.
-
-\begin{definition}
-    Let $\mathcal{H}$ be a family of functions
-    from $\Sigma^*$ to $\Nat$.
-    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
-    a tuple $(Q, q_0, \delta, \lambda, F)$ where
-    \begin{itemize}
-        \item $Q$ is a finite set of states,
-        \item $q_0 \in Q$ is an initial state,
-        \item $\delta \colon Q \times \Sigma \to Q$
-            is a deterministic transition function,
-        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
-            is a correction function,
-        \item $F \colon Q \to \Nat$ is an output function.
-    \end{itemize}
-\end{definition}
-
-\begin{definition}
-    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
-    The function
-    $\aTransd \colon Q \times \Sigma^* \to \Nat$
-    is defined inductively as follows:
-        $\aTransd(q, \varepsilon) \defined F(q)$, and
-        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
-            + \lambda(q,a)(u)$.
-    The function \intro{computed} by a transducer $\aTransd$
-    is $w \mapsto \aTransd(q_0, w)$.
-\end{definition}
 
@@
 \reintro*\resleq{f}{k})$.
-
-\begin{remark}
-    \label{good-residual-ordering:fact}
-    Let $k \in \Nat$, and let $f \colon \Sigma^* \to \Nat$. Then,
-    $(\resleq{f}{k})$ is a quasi-ordering, satisfying the following
-    extra properties:
-    \begin{enumerate}
-        \item For all $u,v,w \in \Sigma^*$, $u \resleq{f}{k} v$
-            implies $uw \resleq{f}{k} vw$,
-        \item If $u \resleq{f}{k} v$ and $v \resleq{f}{k} u$,
-            then $\app{f}{u} = \app{f}{v}$.
-    \end{enumerate}
-\end{remark}
 

### 2024-04-23 parts/prelim.tex
move definitions to preliminaries

@@
 these two output semigroups, and we refer the reader to \cite{BOKL19} for the
-general definition of \kl{polyregular functions}. The following
+general definition of \kl{polyregular functions}. As a model of computation, we
+will use the following definition of transducers, that is based on the marble
+transducers \cite{EHB99}, and follows the general pattern of transducers
+introduced in \cite{CDTL23}. Note that we only consider $\Rel$-valued (resp.
+$\Nat$-valued) functions, but could consider other commutative semigroups.
+
+\begin{definition}
+    Let $\mathcal{H}$ be a family of functions
+    from $\Sigma^*$ to $\Nat$.
+    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
+    a tuple $(Q, q_0, \delta, \lambda, F)$ where
+    \begin{itemize}
+        \item $Q$ is a finite set of states,
+        \item $q_0 \in Q$ is an initial state,
+        \item $\delta \colon Q \times \Sigma \to Q$
+            is a deterministic transition function,
+        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
+            is a correction function,
+        \item $F \colon Q \to \Nat$ is an output function.
+    \end{itemize}
+\end{definition}
+
+\begin{definition}
+    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
+    The function
+    $\aTransd \colon Q \times \Sigma^* \to \Nat$
+    is defined inductively as follows:
+        $\aTransd(q, \varepsilon) \defined F(q)$, and
+        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
+            + \lambda(q,a)(u)$.
+    The function \intro{computed} by a transducer $\aTransd$
+    is $w \mapsto \aTransd(q_0, w)$.
+\end{definition}
+
+\AP It is already known that this notion of transducers exactly captures $\NSF$
+and $\NPoly$, and generalizes to outputs into other commutative semigroups than
+$\Nat$. To make the above statement formal, let us recall that a
+\intro{counter} in an automaton is a pair $u,w$ of words together with an
+integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0, u)$ and
+$\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$; where we used the
+extension of $\delta \colon Q \times \Sigma \to Q$ to words by defining
+$\delta(q, \varepsilon) = q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. An
+automaton is \reintro{counter-free} if it contains no \reintro{counters}. This
+notion can be lifted to \kl{$\mathcal{H}$-transducers} by considering counters
+of their underlying automata.
+
+\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
+    \label{transducer-nsf-npoly:lemma}
+    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
+    Then, 
+    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
+    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
+        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
+        \kl{counter-free}).
+\end{lemma}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{0.49\linewidth}
+        \begin{tikzpicture}[
+            etat/.style={minimum size=2em}
+            ]
+            \node[etat,state,initial,
+                accepting by arrow,
+                accepting text={$1$},
+                accepting where=below,
+                ] (A) at (0,0) {$\varepsilon$};
+            \node[etat,state,
+                accepting by arrow,
+                accepting text={$0$},
+                accepting where=below,
+                ] (B) at (2,0) {$a$};
+
+            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
+
+            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
+        \end{tikzpicture}
+        \caption{This automaton has a \kl{counter}.}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.49\linewidth}
+        \begin{tikzpicture}[
+            etat/.style={minimum size=2em}
+            ]
+            \node[etat,state,initial,
+                accepting by arrow,
+                accepting text={$1$},
+                accepting where=below,
+                ] (A) at (0,0) {$\varepsilon$};
+            \node[etat,state,
+                accepting by arrow,
+                accepting text={$0$},
+                accepting where=below,
+                ] (B) at (2,0) {$a$};
+
+            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
+
+            \draw[->] (B) edge[loop right] node[midway, right]
+                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
+        \end{tikzpicture}
+        \caption{This automaton is \kl{counter-free}.}
+    \end{subfigure}
+    \caption{Two \kl{$\NPoly[0]$-transducers}
+    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
+    \label{non-canonical-transd:fig}
+\end{figure}
+
+\begin{example}
+    \label{non-canonical-transd:ex}
+    Let $\BadExOk \colon \set{a}^* \to \Nat$
+    be defined as
+    $\BadExOk(\varepsilon) = 1$,
+    $\BadExOk(a) = 0$,
+    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
+    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
+    depicted in \cref{non-canonical-transd:fig}.
+\end{example}
+
+\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
+defined as the minimal $k$ such that $\card{f(w)} =
+\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
+$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp.
+$\reintro*\NPoly[k]$) for the functions in $\ZRat$ (resp. $\NRat$) that have
+growth rate at most $k$.
+
+It is clear that a function $f \in \ZPoly$ defined using a production function
+$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
+generally it is known that \kl{polyregular functions} have \kl{polynomial
+growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
+growth}, a typical example being the function $a^n \mapsto 2^n$.
+
+\begin{example}
+    \label{size-of-word-nsf:ex}
+    The map $f \colon w \mapsto \card{w} + 1$
+    belongs to $\NSF$.
+    \proofref{size-of-word-nsf:ex}
+\end{example}
+
+\AP
+The following
 \cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
@@
 \end{definition}
-
-\begin{example}
-    \label{size-of-word-nsf:ex}
-    The map $f \colon w \mapsto \card{w} + 1$
-    belongs to $\NSF$.
-    \proofref{size-of-word-nsf:ex}
-\end{example}
-
-\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
-defined as the minimal $k$ such that $\card{f(w)} =
-\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
-$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp. $\reintro*\NPoly[k]$) for
-the functions in $\ZPoly$ (resp. $\NPoly$) that have growth rate at most $k$.
-It is clear that a function $f \in \ZPoly$ defined using a production function
-$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
-generally it is known that \kl{polyregular functions} have \kl{polynomial
-growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
-growth}, a typical example being the function $a^n \mapsto 2^n$.
 
@@
 
-\AP
-As a model of computation, we will use the following definition of transducers,
-that is based on the marble transducers \cite{EHB99}, and follows the general
-pattern of transducers introduced in \cite{CDTL23}. Note that we only consider
-$\Nat$-valued functions, but could consider other commutative semigroups.
-
-\begin{definition}
-    Let $\mathcal{H}$ be a family of functions
-    from $\Sigma^*$ to $\Nat$.
-    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
-    a tuple $(Q, q_0, \delta, \lambda, F)$ where
-    \begin{itemize}
-        \item $Q$ is a finite set of states,
-        \item $q_0 \in Q$ is an initial state,
-        \item $\delta \colon Q \times \Sigma \to Q$
-            is a deterministic transition function,
-        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
-            is a correction function,
-        \item $F \colon Q \to \Nat$ is an output function.
-    \end{itemize}
-\end{definition}
-
-\begin{definition}
-    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
-    The function
-    $\aTransd \colon Q \times \Sigma^* \to \Nat$
-    is defined inductively as follows:
-        $\aTransd(q, \varepsilon) \defined F(q)$, and
-        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
-            + \lambda(q,a)(u)$.
-    The function \intro{computed} by a transducer $\aTransd$
-    is $w \mapsto \aTransd(q_0, w)$.
-\end{definition}
-
 \subsection{Polynomials} \AP All polynomials considered in this paper have

### 2024-04-23 parts/transducers.tex
move definitions to preliminaries

@@
 \cite{MELL98}.
-
-\begin{example}
-    \label{non-canonical-transd:ex}
-    Let $\BadExOk \colon \set{a}^* \to \Nat$
-    be defined as
-    $\BadExOk(\varepsilon) = 1$,
-    $\BadExOk(a) = 0$,
-    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
-    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
-    depicted in \cref{non-canonical-transd:fig}.
-\end{example}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.49\linewidth}
-        \begin{tikzpicture}[
-            etat/.style={minimum size=2em}
-            ]
-            \node[etat,state,initial,
-                accepting by arrow,
-                accepting text={$1$},
-                accepting where=below,
-                ] (A) at (0,0) {$\varepsilon$};
-            \node[etat,state,
-                accepting by arrow,
-                accepting text={$0$},
-                accepting where=below,
-                ] (B) at (2,0) {$a$};
-
-            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
-
-            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
-        \end{tikzpicture}
-        \caption{This automaton has a \kl{counter}.}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.49\linewidth}
-        \begin{tikzpicture}[
-            etat/.style={minimum size=2em}
-            ]
-            \node[etat,state,initial,
-                accepting by arrow,
-                accepting text={$1$},
-                accepting where=below,
-                ] (A) at (0,0) {$\varepsilon$};
-            \node[etat,state,
-                accepting by arrow,
-                accepting text={$0$},
-                accepting where=below,
-                ] (B) at (2,0) {$a$};
-
-            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
-
-            \draw[->] (B) edge[loop right] node[midway, right]
-                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
-        \end{tikzpicture}
-        \caption{This automaton is \kl{counter-free}.}
-    \end{subfigure}
-    \caption{Two \kl{$\NPoly[0]$-transducers}
-    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
-    \label{non-canonical-transd:fig}
-\end{figure}
-
-\AP It is already known that this notion of transducers exactly captures $\NSF$
-and $\NPoly$, and generalizes to outputs into other commutative semigroups than
-$\Nat$. To make the above statement formal, let us recall that a
-\intro{counter} in an automaton is a pair $u,w$ of words together with an
-integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0, u)$ and
-$\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$; where we used the
-extension of $\delta \colon Q \times \Sigma \to Q$ to words by defining
-$\delta(q, \varepsilon) = q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. An
-automaton is \reintro{counter-free} if it contains no \reintro{counters}. This
-notion can be lifted to \kl{$\mathcal{H}$-transducers} by considering counters
-of their underlying automata.
-
-\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
-    \label{transducer-nsf-npoly:lemma}
-    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
-    Then, 
-    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
-    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
-        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
-        \kl{counter-free}).
-\end{lemma}
 

### 2024-04-23 parts/appendix.tex
move to appendix
move to appendix and clean up

@@
 \section{Proofs of section \ref{preliminaries:sec}}
+
+\AP
+The following
+\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
+and is similar in shape to the \emph{finite counting automata} introduced by
+\cite{SCHU62}.
+
+\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
+    \label{nat-rel-poly:def}
+    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
+    be a morphism, $k \in \Nat$, and 
+    $\pi \colon M^k \to \Rel$ be a function.
+    The \intro{$\Rel$-polyregular function}
+    $\pi^\dagger \colon \Sigma^* \to \Rel$
+    is computed as follows:
+    \begin{equation*}
+        \pi^\dagger (w) \defined
+        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
+        \quad .
+    \end{equation*}
+    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
+    is called \intro{$\Nat$-polyregular}.
+    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
+    When the monoid $M$ is \kl(monoid){aperiodic}
+    then
+    the function $f$ is \intro{star-free $\Rel$-polyregular}
+    (resp. \intro{star-free $\Nat$-polyregular}).
+\end{definition}
+
+\textbf{TODO: remove this example}
+\begin{example}
+    \label{size-of-word-nsf:ex}
+    The map $f \colon w \mapsto \card{w} + 1$
+    belongs to $\NSF$.
+    \proofref{size-of-word-nsf:ex}
+\end{example}
+
+\textbf{TODO: remove this lemma}
+\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
+    \label{transducer-nsf-npoly:lemma}
+    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
+    Then, 
+    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
+    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
+        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
+        \kl{counter-free}).
+\end{lemma}
+
+\AP Let us briefly recall in the following lemmas alternative characterizations
+of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
+analysis of their \kl{commutative} counterpart, and others that provide a nicer
+syntax to construct examples. These characterizations were previously known,
+and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
+\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
+indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
+first-order free variables $\varphi(\vec{x})$, we define
+$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
+a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
+$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
+to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
+$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
+$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
+multiplication with elements $s \in \mathbb{S}$.
+
+\begin{lemma}
+    \label{polynomial-rational-polyreg:fact}
+    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NPoly$ (resp. $\ZPoly$);
+        \item $f$ is a \kl{polyregular function} with output
+            in $\set{1}^*$,
+            post-composed with $\polysum$
+            (resp. with output in $\set{-1,+1}^*$);
+        \item $f$ has \kl{polynomial growth};
+        \item \label{npoly-counting-mso:item} $f$ belongs to
+            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
+            with $\mathbb{S} = \Nat$
+            (resp. $\mathbb{S} = \Rel$).
+    \end{enumerate}
+\end{lemma}
+
+\begin{lemma}
+    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NSF$ (resp. $\ZSF$);
+        \item $f$ is a \kl{star-free polyregular function} with output
+            in $\set{1}^*$,
+            post-composed with $\polysum$
+            (resp. with output in $\set{-1,+1}^*$);
+        \item $f$ belongs to
+            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
+            with $\mathbb{S} = \Nat$
+            (resp. $\mathbb{S} = \Rel$).
+    \end{enumerate}
+\end{lemma}
+
+%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
+%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
+%\begin{example}
+%    \label{regular-language:ex}
+%    Let $\Sigma$ be a finite alphabet, and
+%    $L \subseteq \Sigma^*$. Then,
+%    $L$ is a regular language if and only if
+%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
+%    \kl{$\Nat$-polyregular function}.
+%    Furthermore, $L$ is a star-free regular language
+%    if and only if $\ind{L}$ is a
+%    \kl{star-free $\Nat$-polyregular function}.
+%\end{example}
+%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
+%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
+%of first order free variables used to construct the function as in
+%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.
 
@@
 
+\begin{proofof}[corrected-version:thm]
+    The implications 
+    \cref{corrected-4:item} $\implies$
+    \cref{corrected-3:item} $\implies$
+    \cref{corrected-2:item} are obvious.
+    \cref{lem:correct-to-n-rat} proves
+    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
+    while \cref{n-rat-correct:lem}
+    proves 
+    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
+    Note that the lemmas provide effective conversion procedures,
+    and that \cref{lem:correct-to-n-rat} also provides a decision
+    procedure.
+\end{proofof}
+
 \begin{proofof}[lem:correct-covered-2]
@@
 \section{Proofs of section \ref{beyond-commutative:sec}}
+
+\begin{remark}
+    \label{good-residual-ordering:fact}
+    Let $k \in \Nat$, and let $f \colon \Sigma^* \to \Nat$. Then,
+    $(\resleq{f}{k})$ is a quasi-ordering, satisfying the following
+    extra properties:
+    \begin{enumerate}
+        \item For all $u,v,w \in \Sigma^*$, $u \resleq{f}{k} v$
+            implies $uw \resleq{f}{k} vw$,
+        \item If $u \resleq{f}{k} v$ and $v \resleq{f}{k} u$,
+            then $\app{f}{u} = \app{f}{v}$.
+    \end{enumerate}
+\end{remark}
 

### 2024-04-23 parts/correction.tex
move to appendix and clean up

@@
 
-\AP When multiple indeterminates are involved, the notion of degree must be
-adapted, and to ensure that the decision procedure is effective, one must
-obtain explicit bounds on the \kl{translations} that are needed. To that end,
-let us now introduce the ordering over which the induction of
+\AP Let us now introduce the ordering over which the induction of
 \cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
@@
     \end{enumerate}
-    Furthermore, the membership is decidable, and effective conversion
-    procedures exist between all the representations.
+    Furthermore, the properties are decidable, and conversions effective.
 \end{theorem}
-\begin{proof}
-    The implications 
-    \cref{corrected-4:item} $\implies$
-    \cref{corrected-3:item} $\implies$
-    \cref{corrected-2:item} are obvious.
-    \cref{lem:correct-to-n-rat} proves
-    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
-    while \cref{n-rat-correct:lem}
-    proves 
-    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
-    Note that the lemmas provide effective conversion procedures,
-    and that \cref{lem:correct-to-n-rat} also provides a decision
-    procedure.
-\end{proof}
 

### 2024-04-23 parts/counter.tex
move to appendix and clean up

@@
 non-negative polynomials are not \kl{$\Nat$-rational polynomials}, which is
-illustrated by \cref{non-neg-not-nrat:ex}. Note that the proof scheme of this
-example is almost a counter-example to \cite[Theoerm 3.3]{KARH77} in
-\cref{thm:counter-example}, and leverages a key property of $\NRat$ recalled
-hereafter.
+illustrated by \cref{non-neg-not-nrat:ex}.
 
@@
 
-Let us now design a counter example to \cref{karh:thm} by suitably
-tweaking \cref{non-neg-not-nrat:ex} to ensure that the polynomial not only is
-\kl{non-negative}, but also belongs to
-$\CoveredPoly$.
+Let us now design a counter example to \cref{karh:thm} by suitably tweaking
+\cref{non-neg-not-nrat:ex} to ensure that the polynomial not only is
+\kl{non-negative}, but also belongs to $\CoveredPoly$.
 

### 2024-04-23 parts/prelim.tex
move to appendix and clean up

@@
     Let $\mathcal{H}$ be a family of functions
-    from $\Sigma^*$ to $\Nat$.
+    from $\Sigma^*$ to $\Rel$.
     An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
     a tuple $(Q, q_0, \delta, \lambda, F)$ where
-    \begin{itemize}
-        \item $Q$ is a finite set of states,
-        \item $q_0 \in Q$ is an initial state,
-        \item $\delta \colon Q \times \Sigma \to Q$
+         $Q$ is a finite set of states,
+         $q_0 \in Q$ is an initial state,
+         $\delta \colon Q \times \Sigma \to Q$
             is a deterministic transition function,
-        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
+         $\lambda \colon Q \times \Sigma \to \mathcal{H}$
             is a correction function,
-        \item $F \colon Q \to \Nat$ is an output function.
-    \end{itemize}
+         and $F \colon Q \to \Rel$ is an output function.
 \end{definition}
@@
     The function
-    $\aTransd \colon Q \times \Sigma^* \to \Nat$
+    $\aTransd \colon Q \times \Sigma^* \to \Rel$
     is defined inductively as follows:
@@
 
-\AP It is already known that this notion of transducers exactly captures $\NSF$
-and $\NPoly$, and generalizes to outputs into other commutative semigroups than
-$\Nat$. To make the above statement formal, let us recall that a
-\intro{counter} in an automaton is a pair $u,w$ of words together with an
-integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0, u)$ and
-$\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$; where we used the
-extension of $\delta \colon Q \times \Sigma \to Q$ to words by defining
-$\delta(q, \varepsilon) = q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. An
-automaton is \reintro{counter-free} if it contains no \reintro{counters}. This
-notion can be lifted to \kl{$\mathcal{H}$-transducers} by considering counters
-of their underlying automata.
+\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
+words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
+\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$;
+where we used the extension of $\delta \colon Q \times \Sigma \to Q$ to words
+by defining $\delta(q, \varepsilon) = q$, and $\delta(q,au) =
+\delta(\delta(q,a), u)$. An automaton is \reintro{counter-free} if it contains
+no \reintro{counters}. This notion can be lifted to
+\kl{$\mathcal{H}$-transducers} by considering counters of their underlying
+automata.
 
-\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
-    \label{transducer-nsf-npoly:lemma}
-    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
-    Then, 
-    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
-    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
-        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
-        \kl{counter-free}).
-\end{lemma}
+\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
+define the classes $\intro*\NPoly[k]$, $\intro*\NSF[k]$, $\intro*\ZPoly[k]$ and
+$\intro*\ZSF[k]$ as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
+case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
+\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
+$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
+\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
+$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
+\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
+function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
+where the transducer is furthermore required to be \kl{counter-free}. Then,
+$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
+$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined similarly.
+In the rest of the paper, we will always assume that
+\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
+have an $\Nat$-valued output functions, making a clean separation between
+$\NPoly$ and $\ZPoly$.
+
+\begin{example}
+    \label{non-canonical-transd:ex}
+    Let $\BadExOk \colon \set{a}^* \to \Nat$
+    be defined as
+    $\BadExOk(\varepsilon) = 1$,
+    $\BadExOk(a) = 0$,
+    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
+    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
+    depicted in \cref{non-canonical-transd:fig}.
+\end{example}
 
@@
 
-\begin{example}
-    \label{non-canonical-transd:ex}
-    Let $\BadExOk \colon \set{a}^* \to \Nat$
-    be defined as
-    $\BadExOk(\varepsilon) = 1$,
-    $\BadExOk(a) = 0$,
-    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
-    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
-    depicted in \cref{non-canonical-transd:fig}.
-\end{example}
-
-\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
-defined as the minimal $k$ such that $\card{f(w)} =
-\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
-$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp.
-$\reintro*\NPoly[k]$) for the functions in $\ZRat$ (resp. $\NRat$) that have
-growth rate at most $k$.
-
-It is clear that a function $f \in \ZPoly$ defined using a production function
-$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
-generally it is known that \kl{polyregular functions} have \kl{polynomial
-growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
-growth}, a typical example being the function $a^n \mapsto 2^n$.
-
-\begin{example}
-    \label{size-of-word-nsf:ex}
-    The map $f \colon w \mapsto \card{w} + 1$
-    belongs to $\NSF$.
-    \proofref{size-of-word-nsf:ex}
-\end{example}
-
-\AP
-The following
-\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
-and is similar in shape to the \emph{finite counting automata} introduced by
-\cite{SCHU62}.
-
-\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
-    \label{nat-rel-poly:def}
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        \pi^\dagger (w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
-    When the monoid $M$ is \kl(monoid){aperiodic}
-    then
-    the function $f$ is \intro{star-free $\Rel$-polyregular}
-    (resp. \intro{star-free $\Nat$-polyregular}).
-\end{definition}
-
-\AP Let us briefly recall in the following lemmas alternative characterizations
-of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
-analysis of their \kl{commutative} counterpart, and others that provide a nicer
-syntax to construct examples. These characterizations were previously known,
-and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
-\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
-indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
-first-order free variables $\varphi(\vec{x})$, we define
-$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
-a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
-$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
-to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
-$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
-$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
-multiplication with elements $s \in \mathbb{S}$.
-
-\begin{lemma}
-    \label{polynomial-rational-polyreg:fact}
-    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NPoly$ (resp. $\ZPoly$);
-        \item $f$ is a \kl{polyregular function} with output
-            in $\set{1}^*$,
-            post-composed with $\polysum$
-            (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ has \kl{polynomial growth};
-        \item \label{npoly-counting-mso:item} $f$ belongs to
-            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
-            with $\mathbb{S} = \Nat$
-            (resp. $\mathbb{S} = \Rel$).
-    \end{enumerate}
-\end{lemma}
-
-\begin{lemma}
-    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NSF$ (resp. $\ZSF$);
-        \item $f$ is a \kl{star-free polyregular function} with output
-            in $\set{1}^*$,
-            post-composed with $\polysum$
-            (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ belongs to
-            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
-            with $\mathbb{S} = \Nat$
-            (resp. $\mathbb{S} = \Rel$).
-    \end{enumerate}
-\end{lemma}
-
-%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
-%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
-%\begin{example}
-%    \label{regular-language:ex}
-%    Let $\Sigma$ be a finite alphabet, and
-%    $L \subseteq \Sigma^*$. Then,
-%    $L$ is a regular language if and only if
-%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
-%    \kl{$\Nat$-polyregular function}.
-%    Furthermore, $L$ is a star-free regular language
-%    if and only if $\ind{L}$ is a
-%    \kl{star-free $\Nat$-polyregular function}.
-%\end{example}
-%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
-%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
-%of first order free variables used to construct the function as in
-%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.
+\AP One of the appeals of $\NPoly$ and $\ZPoly$ are the numerous
+characterizations of these classes in terms of logic, weighted automata, and
+the larger class of \kl{polyregular functions} \cite{CDTL23,DOUE23}. In this
+paper, the main focus will be the connection to weighted automata, which is
+based on the notion of \emph{growth rate}. The \intro{growth rate} of a
+function $f \colon \Sigma^* \to \Rel$ is defined as the minimal $k$ such that
+$\card{f(w)} = \bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that
+the function $f$ has \intro{polynomial growth}. It turns out that for all $k
+\in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
+(resp. in $\NRat)$ that have growth rate at most $k$. Not all
+\kl{$\Nat$-rational series} have \kl{polynomial growth}, a typical example
+being the function $a^n \mapsto 2^n$.
 
 Let us briefly state that \kl{commutativity} is a decidable property of
-\kl{$\Rel$-rational series}. The proof is straightforward in the case of
-\kl{$\Rel$-polyregular functions}, and slightly more involved for general
-\kl{$\Rel$-rational series}.
+\kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}. This
+ensures that the present paper, focusing of \kl{commutative} functions,
+considers functions in a decidable subclass. The proof is straightforward in
+the case of \kl{$\Rel$-polyregular functions}, and slightly more involved for
+general \kl{$\Rel$-rational series}.
 

### 2024-04-23 parts/starfree.tex
mise en forme
move to appendix and clean up

@@
 
-The goal of this section is to go beyond polynomials and consider the case of
-\kl{commutative} \kl{$\Rel$-polyregular functions}. The hypothesis of
-\kl{commutativity} reduces $\ZPoly$ to a combination of a finite number of
-polynomials, arranged by counting modulo some numbers. This characterization
-will be proven in \cref{decompose-polynomial:lem}, and will be the key
-ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$
-inside $\NPoly$. Note that \cref{decompose-polynomial:lem} is actually a
-refinement of \cref{n-poly-combinatorics:lem} under the extra assumption of
-\kl{commutativity}.
+In this section, we leverage the decidability results of \cref{polynomials:sec}
+to decide $\NPoly$ inside $\ZPoly$ and $\NSF$ inside $\NPoly$, both under the
+extra assumption of \kl{commutativity}. The hypothesis of \kl{commutativity}
+reduces $\ZPoly$ to a combination of a finite number of polynomials, arranged
+by counting modulo some numbers. This characterization will be proven in
+\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
+procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
+\cref{decompose-polynomial:lem} refines
+\cref{n-poly-combinatorics:lem} for \kl{commutative} functions.
 
@@
     \end{equation*}
+    \proofref{decompose-polynomial:lem}
 \end{lemma}
@@
 In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
-of \emph{polynomial pumping arguments}
-\cite{DOUE21,DOUE22,CDTL23}, we will provide a
-semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
-to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.
+of \emph{polynomial pumping arguments} \cite{DOUE21,DOUE22,CDTL23}, we will
+provide a semantic property in the form of \cref{k-combinatorial:def}, that is
+conjectured to characterize $\NPoly$ inside $\ZPoly$ in the general setting.
 
@@
     and conversions effective.
+    \proofref{decidable-n-poly:thm}
 \end{theorem}
@@
     Furthermore, membership is decidable and conversions are effective.
+    \proofref{zsf-npoly-nsf:thm}
 \end{theorem}

### 2024-04-23 parts/transducers.tex
move to appendix and clean up

@@
 
-\AP
-Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
+\AP Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
 canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
 approach is to consider as states some representatives for the $\resleq{f}{k}$
-minimal elements of $\Sigma^*$ (assuming there are finitely many of them), all
-computing the same function (\cref{good-residual-ordering:fact}), and define
-transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
-\resleq{f}{k} ua$. This naïve approach does not yield a canonical model, as
-illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
+minimal elements of $\Sigma^*$ (assuming there are finitely many of them),
+%all computing the same function (\cref{good-residual-ordering:fact}),
+and define transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such
+that $v \resleq{f}{k} ua$. This naïve approach does not yield a canonical
+model, as illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
 \cref{non-canonical-transd:fig} computing the function $\BadExOk$ of
-\cref{non-canonical-transd:ex}, and having as states minimal elements for $\resleq{\BadExOk}{0}$.
-To obtain a
-canonical notion, we enforce that the set of states is a \intro{downwards
-closed} subset of $\Sigma^*$ for the prefix ordering $(\prefleq)$, that is, if
-$u \in \Sigma^*$ is a state, then every prefix of $u$ must also be a state.
+\cref{non-canonical-transd:ex}, and having as states minimal elements for
+$\resleq{\BadExOk}{0}$. To obtain a canonical notion, we enforce that the set
+of states is a \intro{downwards closed} subset of $\Sigma^*$ for the prefix
+ordering $(\prefleq)$, that is, every prefix of a state is also a state.
 
@@
         \item For all $u,a \in Q$,
-            $\lambda(u,a) = \app{f}{ua} - \app{f}{\delta(u,a)}$.
+            $\lambda(u,a) = (\app{f}{ua} - \app{f}{\delta(u,a)}) \in \NPoly[k-1]$.
     \end{enumerate}
@@
 actually unique when it exists. We now introduce \cref{residual:algo} to
-compute a \kl{$k$-residual transducer} given a function $f$. Notice that this
+compute the \kl{$k$-residual transducer} given a function $f$. Notice that this
 algorithm requires the ability to test if a function belongs to $\NPoly[k]$,
@@
 functions}. However, the termination of this algorithm also proves the
-existence of the \kl{$k$-residual transducer}.
+existence of the \kl{$k$-residual transducer}. The key argument proving the
+termination of \cref{residual:algo} is based on the fact that for a function $f
+\in \NPoly[k]$, the quasi-ordering $(\resleq{f}{k})$ is a
+\kl{well-quasi-ordering}.
 
-\begin{algorithm}
+\begin{algorithm}[t]
     $Q \defined \{ \varepsilon \}$;
@@
 
-The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
-section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
-\Rightarrow \resleq{f}{k})$ is \kl{well}, but the quasi-ordering $(\resleq{f}{k})$ is a
-\kl{well-quasi-ordering}. This has to be compared with the property of being
-\kl{$(k,\Nat)$-combinatorial}, which states that ultimately, the function is
-increasing in every direction.
-
 \begin{theorem}
@@
     \end{enumerate}
-    If $f$ is assumed to be \kl{commutative}, then all the above
+    If $f$ is assumed to be \kl{commutative}, the  
     properties are decidable, and the conversions effective.
+    \proofref{non-commutative-npoly:thm}
 \end{theorem}
 
-\AP
-We can prove that for functions in $\NPoly[0]$, aperiodicity is equivalent to
-the absence of counters in the \kl{residual transducer} of the function (see
-\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
-higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
-In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
-states computing the function is \kl{counter-free}, which suggests that a more
-sophisticated object is needed to observe aperiodicity. In the spirit of the
-characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
-(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
-ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
-ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
-\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
-This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
-the easy implication is already known to hold
-(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
-\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
-defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
-\in \NSF[k-1]$.
+\AP Unfortunately, the \kl{residual transducer} does not seem to provide an
+algorithm to decide $\NSF$ inside $\NPoly$ in the general setting. While we can
+prove that for functions in $\NPoly[0]$, aperiodicity is equivalent to the
+absence of counters in the \kl{residual transducer} of the function (see
+\cref{aperiodic-iff-residual:lem}), this result does not generalize to higher
+growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
 
@@
 
+In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
+states computing the function is \kl{counter-free}, which suggests that a more
+sophisticated object is needed to observe aperiodicity. In the spirit of the
+characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
+(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
+ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
+ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
+\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
+This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
+the easy implication is already known to hold
+(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
+\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
+defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
+\in \NSF[k-1]$.
+
 \begin{lemma}

### 2024-04-23 parts/conclusion.tex
mise en forme
almost to 12 pages

@@
 Let us conclude with a more general remark on the status of commutative input
-in the study of unary output polyregular functions. Notice that all semantic
+in the study of unary output polyregular functions. Notice that semantic
 characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
-polynomial}, pumpable) arise from pre-composition with a \kl{commutative}
+polynomial}) arise from pre-composition with a \kl{commutative}
 \kl{star-free polyregular function}, as formally stated in
@@
 combinations of commutative functions, leading to
-\cref{z-poly-commutative-encoding:conj}. This conjecture would play the role of
-\cref{decompose-polynomial:lem} in the study of functions in $\ZPoly$.
+\cref{z-poly-commutative-encoding:conj}, which would be a non-commutative
+version of \cref{decompose-polynomial:lem}. For the subclass of 
+\emph{polyblind functions} \cite{LENP21,DOUE22}, parts of the conjecture holds,
+because they essentially behave as commutative functions \cite[Theorem
+6.12]{DOUE23}.
+
+%Indeed, a suitable application of Parikh's Theorem \cite{PARI66}
+%guarantees the existence of the functions $\mathsf{enc}$ and $\mathsf{dec}$,
+%although they may fail to be  \kl{polyregular functions}.
 
@@
             $(f \circ h) \in \ZPoly[k+l-1]$.
+    \proofref{pre-compose-growth-commut:lemma}
 \end{lemma}
@@
             $(f \circ h) \in \ZSF$.
+    \proofref{pre-compose-sf-commut:lemma}
 \end{lemma}
@@
 
-In the specific subclass of \emph{polyblind functions}
-\cite{LENP21,DOUE22}, parts of the conjecture holds,
-because they essentially behave as commutative functions \cite[Theorem
-6.12]{DOUE23}. Indeed, a suitable application of Parikh's Theorem
-\cite{PARI66} guarantees the existence of the functions $\mathsf{enc}$
-and $\mathsf{dec}$, although they may fail to be  \kl{polyregular functions}.
-
 \begin{lemma}
@@
     (resp. \kl{star-free polyregular functions}).
+    \proofref{z-poly-commutative-encoding:remark}
 \end{lemma}

### 2024-04-23 parts/correction.tex
almost to 12 pages

@@
 
-Let us now illustrate how the \kl{discrete differentiation} operator interacts
-with \kl{maximal monomials} with respect to the \kl{Hoare ordering}: it
-extracts \emph{sub-maximal} monomials from a given polynomial.
-
-\begin{lemma}
-    \label{derivation-simplifies:lemma}
-    For all $P \in \Rel[\vec{X}]$ that are non-constant,
-    for all $K \in \Nat$,
-    $\MaximalMonomials(\Diff{K}{P}) \hoarele
-    \MaximalMonomials(P)$.
-    \proofref{derivation-simplifies:lemma}
-\end{lemma}
-
-The following \cref{derivation-stabilises-correct:lem} is the main
-combinatorial argument of this section. It leverages the positivity of the
-\kl{maximal monomials} to compute a shift large enough to make lower degree
-coefficients positive. This lemma will be lifted to a given polynomial $P$ by
-noticing that if $K \in \Nat$, and $P = P_1 + P_2$, then $\translate{K}(P) =
-P_1 + \Diff{K}{P_1} + \translate{K}(P_2)$.
+The key lemma of this section is \cref{lem:correct-to-n-rat}, which is proved
+by induction on the \kl{maximal monomials} of a polynomial $P$. The core
+combinatorial argument allowing to perform the induction is
+\cref{derivation-stabilises-correct:lem}, which is proved by a straightforward
+binomial expansion. Using this lemma, the induction is performed as follows:
+after decomposing a polynomial $P$ into a sum of \kl{maximal monomials} $P_1$
+and the rest $P_2$, notice that for all $K \in \Nat$, $\translate{K}(P) = P_1 +
+\Diff{K}{P_1} + \translate{K}(P_2)$. Then, because $P \in \CorrectPoly$, $P_1$
+is a \kl{$\Nat$-rational polynomial}, furthermore, we can use the induction
+hypothesis on $\Diff{K}{P_1} + \translate{k}(P_2)$ since it has lower degree
+monomials and is in $\CorrectPoly$, finally we can also apply the induction
+hypothesis to $\restr{P}{\nu}$ where $\nu$ is a partial function from $\vec{X}$
+to $\set{0, \dots, K}$ fixing at least one variable. It is then just a matter
+of combining these functions to obtain a \kl{$\Nat$-polyregular function} that
+\kl{represents} $P$. Notice that in \cref{lem:correct-to-n-rat}, polynomials
+can be represented using \kl{star-free $\Nat$-polyregular functions}, which is
+a stronger statement than the one in \cref{karh:thm}, but is in line with the
+idea that polynomials are not periodic functions (for instance, in
+\cref{zsf-nsf:conjecture}).
 
@@
 
-%In the actual statement of \cref{corrected-version:thm}, the representation of
-%polynomials in $\CorrectPoly$ is strengthened from \kl{$\Nat$-polyregular
-%functions} to \kl{star-free $\Nat$-polyregular functions}. This is not
-%surprising, given the conjecture that \kl{ultimately polynomial}
-%\kl{$\Nat$-polyregular functions} (the formal definition will be given in
-%\cref{star-free:sec}) were conjectured to be star free
-%(\cref{zsf-nsf:conjecture}), and that polynomials are always \kl{ultimately
-%polynomial}.
-
 \begin{lemma}
@@
     by a \kl{star-free $\Nat$-polyregular function},
-    which can be explicitly constructed from $P$.
+    which is computable given $P$.
 \end{lemma}
@@
 needed in this characterization, as \cref{bad-poly-translate:ex} illustrates.
+Effectiveness was the last missing piece of our main result
+\cref{corrected-version:thm}, which corrects and generalizes \cref{karh:thm}.
 
@@
 
-We are now ready to state the corrected and generalized version of
-\cref{karh:thm}, which is the main technical contribution of the paper.
-
 \begin{theorem}
@@
 
-The counter example of \cref{thm:counter-example} used three indeterminates,
-and this is not a coincidence: in the particular case of at most two
-indeterminates, the classes $\CorrectPoly$ and $\CoveredPoly$ coincide. In
-particular, the examples appearing in \cite{KARH77} are not invalidated, as
-they all use at most two indeterminates. This is because, in the
-univariate case, being non-negative and having non-negative maximal coefficient
-implies being an \kl{$\Nat$-rational polynomial}.
+For completeness, let us remark that the counter example of
+\cref{thm:counter-example} uses three indeterminates, and this is not a
+coincidence: in the particular case of at most two indeterminates, the classes
+$\CorrectPoly$ and $\CoveredPoly$ coincide. In particular, the examples
+appearing in \cite{KARH77} are not invalidated, as they all use at most two
+indeterminates. This is because, in the univariate case, being non-negative and
+having non-negative maximal coefficient implies being an \kl{$\Nat$-rational
+polynomial}.
 

### 2024-04-23 parts/counter.tex
almost to 12 pages

@@
 
-\AP A polynomial $P \in \Rel[X_1, \dots, X_n]$ is an \intro{$\Nat$-rational
-polynomial} if it is \kl{represented} by a \kl{$\Nat$-polyregular function}.
-The analogue notion of \intro{$\Rel$-rational polynomials} is not of particular
-interest, since every polynomial $P$ is \kl{represented} by a
-\kl{$\Rel$-polyregular function}. It is an easy check that polynomials with
+\AP Before deciding $\NPoly$ inside $\ZPoly$ (which will be done in
+\cref{beyond-polynomials:sec}), let us restrict our attention to simpler
+functions, namely polynomials. A polynomial $P \in \Rel[X_1, \dots, X_n]$ is an
+\intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
+\kl{$\Nat$-polyregular function}. It is an easy check that polynomials with
 coefficients in $\Nat$ are \kl{$\Nat$-rational polynomials}
@@
 a polynomial with negative coefficients that is an \kl{$\Nat$-rational
-polynomial}. As a consequence, the class of interest strictly contains
-$\Nat[\vec{X}]$.
+polynomial}.
+
+The analogue notion of \intro{$\Rel$-rational
+polynomials} is not of particular interest, since every polynomial $P$ is
+\kl{represented} by a \kl{$\Rel$-polyregular function}. Our goal is to decide
+if a \kl{$\Rel$-rational polynomial} (so, any polynomial) is an
+\kl{$\Nat$-rational polynomial}, and if so, to compute an
+\kl{$\Nat$-polyregular function} that \kl{represents} it. 
+This particular problem was claimed to be solved in  \cite{KARH77}, 
+using a characterization restated in \cref{karh:thm}.
 
@@
 \end{example}
-
-The goal of this section is to provide semantic characterizations of
-\kl{$\Nat$-rational polynomials}, decide if a polynomial $P$ is an
-\kl{$\Nat$-rational polynomial}, and if so, explicitly compute an
-\kl{$\Nat$-polyregular function} $f$ that \kl{represents} $P$. To that end, let
-us now introduce the characterization of \kl{$\Nat$-rational polynomials} given
-by \cite{KARH77}, which is stated using a class $\CoveredPoly[\vec{X}]$ of
-polynomials with multiple indeterminates.
 

### 2024-04-23 parts/starfree.tex
almost to 12 pages

@@
     is \intro{ultimately polynomial}
-    whenever there exists $N_0 \in \Nat$ such that
+    when there exists $N_0 \in \Nat$ such that
     for all $n \in \Nat$

### 2024-04-23 parts/transducers.tex
almost to 12 pages

@@
 
-\AP The goal of this section is to provide tools that do not require the
-assumption of \kl{commutativity}. To that end, we introduce the notion of
+\AP The goal of this section is to introduce new tools that do not require the
+assumption of \kl{commutativity}. To that end, we propose a notion of
 \kl{$k$-residual transducer}, that generalizes the notion of residual
 transducer introduced for \kl{$\Rel$-polyregular functions} in \cite{CDTL23}.
-This notion of residual transducer has its own interest, as it shifts the
-attention from (compatible) equivalence relations of finite index that are the
-classical tool in automata theory, towards (compatible) order relations that
-are \kl{well-quasi-ordered}, which is the order counterpart of having finite
-index. Let us recall that a sequence $\seqof{u_i}{i \in \Nat}$ of elements in a
-quasi-ordered set $(X, \leq)$ is \intro{good} whenever there exist $i < j$ such
-that $u_i \leq u_j$. The set $X$ is a \intro{well-quasi-ordering} when every
-infinite sequence is \reintro{good}. A sequence is \intro{bad} when it is not
-\reintro{good}. The notion of \reintro{good sequences} can also be applied to
-binary relations that are not orderings, and a binary relation $R$ for which
-every infinite sequence is \reintro{good} is said to be \intro{well}
-\cite{MELL98}.
+This is done by shifting the attention from (compatible) equivalence relations
+of finite index that are the classical tool in automata theory, towards
+(compatible) order relations that are \kl{well-quasi-ordered}, which is the
+order counterpart of having finite index. Let us recall that a sequence
+$\seqof{u_i}{i \in \Nat}$ of elements in a quasi-ordered set $(X, \leq)$ is
+\intro{good} whenever there exist $i < j$ such that $u_i \leq u_j$. The set $X$
+is a \intro{well-quasi-ordering} when every infinite sequence is
+\reintro{good}. A sequence is \intro{bad} when it is not \reintro{good}. The
+notion of \reintro{good sequences} can also be applied to binary relations that
+are not orderings, and a binary relation $R$ for which every infinite sequence
+is \reintro{good} is said to be \intro{well} \cite{MELL98}.
 
@@
 \AP Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
-canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
-approach is to consider as states some representatives for the $\resleq{f}{k}$
-minimal elements of $\Sigma^*$ (assuming there are finitely many of them),
-%all computing the same function (\cref{good-residual-ordering:fact}),
-and define transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such
-that $v \resleq{f}{k} ua$. This naïve approach does not yield a canonical
-model, as illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
+canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The idea is to
+consider as states the minimal elements of $\Sigma^*$ for $\resleq{f}{k}$, and
+define transitions by letting $\delta(u, a)$ be some state $v$ such that $v
+\resleq{f}{k} ua$. To produce a canonical model, this has to be done carefully,
+as illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
 \cref{non-canonical-transd:fig} computing the function $\BadExOk$ of
 \cref{non-canonical-transd:ex}, and having as states minimal elements for
-$\resleq{\BadExOk}{0}$. To obtain a canonical notion, we enforce that the set
-of states is a \intro{downwards closed} subset of $\Sigma^*$ for the prefix
-ordering $(\prefleq)$, that is, every prefix of a state is also a state.
+$\resleq{\BadExOk}{0}$. To ensure unicity, we ask that the set of states is a
+\intro{downwards closed} subset of $\Sigma^*$ for the prefix ordering
+$(\prefleq)$, i.e. that every prefix of a state is also a state.
 
@@
 
-Because of the restriction on prefixes, the \kl{$k$-residual transducer} is
-actually unique when it exists. We now introduce \cref{residual:algo} to
+We now introduce \cref{residual:algo} to
 compute the \kl{$k$-residual transducer} given a function $f$. Notice that this
@@
     \end{enumerate}
-    If $f$ is assumed to be \kl{commutative}, the  
+    If $f$ is \kl{commutative}, the  
     properties are decidable, and the conversions effective.
@@
 \cref{aperiodic-iff-residual:lem}), this result does not generalize to higher
-growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
+growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}, which
+suggests that a more sophisticated object is needed to observe aperiodicity. In
+the spirit of the characterization of $\NPoly$ based the $(\resleq{f}{k})$
+quasi-ordering (\cref{non-commutative-npoly:thm}), we introduce the notion of
+\intro{aperiodic ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is
+\reintro[aperiodic ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$,
+there exists $N_0 \in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$
+is increasing. This leads to the following
+\cref{sf-no-periods-on-sequences:conj}, of which the easy implication is
+already known to hold (\cref{sf-no-periods-on-sequences:lemma}), and requires
+the introduction of the \emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of
+$(\resleq{f}{k})$, defined by $u \reintro*\resleqsf{f}{k} v$ whenever
+$(\app{f}{v} - \app{f}{u}) \in \NSF[k-1]$.
 
@@
 
-In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
-states computing the function is \kl{counter-free}, which suggests that a more
-sophisticated object is needed to observe aperiodicity. In the spirit of the
-characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
-(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
-ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
-ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
-\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
-This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
-the easy implication is already known to hold
-(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
-\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
-defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
-\in \NSF[k-1]$.
-
 \begin{lemma}
     \label{sf-no-periods-on-sequences:lemma}
-    Let $k \in \Nat$, and $f \in \SF[k]$. Then,
+    Let $k \in \Nat$, and $f \in \NPoly[k]$. If $f \in \NSF[k]$, then
     $(\Sigma^*, \resleqsf{f}{k})$ is an
@@
     \label{sf-no-periods-on-sequences:conj}
-    For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
-    $f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
-    \kl{well-quasi-ordering}.
+    The implication of \cref{sf-no-periods-on-sequences:lemma} is actually an equivalence.
+    %For all $k \in \Nat$ and $f \colon \Sigma^* \to \Rel$,
+    %$f \in \NSF[k]$ if and only if $(\Sigma^*, \resleqsf{f}{k})$ is an \kl[aperiodic ordering]{aperiodic}
+    %\kl{well-quasi-ordering}.
 \end{conjecture}

### 2024-04-24 parts/appendix.tex
almost to 12 pages
moving more to appendix

@@
 \end{remark}
+
+\AP Let us now introduce the ordering over which the induction of
+\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
+set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
+order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
+\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
+ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
+\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
+\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
+$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
+\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
+quasi-ordering]{SCSC12}.
 
@@
 
+\begin{lemma}
+    \label{derivation-simplifies:lemma}
+    For all $P \in \Rel[\vec{X}]$ that are non-constant,
+    for all $K \in \Nat$,
+    $\MaximalMonomials(\Diff{K}{P}) \hoarele
+    \MaximalMonomials(P)$.
+    \proofref{derivation-simplifies:lemma}
+\end{lemma}
 \begin{proofof}[derivation-simplifies:lemma]

### 2024-04-24 parts/correction.tex
moving more to appendix

@@
 
-\AP Let us now introduce the ordering over which the induction of
-\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
-set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
-order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
-\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
-ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
-\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
-\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
-$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
-\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
-quasi-ordering]{SCSC12}.
-
-The key lemma of this section is \cref{lem:correct-to-n-rat}, which is proved
-by induction on the \kl{maximal monomials} of a polynomial $P$. The core
+\AP The key lemma of this section is \cref{lem:correct-to-n-rat}, which is
+proved by induction on the \kl{maximal monomials} of a polynomial $P$. The core
 combinatorial argument allowing to perform the induction is

### 2024-04-24 parts/prelim.tex
almost to 12 pages
moving more to appendix

@@
 Let us briefly state that \kl{commutativity} is a decidable property of
-\kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}. This
-ensures that the present paper, focusing of \kl{commutative} functions,
-considers functions in a decidable subclass. The proof is straightforward in
-the case of \kl{$\Rel$-polyregular functions}, and slightly more involved for
-general \kl{$\Rel$-rational series}.
+\kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}.
 
@@
 
-\subsection{Polynomials} \AP All polynomials considered in this paper have
+\subsection{Polynomials} 
+\AP All polynomials considered in this paper have
 coefficients in $\Rel$ unless explicitly stated otherwise. 

### 2024-04-24 parts/appendix.tex
reshaping appendix

@@
 \section{Proofs of section \ref{preliminaries:sec}}
-
-\AP
-The following
-\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
-and is similar in shape to the \emph{finite counting automata} introduced by
-\cite{SCHU62}.
-
-\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
-    \label{nat-rel-poly:def}
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a function.
-    The \intro{$\Rel$-polyregular function}
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is computed as follows:
-    \begin{equation*}
-        \pi^\dagger (w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is called \intro{$\Nat$-polyregular}.
-    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
-    When the monoid $M$ is \kl(monoid){aperiodic}
-    then
-    the function $f$ is \intro{star-free $\Rel$-polyregular}
-    (resp. \intro{star-free $\Nat$-polyregular}).
-\end{definition}
-
-\textbf{TODO: remove this example}
-\begin{example}
-    \label{size-of-word-nsf:ex}
-    The map $f \colon w \mapsto \card{w} + 1$
-    belongs to $\NSF$.
-    \proofref{size-of-word-nsf:ex}
-\end{example}
 
@@
 \end{lemma}
-
-\AP Let us briefly recall in the following lemmas alternative characterizations
-of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
-analysis of their \kl{commutative} counterpart, and others that provide a nicer
-syntax to construct examples. These characterizations were previously known,
-and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
-\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
-indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
-first-order free variables $\varphi(\vec{x})$, we define
-$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
-a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
-$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
-to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
-$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
-$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
-multiplication with elements $s \in \mathbb{S}$.
-
-\begin{lemma}
-    \label{polynomial-rational-polyreg:fact}
-    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NPoly$ (resp. $\ZPoly$);
-        \item $f$ is a \kl{polyregular function} with output
-            in $\set{1}^*$,
-            post-composed with $\polysum$
-            (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ has \kl{polynomial growth};
-        \item \label{npoly-counting-mso:item} $f$ belongs to
-            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
-            with $\mathbb{S} = \Nat$
-            (resp. $\mathbb{S} = \Rel$).
-    \end{enumerate}
-\end{lemma}
-
-\begin{lemma}
-    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NSF$ (resp. $\ZSF$);
-        \item $f$ is a \kl{star-free polyregular function} with output
-            in $\set{1}^*$,
-            post-composed with $\polysum$
-            (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ belongs to
-            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
-            with $\mathbb{S} = \Nat$
-            (resp. $\mathbb{S} = \Rel$).
-    \end{enumerate}
-\end{lemma}
-
-%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
-%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
-%\begin{example}
-%    \label{regular-language:ex}
-%    Let $\Sigma$ be a finite alphabet, and
-%    $L \subseteq \Sigma^*$. Then,
-%    $L$ is a regular language if and only if
-%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
-%    \kl{$\Nat$-polyregular function}.
-%    Furthermore, $L$ is a star-free regular language
-%    if and only if $\ind{L}$ is a
-%    \kl{star-free $\Nat$-polyregular function}.
-%\end{example}
-%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
-%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
-%of first order free variables used to construct the function as in
-%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.
-
-\begin{proofof}[size-of-word-nsf:ex]
-    Let us define $M \defined (\set{1}, \times)$ which is 
-    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
-    defined by $\mu(w) \defined 1$, and
-    $\pi \colon M^2 \to \Nat$
-    that is the constant function equal to $1$.
-    We check that for all $w \in \Sigma^*$:
-    \begin{equation*}
-        \pi^\dagger(w)
-        =
-        \sum_{uv = w} 1
-        =
-        \card{w} + 1
-        = f(w)
-        \quad . 
-        \qedhere
-    \end{equation*}
-\end{proofof}
 
@@
 
+\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
+equivalent definition of \kl{$\Nat$-polyregular functions} that is more
+combinatorial in nature, and is similar in shape to the \emph{finite counting
+automata} introduced by \cite{SCHU62}.
+
+\begin{lemma}[$\Rel$-polyregular functions {\cite{CDTL23}}]
+    \label{nat-rel-poly:def}
+    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
+    be a morphism, $k \in \Nat$, and 
+    $\pi \colon M^k \to \Rel$ be a function.
+    The following function 
+    $\pi^\dagger \colon \Sigma^* \to \Rel$
+    is \kl{$\Rel$-polyregular}:
+    \begin{equation*}
+        \pi^\dagger (w) \defined
+        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
+        \quad .
+    \end{equation*}
+    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
+    is furthermore \kl{$\Nat$-polyregular}.
+    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
+    When the monoid $M$ is \kl(monoid){aperiodic}
+    then
+    the function $f$ is \kl{star-free $\Rel$-polyregular}
+    (resp. \kl{star-free $\Nat$-polyregular}).
+\end{lemma}
+
+In order to illustrate the above definition, let us provide an example of a
+\kl{$\Nat$-polyregular function} computed using a finite monoid.
+
+\begin{example}
+    \label{size-of-word-nsf:ex}
+    The map $f \colon w \mapsto \card{w} + 1$
+    belongs to $\NSF$.
+\end{example}
+\begin{proof}
+    Let us define $M \defined (\set{1}, \times)$ which is 
+    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
+    defined by $\mu(w) \defined 1$, and
+    $\pi \colon M^2 \to \Nat$
+    that is the constant function equal to $1$.
+    We check that for all $w \in \Sigma^*$:
+    \begin{equation*}
+        \pi^\dagger(w)
+        =
+        \sum_{uv = w} 1
+        =
+        \card{w} + 1
+        = f(w)
+        \quad . 
+        \qedhere
+    \end{equation*}
+\end{proof}
+
 \begin{proofof}[n-poly-combinatorics:lem]
+    We leverage the characterization of \kl{$\Nat$-polyregular functions}
+    given in \cref{nat-rel-poly:def}.
     Let $\omega$ be an idempotent power for the finite monoid $M$,
@@
     it is in fact a \kl{$\Nat$-polyregular function}
-    (\cref{polynomial-rational-polyreg:fact}).
+    (\cite[Theorem 5.22]{DOUE23}).
     Using \cref{n-poly-combinatorics:lem},
@@
     \MaximalMonomials(P)$.
-    \proofref{derivation-simplifies:lemma}
 \end{lemma}
-\begin{proofof}[derivation-simplifies:lemma]
+\begin{proof}
     Let $P \in \Rel[\vec{X}]$ and
@@
     \end{align*}
-\end{proofof}
+\end{proof}
 
@@
 \begin{lemma}
+    \label{polynomial-rational-polyreg:fact}
+    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NPoly$ (resp. $\ZPoly$);
+        \item $f$ is a \kl{polyregular function} with output
+            in $\set{1}^*$,
+            post-composed with $\polysum$
+            (resp. with output in $\set{-1,+1}^*$);
+        \item $f$ has \kl{polynomial growth};
+        \item \label{npoly-counting-mso:item} $f$ belongs to
+            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
+            with $\mathbb{S} = \Nat$
+            (resp. $\mathbb{S} = \Rel$).
+    \end{enumerate}
+\end{lemma}
+
+\begin{lemma}
+    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
+    the following are equivalent:
+    \begin{enumerate}
+        \item $f \in \NSF$ (resp. $\ZSF$);
+        \item $f$ is a \kl{star-free polyregular function} with output
+            in $\set{1}^*$,
+            post-composed with $\polysum$
+            (resp. with output in $\set{-1,+1}^*$);
+        \item $f$ belongs to
+            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
+            with $\mathbb{S} = \Nat$
+            (resp. $\mathbb{S} = \Rel$).
+    \end{enumerate}
+\end{lemma}
+
+\begin{lemma}
     \label{n-poly-k-implies-wqo:lemma}
@@
     Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
-    \proofref{n-poly-k-implies-wqo:lemma}
 \end{lemma}
-\begin{proofof}[n-poly-k-implies-wqo:lemma]
+\begin{proof}
     Because $f \in \NPoly[k]$, there exists
@@
     exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
-\end{proofof}
+\end{proof}
 

### 2024-04-24 parts/prelim.tex
reshaping appendix

@@
 $\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
-$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined similarly.
+$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
 In the rest of the paper, we will always assume that

### 2024-04-24 parts/appendix.tex
appendix cleaned up

@@
 \section{Proofs of section \ref{preliminaries:sec}}
-
-\textbf{TODO: remove this lemma}
-\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
-    \label{transducer-nsf-npoly:lemma}
-    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
-    Then, 
-    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
-    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
-        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
-        \kl{counter-free}).
-\end{lemma}
 
@@
     that is a multiple of $K$.
-    \proofref{derivation-monomials:fact}
 \end{fact}
-\begin{proofof}[derivation-monomials:fact]
+\begin{proof}
     Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
@@
     $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
-\end{proofof}
+\end{proof}
 
@@
 \end{fact}
-\begin{proof}[Proof of \cref{unique-res-transducer:fact}]
+\begin{proof}
     Let $\aTransd_1$ and $\aTransd_2$ be two
@@
 \end{fact}
-\begin{proof}[Proof of
-    \cref{q-o-prefix-cool:fact}]
+\begin{proof}
     Let us write $Q_i$ and $O_i$ for the value of the variables
@@
 \end{lemma}
-
-\begin{proofof}[correct-residual:lemma]
+\begin{proof}
     Because of \cref{q-o-prefix-cool:fact},
@@
     is indeed computing correctly.
-\end{proofof}
+\end{proof}
 
@@
 \end{lemma}
-\begin{proofof}[wqo-implies-termination:lemma]
+\begin{proof}
     Assume towards a contradiction that
@@
     Hence, \cref{residual:algo} must terminate.
-\end{proofof}
-
-\begin{lemma}
-    \label{polynomial-rational-polyreg:fact}
-    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NPoly$ (resp. $\ZPoly$);
-        \item $f$ is a \kl{polyregular function} with output
-            in $\set{1}^*$,
-            post-composed with $\polysum$
-            (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ has \kl{polynomial growth};
-        \item \label{npoly-counting-mso:item} $f$ belongs to
-            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
-            with $\mathbb{S} = \Nat$
-            (resp. $\mathbb{S} = \Rel$).
-    \end{enumerate}
-\end{lemma}
-
-\begin{lemma}
-    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
-    the following are equivalent:
-    \begin{enumerate}
-        \item $f \in \NSF$ (resp. $\ZSF$);
-        \item $f$ is a \kl{star-free polyregular function} with output
-            in $\set{1}^*$,
-            post-composed with $\polysum$
-            (resp. with output in $\set{-1,+1}^*$);
-        \item $f$ belongs to
-            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
-            with $\mathbb{S} = \Nat$
-            (resp. $\mathbb{S} = \Rel$).
-    \end{enumerate}
-\end{lemma}
+\end{proof}
 
@@
     such that
-    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$.
+    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$
+    \cite[Theorem 5.15]{DOUE23}.
 
@@
     \cref{n-poly-1-transd:item} implies \cref{n-poly-k:item} by
-    \cref{transducer-nsf-npoly:lemma}. Then,
+    definition. Then,
     \cref{n-poly-k:item} implies \cref{n-poly-wqo:item} by
@@
     transducer} of $f$ is \kl{counter-free}, then $f \in \NSF$
-    using \cref{transducer-nsf-npoly:lemma}.
+    by definition of $\NSF$.
 
@@
     $\phi_i \in \FO$ has quantifier rank at most $q$
-    and $\card{\vec{x}} = k$.
+    and $\card{\vec{x}} = k$
+    \cite[Theorem 7.10]{DOUE23}.
 

### 2024-04-24 parts/correction.tex
appendix cleaned up

@@
 This follows from the correct implication in the statement of \cref{karh:thm},
-but we provide a self-contained proof using a refinement of the
-classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and
-$\NPoly$ \cite[Lemma 5.37]{DOUE23}.
+but we provide a self-contained proof using a refinement of the classical
+combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and $\NPoly$
+\cite[Lemma 5.37]{DOUE23}. The proof is based on a combinatorial
+characterization of $\NPoly$ reminiscent of the \emph{finite counting automata}
+introduced by \cite{SCHU62}.
 
@@
     which is computable given $P$.
+    \proofref{lem:correct-to-n-rat}
 \end{lemma}

### 2024-04-24 parts/transducers.tex
appendix cleaned up

@@
 
-We now introduce \cref{residual:algo} to
-compute the \kl{$k$-residual transducer} given a function $f$. Notice that this
-algorithm requires the ability to test if a function belongs to $\NPoly[k]$,
-which is only known to be feasible for \kl{commutative} \kl{polyregular
-functions}. However, the termination of this algorithm also proves the
-existence of the \kl{$k$-residual transducer}. The key argument proving the
-termination of \cref{residual:algo} is based on the fact that for a function $f
-\in \NPoly[k]$, the quasi-ordering $(\resleq{f}{k})$ is a
-\kl{well-quasi-ordering}.
+Let us now introduce \cref{residual:algo} to compute the \kl{$k$-residual
+transducer} given a function $f$. Notice that this algorithm requires the
+ability to test if a function belongs to $\NPoly[k]$, which is only known to be
+feasible for \kl{commutative} \kl{polyregular functions}. However, the
+termination of this algorithm also proves the existence of the \kl{$k$-residual
+transducer}. The key argument proving the termination of \cref{residual:algo}
+is based on the fact that for a function $f \in \NPoly[k]$, the quasi-ordering
+$(\resleq{f}{k})$ is a \kl{well-quasi-ordering}. This connection is made
+possible through the characterization of $\NPoly$ as counting the number of
+valuations of some $\MSO$ formulas \cite{KRRC13,CDTL23}.
 

### 2024-04-25 parts/correction.tex
knowledges

@@
 \Diff{K}{P_1} + \translate{K}(P_2)$. Then, because $P \in \CorrectPoly$, $P_1$
-is a \kl{$\Nat$-rational polynomial}, furthermore, we can use the induction
+is a \kl{$\Nat$-rational polynomial}. Furthermore, we can use the induction
 hypothesis on $\Diff{K}{P_1} + \translate{k}(P_2)$ since it has lower degree
@@
 a stronger statement than the one in \cref{karh:thm}, but is in line with the
-idea that polynomials are not periodic functions (for instance, in
+idea that polynomials are not periodic functions (this will be made precise in
 \cref{zsf-nsf:conjecture}).

### 2024-04-25 parts/intro.tex
decidability of decision
knowledges

@@
 
-\AP In order to approach the decidability of decision problems on
+\AP In order to approach the decision problems on
 \kl{polyregular functions}, restricting the output alphabet to a single letter
@@
 as having outputs in $\Nat$, and this is why they were introduced respectively
-as \kl{$\Nat$-polyregular functions} ($\intro*\NPoly$) and \kl{star-free
-$\Nat$-polyregular functions} ($\intro*\NSF$) \cite{DOUE21,DOUE22}. From this
+as \kl{$\Nat$-polyregular functions} ($\NPoly$) and \kl{star-free
+$\Nat$-polyregular functions} ($\NSF$) \cite{DOUE21,DOUE22}. From this
 point of view, $\NPoly$ is a subclass of a larger and well-studied class of
@@
 authors introduced the class of \kl{$\Rel$-polyregular functions}
-($\intro*\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
+($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
 negative outputs. This remains a subclass of \kl{rational series}, more
 precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
-this class of functions the aperiodic variant $\intro*\ZSF$ has been shown to
+this class of functions the aperiodic variant $\ZSF$ has been shown to
 be decidable \cite{CDTL23}. Beware that several non-equivalent notions of

### 2024-04-25 parts/prelim.tex
knowledges

@@
 \AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
-define the classes $\intro*\NPoly[k]$, $\intro*\NSF[k]$, $\intro*\ZPoly[k]$ and
-$\intro*\ZSF[k]$ as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
+define the classes of
+\intro{$\Rel $-polyregular functions}
+$\intro*\ZPoly[k]$,
+\intro{$\Nat $-polyregular functions}
+$\intro*\NPoly[k]$,
+\intro{star-free $\Rel $-polyregular functions}
+$\intro*\ZSF[k]$, and
+\intro{star-free $\Nat $-polyregular functions}
+$\intro*\NSF[k]$, 
+as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
 case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =

### 2024-06-26 parts/intro.tex
update intro

@@
 \AP There have been recurrent attempts to generalize the notions of regularity
-from languages to functions, leading to classes of various expressiveness, such
-as\footnote{This list is far from exhaustive.} Mealy Machines ($\intro*\Mealy$)
-\cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
-rational functions ($\intro*\Rational$) \cite{EILE74}, \intro{regular functions}
-($\intro*\Regular$) \cite{ENMA02}, and \intro{polyregular functions} ($\intro*\Poly$)
-\cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
-defined (respectively, $\intro*\AMealy$, $\intro*\ASequential$,
-$\intro*\ARational$, $\intro*\ARegular$, $\intro*\SF$), lifting the
-correspondence between counter-free automata, star-free languages and aperiodic
-monoids to the functional setting
+and aperiodicity from languages to functions, leading to classes of various
+expressiveness, such as\footnote{This list is far from exhaustive.} Mealy
+Machines ($\intro*\Mealy$) \cite{MEAL55}, sequential functions
+($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
+\cite{EILE74}, \intro{regular functions} ($\intro*\Regular$) \cite{ENMA02}, and
+\intro{polyregular functions} ($\intro*\Poly$) \cite{BOKL19}. For all of these
+models, an \emph{aperiodic} counterpart can be defined (respectively,
+$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
+$\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
+counter-free automata, star-free regular languages and aperiodic monoids to the
+functional setting \textbf{TODO: cite more carefully here}
 \cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
@@
 the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
-larger class $\mathsf{C}$. As an example, to conclude that a function can be
-computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
-enough to prove that it is sequential ($\Sequential$), and computed by an
-\intro{aperiodic polyregular function} ($\SF$).
+larger class $\mathsf{C}$. Because such a (strong) statement does not
+explicitly appear in the literature, a standalone proof will be given in
+\cref{orthogonal-aperiodic-power:thm}. As an example, to conclude that a
+function can be computed by an \emph{aperiodic sequential function}
+($\ASequential$) it is enough to prove that it is sequential ($\Sequential$),
+and computed by an \intro{aperiodic polyregular function} ($\SF$).
 
@@
 automaton or the syntactic monoid of regular languages, over which aperiodicity
-is equivalent to a simple syntactic criterion (e.g. having no counters). This
-is the proof scheme that has been implemented for sequential functions
-\cite{CHOF03}, and more recently for rational functions \cite{FGL16,FGLM18}.
-Unfortunately, such a canonical object remains to be defined for regular and
-polyregular functions.
+is equivalent to a simple syntactic criterion, e.g. having no counters
+\cite{MNPA71}. This is the proof scheme that has been implemented for
+sequential functions \cite{CHOF03}, and more recently for rational functions
+\cite{FGL16,FGLM18}. Unfortunately, such a canonical object remains to be
+defined for regular and polyregular functions.
 
@@
 
-\AP In order to approach the decision problems on
-\kl{polyregular functions}, restricting the output alphabet to a single letter
-has proven to be a fruitful approach \cite{DOUE21,DOUE22}. The key advantage of
-single-letter output alphabets is that the particular order in which letters
-are produced by the machine is irrelevant. This allows to obtain results that
-would otherwise fail because of this ordering, such as the so-called
-\emph{pebble minimization theorem} of \cite{DOUE21}, which fails in the general
-case \cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary
-output} in the case of \kl{polyregular functions} can often be lifted to
-outputs in the free commutative semigroup generated by a finite alphabet
-$\Gamma$. In this sense, the shift to \emph{single letter output alphabet} is
-really about enforcing commutativity on the output. However, even with the
-restriction to single letter output alphabets, the decidability of
-\kl{star-free polyregular functions} among \kl{polyregular functions} remains
-an open problem \cite[Conjecture 7.61]{DOUE23}.
+\AP In order to approach the decision problems on \kl{polyregular functions},
+restricting the output alphabet to a single letter has proven to be a fruitful
+approach \cite{DOUE21,DOUE22}. The key advantage of single-letter output
+alphabets is that the particular order in which letters are produced by the
+machine is irrelevant. This restriction greatly simplifies the definition of
+\kl{polyregular functions}, and opens the door to results that would otherwise
+fail because of this ordering, such as the so-called \emph{pebble minimization
+theorem} of \cite{DOUE21}, which fails in the general case
+\cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary output}
+in the case of \kl{polyregular functions} can often be lifted to outputs in the
+free commutative semigroup generated by a finite alphabet $\Gamma$. In this
+sense, the shift to \emph{single letter output alphabet} is really about
+enforcing commutativity on the output. However, even with the restriction to
+single letter output alphabets, the decidability of \kl{star-free polyregular
+functions} among \kl{polyregular functions} remains an open problem
+\cite[Conjecture 7.61]{DOUE23}.
 
@@
 as having outputs in $\Nat$, and this is why they were introduced respectively
-as \kl{$\Nat$-polyregular functions} ($\NPoly$) and \kl{star-free
-$\Nat$-polyregular functions} ($\NSF$) \cite{DOUE21,DOUE22}. From this
-point of view, $\NPoly$ is a subclass of a larger and well-studied class of
-functions called (non-commutative) \intro{rational series}
-\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
-\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
-\intro{$\Nat$-rational series} ($\intro*\NRat$) that was frequently
-rediscovered in the literature \cite{SCHU62,KRRC13}. In \cite{CDTL23}, the
-authors introduced the class of \kl{$\Rel$-polyregular functions}
-($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
-negative outputs. This remains a subclass of \kl{rational series}, more
-precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
-this class of functions the aperiodic variant $\ZSF$ has been shown to
-be decidable \cite{CDTL23}. Beware that several non-equivalent notions of
-aperiodicity coexist for rational series \cite{REUT80,DRGA19,CDTL23}, and that
-we only refer to the one applying to rational series of polynomial growths
-introduced in \cite{CDTL23}.
+as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
+with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$
+\kl{for star-free polyregular functions} with unary output)
+\cite{DOUE21,DOUE22}. From this point of view, $\NPoly$ is a subclass of a
+larger and well-studied class of functions called (non-commutative)
+\intro{rational series} \cite{REUT80,BERE88,BERE10}. It has been noticed in
+\cite{CDTL23} that \kl{$\Nat$-polyregular functions} correspond to a natural
+subclass of \intro{$\Nat$-rational series} ($\intro*\NRat$), another model of
+computation that aims at generalising the notion of regular language to
+functions with outputs in a semiring \cite{BERE88,BERE10}. It was frequently
+rediscovered in the literature that \kl{$\Nat$-polyregular functions} are a
+subclass of \kl{rational series} \cite{SCHU62,KRRC13,CDTL23}.
+
+In \cite{CDTL23}, the authors introduced the class of \kl{$\Rel$-polyregular
+functions} ($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by
+allowing negative outputs. This remains a subclass of \kl{rational series},
+more precisely, of the class of \intro{$\Rel$-rational series}
+($\intro*\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
+been shown to be decidable \cite{CDTL23}. Beware that several non-equivalent
+notions of aperiodicity coexist for rational series
+\cite{REUT80,DRGA19,CDTL23}, and that we only refer to the one applying to
+\kl{rational series} that are actually in $\ZPoly$ \cite{CDTL23}.
 
@@
 \paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
-functions} that are furthermore \kl{commutative}. Remark that this enforces a
-commutativity property both on the input alphabet and on the output alphabet.
+functions} that are furthermore \kl{commutative}, that is where the output of
+the functions remains unchanged when the letters of the input word are permuted.
 In this restricted setting, we answer positively to \cite[Conjecture
@@
 are decidable in the \kl{commutative} setting. Furthermore, we characterize
-\kl{commutative} \kl{$\Nat$-polyregular function} as the class of
+\kl{commutative} \kl{$\Nat$-polyregular functions} as the class of
 \kl{$\Rel$-polyregular functions} that are \kl{combinatorial} (defined page
-\kpageref{combinatorial}), which is a semantic property akin to the one
+\kpageref{combinatorial}), which is a semantic property akin to the notion of
 \kl{ultimate polynomiality} that characterizes $\ZSF$ inside $\ZPoly$

### 2024-06-26 parts/intro.tex
up to line 150

@@
 \paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
-the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
+the multiple characterizations of \kl{$\Nat$-polyregular functions} (resp.
 $\Rel$-polyregular functions), show that one can decide if a function $f \in

### 2024-06-26 parts/prelim.tex
up to line 150

@@
 
-\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
-integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
-denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
-(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
+\AP In the rest of this paper, we use the symbol $\Rel$ to denote the set of
+integers, $\Rat$ to denote the set of rational numbers, and $\Nat$ to denote
+the set of non-negative integers. The capital letters $\Sigma,\Gamma$ denote
+fixed alphabets, i.e. finite set of letters, and $\Sigma^*, \Gamma^*$ (resp.
+$\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
 $\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
 \in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
-$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
-that the reader is familiar with the basics of automata theory, in particular
-the notions of monoid morphisms, idempotents in monoids, monadic second-order
-($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
-\cite{THOM97}). As aperiodicity will be a central notion of this paper, let us
-recall that a monoid $M$ is \intro(monoid){aperiodic} whenever for all $x \in
-M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.
+$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$.
 
-\AP We also assume the reader to be familiar with \kl{rational series} and
-weighted automata, and refer to \cite{BERE10} for a comprehensive introduction.
-Let us recall that one of the equivalent definitions of a $K$-rational series
-(where $K$ is a ring) is given by the functions from $\Sigma^*$ to $K$ computed
-by weighted automata, which are non-deterministic finite automata with
-transitions labelled by elements of $K$, and whose output on a given word is
-the sum over all accepting paths of the products of the weights along the path.
+\AP
+We assume that the reader is familiar with the basics of automata theory, in
+particular the notions of monoid morphisms, idempotents in monoids, monadic
+second-order ($\MSO$) logic and first-order ($\FO$) logic over finite words
+(see e.g. \cite{THOM97}). As aperiodicity will be a central notion of this
+paper, let us recall that a monoid $M$ is \intro(monoid){aperiodic} whenever
+for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.
+
+\AP Let us briefly recall that one of the equivalent definitions of a
+\intro{$K$-rational series} (where $K$ is a ring), and refer to \cite{BERE10}
+for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
+to $K$ are the functions computed by \intro{$K$-weighted automata} which are
+non-deterministic finite automata with transitions labelled by elements of $K$,
+and whose output on a given word is the sum over all accepting paths of the
+products of the weights along the path. For instance, the function $w \mapsto
+2^{\card{w}}$ is computed by an \kl{$\Nat$-weighted automaton} with a single
+state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
+series}.
 
@@
 that counts occurrences of every letter in the input word (that is, computes
-the Parikh vector) namely: $ \commute[w] \defined (a \mapsto \card[a]{w})$.
-Given a set $X$, a function $f \colon \Sigma^* \to X$ is \intro{commutative}
-whenever for all $u \in \Sigma^*$, for all permutations $\sigma$ of $\set{1,
-\dots, \card{w}}$, $f(\sigma(u)) = f(u)$. Equivalently, it is
-\reintro{commutative} whenever there exists a map $g \colon \Nat^\Sigma \to X$
-such that $g \circ \commute = f$. A function $g \colon \Nat^n \to X$ is
-\intro{represented} by a function $f \colon \Sigma^* \to X$ if there exists a
-map $\eta \colon \set{1, \dots, n} \to \Sigma$ such that $g \circ (v \colon
-\Nat^\Sigma \mapsto \seqof{v(\eta(k))}{1 \leq k \leq n} \colon \Nat^n) \circ
-\commute = f$. This notion will be useful to formally state that a polynomial
-``is'' a \kl{polyregular function}.
+the Parikh vector) namely: $ \commute[w] \defined \seqof{a \mapsto
+\card[a]{w}}{a \in \Sigma}$. Given a set $X$, a function $f \colon \Sigma^* \to
+X$ is \intro{commutative} whenever for all $u \in \Sigma^*$, for all
+permutations $\sigma$ of $\set{1, \dots, \card{w}}$, $f(\sigma(u)) = f(u)$.
+Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
+\Nat^\Sigma \to X$ such that $g \circ \commute = f$.
+
+\AP Let $n \in \Nat$, and let $\Sigma$ be a finite alphabet. Given a function
+$\eta \colon \set{1, \dots, n} \to \Sigma$, we define the $\eta^\dagger \colon
+\Nat^n \to \Sigma^*$ as $\eta^\dagger(\vec{x}) \defined \eta(1)^{x_1} \dots
+\eta(n)^{x_n}$. A function $f \colon \Nat^n \to X$ is \intro{represented} by a
+\kl{commutative} function $g \colon \Sigma^* \to X$ if there exists a map $\eta
+\colon \set{1, \dots, n} \to \Sigma$ such that $g \circ \eta^\dagger = f$. This
+notion will be useful to formally state that a polynomial ``is'' a
+\kl{commutative} \kl{polyregular function}. For instance, the polynomial
+function $P(X,Y) = X \times Y$ is \kl{represented} by the \kl{commutative}
+function $g \colon \set{a,b}^* \to \Rel$ defined by $g(w) \defined \card[a]{w}
+\times \card[b]{w}$.
 
@@
          $Q$ is a finite set of states,
-         $q_0 \in Q$ is an initial state,
+         $q_0 \in Q$ is called the initial state of the transducer,
          $\delta \colon Q \times \Sigma \to Q$
-            is a deterministic transition function,
+            is called the deterministic transition function of the transducer,
          $\lambda \colon Q \times \Sigma \to \mathcal{H}$
-            is a correction function,
-         and $F \colon Q \to \Rel$ is an output function.
+            is called the correction function of the transducer,
+         and $F \colon Q \to \Rel$ is called the output function of the transducer.
 \end{definition}
 
+Based on this syntax, let us now define the semantics of a transducer, in terms
+of the function from $\Sigma^*$ to $\Rel$ that it computes. The intuition is
+that it suffices to apply the correction function $\lambda$ on every transition
+of the underlying automaton, sum the results, and add to this number the output
+associated to the final state of the computation.
+
 \begin{definition}
+    \label{transducer-sem:def}
     Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
@@
     depicted in \cref{non-canonical-transd:fig}.
+    As a consequence, $f \in \NPoly[1]$.
+    Furthermore, because the automaton of 
+        \cref{non-canonical-transd:fig:counter-free}
+        has no \kl{counter},
+    and the correction function only uses functions in $\NSF[0]$,
+    $f \in \NSF[1]$.
 \end{example}
@@
 
-            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
+            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid \ind{\card{w} > 1} + \ind{\card{w} \geq 1}$} (A);
         \end{tikzpicture}
@@
         \caption{This automaton is \kl{counter-free}.}
+        \label{non-canonical-transd:fig:counter-free}
     \end{subfigure}

### 2024-06-26 parts/prelim.tex
up to line 152

@@
 
+\begin{example}
+    \label{simple-transd:ex}
+    Let $\mathsf{Const}$ be the set of constant functions from $\Sigma^*$ to $\Rel$,
+    and let us write $\const{n}$ for the constant function equal to $n \in \Rel$.
+    In 
+    \cref{simple-transd:fig}, we provide a graphical depiction of a
+    \kl{$\mathsf{Const}$-transducer} that \kl{computes} the function
+    $f \colon \set{a,b}^* \to \Rel$ defined by
+    $f(\varepsilon) = 1$,
+    $f(aw) = 3 \times \card{w}$,
+    and $f(bw) = 3 \times \card{w} + 1$.
+\end{example}
+
+\begin{figure}
+    \centering
+    \begin{tikzpicture}[
+        etat/.style={minimum size=2em}
+        ]
+        \node[etat,state,initial,
+            accepting by arrow,
+            accepting text={$1$},
+            accepting where=below,
+            ] (A) at (0,0) {$q_0$};
+        \node[etat,state,
+            accepting by arrow,
+            accepting text={$0$},
+            accepting where=below,
+            ] (B) at (2.5,0) {$q_1$};
+
+        \draw[->] (A) to[bend right=20] node[midway, below] {$a \mid \const{0}$} (B);
+        \draw[->] (A) to[bend left=20] node[midway, above] {$b \mid \const{1}$} (B);
+
+        \draw[->] (B) edge[loop right] node[midway, right]
+            {$a,b \mid \const{3}$} (B);
+    \end{tikzpicture}
+    \caption{A \kl{$\mathsf{Const}$-transducer}, where $\mathsf{Const}$ is the set of constant functions
+        from $\set{a,b}^*$ to $\Rel$,
+        with a set of states $Q \defined \set{ q_0, q_1 }$,
+        an initial state $q_0$,
+        an output function $F \colon Q \to \Rel$ such that $F(q_0) = 1$ and $F(q_1) = 0$,
+        and correction functions $\lambda(q,a) = \const{0}$, $\lambda(q,b) = \const{1}$,
+        and $\lambda(q_1,a) = \lambda(q_1,b) = \const{3}$.
+    }
+    \label{simple-transd:fig}
+\end{figure}
+
 \AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
@@
     $\BadExOk(a) = 0$,
-    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
+    and $\BadExOk(a^n) = n-1$ for all $n \geq 2$.
     The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
@@
 \end{example}
+\begin{proofof}[non-canonical-transd:ex]
+    Let us prove that the transducer of
+    \cref{non-canonical-transd:fig:counter-free}
+    computes $f$.
+    We prove by induction on $w$
+    that $A(q_0, w) = f(w)$,
+    and that $A(q_1, w) = f(aw)$.
+    Let us first remark that $f(au) = \card{u}$.
+
+    When $w = \varepsilon$, $A(q_0, \varepsilon) = F(\varepsilon) = 1 = f(\varepsilon)$.
+    Similarly, $A(q_1, \varepsilon) = F(q_1) = 0 = f(a)$.
+
+    Assume that $w = au$. Then:
+    \begin{align*}
+        A(q_0, w) = A(q_0, au) &= \lambda(q_0, a)(u) + A(q_1, u) \\ 
+                               &= 0 + A(q_1,u) \\
+                               &= f(au) & \text{by induction hypothesis} \\
+                               &= f(w) 
+    \end{align*}
+    Similarly,
+    \begin{align*}
+        A(q_1, w) = A(q_1, au) &= \lambda(q_1, a)(u) + A(q_1, u) \\ 
+                               &= 1 + \ind{\card{u} = 0} + A(q_1,u) \\
+                               &= 1 + \ind{\card{u} = 0} + f(au) & \text{by induction hypothesis} \\
+                               &= \begin{cases}
+                                    2 + f(a) & \text{if $u = \varepsilon$} \\
+                                    1 + f(au) & \text{otherwise}
+                               \end{cases}
+                               \\
+                               &= \begin{cases}
+                                    2 & \text{if $u = \varepsilon$} \\
+                                    1 + \card{u} & \text{otherwise}
+                               \end{cases}
+                               \\
+                               &= \begin{cases}
+                                   f(aa) & \text{if $u = \varepsilon$} \\
+                                   f(
+                               \end{cases}
+    \end{align*}
+\end{proofof}
 
@@
 
-            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid \ind{\card{w} > 1} + \ind{\card{w} \geq 1}$} (A);
+            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
         \end{tikzpicture}
@@
             \draw[->] (B) edge[loop right] node[midway, right]
-                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
+                {$a \mid 1$} (B);
         \end{tikzpicture}
@@
     \caption{Two \kl{$\NPoly[0]$-transducers}
-    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
+    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex},
+    that maps $a^n$ to $n$ for all $n \geq 2$, $a$ to $0$, and $\varepsilon$ to $1$.
+    }
     \label{non-canonical-transd:fig}

### 2024-06-27 parts/appendix.tex
up to 227

@@
 \section{Proofs of section \ref{preliminaries:sec}}
+
+\begin{proofof}[non-canonical-transd:ex]
+    Let us prove that the transducer of
+    \cref{non-canonical-transd:fig:counter-free}
+    computes $f$.
+    We prove by induction on $w$
+    that $A(q_0, w) = f(w)$,
+    and that $A(q_1, w) = f(aw)$.
+    To that end, let us first remark that $f(\varepsilon) = 1$,
+    and $f(au) = \card{u}$.
+
+    When $w = \varepsilon$, $A(q_0, \varepsilon) = F(\varepsilon) = 1 = f(\varepsilon)$.
+    Similarly, $A(q_1, \varepsilon) = F(q_1) = 0 = f(a)$.
+
+    Assume that $w = au$. Then:
+    \begin{align*}
+        A(q_0, w) = A(q_0, au) &= \lambda(q_0, a)(u) + A(q_1, u) \\ 
+                               &= 0 + A(q_1,u) \\
+                               &= f(au) & \text{by induction hypothesis} \\
+                               &= f(w) 
+    \end{align*}
+    Similarly,
+    \begin{align*}
+        A(q_1, w) = A(q_1, au) &= \lambda(q_1, a)(u) + A(q_1, u) \\ 
+                               &= 1 + A(q_1,u) \\
+                               &= 1 + f(au) & \text{by induction hypothesis} \\
+                               &= 1 + \card{u} \\
+                               &= \card{au} \\
+                               &= f(aau) \\
+    \end{align*}
+
+    The proof for the other automaton is similar. The
+    key ingredient in the induction hypothesis is that if $\card{u} \geq 1$, then:
+    $f(au) - f(u) = 1$ and otherwise $f(a) - f(\varepsilon) = -1$.
+    Hence, $f(au) - f(u) = 1 - 2 \times \ind{\card{u} = 0} = \lambda(q_1, a)(u)$.
+\end{proofof}
 

### 2024-06-27 parts/counter.tex
up to 227

@@
 \label{polynomials:sec}
-%\subsection{The counter example}
+%\subsection{The counterexample}
 \label{sec:c-example}
@@
 \cref{beyond-polynomials:sec}), let us restrict our attention to simpler
-functions, namely polynomials. A polynomial $P \in \Rel[X_1, \dots, X_n]$ is an
+functions, namely polynomials. A polynomial $P \in \Rat[X_1, \dots, X_n]$ is an
 \intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
@@
 
-To understand why \cref{karh:thm} is false, we need to understand why some
-non-negative polynomials are not \kl{$\Nat$-rational polynomials}, which is
-illustrated by \cref{non-neg-not-nrat:ex}.
+Before giving a counterexample to the above statement, let us first provide via
+\cref{non-neg-not-nrat:ex} simple proof that there exists some
+\kl{non-negative} polynomial that is not an \kl{$\Nat$-rational polynomial}.
+While the example will not be in $\CoveredPoly$, it illustrates the key
+difference between \kl{non-negative} polynomials and \kl{$\Nat$-rational
+polynomials}. In order to derive this example, we will need the following
+result about the pre-image of regular languages by \kl{polyregular functions}.\footnote{
+    In this particular case, one could have
+    considered more generally \kl{$\Nat$-rational series},
+    and replaced regular languages over a unary alphabet
+    by semi-linear sets.
+}
 
-\begin{fact}
+\begin{theorem}[{\cite[Theorem 1.7]{BOJA18}}]
     \label{pre-image-regular:fact}
-    The pre-image of a regular language by an \kl{$\Nat$-rational series}
+    The pre-image of a regular language by a \kl{polyregular function}
     is a regular language.
-\end{fact}
+\end{theorem}
 
@@
     Let $P(X, Y) \defined (X - Y)^2$.
-    Then $P$ is
-    is \kl{non-negative}, but is
+    Then $P$ is \kl{non-negative}, but is
     not an \kl{$\Nat$-rational polynomial}.
-\end{example}
-\begin{proof}
-    Assume by contradiction that
+    Indeed, assume by contradiction that
     $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
@@
     (\cref{pre-image-regular:fact}),
-    but $S^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
+    but $f^{-1}(\set{0}) = \setof{ w \in \Sigma }{ \card[a]{w} = \card[b]{w} }$
     is not.
-\end{proof}
+\end{example}
 
-Let us now design a counter example to \cref{karh:thm} by suitably tweaking
+Let us now design a counterexample to \cref{karh:thm} by suitably tweaking
 \cref{non-neg-not-nrat:ex} to ensure that the polynomial not only is

### 2024-06-27 parts/prelim.tex
up to 227

@@
 \subsection{Polyregular Functions}
+\label{polyregular:sec}
 
@@
     $f \in \NSF[1]$.
+    \proofref{non-canonical-transd:ex}
 \end{example}
-\begin{proofof}[non-canonical-transd:ex]
-    Let us prove that the transducer of
-    \cref{non-canonical-transd:fig:counter-free}
-    computes $f$.
-    We prove by induction on $w$
-    that $A(q_0, w) = f(w)$,
-    and that $A(q_1, w) = f(aw)$.
-    Let us first remark that $f(au) = \card{u}$.
-
-    When $w = \varepsilon$, $A(q_0, \varepsilon) = F(\varepsilon) = 1 = f(\varepsilon)$.
-    Similarly, $A(q_1, \varepsilon) = F(q_1) = 0 = f(a)$.
-
-    Assume that $w = au$. Then:
-    \begin{align*}
-        A(q_0, w) = A(q_0, au) &= \lambda(q_0, a)(u) + A(q_1, u) \\ 
-                               &= 0 + A(q_1,u) \\
-                               &= f(au) & \text{by induction hypothesis} \\
-                               &= f(w) 
-    \end{align*}
-    Similarly,
-    \begin{align*}
-        A(q_1, w) = A(q_1, au) &= \lambda(q_1, a)(u) + A(q_1, u) \\ 
-                               &= 1 + \ind{\card{u} = 0} + A(q_1,u) \\
-                               &= 1 + \ind{\card{u} = 0} + f(au) & \text{by induction hypothesis} \\
-                               &= \begin{cases}
-                                    2 + f(a) & \text{if $u = \varepsilon$} \\
-                                    1 + f(au) & \text{otherwise}
-                               \end{cases}
-                               \\
-                               &= \begin{cases}
-                                    2 & \text{if $u = \varepsilon$} \\
-                                    1 + \card{u} & \text{otherwise}
-                               \end{cases}
-                               \\
-                               &= \begin{cases}
-                                   f(aa) & \text{if $u = \varepsilon$} \\
-                                   f(
-                               \end{cases}
-    \end{align*}
-\end{proofof}
 
@@
 
-            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
+            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 - 2 \times \ind{\card{w} = 0}$} (A);
         \end{tikzpicture}
@@
 \in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
-(resp. in $\NRat)$ that have growth rate at most $k$. Not all
-\kl{$\Nat$-rational series} have \kl{polynomial growth}, a typical example
-being the function $a^n \mapsto 2^n$.
+(resp. in $\NRat)$ that have growth rate at most $k$ \cite{CDTL23}.
 
@@
 For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
-\kl{monomial}, but $X + Y$ or $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
+\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
 \dots, X_n]$ for the set of \kl{monomials} over these indeterminates.

### 2024-06-27 parts/correction.tex
updates

@@
 \kl{$\Nat$-polyregular functions} are. In this section, we prove that closing
-$\CorrectPoly$ under assignments of variable is enough to recover from
+$\CoveredPoly$ under variable assignments is enough to recover from
 \cref{karh:thm}.
@@
     The class $\intro*\CorrectPoly[\vec{X}]$ is the collection of
-    polynomials $P \in \Rel[\vec{X}]$ such that
-    $P$ is 
-    such that, for every partial function $\nu \colon \vec{X} \topartial \Nat$,
+    polynomials $P \in \Rel[\vec{X}]$ such that,
+    for every partial function $\nu \colon \vec{X} \topartial \Nat$,
     every \kl{maximal monomial} of
@@
 proved by induction on the \kl{maximal monomials} of a polynomial $P$. The core
-combinatorial argument allowing to perform the induction is
+combinatorial argument allowing us to perform the induction is
 \cref{derivation-stabilises-correct:lem}, which is proved by a straightforward
@@
 \Diff{K}{P_1} + \translate{K}(P_2)$. Then, because $P \in \CorrectPoly$, $P_1$
-is a \kl{$\Nat$-rational polynomial}. Furthermore, we can use the induction
-hypothesis on $\Diff{K}{P_1} + \translate{k}(P_2)$ since it has lower degree
-monomials and is in $\CorrectPoly$, finally we can also apply the induction
+is a \kl{$\Nat$-rational polynomial}. Furthermore, using a large enough value
+of $K$, the polynomial $\Diff{K}{P_1} + \translate{K}(P_2)$ has lower degree
+than $P$, and belongs to $\CorrectPoly$. As a consequence, we conclude that
+$\translate{K}(P)$ is an \kl{$\Nat$-rational polynomial}. To conclude that $P$
+itself is a \kl{$\Nat$-rational polynomial}, it suffices to apply the induction
 hypothesis to $\restr{P}{\nu}$ where $\nu$ is a partial function from $\vec{X}$
@@
     non-maximal monomials of $P$.
-    There exists a computable $K$,
+    There exists a computable number $K \in \Nat$,
     such that
@@
     \label{derivation-translation:lem}
-    Let $P \in \Rel[\vec{X}]$, 
-    there exists a computable $K \in \Nat$
+    Let $\vec{X}$ be a tuple of indeterminates,
+    and let $P \in \Rel[\vec{X}]$.
+    There exists a computable number $K \in \Nat$
     such that the following are equivalent:
@@
     \end{enumerate}
-    Furthermore, the membership is decidable.
     \proofref{derivation-translation:lem}
@@
     \label{corrected-version:thm}
-    Let $P \in \Rel[\vec{X}]$ be a polynomial.
+    Let $\vec{X}$ be a tuple of indeterminates,
+    and let $P \in \Rel[\vec{X}]$.
     The following are equivalent:
@@
 appearing in \cite{KARH77} are not invalidated, as they all use at most two
-indeterminates. This is because, in the univariate case, being non-negative and
-having non-negative maximal coefficient implies being an \kl{$\Nat$-rational
-polynomial}.
+indeterminates. Note that the equivalence is clear for the univariate case,
+where being non-negative and having non-negative maximal coefficient clearly
+imply being an \kl{$\Nat$-rational polynomial}.
 

### 2024-06-28 parts/starfree.tex
correcting star-free

@@
 \reintro{$\omega$-type} written $\intro*\moduloType[\omega](n)$ which is
-defined as the pair $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest
-of the Euclidian division of $n$ by $\omega$. This notion of type is lifted to
+defined as the pair $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the  remainder
+of the Euclidean division of $n$ by $\omega$. This notion of type is lifted to
 vectors in $\Nat^p$ for any $p \in \Nat$ by pointwise application.
@@
     there exists a polynomial $P \in \CorrectPoly$,  
-    satisfying for all $X_1, \dots, X_k \geq \omega$:
+    satisfying for all $x_1, \dots, x_k \geq \omega$:
     \begin{equation*}
@@
         \left(
-            \alpha_0 \prod_{i = 1}^k u_i^{\omega X_i} \alpha_i
+            \alpha_0 \prod_{i = 1}^k u_i^{\omega x_i} \alpha_i
         \right)
         = 
-        P(X_1, \dots, X_k) \quad .
+        P(x_1, \dots, x_k) \quad .
     \end{equation*}
-    We say that $f$ is \reintro{combinatorial}
-    if it is \reintro{$(k,\Nat)$-combinatorial} for all $k \in \Nat$.
 \end{definition}
@@
 functions by the means of polynomial behaviors (see
-\cref{aperidic-ultimately-polynomial:ex}).
+\cref{aperidic-ultimately-polynomial:ex}). Note that up until now, we only had
+to consider polynomials with coefficients in $\Rel$. However, there are some
+polynomials with coefficients in $\Rat$ that can be \kl{represented} by a
+\kl{$\Nat$-polyregular function}. For instance, the polynomial $P(X) \defined
+X(X+1)/2$. This is why, in the upcoming definition, we have to consider
+polynomials with coefficients in $\Rat$.
 
@@
     and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
-    \in \Sigma^*$, there exists a polynomial $P \in \Rel[X_1, \dots, X_n]$
-    such that
+    \in \Sigma^*$, there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
+    such that:
     \begin{equation*}
         f\left(
-            \alpha_0 \prod_{i = 1}^{n} w_i^{X_i} \alpha_i
+            \alpha_0 \prod_{i = 1}^{n} w_i^{x_i} \alpha_i
         \right)
         = 
-        P(X_1, \dots, X_n)
+        P(x_1, \dots, x_n)
         \quad 
-        \forall X_1, \dots, X_n \geq N_0
+        \forall x_1, \dots, x_n \geq N_0
         \quad .
@@
 \end{definition}
+
+Let us recall that a language $L$ is \intro(language){aperiodic} whenever it is
+recognized by an \kl(monoid){aperiodic} finite monoid, which turns out to be
+equivalent to the following characterisation among regular languages: for all
+$u,v,w$, there exists $n \in \Nat$, $u v^n w \in L$ if and only if $u v^{n+1} w
+\in L$. As a consequence of this characterization, a language $L$ is aperiodic
+if and only if its indicator function $\ind{L}$ is \kl{ultimately polynomial}.
+This was already observed in \cite[\textbf{statement?}]{CDTL23}, and reproven
+in \cite[\textbf{statement?}]{DOUE23}, but we include a proof hereafter for
+readability and didactic purposes.
 
@@
 \end{example}
+\begin{proof}
+    Assume that $\ind{L}$ is \kl{ultimately polynomial}. Let $u,v,w \in
+    \Sigma^*$. There exists a polynomial $P \in \Rat[X]$ and a number $N_0 \in
+    \Nat$, such that $\ind{L}(u v^n w) = P(n)$ for all $n \geq N_0$. Because
+    $\ind{L}$ has co-domain included in $\set{0,1}$, so does $P$, which proves
+    that $P$ is a constant polynomial. Then, $\ind{L}(u v^{n+1} w) = P(n+1) =
+    P(n) = \ind{L}(uv^n w)$ for all $n \geq N_0$. We have proven that $L$
+    is aperiodic.
 
-\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}
+    Conversely, assume that $L$ is aperiodic, let $n \in \Nat$ be a number, and
+    let $\alpha_0$, $w_1$, $\alpha_1$, $\ldots$, $\alpha_{n-1}$, $w_n$, $\alpha_n$
+    be words in $\Sigma^*$. Because $L$ is \kl(language){aperodic}, the language
+    $L$ is recognized by an \kl(monoid){aperiodic} finite monoid $M$, via a
+    morphism $\mu \colon \Sigma^* \to M$ and an accepting part $F \subseteq M$.
+    Because $M$ is a \kl(monoid){aperiodic}, there exists $N_0$
+    such that for all $m \in M$, $m^{N_0+1} = m^{N_0}$. As a consequence,
+    for all $1 \leq i \leq n$,
+    whenever $x_i \geq N_0$, we have $\mu(w_i)^{x_i} = \mu(w_i)^{N_0}$.
+    Therefore, for all $x_1, \dots, x_n \geq N_0$, we have:
+    \begin{align*}
+        \ind{L}\left(\alpha_0 \prod_{i= 1}^n w_i^{x_i} \alpha_i \right) 
+        &= 
+        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{x_i} \cdot \mu(\alpha_i)\right)
+        & \text{ because $M$ recognizes $L$}
+        \\
+        &= 
+        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{N_0} \cdot \mu(\alpha_i)\right)
+        & \text{ by aperiodicity }
+        \\
+        &\defined P(x_1, \dots, x_n) & \text{ constant polynomial }
+    \end{align*}
+    Which proves that $\ind{L}$ is \kl{ultimately polynomial}.
+\end{proof}
+
+\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
+the membership of $\ZSF$ inside $\ZPoly$,
 relies on the construction of a canonical object called the \emph{residual
@@
 
-We answer positively to \cref{zsf-nsf:conjecture} in the commutative case, by
+We answer
+\cref{zsf-nsf:conjecture} 
+positively in the commutative case, by
 leveraging the semantic characterizations respectively of $\ZSF$ inside

### 2024-06-28 parts/appendix.tex
fixes from tito

@@
     for every \kl{polyregular function} $g \in \Poly[\ell]$,
-    $f \circ g \in \ZPoly[k+ \ell - 1]$
+    $f \circ g \in \ZPoly[k \times \ell]$
     \cite{CDTL23}.

### 2024-06-28 parts/conclusion.tex
fixes from tito

@@
             of \kl{growth rate} $l \in \Nat$,
-            $(f \circ h) \in \ZPoly[k+l-1]$.
+            $(f \circ h) \in \ZPoly[k\times l]$.
     \proofref{pre-compose-growth-commut:lemma}

### 2024-06-28 parts/transducers.tex
knowledges
fixes from tito

@@
 is a \intro{well-quasi-ordering} when every infinite sequence is
-\reintro{good}. A sequence is \intro{bad} when it is not \reintro{good}. The
-notion of \reintro{good sequences} can also be applied to binary relations that
+\kl{good}. A sequence is \intro{bad} when it is not \kl{good}. The
+notion of \kl{good sequences} can also be applied to binary relations that
 are not orderings, and a binary relation $R$ for which every infinite sequence
-is \reintro{good} is said to be \intro{well} \cite{MELL98}.
+is \kl{good} is said to be \intro{well} \cite{MELL98}.
 
@@
 function $f \colon \Sigma^* \to \Rel$,  defined by $\intro*\app{f}{u} \defined
-w \mapsto f(uw)$. The collection of \reintro{residuals} of a function $f$ is
-denoted $\intro*\Res(f)$ and is defined as the set of $\reintro*\app{f}{u}$
+w \mapsto f(uw)$. The collection of \kl{residuals} of a function $f$ is
+denoted $\intro*\Res(f)$ and is defined as the set of $\app{f}{u}$
 where $u$ ranges over words in $\Sigma^*$. Given $k \in \Nat$, we define $f
@@
 interest for a function $f \in \ZPoly[k]$ will be $(\Res(f),
-\reintro*\zpolyequiv[k])$, and $(\Res(f), \reintro*\npolyleq[k])$. However, to
-simplify notations, instead of writing $\app{f}{u} \reintro*\npolyleq[k]
+\zpolyequiv[k])$, and $(\Res(f), \npolyleq[k])$. However, to
+simplify notations, instead of writing $\app{f}{u} \npolyleq[k]
 \app{f}{v}$ when $u,v \in \Sigma^*$, we will use the convenient notation $u
 \intro*\resleq{f}{k} v$, and directly consider the space $(\Sigma^*,
-\reintro*\resleq{f}{k})$.
+\resleq{f}{k})$.
 
@@
 $\resleq{\BadExOk}{0}$. To ensure unicity, we ask that the set of states is a
-\intro{downwards closed} subset of $\Sigma^*$ for the prefix ordering
-$(\prefleq)$, i.e. that every prefix of a state is also a state.
+\intro{downwards closed} subset of $\Sigma^*$ for the \intro{prefix ordering}
+$(\intro*\prefleq)$, i.e. that every prefix of a state is also a state.
 
@@
             for $\resleq{f}{k}$;
-        \item \label{n-poly-residual:item} $f$ is \kl{computed} by a
-            \kl{$k$-residual transducer}.
+        \item \label{n-poly-residual:item} The
+            \kl{$k$-residual transducer}
+            of 
+            $f$ exists.
     \end{enumerate}

### 2024-06-28 parts/appendix.tex
fishied basic fixes

@@
     if it belongs to the Zariski closure, which is decidable
-    \cite{HOPW18}.
+    \cite[\textbf{WHERE}]{HOPW18}.
 \end{proof}
@@
     of rational series is decidable 
-    \cite{BERE10}
+    \cite[\textbf{WHERE}]{BERE10}
     we have obtained a decision procedure.
@@
     Remark that $P = (X-1)^2 + 1$. Let $f  \colon \Sigma^* \to \Nat$
-    that maps $aw \mapsto |w|^2$, and $\varepsilon \to 0$.
+    that maps $aw \mapsto |w|^2$, and $\varepsilon \mapsto 0$.
     It is clear that $f \in \NPoly$, and that 
     $P$ is \kl{represented} by
-    $\ind{|w| \geq 1} \times f + \ind{|w| = 0}$.
+    $\ind{|w| \geq 1} \times f + \ind{|w| = 0} + 1$.
 
@@
 \begin{proofof}[thm:counter-example]
-    It is clear that $\BadPoly$ is \kl{non-negative}. We can develop
+    It is clear that $\BadPoly$ is \kl{non-negative}. We can expand
     the expression of $\BadPoly$ to 
@@
 
-    Remark that $\BadPoly(X,Y,Z) = 0$
-    if and only if $Z(X+Y)^2 = -2 (X-Y)^2$. Hence,
-    $\BadPoly(X,Y,Z)=0$ if and only if $Z = 0$ and $X = Y$, or 
-    $Z \neq 0$, and $X = Y = 0$.
+    Remark that for all $x,y,z \geq 0$, $\BadPoly(x,y,z) = 0$
+    if and only if $z(x+y)^2 = -2 (x-y)^2$. Hence,
+    $\BadPoly(x,y,z)=0$ if and only if $z = 0$ and $x = Y$, or 
+    $z \neq 0$, and $x = y = 0$.
 
@@
     We call $\pi$ the \emph{production function} of $\pi^\dagger$.
-    When the monoid $M$ is \kl(monoid){aperiodic}
+    If the monoid $M$ is \kl(monoid){aperiodic}
     then
@@
 \begin{proofof}[n-poly-combinatorics:lem]
+    \textbf{TODO: reprove}
+
     We leverage the characterization of \kl{$\Nat$-polyregular functions}
@@
     is a product of binomials of the form $\binom{X - l}{s}$
-    which has the desired shape.
+    which almost has the desired shape.
+
+    Indeed, the binomial 
+    $\binom{X - l}{s}$ is equal to
+    $(X-l) \cdots (X-l-s+1) / s!$.
+    As a consequence,
+    $\binom{s! X - l}{s} \in \Rel[X]$,
+    and considering $\omega' \defined \omega \times s!$ (for some large enough $s$)
+    allows us to conclude.
 \end{proofof}
@@
 As one would expect, the \kl{discrete derivatives} are linear operations on
-polynomials, that commutes with \kl{translation operators}. However, the
+polynomials, that commute with \kl{translation operators}. However, the
 \kl{translation operators} do not perfectly commute with partial applications
@@
 
-\begin{remark}
+\begin{fact}
     \label{discrete-deriv-linear:fact}
@@
     \end{enumerate}
-\end{remark}
+\end{fact}
 
@@
 order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
-\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
-ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
-\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
-\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
-$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
-\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
-quasi-ordering]{SCSC12}.
+\kl{well-quasi-ordering} that is isomorphic (once quotiented to be a partial
+ordering) to $\Nat^k$ with the product ordering \cite[Dickson’s Lemma]{SCSC12}.
+We endow $\Pfin(\Monomials)$ with the \intro{Hoare ordering}, that is, $S_1
+\hoareleq S_2$ whenever for all \kl{monomials} $M_1 \in S_1$, there exists a
+monomial $M_2 \in S_2$, such that $M_1$ \kl{divides} $M_2$. The set
+$(\Pfin(\Monomials), \hoareleq)$ remains a \kl{well-quasi-ordering}, and is in
+particular well-founded \cite[Hoare quasi-ordering]{SCSC12}.
 
@@
 
+\textbf{TODO: fix this?}
 \begin{fact}
@@
     for all $1 \leq i \leq n$, $t_i \leq k_i$,
-    and there exists $1 \leq p \leq n$ such that$ t_j < k_j$.
+    and there exists $1 \leq p \leq n$ such that $t_j < k_j$.
     In particular, $T$ \kl{divides}
@@
         \MaximalMonomials(P)
-        \quad .
-        \qedhere
+        & \text{ because } \deg(\Diff{K}{P}) < \deg(P)
     \end{align*}
+    Which concludes the proof.
 \end{proof}
 
+\textbf{Redo the proof}.
 \begin{proofof}[derivation-stabilises-correct:lem]
@@
     compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
-    \kl{represents} $\restr{P}{\nu}$.
+    \kl{represents} $\restr{P}{\nu}$. This is possible because we assumed 
+    that all indeterminates in $\vec{X}$ are used in $P$.
 
@@
     $\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
-    a_1^{X_1} \cdots a_n^{X_n}$, with $X_1, \dots, X_n \geq 0$.
+    a_1^{x_1} \cdots a_n^{x_n}$, with $x_1, \dots, x_n \geq 0$.
 
@@
         \begin{cases}
-            f_\nu(w) & \text{ if } \exists i \in \set{1, \dots, n}, X_i \leq K \\
-            f_\Delta(a_1^{X_1 - K} \cdots a_n^{X_n - K}) & \text{ otherwise }
+            f_{[X_i \mapsto x_i]}(w) & \text{ if } \exists i \in \set{1, \dots, n}, x_i \leq K \\
+            f_\Delta(a_1^{x_1 - K} \cdots a_n^{x_n - K}) & \text{ otherwise }
         \end{cases}
@@
 \begin{proofof}[decompose-polynomial:lem]
+    Let us first write $f \in \ZPoly$ as a linear combination 
+    $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
+
     We apply
-    \cref{n-poly-combinatorics:lem} to compute the $\omega$.
+    \cref{n-poly-combinatorics:lem} to compute the $\omega$ that works for both $f_+$ and $f_-$.
     Using this $\omega$,
@@
     For the converse implication, we use \cref{decompose-polynomial:lem} to
-    effectively compute a polynomial representation of $f$. Using the
+    effectively compute a polynomial representation of $f$.
+
+    \textbf{TODO: this is because they coincide en enough points}.
+    Using the
     assumption, all of these polynomials are actually \kl{$\Nat$-rational
@@
     $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
-    \kl{represented} by a function $f_1 \in \NSF[k]$.
+    \kl{represented} by a function $f_1 \in \NSF[k]$,
+    thanks to
+    \cref{lem:correct-to-n-rat}.
 
@@
     \end{equation}
-    Or, the \kl{commutative} \kl{star-free polyregular function} 
+    Now, the \kl{commutative} \kl{star-free polyregular function} 
     $h \colon \set{1, \dots, k}^* \to \Sigma^*$ that maps
@@
     We are going to prove that if $f \in \ZPoly[k]$ is polyblind, 
-    then the conjecture \cref{z-poly-commutative-encoding:conj} holds when allowing 
+    then the \cref{z-poly-commutative-encoding:conj} holds when allowing 
     for non polyregular functions $\mathsf{enc}$ and $\mathsf{dec}$.

### 2024-06-28 parts/appendix.tex
ok: finished simple things

@@
 \end{proof}
-
-\begin{proofof}[n-poly-combinatorics:lem]
-    \textbf{TODO: reprove}
-
-    We leverage the characterization of \kl{$\Nat$-polyregular functions}
-    given in \cref{nat-rel-poly:def}.
-    Let $\omega$ be an idempotent power for the finite monoid $M$,
-    and
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
-    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
-    $f = \pi^\dagger$. That is, 
-    $f(w)$ is the sum over all factorizations of $w$
-    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
-    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
-
-    Let us define an equivalence relation $\equiv$ over words of
-    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
-    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
-    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
-    repetitions of the factors $u_i^{\omega}$ in the words.
-    We lift the equivalence relation $\equiv$ to $k$-tuples of
-    words by pointwise application.
-
-    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then their
-    production are the same, i.e., $\pi(\mu(v_1), \dots, \mu(v_k)) =
-    \pi(\mu(w_1), \dots, \mu(w_k))$. This holds because $\mu(u_i^\omega) =
-    \mu(u_i)^\omega$ is an idempotent element of the monoid $M$. In particular,
-    $f(w)$ can be computed as the sum over equivalence classes for $\equiv$, of
-    the value of $\pi$ on one representative, which belongs to $\Nat$.
-
-    Therefore, it suffices to prove that the number of partitions of $w$ in a
-    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
-    with \kl{non-negative} \kl{maximal monomials}.
-
-    Notice that to produce all elements in a given equivalence class
-    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
-    back to the normalized version
-    of the equivalence class. For all elements in an equivalence class,
-    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
-    and therefore the number of possible choices
-    is either constant, or 
-    is a product of binomials of the form $\binom{X - l}{s}$
-    which almost has the desired shape.
-
-    Indeed, the binomial 
-    $\binom{X - l}{s}$ is equal to
-    $(X-l) \cdots (X-l-s+1) / s!$.
-    As a consequence,
-    $\binom{s! X - l}{s} \in \Rel[X]$,
-    and considering $\omega' \defined \omega \times s!$ (for some large enough $s$)
-    allows us to conclude.
-\end{proofof}
 
@@
 
-\begin{proofof}[lem:correct-to-n-rat]
-    We prove the result by induction on $\MaximalMonomials(P)$. 
-    In the proof, we write $\vec{X}$ for the indeterminates appearing in $P$,
-    that is, we assume without loss of generality that all indeterminates are used.
-
-    \textbf{Base case:} If the (unique) \kl{maximal monomial} of $P$ is a
-    constant term. Since $P \in \CorrectPoly$, $P = n \in \Nat$, and therefore
-    $P$ is \kl{represented} by a constant \kl{star-free $\Nat$-polyregular
-    function}.
-
-    \textbf{Induction:} Assume that $P$ is not a constant polynomial, and let
-    us write $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
-    monomials} of $P$. We compute a bound $K$ such that $Q \defined
-    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$ using
-    \cref{derivation-stabilises-correct:lem}. Thanks to
-    \cref{derivation-simplifies:lemma}, we also know that $\MaximalMonomials(Q)
-    \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} which is
-    effectively computable.
-
-    Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
-    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular
-    function} (\cref{n-poly-n-poly:example}). As a consequence,
-    $\translate{K}(P) = P_1 + Q$ is (effectively) \kl{represented} by a
-    function $f_\Delta$.
-
-    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
-    K}$ fixing at least one indeterminate,  we know that
-    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
-    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
-    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
-    \kl{represents} $\restr{P}{\nu}$. This is possible because we assumed 
-    that all indeterminates in $\vec{X}$ are used in $P$.
-
-    Let us assume that the alphabet over which the (\kl{commutative}) functions
-    $f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
-    representing the indeterminate $X_i$ of the polynomials. Now, let us define
-    by case analysis the following \kl{commutative} \kl{star-free
-    $\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
-    a_1^{x_1} \cdots a_n^{x_n}$, with $x_1, \dots, x_n \geq 0$.
-
-    \begin{equation*}
-        f(w) \defined
-        \begin{cases}
-            f_{[X_i \mapsto x_i]}(w) & \text{ if } \exists i \in \set{1, \dots, n}, x_i \leq K \\
-            f_\Delta(a_1^{x_1 - K} \cdots a_n^{x_n - K}) & \text{ otherwise }
-        \end{cases}
-        \quad .
-    \end{equation*}
-    Remark that
-    $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}
-    that
-    \kl{represents} $P$.
-\end{proofof}
-
 \begin{proofof}[derivation-translation:lem]
@@
 
-\begin{proofof}[corrected-version:thm]
-    The implications 
-    \cref{corrected-4:item} $\implies$
-    \cref{corrected-3:item} $\implies$
-    \cref{corrected-2:item} are obvious.
-    \cref{lem:correct-to-n-rat} proves
-    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
-    while \cref{n-rat-correct:lem}
-    proves 
-    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
-    Note that the lemmas provide effective conversion procedures,
-    and that \cref{lem:correct-to-n-rat} also provides a decision
-    procedure.
-\end{proofof}
-
-\begin{proofof}[lem:correct-covered-2]
-    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
-    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
-    For the converse inclusion, let us consider $P(X,Y)$
-    that is \kl{non-negative}, such that the \kl{maximal monomials}
-    are  \kl{non-negative}.
-
-    If we fix none of the variables, then the \kl{maximal monomials}
-    are \kl{non-negative} by assumption. If we fix one of the
-    variables, we can assume without loss of generality that we 
-    fix $X = k$ for some $k \in \Nat$.
-    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
-    and therefore must have a positive leading coefficient
-    (which is the unique \kl{maximal monomial} in this case)
-    or be constant equal to 0. In both cases, the \kl{maximal monomials}
-    have positive coefficient.
-    The same reasoning applies \emph{a fortiori} in the case where
-    we fix the two indeterminates, leading to a constant polynomial.
-\end{proofof}
-
 \section{Proofs of section \ref{beyond-polynomials:sec}}
@@
     to be below the threshold $\omega^2$.
-\end{proofof}
-
-\begin{proofof}[decidable-n-poly:thm]
-    Let us prove that
-    \cref{f-npoly-combi:item}
-    implies \cref{f-combinatorial:item}. 
-    We compute the $\omega$ by applying 
-    \cref{n-poly-combinatorics:lem}, and then
-    notice that the polynomial $P$ obtained is
-    by definition a \kl{$\Nat$-rational polynomial}.
-
-    For the converse implication, we use \cref{decompose-polynomial:lem} to
-    effectively compute a polynomial representation of $f$.
-
-    \textbf{TODO: this is because they coincide en enough points}.
-    Using the
-    assumption, all of these polynomials are actually \kl{$\Nat$-rational
-    polynomials}. Let us now construct a new way to compute $f$, by first
-    computing the \kl{$\omega$-type} of the input (which is possible in
-    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
-    which is possible in $\NPoly$.
 \end{proofof}

### 2024-06-28 parts/correction.tex
ok: finished simple things

@@
     \end{equation*}
-    \proofref{n-poly-combinatorics:lem}
 \end{lemma}
+\begin{proof}
+    \textbf{TODO: reprove}
+
+    We leverage the characterization of \kl{$\Nat$-polyregular functions}
+    given in \cref{nat-rel-poly:def}.
+    Let $\omega$ be an idempotent power for the finite monoid $M$,
+    and
+    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
+    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
+    $f = \pi^\dagger$. That is, 
+    $f(w)$ is the sum over all factorizations of $w$
+    into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
+    of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
+
+    Let us define an equivalence relation $\equiv$ over words of
+    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
+    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
+    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
+    repetitions of the factors $u_i^{\omega}$ in the words.
+    We lift the equivalence relation $\equiv$ to $k$-tuples of
+    words by pointwise application.
+
+    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then their
+    production are the same, i.e., $\pi(\mu(v_1), \dots, \mu(v_k)) =
+    \pi(\mu(w_1), \dots, \mu(w_k))$. This holds because $\mu(u_i^\omega) =
+    \mu(u_i)^\omega$ is an idempotent element of the monoid $M$. In particular,
+    $f(w)$ can be computed as the sum over equivalence classes for $\equiv$, of
+    the value of $\pi$ on one representative, which belongs to $\Nat$.
+
+    Therefore, it suffices to prove that the number of partitions of $w$ in a
+    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
+    with \kl{non-negative} \kl{maximal monomials}.
+
+    Notice that to produce all elements in a given equivalence class
+    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
+    back to the normalized version
+    of the equivalence class. For all elements in an equivalence class,
+    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
+    and therefore the number of possible choices
+    is either constant, or 
+    is a product of binomials of the form $\binom{X - l}{s}$
+    which almost has the desired shape.
+
+    Indeed, the binomial 
+    $\binom{X - l}{s}$ is equal to
+    $(X-l) \cdots (X-l-s+1) / s!$.
+    As a consequence,
+    $\binom{s! X - l}{s} \in \Rel[X]$,
+    and considering $\omega' \defined \omega \times s!$ (for some large enough $s$)
+    allows us to conclude.
+\end{proof}
 
@@
     which is computable given $P$.
-    \proofref{lem:correct-to-n-rat}
 \end{lemma}
+\begin{proof}
+    We prove the result by induction on $\MaximalMonomials(P)$. 
+    In the proof, we write $\vec{X}$ for the indeterminates appearing in $P$,
+    that is, we assume without loss of generality that all indeterminates are used.
+
+    \textbf{Base case:} If the (unique) \kl{maximal monomial} of $P$ is a
+    constant term. Since $P \in \CorrectPoly$, $P = n \in \Nat$, and therefore
+    $P$ is \kl{represented} by a constant \kl{star-free $\Nat$-polyregular
+    function}.
+
+    \textbf{Induction:} Assume that $P$ is not a constant polynomial, and let
+    us write $P = P_1 + P_2$ where $P_1$ is the sum of the \kl{maximal
+    monomials} of $P$. We compute a bound $K$ such that $Q \defined
+    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$ using
+    \cref{derivation-stabilises-correct:lem}. Thanks to
+    \cref{derivation-simplifies:lemma}, we also know that $\MaximalMonomials(Q)
+    \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
+    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} which is
+    effectively computable.
+
+    Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
+    (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular
+    function} (\cref{n-poly-n-poly:example}). As a consequence,
+    $\translate{K}(P) = P_1 + Q$ is (effectively) \kl{represented} by a
+    function $f_\Delta$.
+
+    For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
+    K}$ fixing at least one indeterminate,  we know that
+    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
+    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
+    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
+    \kl{represents} $\restr{P}{\nu}$. This is possible because we assumed 
+    that all indeterminates in $\vec{X}$ are used in $P$.
+
+    Let us assume that the alphabet over which the (\kl{commutative}) functions
+    $f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
+    representing the indeterminate $X_i$ of the polynomials. Now, let us define
+    by case analysis the following \kl{commutative} \kl{star-free
+    $\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
+    a_1^{x_1} \cdots a_n^{x_n}$, with $x_1, \dots, x_n \geq 0$.
+
+    \begin{equation*}
+        f(w) \defined
+        \begin{cases}
+            f_{[X_i \mapsto x_i]}(w) & \text{ if } \exists i \in \set{1, \dots, n}, x_i \leq K \\
+            f_\Delta(a_1^{x_1 - K} \cdots a_n^{x_n - K}) & \text{ otherwise }
+        \end{cases}
+        \quad .
+    \end{equation*}
+    Remark that
+    $f$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular function}
+    that
+    \kl{represents} $P$.
+\end{proof}
 
@@
 \end{theorem}
+\begin{proof}
+    The implications 
+    \cref{corrected-4:item} $\implies$
+    \cref{corrected-3:item} $\implies$
+    \cref{corrected-2:item} are obvious.
+    \cref{lem:correct-to-n-rat} proves
+    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
+    while \cref{n-rat-correct:lem}
+    proves 
+    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
+    Note that the lemmas provide effective conversion procedures,
+    and that
+    \cref{derivation-translation:lem}
+    also provides a decision
+    procedure.
+\end{proof}
 
@@
     $\CorrectPoly[X,Y] = \CoveredPoly[X,Y]$.
-    \proofref{lem:correct-covered-2}
 \end{lemma}
+\begin{proof}
+    It is clear that $\CorrectPoly[X,Y] \subseteq \CoveredPoly[X,Y]$,
+    by considering the empty valuation $\nu \colon \set{X,Y} \topartial \Nat$.
+    For the converse inclusion, let us consider $P(X,Y)$
+    that is \kl{non-negative}, such that the \kl{maximal monomials}
+    are  \kl{non-negative}.
+
+    If we fix none of the variables, then the \kl{maximal monomials}
+    are \kl{non-negative} by assumption. If we fix one of the
+    variables, we can assume without loss of generality that we 
+    fix $X = k$ for some $k \in \Nat$.
+    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
+    and therefore must have a positive leading coefficient
+    (which is the unique \kl{maximal monomial} in this case)
+    or be constant equal to 0. In both cases, the \kl{maximal monomials}
+    have positive coefficient.
+    The same reasoning applies \emph{a fortiori} in the case where
+    we fix the two indeterminates, leading to a constant polynomial.
+\end{proof}

### 2024-06-28 parts/intro.tex
ok: finished simple things

@@
 defined in terms of deterministic finite automata and regular expressions
-\cite{KLEE56}, Monadic Second Order Logic ($\intro*\MSO$) \cite{BUCH60}, or
-finite monoids \cite{SCHU61}. Among regular languages, there exists a robust
-subclass sharing similar characterizations, namely the \emph{star-free} regular
-languages. These are obtained via counter-free (minimal) automata
-\cite{MNPA71}, represented by star-free regular expressions \cite{SCHU65},
-definable in First Order Logic ($\intro*\FO$) \cite{PEPI86}, and recognized by
-aperiodic monoids \cite{SCHU65}. It is decidable whether a regular language is
-aperiodic, thanks to the characterizations in terms of minimal automaton and/or
-syntactic monoid.
+\cite{KLEE56}, Monadic Second Order Logic ($\intro*\MSO$)
+\cite{BUCH60,ELGO61,TRAK66}, or finite monoids \cite{RABI59,SCHU61}. Among regular
+languages, there exists a robust subclass sharing similar characterizations,
+namely the \emph{star-free} regular languages. These are obtained via
+counter-free (minimal) automata \cite{MNPA71}, represented by star-free regular
+expressions \cite{SCHU65}, definable in First Order Logic ($\intro*\FO$)
+\cite{MNPA71}, and recognized by aperiodic monoids \cite{SCHU65}. It is
+decidable whether a regular language is aperiodic, thanks to the
+characterizations in terms of minimal automaton and/or syntactic monoid.
 
@@
 ($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
-\cite{EILE74}, \intro{regular functions} ($\intro*\Regular$) \cite{ENMA02}, and
-\intro{polyregular functions} ($\intro*\Poly$) \cite{BOKL19}. For all of these
-models, an \emph{aperiodic} counterpart can be defined (respectively,
-$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
+\cite{EILE74}, \intro{regular functions} ($\intro*\Regular$) \cite{ENHO01}, and
+\intro{polyregular functions} ($\intro*\Poly$) \cite{ENMA02,BOJA18,BOKL19}. For
+all of these models, an \emph{aperiodic} counterpart can be defined
+(respectively, $\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
 $\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
 counter-free automata, star-free regular languages and aperiodic monoids to the
-functional setting \textbf{TODO: cite more carefully here}
-\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
-these classes of functions are all known to be strict, and we depicted in
+functional setting: 
+for Mealy Machines \cite{SCHU65,MNPA71},
+for sequential functions \cite{CHOF03}, 
+for rational functions \cite{FGL16,FGLM18}, for regular
+functions \cite{BOJA14,CADA15,DJR16,DGK21}, and for polyregular
+functions \cite{BDK18,BOKL19}. The inclusions between these classes of
+functions are all known to be strict, and we depicted in
 \cref{computational-models:fig} the status of the \emph{membership problem}
 associated to these strict inclusions (that is, given a function $f$, decide if
-$f$ can be computed by a function of a proper subclass). Note that in
-\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
-$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
-the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
-larger class $\mathsf{C}$. Because such a (strong) statement does not
-explicitly appear in the literature, a standalone proof will be given in
-\cref{orthogonal-aperiodic-power:thm}. As an example, to conclude that a
-function can be computed by an \emph{aperiodic sequential function}
-($\ASequential$) it is enough to prove that it is sequential ($\Sequential$),
-and computed by an \intro{aperiodic polyregular function} ($\SF$).
+$f$ can be computed by a function of a proper subclass).
+
+%Note that in
+%\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
+%$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
+%the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
+%larger class $\mathsf{C}$. Because such a (strong) statement does not
+%explicitly appear in the literature, a standalone proof will be given in
+%\cref{orthogonal-aperiodic-power:thm}. As an example, to conclude that a
+%function can be computed by an \emph{aperiodic sequential function}
+%($\ASequential$) it is enough to prove that it is sequential ($\Sequential$),
+%and computed by an \intro{aperiodic polyregular function} ($\SF$).
 

### 2024-06-28 parts/starfree.tex
ok: finished simple things

@@
     and conversions effective.
-    \proofref{decidable-n-poly:thm}
 \end{theorem}
+\begin{proof}
+    Let us prove that
+    \cref{f-npoly-combi:item}
+    implies \cref{f-combinatorial:item}. 
+    We compute the $\omega$ by applying 
+    \cref{n-poly-combinatorics:lem}, and then
+    notice that the polynomial $P$ obtained is
+    by definition a \kl{$\Nat$-rational polynomial}.
+
+    For the converse implication, we use \cref{decompose-polynomial:lem} to
+    effectively compute a polynomial representation of $f$.
+
+    \textbf{TODO: this is because they coincide en enough points}.
+    Using the
+    assumption, all of these polynomials are actually \kl{$\Nat$-rational
+    polynomials}. Let us now construct a new way to compute $f$, by first
+    computing the \kl{$\omega$-type} of the input (which is possible in
+    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
+    which is possible in $\NPoly$.
+\end{proof}
 

### 2024-07-04 parts/starfree.tex
test

@@
 if and only if its indicator function $\ind{L}$ is \kl{ultimately polynomial}.
-This was already observed in \cite[\textbf{statement?}]{CDTL23}, and reproven
-in \cite[\textbf{statement?}]{DOUE23}, but we include a proof hereafter for
-readability and didactic purposes.
+This was already observed in \cite[Claim V.6]{CDTL23}, and reproven in
+\cite[Claim 7.45, Lemma 7.53]{DOUE23}. We include a self-contained proof
+hereafter for readability and didactic purposes.
 
@@
 \end{theorem}
+
+\textbf{TODO: prove the following conjecture to conclude on polynomials in general
+and not just Z coefficients.}
+\begin{conjecture}
+    Let $\vec{X}$ be a vector of variables, and $P \in \Rel[\vec{X}]$ be a
+    polynomial. The following properties are decidable
+    \begin{itemize}
+        \item Is $P$ \kl{represented} by a \kl{$\Rel$-polyregular function}?
+        \item Is $P$ \kl{represented} by a \kl{$\Nat$-polyregular function}?
+    \end{itemize}
+\end{conjecture}

### 2024-07-04 parts/appendix.tex
update for CSL«

@@
     (\cref{pre-image-regular:fact}). 
-\end{proofof}
-
-\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
-equivalent definition of \kl{$\Nat$-polyregular functions} that is more
-combinatorial in nature, and is similar in shape to the \emph{finite counting
-automata} introduced by \cite{SCHU62}.
-
-\begin{lemma}[$\Rel$-polyregular functions {\cite{CDTL23}}]
-    \label{nat-rel-poly:def}
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a function.
-    The following function 
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is \kl{$\Rel$-polyregular}:
-    \begin{equation*}
-        \pi^\dagger (w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
-        \quad .
-    \end{equation*}
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is furthermore \kl{$\Nat$-polyregular}.
-    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
-    If the monoid $M$ is \kl(monoid){aperiodic}
-    then
-    the function $f$ is \kl{star-free $\Rel$-polyregular}
-    (resp. \kl{star-free $\Nat$-polyregular}).
-\end{lemma}
-
-In order to illustrate the above definition, let us provide an example of a
-\kl{$\Nat$-polyregular function} computed using a finite monoid.
-
-\begin{example}
-    \label{size-of-word-nsf:ex}
-    The map $f \colon w \mapsto \card{w} + 1$
-    belongs to $\NSF$.
-\end{example}
-\begin{proof}
-    Let us define $M \defined (\set{1}, \times)$ which is 
-    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
-    defined by $\mu(w) \defined 1$, and
-    $\pi \colon M^2 \to \Nat$
-    that is the constant function equal to $1$.
-    We check that for all $w \in \Sigma^*$:
-    \begin{equation*}
-        \pi^\dagger(w)
-        =
-        \sum_{uv = w} 1
-        =
-        \card{w} + 1
-        = f(w)
-        \quad . 
-        \qedhere
-    \end{equation*}
-\end{proof}
-
-\begin{proofof}[n-rat-correct:lem]
-    Let $f$ be the \kl{commutative}
-    \kl{$\Nat$-rational series}
-    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
-    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
-    it is in fact a \kl{$\Nat$-polyregular function}
-    (\cite[Theorem 5.22]{DOUE23}).
-    Using \cref{n-poly-combinatorics:lem},
-    there exists a number $\omega \in \Nat$
-    and polynomial $Q$ with \kl{non-negative}
-    \kl{maximal monomials} such that
-    for all $X_1, \dots, X_p \geq \omega$:
-    \begin{equation*}
-        f\left(
-            \prod_{i = 1}^p a_i^{\omega X_i}
-        \right)
-        = Q(X_1, \dots, X_p)
-        = P(\omega X_1, \dots, \omega X_p) 
-        \quad .
-    \end{equation*}
-    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
-    as polynomials,
-    and in particular that
-    the \kl{maximal monomials} of 
-    $P$ are \kl{non-negative}.
-
-    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
-    the polynomial $\restr{P}{\nu}$ continues to be represented
-    by a \kl{$\Nat$-polyregular function}, namely
-    $f$ partially applied to a word. As a consequence,
-    the \kl{maximal monomials} of
-    $\restr{P}{\nu}$ are also \kl{non-negative}, 
-    and
-    we have proven that $P \in \CorrectPoly$.
 \end{proofof}

### 2024-07-04 parts/correction.tex
update for CSL«

@@
 \end{corollary}
+\begin{proof}
+    Let $f$ be the \kl{commutative}
+    \kl{$\Nat$-rational series}
+    with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
+    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
+    it is in fact a \kl{$\Nat$-polyregular function}
+    (\cite[Theorem 5.22]{DOUE23}).
+    Using \cref{n-poly-combinatorics:lem},
+    there exists a number $\omega \in \Nat$
+    and polynomial $Q$ with \kl{non-negative}
+    \kl{maximal monomials} such that
+    for all $X_1, \dots, X_p \geq \omega$:
+    \begin{equation*}
+        f\left(
+            \prod_{i = 1}^p a_i^{\omega X_i}
+        \right)
+        = Q(X_1, \dots, X_p)
+        = P(\omega X_1, \dots, \omega X_p) 
+        \quad .
+    \end{equation*}
+    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
+    as polynomials,
+    and in particular that
+    the \kl{maximal monomials} of 
+    $P$ are \kl{non-negative}.
+
+    Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
+    the polynomial $\restr{P}{\nu}$ continues to be represented
+    by a \kl{$\Nat$-polyregular function}, namely
+    $f$ partially applied to a word. As a consequence,
+    the \kl{maximal monomials} of
+    $\restr{P}{\nu}$ are also \kl{non-negative}, 
+    and
+    we have proven that $P \in \CorrectPoly$.
+\end{proof}
 

### 2024-07-04 parts/prelim.tex
update for CSL«

@@
 $\Nat$-valued) functions, but could consider other commutative semigroups.
+
+\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
+equivalent definition of \kl{$\Nat$-polyregular functions} that is more
+combinatorial in nature, and is similar in shape to the \emph{finite counting
+automata} introduced by \cite{SCHU62}.
+
+\begin{lemma}[$\Rel$-polyregular functions {\cite{CDTL23}}]
+    \label{nat-rel-poly:def}
+    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
+    be a morphism, $k \in \Nat$, and 
+    $\pi \colon M^k \to \Rel$ be a function.
+    The following function 
+    $\pi^\dagger \colon \Sigma^* \to \Rel$
+    is \kl{$\Rel$-polyregular}:
+    \begin{equation*}
+        \pi^\dagger (w) \defined
+        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
+        \quad .
+    \end{equation*}
+    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
+    is furthermore \kl{$\Nat$-polyregular}.
+    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
+    If the monoid $M$ is \kl(monoid){aperiodic}
+    then
+    the function $f$ is \kl{star-free $\Rel$-polyregular}
+    (resp. \kl{star-free $\Nat$-polyregular}).
+\end{lemma}
+
+In order to illustrate the above definition, let us provide an example of a
+\kl{$\Nat$-polyregular function} computed using a finite monoid.
+
+\begin{example}
+    \label{size-of-word-nsf:ex}
+    The map $f \colon w \mapsto \card{w} + 1$
+    belongs to $\NSF$.
+\end{example}
+\begin{proof}
+    Let us define $M \defined (\set{1}, \times)$ which is 
+    a finite \kl{aperiodic monoid}, $\mu \colon \Sigma^* \to M$
+    defined by $\mu(w) \defined 1$, and
+    $\pi \colon M^2 \to \Nat$
+    that is the constant function equal to $1$.
+    We check that for all $w \in \Sigma^*$:
+    \begin{equation*}
+        \pi^\dagger(w)
+        =
+        \sum_{uv = w} 1
+        =
+        \card{w} + 1
+        = f(w)
+        \quad . 
+        \qedhere
+    \end{equation*}
+\end{proof}
 

### 2024-07-04 parts/appendix.tex
test

@@
 
-\begin{proof}[A simpler proof of \cref{decidable-commutative-rat:lemma} in the general case]
-    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
-    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
-    As a consequence, a function $f$ is commutative if and only if
-    $f \circ c = f = f \circ t$.
-    When $f$ is a rational series,
-    $f \circ c$ and $f \circ t$ are both rational series that can be
-    effectively computed from $f$, and since equivalence
-    of rational series is decidable 
-    \cite[\textbf{WHERE}]{BERE10}
-    we have obtained a decision procedure.
-\end{proof}
-
 \section{Proofs of section \ref{polynomials:sec}}
@@
     are non-negative.
-\end{proofof}
-
-\begin{proofof}[thm:counter-example]
-    It is clear that $\BadPoly$ is \kl{non-negative}. We can expand
-    the expression of $\BadPoly$ to 
-    obtain
-    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
-    The \kl{maximal monomials} of $P$
-    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
-    \kl{non-negative}.
-
-    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
-    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
-    There exists a \kl{commutative}
-    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
-    such that for all $w \in \Sigma^*$,
-    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-
-    Remark that for all $x,y,z \geq 0$, $\BadPoly(x,y,z) = 0$
-    if and only if $z(x+y)^2 = -2 (x-y)^2$. Hence,
-    $\BadPoly(x,y,z)=0$ if and only if $z = 0$ and $x = Y$, or 
-    $z \neq 0$, and $x = y = 0$.
-
-    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
-    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
-    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ not a regular language,
-    we
-    conclude that $L$ is not a regular language.
-    However, $L = f^{-1}(\set{0})$ is a regular language
-    (\cref{pre-image-regular:fact}). 
 \end{proofof}
@@
     to be below the threshold $\omega^2$.
-\end{proofof}
-
-\begin{proofof}[zsf-npoly-nsf:thm]
-    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
-
-    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
-    only the empty word $\varepsilon$, and because $f$ is a
-    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
-    Note that $f$ is therefore a constant function, equal to a natural
-    number, hence belongs to $\NSF$.
-
-    \textbf{Induction.}
-    Because $f$ is \kl{ultimately polynomial},
-    there exists $N_0 \in \Nat$, 
-    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
-    such that
-    \begin{equation*}
-        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
-        = 
-        P(\seqof{X_a}{a \in \Sigma})
-        \quad
-        \text{ when }
-        \forall a \in \Sigma, X_a \geq N_0
-        \quad .
-    \end{equation*}
-    Furthermore, leveraging \cref{decompose-polynomial:lem},
-    we obtain $\omega \in \Nat$
-    and a family of polynomials 
-    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^\Sigma}$
-    that represent $f$.
-    For all $t \in \ModuloTypes[\omega]^\Sigma$
-    that have threshold greater than $\omega^2$,
-    we conclude that $P_t$ and $P$ coincide (up to translation) over an infinite grid,
-    hence we conclude that
-    if all letters appear at least $N_1 \defined \max (\omega^2, N_0)$ times,
-    then $f$ computes $P$. But this also shows that
-    $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
-    \kl{represented} by a function $f_1 \in \NSF[k]$,
-    thanks to
-    \cref{lem:correct-to-n-rat}.
-
-    To compute $f$ on words where some letter appears fewer than $N_1$ times,
-    we partially apply $f$ to obtain a function $g$ over a restricted input
-    alphabet. Using the induction hypothesis, $g$ belongs to $\NSF[k]$, and we
-    can therefore (effectively) compute a representation of $f$ in $\NSF[k]$.
-
-    For the effective part of the theorem, 
-    we know from \cite[Theorem V.13]{CDTL23}, it is decidable whether
-    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
-    it is decidable whether $f \in \NPoly$.
-    Furthermore, one can enumerate functions in $\NSF[k]$,
-    and check equivalence to effectively represent $f$ as a
-    \kl{star-free $\Nat$-polyregular function}.
 \end{proofof}

### 2024-07-04 parts/prelim.tex
test

@@
 \end{lemma}
+\begin{proof}
+    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
+    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
+    As a consequence, a function $f$ is commutative if and only if
+    $f \circ c = f = f \circ t$.
+    When $f$ is a rational series,
+    $f \circ c$ and $f \circ t$ are both rational series that can be
+    effectively computed from $f$, and since equivalence
+    of rational series is decidable 
+    \cite[\textbf{WHERE}]{BERE10}
+    we have obtained a decision procedure.
+\end{proof}
 

### 2024-07-04 globals/metadata.tex
big edits
publish to arxiv
update for CSL«

--- 2024-02-12 globals/metadata.tex
@@
 %TODO mandatory; please add comma-separated list of keywords
-\keywords{Rational series, weighted automata, polyregular function, commutative input} 
+\keywords{Rational series, weighted automata, polyregular function, commutative} 
 \category{} %optional, e.g. invited paper
@@
 \newcommand{\acknowledge}{
-    Acknowledge Gaëtan
+    I would like to thank Gaëtan Douéneau for his help and support.
 }
@@
     \begin{abstract}
-        This paper answers two questions about 
-        commutative unary output polyregular functions, called
-        $\Nat$-polyregular functions, that forms a proper subset
-        of $\Nat$-rational series.
-        We show that given a $\Rel$-rational series,
-        it is decidable whether it is computable via a commutative
-        $\Nat$-polyregular function, and provide a counter-example
-        to the theorem of Karhumäki that studied the same question
-        in the case of polynomials.
-
-        We also prove that it is decidable whether a commutative $\Nat$-polyregular
-        function is star-free, by proving the stronger statement
-        that star-free $\Rel$-polyregular functions that are
-        $\Nat$-polyregular are computable using a star-free $\Nat$-polyregular function.
-        This answers positively to previous conjectures on $\Nat$-polyregular functions,
-        and the study of the commutative case seems to be useful
-        towards a proof in the non-commutative case.
+This paper studies which functions computed by $\Rel$-weighted automata can be
+realised by $\Nat$-weighted automata, under two extra assumptions:
+commutativity (the order of letters in the input does not matter) and
+polynomial growth (the ouptut of the function is bounded by a polynomial in the
+size of the input). This continues the work of Karhumäki, who studied the same
+question under the assumption that the computed function was a polynomial,
+which is more restrictive. Answering this question starts by exhibiting a
+counterexample to the theorem of Karhumäki, proposing a new and valid
+characterization in the case of polynomials, that lifts to the more general
+setting of polynomial growth and commutativity. Surprisingly, we manage to
+leverage this effective characterization to decide whether a function computed
+by a commutative $\Nat$-weighted automaton of polynomial growth is star-free, a
+notion borrowed from the theory of regular languages that has been the subject
+of many investigations in the context of string-to-string functions
+in the last decade. 
     \end{abstract}
@@
 \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{ICALP 2023}
+\EventShortTitle{CSL 2025}
 \EventAcronym{ICALP}
-\EventYear{2023}
+\EventYear{2024}
 \EventDate{December 24--27, 2016}

### 2024-07-04 parts/appendix.tex
update for CSL«

@@
     \cite[\textbf{WHERE}]{HOPW18}.
+\end{proof}
+
+\begin{proof}[A simpler proof of \cref{decidable-commutative-rat:lemma} in the general case]
+    Remark that the group of permutations of $\set{1, \dots, n}$ is generated by
+    the cycle $c \defined (n,1, \dots, n-1)$ and the transposition $t \defined (1, 2)$.
+    As a consequence, a function $f$ is commutative if and only if
+    $f \circ c = f = f \circ t$.
+    When $f$ is a rational series,
+    $f \circ c$ and $f \circ t$ are both rational series that can be
+    effectively computed from $f$, and since equivalence
+    of rational series is decidable 
+    \cite[\textbf{WHERE}]{BERE10}
+    we have obtained a decision procedure.
 \end{proof}

### 2024-07-04 parts/correction.tex
update for CSL«

@@
     Let $f$ be an \kl{$\Nat$-polyregular} function. 
-    There exists a computable $\omega \in \Nat$
+    There exists a computable $\omega \in \Nat_{\geq 1}$
     such that for all $p \in \Nat$,
@@
     Using \cref{n-poly-combinatorics:lem},
-    there exists a number $\omega \in \Nat$
+    there exists a number $\omega \in \Nat_{\geq 1}$
     and polynomial $Q$ with \kl{non-negative}

### 2024-07-04 parts/counter.tex
update for CSL«

@@
 \end{lemma}
+\begin{proof}
+    It is clear that $\BadPoly$ is \kl{non-negative}. We can expand
+    the expression of $\BadPoly$ to 
+    obtain
+    $\BadPoly = ZX^2 + ZY^2 + 2ZXY + 2X^2 -4XY + 2Y^2$.
+    The \kl{maximal monomials} of $P$
+    are $ZX^2$, $ZY^2$, and $2ZXY$, all of which are
+    \kl{non-negative}.
+
+    Assume by contradiction that $\BadPoly$ is an \kl{$\Nat$-rational polynomial}.
+    Let $\Sigma \defined \set{a,b,c}$ be a finite alphabet.
+    There exists a \kl{commutative}
+    \kl{$\Nat$-polyregular function} $f \colon \Sigma^* \to \Nat$
+    such that for all $w \in \Sigma^*$,
+    $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
+
+    Remark that for all $x,y,z \geq 0$, $\BadPoly(x,y,z) = 0$
+    if and only if $z(x+y)^2 = -2 (x-y)^2$. Hence,
+    $\BadPoly(x,y,z)=0$ if and only if $z = 0$ and $x = Y$, or 
+    $z \neq 0$, and $x = y = 0$.
+
+    Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the
+    above computation, we conclude that $L = \setof{ w \in \set{a,b}^* }{
+    \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
+    Because $L \cap \set{a,b}^*$ not a regular language,
+    we
+    conclude that $L$ is not a regular language.
+    However, $L = f^{-1}(\set{0})$ is a regular language
+    (\cref{pre-image-regular:fact}). 
+\end{proof}

### 2024-07-04 parts/intro.tex
update for CSL«

@@
 ($\intro*\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
-been shown to be decidable \cite{CDTL23}. Beware that several non-equivalent
-notions of aperiodicity coexist for rational series
+been shown to be decidable \cite{CDTL23}. Let us underline that several
+non-equivalent notions of aperiodicity coexist for rational series
 \cite{REUT80,DRGA19,CDTL23}, and that we only refer to the one applying to
@@
 restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
-was conjectured that, analogously to classes appearing in
-\cref{computational-models:fig}, the following equation would hold: $\NSF =
-\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
-decidability of aperiodicity for single letter output polyregular functions
-since aperiodicity is decidable for $\ZSF$. We refer the reader to
-\cref{previously-known-inclusions:fig} to get an overview of the different
-classes of functions appearing in the paper.
+was conjectured that, being ``computable using $\Nat$ coefficients'' and being
+``star-free'' were two orthogonal properties, and therefore that the following
+equation would hold: $\NSF = \NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}.
+This would imply the decidability of aperiodicity for single letter output
+polyregular functions since aperiodicity is decidable for $\ZSF$. We refer the
+reader to \cref{previously-known-inclusions:fig} to get an overview of the
+different classes of functions appearing in the paper.
 
@@
 
-\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
+\subparagraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
 functions} that are furthermore \kl{commutative}, that is where the output of
@@
 
-Finally, we make a first step towards understanding the non-commutative case by
-introducing a canonical model of computation for \kl{$\Nat$-polyregular
-functions} built upon residuals, called the \kl{residual transducer}. This
-residual transducer is obtained by shifting attention from the notion of
-equivalence relations of finite index over $\Sigma^*$, used in traditional
-automata theory and in the study of $\ZPoly$, to the notion of
-\kl{well-quasi-orderings}. The residual transducer is effectively computable if
-$f$ is \kl{commutative} and \kl{$\Nat$-polyregular}, and it is conjectured to
-be computable for all functions in $\NPoly$.
-
 The approach taken in this paper avoids the use of tools from the theory of
@@
 traditional method used to understand (commutative output) \kl{polyregular
-functions} \cite{DOUE21,DOUE22,CDTL23}, and instead
-focuses on intrinsic description of the functions themselves (via canonical
-computational models such as the \kl{residual transducer} of
-\cref{residual-transducer:def}, or explicit encoding/decoding into polynomials
-via \cref{decompose-polynomial:lem}).
+functions} \cite{DOUE21,DOUE22,CDTL23}, and instead focuses on intrinsic
+description of the functions themselves, via  explicit encoding/decoding into
+polynomials (see for instance \cref{decompose-polynomial:lem}).
 
-\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
+\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
 the multiple characterizations of \kl{$\Nat$-polyregular functions} (resp.
@@
 \cref{thm:counter-example}, and an alternative characterization in
-\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
+\cref{corrected-version:thm}. Finally, in \cref{beyond-polynomials:sec}, we leverage the
 results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
@@
 holds for \kl{commutative} functions
-(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
-introduce the notion of \kl{residual transducer}
-(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
-that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
-\kl{well-quasi-ordered} collection of residuals.
+(\cref{zsf-npoly-nsf:thm}).

### 2024-07-04 parts/starfree.tex
update for CSL«

@@
 \end{theorem}
+\begin{proof}
+    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
+
+    \textbf{Base Case.} When $\Sigma$ is empty, $\Sigma^*$ contains
+    only the empty word $\varepsilon$, and because $f$ is a
+    \kl{$\Nat$-polyregular function}, $f(\varepsilon) \in \Nat$.
+    Note that $f$ is therefore a constant function, equal to a natural
+    number, hence belongs to $\NSF$.
+
+    \textbf{Induction.}
+    Because $f$ is \kl{ultimately polynomial},
+    there exists $N_0 \in \Nat$, 
+    and $P \in \Rel[\seqof{X_a}{a \in \Sigma}]$ 
+    such that
+    \begin{equation*}
+        f\left( \prod_{a \in \Sigma} a^{X_a} \right)
+        = 
+        P(\seqof{X_a}{a \in \Sigma})
+        \quad
+        \text{ when }
+        \forall a \in \Sigma, X_a \geq N_0
+        \quad .
+    \end{equation*}
+    Furthermore, leveraging \cref{decompose-polynomial:lem},
+    we obtain $\omega \in \Nat$
+    and a family of polynomials 
+    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^\Sigma}$
+    that represent $f$.
+    For all $t \in \ModuloTypes[\omega]^\Sigma$
+    that have threshold greater than $\omega^2$,
+    we conclude that $P_t$ and $P$ coincide (up to translation) over an infinite grid,
+    hence we conclude that
+    if all letters appear at least $N_1 \defined \max (\omega^2, N_0)$ times,
+    then $f$ computes $P$. But this also shows that
+    $\translate{N_1}(P)$ is a \kl{$\Nat$-rational polynomial}, hence is
+    \kl{represented} by a function $f_1 \in \NSF[k]$,
+    thanks to
+    \cref{lem:correct-to-n-rat}.
+
+    To compute $f$ on words where some letter appears fewer than $N_1$ times,
+    we partially apply $f$ to obtain a function $g$ over a restricted input
+    alphabet. Using the induction hypothesis, $g$ belongs to $\NSF[k]$, and we
+    can therefore (effectively) compute a representation of $f$ in $\NSF[k]$.
+
+    For the effective part of the theorem, 
+    we know from \cite[Theorem V.13]{CDTL23}, it is decidable whether
+    $f \in \ZSF[k]$ and from \cref{decidable-n-poly:thm}
+    it is decidable whether $f \in \NPoly$.
+    Furthermore, one can enumerate functions in $\NSF[k]$,
+    and check equivalence to effectively represent $f$ as a
+    \kl{star-free $\Nat$-polyregular function}.
+\end{proof}
 

### 2024-07-05 parts/appendix.tex
starving the beast

--- 2024-07-04 parts/appendix.tex
@@
 
-\begin{proofof}[n-poly-n-poly:example]
-    The polynomials of $\Nat[\vec{X}]$
-    are obtained from basic functions (constant in $\Nat$,
-    $X_i$ for some $1 \leq i \leq n$)
-    by products and sums. Because the basic functions are
-    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
-    that are closed under these operations, we conclude.
-\end{proofof}
-
 \begin{proofof}[negative-not-nrat:ex]
@@
 \end{proofof}
-
-As one would expect, the \kl{discrete derivatives} are linear operations on
-polynomials, that commute with \kl{translation operators}. However, the
-\kl{translation operators} do not perfectly commute with partial applications
-$\restr{\cdot}{\nu}$.
-
-\begin{fact}
-    \label{discrete-deriv-linear:fact}
-    For all tuples $\vec{X}$ of indeterminates,
-    for all $K \in \Nat$, $L \in \Rel$, for all partial functions
-    $\nu \colon \vec{X} \topartial \Nat$:
-    \begin{enumerate}
-        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
-        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
-        and
-        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
-            are linear operators,
-        \item $\Diff{K}{ \cdot } \circ \translate{L}
-            = \translate{L} \circ \Diff{K}{\cdot}$,
-        \item $\restr{\cdot}{\nu} \circ \translate{L}
-            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
-    \end{enumerate}
-\end{fact}
-
-\AP Let us now introduce the ordering over which the induction of
-\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
-set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
-order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
-\kl{well-quasi-ordering} that is isomorphic (once quotiented to be a partial
-ordering) to $\Nat^k$ with the product ordering \cite[Dickson’s Lemma]{SCSC12}.
-We endow $\Pfin(\Monomials)$ with the \intro{Hoare ordering}, that is, $S_1
-\hoareleq S_2$ whenever for all \kl{monomials} $M_1 \in S_1$, there exists a
-monomial $M_2 \in S_2$, such that $M_1$ \kl{divides} $M_2$. The set
-$(\Pfin(\Monomials), \hoareleq)$ remains a \kl{well-quasi-ordering}, and is in
-particular well-founded \cite[Hoare quasi-ordering]{SCSC12}.
-
-\begin{fact}
-    \label{maxi-monomials-submodular:fact}
-    Let $P, Q$ be two polynomials.
-    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
-    \cup \MaximalMonomials(Q)$.
-\end{fact}
-
-\begin{fact}
-    \label{translation-maximal:fact}
-    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
-    \begin{equation*}
-        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
-        \quad .
-    \end{equation*}
-\end{fact}
-
-\textbf{TODO: fix this?}
-\begin{fact}
-    \label{derivation-monomials:fact}
-    Let $K > 0$,
-    and let $M,T$ be two \kl{monomials}, such that
-    $T$ \kl{strictly divides} $M$.
-    Then,
-    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
-    and 
-    $T$ \kl{divides} some \kl{maximal monomial}
-    of $\Diff{K}{M}$, which has a coefficient
-    that is a multiple of $K$.
-\end{fact}
-\begin{proof}
-    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
-    where $k_1, \dots, k_n > 0$ and $n > 0$.
-    Let us compute
-    \begin{align*}
-        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
-                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
-    \end{align*}
-    Therefore, 
-    the \kl{maximal monomials} of $\Diff{K}{M}$
-    are precisely
-    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
-    which \kl{strictly divides} $M$.
-    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
-    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
-    for all $1 \leq i \leq n$, $t_i \leq k_i$,
-    and there exists $1 \leq p \leq n$ such that $t_j < k_j$.
-    In particular, $T$ \kl{divides}
-    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
-\end{proof}
-
-\begin{lemma}
-    \label{derivation-simplifies:lemma}
-    For all $P \in \Rel[\vec{X}]$ that are non-constant,
-    for all $K \in \Nat$,
-    $\MaximalMonomials(\Diff{K}{P}) \hoarele
-    \MaximalMonomials(P)$.
-\end{lemma}
-\begin{proof}
-    Let $P \in \Rel[\vec{X}]$ and
-    $K \in \Nat$.
-    Leveraging 
-    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
-    \begin{align*}
-        \MaximalMonomials\left(\Diff{K}{P}\right)
-        &=
-        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
-        &\hoareleq
-        \bigcup_{M \in \Monomials(P)} 
-        \MaximalMonomials(\Diff{K}{M}) \\
-        &\hoarele
-        \MaximalMonomials(P)
-        & \text{ because } \deg(\Diff{K}{P}) < \deg(P)
-    \end{align*}
-    Which concludes the proof.
-\end{proof}
 

### 2024-07-05 parts/conclusion.tex
starving the beast

@@
 \end{lemma}
-
-\begin{conjecture}
-    \label{z-poly-commutative-encoding:conj}
-    Let $k \in \Nat$, and let $f \in \ZPoly[k]$ (resp. $\ZSF[k]$).
-    There exists a finite set $Q$, 
-    parameters $l_q \in \Nat$ for all $q \in Q$,
-    \kl{commutative} polyregular functions (resp. \kl{star-free polyregular})
-    $(g_q)_{q \in Q}$
-    from $\Nat^{l_q}$ to $\Rel$,
-    a surjective polyregular function (resp. \kl{star-free polyregular})
-    $\mathsf{dec} \colon \Sigma^* \tosurj \sum_{q \in Q} \Nat^{l_q}$,
-    and a polyregular function (resp. \kl{star-free polyregular})
-    $\mathsf{enc} \colon \sum_{q \in Q} \Nat^{l_q} \to \Sigma^*$,
-    such that
-    $f = (\sum_{q \in Q} g_q) \circ \mathsf{dec}$
-    and $\mathsf{dec} \circ \mathsf{enc} = \mathsf{id}$.
-\end{conjecture}
-
-\begin{lemma}
-    \label{z-poly-commutative-encoding:remark}
-    The \cref{z-poly-commutative-encoding:conj} holds for the
-    subclass of polyblind functions when dropping the assumption that
-    $\mathsf{enc}$ and $\mathsf{dec}$ are \kl{polyregular functions}
-    (resp. \kl{star-free polyregular functions}).
-    \proofref{z-poly-commutative-encoding:remark}
-\end{lemma}

### 2024-07-05 parts/correction.tex
starving the beast

@@
 combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and $\NPoly$
-\cite[Lemma 5.37]{DOUE23}. The proof is based on a combinatorial
-characterization of $\NPoly$ reminiscent of the \emph{finite counting automata}
-introduced by \cite{SCHU62}.
+\cite[Lemma 5.37]{DOUE23}.
 
@@
     Let $f$ be the \kl{commutative}
-    \kl{$\Nat$-rational series}
+    \kl{$\Nat$-polyregular function}
     with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
-    that \kl{represents} $P$. Because $f$ has \kl{polynomial growth},
-    it is in fact a \kl{$\Nat$-polyregular function}
-    (\cite[Theorem 5.22]{DOUE23}).
+    that \kl{represents} $P$. 
     Using \cref{n-poly-combinatorics:lem},
@@
     by a \kl{$\Nat$-polyregular function}, namely
-    $f$ partially applied to a word. As a consequence,
+    $f$ partially applied to a word $u$ and using a restricted input alphabet. As a consequence,
     the \kl{maximal monomials} of
@@
 
-\AP The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
+\AP
+We have proven that every \kl{$\Nat$-rational polynomial} is in $\CorrectPoly$,
+and the rest of this section is devoted to proving the converse implication.
+The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
 \intro{discrete derivative} to perform an induction on the \kl{maximal monomials}.
@@
 \cref{zsf-nsf:conjecture}).
+
+As one would expect, the \kl{discrete derivatives} are linear operations on
+polynomials, that commute with \kl{translation operators}. However, the
+\kl{translation operators} do not perfectly commute with partial applications
+$\restr{\cdot}{\nu}$.
+
+\begin{fact}
+    \label{discrete-deriv-linear:fact}
+    For all tuples $\vec{X}$ of indeterminates,
+    for all $K \in \Nat$, $L \in \Rel$, for all partial functions
+    $\nu \colon \vec{X} \topartial \Nat$:
+    \begin{enumerate}
+        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
+        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
+        and
+        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
+            are linear operators,
+        \item $\Diff{K}{ \cdot } \circ \translate{L}
+            = \translate{L} \circ \Diff{K}{\cdot}$,
+        \item $\restr{\cdot}{\nu} \circ \translate{L}
+            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
+    \end{enumerate}
+\end{fact}
+
+\AP Let us now introduce the ordering over which the induction of
+\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
+set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
+order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
+\kl{well-quasi-ordering} that is isomorphic (once quotiented to be a partial
+ordering) to $\Nat^k$ with the product ordering \cite[Dickson’s Lemma]{SCSC12}.
+We endow $\Pfin(\Monomials)$ with the \intro{Hoare ordering}, that is, $S_1
+\hoareleq S_2$ whenever for all \kl{monomials} $M_1 \in S_1$, there exists a
+monomial $M_2 \in S_2$, such that $M_1$ \kl{divides} $M_2$. The set
+$(\Pfin(\Monomials), \hoareleq)$ remains a \kl{well-quasi-ordering}, and is in
+particular well-founded \cite[Hoare quasi-ordering]{SCSC12}.
+
+\begin{fact}
+    \label{maxi-monomials-submodular:fact}
+    Let $P, Q$ be two polynomials.
+    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
+    \cup \MaximalMonomials(Q)$.
+\end{fact}
+
+\begin{fact}
+    \label{translation-maximal:fact}
+    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
+    \begin{equation*}
+        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
+        \quad .
+    \end{equation*}
+\end{fact}
+
+\textbf{TODO: fix this?}
+\begin{fact}
+    \label{derivation-monomials:fact}
+    Let $K > 0$,
+    and let $M,T$ be two \kl{monomials}, such that
+    $T$ \kl{strictly divides} $M$.
+    Then,
+    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
+    and 
+    $T$ \kl{divides} some \kl{maximal monomial}
+    of $\Diff{K}{M}$, which has a coefficient
+    that is a multiple of $K$.
+\end{fact}
+\begin{proof}
+    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
+    where $k_1, \dots, k_n > 0$ and $n > 0$.
+    Let us compute
+    \begin{align*}
+        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
+                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
+    \end{align*}
+    Therefore, 
+    the \kl{maximal monomials} of $\Diff{K}{M}$
+    are precisely
+    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
+    which \kl{strictly divides} $M$.
+    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
+    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
+    for all $1 \leq i \leq n$, $t_i \leq k_i$,
+    and there exists $1 \leq p \leq n$ such that $t_j < k_j$.
+    In particular, $T$ \kl{divides}
+    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
+\end{proof}
+
+\begin{lemma}
+    \label{derivation-simplifies:lemma}
+    For all $P \in \Rel[\vec{X}]$ that are non-constant,
+    for all $K \in \Nat$,
+    $\MaximalMonomials(\Diff{K}{P}) \hoarele
+    \MaximalMonomials(P)$.
+\end{lemma}
+\begin{proof}
+    Let $P \in \Rel[\vec{X}]$ and
+    $K \in \Nat$.
+    Leveraging 
+    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
+    \begin{align*}
+        \MaximalMonomials\left(\Diff{K}{P}\right)
+        &=
+        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
+        &\hoareleq
+        \bigcup_{M \in \Monomials(P)} 
+        \MaximalMonomials(\Diff{K}{M}) \\
+        &\hoarele
+        \MaximalMonomials(P)
+        & \text{ because } \deg(\Diff{K}{P}) < \deg(P)
+    \end{align*}
+    Which concludes the proof.
+\end{proof}
 
@@
 \end{proof}
+
+\begin{corollary}
+    \label{decide-rat-poly-npoly:cor}
+    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
+    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
+    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
+    \begin{enumerate}
+        \item $P$ is a \kl{$\Nat$-rational polynomial}.
+        \item $\alpha P \in \CorrectPoly$,
+            and for all $\vec{x} \in \Nat^{\vec{X}}$,
+            $P(\vec{x})$ is a multiple of $\alpha$.
+    \end{enumerate}
+    In particular, the properties are decidable.
+\end{corollary}
+\begin{proof}
+    If $P$ is a \kl{$\Nat$-rational polynomial}, then ...
+    If $\alpha P \in \CorrectPoly$, then  it is an \kl{$\Nat$-rational polynomial}.
+    But in particular, there exists a polyregular function $f$ that represents $\alpha P$
+    with outputs in $\set{ a }$.
+    Now, let us post-compose this function by the one that
+    maps $a^{k \alpha + r}$ to $a^{k}$, which is a \kl{polyregular function}.
+    We conclude that $P$ is \kl{represented} by an \kl{$\Nat$-polyregular function}.
+\end{proof}

### 2024-07-05 parts/counter.tex
starving the beast

@@
 \cref{beyond-polynomials:sec}), let us restrict our attention to simpler
-functions, namely polynomials. A polynomial $P \in \Rat[X_1, \dots, X_n]$ is an
+functions, namely polynomials. A polynomial $P \in \Rel[X_1, \dots, X_n]$ is an
 \intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
-\kl{$\Nat$-polyregular function}. It is an easy check that polynomials with
-coefficients in $\Nat$ are \kl{$\Nat$-rational polynomials}
+\kl{$\Nat$-rational series}, in which case it is also \kl{represented} by a
+\kl{$\Nat$-polyregular function}, because it has \kl{polynomial growth}
+(\cref{polyregular-polynomial-growth:lemma}). It is an easy check that
+polynomials with coefficients in $\Nat$ are \kl{$\Nat$-rational polynomials}
 (\cref{n-poly-n-poly:example}). However, \cref{negative-but-npoly:ex} provides
 a polynomial with negative coefficients that is an \kl{$\Nat$-rational
-polynomial}.
-
-The analogue notion of \intro{$\Rel$-rational
-polynomials} is not of particular interest, since every polynomial $P$ is
-\kl{represented} by a \kl{$\Rel$-polyregular function}. Our goal is to decide
-if a \kl{$\Rel$-rational polynomial} (so, any polynomial) is an
-\kl{$\Nat$-rational polynomial}, and if so, to compute an
-\kl{$\Nat$-polyregular function} that \kl{represents} it. 
-This particular problem was claimed to be solved in  \cite{KARH77}, 
-using a characterization restated in \cref{karh:thm}.
+polynomial}. The analogue notion of \intro{$\Rel$-rational polynomials} is not
+of particular interest, since every polynomial $P$ is \kl{represented} by a
+\kl{$\Rel$-polyregular function}. Our goal is to decide if a
+\kl{$\Rel$-rational polynomial} (so, any polynomial) is an \kl{$\Nat$-rational
+polynomial}, and if so, to compute an \kl{$\Nat$-polyregular function} that
+\kl{represents} it. This particular problem was claimed to be solved in
+\cite{KARH77}, using a characterization restated in \cref{karh:thm}.
 
@@
 \end{lemma}
+\begin{proof}
+    The polynomials of $\Nat[\vec{X}]$
+    are obtained from basic functions (constant in $\Nat$,
+    $X_i$ for some $1 \leq i \leq n$)
+    by products and sums. Because the basic functions are
+    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
+    that are closed under these operations, we conclude.
+\end{proof}
 
@@
     \label{karh:thm}
-    Let $\Sigma$ be a finite alphabet
-    and $P$ be a polynomial. Then,
+    Let $P \in \Rel[\vec{X}]$ be a polynomial. Then,
     $P$ is an \kl{$\Nat$-rational polynomial}
@@
 \end{proof}
+
+We will discuss at the end of \cref{sec:proof} why the example of
+\cref{thm:counter-example} is minimal in the number of indeterminates, which
+first requires us to provide a \emph{correct} analogue of \cref{karh:thm}.

### 2024-07-05 parts/intro.tex
starving the beast

@@
 counter-free automata, star-free regular languages and aperiodic monoids to the
-functional setting: 
-for Mealy Machines \cite{SCHU65,MNPA71},
-for sequential functions \cite{CHOF03}, 
-for rational functions \cite{FGL16,FGLM18}, for regular
-functions \cite{BOJA14,CADA15,DJR16,DGK21}, and for polyregular
+functional setting: for Mealy Machines \cite{SCHU65,MNPA71}, for sequential
+functions \cite{CHOF03}, for rational functions \cite{FGL16,FGLM18}, for
+regular functions \cite{BOJA14,CADA15,DJR16,DGK21}, and for polyregular
 functions \cite{BDK18,BOKL19}. The inclusions between these classes of
@@
 associated to these strict inclusions (that is, given a function $f$, decide if
-$f$ can be computed by a function of a proper subclass).
+$f$ can be computed by a function of a proper subclass). We also redirect the
+reader to the survey of Muscholl and Puppis for a more comprehensive
+introduction to the status of various membership problems between classes of
+string-to-string functions \cite{MUSC19}.
 
@@
 subclass of \intro{$\Nat$-rational series} ($\intro*\NRat$), another model of
-computation that aims at generalising the notion of regular language to
+computation that aims at generalizing the notion of regular language to
 functions with outputs in a semiring \cite{BERE88,BERE10}. It was frequently
@@
 restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
-was conjectured that, being ``computable using $\Nat$ coefficients'' and being
+was conjectured that, being ``computable using coefficients in $\Nat$'' and being
 ``star-free'' were two orthogonal properties, and therefore that the following
@@
 
-The approach taken in this paper avoids the use of tools from the theory of
-factorizations forests in finite monoids \cite{SIMO90}, which has been the
-traditional method used to understand (commutative output) \kl{polyregular
-functions} \cite{DOUE21,DOUE22,CDTL23}, and instead focuses on intrinsic
-description of the functions themselves, via  explicit encoding/decoding into
-polynomials (see for instance \cref{decompose-polynomial:lem}).
-
 \subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce

### 2024-07-05 parts/prelim.tex
starving the beast

@@
 these two output semigroups, and we refer the reader to \cite{BOKL19} for the
-general definition of \kl{polyregular functions}. As a model of computation, we
-will use the following definition of transducers, that is based on the marble
-transducers \cite{EHB99}, and follows the general pattern of transducers
-introduced in \cite{CDTL23}. Note that we only consider $\Rel$-valued (resp.
-$\Nat$-valued) functions, but could consider other commutative semigroups.
+general definition of \kl{polyregular functions}.
 
@@
 combinatorial in nature, and is similar in shape to the \emph{finite counting
-automata} introduced by \cite{SCHU62}.
+automata} introduced by \cite{SCHU62}. Note that we only consider $\Rel$-valued (resp.
+$\Nat$-valued) functions, but could consider other commutative semigroups.
 
-\begin{lemma}[$\Rel$-polyregular functions {\cite{CDTL23}}]
+\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
     \label{nat-rel-poly:def}
@@
     (resp. \kl{star-free $\Nat$-polyregular}).
-\end{lemma}
+\end{definition}
 
@@
 
-\begin{definition}
-    Let $\mathcal{H}$ be a family of functions
-    from $\Sigma^*$ to $\Rel$.
-    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
-    a tuple $(Q, q_0, \delta, \lambda, F)$ where
-         $Q$ is a finite set of states,
-         $q_0 \in Q$ is called the initial state of the transducer,
-         $\delta \colon Q \times \Sigma \to Q$
-            is called the deterministic transition function of the transducer,
-         $\lambda \colon Q \times \Sigma \to \mathcal{H}$
-            is called the correction function of the transducer,
-         and $F \colon Q \to \Rel$ is called the output function of the transducer.
-\end{definition}
-
-Based on this syntax, let us now define the semantics of a transducer, in terms
-of the function from $\Sigma^*$ to $\Rel$ that it computes. The intuition is
-that it suffices to apply the correction function $\lambda$ on every transition
-of the underlying automaton, sum the results, and add to this number the output
-associated to the final state of the computation.
-
-\begin{definition}
-    \label{transducer-sem:def}
-    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
-    The function
-    $\aTransd \colon Q \times \Sigma^* \to \Rel$
-    is defined inductively as follows:
-        $\aTransd(q, \varepsilon) \defined F(q)$, and
-        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
-            + \lambda(q,a)(u)$.
-    The function \intro{computed} by a transducer $\aTransd$
-    is $w \mapsto \aTransd(q_0, w)$.
-\end{definition}
-
-\begin{example}
-    \label{simple-transd:ex}
-    Let $\mathsf{Const}$ be the set of constant functions from $\Sigma^*$ to $\Rel$,
-    and let us write $\const{n}$ for the constant function equal to $n \in \Rel$.
-    In 
-    \cref{simple-transd:fig}, we provide a graphical depiction of a
-    \kl{$\mathsf{Const}$-transducer} that \kl{computes} the function
-    $f \colon \set{a,b}^* \to \Rel$ defined by
-    $f(\varepsilon) = 1$,
-    $f(aw) = 3 \times \card{w}$,
-    and $f(bw) = 3 \times \card{w} + 1$.
-\end{example}
-
-\begin{figure}
-    \centering
-    \begin{tikzpicture}[
-        etat/.style={minimum size=2em}
-        ]
-        \node[etat,state,initial,
-            accepting by arrow,
-            accepting text={$1$},
-            accepting where=below,
-            ] (A) at (0,0) {$q_0$};
-        \node[etat,state,
-            accepting by arrow,
-            accepting text={$0$},
-            accepting where=below,
-            ] (B) at (2.5,0) {$q_1$};
-
-        \draw[->] (A) to[bend right=20] node[midway, below] {$a \mid \const{0}$} (B);
-        \draw[->] (A) to[bend left=20] node[midway, above] {$b \mid \const{1}$} (B);
-
-        \draw[->] (B) edge[loop right] node[midway, right]
-            {$a,b \mid \const{3}$} (B);
-    \end{tikzpicture}
-    \caption{A \kl{$\mathsf{Const}$-transducer}, where $\mathsf{Const}$ is the set of constant functions
-        from $\set{a,b}^*$ to $\Rel$,
-        with a set of states $Q \defined \set{ q_0, q_1 }$,
-        an initial state $q_0$,
-        an output function $F \colon Q \to \Rel$ such that $F(q_0) = 1$ and $F(q_1) = 0$,
-        and correction functions $\lambda(q,a) = \const{0}$, $\lambda(q,b) = \const{1}$,
-        and $\lambda(q_1,a) = \lambda(q_1,b) = \const{3}$.
-    }
-    \label{simple-transd:fig}
-\end{figure}
-
-\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
-words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
-\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$;
-where we used the extension of $\delta \colon Q \times \Sigma \to Q$ to words
-by defining $\delta(q, \varepsilon) = q$, and $\delta(q,au) =
-\delta(\delta(q,a), u)$. An automaton is \reintro{counter-free} if it contains
-no \reintro{counters}. This notion can be lifted to
-\kl{$\mathcal{H}$-transducers} by considering counters of their underlying
-automata.
-
-\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
-define the classes of
-\intro{$\Rel $-polyregular functions}
-$\intro*\ZPoly[k]$,
-\intro{$\Nat $-polyregular functions}
-$\intro*\NPoly[k]$,
-\intro{star-free $\Rel $-polyregular functions}
-$\intro*\ZSF[k]$, and
-\intro{star-free $\Nat $-polyregular functions}
-$\intro*\NSF[k]$, 
-as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
-case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
-\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
-$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
-\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
-$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
-\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
-function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
-where the transducer is furthermore required to be \kl{counter-free}. Then,
-$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
-$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
-In the rest of the paper, we will always assume that
-\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
-have an $\Nat$-valued output functions, making a clean separation between
-$\NPoly$ and $\ZPoly$.
-
-\begin{example}
-    \label{non-canonical-transd:ex}
-    Let $\BadExOk \colon \set{a}^* \to \Nat$
-    be defined as
-    $\BadExOk(\varepsilon) = 1$,
-    $\BadExOk(a) = 0$,
-    and $\BadExOk(a^n) = n-1$ for all $n \geq 2$.
-    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
-    depicted in \cref{non-canonical-transd:fig}.
-    As a consequence, $f \in \NPoly[1]$.
-    Furthermore, because the automaton of 
-        \cref{non-canonical-transd:fig:counter-free}
-        has no \kl{counter},
-    and the correction function only uses functions in $\NSF[0]$,
-    $f \in \NSF[1]$.
-    \proofref{non-canonical-transd:ex}
-\end{example}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.49\linewidth}
-        \begin{tikzpicture}[
-            etat/.style={minimum size=2em}
-            ]
-            \node[etat,state,initial,
-                accepting by arrow,
-                accepting text={$1$},
-                accepting where=below,
-                ] (A) at (0,0) {$\varepsilon$};
-            \node[etat,state,
-                accepting by arrow,
-                accepting text={$0$},
-                accepting where=below,
-                ] (B) at (2,0) {$a$};
-
-            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
-
-            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 - 2 \times \ind{\card{w} = 0}$} (A);
-        \end{tikzpicture}
-        \caption{This automaton has a \kl{counter}.}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.49\linewidth}
-        \begin{tikzpicture}[
-            etat/.style={minimum size=2em}
-            ]
-            \node[etat,state,initial,
-                accepting by arrow,
-                accepting text={$1$},
-                accepting where=below,
-                ] (A) at (0,0) {$\varepsilon$};
-            \node[etat,state,
-                accepting by arrow,
-                accepting text={$0$},
-                accepting where=below,
-                ] (B) at (2,0) {$a$};
-
-            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);
-
-            \draw[->] (B) edge[loop right] node[midway, right]
-                {$a \mid 1$} (B);
-        \end{tikzpicture}
-        \caption{This automaton is \kl{counter-free}.}
-        \label{non-canonical-transd:fig:counter-free}
-    \end{subfigure}
-    \caption{Two \kl{$\NPoly[0]$-transducers}
-    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex},
-    that maps $a^n$ to $n$ for all $n \geq 2$, $a$ to $0$, and $\varepsilon$ to $1$.
-    }
-    \label{non-canonical-transd:fig}
-\end{figure}
-
 \AP One of the appeals of $\NPoly$ and $\ZPoly$ are the numerous
@@
 (resp. in $\NRat)$ that have growth rate at most $k$ \cite{CDTL23}.
+
+\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
+define the classes of \intro{$\Rel $-polyregular functions} $\intro*\ZPoly[k]$,
+\intro{$\Nat $-polyregular functions} $\intro*\NPoly[k]$, \intro{star-free
+$\Rel $-polyregular functions} $\intro*\ZSF[k]$, and \intro{star-free $\Nat
+$-polyregular functions} $\intro*\NSF[k]$, as follows \cite[Theorems 5.15 and
+7.10]{DOUE23}: in the base case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  =
+\reintro*\ZSF[-1] = \reintro*\ZPoly[-1]$ is the singleton containing the
+constant function equal to $0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the
+set of functions \kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
+$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
+\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
+function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
+where the transducer is furthermore required to be \kl{counter-free}. Then,
+$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
+$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
+In the rest of the paper, we will always assume that
+\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
+have an $\Nat$-valued output functions, making a clean separation between
+$\NPoly$ and $\ZPoly$. that \kl{represents} $P$.
+
+Because $f$ has \kl{polynomial
+growth}, it is in fact a \kl{$\Nat$-polyregular function} (\cite[Theorem
+5.22]{DOUE23}).
+\begin{lemma}[CITE]
+    \label{polyregular-polynomial-growth:lemma}
+    Let $f \in \ZRat$. The following are equivalent:
+    \begin{enumerate}
+        \item $f \in \ZPoly[k]$.
+        \item $f$ has polynomial growth of degree at most $k$.
+    \end{enumerate}
+    And similarly for $\NPoly$ and $\NRat$.
+\end{lemma}
+
+Let us state the crucial compositional properties of
+\kl{$\Rel$-polyregular functions} that will be used in this paper
+to construct \kl{$\Rel$-polyregular functions}.
+\begin{lemma}[CITE]
+    \label{stability-polyregular:lemma}
+    The class $\NPoly[k]$ (resp. $\ZPoly[k]$) is stable under the operations of
+    addition, multiplication, Cauchy product,
+    and ``value pinning''.
+\end{lemma}
 

### 2024-07-05 parts/starfree.tex
starving the beast

@@
     There exists a computable
-    $\omega \in \Nat$,
+    $\omega \in \Nat_{\geq 1}$,
     and computable 
     polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
-    such that for all $w \in \Sigma^*$,
+    such that for all $x_1, \dots, x_n \geq \omega$:
     \begin{equation*}
-        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
-        = P_{\moduloType[\omega](X_1, \dots, X_n)}
+        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
+        = P_{\moduloType[\omega](x_1, \dots, x_n)}
         \left(
-            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
+            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
         \right)
@@
 \end{proof}
+
+\begin{conjecture}
+    \label{npoly-zpoly:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \ZPoly[k]$
+    belongs to $\NPoly[k]$ if and only if
+    it is \kl{$(k,\Nat)$-combinatorial}.
+    In particular,
+    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
+\end{conjecture}
 
@@
 \begin{proof}
+    Let $f$ be a \kl{commutative} \kl{$\Nat$-polyregular function}. If $f \in \NSF \subseteq \ZSF$
+    then $f$ is \kl{ultimately polynomial} by \cite{CDTL23}.
+
+    % FIXME:
+    % -
+
     We proceed by induction on the size of the alphabet $\card{\Sigma}$.

### 2024-07-05 parts/conclusion.tex
starving the beast

@@
 \kl{star-free polyregular function}, as formally stated in
-\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}. Following
-this idea, we conjecture that $\ZPoly$ and $\NPoly$ can be described as
-combinations of commutative functions, leading to
-\cref{z-poly-commutative-encoding:conj}, which would be a non-commutative
-version of \cref{decompose-polynomial:lem}. For the subclass of 
-\emph{polyblind functions} \cite{LENP21,DOUE22}, parts of the conjecture holds,
-because they essentially behave as commutative functions \cite[Theorem
-6.12]{DOUE23}.
+\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
 
@@
 \end{lemma}
+
+Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
+functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
+immediately, and $\NSF$ inside $\NPoly$ would immediately follow. On the one
+hand, one can guess a candidate function and check for equivalence, on the
+other hand, one can guess a \kl{commutative} \kl{star-free polyregular
+function} and check membership (which is decidable thanks to this paper).

### 2024-07-05 parts/starfree.tex
starving the beast

--- 2024-07-04 parts/starfree.tex
@@
 \AP To actually state the lemma, we will need some notations that will
-formalize the intuitive idea of counting tuples of integers modulo some
-constant. Let $\omega \in \Nat$, we define the set $\intro*\ModuloTypes[\omega]
-\defined \set{0, \dots, \omega^2} \times \set{0, \dots, \omega - 1}$ of
-\intro{types modulo $\omega$}. To a number $n \in \Nat$, we associate its
-\reintro{$\omega$-type} written $\intro*\moduloType[\omega](n)$ which is
-defined as the pair $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the  remainder
-of the Euclidean division of $n$ by $\omega$. This notion of type is lifted to
-vectors in $\Nat^p$ for any $p \in \Nat$ by pointwise application.
+formalize the intuitive idea of identifying tuples of integers modulo some
+constant. Let $\omega \in \Nat_{\geq 1}$, we define the set
+$\intro*\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
+\dots, \omega - 1}$ of \intro{types modulo $\omega$}. To a number $n \in \Nat$,
+we associate its \reintro{$\omega$-type} written
+$\intro*\moduloType[\omega](n)$ which is defined as the pair $(t, r)$ where $t
+= \min (n, \omega^2)$ and $r$ is the  remainder of the Euclidean division of
+$n$ by $\omega$. This notion of type is lifted to vectors in $\Nat^p$ for any
+$p \in \Nat$ by pointwise application.
 
@@
     \begin{enumerate}
-        \item $f$ is \kl{ultimately polynomial},
-        \item $f \in \ZSF$,
-        \item $f \in \NSF$.
+        \item \label{zsf-npoly-nsf:thm:up:item} $f$ is \kl{ultimately polynomial},
+        \item \label{zsf-npoly-nsf:thm:zsf:item} $f \in \ZSF$,
+        \item \label{zsf-npoly-nsf:thm:nsf:item} $f \in \NSF$.
     \end{enumerate}
@@
 \end{theorem}
+\begin{proof}
+    The implication \cref{zsf-npoly-nsf:thm:zsf:item} $\Rightarrow$
+    \cref{zsf-npoly-nsf:thm:zsf:item} is immediate since $\NSF \subseteq \ZSF$.
+    Furthermore,
+    \cref{zsf-npoly-nsf:thm:zsf:item} implies \cref{zsf-npoly-nsf:thm:up:item}
+    following previous results for \kl{star-free $\Rel$-polyregular functions}
+    \cite{CDTL23}.
 
-\textbf{TODO: prove the following conjecture to conclude on polynomials in general
-and not just Z coefficients.}
-\begin{conjecture}
-    Let $\vec{X}$ be a vector of variables, and $P \in \Rel[\vec{X}]$ be a
-    polynomial. The following properties are decidable
-    \begin{itemize}
-        \item Is $P$ \kl{represented} by a \kl{$\Rel$-polyregular function}?
-        \item Is $P$ \kl{represented} by a \kl{$\Nat$-polyregular function}?
-    \end{itemize}
-\end{conjecture}
+    For the implication \cref{zsf-npoly-nsf:thm:up:item} $\Rightarrow$
+    \cref{zsf-npoly-nsf:thm:nsf:item}, let us assume that $f$ is \kl{ultimately
+    polynomial}. We prove the result by induction on the size of the alphabet
+    $\Sigma$. By definition, there exists $N_0 \in \Nat$, and $P \in
+    \Rat[\seqof{X_a}{a \in \Sigma}]$ such that: \begin{equation*} f\left(
+        \prod_{a \in \Sigma} a^{X_a} \right) = P(\seqof{X_a}{a \in \Sigma})
+        \quad \text{ when } \forall a \in \Sigma, X_a \geq N_0 \quad .
+    \end{equation*} It is clear that $P$ is \kl{represented} by a
+    \kl{$\Nat$-polyregular function}, namely, $f$ partially applied to the word
+    $\prod_{a \in \Sigma} a^{N_0}$, and therefore is in fact represented by a
+    \kl{star-free $\Nat$-polyregular function} thanks to
+    \cref{decide-rat-poly-npoly:cor}.
+    For every letter $a \in \Sigma$ and number $0 \leq n \leq N_0$, there exists, by induction hypothesis,
+    a \kl{star-free $\Nat$-polyregular function} $g_a^n$ that represents the
+    function $f_{a^n} \colon (\Sigma \setminus \set{a})^* \to \Rel$
+    that maps $u \in \Sigma \setminus \set{a}$ to $f(a^n u)$.
+
+    Let us now define the function $g \colon \Sigma^* \to \Rel$ as follows:
+    \begin{equation*}
+        g \colon w \mapsto \begin{cases}
+            g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
+            P(\seqof{\card[a]{w}}{a \in \Sigma}) & \text{ otherwise }
+        \end{cases}
+    \end{equation*}
+    We conclude that $g$ is a \kl{star-free $\Nat$-polyregular function}, and by definition
+    $f = g$.
+\end{proof}

### 2024-07-17 parts/prelim.tex
update citations

@@
     \label{nat-rel-poly:def}
-    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
-    be a morphism, $k \in \Nat$, and 
-    $\pi \colon M^k \to \Rel$ be a function.
-    The following function 
-    $\pi^\dagger \colon \Sigma^* \to \Rel$
-    is \kl{$\Rel$-polyregular}:
+    Let $k \in \Nat$. The set $\intro*\ZPoly[k]$ of polyregular
+    \intro{$\Rel$-polyregular functions} of degree at most $k$,
+    is the set of functions $f \colon \Sigma^* \to \Rel$
+    such that
+    there exists a finite monoid $M$,
+    a morphism $\mu \colon \Sigma^* \to M$,
+    and a function $\pi \colon M^{k+1} \to \Rel$
+    satisfying:
     \begin{equation*}
-        \pi^\dagger (w) \defined
-        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
+        f \defined
+        \sum_{w = u_1 \cdots u_{k+1}} \pi(\mu(u_1), \dots, \mu(u_{k+1}))
         \quad .
     \end{equation*}
-    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
-    is furthermore \kl{$\Nat$-polyregular}.
-    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
+    We call $\pi$ the \emph{production function} of $f$.
+    If the function $\pi$ has codomain $\Nat$,
+    then $f$ is \intro{$\Nat$-polyregular} of degree at most $k$,
+    i.e., $f \in \intro*\NPoly[k]$.
     If the monoid $M$ is \kl(monoid){aperiodic}
@@
     the function $f$ is \kl{star-free $\Rel$-polyregular}
-    (resp. \kl{star-free $\Nat$-polyregular}).
+    ($\intro*\ZSF[k]$), resp. \kl{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
 \end{definition}
 
-In order to illustrate the above definition, let us provide an example of a
-\kl{$\Nat$-polyregular function} computed using a finite monoid.
+\AP We complete definition \cref{nat-rel-poly:def} by defining $\intro*\NPoly =
+\bigcup_{k \in \Nat} \intro*\NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
+$\ZSF$. In order to illustrate these definitions, let us provide an example of
+a \kl{$\Nat$-polyregular function} computed using a finite monoid.
 
@@
     The map $f \colon w \mapsto \card{w} + 1$
-    belongs to $\NSF$.
+    belongs to $\NSF[1]$.
 \end{example}
@@
 \in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
-(resp. in $\NRat)$ that have growth rate at most $k$ \cite{CDTL23}.
+(resp. in $\NRat)$ that have growth rate at most $k$.
 
-\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
-define the classes of \intro{$\Rel $-polyregular functions} $\intro*\ZPoly[k]$,
-\intro{$\Nat $-polyregular functions} $\intro*\NPoly[k]$, \intro{star-free
-$\Rel $-polyregular functions} $\intro*\ZSF[k]$, and \intro{star-free $\Nat
-$-polyregular functions} $\intro*\NSF[k]$, as follows \cite[Theorems 5.15 and
-7.10]{DOUE23}: in the base case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  =
-\reintro*\ZSF[-1] = \reintro*\ZPoly[-1]$ is the singleton containing the
-constant function equal to $0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the
-set of functions \kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
-$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
-\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
-function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
-where the transducer is furthermore required to be \kl{counter-free}. Then,
-$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
-$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
-In the rest of the paper, we will always assume that
-\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
-have an $\Nat$-valued output functions, making a clean separation between
-$\NPoly$ and $\ZPoly$. that \kl{represents} $P$.
-
-Because $f$ has \kl{polynomial
-growth}, it is in fact a \kl{$\Nat$-polyregular function} (\cite[Theorem
-5.22]{DOUE23}).
-\begin{lemma}[CITE]
+\begin{lemma}[{\cite[Theorem 5.22]{DOUE23}}]
     \label{polyregular-polynomial-growth:lemma}
@@
 to construct \kl{$\Rel$-polyregular functions}.
-\begin{lemma}[CITE]
+\begin{lemma}[{\cite[Theorem II.20]{CDTL23}}]
     \label{stability-polyregular:lemma}
@@
     of rational series is decidable 
-    \cite[\textbf{WHERE}]{BERE10}
+    \cite[Corollary 3.6]{BERE10},
     we have obtained a decision procedure.

### 2024-08-12 parts/appendix.tex
removing useless appendix parts

@@
 \section{Proofs of section \ref{preliminaries:sec}}
-
-\begin{proofof}[non-canonical-transd:ex]
-    Let us prove that the transducer of
-    \cref{non-canonical-transd:fig:counter-free}
-    computes $f$.
-    We prove by induction on $w$
-    that $A(q_0, w) = f(w)$,
-    and that $A(q_1, w) = f(aw)$.
-    To that end, let us first remark that $f(\varepsilon) = 1$,
-    and $f(au) = \card{u}$.
-
-    When $w = \varepsilon$, $A(q_0, \varepsilon) = F(\varepsilon) = 1 = f(\varepsilon)$.
-    Similarly, $A(q_1, \varepsilon) = F(q_1) = 0 = f(a)$.
-
-    Assume that $w = au$. Then:
-    \begin{align*}
-        A(q_0, w) = A(q_0, au) &= \lambda(q_0, a)(u) + A(q_1, u) \\ 
-                               &= 0 + A(q_1,u) \\
-                               &= f(au) & \text{by induction hypothesis} \\
-                               &= f(w) 
-    \end{align*}
-    Similarly,
-    \begin{align*}
-        A(q_1, w) = A(q_1, au) &= \lambda(q_1, a)(u) + A(q_1, u) \\ 
-                               &= 1 + A(q_1,u) \\
-                               &= 1 + f(au) & \text{by induction hypothesis} \\
-                               &= 1 + \card{u} \\
-                               &= \card{au} \\
-                               &= f(aau) \\
-    \end{align*}
-
-    The proof for the other automaton is similar. The
-    key ingredient in the induction hypothesis is that if $\card{u} \geq 1$, then:
-    $f(au) - f(u) = 1$ and otherwise $f(a) - f(\varepsilon) = -1$.
-    Hence, $f(au) - f(u) = 1 - 2 \times \ind{\card{u} = 0} = \lambda(q_1, a)(u)$.
-\end{proofof}
 
@@
 
-\section{Proofs of section \ref{beyond-commutative:sec}}
-
-\begin{remark}
-    \label{good-residual-ordering:fact}
-    Let $k \in \Nat$, and let $f \colon \Sigma^* \to \Nat$. Then,
-    $(\resleq{f}{k})$ is a quasi-ordering, satisfying the following
-    extra properties:
-    \begin{enumerate}
-        \item For all $u,v,w \in \Sigma^*$, $u \resleq{f}{k} v$
-            implies $uw \resleq{f}{k} vw$,
-        \item If $u \resleq{f}{k} v$ and $v \resleq{f}{k} u$,
-            then $\app{f}{u} = \app{f}{v}$.
-    \end{enumerate}
-\end{remark}
-
-\begin{fact}
-    \label{unique-res-transducer:fact}
-    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
-    Then $f$ has at most one \kl{$k$-residual transducer}.
-\end{fact}
-\begin{proof}
-    Let $\aTransd_1$ and $\aTransd_2$ be two
-    \kl{$k$-residual transducers} for $f$.
-    The two initial states must be $\varepsilon$.
-    Let us prove by induction on $u \in \Sigma^*$ that
-    $\delta_1(\varepsilon, u) = \delta_2(\varepsilon, u)$
-    and that $Q_1$ equals $Q_2$ over prefixes of $u$.
-    This will prove that 
-    $Q_1 = Q_2$, hence that $\aTransd_1 = \aTransd_2$.
-
-    Let $u \in \Sigma^* \cap Q_1 \cap Q_2$ and $a \in \Sigma$, $v_1 \in Q_1$ be
-    defined as $v \defined \delta_1(u,a)$, and $v_2 \defined \delta_2(u,a)$.
-    Remark that by induction hypothesis, for all $v \prefle u$, $v \in Q_1 \cap
-    Q_2$. If $\delta_1(u,a) = Q_1$, it means that for all $v \in Q_2$ such that
-    $v \prefle ua$, we have $\neg( v \resleq{f}{k} ua )$. The only possible
-    transition in $\aTransd_2$ is therefore $\delta_2(u,a) = ua$, and $ua \in
-    Q_2$. Similarly, if $\delta_1(u,a) \prefleq u$, then $\delta_2(u,a) =
-    \delta_1(u,a)$ by definition of $\delta_2$ as a maximum.
-\end{proof}
-
-\begin{fact}
-    \label{q-o-prefix-cool:fact}
-    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
-    At each step of the \texttt{while loop}
-    of \cref{residual:algo}, the sets
-    $Q$ and $O$ are such that
-    \begin{enumerate}
-        \item $Q \cup O$ is a \kl{downwards closed} subset of 
-            $\Sigma^*$ for $\prefleq$;
-        \item elements in $O$ are pairwise incomparable
-            for $\prefleq$, and are maximal
-            for $\prefleq$ inside $Q \cup O$.
-    \end{enumerate}
-\end{fact}
-\begin{proof}
-    Let us write $Q_i$ and $O_i$ for the value of the variables
-    $Q$ and $O$ at step $i$ of the \texttt{while loop}.
-    We prove the desired property by induction on $i$.
-
-    For $i=0$, the property is true because
-    $Q_0 = \set{\varepsilon}$ and $O_0 = \setof{a}{a \in \Sigma}$.
-
-    For $i+1$. Either the \texttt{if} branch was taken, in which case $Q_{i+1}
-    \cup O_{i+1} = (Q_i \cup O_i) \setminus \set{u}$ for some $u \in O_i$. This
-    set remains \kl{downwards closed}, and elements in $O_{i+1}$ remain maximal
-    elements.
-
-    If the \texttt{else} branch was taken, then there exists $u \in O_i$ such
-    that $Q_{i+1} = Q_i \cup \set{u}$ and $O_{i+1} = O_i \setminus \set{ u }
-    \cup \setof{ ua }{ a \in \Sigma}$. We conclude that $Q_{i+1} \cup O_{i+1} =
-    Q_i \cup O_i \cup \setof{ ua }{a \in \Sigma}$ continues to be \kl{downwards
-    closed} for $\prefleq$. Let $v \in Q_{i+1} \cup O_{i+1}$ be such that $ua
-    \prefleq v$ for some $a \in \Sigma$. Then $u \prefleq v$, and $u = v$ since
-    $u$ was a maximal element. As a consequence, $ua$ is a maximal element for
-    all $a \in \Sigma$. Assume by contradiction that $ua$ is comparable with
-    some $v \in O_{i+1}$ with $ua \neq v$, it cannot be that $ua \prefleq v$ by
-    the above argument, and if $v \prefleq ua$ with $v \neq ua$, then $v
-    \prefleq u$ and $u = v$, which is absurd since $v \not \in O{i+1}$.
-    We have concluded that $O_{i+1}$ continues to have pairwise incomparable
-    elements.
-\end{proof}
-
-\begin{lemma}
-    \label{correct-residual:lemma}
-    If \cref{residual:algo} terminates on 
-    an input $f \colon \Sigma^* \to \Nat$, $k \in \Nat$,
-    then it computes the \kl{$k$-residual transducer} of $f$.
-    \proofref{correct-residual:lemma}
-\end{lemma}
-\begin{proof}
-    Because of \cref{q-o-prefix-cool:fact},
-    we already know that $q_0 = \varepsilon$,
-    $Q$ is a \kl{downwards closed} subset of $\Sigma^*$
-    for $\prefleq$, 
-    that every state of $Q$ is accessible from $q_0$.
-    Notice that at every step,
-    $\lambda(u,a)$ is defined as
-    $\app{f}{ua} - \app{f}{\delta(u,a)}$.
-    Finally, since $Q \cup O$ is a \kl{downwards closed} subset of $\Sigma^*$
-    at every step,
-    we have that at step $i$,
-    for all $ua \in O_i$,
-    $\setof{w \in Q}{w \prefleq ua} = \setof{w \in Q_i}{ w \prefleq ua}$,
-    which proves that the maximum considered in the algorithm
-    is indeed computing correctly.
-\end{proof}
-
-\begin{lemma}
-    \label{wqo-implies-termination:lemma}
-    Let $f \colon \Sigma^* \to \Nat$, and $k \in \Nat$ be such that
-    every infinite, $\prefleq$-increasing sequence is \kl{good}
-    in $(\Sigma^*, \resleq{f}{k})$
-    (or equivalently, such that the relation $(\prefleq \Rightarrow \resleq{f}{k})$
-    is \kl{well}).
-    Then, \cref{residual:algo} terminates on the input $(f,k)$.
-    \proofref{wqo-implies-termination:lemma}
-\end{lemma}
-\begin{proof}
-    Assume towards a contradiction that
-    \cref{residual:algo} does not terminate.
-    Then, the \texttt{else} branch in the \texttt{while loop}
-    must be taken infinitely often.
-    This means that the set $Q$ of states grows arbitrarily large.
-
-    Let us write $\seqof{Q_i}{i \in \Nat}$ for the set of states $Q$ at step
-    $i$ of the execution of \cref{residual:algo}. Applying
-    \cref{q-o-prefix-cool:fact}, we know that for all $i \in \Nat$, $Q_i$ is
-    \kl{downwards closed} for $\prefleq$. Let us write $Q_\infty \defined
-    \bigcup_{i \in \Nat} Q_i$. The set $Q_\infty$ is infinite, and is
-    \kl{downwards closed} for $\prefleq$. As a consequence, it is an infinite
-    tree with a finite branching (at most $\card{\Sigma}$), and has an infinite
-    branch $\seqof{u_j}{j \in \Nat}$ thanks to König' s lemma.
-
-    Let us prove that this infinite branch is a \kl{bad sequence} for the
-    ordering $\resleq{f}{k}$.
-    Let $j < p$, and assume by contradiction that $u_j \resleq{f}{k} u_p$. We
-    know that $u_j \in Q_j$ and $u_p \in Q_p$. Then, at step $p-1$ of the
-    algorithm, $u_j \in Q_{p-1}$, since $u_j \in Q_j \subseteq Q_{p-1}$.
-    Because $u_j \prefleq u_p$ and $u_j \resleq{f}{k} u_p$,
-    \cref{residual:algo} must take the \texttt{if} branch at step $p-1$. As a
-    consequence, $u_p \not\in Q_{p}$, which is absurd.
-
-    We have proven that the infinite branch is a \kl{bad sequence}
-    for $\resleq{f}{k}$, which contradicts the assumption.
-    Hence, \cref{residual:algo} must terminate.
-\end{proof}
-
-\begin{lemma}
-    \label{n-poly-k-implies-wqo:lemma}
-    Let $k \in \Nat$, and let $f \in \NPoly[k]$.
-    Then, $(\Sigma^* \resleq{f}{k})$ is a \kl{well-quasi-ordering}.
-\end{lemma}
-\begin{proof}
-    Because $f \in \NPoly[k]$, there exists
-    a tuple $\vec{x}$ of first order free variables,
-    $\MSO$ formulas $\seqof{\psi_i(\vec{x})}{1 \leq i \leq n}$,
-    and positive coefficients $\seqof{m_i}{1 \leq i \leq n}$,
-    such that
-    $f = \sum_{i = 1}^n m_i \times \vcount{\varphi_i(\vec{x})}$
-    \cite[Theorem 5.15]{DOUE23}.
-
-    Let $q$ be the maximal quantifier rank of formulas $\seqof{\psi_i}{1 \leq i
-    \leq n}$. To a word $u \in \Sigma^*$, we associate the vector $\MSO^q(u)$
-    that maps an $\MSO$-type with $\ell \leq |\vec{x}|$ free variables to the
-    number of realizations of this type in $u$.
-
-    Let $u, v \in \Sigma^*$ such that $\MSO^q(u) \leq \MSO^q(v)$, which means
-    that every $\MSO$ type (of quantifier rank $q$ and with at most $n$ free
-    variables) has at least as many realizations in $v$ than it has in $u$.
-    Remark that by the compositionality of $\MSO$ over words (for instance, see
-    the Feferman-Vaught theorem \cite{FEVAU59,MAKOW04}), for all $\MSO^q$ types
-    $t(\vec{x})$, there are finitely many $\MSO^q$ types $t_l^j(\vec{y_i}),
-    t_r^j(\vec{z_i})$ with $\vec{x} = \vec{y_i} \uplus \vec{z_i}$
-    for $1 \leq j \leq N_0$, such that for every
-    tuple $\vec{a}$ of elements in a word $uv$, $\MSO^q(\vec{a} / uv) =
-    t(\vec{x})$ if and only if there exists $1 \leq j \leq N_0$,
-    such that $\vec{a} = \vec{b} \uplus \vec{c}$,
-    $\MSO^q(\vec{b} / u) =
-    t_l^j(\vec{y_i})$, and $\MSO^q(\vec{c} / v) = t_r^j(\vec{z_i})$.
-    We write $t = t_l \odot t_r$ to signify
-    that $\MSO^q(\vec{bc} / uv) = t$
-    if and only if $\MSO^q(\vec{b}/u) = t_l$
-    and $\MSO^q(\vec{c}/v) = t_r$.
-
-    As a consequence, if $\MSO^q(u) \leq \MSO^q(v)$, then 
-    for all $w \in \Sigma^*$:
-    \begin{align*}
-        & f(vw) - f(uw) \\
-        &= 
-        \sum_{i = 1}^n m_i
-        \left[
-            \vcount{\phi_i(\vec{x})} (vw) -
-            \vcount{\phi_i(\vec{x})} (uw)
-        \right] \\
-        &= 
-        \sum_{i = 1}^n
-        m_i
-            \sum_{\phi_i \in t(\vec{x})}
-        \left[
-            \vcount{t(\vec{x})}(vw)
-            -
-            \vcount{t(\vec{x})}(uw)
-        \right] \\
-        &= 
-        \sum_{i = 1}^n
-        m_i
-        \sum_{1 \leq j \leq N_0}
-        \sum_{\phi_i \in t_l^j(\vec{y}) \odot t_r^j(\vec{z})}
-        \underbrace{
-        \left[
-            \vcount{t_r^j(\vec{y})}(v)
-            -
-            \vcount{t_r^j(\vec{y})}(u)
-        \right] 
-    }_{ \in \Nat }
-            \times 
-            \vcount{t_l^j(\vec{z})}(w)
-    \end{align*}
-
-    We have proven that if $\MSO^q(u) \leq \MSO^q(v)$, then $u \resleq{f}{k}
-    v$. Recall that $\Nat^p$ is a \kl{well-quasi-ordering} when endowed with
-    the product ordering, and therefore that $\setof{\MSO^q(u)}{u \in
-    \Sigma^*}$ is a \kl{well-quasi-ordering}.
-
-    Let $\seqof{u_i}{i \in \Nat}$ be an infinite sequence of $\Sigma^*$.
-    Without loss of generality, one can assume that for all $i \neq j$, $u_i
-    \equiv_k u_j$, i.e., that the difference $\app{f}{u_i} - \app{f}{u_j}$
-    belongs to $\ZPoly[k-1]$, since the latter has finite index. Thanks to the
-    above remarks, there exists $i < j$ such that $\MSO^q(u_i) \leq
-    \MSO^q(u_j)$. As a consequence, $g \defined \app{f}{u_j} - \app{f}{u_i} \in
-    \NPoly$, and therefore $g \in \NPoly[k-1]$. We have proven that there
-    exists $i < j$ such that $u_i \resleq{f}{k} u_j$.
-\end{proof}
-
-\begin{proofof}[non-commutative-npoly:thm]
-    \cref{n-poly-1-transd:item} implies \cref{n-poly-k:item} by
-    definition. Then,
-    \cref{n-poly-k:item} implies \cref{n-poly-wqo:item} by
-    \cref{n-poly-k-implies-wqo:lemma}.
-    The implication \cref{n-poly-wqo:item} $\Rightarrow$ \cref{n-poly-well:item}
-    is obvious.
-    Then, \cref{wqo-implies-termination:lemma} proves
-    that \cref{n-poly-well:item} implies \cref{n-poly-residual:item}.
-    Finally, because a \kl{$k$-residual transducer} is a \kl{$\NPoly[k-1]$-transducer},
-    \cref{n-poly-residual:item} implies \cref{n-poly-1-transd:item}.
-\end{proofof}
-
-\begin{proofof}[aperiodic-iff-residual:lem]
-    It is clear that $\NSF \subseteq \ZSF$, and known that if $f \in \ZSF$
-    then it is \kl{ultimately polynomial}. Furthermore, if the \kl{$0$-residual
-    transducer} of $f$ is \kl{counter-free}, then $f \in \NSF$
-    by definition of $\NSF$.
-
-    Assume that $f$ is \kl{ultimately polynomial}, let us prove that the
-    \kl{$0$-residual transducer} of $f$ is \kl{counter-free}. 
-    Note that because $f \in \NPoly[0]$,
-    $u \resleq{f}{0} v$ if and only if $\app{f}{u} = \app{f}{v}$.
-    In particular, in a \kl{$0$-residual transducer} of $f$,  two states that
-    represent the same residual must be incomparable for the prefix relation.
-
-    Let $(q,w^n)$ be
-    a counter with $n \geq 1$. This means that $\delta(q, w^n) = q$ in the
-    automaton, and implies that $q \resleq{f}{0} qw^n$, hence that $\app{f}{q}
-    = \app{f}{qw^n}$. Because $f$ is \kl{ultimately polynomial},
-    we know that $\app{f}{qw^n} = \app{f}{qw^{n+1}}$,
-    hence that $\app{f}{qw} = \app{f}{q}$.
-
-    Let us write $t \defined \delta(q,w) = \delta(q,w^{n+1})$. We know that
-    $\app{f}{q} = \app{f}{t}$. Assume by contradiction that $t$ and $q$ are
-    incomparable for the prefix relation. Let us split $w = w_1 w_2$ where
-    $w_1$ is the shortest prefix of $w$ such that $s_0 \defined \delta(q,w_1)$
-    is an ancestor of $q$ and of $t$ for the prefix relation, it must exist
-    because $\delta(q,w_1 w_2) = t$.
-
-    Now, consider $s_1 \defined \delta(t, w_1)$. Assume by contradiction that
-    $s_0$ is not comparable with $s_1$ for the prefix relation. Then, consider
-    the smallest prefix $v$ of $w_1$ such that $\delta(t, v)$ is a strict
-    prefix of $s_0$. It must exist, otherwise $s_0$ is always a prefix of
-    $s_1$. Because $\app{f}{t} = \app{f}{q}$, we conclude that $\app{f}{tv} =
-    \app{f}{qv}$. However, this contradicts the minimality of $w_1$, since
-    $\delta(t,v)$ is an ancestor of $q$ and $t$.
-
-    We have proven that $s_0$ and $s_1$ are comparable, hence they are equal,
-    since $\app{f}{s_1} = \app{f}{t w_1} = \app{f}{q w_1} = \app{f}{s_0}$.
-    Finally, we have proven that $\delta(q, w_1) = s_0$, $\delta(s_0, w_2) =
-    t$, and $\delta(s_0, w_2) = \delta(t, w_1w_2) = q$ which is absurd.
-    As a consequence $t$ and $q$ were comparable for the prefix relation,
-    hence equal, and therefore $\delta(q, w) = q$.
-\end{proofof}
-
-\begin{proofof}[sf-no-periods-on-sequences:lemma]
-    Let $f \in \NSF[k]$, $q \in \Nat$, and 
-    write $f = \sum_{i=1}^n m_i \times \vcount{\phi_i(\vec{x})}$, where
-    $\phi_i \in \FO$ has quantifier rank at most $q$
-    and $\card{\vec{x}} = k$
-    \cite[Theorem 7.10]{DOUE23}.
-
-    As in the proof of \cref{n-poly-k-implies-wqo:lemma}, we are going to
-    assign a tuple of integers to a word $u \in \Sigma^*$ by counting the
-    number of realizations of each $\FO^q$ type of at most $k$ variables in
-    $u$. To that end, let us write $\FO^q(u)$ this vector.
-
-    First, let us notice that it suffices to prove that for some $n \in \Nat$,
-    $\FO^q(uw^n) \leq \FO^q(uw^{n+1})$ to conclude, since $\app{f}{uw^{n+1}} -
-    \app{f}{uw^n}$ will be obtained as a positive combination of counting
-    first-order types in the argument.
-
-    Let $t$ be a first order type with at most $k$ free variables and
-    quantifier rank at most $q$. The map $g_t \colon X \mapsto
-    \vcount{t}(uw^X)$ is a \kl{commutative} \kl{star-free $\Nat$-polyregular
-    function}. As a consequence, there exists $N_0$ and a polynomial $P \in
-    \CorrectPoly$ such that for all $X \geq N_0$, $g_t(X) = P(X)$. Now, because
-    $P \in \CorrectPoly$, there exists a $K \in \Nat$ such that
-    $\translate{K}(\Diff{1}{P}) \in \Nat[X]$
-    (\cref{derivation-translation:lem}), and in particular $g_t(X + N_0 + K+1)
-    - g_t(X + N_0 + K) \in \Nat$, for all $X \geq 0$.
-
-    Because there are finitely many non-equivalent $\FO^q$ types with at most 
-    $k$ free variables, we can take the maximum of the $K$'s obtained for each 
-    of those, and conclude.
-\end{proofof}
-
 \section{Proofs of section \ref{sec:ccl}}

### 2024-08-13 parts/correction.tex
starving the beast
correct proof of the lemma 15

@@
 
-Let us now prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
-This follows from the correct implication in the statement of \cref{karh:thm},
-but we provide a self-contained proof using a refinement of the classical
-combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and $\NPoly$
-\cite[Lemma 5.37]{DOUE23}.
+\AP Let us now prove that \kl{$\Nat$-rational polynomials} are in
+$\CorrectPoly$. This follows from the correct implication in the statement of
+\cref{karh:thm}, but we provide a self-contained proof using a refinement of
+the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
+and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We believe that the formulation of such
+pumping arguments can be made clearer through some extra notations, with the
+added property of highlighting the central role of \kl{commutative} polyregular
+functions. Let us define that a function $q$ is a \intro{(noncommutative output)
+polynomial function} from $\Nat^p$ to $\Sigma^*$ whenever there exists words
+$\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words $u_1, \dots, u_p \in
+\Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0 \prod_{i = 1}^p u_i^{X_i}
+\alpha_i$. That is, $q$ is syntactically defined by a non-commutative
+polynomial function over the semi-ring $\Sigma^*$. Almost by definition,
+\kl{polynomial functions} are \kl{commutative} \kl{polyregular functions}.
+
+\AP There is still a minor detail that remains to be smoothened before proving
+our combinatorial \cref{n-poly-combinatorics:lem}, which is that some
+polyregular functions can exhibit a limited form of division. For instance, the
+function computing the sum of the lengths of the suffixes of a word $w$
+actually computes $\card{w} \times (\card{w} - 1) / 2$. This will be at odds
+with our restriction to polynomials with coefficients in $\Rel$. However, we
+argue that up to a suitable change of basis for the ring of polynomials, all
+coefficients will actually be in $\Nat$. This relatively simple trick works by
+considering as ``basic polynomials" the \intro{binomial polynomials} of the
+form $\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural numbers. This is
+why we introduce the notion of \intro{natural binomial polynomial} as
+$\Nat$-linear combinations of products of \kl{binomial polynomials}.
 
@@
     There exists a computable $\omega \in \Nat_{\geq 1}$
-    such that for all $p \in \Nat$,
-    for all $\alpha_0, \dots, \alpha_p \in \Sigma^*$,
-    for all $u_1, \dots, u_p \in \Sigma^*$,
-    there exists a polynomial $P \in \Rel[X_1, \dots, X_p]$
-    whose \kl{maximal monomials} are \kl{non-negative},
-    and such that for all $X_1, \dots, X_p \geq \omega$:
+    such that for all \kl{polynomial functions}
+    $q \colon \Nat^p \to \Sigma^*$,
+    there exists a computable \kl{natural binomial polynomial} $P$
+    such that:
     \begin{equation*}
-        f\left(
-            \alpha_0 \prod_{i = 1}^p u_i^{\omega X_i} \alpha_i
-        \right)
-        = P(X_1, \dots, X_p) \quad .
+        f \circ q(\omega X_1, \dots, \omega X_p)
+        =
+        P
+        \quad 
+        \text{ over } (\Nat_{\geq 1})^p
     \end{equation*}
@@
 \begin{proof}
-    \textbf{TODO: reprove}
-
-    We leverage the characterization of \kl{$\Nat$-polyregular functions}
-    given in \cref{nat-rel-poly:def}.
-    Let $\omega$ be an idempotent power for the finite monoid $M$,
-    and
-    $w \defined \alpha_0 \prod_{i = 1}^p (u_i^{\omega \times X_i} \alpha_i)$.
-    Let $k \in \Nat$, and $\pi \colon M^k \topartial \Nat$ be such that
-    $f = \pi^\dagger$. That is, 
+    Let us fix a finite monoid $M$, a morphism $\mu \colon \Sigma^* \to M$, and 
+    let $\omega$ be an idempotent power for the finite monoid $M$,
+    i.e., a number such that for all $x \in M$,
+    $x^{\omega+1} = x^\omega$.
+    We will prove by induction on $k \in \Nat_{\geq 1}$
+    that for all
+    $\pi \colon M^k \to \Nat$,
+    the lemma holds for the function 
+    $f = \pi^\dagger$, i.e., such that
     $f(w)$ is the sum over all factorizations of $w$
@@
     of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
+    Since every \kl{$\Nat$-polyregular function}
+    is obtained via some choice of $M, \mu, \pi$ (\cref{nat-rel-poly:def})
+    we derive the desired result.
 
-    Let us define an equivalence relation $\equiv$ over words of
-    $\Sigma^*$ as follows: two words $v_1, v_2$ are equivalent if they are
-    equal when normalizing them with the rules $u_i^{\omega} u_i^{\omega} \to
-    u_i^{\omega}$ for all $1 \leq i \leq p$. That is, we disregard
-    repetitions of the factors $u_i^{\omega}$ in the words.
-    We lift the equivalence relation $\equiv$ to $k$-tuples of
-    words by pointwise application.
+    In all cases, we will use a \kl{polynomial function} $q \colon \Nat^p \to
+    \Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
+    (u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ ($0 \leq i
+    \leq p$) and $u_i$ ($1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
+    implicitly multiply all variables by $\omega$ in order to make the
+    equations more readable. Note that when using an induction hypothesis, we
+    will therefore have to check that the polynomial $q$ will have this
+    specific form.
 
-    Remark that if $(v_1, \dots, v_k) \equiv (w_1, \dots, w_k)$ then their
-    production are the same, i.e., $\pi(\mu(v_1), \dots, \mu(v_k)) =
-    \pi(\mu(w_1), \dots, \mu(w_k))$. This holds because $\mu(u_i^\omega) =
-    \mu(u_i)^\omega$ is an idempotent element of the monoid $M$. In particular,
-    $f(w)$ can be computed as the sum over equivalence classes for $\equiv$, of
-    the value of $\pi$ on one representative, which belongs to $\Nat$.
+    \textbf{Base case: $k = 1$.}
+    Let us show that $f \circ q$ is
+    actually a constant (positive) function (and in particular a \kl{natural binomial polynomial}), as for
+    all $n_1, \dots, n_p \geq 1$,
+    \begin{align*}
+        f(q(n_1, \dots, n_p))
+        &\eqdef \pi\left(\mu(q(n_1, \dots, n_p))\right) \\
+        &\eqdef \pi\left(\mu(\alpha_0 \prod_{i = 1}^p u_i^{\omega n_i} \alpha_i)\right)
+        \\
+        &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i)^{\omega n_i} \mu(\alpha_i)\right) 
+        & \text{ morphism }
+        \\
+        &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
+        & \text{ idempotent and } n_i \geq 1
+        &= c \geq 0
+    \end{align*}
 
-    Therefore, it suffices to prove that the number of partitions of $w$ in a
-    single equivalence class for $\equiv$ is a polynomial $P(X_1, \dots, X_p)$
-    with \kl{non-negative} \kl{maximal monomials}.
+    \textbf{Induction hypothesis.}
+    The proof will essentially follow the same schema as the base case,
+    with extra care needed to properly handle the summation of polynomials
+    that arises when $k \geq 2$.
 
-    Notice that to produce all elements in a given equivalence class
-    means adding the factors $u_i^\omega$ where $1 \leq i \leq p$
-    back to the normalized version
-    of the equivalence class. For all elements in an equivalence class,
-    one has to add the same amount of $u_i^\omega$ for $1 \leq i \leq p$,
-    and therefore the number of possible choices
-    is either constant, or 
-    is a product of binomials of the form $\binom{X - l}{s}$
-    which almost has the desired shape.
+    We will use the fact that given $n_1, \dots, n_p \geq 0$,
+    there exists a bijection between
+    partitions of $q(\omega n_1, \dots, \omega n_p)$
+    into $k+1$ words, and 
+    triples composed of
+    a position  $1 \leq \ell \leq p$,
+    a ratio     $0 \leq b \leq n_\ell$,
+    a remainder $0 \leq r < \omega \times \card{u_\ell}$,
+    and a partition of
+    $u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
+    \alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
+    into $k$ words,
+    where we have used the notation 
+    $v\mid_{[-r:]}$ to denote the word obtained from $v$ 
+    by keeping only the last $r$ letters, in a sort of Pythonic syntax.
+    That is, we look at where the first ``cut" is made.
 
-    Indeed, the binomial 
-    $\binom{X - l}{s}$ is equal to
-    $(X-l) \cdots (X-l-s+1) / s!$.
-    As a consequence,
-    $\binom{s! X - l}{s} \in \Rel[X]$,
-    and considering $\omega' \defined \omega \times s!$ (for some large enough $s$)
-    allows us to conclude.
+    Now, given $(\ell, b, r)$ respecting the inequalities 
+    mentioned above,
+    one can define the \kl{polynomial functions}
+    $s_{\ell,b,r} \colon \Nat^{\ell - 1} \to \Sigma^*$
+    and
+    $q_{\ell,b,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
+    so that
+    $q(X_1, \dots, X_p) = s_{\ell, b, r}(X_1, \dots, X_{\ell-1}) 
+    q_{\ell, b, r}(X_{\ell}, \dots, X_p)$
+    under the assumption that $X_1 \geq b$:
+    \begin{equation*}
+        q_{\ell,b,r}(X_{1}, \dots, X_{p - \ell + 1})
+        \defined
+        u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times (X_1 - b)}
+        \alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
+        \quad .
+    \end{equation*}
+    Similarly, let us define 
+    $s_{\ell, b, r}(X_1, \dots, X_{\ell - 1})$
+    as 
+    $\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1} 
+    u_\ell^{\omega b} u_{\ell}^\omega \mid_{[1:r]}$,
+    Remark that $q_{\ell,b,r}$ has all of its indeterminates
+    multiplied by $\omega$, so that the induction hypothesis 
+    can be used.
+
+    The key remark is that $\mu(s_{\ell,b,r}(X_1, \dots, X_{\ell - 1}))$ is a
+    constant function from $\Nat^{\ell-1}$ to $M$, and that its value is also
+    constant when $b$ varies among numbers greater than $0$. This holds with
+    the exact same proof as the base case of this induction. Let us call
+    $m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
+    obtained by $\mu(s_{\ell,b,r}(X_1, \dots, X_{\ell-1}))$ respectively when
+    $b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
+    \to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
+    \pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$ and similarly for $\pi_{\ell,r,1}
+    \colon M^k \to \Nat$.
+
+    With all of this preliminary work done, we are now ready to conclude by
+    simply unrolling the definition of 
+    a function
+    $f$ defined using a \kl{production function}
+    $\pi \colon M^{k+1} \to \Nat$.
+    In the following equations, we omit the precise
+    range for the sum over $(\ell, b, r)$ to save space.
+    Similarly, we will use 
+    $q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
+    \defined q_{\ell,b,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
+    at the end of the computation.
+    \begin{align*}
+        &f(q(n_1, \dots, n_p))\\
+        &\eqdef 
+        \sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
+        \pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
+        &= \sum_{(\ell, b, r)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
+           \pi(\mu(s_{\ell,b,r}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
+        \\
+        &= 
+           \sum_{(\ell, 0, r)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,0,r}(n_\ell, \dots, n_p)}
+           \pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
+           \\
+        &+ 
+           \sum_{(\ell, b > 0, r)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
+           \pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
+        \\
+        &= 
+           \sum_{(\ell, 0, r)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,0,r}(n_\ell, \dots, n_p)}
+           \pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
+           \\
+        &+ 
+           \sum_{(\ell, b > 0, r)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
+           \pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
+        \\
+        &= 
+           \sum_{1 \leq \ell \leq p}
+           \sum_{0 \leq r < \omega \times \card{u_\ell}}
+           (\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell, \dots, n_p))
+           \\
+        &+ 
+           \sum_{1 \leq \ell \leq p}
+           \sum_{0 \leq r < \omega \times \card{u_\ell}}
+           \sum_{0 < b < n_\ell}
+           (\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
+    \end{align*}
+    Now, using the induction hypothesis,
+    we have \kl{natural binomial polynomials} $P_{\ell,r,0}$ and $P_{\ell,r,1}$
+    satisfying 
+    \begin{equation*}
+       (\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
+       = P_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
+       \quad 
+       \forall n_\ell - b, \dots, n_p \geq 1
+       \quad .
+   \end{equation*}
+   And similarly
+    \begin{equation*}
+        (\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
+        = P_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
+       \quad 
+       \forall n_1, \dots, n_{\ell - 1} \geq 1
+       \quad .
+   \end{equation*}
+
+   It is an easy and folklore fact that \kl{natural binomial polynomials} are
+   stable under the operation $\Sigma_Y \colon Q(X_1,\dots,X_n,Y) \mapsto \sum_{0
+   < b < Y} Q(X_1, \dots, X_n, b)$, and by pre-composition with 
+   a multiplication by a constant $\omega \geq 1$.
+
+   This allows us to conclude by defining 
+   \begin{equation*}
+       P(X_1, \dots, X_p) \defined
+       \sum_{1 \leq \ell \leq p} \sum_{0 \leq r < \omega \card{u_\ell}}
+       P_{l,r,0}(\omega X_{\ell}, \dots, \omega X_p) + \Sigma_Y P_{l,r,1}(\omega Y,
+       \omega X_{\ell+1}, \dots, \omega X_p)
+       \quad .
+   \end{equation*}
+   The latter being a \kl{natural binomial polynomial} satisfying
+   the desired equality.
 \end{proof}
+
+Let us highlight that the multiplicative factor $\omega$ is necessary in
+\cref{n-poly-combinatorics:lem}. Indeed, the function $f \colon \set{a}^* \to
+\Nat$ defined as $0$ when the input is of odd length and $1$ when the input is
+of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial.
 
@@
     \begin{enumerate}
-        \item $P$ is a \kl{$\Nat$-rational polynomial}.
+        \item $P$ is a \kl{$\Nat$-rational polynomial},
+        \item $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
         \item $\alpha P \in \CorrectPoly$,

### 2024-08-14 parts/appendix.tex
gigantic rewrite

@@
     $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
+    We 
+    compute $\omega_+$ (resp. $\omega_-$)
+    by
+    applying
+    \cref{n-poly-combinatorics:lem} 
+    to $f_+$ (resp. to $f_-$).
+    Finally, let $\omega \defined \ppcm(\omega_+, \omega_-)$.
 
-    We apply
-    \cref{n-poly-combinatorics:lem} to compute the $\omega$ that works for both $f_+$ and $f_-$.
     Using this $\omega$,
@@
     $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
-    there exists a polynomial $P_r \in \Rel[X_1, \dots, X_p]$
-    such that $X_1, \dots, X_p \geq \omega$:
+    there exists an \kl{integer binomial polynomial}
+    $P_r \in \Rel[X_1, \dots, X_p]$,
+    obtained as
+    the difference of the two \kl{natural binomial polynomial} $P_r^+$ and $P_r^-$
+    associated to $f_+$ and $f_-$,
+    such that $X_1, \dots, X_p \geq 1$:
     \begin{equation*}
@@
     \end{equation*}
-    To get the full family of polynomials needed requires
-    just to only iterate subsets of letters, and fix the other ones
-    to be below the threshold $\omega^2$.
+    We conclude by selecting different values of $r$ and
+    setting some subsets of variables to be assigned to the constant $0$.
 \end{proofof}

### 2024-08-14 parts/correction.tex
gigantic rewrite

@@
 coefficients will actually be in $\Nat$. This relatively simple trick works by
-considering as ``basic polynomials" the \intro{binomial polynomials} of the
-form $\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural numbers. This is
-why we introduce the notion of \intro{natural binomial polynomial} as
-$\Nat$-linear combinations of products of \kl{binomial polynomials}.
+considering as ``basic polynomials" the \intro{binomial monomial} of the form
+$\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural numbers. Beware that
+$\binom{X - \ell}{k}$ is defined to be $0$ when $X \leq \ell$, and is therefore
+a ``true polynomial" only when $X \geq \ell$. This is why we introduce the
+notion of \intro{natural binomial polynomials} as $\Nat$-linear combinations of
+products of \kl{binomial monomials}.
 
@@
     \Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
-    (u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ ($0 \leq i
-    \leq p$) and $u_i$ ($1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
+    (u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ (for $0 \leq i
+    \leq p$) and $u_i$ (for $1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
     implicitly multiply all variables by $\omega$ in order to make the
@@
 of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial.
+Let us remark that the \kl{maximal monomials} of the polynomial $\binom{X -
+p}{k}$ have coefficient $1/k!$, and are therefore \kl{non-negative}.
 
@@
     there exists a number $\omega \in \Nat_{\geq 1}$
-    and polynomial $Q$ with \kl{non-negative}
-    \kl{maximal monomials} such that
-    for all $X_1, \dots, X_p \geq \omega$:
+    and \kl{natural binomial polynomial} $Q$
+    such that
+    for all $n_1, \dots, n_p \geq 1$:
     \begin{equation*}
         f\left(
-            \prod_{i = 1}^p a_i^{\omega X_i}
+            \prod_{i = 1}^p a_i^{\omega n_i}
         \right)
-        = Q(X_1, \dots, X_p)
-        = P(\omega X_1, \dots, \omega X_p) 
+        = Q(n_1, \dots, n_p)
+        = P(\omega n_1, \dots, \omega n_p) 
         \quad .
@@
 
-\begin{fact}
-    \label{maxi-monomials-submodular:fact}
-    Let $P, Q$ be two polynomials.
-    Then, $\MaximalMonomials(P+Q) \hoareleq \MaximalMonomials(P)
-    \cup \MaximalMonomials(Q)$.
-\end{fact}
-
-\begin{fact}
-    \label{translation-maximal:fact}
-    For all $P \in \Rel[\vec{X}]$, and $K \in \Nat$,
-    \begin{equation*}
-        \MaximalMonomials(P) = \MaximalMonomials(\translate{K}(P))
-        \quad .
-    \end{equation*}
-\end{fact}
-
-\textbf{TODO: fix this?}
-\begin{fact}
-    \label{derivation-monomials:fact}
-    Let $K > 0$,
-    and let $M,T$ be two \kl{monomials}, such that
-    $T$ \kl{strictly divides} $M$.
-    Then,
-    every \kl{monomial} of $\Diff{K}{M}$ \kl{strictly divides} $M$,
-    and 
-    $T$ \kl{divides} some \kl{maximal monomial}
-    of $\Diff{K}{M}$, which has a coefficient
-    that is a multiple of $K$.
-\end{fact}
-\begin{proof}
-    Let $M \defined X_1^{k_1} \cdots X_n^{k_n}$,
-    where $k_1, \dots, k_n > 0$ and $n > 0$.
-    Let us compute
-    \begin{align*}
-        \Diff{K}{M} &= \prod_{i = 1}^n X_i^{k_i} - \prod_{i = 1}^n (X_i - K)^{k_i} \\
-                    &= \sum_{p = 1}^n \frac{K}{X_p - K} \prod_{i = 1}^n (X_i - K)^{k_i} 
-    \end{align*}
-    Therefore, 
-    the \kl{maximal monomials} of $\Diff{K}{M}$
-    are precisely
-    of the form $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$ for some $1 \leq p \leq i$,
-    which \kl{strictly divides} $M$.
-    Let $T$ be a \kl{monomial} that \kl{strictly divides} $M$,
-    then, $T = \prod_{i = 1}^n X_i^{t_i}$, 
-    for all $1 \leq i \leq n$, $t_i \leq k_i$,
-    and there exists $1 \leq p \leq n$ such that $t_j < k_j$.
-    In particular, $T$ \kl{divides}
-    $K \frac{\prod_{i = 1}^n X_i^{k_i}}{X_p}$.
-\end{proof}
-
 \begin{lemma}
-    \label{derivation-simplifies:lemma}
-    For all $P \in \Rel[\vec{X}]$ that are non-constant,
-    for all $K \in \Nat$,
-    $\MaximalMonomials(\Diff{K}{P}) \hoarele
-    \MaximalMonomials(P)$.
+    \label{all-positive-derivative:lem}
+    Let $P \in \Nat[\vec{X}]$ that is non-constant, and $K \in \Nat$,
+    then $\Diff{K}{P} \in \Nat[\vec{X}]$ and all of its
+    coefficients are (positive) multiples of $K$.
+    Furthermore, every monomial that strictly divides some monomial of $P$
+    appears in $\Diff{K}{P}$.
 \end{lemma}
 \begin{proof}
-    Let $P \in \Rel[\vec{X}]$ and
-    $K \in \Nat$.
-    Leveraging 
-    \cref{derivation-monomials:fact,discrete-deriv-linear:fact,maxi-monomials-submodular:fact}:
-    \begin{align*}
-        \MaximalMonomials\left(\Diff{K}{P}\right)
-        &=
-        \MaximalMonomials\left(\sum_{M \in \Monomials(P)} \Diff{K}{M}\right) \\
-        &\hoareleq
-        \bigcup_{M \in \Monomials(P)} 
-        \MaximalMonomials(\Diff{K}{M}) \\
-        &\hoarele
-        \MaximalMonomials(P)
-        & \text{ because } \deg(\Diff{K}{P}) < \deg(P)
-    \end{align*}
-    Which concludes the proof.
+    We prove the result for monomials, as it extends
+    to $\Nat$-linear combinations by linearity.
+    Let $P = \prod_{i = 1}^k X_i^{\alpha_i}$ be a monomial.
+    Notice that $\translate{K}(P) = \prod_{i = 1}^n (X_i + K)^{\alpha_i}$,
+    and using a binomial expansion
+    we list all the possible divisors of $P$,
+    all of which with coefficients that are positive integers and multiples of $K$ except the coefficient
+    of the maximal monomial (equal to $P$ itself) which is $1$.
+    As a consequence, $\translate{K}(P) - P$ is simply
+    obtained by removing this maximal monomial, which concludes the proof.
 \end{proof}
@@
     such that
-    $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$.
-    \proofref{derivation-stabilises-correct:lem}
+    $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2)) \in \Nat[\vec{X}]$.
 \end{lemma}
+\begin{proof}
+    Let us first tackle the specific case where $P$ is a constant polynomial.
+    In this case, $P_1 = P$ and $P_2 = 0$.
+    Furthermore,
+    $\translate{K}(P_1) = P_1$ and $\Diff{K}{P_1} = 0$ for all $K \in \Nat$.
+    We conclude that $\Diff{K}{P_1} + \translate{K}(P_2) = 0$
+    for all $K \in \Nat$, hence belongs to $\Nat[\vec{X}]$. Selecting $K = 0$
+    we conclude.
+
+    Assume now that $P$ is not a constant polynomial. We will use
+    \cref{all-positive-derivative:lem} on a well-selected value of $K$. Let us
+    write $\alpha$ to be the maximal absolute value of a coefficient in $P$.
+    Let $D$ be the number of unitary monomials that divide some monomial
+    appearing in $P$. We can now define $K \defined D \times \alpha$,
+    and let
+    $Q \defined (\Diff{K}{P_1} + \translate{K}(P_2))$.
+    Remark that $\Diff{K}{P_1}$ is already in $\Nat[\vec{X}]$,
+    and the constant term of $\translate{K}(P_2)$ is also
+    in $\Nat$.
+    For any other monomial of $P_2$, by the maximality of $P_1$,
+    it strictly divides some monomial of $P_1$,
+    equals some monomial of $\Diff{K}{P_1}$. Because every monomial
+    of $\Diff{K}{P_1}$ has a coefficient that is a multiple of $K = \alpha \times D$, we can
+    match every monomial of $P_2$ to $\alpha$ copies of a unitary
+    monomial appearing in $\Diff{K}{P_1}$, and because $\alpha$ is large enough
+    we conclude that 
+    $Q \in \Nat[\vec{X}]$.
+\end{proof}
 
@@
 \begin{proof}
-    We prove the result by induction on $\MaximalMonomials(P)$. 
+    We prove the result by induction on the number of indeterminates of $P$.
     In the proof, we write $\vec{X}$ for the indeterminates appearing in $P$,
@@
     monomials} of $P$. We compute a bound $K$ such that $Q \defined
-    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \CorrectPoly$ using
-    \cref{derivation-stabilises-correct:lem}. Thanks to
-    \cref{derivation-simplifies:lemma}, we also know that $\MaximalMonomials(Q)
-    \hoarele \MaximalMonomials(P)$. By induction hypothesis, $Q$ is
-    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} which is
-    effectively computable.
-
-    Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
+    (\Diff{K}{P_1} + \translate{K}(P_2)) \in \Nat[\vec{X}]$ using
+    \cref{derivation-stabilises-correct:lem}. In particular, $Q$ is
+    \kl{represented} by a \kl{star-free $\Nat$-polyregular function} using
+    \cref{n-poly-n-poly:example}, the latter being effectively computable from
+    $Q$. Let us now remark that $P_1 \in \Nat[\vec{X}]$, and is therefore
     (effectively) \kl{represented} by a \kl{star-free $\Nat$-polyregular
-    function} (\cref{n-poly-n-poly:example}). As a consequence,
+    function} (using again \cref{n-poly-n-poly:example}). As a consequence,
     $\translate{K}(P) = P_1 + Q$ is (effectively) \kl{represented} by a
@@
     For all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots,
-    K}$ fixing at least one indeterminate,  we know that
-    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials{P}$. Because
-    $\restr{P}{\nu} \in \CorrectPoly$, one can use the induction hypothesis to
-    compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
-    \kl{represents} $\restr{P}{\nu}$. This is possible because we assumed 
-    that all indeterminates in $\vec{X}$ are used in $P$.
+    K}$ fixing at least one indeterminate, one can use the induction hypothesis
+    to compute a \kl{star-free $\Nat$-polyregular function} $f_\nu$ that
+    \kl{represents} $\restr{P}{\nu}$. This is possible because we assumed that
+    all indeterminates in $\vec{X}$ are used in $P$.
 
@@
     \end{enumerate}
+    In particular, the above properties are decidable.
     \proofref{derivation-translation:lem}
@@
     \begin{enumerate}
+        \item \label{corrected-0:item} $P$ is a \kl{natural binomial polynomial},
         \item \label{corrected-1:item} $P \in \CorrectPoly$,
@@
 \end{proof}
+
+\AP As a consequence, we can also fully characterize polynomials that can be
+computed using \kl{$\Rel$-rational series}, as they are precisely those
+computable by a difference of two \kl{$\Nat$-rational polynomials}. This drives
+us to introduce the class of \intro{integer binomial polynomials}, which are
+obtained by $\Rel$-linear combinations of \kl{binomial monomials} of the form
+$\binom{X - p}{k}$ for some $k,p \in \Nat$.
+
+\begin{corollary}
+    \label{integer-binomial-polynomials:cor}
+    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
+    if and only if $P$ is a \kl{integer binomial polynomial}.
+\end{corollary}

### 2024-08-14 parts/intro.tex
gigantic rewrite

@@
 with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$
-\kl{for star-free polyregular functions} with unary output)
+for \kl{star-free polyregular functions} with unary output)
 \cite{DOUE21,DOUE22}. From this point of view, $\NPoly$ is a subclass of a

### 2024-08-14 parts/starfree.tex
gigantic rewrite

@@
 procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
-\cref{decompose-polynomial:lem} refines
+\cref{decompose-polynomial:lem} can be understood as a refinement of
 \cref{n-poly-combinatorics:lem} for \kl{commutative} functions.
@@
     and computable 
-    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
-    such that for all $x_1, \dots, x_n \geq \omega$:
+    \kl{integer binomial polynomials} $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
+    such that for all $x_1, \dots, x_n \geq 0$:
     \begin{equation*}
@@
     for all
-    $\alpha_0, \dots, \alpha_k \in \Sigma^*$,
-    for all $u_1, \dots, u_k \in \Sigma^*$,
+    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
     there exists a polynomial $P \in \CorrectPoly$,
-    satisfying for all $x_1, \dots, x_k \geq \omega$:
+    satisfying:
     \begin{equation*}
-        f
-        \left(
-            \alpha_0 \prod_{i = 1}^k u_i^{\omega x_i} \alpha_i
-        \right)
+        f \circ q(\omega X_1,\dots, \omega X_k)
         = 
-        P(x_1, \dots, x_k) \quad .
+        P
+        \quad 
+        \text{ over } (\Nat_{\geq 1})^k
+        \quad .
     \end{equation*}
@@
 \begin{proof}
-    Let us prove that
+    The implication
     \cref{f-npoly-combi:item}
-    implies \cref{f-combinatorial:item}. 
-    We compute the $\omega$ by applying 
-    \cref{n-poly-combinatorics:lem}, and then
-    notice that the polynomial $P$ obtained is
-    by definition a \kl{$\Nat$-rational polynomial}.
+    $\implies$ \cref{f-combinatorial:item} immediately follows
+    from 
+    \cref{n-poly-combinatorics:lem} and \cref{corrected-version:thm}.
 
     For the converse implication, we use \cref{decompose-polynomial:lem} to
-    effectively compute a polynomial representation of $f$.
+    effectively compute a polynomial representation of $f$. That is, we obtain
+    an $\omega_1 \in \Nat$ and \kl{integer binomial polynomials}
+    $\seqof{P_{t}}{t \in \ModuloTypes[\omega_1]^k}$. Now, because $f$ is
+    \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
+    such that for all \kl{polynomial functions} $q$, there exists an
+    \kl{integer binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
+    = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
+    \omega_1 \times \omega_2$, we can without loss of generality assume that
+    $\omega_1 = \omega_2$ in the rest of this proof.
 
-    \textbf{TODO: this is because they coincide en enough points}.
-    Using the
-    assumption, all of these polynomials are actually \kl{$\Nat$-rational
-    polynomials}. Let us now construct a new way to compute $f$, by first
+    Now, consider an \kl{$\omega$-type} $t \in \ModuloTypes[\omega]^k$. By
+    definition, this provides us with a selection function $\ind{S} \colon \set{1,
+    \dots, k} \to \set{0,1}$, and numbers $r_1, \dots, r_k$ between $0$ and
+    $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
+    \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has type $t$.
+    In particular, for all $x_1, \dots, x_k \geq 1$, the following
+    equality holds:
+    \begin{equation*}
+        f\left(\prod_{i = 1}^k 
+            a_i^{ x_i \omega \ind{S}(i) + r_i}
+        \right)
+        =
+        P_t(\seqof{x_i}{i \in S})
+    \end{equation*}
+
+    Let us therefore consider the \kl{polynomial function} $q \colon \Nat^k \to
+    \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
+    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
+    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists
+    a polynomial $P \in \CorrectPoly$ such that for all $x_1, \dots, x_n
+    \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k) = P(x_1,
+    \dots, x_k)$ holds. In particular, we conclude that over $(\Nat_{\geq
+    1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They are therefore
+    equal as multivariate polynomials, and we conclude that $P_t$ belongs to
+    $\CorrectPoly$, and in particular is an \kl{$\Nat$-rational polynomial}
+    thanks to \cref{corrected-version:thm}.
+
+    Let us now construct a new way to compute $f$, by first
     computing the \kl{$\omega$-type} of the input (which is possible in
@@
 \kl{$\Nat$-polyregular function}. For instance, the polynomial $P(X) \defined
-X(X+1)/2$. This is why, in the upcoming definition, we have to consider
-polynomials with coefficients in $\Rat$.
+X(X+1)/2$. Historically, the definition used polynomials with coefficients in
+$\Rat$, which is why we use such polynomials here, even though we know that all
+of those polynomials will actually be \kl{integer binomial polynomials} thanks
+to \cref{integer-binomial-polynomials:cor}.
 
@@
     when there exists $N_0 \in \Nat$ such that
-    for all $n \in \Nat$
-    and for all words $\alpha_0, w_1, \alpha_1, \cdots, \alpha_{n-1}, w_n, \alpha_n
-    \in \Sigma^*$, there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
+    for all $n \in \Nat$,
+    for all \kl{polynomial function} $q \colon \Nat^n \to \Sigma^*$,
+    there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
     such that:
     \begin{equation*}
-        f\left(
-            \alpha_0 \prod_{i = 1}^{n} w_i^{x_i} \alpha_i
-        \right)
-        = 
-        P(x_1, \dots, x_n)
+        f \circ p = P
         \quad 
-        \forall x_1, \dots, x_n \geq N_0
+        \text{ over } (\Nat_{\geq N_0})^n
         \quad .
@@
 recognized by an \kl(monoid){aperiodic} finite monoid, which turns out to be
-equivalent to the following characterisation among regular languages: for all
+equivalent to the following characterization among regular languages: for all
 $u,v,w$, there exists $n \in \Nat$, $u v^n w \in L$ if and only if $u v^{n+1} w
@@
 if and only if its indicator function $\ind{L}$ is \kl{ultimately polynomial}.
-This was already observed in \cite[Claim V.6]{CDTL23}, and reproven in
+This, was already observed in \cite[Claim V.6]{CDTL23}, and restated in
 \cite[Claim 7.45, Lemma 7.53]{DOUE23}. We include a self-contained proof
-hereafter for readability and didactic purposes.
+hereafter for readability and didactic purposes. The proof follows essentially
+the same scheme as the base case of \cref{n-poly-combinatorics:lem}.
 
@@
 relies on the construction of a canonical object called the \emph{residual
+,
 transducer}, the latter being essentially based on differences between
 functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
-proof method does not carry from $\Rel$-output functions to $\Nat$-output
+proof method does not carry from $\Rel$-output functions to $\Nat$-output,
 functions, it was conjectured that the semantic property of being

### 2024-08-14 parts/correction.tex
typos

@@
 that $\CoveredPoly$ is not stable under fixing indeterminates, while
-\kl{$\Nat$-polyregular functions} are. In this section, we prove that closing
+\kl{$\Nat$-polyregular functions} are. 
+Indeed, the polynomial $\BadPoly$ introduced in
+\cref{def:bad-polynomial} satisfies $\BadPoly(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
+coefficient for a \kl{maximal monomial}, namely $-2XY$.
+In this section, we prove that closing
 $\CoveredPoly$ under variable assignments is enough to recover from
@@
 First, let us remark that $\CorrectPoly \subseteq \CoveredPoly$, because
-polynomials in $\CorrectPoly$ are \kl{non-negative}. Let us also check that the
-counter example provided in \cref{thm:counter-example} is not in
-$\CorrectPoly$. For that, notice that for $\BadPoly$ introduced in
-\cref{def:bad-polynomial}, $\BadPoly(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$.
+polynomials in $\CorrectPoly$ are \kl{non-negative}. We also remarked at the
+beginning of this section that our counter example provided in
+\cref{thm:counter-example} is not in $\CorrectPoly$.
 
@@
 As one would expect, the \kl{discrete derivatives} are linear operations on
-polynomials, that commute with \kl{translation operators}. However, the
+polynomials, and they commute with \kl{translation operators}. However, the
 \kl{translation operators} do not perfectly commute with partial applications
@@
 
-While \cref{lem:correct-to-n-rat} provides effective conversion, it does not
-explicitly state that the membership is decidable to keep the proof clearer. A
-similar proof scheme can be followed to conclude that membership is decidable,
-and even show that elements in $\CorrectPoly$ are, up to suitable translations,
-polynomials in $\Nat[\vec{X}]$. Beware that partial applications are still
-needed in this characterization, as \cref{bad-poly-translate:ex} illustrates.
-Effectiveness was the last missing piece of our main result
+While \cref{lem:correct-to-n-rat} provides an effective conversion procedure,
+it does not explicitly state that the membership is decidable to keep the proof
+clearer. A similar proof scheme can be followed to conclude that membership is
+decidable, and even show that elements in $\CorrectPoly$ are, up to suitable
+translations, polynomials in $\Nat[\vec{X}]$. Beware that partial applications
+are still needed in this characterization, as \cref{bad-poly-translate:ex}
+illustrates. Effectiveness was the last missing piece of our main result
 \cref{corrected-version:thm}, which corrects and generalizes \cref{karh:thm}.
@@
 
-For completeness, let us remark that the counter example of
+For completeness, let us remark that the counterexample of
 \cref{thm:counter-example} uses three indeterminates, and this is not a

### 2024-08-14 parts/counter.tex
typos

@@
 \kl{represents} it. This particular problem was claimed to be solved in
-\cite{KARH77}, using a characterization restated in \cref{karh:thm}.
+\cite{KARH77}, using the \cref{karh:def} to characterize \kl{$\Nat$-polyregular
+functions}, as restated in \cref{karh:thm}.
 
@@
 \begin{definition}[{\cite[Section 3, page 3]{KARH77}}]
+    \label{karh:def}
     Let $\vec{X}$ be a finite tuple of indeterminates.
@@
 polynomials}. In order to derive this example, we will need the following
+fundamental
 result about the pre-image of regular languages by \kl{polyregular functions}.\footnote{

### 2024-08-14 parts/prelim.tex
typos

@@
     \label{stability-polyregular:lemma}
-    The class $\NPoly[k]$ (resp. $\ZPoly[k]$) is stable under the operations of
-    addition, multiplication, Cauchy product,
-    and ``value pinning''.
+    Let $k \geq 1$ and 
+    $f,g \in \NPoly[k]$ (resp. $\ZPoly[k]$, $\NSF[k]$, $\ZSF[k]$),
+    $L$ be a \kl{star-free language} over $\Sigma^*$.
+    Then, the following 
+    are also in $\NPoly[k]$ (resp. $\ZPoly[k]$,
+    $\NSF[k]$, $\ZSF[k]$):
+    \begin{itemize}
+        \item $f + g \defined w \mapsto f(w) + g(w)$,
+        \item $f \times g \defined w \mapsto f(w) \times g(w)$,
+        \item $\ind{L} \times f$ 
+    \end{itemize}
+    Furthermore, the above constructions preserve \kl{commutativity}.
 \end{lemma}
@@
 Let us briefly state that \kl{commutativity} is a decidable property of
-\kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}.
+\kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}. As a
+consequence, we are working inside a relatively robust and decidable class of
+\kl{$\Rel$-rational series}.
 

### 2024-08-14 parts/starfree.tex
typos

@@
 polynomials with coefficients in $\Rat$ that can be \kl{represented} by a
-\kl{$\Nat$-polyregular function}. For instance, the polynomial $P(X) \defined
-X(X+1)/2$. Historically, the definition used polynomials with coefficients in
-$\Rat$, which is why we use such polynomials here, even though we know that all
-of those polynomials will actually be \kl{integer binomial polynomials} thanks
-to \cref{integer-binomial-polynomials:cor}.
+\kl{$\Nat$-polyregular function}. Recall that the polynomial $P(X) \defined
+X(X-1)/2$ can be computed by a \kl{$\Nat$-polyregular function}. Historically,
+the definition used polynomials with coefficients in $\Rat$, which is why we
+use such polynomials here, even though we know that all of those polynomials
+will actually be \kl{integer binomial polynomials} thanks to
+\cref{integer-binomial-polynomials:cor}.
 
@@
     function $f_{a^n} \colon (\Sigma \setminus \set{a})^* \to \Rel$
-    that maps $u \in \Sigma \setminus \set{a}$ to $f(a^n u)$.
+    that maps $u \in (\Sigma \setminus \set{a})^*$ to $f(a^n u)$.
 

### 2024-08-15 parts/appendix.tex
enormous rewrite

@@
 %! lang = en-US
-\section{Proofs of section \ref{preliminaries:sec}}
-
-\begin{proof}[Proof of \cref{decidable-commutative-poly:lemma}, in the polyregular case]
-    Let $\Sigma$ be a finite alphabet, endowed with a
-    total ordering over its letters,
-    and let $f \colon \Sigma^* \to \Rel$ be a
-    \kl{$\Rel$-polyregular function}.
-    The map $\polysort \colon \Sigma^* \to \Sigma^*$
-    that sorts the letters in the input word $w$ according
-    to the chosen ordering is a \kl{polyregular function}
-    \cite[Proposition II.12]{CDTL23}.
-    As a consequence,
-    $(f \circ \polysort) \in \ZPoly$.
-    Finally, $f$ is \kl{commutative} if and only
-    if $f = f \circ \polysort$, which is decidable
-    in the case of $\ZPoly$
-    \cite[Corollary II.24]{CDTL23}, and more generally in the case 
-    of $\ZRat$ \cite[Corollary 3.6 p 38]{BERE10}.
-\end{proof}
-
-\begin{proof}[Proof of \cref{decidable-commutative-rat:lemma} in the general case]
-    Because $f$ is a \kl{$\Rel$-rational series},
-    there exists a $n \in \Nat$,
-    vectors $v_1$ and $v_2$,
-    and $n$ by $n$ integer matrices $\seqof{M_a}{a \in \Sigma}$,
-    such that
-    \begin{equation*}
-        \forall w \in \Sigma^*,
-        f(w) = {}^t v_1 \left(\prod_{i = 1}^{|w|} M_{w_i}\right) v_2 \quad .
-    \end{equation*}
-    To simplify notations, 
-    let us write $M_w \defined \prod_{i = 1}^{|w|} M_{w_i}$,
-    and 
-    $S \defined \setof{M_w}{w \in \Sigma^+}$ be the
-    semigroup generated by $\seqof{M_a}{a \in \Sigma}$.
-
-    The function $f$ is \kl{commutative} if and only if for all $u,v \in
-    \Sigma^+$, ${}^t v_1 (M_u M_v  - M_v M_u) v_2 = 0$. The above is a
-    polynomial equation involving the coefficients of $M_u$ and $M_v$.
-    Let us compute $\overline{S \times S}$ the Zariski closure of $S \times S$.
-    The above polynomial vanishes over $S \times S$ if and only
-    if it belongs to the Zariski closure, which is decidable
-    \cite[\textbf{WHERE}]{HOPW18}.
-\end{proof}
 
@@
 
-\textbf{Redo the proof}.
-\begin{proofof}[derivation-stabilises-correct:lem]
-    If $P$ is constant, then $P_2 = 0$, and $\Diff{K}{P_1} = 0$,
-    hence for $K = 0$, $Q \in \CorrectPoly$.
-
-    Otherwise, let $\alpha$ be the maximal absolute value 
-    of coefficients appearing in $P$,
-    and $D$ be the maximal number of divisors for
-    a monomial in $P$.
-    We define $K \defined \alpha \times D + 1$.
-
-    Let $\nu \colon \vec{X} \topartial \Nat$, we are going to prove that
-    $\restr{Q}{\nu}$ has \kl{non-negative} \kl{maximal monomials}. To that end,
-    let us consider $T$ such a maximal monomial, and let us write $T = \beta
-    T_1$ so that $T_1$ has coefficient $1$ and $\beta \in \Rel$. Because we
-    will need to normalize other monomials in this way, we define for the rest
-    of the proof the function $\mathsf{unit}$ that maps a monomial to the
-    monomial obtained by replacing its constant term by $1$ if the monomial is
-    non-zero, and leaving the monomial unchanged otherwise.
-
-    Let us write $\mathcal{D}_1$ to be the set of \kl{monomials} of
-    $\Diff{K}{P_1}$ that are \kl{divided} by $T$. Similarly, let us write
-    $\mathcal{D}_2$ for the set of monomials of $\translate{K}(P_2)$ that are
-    \kl{divided} by $T$. Thanks to \cref{derivation-monomials:fact}, we know
-    that every element of $\mathcal{D}_2$ divides some element in
-    $\mathcal{D}_1$. We write $\mathcal{D}_2^M$ for the set of monomials $S \in
-    \mathcal{D}_2$ such that $S$ \kl{divides} $M$.
-
-    Now, remark that the coefficient $\beta$ is obtained as follows:
-    \begin{align*}
-        \beta &= \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
-               + \sum_{M \in \mathcal{D}_2} \restr{M / T_1}{\nu} \\
-              &\geq \sum_{M \in \mathcal{D}_1} \restr{M / T_1}{\nu}
-              - \sum_{M \in \mathcal{D}_2} \left|\restr{M / T_1}{\nu}\right|
-              \\
-              &\geq
-              \sum_{M \in \mathcal{D}_1}
-              \left[
-              \restr{M / T_1}{\nu}
-              -
-              \sum_{S \in \mathcal{D}_2^M}
-              \left|\restr{S/T_1}{\nu}\right|
-              \right]
-              \\
-              &\geq
-              \sum_{M \in \mathcal{D}_1}
-              \restr{\mathsf{unit}(M)/T_1}{\nu}
-              \left[
-              K
-              - \alpha D
-              \right]
-              \\
-              &\geq 0 \quad . \qedhere
-    \end{align*}
-\end{proofof}
-
 \begin{proofof}[derivation-translation:lem]
@@
     $\translate{K}(\restr{P}{\nu})$ are exactly the \kl{maximal monomials} of
-    $\restr{P}{\nu}$ thanks to \cref{translation-maximal:fact}, and are
-    therefore \kl{non-negative}.
+    $\restr{P}{\nu}$, and are therefore \kl{non-negative}.
 
     Conversely, the implication \cref{d-t-correct:item} $\implies$
-    \cref{d-t-transl:item} is proven by induction on $\MaximalMonomials(P)$. If
-    $P$ is a constant polynomial, then $P = n$ for some $n \in \Nat$, and we
-    conclude that $\translate{0}(P) \in \Nat[\vec{X}]$. Otherwise, $P = P_1 +
-    P_2$ where $P_1$ is the sum of the \kl{maximal monomials} of $P$. Using
+    \cref{d-t-transl:item} is proven by induction on the number of
+    indeterminates of $P$. If $P$ is a constant polynomial, then $P = n$ for
+    some $n \in \Nat$, and we conclude that $\translate{0}(P) \in
+    \Nat[\vec{X}]$. Otherwise, $P = P_1 + P_2$ where $P_1$ is the sum of the
+    \kl{maximal monomials} of $P$. Using
     \cref{derivation-stabilises-correct:lem}, there exists a computable $K \in
     \Nat$ such that $Q \defined \Diff{K}{P_1} + \translate{K}(P_2)$ belongs to
-    $\CorrectPoly$. Since, $\MaximalMonomials(Q) \hoarele
-    \MaximalMonomials(P)$, we conclude by induction hypothesis that there
-    exists $L \in \Nat$ such that for every partial valuation $\nu \colon
-    \vec{X} \topartial \Nat$, $\translate{L}(\restr{Q}{\nu}) \in
-    \Nat[\vec{X}]$. Similarly, for all partial functions $\nu \colon \vec{X}
-    \topartial \set{0, \dots, K}$ fixing at least one indeterminate of $P$,
-    $\MaximalMonomials(\restr{P}{\nu}) \hoarele \MaximalMonomials(P)$, hence
-    there exists $L_\nu \in \Nat$ such that $\translate{L_\nu}(\restr{P}{\nu})
-    \in \Nat[\vec{X}]$. Let us define $L_m$ to be the maximum of $L$, and
-    $L_\nu$ where $\nu$ ranges over partial functions described above.
+    $\Nat[\vec{X}]$. Furthermore, for all partial functions $\nu \colon \vec{X}
+    \topartial \set{0, \dots, K}$ fixing at least one indeterminate of $P$, the
+    induction hypothesis yields a number $L_\nu \in \Nat$ such that
+    $\translate{L_\nu}(\restr{P}{\nu}) \in \Nat[\vec{X}]$. Let us define $L_m$
+    to be the maximum of $L_\nu$ where $\nu$ ranges over partial
+    functions described above.
 
@@
     In this case,
-    $\translate{K}(\restr{P}{\translate{K}(\mu)}
-    = \restr{\translate{K}(P)}{\mu}$
-    thanks to 
-    \cref{discrete-deriv-linear:fact}.
+    $\translate{K}(\restr{P}{\translate{K}(\mu)})
+    = \restr{\translate{K}(P)}{\mu}$.
     Remark that by definition,
@@
     $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
-    there exists an \kl{integer binomial polynomial}
-    $P_r \in \Rel[X_1, \dots, X_p]$,
-    obtained as
-    the difference of the two \kl{natural binomial polynomial} $P_r^+$ and $P_r^-$
+    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
     associated to $f_+$ and $f_-$,
@@
         a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
-        = P_r(X_1, \dots, X_p) \quad .
+        = F_r^+(X_1, \dots, X_p) - F_r^-(X_1, \dots, X_p) \quad .
     \end{equation*}
-    We conclude by selecting different values of $r$ and
+    Now, for large enough values of $x_1, \dots, x_p$,
+    this means that 
+    $f\left(
+        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right)$
+    coincides with an \kl{integer binomial polynomial},
+    and we can therefore conclude selecting different values of $r$ and
     setting some subsets of variables to be assigned to the constant $0$.
@@
 \end{proofof}
-
-\begin{proofof}[z-poly-commutative-encoding:remark]
-    We are going to prove that if $f \in \ZPoly[k]$ is polyblind, 
-    then the \cref{z-poly-commutative-encoding:conj} holds when allowing 
-    for non polyregular functions $\mathsf{enc}$ and $\mathsf{dec}$.
-
-    Let us consider $f \in \ZPoly[k]$ that is polyblind.
-    Using \cite[Theorem 6.12]{DOUE23},
-    we know that there exists a family of $\MSO$ formulas
-    $\seqof{\psi_{i,j}(x)}{1 \leq i,j \leq n}$,
-    and coefficients $\seqof{z_i}{1 \leq i \leq n}$,
-    such that:
-    \begin{equation*}
-        f = \sum_{i = 1}^n z_i \times \prod_{j = 1}^n \vcount{\psi_{ij}(x)}
-        \quad .
-    \end{equation*}
-
-    Let us now consider a monoid $M$ that recognizes all formulas $\psi_{ij}$
-    for $1 \leq i,j \leq n$ with a morphism $\mu \colon \Sigma^* \to M$. Let us
-    consider the map $\mathsf{t} \colon \Sigma^* \to (M^3)^*$, that associates
-    to a word $w \in \Sigma^*$ the word $\mathsf{t}(w) \defined \prod_{i =
-    1}^{|w|} ( \mu(w_{< i}), \mu(w_i), \mu(w_{> i}))$. That is, to every letter
-    one associates its left and right context in the monoid.
-
-    Let us define $\mathsf{s} \colon \Sigma^* \to (\Nat)^{M^3}$
-    by letting $\mathsf{s}(w)$ be the Parikh image of $\mathsf{t}$, that is,
-    counting the number of occurrences of every letter.
-    It is clear how to define the polyregular function $g \colon (\Nat)^{M^3} \to \Rel$
-    that makes the following diagram commute, at is it even a polynomial function
-    with $M^3$ indeterminates:
-    \begin{center}
-        \begin{tikzcd}
-            (M^3)^* 
-            \arrow{dr}[below,anchor=center,rotate=-39,yshift=-3pt]{\mathsf{count}}
-            & \Sigma^* \arrow{l}[above]{\mathsf{t}} \arrow{r}{f} \arrow{d}[left]{\mathsf{s}} & \Rel \\
-            & \Nat^{M^3} \arrow{ur}[below]{g} & \\
-        \end{tikzcd}
-    \end{center}
-    One problem is that the map $\mathsf{s}$ is not surjective, because of
-    potential dependencies between the types encountered in $w$.
-    Let us however notice that one can build a context-free grammar
-    recognizing exactly the image of $\mathsf{t}$. Namely,
-    \begin{align*}
-        S &\to S^{1,m,1} &  m \in M \\
-        S^{m_r, m, m_l} &\to (m_l, m, m_r) & \exists a \in \Sigma, \mu(a) = m \\
-        S^{m_r, m, m_l} &\to 
-        S^{m_r, m_x, m_y m_l}
-        S^{m_r m_x, m_y, m_l}
-                        & m_x m_y = m
-    \end{align*}
-
-    Using Parikh's Theorem \cite{PARI66},
-    we conclude that the image of $\mathsf{s}$ is a semilinear set,
-    i.e., that there exists a finite set $Q$,
-    numbers $\seqof{d_q}{q \in Q}$,
-    vectors $\seqof{v_{i,q}}{0 \leq i \leq d_q, q \in Q}$ in $\Nat^{M^3}$,
-    such that
-    \begin{equation*}
-        \mathsf{s}(\Sigma^*) =
-        \bigcup_{q \in Q}
-        \setof{
-            v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
-        }{
-            (\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}
-        } \quad .
-    \end{equation*}
-
-    We can therefore define a function $\mathsf{decode} \colon \Sigma^* \to
-    \sum_{q \in Q} \Nat^{d_q}$ that maps to a word $w$ to \emph{some} $q \in Q$
-    and parameters $(\lambda_1, \dots, \lambda_{d_q}) \in \Nat^{d_q}$ such that
-    $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$. By taming the
-    semilinear set prior to this step, we can ensure that the function is in
-    fact uniquely defined, this is done by computing an equivalent semilinear
-    set that is \emph{unambiguous} and \emph{proper} \cite{CHHA16}.
-    Similarly, let us define $\mathsf{encode} \colon \sum_{q \in Q} \Nat^{d_q}
-    \to \Sigma^*$ that maps an element $q \in Q$ with parameters $(\lambda_1,
-    \dots, \lambda_{d_q}) \in \Nat^{d_q}$ to \emph{some} word $w \in \Sigma^*$
-    such that $\mathsf{s}(w) = v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}$.
-    Notice that $\mathsf{decode} \circ \mathsf{encode} = \mathsf{id}$, by the
-    unicity of the representation in $\sum_{q \in Q} \Nat^{d_q}$.
-    We can define the map $\mathsf{repr} \colon \sum_{q \in Q} \Nat^{d_q} \to \Nat^{M^3}$
-    as follows:
-    \begin{equation*}
-        \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \defined
-        v_0 + \sum_{i = 1}^{d_q} \lambda_i v_{i,q}
-        \quad .
-    \end{equation*}
-    Finally, let us define for $q \in Q$ the map $g_q \colon \Nat^{d_q} \to \Rel$ as the polynomial
-    expression:
-    \begin{equation*}
-        g_q(\lambda_1, \dots, \lambda_{d_q}) \defined \mathsf{repr}(q, \lambda_1, \dots, \lambda_{d_q}) \quad .
-    \end{equation*}
-
-    We have proven that the following diagram commutes:
-    \begin{center}
-        \begin{tikzcd}
-            \Sigma^* \arrow{rr}{f} 
-                     \arrow{dr}{\mathsf{s}}
-                     \arrow[bend right=20]{dd}[rotate=90,anchor=center,yshift=4pt]{\mathsf{decode}}
-                     &        & \Rel \\
-                     & \Nat^{M^3} \arrow{ur}[below]{g} & \\
-            \sum_{q \in Q} \Nat^{d_q}
-            \arrow[bend right=20]{uu}[right,rotate=90,anchor=center,yshift=-4pt]{\mathsf{encode}}
-            \arrow{ur}[left]{\mathsf{repr}}
-            \arrow[bend right=45]{uurr}[right]{\seqof{g_q}{q \in Q}}
-                     & & \\
-        \end{tikzcd}
-    \end{center}
-
-    The only missing part of this proof is the fact that the maps $\mathsf{encode}$ and
-    $\mathsf{decode}$ can be computed as \kl{polyregular functions}. Notice
-    that \emph{mutatis mutandis}, the same proof scheme applies to $f \in
-    \ZSF[k]$, by assuming that the monoid is aperiodic, since the functions
-    constructed (apart from $\mathsf{encode}$ and $\mathsf{decode}$) will be
-    \kl{star-free}.
-\end{proofof}

### 2024-08-15 parts/conclusion.tex
gigantic rewrite
enormous rewrite

@@
 Let us conclude with a more general remark on the status of commutative input
-in the study of unary output polyregular functions. Notice that semantic
-characterizations of subclasses of $\ZPoly$ (\kl{combinatorial}, \kl{ultimately
-polynomial}) arise from pre-composition with a \kl{commutative}
-\kl{star-free polyregular function}, as formally stated in
-\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
-
-%Indeed, a suitable application of Parikh's Theorem \cite{PARI66}
-%guarantees the existence of the functions $\mathsf{enc}$ and $\mathsf{dec}$,
-%although they may fail to be  \kl{polyregular functions}.
+in the study of unary output polyregular functions. Throughout the paper, we
+have tried to highlight that results on polyregular functions often have the
+following form: a polyregular function $f \colon \Sigma^* \to \Gamma^*$
+satisfies property $X$ if and only if for all \kl{commutative} \kl{star-free
+polyregular function} $g \colon \Delta^* \to \Sigma^*$, $f \circ g$ satisfies
+$X$. Let us formally restate this for two earlier results in the case of
+\kl{$\Rel$-polyregular functions}
+in \cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
 
 \begin{lemma}
+    \proofref{pre-compose-growth-commut:lemma}
     \label{pre-compose-growth-commut:lemma}
@@
             $(f \circ h) \in \ZPoly[k\times l]$.
-    \proofref{pre-compose-growth-commut:lemma}
 \end{lemma}
@@
 \begin{lemma}
+    \proofref{pre-compose-sf-commut:lemma}
     \label{pre-compose-sf-commut:lemma}
@@
             $(f \circ h) \in \ZSF$.
-    \proofref{pre-compose-sf-commut:lemma}
 \end{lemma}
@@
 functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
-immediately, and $\NSF$ inside $\NPoly$ would immediately follow. On the one
-hand, one can guess a candidate function and check for equivalence, on the
+immediately, and $\NSF$ inside $\NPoly$ would also immediately follow. On the
+one hand, one can guess a candidate function and check for equivalence, on the
 other hand, one can guess a \kl{commutative} \kl{star-free polyregular

### 2024-08-15 parts/correction.tex
enormous rewrite

@@
 \AP
-The counter example provided by \cref{def:bad-polynomial} relies on the fact
+The counterexample provided by \cref{def:bad-polynomial} relies on the fact
 that $\CoveredPoly$ is not stable under fixing indeterminates, while
@@
 \begin{definition}
-    Let $\vec{X}$ be a finite tuple of indeterminates.
     The class $\intro*\CorrectPoly[\vec{X}]$ is the collection of
@@
 polynomials in $\CorrectPoly$ are \kl{non-negative}. We also remarked at the
-beginning of this section that our counter example provided in
+beginning of this section that our counterexample provided in
 \cref{thm:counter-example} is not in $\CorrectPoly$.
@@
 the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
-and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We believe that the formulation of such
-pumping arguments can be made clearer through some extra notations, with the
-added property of highlighting the central role of \kl{commutative} polyregular
-functions. Let us define that a function $q$ is a \intro{(noncommutative output)
-polynomial function} from $\Nat^p$ to $\Sigma^*$ whenever there exists words
-$\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words $u_1, \dots, u_p \in
-\Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0 \prod_{i = 1}^p u_i^{X_i}
-\alpha_i$. That is, $q$ is syntactically defined by a non-commutative
-polynomial function over the semi-ring $\Sigma^*$. Almost by definition,
-\kl{polynomial functions} are \kl{commutative} \kl{polyregular functions}.
+and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in
+\cref{n-poly-combinatorics:lem} a finer statement that has two main goals:
+first, it formulates the traditional \emph{pumping argument} as a composition
+property using \kl{(noncommutative output) polynomial functions}, allowing us
+to use the same formalism in \cref{k-combinatorial:def} and
+\cref{ultimately-polynomial:def} and highlight the role of \kl{commutative}
+\kl{polyregular functions} in the broader study of \kl{polyregular functions};
+then we also give a precise shape of the polynomials that arise from such
+\emph{pumping arguments}, which was lacking in former similar statements where
+$\Rat[X_1, \dots, X_n]$ was used as an over approximation.
 
-\AP There is still a minor detail that remains to be smoothened before proving
-our combinatorial \cref{n-poly-combinatorics:lem}, which is that some
-polyregular functions can exhibit a limited form of division. For instance, the
-function computing the sum of the lengths of the suffixes of a word $w$
-actually computes $\card{w} \times (\card{w} - 1) / 2$. This will be at odds
-with our restriction to polynomials with coefficients in $\Rel$. However, we
-argue that up to a suitable change of basis for the ring of polynomials, all
-coefficients will actually be in $\Nat$. This relatively simple trick works by
-considering as ``basic polynomials" the \intro{binomial monomial} of the form
-$\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural numbers. Beware that
-$\binom{X - \ell}{k}$ is defined to be $0$ when $X \leq \ell$, and is therefore
-a ``true polynomial" only when $X \geq \ell$. This is why we introduce the
-notion of \intro{natural binomial polynomials} as $\Nat$-linear combinations of
-products of \kl{binomial monomials}.
+\AP To address our first goal, let us define that a function $q$ is a
+\intro{(noncommutative output) polynomial function} from $\Nat^p$ to $\Sigma^*$
+whenever there exists words $\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words
+$u_1, \dots, u_p \in \Sigma^*$, such that $q(X_1, \dots, X_p) = \alpha_0
+\prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
+non-commutative monomial over the semi- $\Sigma^*$. Almost by
+definition, \kl{polynomial functions} are \kl{commutative} \kl{polyregular
+functions}.
+
+\AP Remark that our second goal is at odds with our restriction to polynomials
+with coefficients in $\Rel$, because of the polynomial $X(X-1)/2$
+(\cref{q-polynomial-n-poly:ex}). However, we argue that up to a suitable change
+of basis for the ring of polynomials, all coefficients will actually be in
+$\Nat$. This relatively simple trick works by considering as ``basic
+polynomials" the \intro{binomial function} of the form $\binom{X - \ell}{k}$,
+where $\ell$ and $k$ are natural numbers. We extend this to \intro{natural
+binomial functions} that are obtained by considering $\Nat$-linear combinations
+of products of \kl{binomial functions}, that is, we consider functions that
+have the following shape: $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i \prod_{j =
+1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is
+defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial. For
+this reason, we also introduce its ``polynomial counterpart" $\intro*\pbinom{X
+- \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
+\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
+as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
+the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
+- p_{i,j}}{k_{i,j}}$.
+
+\begin{lemma}
+    \label{binomial-function-star-free:lem}
+    Let $F$ be a \kl{binomial function} from $\Nat^k$ to $\Nat$.
+    Then it is \kl{represented} by a \kl{star-free polyregular function}.
+\end{lemma}
+\begin{proof}
+    Because of the stability properties of 
+    $\NSF$ (\cref{stability-polyregular:lemma}), we only need to 
+    check that given $r,s \in \Nat$,
+    the function $x \mapsto \binom{x - r}{s}$ is \kl{represented}
+    by a function $f_{r,s} \in \NSF$.
+    Let us prove it when $r = 0$, since the other functions 
+    can be obtained by translating $f_{r,s}$.
+    Remark that 
+    $\binom{x}{s} = \sum_{u_1 \cdots u_{s+1} = a^x} 1$,
+    which is clearly expressible as a \kl{star-free polyregular function}.
+\end{proof}
 
@@
     $q \colon \Nat^p \to \Sigma^*$,
-    there exists a computable \kl{natural binomial polynomial} $P$
+    there exists a computable \kl{natural binomial function} $F$
     such that:
@@
         =
-        P
+        F
         \quad 
         \text{ over } (\Nat_{\geq 1})^p
+        \quad .
     \end{equation*}
@@
     i.e., a number such that for all $x \in M$,
-    $x^{\omega+1} = x^\omega$.
+    $(x^{\omega})^2 = x^{\omega}$.
     We will prove by induction on $k \in \Nat_{\geq 1}$
@@
     $\pi \colon M^k \to \Nat$,
-    the lemma holds for the function 
+    the lemma holds with $s = 0$ for the function 
     $f = \pi^\dagger$, i.e., such that
@@
         &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
-        & \text{ idempotent and } n_i \geq 1
+        & \text{ idempotent and } n_i \geq 1 \\
         &= c \geq 0
@@
     and a partition of
-    $u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
+    $)u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
     \alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
@@
     where we have used the notation 
-    $v\mid_{[-r:]}$ to denote the word obtained from $v$ 
+    $(v)_{[-r:]}$ to denote the word obtained from $v$ 
     by keeping only the last $r$ letters, in a sort of Pythonic syntax.
@@
     one can define the \kl{polynomial functions}
-    $s_{\ell,b,r} \colon \Nat^{\ell - 1} \to \Sigma^*$
+    $s_{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
     and
-    $q_{\ell,b,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
+    $q_{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
     so that
-    $q(X_1, \dots, X_p) = s_{\ell, b, r}(X_1, \dots, X_{\ell-1}) 
-    q_{\ell, b, r}(X_{\ell}, \dots, X_p)$
-    under the assumption that $X_1 \geq b$:
+    $q(X_1, \dots, X_p) = s_{\ell,r,b}(X_1, \dots, X_{\ell-1}) 
+    q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
+    under the assumption that $X_\ell \geq b$:
     \begin{equation*}
-        q_{\ell,b,r}(X_{1}, \dots, X_{p - \ell + 1})
+        q_{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
         \defined
-        u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times (X_1 - b)}
+        u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times X_1}
         \alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
@@
     Similarly, let us define 
-    $s_{\ell, b, r}(X_1, \dots, X_{\ell - 1})$
+    $s_{\ell, r, b}(X_1, \dots, X_{\ell - 1})$
     as 
     $\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1} 
-    u_\ell^{\omega b} u_{\ell}^\omega \mid_{[1:r]}$,
-    Remark that $q_{\ell,b,r}$ has all of its indeterminates
+    u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}$,
+    Remark that $q_{\ell,r}$ has all of its indeterminates
     multiplied by $\omega$, so that the induction hypothesis 
@@
 
-    The key remark is that $\mu(s_{\ell,b,r}(X_1, \dots, X_{\ell - 1}))$ is a
-    constant function from $\Nat^{\ell-1}$ to $M$, and that its value is also
-    constant when $b$ varies among numbers greater than $0$. This holds with
+    The key remark is that $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
+    constant function from $\Nat^{\ell-1}$ to $M$, and that its value
+    only depends on whether $b = 0$ or $b > 0$.
+    This holds with
     the exact same proof as the base case of this induction. Let us call
     $m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
-    obtained by $\mu(s_{\ell,b,r}(X_1, \dots, X_{\ell-1}))$ respectively when
+    obtained by $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
     $b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
     \to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
-    \pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$ and similarly for $\pi_{\ell,r,1}
-    \colon M^k \to \Nat$.
+    \pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$.
 
@@
     In the following equations, we omit the precise
-    range for the sum over $(\ell, b, r)$ to save space.
+    range for the sum over $(\ell, r, b)$ to save space.
     Similarly, we will use 
     $q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
-    \defined q_{\ell,b,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
+    \defined q_{\ell,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
     at the end of the computation.
@@
         &= \sum_{(\ell, b, r)}
-           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
-           \pi(\mu(s_{\ell,b,r}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
+           \pi(\mu(s_{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
         \\
@@
            \sum_{(\ell, 0, r)}
-           \sum_{v_2 \dots v_{k+1} = q_{\ell,0,r}(n_\ell, \dots, n_p)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
            \pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
@@
            \sum_{(\ell, b > 0, r)}
-           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
            \pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
@@
            \sum_{(\ell, 0, r)}
-           \sum_{v_2 \dots v_{k+1} = q_{\ell,0,r}(n_\ell, \dots, n_p)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
            \pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
@@
            \sum_{(\ell, b > 0, r)}
-           \sum_{v_2 \dots v_{k+1} = q_{\ell,b,r}(n_\ell, \dots, n_p)}
+           \sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell-b, \dots, n_p)}
            \pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
@@
            \sum_{0 \leq r < \omega \times \card{u_\ell}}
-           (\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell, \dots, n_p))
+           (\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell-b, \dots, n_p))
            \\
@@
     Now, using the induction hypothesis,
-    we have \kl{natural binomial polynomials} $P_{\ell,r,0}$ and $P_{\ell,r,1}$
+    we have \kl{natural binomial functions} $F_{\ell,r,0}$ and $F_{\ell,r,1}$
     satisfying 
@@
        (\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
-       = P_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
+       = F_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
        \quad 
@@
         (\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
-        = P_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
+        = F_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
        \quad 
@@
 
-   It is an easy and folklore fact that \kl{natural binomial polynomials} are
-   stable under the operation $\Sigma_Y \colon Q(X_1,\dots,X_n,Y) \mapsto \sum_{0
-   < b < Y} Q(X_1, \dots, X_n, b)$, and by pre-composition with 
-   a multiplication by a constant $\omega \geq 1$.
-
-   This allows us to conclude by defining 
+   It is a well-known equality that for all $s,n \in \Nat$,
+   $\sum_{0 < i < n} \binom{s}{i} = \binom{s+1}{n}$.
+   As a consequence, \kl{natural binomial functions} are
+   stable under the following summation operation:
+   $\Sigma_Y \colon Q(x_1,\dots,x_n,y) \mapsto \sum_{0 < b < y} Q(x_1, \dots, x_n, b)$.
+   This allows us to conclude by defining:
    \begin{equation*}
-       P(X_1, \dots, X_p) \defined
+       F(x_1, \dots, x_p) \defined
        \sum_{1 \leq \ell \leq p} \sum_{0 \leq r < \omega \card{u_\ell}}
-       P_{l,r,0}(\omega X_{\ell}, \dots, \omega X_p) + \Sigma_Y P_{l,r,1}(\omega Y,
-       \omega X_{\ell+1}, \dots, \omega X_p)
+       F_{l,r,0}(x_{\ell}, \dots, \omega x_p) + \Sigma_Y F_{l,r,1}(Y,
+       x_{\ell+1}, \dots, x_p)
        \quad .
    \end{equation*}
-   The latter being a \kl{natural binomial polynomial} satisfying
+   The latter being a \kl{natural binomial function} satisfying
    the desired equality.
+
 \end{proof}
@@
 of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial.
-Let us remark that the \kl{maximal monomials} of the polynomial $\binom{X -
-p}{k}$ have coefficient $1/k!$, and are therefore \kl{non-negative}.
+We can trade-off this multiplicative factor for a constant term 
+addition under the extra assumption that the function 
+is \kl{star-free polyregular}, as described in the following lemma.
+
+\begin{lemma}
+    \label{n-sf-combinatorics:lem}
+    Let $f$ be a \kl{star-free $\Nat$-polyregular} function. 
+    There exists a computable $s \in \Nat_{\geq 1}$
+    such that for all \kl{polynomial functions}
+    $q \colon \Nat^p \to \Sigma^*$,
+    there exists a computable \kl{natural binomial function} $F$
+    such that:
+    \begin{equation*}
+        f \circ q(X_1 + s, \dots, X_p + s)
+        =
+        F
+        \quad 
+        \text{ over } (\Nat_{\geq 1})^p
+        \quad .
+    \end{equation*}
+\end{lemma}
+\begin{proof}[Proof Sketch]
+    We redo the proof of \cref{n-poly-combinatorics:lem}, with the extra
+    assumption that the monoid $M$ is \kl(monoid){aperiodic}, hence that there
+    exists $s \in \Nat$ such that $x^{s+1} = x^s$ for all $x \in M$. As a
+    consequence, we can define $\omega = 1$, and
+    translate inputs by $s$, instead of multiplying them by $\omega$.
+\end{proof}
+
+Let us also address the fact that the distinction between \kl{natural binomial
+functions} and \kl{natural binomial polynomials} vanishes when we are
+interested in polynomials. As a consequence, to a \kl{natural binomial
+function} corresponds exactly one \kl{natural binomial polynomial} that
+coincides with it on sufficiently large inputs.
+
+\begin{lemma}
+    \label{natural-binomial-polynomial:lem}
+    Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial,
+    $\ell \in \Nat$, 
+    and let $f$ 
+    be a 
+    \kl{natural binomial function} such that
+    $P = f$ over $(\Nat_{\geq \ell})^n$. If
+    $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i 
+    \prod_{j = 1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$,
+    then
+    $P =
+    \sum_{i = 1}^n n_i 
+    \prod_{j = 1}^k \pbinom{X_j - p_{i,j}}{k_{i,j}}$.
+\end{lemma}
+\begin{proof}
+    Let $m$ be the maximal value of all $p_{i,j}$, $k_{i,j}$, and $\ell$.
+    Over $(\Nat_{\geq 2m})^k$, $f$ coincides with
+    the desired polynomial. Now, because $f$,$P$, and the desired
+    polynomial coincide
+    over a large enough subset of $\Nat^k$,
+    this means that $P$ has the desired shape.
+\end{proof}
 
@@
     $P \in \CorrectPoly$.
-    \proofref{n-rat-correct:lem}
 \end{corollary}
@@
     there exists a number $\omega \in \Nat_{\geq 1}$
-    and \kl{natural binomial polynomial} $Q$
+    and \kl{natural binomial function} $Q$
     such that
@@
     \end{equation*}
-    We conclude that $P(\omega X_1, \dots, \omega X_p) = Q(X_1, \dots, X_p)$
-    as polynomials,
-    and in particular that
+    As a consequence of \cref{natural-binomial-polynomial:lem}, $P(\omega X_1, \dots, \omega X_p)$ equals the corresponding 
+    \kl{natural binomial polynomial}.
+    Let us remark that the (unique) \kl{maximal monomial} of the 
+    \kl{binomial monomial} $\pbinom{X -
+    p}{k}$ has coefficient $1/k!$, and is therefore \kl{non-negative}.
+    We conclude that
     the \kl{maximal monomials} of 
-    $P$ are \kl{non-negative}.
+    $P(\omega X_1, \dots, \omega X_p)$ are \kl{non-negative},
+    and since $\omega \geq 1$, we conclude that
+    the \kl{maximal monomials} of $P$ have \kl{non-negative} coefficients.
 
@@
 
-\AP
-We have proven that every \kl{$\Nat$-rational polynomial} is in $\CorrectPoly$,
-and the rest of this section is devoted to proving the converse implication.
-The core of the upcoming \cref{lem:correct-to-n-rat} leverages a notion of
-\intro{discrete derivative} to perform an induction on the \kl{maximal monomials}.
-This notion of derivation is built by translating the domain of the polynomial.
-To that end, let us write $\intro*\translate{K}$ for the \intro{translation function}
-that maps a polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 +
-K, \dots, X_n + K)$.
+\AP We have proven that every \kl{$\Nat$-rational polynomial} is in
+$\CorrectPoly$, and the rest of this section is devoted to proving the converse
+implication. The core of the upcoming \cref{lem:correct-to-n-rat} leverages a
+notion of \intro{discrete derivative} which is built by translating the domain
+of the polynomial. To that end, let us write $\intro*\translate{K}$ for the
+\intro{translation function} that maps a polynomial $P \in \Rel[X_1, \dots,
+X_n]$ to the polynomial $P(X_1 + K, \dots, X_n + K)$.
 
@@
 \AP The key lemma of this section is \cref{lem:correct-to-n-rat}, which is
-proved by induction on the \kl{maximal monomials} of a polynomial $P$. The core
-combinatorial argument allowing us to perform the induction is
-\cref{derivation-stabilises-correct:lem}, which is proved by a straightforward
-binomial expansion. Using this lemma, the induction is performed as follows:
-after decomposing a polynomial $P$ into a sum of \kl{maximal monomials} $P_1$
-and the rest $P_2$, notice that for all $K \in \Nat$, $\translate{K}(P) = P_1 +
-\Diff{K}{P_1} + \translate{K}(P_2)$. Then, because $P \in \CorrectPoly$, $P_1$
-is a \kl{$\Nat$-rational polynomial}. Furthermore, using a large enough value
-of $K$, the polynomial $\Diff{K}{P_1} + \translate{K}(P_2)$ has lower degree
-than $P$, and belongs to $\CorrectPoly$. As a consequence, we conclude that
-$\translate{K}(P)$ is an \kl{$\Nat$-rational polynomial}. To conclude that $P$
-itself is a \kl{$\Nat$-rational polynomial}, it suffices to apply the induction
-hypothesis to $\restr{P}{\nu}$ where $\nu$ is a partial function from $\vec{X}$
-to $\set{0, \dots, K}$ fixing at least one variable. It is then just a matter
-of combining these functions to obtain a \kl{$\Nat$-polyregular function} that
-\kl{represents} $P$. Notice that in \cref{lem:correct-to-n-rat}, polynomials
-can be represented using \kl{star-free $\Nat$-polyregular functions}, which is
-a stronger statement than the one in \cref{karh:thm}, but is in line with the
-idea that polynomials are not periodic functions (this will be made precise in
-\cref{zsf-nsf:conjecture}).
-
-As one would expect, the \kl{discrete derivatives} are linear operations on
-polynomials, and they commute with \kl{translation operators}. However, the
-\kl{translation operators} do not perfectly commute with partial applications
-$\restr{\cdot}{\nu}$.
-
-\begin{fact}
-    \label{discrete-deriv-linear:fact}
-    For all tuples $\vec{X}$ of indeterminates,
-    for all $K \in \Nat$, $L \in \Rel$, for all partial functions
-    $\nu \colon \vec{X} \topartial \Nat$:
-    \begin{enumerate}
-        \item The maps $\translate{K} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
-        $\Diff{K}{ \cdot } \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$,
-        and
-        $\restr{\cdot}{\nu} \colon \Rel[\vec{X}] \to \Rel[\vec{X}]$
-            are linear operators,
-        \item $\Diff{K}{ \cdot } \circ \translate{L}
-            = \translate{L} \circ \Diff{K}{\cdot}$,
-        \item $\restr{\cdot}{\nu} \circ \translate{L}
-            = \translate{L} \circ \restr{\cdot}{\translate{L}(\nu)}$.
-    \end{enumerate}
-\end{fact}
-
-\AP Let us now introduce the ordering over which the induction of
-\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
-set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
-order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
-\kl{well-quasi-ordering} that is isomorphic (once quotiented to be a partial
-ordering) to $\Nat^k$ with the product ordering \cite[Dickson’s Lemma]{SCSC12}.
-We endow $\Pfin(\Monomials)$ with the \intro{Hoare ordering}, that is, $S_1
-\hoareleq S_2$ whenever for all \kl{monomials} $M_1 \in S_1$, there exists a
-monomial $M_2 \in S_2$, such that $M_1$ \kl{divides} $M_2$. The set
-$(\Pfin(\Monomials), \hoareleq)$ remains a \kl{well-quasi-ordering}, and is in
-particular well-founded \cite[Hoare quasi-ordering]{SCSC12}.
+proved by induction on the number of indeterminates of a given polynomial $P$.
+Notice that in \cref{lem:correct-to-n-rat}, polynomials can be represented
+using \kl{star-free $\Nat$-polyregular functions}, which is a stronger
+statement than the one in \cref{karh:thm}, but is in line with the idea that
+polynomials are not periodic functions (this will be made precise in
+\cref{zsf-nsf:conjecture}). In order to prove that result, we use the
+combinatorial \cref{derivation-stabilises-correct:lem} that allows us to
+transform a polynomial $P \in \CorrectPoly$ into a polynomial in
+$\Nat[\vec{X}]$ through a well-chosen translation of the indeterminates.
 
@@
     For any other monomial of $P_2$, by the maximality of $P_1$,
-    it strictly divides some monomial of $P_1$,
+    it strictly divides some monomial of $P_1$, and
     equals some monomial of $\Diff{K}{P_1}$. Because every monomial
     of $\Diff{K}{P_1}$ has a coefficient that is a multiple of $K = \alpha \times D$, we can
-    match every monomial of $P_2$ to $\alpha$ copies of a unitary
-    monomial appearing in $\Diff{K}{P_1}$, and because $\alpha$ is large enough
+    compensate every monomial of $P_2$ by a monomial of $\Diff{K}{P_1}$.
+    Therefore,
     we conclude that 
@@
     \label{lem:correct-to-n-rat}
-    Let $\vec{X}$ be a tuple of indeterminates,
-    and let $P \in \Rel[\vec{X}]$.
+    Let $P \in \Rel[\vec{X}]$.
     If $P \in \CorrectPoly$, then $P$ is \kl{represented}
@@
 \cref{corrected-version:thm}, which corrects and generalizes \cref{karh:thm}.
+\begin{example}
+    \label{bad-poly-translate:ex}
+    The polynomial $\BadPoly$ is not a 
+    \kl{$\Nat$-rational polynomial},
+    but is \kl{non-negative} and satisfies
+    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
+\end{example}
 
@@
     \label{derivation-translation:lem}
-    Let $\vec{X}$ be a tuple of indeterminates,
-    and let $P \in \Rel[\vec{X}]$.
+    \proofref{derivation-translation:lem}
+    Let $P \in \Rel[\vec{X}]$.
     There exists a computable number $K \in \Nat$
@@
     In particular, the above properties are decidable.
-    \proofref{derivation-translation:lem}
 \end{lemma}
-
-\begin{example}
-    \label{bad-poly-translate:ex}
-    The polynomial $\BadPoly$ is not a 
-    \kl{$\Nat$-rational polynomial},
-    but is \kl{non-negative} and satisfies
-    $\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
-\end{example}
 
@@
     \label{corrected-version:thm}
-    Let $\vec{X}$ be a tuple of indeterminates,
-    and let $P \in \Rel[\vec{X}]$.
+    Let $P \in \Rel[\vec{X}]$.
     The following are equivalent:
     \begin{enumerate}
-        \item \label{corrected-0:item} $P$ is a \kl{natural binomial polynomial},
         \item \label{corrected-1:item} $P \in \CorrectPoly$,
@@
     procedure.
+
 \end{proof}
@@
 \cref{thm:counter-example} uses three indeterminates, and this is not a
-coincidence: in the particular case of at most two indeterminates, the classes
+coincidence: in the particular cases of one or two indeterminates,  the classes
 $\CorrectPoly$ and $\CoveredPoly$ coincide. In particular, the examples
@@
 
+Let us conclude this section by considering polynomials with coefficients in
+$\Rat$, and justify that all the combinatorial work has already happened in
+$\Rel$ and $\Nat$.
+
 \begin{corollary}
@@
     \begin{enumerate}
-        \item $P$ is a \kl{$\Nat$-rational polynomial},
-        \item $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
-        \item $\alpha P \in \CorrectPoly$,
+        \item \label{rat-npoly-0:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
+        \item \label{rat-npoly-1:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
+        \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
+        \item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
             and for all $\vec{x} \in \Nat^{\vec{X}}$,
@@
 \begin{proof}
-    If $P$ is a \kl{$\Nat$-rational polynomial}, then ...
+
+    Proof of \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}.
     If $\alpha P \in \CorrectPoly$, then  it is an \kl{$\Nat$-rational polynomial}.
@@
     We conclude that $P$ is \kl{represented} by an \kl{$\Nat$-polyregular function}.
+
+    Proof of \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-0:item}.
+    If $P$ is a \kl{natural binomial polynomial} with $P(0,\dots,0) \geq 0$,
+    then it is \kl{non-negative}.
+
+    Proof of \cref{rat-npoly-0:item} $\implies$ \cref{rat-npoly-3:item}.
+    If $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
+    then for large enough values, it is a \kl{natural binomial function},
+    hence computable, and for small values we can hardcode.
+
+    Proof of \cref{rat-npoly-2:item} $\implies$ \cref{rat-npoly-4:item}.
+    If $P$ is a \kl{$\Nat$-rational polynomial}, then $\alpha P$ is one too,
+    and in particular $\alpha P$ belongs to $\CorrectPoly$.
+
+    Proof of \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}.
+    This is because $\NSF \subseteq \NPoly$.
+
+    Proof of \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-0:item}.
+    Because $P$ is computed by a function $f \in \NSF$,
+    one can leverage \cref{n-sf-combinatorics:lem}
+    to conclude that $P$ is actually a \kl{natural binomial polynomial}.
 \end{proof}
 
-\AP As a consequence, we can also fully characterize polynomials that can be
-computed using \kl{$\Rel$-rational series}, as they are precisely those
-computable by a difference of two \kl{$\Nat$-rational polynomials}. This drives
-us to introduce the class of \intro{integer binomial polynomials}, which are
-obtained by $\Rel$-linear combinations of \kl{binomial monomials} of the form
-$\binom{X - p}{k}$ for some $k,p \in \Nat$.
+\AP Notice that we can also fully characterize polynomials that can be computed
+using \kl{$\Rel$-rational series}, as they are precisely those computable by a
+difference of two \kl{$\Nat$-rational polynomials}. This drives us to introduce
+the class of \intro{integer binomial polynomials}, which are obtained by
+$\Rel$-linear combinations of products of \kl{binomial monomials} of the form
+$\pbinom{X - p}{k}$ for some $k,p \in \Nat$.
 
@@
     Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
-    if and only if $P$ is a \kl{integer binomial polynomial}.
+    if and only if $P$ is an \kl{integer binomial polynomial}.
 \end{corollary}
+\begin{proof}
+    Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
+    \kl{$\Nat$-polyregular functions}.
+    Let us argue that $f_+$ and $f_-$ are \kl{star-free polyregular functions},
+    because they are \kl{ultimately polynomial}. (todo later).
+    As a consequence, one can apply \cref{n-sf-combinatorics:lem}
+    to conclude that $P$ coincides with an 
+    \kl{integer binomial polynomial} over large enough inputs,
+    hence that $P$ is such a polynomial.
+\end{proof}

### 2024-08-15 parts/counter.tex
enormous rewrite

@@
 polynomial}. The analogue notion of \intro{$\Rel$-rational polynomials} is not
-of particular interest, since every polynomial $P$ is \kl{represented} by a
-\kl{$\Rel$-polyregular function}. Our goal is to decide if a
-\kl{$\Rel$-rational polynomial} (so, any polynomial) is an \kl{$\Nat$-rational
-polynomial}, and if so, to compute an \kl{$\Nat$-polyregular function} that
-\kl{represents} it. This particular problem was claimed to be solved in
-\cite{KARH77}, using the \cref{karh:def} to characterize \kl{$\Nat$-polyregular
-functions}, as restated in \cref{karh:thm}.
+of particular interest, since every polynomial $P \in \Rel[X_1, \dots, X_n]$ is
+\kl{represented} by a \kl{$\Rel$-polyregular function}. Our goal is to decide
+if a \kl{$\Rel$-rational polynomial} (so, any polynomial) is an
+\kl{$\Nat$-rational polynomial}, and if so, to compute an
+\kl{$\Nat$-polyregular function} that \kl{represents} it. This particular
+problem was claimed to be solved in \cite{KARH77}, using the \cref{karh:def} to
+characterize \kl{$\Nat$-rational polynomials}, as restated in \cref{karh:thm}.
 
@@
     \label{n-poly-n-poly:example}
+    \proofref{n-poly-n-poly:example}
     Let $P \in \Nat[\vec{X}]$. Then, $P$
     is an \kl{$\Nat$-rational polynomial}.
-    \proofref{n-poly-n-poly:example}
 \end{lemma}
@@
     $X_i$ for some $1 \leq i \leq n$)
-    by products and sums. Because the basic functions are
-    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions,
+    by products and sums (\cref{stability-polyregular:lemma}). Because the basic functions are
+    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions (see
+    \cref{size-of-word-nsf:ex}),
     that are closed under these operations, we conclude.
@@
     \label{negative-but-npoly:ex}
+    \proofref{negative-not-nrat:ex}
     The polynomials $X$, $X^2 + 3$,
@@
     not an \kl{$\Nat$-rational polynomial}.
-    \proofref{negative-not-nrat:ex}
 \end{example}
@@
     \label{karh:def}
-    Let $\vec{X}$ be a finite tuple of indeterminates.
     The class $\intro*\CoveredPoly[\vec{X}]$
@@
 
+\AP
 Let us now design a counterexample to \cref{karh:thm} by suitably tweaking
@@
 \kl{non-negative}, but also belongs to $\CoveredPoly$.
-
-\begin{definition}
 \label{def:bad-polynomial}
 We define $\intro*\BadPoly(X,Y,Z) \defined Z (X + Y)^2 + 2 (X - Y)^2$.
-\end{definition}
 

### 2024-08-15 parts/intro.tex
enormous rewrite

--- 2024-07-05 parts/intro.tex
@@
 as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
-with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$
-\kl{for star-free polyregular functions} with unary output)
-\cite{DOUE21,DOUE22}. From this point of view, $\NPoly$ is a subclass of a
-larger and well-studied class of functions called (non-commutative)
-\intro{rational series} \cite{REUT80,BERE88,BERE10}. It has been noticed in
-\cite{CDTL23} that \kl{$\Nat$-polyregular functions} correspond to a natural
-subclass of \intro{$\Nat$-rational series} ($\intro*\NRat$), another model of
-computation that aims at generalizing the notion of regular language to
-functions with outputs in a semiring \cite{BERE88,BERE10}. It was frequently
-rediscovered in the literature that \kl{$\Nat$-polyregular functions} are a
+with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$ for
+\kl{star-free polyregular functions} with unary output) \cite{DOUE21,DOUE22}.
+From this point of view, $\NPoly$ is a subclass of a larger and well-studied
+class of functions called (non-commutative) \kl{rational series}
+\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
+\kl{$\Nat$-polyregular functions} correspond to a natural (strict) subclass of
+\kl{$\Nat$-rational series} ($\NRat$), another model of computation that aims
+at generalizing the notion of regular language to functions with outputs in a
+\emph{commutative} semiring \cite{BERE88,BERE10}.\footnote{In particular, the
+    notion of \kl{rational series} with output in $\Gamma^*$ is not well-defined.} It was frequently rediscovered in the literature that
+\kl{$\Nat$-polyregular functions} (although not often called this way) are a
 subclass of \kl{rational series} \cite{SCHU62,KRRC13,CDTL23}.
@@
 functions} that are furthermore \kl{commutative}, that is where the output of
-the functions remains unchanged when the letters of the input word are permuted.
-In this restricted setting, we answer positively to \cite[Conjecture
-7.61]{DOUE23} (restated in \cref{zsf-nsf:conjecture}), and prove that the
+the functions remains unchanged when the letters of the input word are
+permuted. In this restricted setting, we answer positively to \cite[Conjecture
+7.61]{DOUE23} (stating that $\NSF = \NPoly \cap \ZSF$), and prove that the
 membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
@@
 
+As a byproduct of our fine-grained of (commutative) $\Nat$ and $\Rel$
+polyregular functions, we fully characterize the polynomials that they can
+express. Note that even \kl{$\Nat$-polyregular functions} can compute
+polynomials with coefficients in $\Rat$, such as the polynomial $X(X-1)/2$ (see
+\cref{q-polynomial-n-poly:ex}). However, not all polynomials $P \in
+\Rat[\vec{X}]$ can be expressed using polyregular functions, and we prove that
+it is the case only for those obtained as $\Nat$ (resp. $\Rel$) linear
+combinations of products binomial coefficients $\binom{X_1 - p_1}{k_1} \times
+\cdots \times \binom{X_n - p_n}{k_n}$.\footnote{For the reader perplexed by the
+    fact that $\binom{X_n - p_n}{k_n}$ is \emph{not} a polynomial, we redirect to page 
+\kpageref{binomial monomial}.}
+
 \subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
-the multiple characterizations of \kl{$\Nat$-polyregular functions} (resp.
-$\Rel$-polyregular functions), show that one can decide if a function $f \in
-\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
-the basic definitions used on multivariate polynomials. In
-\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
-\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
-\cref{thm:counter-example}, and an alternative characterization in
-\cref{corrected-version:thm}. Finally, in \cref{beyond-polynomials:sec}, we leverage the
-results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
-\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
-\cref{zsf-nsf:conjecture}
-\cite[Conjecture 7.61]{DOUE23} 
-holds for \kl{commutative} functions
+\kl{$\Nat$-polyregular functions} (resp. $\Rel$-polyregular functions), show
+that one can decide if a function $f \in \ZPoly$ is \kl{commutative}
+(\cref{decidable-commutative-poly:lemma}), and give the basic definitions used
+on multivariate polynomials. In \cref{polynomials:sec}, we focus on
+\emph{polynomials}, by introducing the \cref{karh:thm} of \cite{KARH77},
+providing a counter-example in \cref{thm:counter-example}, and an alternative
+characterization in \cref{corrected-version:thm}. Also in
+\cref{polynomials:sec}, we provide characterizations of $\Rat$-polynomials that
+can be expressed respectively by \kl{$\Nat$-polyregular functions}
+(\cref{corrected-version:thm} \cref{corrected-0:item}) and
+\kl{$\Rel$-polyregular functions} (\cref{integer-binomial-polynomials:cor}).
+Finally, in \cref{beyond-polynomials:sec}, we leverage the results over
+polynomials to decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
+(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture}
+\cite[Conjecture 7.61]{DOUE23} holds for \kl{commutative} functions
 (\cref{zsf-npoly-nsf:thm}).

### 2024-08-15 parts/prelim.tex
enormous rewrite

@@
 
-\AP
-We assume that the reader is familiar with the basics of automata theory, in
-particular the notions of monoid morphisms, idempotents in monoids, monadic
+\AP We assume that the reader is familiar with the basics of automata theory,
+in particular the notions of monoid morphisms, idempotents in monoids, monadic
 second-order ($\MSO$) logic and first-order ($\FO$) logic over finite words
@@
 paper, let us recall that a monoid $M$ is \intro(monoid){aperiodic} whenever
-for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.
+for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$. If the
+monoid $M$ is finite, this $n$ can be uniformly chosen for all elements in $M$.
 
@@
 
-\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
-equivalent definition of \kl{$\Nat$-polyregular functions} that is more
-combinatorial in nature, and is similar in shape to the \emph{finite counting
-automata} introduced by \cite{SCHU62}. Note that we only consider $\Rel$-valued (resp.
-$\Nat$-valued) functions, but could consider other commutative semigroups.
+\AP We chose in this paper to provide a very combinatorial description of
+polyregular functions with commutative outputs because it will play nicely with
+our analysis on polynomials. This description is very similar in shape to the
+\emph{finite counting automata} introduced by \cite{SCHU62}. For the sake of
+simplicity, we only consider $\Rel$-valued (resp. $\Nat$-valued) functions, but
+could consider other commutative semigroups.
 
@@
     and a function $\pi \colon M^{k+1} \to \Rel$
-    satisfying:
+    satisfying for all $w \in \Sigma^*$:
     \begin{equation*}
-        f \defined
+        f (w) = \pi^\dagger (w) \defined
         \sum_{w = u_1 \cdots u_{k+1}} \pi(\mu(u_1), \dots, \mu(u_{k+1}))
@@
     \end{equation*}
-    We call $\pi$ the \emph{production function} of $f$.
+    We call $\pi$ the \intro{production function} of $f$.
     If the function $\pi$ has codomain $\Nat$,
@@
 
-\AP We complete definition \cref{nat-rel-poly:def} by defining $\intro*\NPoly =
+\AP We complete \cref{nat-rel-poly:def} by letting $\intro*\NPoly \defined
 \bigcup_{k \in \Nat} \intro*\NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
 $\ZSF$. In order to illustrate these definitions, let us provide an example of
-a \kl{$\Nat$-polyregular function} computed using a finite monoid.
+a \kl{$\Nat$-polyregular function} computed using a finite monoid in
+\cref{size-of-word-nsf:ex}. Let us also introduce in
+\cref{q-polynomial-n-poly:ex} a function that will appear frequently in this
+paper as an example of \emph{division} computed by a \kl{$\Nat$-polyregular
+function}.
 
@@
     We check that for all $w \in \Sigma^*$:
-    \begin{equation*}
+    $
         \pi^\dagger(w)
@@
         = f(w)
-        \quad . 
-        \qedhere
-    \end{equation*}
+        $.
+\end{proof}
+
+\begin{example}
+    \label{q-polynomial-n-poly:ex}
+    Let $f \colon \Sigma^* \to \Nat$ that maps a word $w$
+    to the sum of the length of its prefixes,
+    i.e., $f(w) \defined \sum_{uv = w} \card{u}$.
+    The function $f$ is an \kl{$\Nat$-polyregular function}
+    and 
+    $f(w) = \card{w}( \card{w} - 1) / 2$.
+\end{example}
+\begin{proof}
+    Let us write $M \defined (\set{0,1}, \max)$
+    which is a finite aperiodic monoid, and $\mu \colon \Sigma^* \to M$
+    that maps $\varepsilon$ to $0$ and the other words to $1$.
+    Then, let us define $\pi \colon M^3 \to \Nat$
+    via $\pi(x,y,z) = y$.
+    Remark that 
+    \begin{align*}
+        \pi^\dagger(w) 
+        &\defined
+        \sum_{xyz = w} \pi(\mu(x), \mu(y), \mu(z)) 
+        = 
+        \sum_{uz = w} \sum_{xy = u} \pi(\mu(x), \mu(y), \mu(z)) 
+        = 
+        \sum_{uz = w} \sum_{xy = u} \mu(y) \\
+        &= 
+        \sum_{uz = w} \card{u} 
+        = f(w) \quad .
+    \end{align*}
+    Finally, the fact that 
+    $\sum_{i = 0}^n i = n(n-1) / 2$ justifies the last equality.
 \end{proof}
@@
 the larger class of \kl{polyregular functions} \cite{CDTL23,DOUE23}. In this
-paper, the main focus will be the connection to weighted automata, which is
+paper, the main focus will be the connection to \kl{weighted automata}, which is
 based on the notion of \emph{growth rate}. The \intro{growth rate} of a
@@
         \item $f \in \ZPoly[k]$.
-        \item $f$ has polynomial growth of degree at most $k$.
+        \item $f$ has \kl{polynomial growth} of degree at most $k$.
     \end{enumerate}
@@
 
-Let us state the crucial compositional properties of
+Let us introduce some compositional properties of
 \kl{$\Rel$-polyregular functions} that will be used in this paper
@@
 \kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}. As a
-consequence, we are working inside a relatively robust and decidable class of
+consequence, we are working inside a relatively robust and decidable subclass of
 \kl{$\Rel$-rational series}.
@@
 
-\subsection{Polynomials} 
-\AP All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. 
-A polynomial $P \in \Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for
-all non-negative integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1,
-\dots, n_k)$ of the polynomial is non-negative. Beware that we do not consider
-negative values as input, as the numbers $n_i$ will ultimately count the number
-of occurrences of a letter in a word. As an example, the polynomial $(X - Y)^2$
-is \kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
+\subsection{Polynomials} \AP All polynomials considered in this paper have
+coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
+\Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for all non-negative
+integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1, \dots, n_k)$ of
+the polynomial is non-negative. Beware that we do not consider negative values
+as input, as the numbers $n_i$ will ultimately count the number of occurrences
+of a letter in a word. As an example, the polynomial $(X - Y)^2$ is
+\kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
 2X$ is not.

### 2024-08-15 parts/starfree.tex
enormous rewrite

@@
     \label{decompose-polynomial:lem}
+    \proofref{decompose-polynomial:lem}
     Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
@@
     \end{equation*}
-    \proofref{decompose-polynomial:lem}
 \end{lemma}
@@
 \end{definition}
+\begin{conjecture}
+    \label{npoly-zpoly:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \ZPoly[k]$
+    belongs to $\NPoly[k]$ if and only if
+    it is \kl{$(k,\Nat)$-combinatorial}.
+    In particular,
+    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
+\end{conjecture}
 
@@
 
-    Now, consider an \kl{$\omega$-type} $t \in \ModuloTypes[\omega]^k$. By
-    definition, this provides us with a selection function $\ind{S} \colon \set{1,
-    \dots, k} \to \set{0,1}$, and numbers $r_1, \dots, r_k$ between $0$ and
+    Now, consider $t \in \ModuloTypes[\omega]^k$. By
+    definition, this provides us with a set $S \subseteq \set{1, \dots, k}$,
+    and numbers $r_1, \dots, r_k$ between $0$ and
     $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
@@
 
-\begin{conjecture}
-    \label{npoly-zpoly:conjecture}
-    Let $k \in \Nat$.
-    A function $f \in \ZPoly[k]$
-    belongs to $\NPoly[k]$ if and only if
-    it is \kl{$(k,\Nat)$-combinatorial}.
-    In particular,
-    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
-\end{conjecture}
-
 Let us now prove that the above characterizations of \kl{commutative}
@@
 functions by the means of polynomial behaviors (see
-\cref{aperidic-ultimately-polynomial:ex}). Note that up until now, we only had
-to consider polynomials with coefficients in $\Rel$. However, there are some
-polynomials with coefficients in $\Rat$ that can be \kl{represented} by a
-\kl{$\Nat$-polyregular function}. Recall that the polynomial $P(X) \defined
-X(X-1)/2$ can be computed by a \kl{$\Nat$-polyregular function}. Historically,
-the definition used polynomials with coefficients in $\Rat$, which is why we
-use such polynomials here, even though we know that all of those polynomials
-will actually be \kl{integer binomial polynomials} thanks to
+\cref{aperidic-ultimately-polynomial:ex}). Historically, the definition used
+polynomials with coefficients in $\Rat$, which is why we use such polynomials
+here, even though we know that all of those polynomials will actually be
+\kl{integer binomial polynomials} thanks to
 \cref{integer-binomial-polynomials:cor}.
@@
 
-    Conversely, assume that $L$ is aperiodic, let $n \in \Nat$ be a number, and
+    Conversely, assume that $L$ is \kl(language){aperiodic}, let $n \in \Nat$ be a number, and
     let $\alpha_0$, $w_1$, $\alpha_1$, $\ldots$, $\alpha_{n-1}$, $w_n$, $\alpha_n$
-    be words in $\Sigma^*$. Because $L$ is \kl(language){aperodic}, the language
+    be words in $\Sigma^*$. Because $L$ is \kl(language){aperiodic}, the language
     $L$ is recognized by an \kl(monoid){aperiodic} finite monoid $M$, via a

### 2024-08-20 parts/correction.tex
typo
test

@@
 polynomials in $\CorrectPoly$ are \kl{non-negative}. We also remarked at the
-beginning of this section that our counterexample provided in
-\cref{thm:counter-example} is not in $\CorrectPoly$.
+beginning of this section that our counterexample $\BadPoly$ provided in
+\cref{thm:counter-example} is not in $\CorrectPoly$. The rest of the
+section is mainly concerned with proving the following corrected version of
+\cref{karh:thm}.
+
+\begin{theorem}[restate=corrected-version:thm,label=corrected-version:thm]
+    Let $P \in \Rel[\vec{X}]$.
+    The following are equivalent:
+    \begin{enumerate}
+        \item \label{corrected-1:item} $P \in \CorrectPoly$,
+        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
+        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
+        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+    \end{enumerate}
+    Furthermore, the properties are decidable, and conversions effective.
+\end{theorem}
 
@@
 the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
-and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in
-\cref{n-poly-combinatorics:lem} a finer statement that has two main goals:
-first, it formulates the traditional \emph{pumping argument} as a composition
-property using \kl{(noncommutative output) polynomial functions}, allowing us
-to use the same formalism in \cref{k-combinatorial:def} and
-\cref{ultimately-polynomial:def} and highlight the role of \kl{commutative}
-\kl{polyregular functions} in the broader study of \kl{polyregular functions};
-then we also give a precise shape of the polynomials that arise from such
-\emph{pumping arguments}, which was lacking in former similar statements where
-$\Rat[X_1, \dots, X_n]$ was used as an over approximation.
+and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in our
+upcoming \cref{n-poly-combinatorics:lem} a finer statement that has two main
+goals: first, it formulates the traditional \emph{pumping argument} as a
+composition property using \kl{(noncommutative output) polynomial functions},
+allowing us to use the same formalism in the upcoming
+\cref{k-combinatorial:def,ultimately-polynomial:def} and highlight the role of
+\kl{commutative} \kl{polyregular functions} in the broader study of
+\kl{polyregular functions}; then we also give a precise shape of the
+polynomials that arise from such \emph{pumping arguments}, which was lacking in
+former similar statements where $\Rat[X_1, \dots, X_n]$ was used as an over
+approximation.
 
@@
 \prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
-non-commutative monomial over the semi- $\Sigma^*$. Almost by
+non-commutative monomial over the monoid $\Sigma^*$. Almost by
 definition, \kl{polynomial functions} are \kl{commutative} \kl{polyregular
@@
 \Nat$ defined as $0$ when the input is of odd length and $1$ when the input is
-of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial.
-We can trade-off this multiplicative factor for a constant term 
-addition under the extra assumption that the function 
-is \kl{star-free polyregular}, as described in the following lemma.
+of even length is \kl{$\Nat$-polyregular}, but $f(a^X)$ is not a polynomial. We
+can trade off this multiplicative factor for a constant term addition under the
+extra assumption that the function is \kl{star-free polyregular}, as described
+in the following \cref{n-sf-combinatorics:lem}.
 
@@
 function} corresponds exactly one \kl{natural binomial polynomial} that
-coincides with it on sufficiently large inputs.
+coincides with it on sufficiently large inputs (\cref{natural-binomial-polynomial:lem})
+which allows us to conclude that \kl{$\Nat$-rational polynomials} are in
+$\CorrectPoly$ (\cref{n-rat-correct:lem}).
 
@@
 
-\begin{theorem}
-    \label{corrected-version:thm}
-    Let $P \in \Rel[\vec{X}]$.
-    The following are equivalent:
-    \begin{enumerate}
-        \item \label{corrected-1:item} $P \in \CorrectPoly$,
-        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
-        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
-    \end{enumerate}
-    Furthermore, the properties are decidable, and conversions effective.
-\end{theorem}
-\begin{proof}
+\begin{proofof}[corrected-version:thm]
     The implications 
@@
     procedure.
-
-\end{proof}
+\end{proofof}
 
@@
 
+\begin{lemma}
+    \label{natural-binomia-correct:lem}
+    Let $P$ be a \kl{non-negative} \kl{natural binomial polynomial},
+    and let $\alpha$ be such that $\alpha P \in \Rel[\vec{X}]$.
+    Then, $\alpha P \in \CorrectPoly$.
+\end{lemma}
+\begin{proof}
+    The only think to remark is that
+    $\restr{P}{\nu}$ remains a \kl{natural binomial polynomial}
+    for all partial valuations $\nu$, and furthermore remains positive.
+\end{proof}
+
 \begin{corollary}
@@
     \begin{enumerate}
-        \item \label{rat-npoly-0:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
-        \item \label{rat-npoly-1:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
+        \item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
+        \item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
         \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
@@
 \begin{proof}
+    First, let us remark that 
+    \cref{rat-npoly-0:item} $\iff$ \cref{rat-npoly-1:item}
+    because \kl{natural binomial polynomials} are increasing.
+    Furthermore, \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}
+    because $\NSF \subseteq \NPoly$.
 
-    Proof of \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}.
-    If $\alpha P \in \CorrectPoly$, then  it is an \kl{$\Nat$-rational polynomial}.
-    But in particular, there exists a polyregular function $f$ that represents $\alpha P$
-    with outputs in $\set{ a }$.
-    Now, let us post-compose this function by the one that
-    maps $a^{k \alpha + r}$ to $a^{k}$, which is a \kl{polyregular function}.
-    We conclude that $P$ is \kl{represented} by an \kl{$\Nat$-polyregular function}.
+    Then, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
+    if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
+    polynomial}. But in particular, there exists a polyregular function $f$
+    that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
+    post-compose this function by the one that maps $a^{k \alpha + r}$ to
+    $a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
+    \kl{represented} by an \kl{$\Nat$-polyregular function}.
 
-    Proof of \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-0:item}.
-    If $P$ is a \kl{natural binomial polynomial} with $P(0,\dots,0) \geq 0$,
-    then it is \kl{non-negative}.
+    We also know that \cref{rat-npoly-2:item} $\implies$
+    \cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
+    polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
+    \CorrectPoly$ by \cref{corrected-version:thm}.
 
-    Proof of \cref{rat-npoly-0:item} $\implies$ \cref{rat-npoly-3:item}.
-    If $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
-    then for large enough values, it is a \kl{natural binomial function},
-    hence computable, and for small values we can hardcode.
+    As a consequence, we have proven that \cref{rat-npoly-0:item} $\iff$
+    \cref{rat-npoly-1:item} and \cref{rat-npoly-2:item} $\iff$
+    \cref{rat-npoly-3:item} $\iff$ \cref{rat-npoly-4:item}.
 
-    Proof of \cref{rat-npoly-2:item} $\implies$ \cref{rat-npoly-4:item}.
-    If $P$ is a \kl{$\Nat$-rational polynomial}, then $\alpha P$ is one too,
-    and in particular $\alpha P$ belongs to $\CorrectPoly$.
-
-    Proof of \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}.
-    This is because $\NSF \subseteq \NPoly$.
-
-    Proof of \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-0:item}.
-    Because $P$ is computed by a function $f \in \NSF$,
-    one can leverage \cref{n-sf-combinatorics:lem}
-    to conclude that $P$ is actually a \kl{natural binomial polynomial}.
+    Finally, \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-4:item} is
+    exactly \cref{natural-binomia-correct:lem}. And \cref{rat-npoly-3:item}
+    $\implies$ \cref{rat-npoly-1:item} is obtained
+    by combining \cref{n-sf-combinatorics:lem} and \cref{natural-binomial-polynomial:lem}.
 \end{proof}
@@
     Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
-    \kl{$\Nat$-polyregular functions}.
-    Let us argue that $f_+$ and $f_-$ are \kl{star-free polyregular functions},
-    because they are \kl{ultimately polynomial}. (todo later).
-    As a consequence, one can apply \cref{n-sf-combinatorics:lem}
+    \kl{star-free $\Nat$-polyregular functions}.
+    One can apply \cref{n-sf-combinatorics:lem}
     to conclude that $P$ coincides with an 

### 2024-08-20 parts/intro.tex
test

@@
 
-\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
+\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we provide
+a combinatorial definition of
 \kl{$\Nat$-polyregular functions} (resp. $\Rel$-polyregular functions), show
@@
 can be expressed respectively by \kl{$\Nat$-polyregular functions}
-(\cref{corrected-version:thm} \cref{corrected-0:item}) and
+(\cref{decide-rat-poly-npoly:cor}) and
 \kl{$\Rel$-polyregular functions} (\cref{integer-binomial-polynomials:cor}).

### 2024-08-20 parts/starfree.tex
test

@@
 relies on the construction of a canonical object called the \emph{residual
-,
 transducer}, the latter being essentially based on differences between

### 2024-08-26 parts/correction.tex
counterexamples

@@
 
+\subsubsection{From $\Nat$-polyregular functions to polynomials}
+
 \AP Let us now prove that \kl{$\Nat$-rational polynomials} are in
@@
 1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is
-defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial. For
-this reason, we also introduce its ``polynomial counterpart" $\intro*\pbinom{X
-- \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
-\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
-as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
-the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
-- p_{i,j}}{k_{i,j}}$.
+defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial.
 
@@
 
-Let us also address the fact that the distinction between \kl{natural binomial
-functions} and \kl{natural binomial polynomials} vanishes when we are
-interested in polynomials. As a consequence, to a \kl{natural binomial
-function} corresponds exactly one \kl{natural binomial polynomial} that
-coincides with it on sufficiently large inputs (\cref{natural-binomial-polynomial:lem})
-which allows us to conclude that \kl{$\Nat$-rational polynomials} are in
-$\CorrectPoly$ (\cref{n-rat-correct:lem}).
-
-\begin{lemma}
-    \label{natural-binomial-polynomial:lem}
-    Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial,
-    $\ell \in \Nat$, 
-    and let $f$ 
-    be a 
-    \kl{natural binomial function} such that
-    $P = f$ over $(\Nat_{\geq \ell})^n$. If
-    $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i 
-    \prod_{j = 1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$,
-    then
-    $P =
-    \sum_{i = 1}^n n_i 
-    \prod_{j = 1}^k \pbinom{X_j - p_{i,j}}{k_{i,j}}$.
-\end{lemma}
-\begin{proof}
-    Let $m$ be the maximal value of all $p_{i,j}$, $k_{i,j}$, and $\ell$.
-    Over $(\Nat_{\geq 2m})^k$, $f$ coincides with
-    the desired polynomial. Now, because $f$,$P$, and the desired
-    polynomial coincide
-    over a large enough subset of $\Nat^k$,
-    this means that $P$ has the desired shape.
-\end{proof}
-
 \begin{corollary}
@@
 \end{proof}
+
+\subsubsection{From polynomials to $\Nat$-polyregular functions}
 
@@
 
+\subsubsection{From $\Rel$ to $\Rat$ and $\Real$}
+
 Let us conclude this section by considering polynomials with coefficients in
@@
 $\Rel$ and $\Nat$.
+
+For
+this reason, we also introduce its ``polynomial counterpart" $\intro*\pbinom{X
+- \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
+\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
+as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
+the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
+- p_{i,j}}{k_{i,j}}$. 
+Let us also address the fact that the distinction between \kl{natural binomial
+functions} and \kl{natural binomial polynomials} vanishes when we are
+interested in polynomials. As a consequence, to a \kl{natural binomial
+function} corresponds exactly one \kl{natural binomial polynomial} that
+coincides with it on sufficiently large inputs (\cref{natural-binomial-polynomial:lem})
+which allows us to conclude that \kl{$\Nat$-rational polynomials} are in
+$\CorrectPoly$ (\cref{n-rat-correct:lem}).
+
+\begin{lemma}
+    \label{natural-binomial-polynomial:lem}
+    Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial,
+    $\ell \in \Nat$, 
+    and let $f$ 
+    be a 
+    \kl{natural binomial function} such that
+    $P = f$ over $(\Nat_{\geq \ell})^n$. If
+    $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i 
+    \prod_{j = 1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$,
+    then
+    $P =
+    \sum_{i = 1}^n n_i 
+    \prod_{j = 1}^k \pbinom{X_j - p_{i,j}}{k_{i,j}}$.
+\end{lemma}
+\begin{proof}
+    Let $m$ be the maximal value of all $p_{i,j}$, $k_{i,j}$, and $\ell$.
+    Over $(\Nat_{\geq 2m})^k$, $f$ coincides with
+    the desired polynomial. Now, because $f$,$P$, and the desired
+    polynomial coincide
+    over a large enough subset of $\Nat^k$,
+    this means that $P$ has the desired shape.
+\end{proof}
+
+\begin{conjecture}
+    Let $P \in \Real[\vec{X}]$. Then $P \in \CorrectPoly$ if and only if $P$ 
+    there exists $K$ such that
+    $\restr{P}{\nu}$ is a \kl{natural binomial polynomial} \kl{in base $K$}
+    for all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
+    That is, for all $\nu$, there exists $\alpha_i \geq 0$ and $d_{i,j} \geq 0$ such that
+    $\restr{P}{\nu} = \sum_i \alpha_i \prod_j \pbinom{X_j - K}{d_{i,j}}$.
+\end{conjecture}
 
@@
     \begin{enumerate}
-        \item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
-        \item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
+        %\item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
+        %\item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
         \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
@@
 \begin{proof}
-    First, let us remark that 
-    \cref{rat-npoly-0:item} $\iff$ \cref{rat-npoly-1:item}
-    because \kl{natural binomial polynomials} are increasing.
-    Furthermore, \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}
-    because $\NSF \subseteq \NPoly$.
-
-    Then, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
+    First, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
     if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
@@
     \CorrectPoly$ by \cref{corrected-version:thm}.
-
-    As a consequence, we have proven that \cref{rat-npoly-0:item} $\iff$
-    \cref{rat-npoly-1:item} and \cref{rat-npoly-2:item} $\iff$
-    \cref{rat-npoly-3:item} $\iff$ \cref{rat-npoly-4:item}.
-
-    Finally, \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-4:item} is
-    exactly \cref{natural-binomia-correct:lem}. And \cref{rat-npoly-3:item}
-    $\implies$ \cref{rat-npoly-1:item} is obtained
-    by combining \cref{n-sf-combinatorics:lem} and \cref{natural-binomial-polynomial:lem}.
 \end{proof}
@@
 
+It is a longstanding result that polynomials with $\Rel$-output are exactly
+\kl{integer binomial polynomials} \cite{POLYA1915,CACHA1996}. However, its
+generalization to $\Nat$ is not immediate. First, having non-negative
+coefficients depends on a choice of basis, and second, the non-negativity is
+not enough to ensure that the polynomial is a \kl{$\Nat$-rational polynomial}.
+
+\begin{example}
+    The polynomial $\pbinom{X - 2}{2} + 1$ is a \kl{non-negative} \kl{natural binomial polynomial}
+    but in the standard basis, it is written 
+    $\pbinom{X}{2} - \pbinom{X}{1} + 2$, which has a \kl{negative} coefficient.
+\end{example}
+
+\begin{example}
+    The polynomial
+    $Q(X,Y,Z) \defined \pbinom{X - 4}{1} \pbinom{Y}{1} \pbinom{Z}{1}
+    + 8 \pbinom{Y}{2} + 8 \pbinom{Z}{2} + 4$
+    is a \kl{non-negative} \kl{natural binomial polynomial},
+    but cannot be computed by a \kl{star-free $\Nat$-polyregular function}.
+\end{example}
+
 \begin{corollary}
     \label{integer-binomial-polynomials:cor}
-    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
-    if and only if $P$ is an \kl{integer binomial polynomial}.
+    Let $P \in \Real[\vec{X}]$. Then, the following propreties are equivalent:
+    \begin{enumerate}
+        \item \label{int-bin-0:item} $P$ is an \kl{integer binomial polynomial},
+        \item \label{int-bin-1:item} $P$ \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+        \item \label{int-bin-2:item} $P(\Rel^k) \subseteq \Rel$.
+    \end{enumerate}
 \end{corollary}

### 2024-08-26 parts/prelim.tex
test
counterexamples

@@
 for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
-to $K$ are the functions computed by \intro{$K$-weighted automata} which are
+to $K$ are the functions computed by \kl{$K$-weighted automata} which are
 non-deterministic finite automata with transitions labelled by elements of $K$,
@@
 state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
+series}. We use $\intro*\NRat$ and $\intro*\ZRat$ to denote respectively
+the set of \kl{$\Nat$-rational series} and the set of \kl{$\Rel$-rational
 series}.
@@
     then
-    the function $f$ is \kl{star-free $\Rel$-polyregular}
-    ($\intro*\ZSF[k]$), resp. \kl{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
+    the function $f$ is \intro{star-free $\Rel$-polyregular}
+    ($\intro*\ZSF[k]$), resp. \intro{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
 \end{definition}
 
-\AP We complete \cref{nat-rel-poly:def} by letting $\intro*\NPoly \defined
-\bigcup_{k \in \Nat} \intro*\NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
+\AP We complete \cref{nat-rel-poly:def} by letting $\reintro*\NPoly \defined
+\bigcup_{k \in \Nat} \NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
 $\ZSF$. In order to illustrate these definitions, let us provide an example of
@@
 \end{proof}
+
+\AP In full generality, \cref{size-of-word-nsf:ex,q-polynomial-n-poly:ex} can be generalized to demonstrate that functions
+of the following functions can be \kl{represented} by a \kl{commutative}
+\kl{star-free $\Nat$-polyregular function}: $x \mapsto \binom{x - p}{k}$ for $p
+\in \Nat$ and $k \in \Nat$, where $\binom{n}{k}$ is the binomial coefficient,
+that is the number of ways to choose $k$ elements among $n$ elements.
 

### 2024-08-27 parts/correction.tex
moving polynomials to another section

--- 2024-08-20 parts/correction.tex
@@
     \label{integer-binomial-polynomials:cor}
-    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
-    if and only if $P$ is an \kl{integer binomial polynomial}.
+    Let $P \in \Real[\vec{X}]$. Then, the following propreties are equivalent:
+    \begin{enumerate}
+        \item \label{int-bin-0:item} $P$ is an \kl{integer binomial polynomial},
+        \item \label{int-bin-1:item} $P$ \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+        \item \label{int-bin-2:item} $P(\Rel^k) \subseteq \Rel$.
+    \end{enumerate}
 \end{corollary}

### 2024-08-27 parts/correction.tex
integer binomial polynomials

--- 2024-08-20 parts/correction.tex
@@
 		1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is
-defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial. For
-this reason, we also introduce its ``polynomial counterpart" $\intro*\pbinom{X
-- \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
-\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
-as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
-the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
-- p_{i,j}}{k_{i,j}}$.
+defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial.
 
@@
 	Let us show that $f \circ q$ is
-    actually a constant (positive) function (and in particular a \kl{natural binomial polynomial}), as for
+	actually a constant (positive) function (and in particular a \kl{natural binomial function}), as for
 	all $n_1, \dots, n_p \geq 1$,
@@
 		f(q(n_1, \dots, n_p))
-        &\eqdef \pi\left(\mu(q(n_1, \dots, n_p))\right) \\
-        &\eqdef \pi\left(\mu(\alpha_0 \prod_{i = 1}^p u_i^{\omega n_i} \alpha_i)\right)
+		 & \eqdef \pi\left(\mu(q(n_1, \dots, n_p))\right)                                      \\
+		 & \eqdef \pi\left(\mu(\alpha_0 \prod_{i = 1}^p u_i^{\omega n_i} \alpha_i)\right)
 		\\
-        &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i)^{\omega n_i} \mu(\alpha_i)\right) 
+		 & = \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i)^{\omega n_i} \mu(\alpha_i)\right)
 		 & \text{ morphism }
 		\\
-        &= \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
+		 & = \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
 		 & \text{ idempotent and } n_i \geq 1                                                  \\
-        &= c \geq 0
+		 & = c \geq 0
 	\end{align*}
@@
 	\begin{align*}
-        &f(q(n_1, \dots, n_p))\\
-        &\eqdef 
+		 & f(q(n_1, \dots, n_p))           \\
+		 & \eqdef
 		\sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
 		\pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
-        &= \sum_{(\ell, b, r)}
+		 & = \sum_{(\ell, b, r)}
 		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
@@
 		\\
-        &= 
+		 & =
 		\sum_{(\ell, 0, r)}
@@
 		\\
-        &+ 
+		 & +
 		\sum_{(\ell, b > 0, r)}
@@
 		\\
-        &= 
+		 & =
 		\sum_{1 \leq \ell \leq p}
@@
 		\\
-        &+ 
+		 & +
 		\sum_{1 \leq \ell \leq p}
@@
 	the desired equality.
-
 \end{proof}
@@
 	\end{equation*}
-    As a consequence of \cref{natural-binomial-polynomial:lem}, $P(\omega X_1, \dots, \omega X_p)$ equals the corresponding 
-    \kl{natural binomial polynomial}.
-    Let us remark that the (unique) \kl{maximal monomial} of the 
-    \kl{binomial monomial} $\pbinom{X -
-    p}{k}$ has coefficient $1/k!$, and is therefore \kl{non-negative}.
+	Let us remark that for large enough values of $X$, the \kl{binomial function}
+	$\binom{X - p}{k}$ coincides with a polynomial whose dominant coefficient
+	is $1/k!$ which is \kl{non-negative}.
 	We conclude that
@@
 
-Let us conclude this section by considering polynomials with coefficients in
-$\Rat$, and justify that all the combinatorial work has already happened in
-$\Rel$ and $\Nat$.
+\subsubsection{From $\Rel$ to $\Rat$}
+
+\AP Let us conclude this section by considering polynomials with
+coefficients in $\Rat$, and justify that all the combinatorial work has
+already happened in $\Rel$ and $\Nat$. A first observation is that from
+\cref{n-poly-combinatorics:lem,n-sf-combinatorics:lem}, we know that the
+functions that can be computed by \kl{star-free $\Nat$-polyregular
+	functions} are going to be suitable combinations of \kl{natural binomial
+	functions}. This motivates the following definition of the ``polynomial
+counterpart" of a binomial coefficient, $\intro*\pbinom{X - \ell}{k}$
+defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
+\intro{binomial monomial}, and introduce \intro{natural binomial
+	polynomials} as $\Nat$-linear combinations of products of \kl{binomial
+	monomials}, i.e., of the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i
+	\prod_{j = 1}^k \pbinom{X_j - p_{i,j}}{k_{i,j}}$. Similarly, we introduce
+the class of \intro{integer binomial polynomials}, which are obtained by
+$\Rel$-linear combinations of products of \kl{binomial monomials}.
+
+\AP
+These definitions are motivated by the classical result of Pólya that
+characterizes polynomials $P$ in $\Rat[X]$ that \intro{integer-valued} (i.e., are such
+that $P(\Rel) \subseteq \Rel$) as \kl{integer binomial polynomials}
+\cite{POLYA1915,CACHA1996}, which are often called \emph{Newton Polynomials}. Note
+that this result straightforwardly extends to multiple indeterminates as we
+prove in \cref{integer-binomial-polynomial:lem}.
 
 \begin{lemma}
-    \label{natural-binomia-correct:lem}
-    Let $P$ be a \kl{non-negative} \kl{natural binomial polynomial},
-    and let $\alpha$ be such that $\alpha P \in \Rel[\vec{X}]$.
-    Then, $\alpha P \in \CorrectPoly$.
+	\label{integer-binomial-polynomial:lem}
+	Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial.
+	Then, $P$ is an \kl{integer binomial polynomial} if and only if
+	$P(\Rel^k) \subseteq \Rel$.
 \end{lemma}
 \begin{proof}
-    The only think to remark is that
-    $\restr{P}{\nu}$ remains a \kl{natural binomial polynomial}
-    for all partial valuations $\nu$, and furthermore remains positive.
+	Let us first notice that $\pbinom{X}{k}$ is \kl{integer valued}
+	when $k \in \Nat$. One way to notice this is that when $X \geq 0$,
+	$\pbinom{X}{k} = \binom{X}{k}$ is the number of ways to choose $k$ elements among $X$,
+	which is an integer. When
+	$X < 0$, then $\pbinom{X}{k} = (-1)^k \pbinom{k - X + 1}{k} = (-1)^k \binom{k - X + 1}{k}$
+	is also an integer. Because \kl{integer-valued} polynomials are closed under products
+	and $\Rel$-linear combination, we conclude that all \kl{integer binomial polynomials} are
+	\kl{integer-valued}.
+
+	For the converse direction, we follow the original proof of Pólya \cite{POLYA1915,CACHA1996}.
+	The only difference is that we leverage differential operators on multiple indeterminates.
+	To that end, let us consider the operator $\Delta_i$ that maps
+	a polynomial $P$ to the polynomial $\Delta_i(P) \defined P(X_1, \dots, X_i + 1, \dots, X_k) - P$.
+	That is, the discrete partial derivative with respect to the $i$-th indeterminate.
+	Let us notice that $\Delta_i \pbinom{X_i}{\alpha} = \pbinom{X_i}{\alpha - 1}$ when $\alpha > 0$,
+	indeed:
+	\begin{align*}
+		\Delta_i \pbinom{X_i}{\alpha} & = \pbinom{X_i + 1}{\alpha} - \pbinom{X_i}{\alpha}                                                         \\
+		                              & = \frac{(X_i + 1) \cdots (X_i + 1 - \alpha + 1)}{\alpha!} - \frac{X_i \cdots (X_i - \alpha + 1)}{\alpha!} \\
+		                              & = \frac{X_i \cdots (X_i - \alpha + 2)}{\alpha!} \alpha                                                    \\
+		                              & = \pbinom{X_i}{\alpha - 1}
+		\quad .
+	\end{align*}
+	Furthermore, when $\alpha = 0$, $\Delta_i \pbinom{X_i}{0} = 0$.
+
+	Let us now remark that a simple expansion proves $\Delta_i \Delta_j P = \Delta_j \Delta_i P$ for all
+	$1 \leq i,j \leq k$.
+	This justifies the soundness of the notation
+	$\Delta_\alpha$ where $\alpha \colon \set{1, \dots, k} \to \Nat$ is a
+	multi-index, where $\Delta_{\alpha} \defined \Delta_1^{\alpha_1} \cdots
+		\Delta_k^{\alpha_k}$.
+	Let us write $\card{\alpha} \defined \sum_{i = 1}^k \alpha_i$ for the
+	size of a multi-index.
+	We claim that for all $P \in \Rat[X_1, \dots, X_k]$ of degree at most $d$,
+	the following \emph{discrete Taylor expansion} holds:
+	\begin{equation}
+		\label{discrete-taylor:eq}
+		P =
+		\sum_{\card{\alpha} \leq d}
+		\Delta_{\alpha} P (0, \dots, 0)
+		\times
+		\prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
+		\quad .
+	\end{equation}
+
+	To obtain \cref{discrete-taylor:eq}, we proceed by induction on the degree $d$.
+	When $d = 0$, the result is holds because $P = P(0, \dots, 0)$.
+	Assume that the result holds for all polynomials of degree at most $d$.
+	Let $P$ of degree at most $d+1$.
+	Then, there exists coefficients $\seqof{c_{\alpha}}{\card{\alpha} \leq d+1}$
+	in $\Rat$
+	such that
+	\begin{equation*}
+		P = \sum_{\card{\alpha} \leq d+1} c_{\alpha} \prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
+		\quad .
+	\end{equation*}
+	This is because the monomials of the form $\prod_{i = 1}^k \pbinom{X_i}{\alpha_i}$
+	form a basis of $\Rat[X_1, \dots, X_k]$.
+	In particular, when applying one partial derivative, one can leverage the induction hypothesis to obtain:
+	\begin{align*}
+		\Delta_1 P & = \sum_{\card{\alpha} \leq d+1} c_{\alpha} \Delta_1 \prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
+		=
+		\sum_{\card{\alpha} \leq d+1 \wedge \alpha_1 > 0} c_{\alpha}  \pbinom{X_1}{\alpha_1 - 1} \prod_{i = 2}^k \pbinom{X_i}{\alpha_i}
+		\\
+		           & = \sum_{\card{\beta} \leq d} \Delta_{\beta}(\Delta_1 P)(0,\dots,0) \times \pbinom{X_1}{\beta_1} \prod_{i = 2}^k \pbinom{X_i}{\beta_i}
+		\quad .
+	\end{align*}
+
+	Using again the fact that the monomials form a basis, we conclude by
+	identification that for all $\card{\alpha} \leq d+1$ such that $\alpha_1 >
+		0$, we have $c_{\alpha} = \Delta_{\alpha}P (0,\dots,0)$. We can repeat the
+	same process with $\Delta_i$ for $1 \leq i \leq k$ to obtain that
+	$c_{\alpha} = \Delta_{\alpha} P (0,\dots, 0)$ for all $\alpha$ such that $1
+		\leq \card{\alpha} \leq d+1$. Now, for the specific case of $\card{\alpha}
+		= 0$, which is the constant term of $P$, we simply conclude that
+	$P(0,\dots,0) = c_{\alpha}$, and notice that $\Delta_{\alpha} P = P$ in
+	this case.
+
+	Let us now conclude by noticing that if $P$ is \kl{integer-valued},
+	then so is $\Delta_i P$ for all $1 \leq i \leq k$. As a consequence,
+	so are $\Delta_{\alpha} P$, for all multi-indexes $\alpha$, and we conclude
+	that $\Delta_{\alpha}P (0,\dots,0) \in \Rel$ for all $\alpha$.
+	Combined with \cref{discrete-taylor:eq}, we conclude that $P$ is an \kl{integer binomial polynomial}.
 \end{proof}
+
+As an immediate corollary, we completely characterize the class of
+polynomials with rational coefficients that can be computed by
+\kl{star-free $\Nat$-polyregular functions}.
+
+\begin{corollary}
+	\label{integer-binomial-polynomials:cor}
+	Let $P \in \Rat[\vec{X}]$, and let $\alpha$ be the minimal number in $\Nat$ such that
+	$\alpha P \in \Rel[\vec{X}]$. Then, the following properties are equivalent:
+	\begin{enumerate}
+		\item \label{int-bin-0:item} $P$ is an \kl{integer binomial polynomial},
+		\item \label{int-bin-1:item} $P$ \kl{represented} by a \kl{star-free $\Rel$-polyregular function},
+		\item \label{int-bin-2:item} $P$ \kl{represented} by a \kl{$\Rel$-polyregular function},
+		\item \label{int-bin-3:item} $P(\Rel^k) \subseteq \Rel$.
+	\end{enumerate}
+	These properties are furthermore decidable.
+\end{corollary}
+\begin{proof}
+	Let us first note that implications \cref{int-bin-1:item}
+	$\implies$ \cref{int-bin-2:item} $\implies$ \cref{int-bin-3:item} are obvious.
+	Now, \cref{int-bin-3:item} $\implies$ \cref{int-bin-0:item} is a direct
+	consequence of \cref{integer-binomial-polynomial:lem}.
+	Finally, \cref{int-bin-0:item} $\implies$ \cref{int-bin-1:item} follows from the fact
+	that $\pbinom{X - p}{k}$ is \kl{represented} by a \kl{star-free $\Rel$-polyregular function}
+	defined by hardcoding the output values (in $\Rel$) when $0 \leq X \leq p$, and
+	using a \kl{star-free $\Nat$-polyregular function} when $X > p$
+	(\cref{binomial-function-star-free:lem}).
+	Because
+	$\ZSF$ is closed under products and $\Rel$-linear combinations, we conclude.
+\end{proof}
+
+Obtaining an analogue of \cref{integer-binomial-polynomials:cor}
+for \kl{$\Nat$-polyregular functions} is more challenging. The
+naïve approach would be to consider some variation around \kl{natural binomial
+	polynomials}. A first observation is that the class of products of the form
+$\prod_{i = 1}^n \pbinom{X_i - p_i}{k_i}$ is a generating family for
+$\Rat[\vec{X}]$ but is not linearly independent. As a consequence, there may be
+several ways to write a given polynomial, and as illustrated by
+\cref{natural-binomial-polynomial-basis:ex}, some of them will
+exhibit negative coefficients.\footnote{This was not a problem for \kl{integer
+		binomial polynomials}, for which the change of basis mapped coefficients in
+	$\Rel$ to coefficients $\Rel$.} Furthermore, there are some \kl{natural
+	binomial polynomials} that take negative values on positive inputs, as
+illustrated by \cref{natural-binomial-polynomial-negative:ex}.
+Finally, restricting our attention to \kl{non-negative} \kl{natural binomial
+	polynomials} suffers from the same kind of issues as
+\cref{karh:thm},
+witnessed by \cref{natural-binomial-polynomial-positive-bad:ex}.
+
+\begin{example}
+	\label{natural-binomial-polynomial-negative:ex}
+	The polynomial $\pbinom{X - 2}{2}$ takes negative values for $X = 0$.
+\end{example}
+
+\begin{example}
+	\label{natural-binomial-polynomial-basis:ex}
+	The polynomial $\pbinom{X - 2}{2} + 1$ is a \kl{non-negative} \kl{natural binomial polynomial}
+	but in the standard basis, it is written
+	$\pbinom{X}{2} - \pbinom{X}{1} + 2$, which has a negative coefficient.
+\end{example}
+
+\begin{example}
+	\label{natural-binomial-polynomial-positive-bad:ex}
+	The polynomial
+	$Q(X,Y,Z) \defined \pbinom{X - 4}{1} \pbinom{Y}{1} \pbinom{Z}{1}
+		+ 8 \pbinom{Y}{2} + 8 \pbinom{Z}{2} + 4$
+	is a \kl{non-negative} \kl{natural binomial polynomial},
+	but cannot be computed by a \kl{star-free $\Nat$-polyregular function}.
+\end{example}
+
+\AP We say that a \kl{natural binomial polynomial} is \intro{in base $K$} if
+all its \kl{binomial monomials} are of the form $\pbinom{X - K}{d}$ for some $d
+	\in \Nat$. To address the issues raised by the previous examples, we introduce
+the notion of \intro{strongly natural binomial polynomials} \kl{in base $K$},
+as the polynomials $P \in \Rat[X]$ such that for all partial valuation $\nu
+	\colon \topartial \set{0, \dots, K}$, $\restr{P}{\nu}$ is a \kl{natural
+	binomial polynomial} \kl{in base $K$}. This leads to the following
+characterization.
 
@@
 	coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
-    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
+	such that $\alpha P \in \Rel[\vec{X}]$. Then, there exists a
+	computable $K \in \Nat$, such that the following are equivalent:
 	\begin{enumerate}
-        \item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
-        \item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
-        \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
+		\item \label{rat-npoly-1:item}
+		      $P$ is a \kl{strongly natural binomial polynomial} \kl{in base $K$},
+		\item \label{rat-npoly-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
 		\item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
@@
 \begin{proof}
-    First, let us remark that 
-    \cref{rat-npoly-0:item} $\iff$ \cref{rat-npoly-1:item}
-    because \kl{natural binomial polynomials} are increasing.
-    Furthermore, \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}
-    because $\NSF \subseteq \NPoly$.
-
-    Then, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
+	First, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
 	if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
@@
 		\CorrectPoly$ by \cref{corrected-version:thm}.
-
-    As a consequence, we have proven that \cref{rat-npoly-0:item} $\iff$
-    \cref{rat-npoly-1:item} and \cref{rat-npoly-2:item} $\iff$
-    \cref{rat-npoly-3:item} $\iff$ \cref{rat-npoly-4:item}.
-
-    Finally, \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-4:item} is
-    exactly \cref{natural-binomia-correct:lem}. And \cref{rat-npoly-3:item}
-    $\implies$ \cref{rat-npoly-1:item} is obtained
-    by combining \cref{n-sf-combinatorics:lem} and \cref{natural-binomial-polynomial:lem}.
 \end{proof}
-
-\AP Notice that we can also fully characterize polynomials that can be computed
-using \kl{$\Rel$-rational series}, as they are precisely those computable by a
-difference of two \kl{$\Nat$-rational polynomials}. This drives us to introduce
-the class of \intro{integer binomial polynomials}, which are obtained by
-$\Rel$-linear combinations of products of \kl{binomial monomials} of the form
-$\pbinom{X - p}{k}$ for some $k,p \in \Nat$.
-
-\begin{corollary}
-    \label{integer-binomial-polynomials:cor}
-    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
-    if and only if $P$ is an \kl{integer binomial polynomial}.
-\end{corollary}
-\begin{proof}
-    Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
-    \kl{star-free $\Nat$-polyregular functions}.
-    One can apply \cref{n-sf-combinatorics:lem}
-    to conclude that $P$ coincides with an 
-    \kl{integer binomial polynomial} over large enough inputs,
-    hence that $P$ is such a polynomial.
-\end{proof}

### 2024-08-27 parts/correction.tex
polish

@@
 	\begin{enumerate}
-		\item \label{rat-npoly-1:item}
+		\item \label{rat-npoly-1:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+		\item \label{rat-npoly-2:item}
 		      $P$ is a \kl{strongly natural binomial polynomial} \kl{in base $K$},
-		\item \label{rat-npoly-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-		\item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+		\item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
 		\item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
-		      and for all $\vec{x} \in \Nat^{\vec{X}}$,
-		      $P(\vec{x})$ is a multiple of $\alpha$.
+		      and $P(\Nat^k) \subseteq \Nat$.
 	\end{enumerate}
@@
 		polynomial}. But in particular, there exists a polyregular function $f$
-	that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
+	that represents $\alpha P$ with outputs in $\set{ a }$.
+	Now, let us
 	post-compose this function by the one that maps $a^{k \alpha + r}$ to
-	$a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
+	$a^{k}$, which is a \kl{polyregular function} (but is not
+	a \kl{star-free polyregular function}). We conclude that $P$ is
 	\kl{represented} by an \kl{$\Nat$-polyregular function}.
 
-	We also know that \cref{rat-npoly-2:item} $\implies$
-	\cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
-		polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
-		\CorrectPoly$ by \cref{corrected-version:thm}.
+	Let us now prove by induction on the number of indeterminates that
+	\cref{rat-npoly-2:item} $\implies$
+	\cref{rat-npoly-1:item}. When the input values of $P$ are all
+	greater than $K$, $P$ coincides with a \kl{natural binomial function},
+	which is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}. If some
+	input value $X_i$ is set to a number $x_i \leq K$, then one can leverage
+	the fact that $\restr{P}{X_i = x_i}$ remains a \kl{strongly natural
+		binomial polynomial} \kl{in base $K$}, and we conclude by induction that
+	$\restr{P}{X_i = x_i}$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular
+		function}. Combining these functions, we obtain a \kl{star-free $\Nat$-polyregular
+		function} \kl{representing} $P$.
+
+	Finally, \cref{rat-npoly-1:item} $\implies$
+	\cref{rat-npoly-4:item} is obtained via
+	\cref{corrected-version:thm} by
+	remarking that the
+	\kl{star-free $\Nat$-polyregular functions} have output in $\Nat$ and are
+	closed under multiplication by a constant $\alpha \in \Nat$.
+
+	To conclude, it remains to show that
+	\cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}.
 \end{proof}

### 2024-08-28 parts/correction.tex
youhouhou

--- 2024-08-27 parts/correction.tex
@@
 		\quad
-		\text{ over } (\Nat_{\geq 1})^p
+		\text{ over } (\Nat)^p
 		\quad .
@@
 
-\AP Let us conclude this section by considering polynomials with
-coefficients in $\Rat$, and justify that all the combinatorial work has
-already happened in $\Rel$ and $\Nat$. A first observation is that from
-\cref{n-poly-combinatorics:lem,n-sf-combinatorics:lem}, we know that the
-functions that can be computed by \kl{star-free $\Nat$-polyregular
-	functions} are going to be suitable combinations of \kl{natural binomial
-	functions}. This motivates the following definition of the ``polynomial
-counterpart" of a binomial coefficient, $\intro*\pbinom{X - \ell}{k}$
-defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
-\intro{binomial monomial}, and introduce \intro{natural binomial
-	polynomials} as $\Nat$-linear combinations of products of \kl{binomial
-	monomials}, i.e., of the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i
-	\prod_{j = 1}^k \pbinom{X_j - p_{i,j}}{k_{i,j}}$. Similarly, we introduce
-the class of \intro{integer binomial polynomials}, which are obtained by
-$\Rel$-linear combinations of products of \kl{binomial monomials}.
+\AP Let us conclude this section by considering polynomials with coefficients
+in $\Rat$, and justify that all the combinatorial work has already happened in
+$\Rel$ and $\Nat$. A first observation is that from
+\cref{n-poly-combinatorics:lem,n-sf-combinatorics:lem},
+we know that the functions that can be computed by \kl{star-free
+$\Nat$-polyregular functions} are going to be suitable combinations of
+\kl{natural binomial functions}. This motivates the following definition of the
+``polynomial counterpart" of a binomial coefficient, given two numbers $\ell,k \in \Nat$,
+$\intro*\pbinom{X -
+\ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$,\footnote{
+    In particular, $\pbinom{X - \ell}{k}$ is defined to be $1$ when $k = 0$, 
+    and $X - \ell$ when $k = 1$.
+} that we call a
+\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
+as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
+the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
+- p_{i,j}}{k_{i,j}}$. Similarly, we introduce the class of \intro{integer
+binomial polynomials}, which are obtained by $\Rel$-linear combinations of
+products of \kl{binomial monomials}.
 
@@
 	Then, $P$ is an \kl{integer binomial polynomial} if and only if
-	$P(\Rel^k) \subseteq \Rel$.
+	$P(\Rel^k) \subseteq \Rel$, if and only if $P(\Nat^k) \subseteq \Rel$.
 \end{lemma}
 \begin{proof}
-	Let us first notice that $\pbinom{X}{k}$ is \kl{integer valued}
+	Let us first notice that $\pbinom{X}{k}$ is \kl{integer-valued}
 	when $k \in \Nat$. One way to notice this is that when $X \geq 0$,
@@
 	$X < 0$, then $\pbinom{X}{k} = (-1)^k \pbinom{k - X + 1}{k} = (-1)^k \binom{k - X + 1}{k}$
-	is also an integer. Because \kl{integer-valued} polynomials are closed under products
-	and $\Rel$-linear combination, we conclude that all \kl{integer binomial polynomials} are
+	is also an integer. Because \kl{integer-valued} polynomials are closed under products,
+	$\Rel$-linear combination, and 
+    shifting of domain by an integer,
+    we conclude that all \kl{integer binomial polynomials} are
 	\kl{integer-valued}.
@@
 	Combined with \cref{discrete-taylor:eq}, we conclude that $P$ is an \kl{integer binomial polynomial}.
+    Remark that $\Delta_{\alpha} P (0, \dots, 0)$ only ever involves evaluation of $P$ on 
+    $\Nat^k$, justifying the weakening of the hypothesis.
 \end{proof}
@@
 
-\AP We say that a \kl{natural binomial polynomial} is \intro{in base $K$} if
-all its \kl{binomial monomials} are of the form $\pbinom{X - K}{d}$ for some $d
-	\in \Nat$. To address the issues raised by the previous examples, we introduce
-the notion of \intro{strongly natural binomial polynomials} \kl{in base $K$},
-as the polynomials $P \in \Rat[X]$ such that for all partial valuation $\nu
-	\colon \topartial \set{0, \dots, K}$, $\restr{P}{\nu}$ is a \kl{natural
-	binomial polynomial} \kl{in base $K$}. This leads to the following
-characterization.
+\AP To address the issues raised by the previous examples, we introduce the
+notion of \intro{strongly natural binomial polynomials}, as the polynomials $P
+\in \Rat[X]$ such that for all partial valuation $\nu \colon \topartial \Nat$,
+$\restr{P}{\nu}$ is a \kl{natural binomial polynomial}. This leads to the
+following characterization.
 
-\begin{corollary}
+\begin{theorem}
 	\label{decide-rat-poly-npoly:cor}
@@
 	coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
-	such that $\alpha P \in \Rel[\vec{X}]$. Then, there exists a
-	computable $K \in \Nat$, such that the following are equivalent:
+	such that $\alpha P \in \Rel[\vec{X}]$. Then, 
+    the following are equivalent:
 	\begin{enumerate}
-		\item \label{rat-npoly-1:item}
-		      $P$ is a \kl{strongly natural binomial polynomial} \kl{in base $K$},
-		\item \label{rat-npoly-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
-		\item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
-		\item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
-		      and for all $\vec{x} \in \Nat^{\vec{X}}$,
-		      $P(\vec{x})$ is a multiple of $\alpha$.
+		\item \label{rat-npoly-1:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
+		\item \label{rat-npoly-2:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
+		\item \label{rat-npoly-3:item}
+		      $P$ is a \kl{strongly natural binomial polynomial},
+        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$ 
+            and $P$ is \kl{integer-valued}.
 	\end{enumerate}
 	In particular, the properties are decidable.
-\end{corollary}
+\end{theorem}
 \begin{proof}
-	First, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
-	if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
-		polynomial}. But in particular, there exists a polyregular function $f$
-	that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
-	post-compose this function by the one that maps $a^{k \alpha + r}$ to
-	$a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
-	\kl{represented} by an \kl{$\Nat$-polyregular function}.
 
-	We also know that \cref{rat-npoly-2:item} $\implies$
-	\cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
-		polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
-		\CorrectPoly$ by \cref{corrected-version:thm}.
+    For the implication \cref{rat-npoly-1:item} $\implies$
+    \cref{rat-npoly-4:item}, we obtain $\alpha P \in \CorrectPoly$ via
+    \cref{corrected-version:thm} by remarking that the \kl{star-free
+    $\Nat$-polyregular functions} have output in $\Nat$ and are closed under
+    multiplication by a constant $\alpha \in \Nat$. The fact that $P$ is
+    \kl{integer-valued} follows from
+    \cref{integer-binomial-polynomials:cor} and the fact that
+    $\NSF \subseteq \ZPoly$.
+
+    Then, \cref{rat-npoly-4:item} $\implies$
+    \cref{rat-npoly-3:item}, because if $\alpha P \in \CorrectPoly$,
+    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ by
+    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
+    to obtain a number $s \geq 1$ (that only depends on $P$) such
+    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
+    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
+    over $\Nat^k$. Let us write the
+    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
+    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
+    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
+    $\card{\alpha} \leq d_2$.
+    Now, because of the equality 
+    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
+    we conclude that we can find a $K \in \Nat$ and coefficients
+    $f_{\alpha} \in \Nat$ such that 
+    $F(X_1, \dots, X_k) =
+    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
+    Now, remark that $P$ is an \kl{integer-valued} polynomial of
+    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
+    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
+    \in \Rel$ for all $\card{\alpha} \leq d$
+    (\cref{integer-binomial-polynomial:lem}). 
+    As a consequence,
+    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
+    and 
+    by identifying coefficients in the previous two expressions, we obtain
+    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
+    Finally,
+    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
+    and therefore
+    $\lambda P (X_1 + s, \dots, X_k + s) =
+    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
+    In particular,
+    we get
+    \begin{equation*}
+        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
+        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
+    \end{equation*}
+    And we conclude by term identification
+    by term identification
+    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
+    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
+    and $P$
+    is a \kl{natural binomial polynomial}.
+    Remark that this reasoning remains true whenever we partially apply $P$ to some values
+    in $\Nat$, as it amounts to 
+    partially applying the function $f$. We have proven that 
+    $P$ is a \kl{strongly natural binomial polynomial}.
+
+    Let us now prove by induction on the number of indeterminates that
+    \cref{rat-npoly-3:item} $\implies$
+    \cref{rat-npoly-2:item}. Note that by construction, there exists a
+    number $K \in \Nat$ such that when the input values of $P$ are all greater
+    than $K$, $P$ coincides with a \kl{natural binomial function}, which is
+    itself \kl{represented} by a \kl{star-free $\Nat$-polyregular function}. If
+    some input value $X_i$ is set to a number $x_i \leq K$, then one can
+    leverage the fact that $\restr{P}{X_i = x_i}$ remains a \kl{strongly
+    natural binomial polynomial} to conclude by induction that $\restr{P}{X_i =
+    x_i}$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function}.
+    Combining these, we obtain a \kl{star-free $\Nat$-polyregular function}
+    \kl{representing} $P$.
+
+    Finally, the implication  \cref{rat-npoly-2:item} $\implies$ \cref{rat-npoly-1:item} is
+    immediate as $\NSF \subseteq \NPoly$.
 \end{proof}

### 2024-09-02 globals/metadata.tex
wohooh

@@
 \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+\relatedversiondetails[linktext={arXiv:2404.02232}, cite={lopez2024com:arxiv}]{Full Version}{https://arxiv.org/abs/2404.02232}
 
@@
 \newcommand{\acknowledge}{
-    I would like to thank Gaëtan Douéneau for his help and support.
 }
@@
 of many investigations in the context of string-to-string functions
-in the last decade. 
+during the last decade. 
     \end{abstract}
@@
 \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CSL 2025}
-\EventAcronym{ICALP}
-\EventYear{2024}
+\EventShortTitle{Event Short Title}
+\EventAcronym{Event Acronym}
+\EventYear{2025}
 \EventDate{December 24--27, 2016}

### 2024-09-02 parts/appendix.tex
wohooh

@@
     to $f_+$ (resp. to $f_-$).
-    Finally, let $\omega \defined \ppcm(\omega_+, \omega_-)$.
+    Finally, let $\omega$ be the least common multiple
+    of $\omega_+$ and $\omega_-$, i.e. $\omega \defined \ppcm(\omega_+, \omega_-)$.
 
@@
     associated to $f_+$ and $f_-$,
-    such that $X_1, \dots, X_p \geq 1$:
+    such that $x_1, \dots, x_p \geq 1$:
     \begin{equation*}
         f\left(
-        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right) 
-        = F_r^+(X_1, \dots, X_p) - F_r^-(X_1, \dots, X_p) \quad .
+        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right) 
+        = F_r^+(x_1, \dots, x_p) - F_r^-(x_1, \dots, x_p) \quad .
     \end{equation*}
@@
     $f\left(
-        a_1^{\omega X_1 + r_1} \cdots a_n ^{\omega X_n + r_n}\right)$
+        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right)$
     coincides with an \kl{integer binomial polynomial},

### 2024-09-02 parts/correction.tex
wohooh

@@
 %! lang = en-US
-\subsection{The Corrected Theorem}
-\label{sec:proof}
 
-\AP
-The counterexample provided by \cref{def:bad-polynomial} relies on the fact
-that $\CoveredPoly$ is not stable under fixing indeterminates, while
-\kl{$\Nat$-polyregular functions} are.
-Indeed, the polynomial $\BadPoly$ introduced in
-\cref{def:bad-polynomial} satisfies $\BadPoly(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative
-coefficient for a \kl{maximal monomial}, namely $-2XY$.
-In this section, we prove that closing
-$\CoveredPoly$ under variable assignments is enough to recover from
-\cref{karh:thm}.
-We use the following notation to fix the value of some indeterminate, if
-$P(X,Y)$ is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is
-the polynomial $P(1,Y) \in \Rel[Y]$. More generally, if $\nu$ is a partial
-function from $\vec{X}$ to $\Nat$, written $\nu \colon \vec{X} \topartial
-	\Nat$, the restriction $\restr{P(\vec{X})}{\nu}$ is the polynomial with
-indeterminates $\vec{Y} \defined \vec{X} - \dom(\nu)$ obtained by fixing the
-variables of the domain of $\nu$.
+\AP We will discuss at the end of \cref{sec:poly-to-n-poly}
+why the
+example of \cref{thm:counter-example} is minimal in the number of
+indeterminates, which first requires us to provide a \emph{correct} analogue of
+\cref{karh:thm}. The
+counterexample provided by \cref{def:bad-polynomial}
+relies on the fact that
+$\CoveredPoly$ is not stable under fixing indeterminates, while
+\kl{$\Nat$-rational polynomials} are. Indeed, the polynomial $\BadPoly$
+introduced in \cref{def:bad-polynomial}
+satisfies $\BadPoly(X,Y,1) = 3X^2 +
+3Y^2 - 2XY$, which has a negative coefficient for a \kl{maximal monomial},
+namely $-2XY$. In this section, we prove that closing $\CoveredPoly$ under
+variable assignments is enough to recover from
+\cref{karh:thm}. We
+use the following notation to fix the value of some indeterminate, if $P(X,Y)$
+is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is the
+polynomial $P(1,Y) \in \Rel[Y]$. More generally, if $\nu$ is a partial function
+from $\vec{X}$ to $\Nat$, written $\nu \colon \vec{X} \topartial \Nat$, the
+restriction $\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminates
+$\vec{Y} \defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the
+domain of $\nu$.
 
@@
 
-\subsubsection{From $\Nat$-polyregular functions to polynomials}
+Let us briefly remark that \cref{corrected-version:thm} is surprising given the
+fact that is not possible to decide whether a polynomial $P \in \Rel[\vec{X}]$
+is \kl{non-negative} (\cref{undecidable-non-negative:lem}), by
+reduction to the undecidability of Hilbert's Tenth Problem
+\cite{HILB1902,MATI1970}.
 
-\AP Let us now prove that \kl{$\Nat$-rational polynomials} are in
-$\CorrectPoly$. This follows from the correct implication in the statement of
-\cref{karh:thm}, but we provide a self-contained proof using a refinement of
-the classical combinatorial arguments for $\ZPoly$ \cite[Lemma 4.16]{CDTL23}
-and $\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in our
-upcoming \cref{n-poly-combinatorics:lem} a finer statement that has two main
-goals: first, it formulates the traditional \emph{pumping argument} as a
-composition property using \kl{(noncommutative output) polynomial functions},
-allowing us to use the same formalism in the upcoming
-\cref{k-combinatorial:def,ultimately-polynomial:def} and highlight the role of
-\kl{commutative} \kl{polyregular functions} in the broader study of
-\kl{polyregular functions}; then we also give a precise shape of the
-polynomials that arise from such \emph{pumping arguments}, which was lacking in
-former similar statements where $\Rat[X_1, \dots, X_n]$ was used as an over
-approximation.
+\begin{lemma}
+    \label{undecidable-non-negative:lem}
+    Let $P \in \Rel[\vec{X}]$ be a polynomial. Then, checking whether $P$ is
+    \kl{non-negative} is undecidable.
+\end{lemma}
+\begin{proof}
+    Recall that by the undecidability of the Hilbert's Tenth Problem \cite[Problem 10 page 22]{HILB1902}, it is
+    undecidable whether a given polynomial $P \in \Rel[\vec{X}]$ has a solution in
+    $\Rel^n$ \cite{MATI1970,DAVIS1973}.
+    The result follows by observing that
+    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^n$ if and only 
+    if $P^2 - 1$ is \kl{non-negative}. 
+\end{proof}
+
+The proof of \cref{corrected-version:thm} is divided into two parts. First, we
+provide in \cref{sec:n-poly-to-poly}
+a fine
+combinatorial understanding of what functions can be computed by functions in
+$\NPoly$ and $\ZPoly$. This will allow us to prove that \kl{$\Nat$-polyregular
+functions} are in $\CorrectPoly$ (\cref{n-rat-correct:lem}).
+Then, in \cref{sec:poly-to-n-poly}
+we will
+show how to compute polynomials in $\CorrectPoly$ using $\NPoly$
+(\cref{lem:correct-to-n-rat}).
+With this in hand, we will next generalize these results
+to polynomials in $\Rat[\vec{X}]$ in \cref{sec:rel-to-rat}.
+
+\subsection{From $\Nat$-polyregular functions to polynomials}
+\label{sec:n-poly-to-poly}
+
+\AP Let us prove that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
+This fact follows from the correct implication in the statement of
+\cref{karh:thm}, but
+we provide a self-contained proof using a refinement of the classical
+combinatorial \emph{pumping arguments} for $\ZPoly$ \cite[Lemma 4.16]{CDTL23} and
+$\NPoly$ \cite[Lemma 5.37]{DOUE23}. We take extra care to reprove in our
+upcoming \cref{n-poly-combinatorics:lem} a strong statement that has
+two main goals. Our first goal is to highlight the role of \kl{commutative}
+\kl{polyregular functions} in the broader study of \kl{polyregular functions},
+which is done by reformulating the traditional pumping argument as a
+composition property involving said functions, which will be reused in the
+upcoming \cref{k-combinatorial:def,ultimately-polynomial:def}
+in
+\cref{beyond-polynomials:sec}. Our second goal
+is to give a precise shape of the functions that arise from such \emph{pumping
+arguments}, which was lacking in former similar statements.
 
@@
 
-\AP Remark that our second goal is at odds with our restriction to polynomials
-with coefficients in $\Rel$, because of the polynomial $X(X-1)/2$
-(\cref{q-polynomial-n-poly:ex}). However, we argue that up to a suitable change
-of basis for the ring of polynomials, all coefficients will actually be in
-$\Nat$. This relatively simple trick works by considering as ``basic
-polynomials" the \intro{binomial function} of the form $\binom{X - \ell}{k}$,
-where $\ell$ and $k$ are natural numbers. We extend this to \intro{natural
-	binomial functions} that are obtained by considering $\Nat$-linear combinations
-of products of \kl{binomial functions}, that is, we consider functions that
-have the following shape: $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i \prod_{j =
-		1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is
-defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial.
+\AP Our second goal is achieved by understanding that \kl{$\Nat$-polyregular
+functions} essentially compute binomial coefficients, as illustrated by the
+polynomial $X(X-1)/2 = \binom{X}{2}$.
+of \cref{q-polynomial-n-poly:ex}. A \intro{binomial function} is a
+function of the form $\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural
+numbers. We extend this to \intro{natural binomial functions} that are obtained
+by considering $\Nat$-linear combinations of products of \kl{binomial
+functions}, that is, we consider functions that have the following shape:
+$f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \binom{x_j -
+p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is defined to be $0$ when
+$X \leq \ell$, and is therefore not a polynomial. Let us immediately
+prove that \kl{binomial functions}  can be \kl{represented} in $\NSF$,
+generalizing \cref{q-polynomial-n-poly:ex}.
 
@@
 		\quad
-		\text{ over } (\Nat)^p
+		\text{ over } \Nat^p
 		\quad .
@@
 \end{proof}
+
+Because \kl{natural binomial functions} behave as polynomials with
+\kl{non-negative} \kl{maximal monomials} on large enough inputs, we can
+conclude from \cref{n-poly-combinatorics:lem}
+that \kl{$\Nat$-rational polynomials} are in $\CorrectPoly$.
 
@@
 
-\subsubsection{From polynomials to $\Nat$-polyregular functions}
+\subsection{From polynomials to $\Nat$-polyregular functions}
+\label{sec:poly-to-n-poly}
 
@@
 $\CorrectPoly$, and the rest of this section is devoted to proving the converse
-implication. The core of the upcoming \cref{lem:correct-to-n-rat} leverages a
-notion of \intro{discrete derivative} which is built by translating the domain
-of the polynomial. To that end, let us write $\intro*\translate{K}$ for the
-\intro{translation function} that maps a polynomial $P \in \Rel[X_1, \dots,
-		X_n]$ to the polynomial $P(X_1 + K, \dots, X_n + K)$.
+implication. The core of the upcoming \cref{lem:correct-to-n-rat}
+leverages a notion of \intro{discrete derivative} which is built by
+translating the domain of the polynomial. To that end, let us write
+$\intro*\translate{K}$ for the \intro{translation function} that maps a
+polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 + K, \dots,
+X_n + K)$.
 
@@
 
-\subsubsection{From $\Rel$ to $\Rat$}
+\subsection{From $\Rel$ to $\Rat$}
+\label{sec:rel-to-rat}
 
-\AP Let us conclude this section by considering polynomials with coefficients
-in $\Rat$, and justify that all the combinatorial work has already happened in
-$\Rel$ and $\Nat$. A first observation is that from
-\cref{n-poly-combinatorics:lem,n-sf-combinatorics:lem},
-we know that the functions that can be computed by \kl{star-free
+\AP Let us complete our analysis of polynomials \kl{represented} by $\NPoly$ or
+$\ZPoly$ by considering polynomials with coefficients in $\Rat$, and justify
+that all the combinatorial work has already happened in $\Rel$ and $\Nat$. A
+first observation is that from \cref{n-sf-combinatorics:lem}, we
+know that the polynomials that can be computed by \kl{star-free
 $\Nat$-polyregular functions} are going to be suitable combinations of
 \kl{natural binomial functions}. This motivates the following definition of the
-``polynomial counterpart" of a binomial coefficient, given two numbers $\ell,k \in \Nat$,
-$\intro*\pbinom{X -
-\ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$,\footnote{
-    In particular, $\pbinom{X - \ell}{k}$ is defined to be $1$ when $k = 0$, 
-    and $X - \ell$ when $k = 1$.
-} that we call a
+``polynomial counterpart" of a binomial coefficient, given two numbers $\ell,k
+\in \Nat$, $\intro*\pbinom{X - \ell}{k}$ defined as $(X - \ell) \cdots (X -
+\ell - k) / k!$,\footnote{ In particular, $\pbinom{X - \ell}{k}$ is defined to
+be $1$ when $k = 0$, and $X - \ell$ when $k = 1$. } that we call a
 \intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
@@
 
-As an immediate corollary, we completely characterize the class of
-polynomials with rational coefficients that can be computed by
-\kl{star-free $\Nat$-polyregular functions}.
+As an immediate corollary,  we completely characterize the class of polynomials
+in $\Rat[\vec{X}]$ that are \kl{represented} by $\ZPoly$ as the \kl{integer
+binomial polynomials}.
 
-\begin{corollary}
+\begin{theorem}
 	\label{integer-binomial-polynomials:cor}
@@
 		\item \label{int-bin-2:item} $P$ \kl{represented} by a \kl{$\Rel$-polyregular function},
-		\item \label{int-bin-3:item} $P(\Rel^k) \subseteq \Rel$.
+        \item \label{int-bin-3:item} $P$ is \kl{integer-valued}.
 	\end{enumerate}
 	These properties are furthermore decidable.
-\end{corollary}
+\end{theorem}
 \begin{proof}
@@
 
-Obtaining an analogue of \cref{integer-binomial-polynomials:cor}
-for \kl{$\Nat$-polyregular functions} is more challenging. The
-naïve approach would be to consider some variation around \kl{natural binomial
-	polynomials}. A first observation is that the class of products of the form
-$\prod_{i = 1}^n \pbinom{X_i - p_i}{k_i}$ is a generating family for
-$\Rat[\vec{X}]$ but is not linearly independent. As a consequence, there may be
-several ways to write a given polynomial, and as illustrated by
-\cref{natural-binomial-polynomial-basis:ex}, some of them will
-exhibit negative coefficients.\footnote{This was not a problem for \kl{integer
-		binomial polynomials}, for which the change of basis mapped coefficients in
-	$\Rel$ to coefficients $\Rel$.} Furthermore, there are some \kl{natural
-	binomial polynomials} that take negative values on positive inputs, as
-illustrated by \cref{natural-binomial-polynomial-negative:ex}.
-Finally, restricting our attention to \kl{non-negative} \kl{natural binomial
-	polynomials} suffers from the same kind of issues as
-\cref{karh:thm},
-witnessed by \cref{natural-binomial-polynomial-positive-bad:ex}.
+\AP Obtaining an analogue of \cref{integer-binomial-polynomials:cor}
+for \kl{$\Nat$-polyregular functions} requires a bit more work,
+as polynomials in $\Rat[\vec{X}]$ that are \kl{represented} by $\NPoly$ are not
+exactly \kl{natural binomial polynomials}
+(\cref{natural-binomial-polynomial-negative:ex,natural-binomial-polynomial-basis:ex,natural-binomial-polynomial-positive-bad:ex}).
+To address the issues raised by these examples, we introduce the notion of
+\intro{strongly natural binomial polynomials}, as the polynomials $P \in
+\Rat[X]$ such that for all partial valuation $\nu \colon \topartial \Nat$,
+$\restr{P}{\nu}$ is a \kl{natural binomial polynomial}, which 
+characterizes the class of polynomials that are \kl{represented} by $\NPoly$
+(\cref{decide-rat-poly-npoly:cor}).
 
@@
 \end{example}
-
-\AP To address the issues raised by the previous examples, we introduce the
-notion of \intro{strongly natural binomial polynomials}, as the polynomials $P
-\in \Rat[X]$ such that for all partial valuation $\nu \colon \topartial \Nat$,
-$\restr{P}{\nu}$ is a \kl{natural binomial polynomial}. This leads to the
-following characterization.
 
@@
 \end{proof}
+
+Let us remark that \cref{decide-rat-poly-npoly:cor} shows that the
+class of polynomials \kl{represented} by $\NPoly$ is the same as the class of
+polynomials \kl{represented} by $\NSF$, which is a non-trivial statement that
+will be reused in the study of more general \kl{commutative} functions in
+$\ZPoly$.

### 2024-09-02 parts/counter.tex
wohooh

@@
 \label{polynomials:sec}
-%\subsection{The counterexample}
-\label{sec:c-example}
 
-\AP Before deciding $\NPoly$ inside $\ZPoly$ (which will be done in
-\cref{beyond-polynomials:sec}), let us restrict our attention to simpler
-functions, namely polynomials. A polynomial $P \in \Rel[X_1, \dots, X_n]$ is an
+\AP In this section, we will completely characterize which polynomials in
+$\Rat[\vec{X}]$ that are \kl{represented} by \kl{$\Nat$-polyregular functions}
+(resp. \kl{$\Rel$-polyregular functions}). To that end, we start by
+characterizing these classes for polynomials in $\Rel[\vec{X}]$.
+We say that a polynomial $P \in \Rel[X_1, \dots, X_n]$ is an
 \intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
@@
 \kl{$\Nat$-polyregular function}, because it has \kl{polynomial growth}
-(\cref{polyregular-polynomial-growth:lemma}). It is an easy check that
+(\cref{polyregular-polynomial-growth:lemma}).
+The analogue notion of \intro{$\Rel$-rational polynomials} is not
+of particular interest, since every polynomial $P \in \Rel[X_1, \dots, X_n]$ is
+\kl{represented} by a \kl{$\Rel$-polyregular function}. 
+It is an easy check that
 polynomials with coefficients in $\Nat$ are \kl{$\Nat$-rational polynomials}
@@
 a polynomial with negative coefficients that is an \kl{$\Nat$-rational
-polynomial}. The analogue notion of \intro{$\Rel$-rational polynomials} is not
-of particular interest, since every polynomial $P \in \Rel[X_1, \dots, X_n]$ is
-\kl{represented} by a \kl{$\Rel$-polyregular function}. Our goal is to decide
-if a \kl{$\Rel$-rational polynomial} (so, any polynomial) is an
-\kl{$\Nat$-rational polynomial}, and if so, to compute an
-\kl{$\Nat$-polyregular function} that \kl{represents} it. This particular
-problem was claimed to be solved in \cite{KARH77}, using the \cref{karh:def} to
-characterize \kl{$\Nat$-rational polynomials}, as restated in \cref{karh:thm}.
+polynomial}. 
+The problem of characterizing \kl{$\Nat$-rational polynomials} was claimed to
+be solved in \cite{KARH77}, using the \cref{karh:def}
+to  characterize
+\kl{$\Nat$-rational polynomials}, as restated in \cref{karh:thm}.
 
@@
 \begin{example}
-    \label{negative-not-nrat:ex}
     \label{negative-but-npoly:ex}
@@
 \end{proof}
-
-We will discuss at the end of \cref{sec:proof} why the example of
-\cref{thm:counter-example} is minimal in the number of indeterminates, which
-first requires us to provide a \emph{correct} analogue of \cref{karh:thm}.

### 2024-09-02 parts/intro.tex
counterexamples
wohooh

--- 2024-08-15 parts/intro.tex
@@
 theorem} of \cite{DOUE21}, which fails in the general case
-\cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary output}
-in the case of \kl{polyregular functions} can often be lifted to outputs in the
-free commutative semigroup generated by a finite alphabet $\Gamma$. In this
-sense, the shift to \emph{single letter output alphabet} is really about
-enforcing commutativity on the output. However, even with the restriction to
+\cite{BOJA22,KLEP23}. However, even with the restriction to
 single letter output alphabets, the decidability of \kl{star-free polyregular
@@
 allowing negative outputs. This remains a subclass of \kl{rational series},
-more precisely, of the class of \intro{$\Rel$-rational series}
-($\intro*\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
+more precisely, of the class of \kl{$\Rel$-rational series}
+($\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
 been shown to be decidable \cite{CDTL23}. Let us underline that several
@@
 \cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
-is an open problem (see e.g. \cite{KARH77}), that remains open even in the
+is also an open problem (see e.g. \cite{KARH77}), that remains open even in the
 restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
@@
 the functions remains unchanged when the letters of the input word are
-permuted. In this restricted setting, we answer positively to \cite[Conjecture
-7.61]{DOUE23} (stating that $\NSF = \NPoly \cap \ZSF$), and prove that the
+permuted. These functions from $\Sigma^*$ to $\Rel$ can be understood as
+functions from $\Nat^{\card{\Sigma}}$ to $\Rel$, and are therefore easier to
+study using classical mathematical tools. The simplest kind of functions in
+this class are actual polynomial functions, which were studied by Karhumäki in
+\cite{KARH77}. Our first contribution is to provide a counter-example to the
+main theorem of \cite{KARH77} that claimed to characterize the (multivariate)
+polynomials in $\Rat[\vec{X}]$ that can be computed by some \kl{$\Nat$-rational
+series} (\cref{thm:counter-example}). Our second contribution is to correct
+this error, and provide (effective) characterizations of polynomials in
+$\Rat[\vec{X}]$ that can be computed by \kl{$\Rel$-polyregular functions}
+(\cref{integer-binomial-polynomials:cor}), and by \kl{$\Nat$-polyregular
+functions} (\cref{decide-rat-poly-npoly:cor}). Our third contribution
+of this paper is to answer \cite[Conjecture 7.61]{DOUE23} (stating that $\NSF =
+\NPoly \cap \ZSF$) positively in the \kl{commutative} case, and prove that the
 membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
@@
 all arrows in the upper left square of \cref{previously-known-inclusions:fig}
-are decidable in the \kl{commutative} setting. Furthermore, we characterize
-\kl{commutative} \kl{$\Nat$-polyregular functions} as the class of
-\kl{$\Rel$-polyregular functions} that are \kl{combinatorial} (defined page
-\kpageref{combinatorial}), which is a semantic property akin to the notion of
-\kl{ultimate polynomiality} that characterizes $\ZSF$ inside $\ZPoly$
-\cite{CDTL23}. As a byproduct of our analysis, we provide a counter example to
-the characterization of polynomials computed by \kl{$\Nat$-rational series}
-given by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
+are decidable in the \kl{commutative} setting.
 
-As a byproduct of our fine-grained of (commutative) $\Nat$ and $\Rel$
-polyregular functions, we fully characterize the polynomials that they can
-express. Note that even \kl{$\Nat$-polyregular functions} can compute
-polynomials with coefficients in $\Rat$, such as the polynomial $X(X-1)/2$ (see
-\cref{q-polynomial-n-poly:ex}). However, not all polynomials $P \in
-\Rat[\vec{X}]$ can be expressed using polyregular functions, and we prove that
-it is the case only for those obtained as $\Nat$ (resp. $\Rel$) linear
-combinations of products binomial coefficients $\binom{X_1 - p_1}{k_1} \times
-\cdots \times \binom{X_n - p_n}{k_n}$.\footnote{For the reader perplexed by the
-    fact that $\binom{X_n - p_n}{k_n}$ is \emph{not} a polynomial, we redirect to page 
-\kpageref{binomial monomial}.}
-
-\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
-\kl{$\Nat$-polyregular functions} (resp. $\Rel$-polyregular functions), show
-that one can decide if a function $f \in \ZPoly$ is \kl{commutative}
-(\cref{decidable-commutative-poly:lemma}), and give the basic definitions used
-on multivariate polynomials. In \cref{polynomials:sec}, we focus on
-\emph{polynomials}, by introducing the \cref{karh:thm} of \cite{KARH77},
-providing a counter-example in \cref{thm:counter-example}, and an alternative
-characterization in \cref{corrected-version:thm}. Also in
-\cref{polynomials:sec}, we provide characterizations of $\Rat$-polynomials that
-can be expressed respectively by \kl{$\Nat$-polyregular functions}
-(\cref{corrected-version:thm} \cref{corrected-0:item}) and
-\kl{$\Rel$-polyregular functions} (\cref{integer-binomial-polynomials:cor}).
-Finally, in \cref{beyond-polynomials:sec}, we leverage the results over
-polynomials to decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
+\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we provide a
+combinatorial definition of \kl{$\Nat$-polyregular functions} (resp.
+$\Rel$-polyregular functions), show that one can decide if a function $f \in
+\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
+the basic definitions used on multivariate polynomials. In
+\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
+\cref{karh:thm} of \cite{KARH77}, and providing a counter-example in
+\cref{thm:counter-example}. We then correct this result in
+\cref{sec:n-poly-to-poly,sec:poly-to-n-poly,sec:rel-to-rat} by providing first
+a characterization of polynomials in $\Rel[\vec{X}]$ that can be computed by
+\kl{$\Nat$-polyregular functions} (\cref{corrected-version:thm}), and then
+extending this result to $\Rat[\vec{X}]$ (\cref{decide-rat-poly-npoly:cor}).
+Along the way, we obtain a characterization of polynomials in $\Rat[\vec{X}]$
+that can be obtained through \kl{$\Rel$-rational series}
+(\cref{integer-binomial-polynomials:cor}). Finally, in
+\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
+decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
 (\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture}

### 2024-09-02 parts/prelim.tex
wohooh

@@
 \times \card[b]{w}$.
+
+\subsection{Polynomials} \AP All polynomials considered in this paper have
+coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
+\Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for all non-negative
+integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1, \dots, n_k)$ of
+the polynomial is non-negative. Beware that we do not consider negative values
+as input, as the numbers $n_i$ will ultimately count the number of occurrences
+of a letter in a word. As an example, the polynomial $(X - Y)^2$ is
+\kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
+2X$ is not.
+
+\AP Let us now give some vocabulary on polynomials with multiple indeterminates
+over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.
+For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
+\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
+\dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
+Every polynomial $P \in \Rel[X_1, \dots, X_n]$ decomposes uniquely
+into a sum of \kl{monomials}.
+
+\AP A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
+$T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
+\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
+decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
+\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
+ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
++ X + Y$, the set of \reintro{maximal monomials}, written
+$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  Y^2}$.  In the polynomial
+$P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials} are $X^2$ and
+$Y^2$.
 
@@
 
-\AP In full generality, \cref{size-of-word-nsf:ex,q-polynomial-n-poly:ex} can be generalized to demonstrate that functions
-of the following functions can be \kl{represented} by a \kl{commutative}
-\kl{star-free $\Nat$-polyregular function}: $x \mapsto \binom{x - p}{k}$ for $p
-\in \Nat$ and $k \in \Nat$, where $\binom{n}{k}$ is the binomial coefficient,
-that is the number of ways to choose $k$ elements among $n$ elements.
-
 \AP One of the appeals of $\NPoly$ and $\ZPoly$ are the numerous
@@
     $\NSF[k]$, $\ZSF[k]$):
-    \begin{itemize}
-        \item $f + g \defined w \mapsto f(w) + g(w)$,
-        \item $f \times g \defined w \mapsto f(w) \times g(w)$,
-        \item $\ind{L} \times f$ 
-    \end{itemize}
+    $f + g \defined w \mapsto f(w) + g(w)$,
+    $f \times g \defined w \mapsto f(w) \times g(w)$,
+    $\ind{L} \times f$.
     Furthermore, the above constructions preserve \kl{commutativity}.
@@
     $f \circ c = f = f \circ t$.
-    When $f$ is a rational series,
-    $f \circ c$ and $f \circ t$ are both rational series that can be
+    When $f$ is a \kl{rational series},
+    $f \circ c$ and $f \circ t$ are both \kl{rational series} that can be
     effectively computed from $f$, and since equivalence
@@
 \end{proof}
-
-\subsection{Polynomials} \AP All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
-\Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for all non-negative
-integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1, \dots, n_k)$ of
-the polynomial is non-negative. Beware that we do not consider negative values
-as input, as the numbers $n_i$ will ultimately count the number of occurrences
-of a letter in a word. As an example, the polynomial $(X - Y)^2$ is
-\kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
-2X$ is not.
-
-\AP Let us now give some vocabulary on polynomials with multiple indeterminates
-over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.
-For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
-\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
-\dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
-Every polynomial $P \in \Rel[X_1, \dots, X_n]$ decomposes uniquely
-into a sum of \kl{monomials}.
-
-\AP A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
-$T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
-\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
-decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
-\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
-ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
-+ X + Y$, the set of \reintro{maximal monomials}, written
-$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  Y^2}$.  In the polynomial
-$P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials} are $X^2$ and
-$Y^2$.

### 2024-09-02 parts/starfree.tex
wohooh

@@
 In this section, we leverage the decidability results of \cref{polynomials:sec}
-to decide $\NPoly$ inside $\ZPoly$ and $\NSF$ inside $\NPoly$, both under the
-extra assumption of \kl{commutativity}. The hypothesis of \kl{commutativity}
-reduces $\ZPoly$ to a combination of a finite number of polynomials, arranged
-by counting modulo some numbers. This characterization will be proven in
-\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
-procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
-\cref{decompose-polynomial:lem} can be understood as a refinement of
-\cref{n-poly-combinatorics:lem} for \kl{commutative} functions.
+to decide membership in $\NPoly$
+inside $\ZPoly$ and membership in $\NSF$ inside $\NPoly$, both under the extra
+assumption of \kl{commutativity}.
 
-\AP To actually state the lemma, we will need some notations that will
+\subsection{Deciding $\NPoly$ inside $\ZPoly$}
+
+To characterize $\NPoly$ inside $\ZPoly$ we introduce the notion of
+\kl{$(k,\Nat)$-combinatorial} function
+\cref{k-combinatorial:def}, following the spirit of previous
+characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
+pumping arguments} \cite{DOUE21,DOUE22,CDTL23}.
+
+\begin{definition}
+    \label{k-combinatorial:def}
+    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
+    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
+    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
+    such that
+    for all
+    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
+    there exists a \kl{strongly natural binomial polynomial} $P$
+    satisfying:
+    \begin{equation*}
+        f \circ q(\omega X_1,\dots, \omega X_k)
+        = 
+        P
+        \quad 
+        \text{ over } (\Nat_{\geq 1})^k
+        \quad .
+    \end{equation*}
+\end{definition}
+
+\AP
+\Cref{k-combinatorial:def} is meaningful under the hypothesis of
+\kl{commutativity}, because the latter reduces $\ZPoly$ to a combination of a
+finite number of polynomials, arranged by counting modulo some numbers. This
+characterization is stated in \cref{decompose-polynomial:lem}, and will
+be the key ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$.
+To actually state the lemma, we will need some notations that will
 formalize the intuitive idea of identifying tuples of integers modulo some
@@
 
-In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
-of \emph{polynomial pumping arguments} \cite{DOUE21,DOUE22,CDTL23}, we will
-provide a semantic property in the form of \cref{k-combinatorial:def}, that is
-conjectured to characterize $\NPoly$ inside $\ZPoly$ in the general setting.
-
-\begin{definition}
-    \label{k-combinatorial:def}
-    Let $k \in \Nat$, and $f \colon \Sigma^* \to \Rel$
-    be a \kl{$\Rel$-polyregular function}. The function $f$ is 
-    \intro{$(k,\Nat)$-combinatorial} if there exists $\omega \in \Nat$,
-    such that
-    for all
-    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
-    there exists a polynomial $P \in \CorrectPoly$,
-    satisfying:
-    \begin{equation*}
-        f \circ q(\omega X_1,\dots, \omega X_k)
-        = 
-        P
-        \quad 
-        \text{ over } (\Nat_{\geq 1})^k
-        \quad .
-    \end{equation*}
-\end{definition}
-\begin{conjecture}
-    \label{npoly-zpoly:conjecture}
-    Let $k \in \Nat$.
-    A function $f \in \ZPoly[k]$
-    belongs to $\NPoly[k]$ if and only if
-    it is \kl{$(k,\Nat)$-combinatorial}.
-    In particular,
-    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
-\end{conjecture}
-
 \begin{theorem}
     \label{decidable-n-poly:thm}
-    Let $k \in \Nat$, and 
-    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
-    The following are equivalent:
+    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
+    Then, the following are equivalent:
     \begin{enumerate}
@@
     from 
-    \cref{n-poly-combinatorics:lem} and \cref{corrected-version:thm}.
-
-    For the converse implication, we use \cref{decompose-polynomial:lem} to
+    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
+    For the converse implication, we use \cref{decompose-polynomial:lem} 
     effectively compute a polynomial representation of $f$. That is, we obtain
@@
     \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
-    such that for all \kl{polynomial functions} $q$, there exists an
-    \kl{integer binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
+    such that for all \kl{polynomial functions} $q$, there exists a
+    \kl{strongly natural binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
     = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
@@
     1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
-    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists
-    a polynomial $P \in \CorrectPoly$ such that for all $x_1, \dots, x_n
-    \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k) = P(x_1,
-    \dots, x_k)$ holds. In particular, we conclude that over $(\Nat_{\geq
-    1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They are therefore
-    equal as multivariate polynomials, and we conclude that $P_t$ belongs to
-    $\CorrectPoly$, and in particular is an \kl{$\Nat$-rational polynomial}
-    thanks to \cref{corrected-version:thm}.
+    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists a
+    \kl{strongly natural binomial polynomial} $P$ such that for all $x_1,
+    \dots, x_n \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k)
+    = P(x_1, \dots, x_k)$ holds. In particular, we conclude that over
+    $(\Nat_{\geq 1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They
+    are therefore equal as multivariate polynomials, and we conclude that $P_t$
+    is a \kl{strongly natural binomial polynomial} \kl{$\Nat$-rational
+    polynomial}.
 
@@
     computing the \kl{$\omega$-type} of the input (which is possible in
-    $\MSO$), and then computing the suitable \kl{$\Nat$-rational polynomial},
-    which is possible in $\NPoly$.
+    $\MSO$), and then computing the suitable \kl{strongly natural binomial polynomials}
+    which is possible in $\NPoly$
+    thanks to \cref{decide-rat-poly-npoly:cor}.
 \end{proof}
+
+\subsection{Deciding $\NSF$ inside $\NPoly$}
 
@@
 \cite[Claim 7.45, Lemma 7.53]{DOUE23}. We include a self-contained proof
-hereafter for readability and didactic purposes. The proof follows essentially
-the same scheme as the base case of \cref{n-poly-combinatorics:lem}.
+hereafter for readability and didactic purposes.
 
@@
 functions, it was conjectured that the semantic property of being
-\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$, a
-conjecture that is restated hereafter.
+\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$ (\cref{zsf-nsf:conjecture}).
 
@@
 \begin{proof}
-    The implication \cref{zsf-npoly-nsf:thm:zsf:item} $\Rightarrow$
+    The implication \cref{zsf-npoly-nsf:thm:nsf:item} $\Rightarrow$
     \cref{zsf-npoly-nsf:thm:zsf:item} is immediate since $\NSF \subseteq \ZSF$.

### 2024-09-04 reviews/rebuttal-icalp.md
rebuttal

--- 2024-03-27 reviews/rebuttal.md
@@
 
-Before answering to the questions of the reviewers, we would like to point out
-that the summary of Reviewer 1 is slightly misleading as the paper does not
-manage to decide whether a function is "$\mathbb{N}$-polyregular" in general,
-but only does so in the case where the function is assumed to be *commutative*
-(which is a strong restriction). Without the assumption of commutativity, the
-problem remains open. We will make sure to put more emphasis on this point.
-
----
-
 It is our understanding that a significant part of the reviewers' concerns
-relates to the high number of typos and grammatical errors in the manuscript.
-We sincerely apologize for these mistakes that made reading this paper
-difficult, and we acknowledge the lack of consistency in the bibliographic
-section and in particular the mix of unicode and non unicode bibtex entries
-which butchered the name of Lê Thành Dũng Tito Nguyên. We will definitely
-correct these issues in the revised version of the manuscript. We decided to
-not respond to specific typos and grammatical errors in this rebuttal and focus
-on more global issues.
-
----
+relates to quality of writing, and especially the high number of typos and
+grammatical errors in the manuscript. We sincerely apologize for these mistakes
+that made reading this paper difficult, and we acknowledge the lack of
+consistency in the bibliographic section and character encoding errors which we
+hope was not perceived as disrespectful. We will definitely correct these
+issues in the revised version of the manuscript.
 
@@
 
-However, it seems reasonable for us to assume that the reader is familiar at
-least with finite monoids, monadic second order logic on words, and weighted
-automata (hence, with rational series) which are well established theories:
-describing them in a few sentences with a reference to a standard book should
-already greatly increase the readability of the paper.
+It seems reasonable for us to assume that the reader is familiar with finite
+monoids, monadic second order logic on words, and weighted automata (hence,
+with rational series) which are well established theories: describing them in a
+few sentences with a reference to a standard book should also greatly
+increase the readability of the paper.
 
-The informal uses of notions such as "growth rate" and "commutativity" in the
-introduction were intended to provide intuitions to the reader, but we
-understand that it was counterproductive and will be more careful in the
-revised version. This is not an isolated issue, and we will make sure to avoid
-such informal statements.
-
-We also understand that whether a proof belongs to the appendix or the main
-body of the paper felt arbitrary, and that the key ingredients of the proofs
-where not clearly identified. While it is difficult to answer to such a comment
-without rewriting the paper, we will try to illustrate on a few examples.
+A third point raised by the reviewers is that whether a proof belongs to the
+appendix or the main body of the paper felt arbitrary, and that the key
+ingredients of the proofs where not clearly identified. While it is difficult
+to answer this question, we will try to illustrate on a few examples.
 
 - Lemma 7 is a simple proof in the case of polyregular functions, and has
-  a sligthly more technical proof[^1] in the case of rational series, but the
+  a slightly more technical proof[^1] in the case of rational series, but the
   main interest of the paper is in polyregular functions, and therefore the
@@
 
-However, we agree that all the proofs of sections 5 and 6 are in appendix
-except for the ones that combine technical lemmas, and a better trade-off
-between the sections will be found in the revised version of the paper. We
-believe that the interesting proofs are those of Lemmas 44, and 45, which
-really leverage the notion of well-quasi-ordering.
+All the proofs of sections 5 and 6 are in appendix except for the ones that
+combine technical lemmas, and a better trade-off between the sections will be
+found in the revised version of the paper. We believe that the interesting
+proofs are those of Lemmas 44, and 45, which really leverage the notion of
+well-quasi-ordering.
 
@@
 
----
+A fourth and important question raised by the reviewers was the relationship
+between this paper and [9], which should be made clearer. Let us try to
+put both papers in context.
 
-To conclude on the relationship between this paper and [9], which should be
-made clearer, let us try to replace both papers in context.
-
-The concern of [9] is about $\mathbb{Z}$-polyregular functions (which are
-a subclass of $\mathbb{Z}$-rational series), defining a notion of
-"aperiodicity" for such functions, and proving that aperiodicity is decidable.
-The work of [9] could be put into context as follows: since the introduction of
-polyregular functions [4,5], there has been a corresponding notion of
-"aperiodic" polyregular functions. However, decidability of aperiodicity
-remains an open question, even in the specific case of unary output (the
-function output words over a single letter alphabet). Notice that polyregular
-functions with unary output are in fact functions from $\Sigma^*$ to
-$\mathbb{N}$ (we write a number in unary).
+The concern of [9] is about $\mathbb{Z}$-polyregular functions (which are a
+subclass of $\mathbb{Z}$-rational series), defining a notion of "aperiodicity"
+for such functions, and proving that aperiodicity is decidable. The work of [9]
+could be put into context as follows: since the introduction of polyregular
+functions [4,5], there has been a corresponding notion of "aperiodic"
+polyregular functions. However, decidability of aperiodicity remains an open
+question, even in the specific case of unary output (the function output words
+over a single letter alphabet). Notice that polyregular functions with unary
+output are in fact functions from $\Sigma^*$ to $\mathbb{N}$ (we write a number
+in unary).
 
@@
 series, it is really about deciding aperiodicity of a "variation" of
-polyregular functions with unary output that allows "negative outputs".
-Unfortunately this "variation" is substantial, and the relationship between
-$\mathbb{Z}$-polyregular functions (of [9]) and $\mathbb{N}$-polyregular
-functions (unary output) remained unclear up to this paper. In particular the
-decidability of aperiodicity for $\mathbb{N}$-polyregular functions could not
-be deduced directly from [9].
+polyregular functions with unary output (elements of $\mathbb{N}$) that allows
+negative outputs (elements of $\mathbb{Z}$). Unfortunately this "variation" is
+substantial, and the relationship between $\mathbb{Z}$-polyregular functions
+(of [9]) and $\mathbb{N}$-polyregular functions (unary output) remained unclear
+up to this paper. In particular the decidability of aperiodicity for
+$\mathbb{N}$-polyregular functions could not be deduced directly from [9].
 
@@
 assumption of commutativity, and then to allow the outputs to non-unary
-alphabets. Both steps are beyond the scope of the current paper.
+alphabets. Both of these steps are beyond the scope of the current paper.
 

### 2024-09-05 parts/appendix.tex
putting things in appendix

@@
 
-\begin{proofof}[negative-not-nrat:ex]
+\begin{proofof}[negative-but-npoly:ex]
     The function $w \mapsto |w|$ is a \kl{$\Nat$-polyregular function}.
@@
     are non-negative.
+\end{proofof}
+
+\begin{proofof}[undecidable-non-negative:lem]
+    Recall that by the undecidability of the Hilbert's Tenth Problem \cite[Problem 10 page 22]{HILB1902}, it is
+    undecidable whether a given polynomial $P \in \Rel[\vec{X}]$ has a solution in
+    $\Rel^n$ \cite{MATI1970,DAVIS1973}.
+    The result follows by observing that
+    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^n$ if and only 
+    if $P^2 - 1$ is \kl{non-negative}. 
+\end{proofof}
+
+\begin{proofof}[n-poly-combinatorics:lem]
+	Let us fix a finite monoid $M$, a morphism $\mu \colon \Sigma^* \to M$, and
+	let $\omega$ be an idempotent power for the finite monoid $M$,
+	i.e., a number such that for all $x \in M$,
+	$(x^{\omega})^2 = x^{\omega}$.
+	We will prove by induction on $k \in \Nat_{\geq 1}$
+	that for all
+	$\pi \colon M^k \to \Nat$,
+	the lemma holds with $s = 0$ for the function
+	$f = \pi^\dagger$, i.e., such that
+	$f(w)$ is the sum over all factorizations of $w$
+	into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
+	of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
+	Since every \kl{$\Nat$-polyregular function}
+	is obtained via some choice of $M, \mu, \pi$ (\cref{nat-rel-poly:def})
+	we derive the desired result.
+
+	In all cases, we will use a \kl{polynomial function} $q \colon \Nat^p \to
+		\Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
+		(u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ (for $0 \leq i
+		\leq p$) and $u_i$ (for $1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
+	implicitly multiply all variables by $\omega$ in order to make the
+	equations more readable. Note that when using an induction hypothesis, we
+	will therefore have to check that the polynomial $q$ will have this
+	specific form.
+
+	\textbf{Base case: $k = 1$.}
+	Let us show that $f \circ q$ is
+	actually a constant (positive) function (and in particular a \kl{natural binomial function}), as for
+	all $n_1, \dots, n_p \geq 1$,
+	\begin{align*}
+		f(q(n_1, \dots, n_p))
+		 & \eqdef \pi\left(\mu(q(n_1, \dots, n_p))\right)                                      \\
+		 & \eqdef \pi\left(\mu(\alpha_0 \prod_{i = 1}^p u_i^{\omega n_i} \alpha_i)\right)
+		\\
+		 & = \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i)^{\omega n_i} \mu(\alpha_i)\right)
+		 & \text{ morphism }
+		\\
+		 & = \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
+		 & \text{ idempotent and } n_i \geq 1                                                  \\
+		 & = c \geq 0
+	\end{align*}
+
+	\textbf{Induction hypothesis.}
+	The proof will essentially follow the same schema as the base case,
+	with extra care needed to properly handle the summation of polynomials
+	that arises when $k \geq 2$.
+
+	We will use the fact that given $n_1, \dots, n_p \geq 0$,
+	there exists a bijection between
+	partitions of $q(\omega n_1, \dots, \omega n_p)$
+	into $k+1$ words, and
+	triples composed of
+	a position  $1 \leq \ell \leq p$,
+	a ratio     $0 \leq b \leq n_\ell$,
+	a remainder $0 \leq r < \omega \times \card{u_\ell}$,
+	and a partition of
+	$)u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
+		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
+	into $k$ words,
+	where we have used the notation
+	$(v)_{[-r:]}$ to denote the word obtained from $v$
+	by keeping only the last $r$ letters, in a sort of Pythonic syntax.
+	That is, we look at where the first ``cut" is made.
+
+	Now, given $(\ell, b, r)$ respecting the inequalities
+	mentioned above,
+	one can define the \kl{polynomial functions}
+	$s_{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
+	and
+	$q_{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
+	so that
+	$q(X_1, \dots, X_p) = s_{\ell,r,b}(X_1, \dots, X_{\ell-1})
+		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
+	under the assumption that $X_\ell \geq b$:
+	\begin{equation*}
+		q_{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
+		\defined
+		u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times X_1}
+		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
+		\quad .
+	\end{equation*}
+	Similarly, let us define
+	$s_{\ell, r, b}(X_1, \dots, X_{\ell - 1})$
+	as
+	$\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
+		u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}$,
+	Remark that $q_{\ell,r}$ has all of its indeterminates
+	multiplied by $\omega$, so that the induction hypothesis
+	can be used.
+
+	The key remark is that $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
+	constant function from $\Nat^{\ell-1}$ to $M$, and that its value
+	only depends on whether $b = 0$ or $b > 0$.
+	This holds with
+	the exact same proof as the base case of this induction. Let us call
+	$m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
+	obtained by $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
+	$b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
+		\to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
+		\pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$.
+
+	With all of this preliminary work done, we are now ready to conclude by
+	simply unrolling the definition of
+	a function
+	$f$ defined using a \kl{production function}
+	$\pi \colon M^{k+1} \to \Nat$.
+	In the following equations, we omit the precise
+	range for the sum over $(\ell, r, b)$ to save space.
+	Similarly, we will use
+	$q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
+		\defined q_{\ell,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
+	at the end of the computation.
+	\begin{align*}
+		 & f(q(n_1, \dots, n_p))           \\
+		 & \eqdef
+		\sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
+		\pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
+		 & = \sum_{(\ell, b, r)}
+		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
+		\pi(\mu(s_{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
+		\\
+		 & =
+		\sum_{(\ell, 0, r)}
+		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
+		\pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
+		\\
+		 & +
+		\sum_{(\ell, b > 0, r)}
+		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
+		\pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
+		\\
+		 & =
+		\sum_{(\ell, 0, r)}
+		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
+		\pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
+		\\
+		 & +
+		\sum_{(\ell, b > 0, r)}
+		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell-b, \dots, n_p)}
+		\pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
+		\\
+		 & =
+		\sum_{1 \leq \ell \leq p}
+		\sum_{0 \leq r < \omega \times \card{u_\ell}}
+		(\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell-b, \dots, n_p))
+		\\
+		 & +
+		\sum_{1 \leq \ell \leq p}
+		\sum_{0 \leq r < \omega \times \card{u_\ell}}
+		\sum_{0 < b < n_\ell}
+		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
+	\end{align*}
+	Now, using the induction hypothesis,
+	we have \kl{natural binomial functions} $F_{\ell,r,0}$ and $F_{\ell,r,1}$
+	satisfying
+	\begin{equation*}
+		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
+		= F_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
+		\quad
+		\forall n_\ell - b, \dots, n_p \geq 1
+		\quad .
+	\end{equation*}
+	And similarly
+	\begin{equation*}
+		(\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
+		= F_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
+		\quad
+		\forall n_1, \dots, n_{\ell - 1} \geq 1
+		\quad .
+	\end{equation*}
+
+	It is a well-known equality that for all $s,n \in \Nat$,
+	$\sum_{0 < i < n} \binom{s}{i} = \binom{s+1}{n}$.
+	As a consequence, \kl{natural binomial functions} are
+	stable under the following summation operation:
+	$\Sigma_Y \colon Q(x_1,\dots,x_n,y) \mapsto \sum_{0 < b < y} Q(x_1, \dots, x_n, b)$.
+	This allows us to conclude by defining:
+	\begin{equation*}
+		F(x_1, \dots, x_p) \defined
+		\sum_{1 \leq \ell \leq p} \sum_{0 \leq r < \omega \card{u_\ell}}
+		F_{l,r,0}(x_{\ell}, \dots, \omega x_p) + \Sigma_Y F_{l,r,1}(Y,
+		x_{\ell+1}, \dots, x_p)
+		\quad .
+	\end{equation*}
+	The latter being a \kl{natural binomial function} satisfying
+	the desired equality.
+\end{proofof}
+
+\begin{proofof}[n-sf-combinatorics:lem]
+	We redo the proof of \cref{n-poly-combinatorics:lem}, with the extra
+	assumption that the monoid $M$ is \kl(monoid){aperiodic}, hence that there
+	exists $s \in \Nat$ such that $x^{s+1} = x^s$ for all $x \in M$. As a
+	consequence, we can define $\omega = 1$, and
+	translate inputs by $s$, instead of multiplying them by $\omega$.
 \end{proofof}
@@
     belong to $\Nat[\vec{X}]$, which is decidable.
+\end{proofof}
+
+\begin{proofof}[integer-binomial-polynomial:lem]
+	Let us first notice that $\pbinom{X}{k}$ is \kl{integer-valued}
+	when $k \in \Nat$. One way to notice this is that when $X \geq 0$,
+	$\pbinom{X}{k} = \binom{X}{k}$ is the number of ways to choose $k$ elements among $X$,
+	which is an integer. When
+	$X < 0$, then $\pbinom{X}{k} = (-1)^k \pbinom{k - X + 1}{k} = (-1)^k \binom{k - X + 1}{k}$
+	is also an integer. Because \kl{integer-valued} polynomials are closed under products,
+	$\Rel$-linear combination, and 
+    shifting of domain by an integer,
+    we conclude that all \kl{integer binomial polynomials} are
+	\kl{integer-valued}.
+
+	For the converse direction, we follow the original proof of Pólya \cite{POLYA1915,CACHA1996}.
+	The only difference is that we leverage differential operators on multiple indeterminates.
+	To that end, let us consider the operator $\Delta_i$ that maps
+	a polynomial $P$ to the polynomial $\Delta_i(P) \defined P(X_1, \dots, X_i + 1, \dots, X_k) - P$.
+	That is, the discrete partial derivative with respect to the $i$-th indeterminate.
+	Let us notice that $\Delta_i \pbinom{X_i}{\alpha} = \pbinom{X_i}{\alpha - 1}$ when $\alpha > 0$,
+	indeed:
+	\begin{align*}
+		\Delta_i \pbinom{X_i}{\alpha} & = \pbinom{X_i + 1}{\alpha} - \pbinom{X_i}{\alpha}                                                         \\
+		                              & = \frac{(X_i + 1) \cdots (X_i + 1 - \alpha + 1)}{\alpha!} - \frac{X_i \cdots (X_i - \alpha + 1)}{\alpha!} \\
+		                              & = \frac{X_i \cdots (X_i - \alpha + 2)}{\alpha!} \alpha                                                    \\
+		                              & = \pbinom{X_i}{\alpha - 1}
+		\quad .
+	\end{align*}
+	Furthermore, when $\alpha = 0$, $\Delta_i \pbinom{X_i}{0} = 0$.
+
+	Let us now remark that a simple expansion proves $\Delta_i \Delta_j P = \Delta_j \Delta_i P$ for all
+	$1 \leq i,j \leq k$.
+	This justifies the soundness of the notation
+	$\Delta_\alpha$ where $\alpha \colon \set{1, \dots, k} \to \Nat$ is a
+	multi-index, where $\Delta_{\alpha} \defined \Delta_1^{\alpha_1} \cdots
+		\Delta_k^{\alpha_k}$.
+	Let us write $\card{\alpha} \defined \sum_{i = 1}^k \alpha_i$ for the
+	size of a multi-index.
+	We claim that for all $P \in \Rat[X_1, \dots, X_k]$ of degree at most $d$,
+	the following \emph{discrete Taylor expansion} holds:
+	\begin{equation}
+		\label{discrete-taylor:eq}
+		P =
+		\sum_{\card{\alpha} \leq d}
+		\Delta_{\alpha} P (0, \dots, 0)
+		\times
+		\prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
+		\quad .
+	\end{equation}
+
+	To obtain \cref{discrete-taylor:eq}, we proceed by induction on the degree $d$.
+	When $d = 0$, the result is holds because $P = P(0, \dots, 0)$.
+	Assume that the result holds for all polynomials of degree at most $d$.
+	Let $P$ of degree at most $d+1$.
+	Then, there exists coefficients $\seqof{c_{\alpha}}{\card{\alpha} \leq d+1}$
+	in $\Rat$
+	such that
+	\begin{equation*}
+		P = \sum_{\card{\alpha} \leq d+1} c_{\alpha} \prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
+		\quad .
+	\end{equation*}
+	This is because the monomials of the form $\prod_{i = 1}^k \pbinom{X_i}{\alpha_i}$
+	form a basis of $\Rat[X_1, \dots, X_k]$.
+	In particular, when applying one partial derivative, one can leverage the induction hypothesis to obtain:
+	\begin{align*}
+		\Delta_1 P & = \sum_{\card{\alpha} \leq d+1} c_{\alpha} \Delta_1 \prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
+		=
+		\sum_{\card{\alpha} \leq d+1 \wedge \alpha_1 > 0} c_{\alpha}  \pbinom{X_1}{\alpha_1 - 1} \prod_{i = 2}^k \pbinom{X_i}{\alpha_i}
+		\\
+		           & = \sum_{\card{\beta} \leq d} \Delta_{\beta}(\Delta_1 P)(0,\dots,0) \times \pbinom{X_1}{\beta_1} \prod_{i = 2}^k \pbinom{X_i}{\beta_i}
+		\quad .
+	\end{align*}
+
+	Using again the fact that the monomials form a basis, we conclude by
+	identification that for all $\card{\alpha} \leq d+1$ such that $\alpha_1 >
+		0$, we have $c_{\alpha} = \Delta_{\alpha}P (0,\dots,0)$. We can repeat the
+	same process with $\Delta_i$ for $1 \leq i \leq k$ to obtain that
+	$c_{\alpha} = \Delta_{\alpha} P (0,\dots, 0)$ for all $\alpha$ such that $1
+		\leq \card{\alpha} \leq d+1$. Now, for the specific case of $\card{\alpha}
+		= 0$, which is the constant term of $P$, we simply conclude that
+	$P(0,\dots,0) = c_{\alpha}$, and notice that $\Delta_{\alpha} P = P$ in
+	this case.
+
+	Let us now conclude by noticing that if $P$ is \kl{integer-valued},
+	then so is $\Delta_i P$ for all $1 \leq i \leq k$. As a consequence,
+	so are $\Delta_{\alpha} P$, for all multi-indexes $\alpha$, and we conclude
+	that $\Delta_{\alpha}P (0,\dots,0) \in \Rel$ for all $\alpha$.
+	Combined with \cref{discrete-taylor:eq}, we conclude that $P$ is an \kl{integer binomial polynomial}.
+    Remark that $\Delta_{\alpha} P (0, \dots, 0)$ only ever involves evaluation of $P$ on 
+    $\Nat^k$, justifying the weakening of the hypothesis.
+\end{proofof}
+
+\begin{proofof}[natural-binomial-polynomial-positive-bad:ex]
+    $Q$ belongs to $\Rel[\vec{X}]$, but not to $\CorrectPoly$ because
+    $Q(0,Y,Z)$ has a negative maximal monomial,
+    hence it cannot be \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
+    thanks to \cref{corrected-version:thm}.
 \end{proofof}

### 2024-09-05 parts/conclusion.tex
wohooh
putting things in appendix

@@
 
-Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
-functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
-immediately, and $\NSF$ inside $\NPoly$ would also immediately follow. On the
-one hand, one can guess a candidate function and check for equivalence, on the
-other hand, one can guess a \kl{commutative} \kl{star-free polyregular
-function} and check membership (which is decidable thanks to this paper).
+Remark that if \cref{pre-compose-sf-commut:lemma} were to hold for
+\kl{$\Nat$-polyregular functions}, then the decidability of $\NPoly$ inside
+$\ZPoly$ would follow immediately, and $\NSF$ inside $\NPoly$ would also
+immediately follow. On the one hand, one can guess a candidate function and
+check for equivalence, on the other hand, one can guess a \kl{commutative}
+\kl{star-free polyregular function} and check membership (which is decidable
+thanks to this paper). This is restated in our concluding conjecture.
+
+\begin{conjecture}
+    \label{npoly-zpoly:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \ZPoly[k]$
+    belongs to $\NPoly[k]$ if and only if
+    it is \kl{$(k,\Nat)$-combinatorial}.
+    In particular,
+    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
+\end{conjecture}

### 2024-09-05 parts/correction.tex
putting things in appendix

@@
 fact that is not possible to decide whether a polynomial $P \in \Rel[\vec{X}]$
-is \kl{non-negative} (\cref{undecidable-non-negative:lem}), by
-reduction to the undecidability of Hilbert's Tenth Problem
-\cite{HILB1902,MATI1970}.
+is \kl{non-negative} (\cref{undecidable-non-negative:lem}), by reduction to the
+undecidability of Hilbert's Tenth Problem \cite{HILB1902,MATI1970}. That is,
+$\CorrectPoly$ is a decidable class that interpolates between the undecidable
+class of non-negative polynomials and strictly generalize polynomials in
+$\Nat[\vec{X}]$.
 
-\begin{lemma}
+\begin{remark}
     \label{undecidable-non-negative:lem}
-    Let $P \in \Rel[\vec{X}]$ be a polynomial. Then, checking whether $P$ is
-    \kl{non-negative} is undecidable.
-\end{lemma}
-\begin{proof}
-    Recall that by the undecidability of the Hilbert's Tenth Problem \cite[Problem 10 page 22]{HILB1902}, it is
-    undecidable whether a given polynomial $P \in \Rel[\vec{X}]$ has a solution in
-    $\Rel^n$ \cite{MATI1970,DAVIS1973}.
-    The result follows by observing that
-    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^n$ if and only 
-    if $P^2 - 1$ is \kl{non-negative}. 
-\end{proof}
+    \proofref{undecidable-non-negative:lem}
+    Checking whether a polynomial
+    $P \in \Rel[\vec{X}]$ is \kl{non-negative} is undecidable.
+\end{remark}
 
 The proof of \cref{corrected-version:thm} is divided into two parts. First, we
-provide in \cref{sec:n-poly-to-poly}
-a fine
-combinatorial understanding of what functions can be computed by functions in
-$\NPoly$ and $\ZPoly$. This will allow us to prove that \kl{$\Nat$-polyregular
-functions} are in $\CorrectPoly$ (\cref{n-rat-correct:lem}).
-Then, in \cref{sec:poly-to-n-poly}
-we will
-show how to compute polynomials in $\CorrectPoly$ using $\NPoly$
-(\cref{lem:correct-to-n-rat}).
-With this in hand, we will next generalize these results
-to polynomials in $\Rat[\vec{X}]$ in \cref{sec:rel-to-rat}.
+provide in \cref{sec:n-poly-to-poly} a fine combinatorial understanding of what
+functions can be computed in $\NPoly$ and $\ZPoly$. This allows us to prove
+that \kl{$\Nat$-polyregular functions} are in $\CorrectPoly$
+(\cref{n-rat-correct:lem}). Then, in \cref{sec:poly-to-n-poly} we will show how
+to compute polynomials in $\CorrectPoly$ using $\NPoly$
+(\cref{lem:correct-to-n-rat}). With this in hand, we will next generalize these
+results to polynomials in $\Rat[\vec{X}]$ in \cref{sec:rel-to-rat}.
 
@@
 functions} essentially compute binomial coefficients, as illustrated by the
-polynomial $X(X-1)/2 = \binom{X}{2}$.
+polynomial $X(X-1)/2 = \binom{X}{2}$
 of \cref{q-polynomial-n-poly:ex}. A \intro{binomial function} is a
@@
 
-\begin{lemma}
-	\label{n-poly-combinatorics:lem}
+\begin{lemma}[restate=n-poly-combinatorics:lem,label=n-poly-combinatorics:lem]
+    \proofref{n-poly-combinatorics:lem}
 	Let $f$ be an \kl{$\Nat$-polyregular} function.
@@
 \end{lemma}
-\begin{proof}
-	Let us fix a finite monoid $M$, a morphism $\mu \colon \Sigma^* \to M$, and
-	let $\omega$ be an idempotent power for the finite monoid $M$,
-	i.e., a number such that for all $x \in M$,
-	$(x^{\omega})^2 = x^{\omega}$.
-	We will prove by induction on $k \in \Nat_{\geq 1}$
-	that for all
-	$\pi \colon M^k \to \Nat$,
-	the lemma holds with $s = 0$ for the function
-	$f = \pi^\dagger$, i.e., such that
-	$f(w)$ is the sum over all factorizations of $w$
-	into $k$ words $(w_1, \dots, w_k) \in \Sigma^*$
-	of the value $\pi(\mu(w_1), \dots, \mu(w_k))$.
-	Since every \kl{$\Nat$-polyregular function}
-	is obtained via some choice of $M, \mu, \pi$ (\cref{nat-rel-poly:def})
-	we derive the desired result.
-
-	In all cases, we will use a \kl{polynomial function} $q \colon \Nat^p \to
-		\Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
-		(u_i^{\omega \times X_i} \alpha_i)$ using some words $\alpha_i$ (for $0 \leq i
-		\leq p$) and $u_i$ (for $1 \leq i \leq p)$ in $\Sigma^*$. That is, we will
-	implicitly multiply all variables by $\omega$ in order to make the
-	equations more readable. Note that when using an induction hypothesis, we
-	will therefore have to check that the polynomial $q$ will have this
-	specific form.
-
-	\textbf{Base case: $k = 1$.}
-	Let us show that $f \circ q$ is
-	actually a constant (positive) function (and in particular a \kl{natural binomial function}), as for
-	all $n_1, \dots, n_p \geq 1$,
-	\begin{align*}
-		f(q(n_1, \dots, n_p))
-		 & \eqdef \pi\left(\mu(q(n_1, \dots, n_p))\right)                                      \\
-		 & \eqdef \pi\left(\mu(\alpha_0 \prod_{i = 1}^p u_i^{\omega n_i} \alpha_i)\right)
-		\\
-		 & = \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i)^{\omega n_i} \mu(\alpha_i)\right)
-		 & \text{ morphism }
-		\\
-		 & = \pi\left(\mu(\alpha_0) \prod_{i = 1}^p \mu(u_i) \mu(\alpha_i)\right)
-		 & \text{ idempotent and } n_i \geq 1                                                  \\
-		 & = c \geq 0
-	\end{align*}
-
-	\textbf{Induction hypothesis.}
-	The proof will essentially follow the same schema as the base case,
-	with extra care needed to properly handle the summation of polynomials
-	that arises when $k \geq 2$.
-
-	We will use the fact that given $n_1, \dots, n_p \geq 0$,
-	there exists a bijection between
-	partitions of $q(\omega n_1, \dots, \omega n_p)$
-	into $k+1$ words, and
-	triples composed of
-	a position  $1 \leq \ell \leq p$,
-	a ratio     $0 \leq b \leq n_\ell$,
-	a remainder $0 \leq r < \omega \times \card{u_\ell}$,
-	and a partition of
-	$)u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
-		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
-	into $k$ words,
-	where we have used the notation
-	$(v)_{[-r:]}$ to denote the word obtained from $v$
-	by keeping only the last $r$ letters, in a sort of Pythonic syntax.
-	That is, we look at where the first ``cut" is made.
-
-	Now, given $(\ell, b, r)$ respecting the inequalities
-	mentioned above,
-	one can define the \kl{polynomial functions}
-	$s_{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
-	and
-	$q_{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
-	so that
-	$q(X_1, \dots, X_p) = s_{\ell,r,b}(X_1, \dots, X_{\ell-1})
-		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
-	under the assumption that $X_\ell \geq b$:
-	\begin{equation*}
-		q_{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
-		\defined
-		u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times X_1}
-		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
-		\quad .
-	\end{equation*}
-	Similarly, let us define
-	$s_{\ell, r, b}(X_1, \dots, X_{\ell - 1})$
-	as
-	$\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
-		u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}$,
-	Remark that $q_{\ell,r}$ has all of its indeterminates
-	multiplied by $\omega$, so that the induction hypothesis
-	can be used.
-
-	The key remark is that $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
-	constant function from $\Nat^{\ell-1}$ to $M$, and that its value
-	only depends on whether $b = 0$ or $b > 0$.
-	This holds with
-	the exact same proof as the base case of this induction. Let us call
-	$m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
-	obtained by $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
-	$b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
-		\to \Nat$ via $\pi_{\ell,r,0}(m_2, \dots, m_{k+1}) \defined
-		\pi(m_{\ell,r,0}, m_1, \dots, m_{k+1})$.
-
-	With all of this preliminary work done, we are now ready to conclude by
-	simply unrolling the definition of
-	a function
-	$f$ defined using a \kl{production function}
-	$\pi \colon M^{k+1} \to \Nat$.
-	In the following equations, we omit the precise
-	range for the sum over $(\ell, r, b)$ to save space.
-	Similarly, we will use
-	$q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
-		\defined q_{\ell,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
-	at the end of the computation.
-	\begin{align*}
-		 & f(q(n_1, \dots, n_p))           \\
-		 & \eqdef
-		\sum_{v_1 \dots v_{k+1} = q(n_1, \dots, n_p)}
-		\pi(\mu(v_1), \dots, \mu(v_{k+1})) \\
-		 & = \sum_{(\ell, b, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
-		\pi(\mu(s_{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
-		\\
-		 & =
-		\sum_{(\ell, 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
-		\pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
-		\\
-		 & +
-		\sum_{(\ell, b > 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
-		\pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
-		\\
-		 & =
-		\sum_{(\ell, 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
-		\pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
-		\\
-		 & +
-		\sum_{(\ell, b > 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell-b, \dots, n_p)}
-		\pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
-		\\
-		 & =
-		\sum_{1 \leq \ell \leq p}
-		\sum_{0 \leq r < \omega \times \card{u_\ell}}
-		(\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell-b, \dots, n_p))
-		\\
-		 & +
-		\sum_{1 \leq \ell \leq p}
-		\sum_{0 \leq r < \omega \times \card{u_\ell}}
-		\sum_{0 < b < n_\ell}
-		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
-	\end{align*}
-	Now, using the induction hypothesis,
-	we have \kl{natural binomial functions} $F_{\ell,r,0}$ and $F_{\ell,r,1}$
-	satisfying
-	\begin{equation*}
-		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
-		= F_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
-		\quad
-		\forall n_\ell - b, \dots, n_p \geq 1
-		\quad .
-	\end{equation*}
-	And similarly
-	\begin{equation*}
-		(\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
-		= F_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
-		\quad
-		\forall n_1, \dots, n_{\ell - 1} \geq 1
-		\quad .
-	\end{equation*}
-
-	It is a well-known equality that for all $s,n \in \Nat$,
-	$\sum_{0 < i < n} \binom{s}{i} = \binom{s+1}{n}$.
-	As a consequence, \kl{natural binomial functions} are
-	stable under the following summation operation:
-	$\Sigma_Y \colon Q(x_1,\dots,x_n,y) \mapsto \sum_{0 < b < y} Q(x_1, \dots, x_n, b)$.
-	This allows us to conclude by defining:
-	\begin{equation*}
-		F(x_1, \dots, x_p) \defined
-		\sum_{1 \leq \ell \leq p} \sum_{0 \leq r < \omega \card{u_\ell}}
-		F_{l,r,0}(x_{\ell}, \dots, \omega x_p) + \Sigma_Y F_{l,r,1}(Y,
-		x_{\ell+1}, \dots, x_p)
-		\quad .
-	\end{equation*}
-	The latter being a \kl{natural binomial function} satisfying
-	the desired equality.
-\end{proof}
 
@@
 
-\begin{lemma}
-	\label{n-sf-combinatorics:lem}
+\begin{lemma}[restate=n-sf-combinatorics:lem,label=n-sf-combinatorics:lem]
+    \proofref{n-sf-combinatorics:lem}
 	Let $f$ be a \kl{star-free $\Nat$-polyregular} function.
@@
 \end{lemma}
-\begin{proof}[Proof Sketch]
-	We redo the proof of \cref{n-poly-combinatorics:lem}, with the extra
-	assumption that the monoid $M$ is \kl(monoid){aperiodic}, hence that there
-	exists $s \in \Nat$ such that $x^{s+1} = x^s$ for all $x \in M$. As a
-	consequence, we can define $\omega = 1$, and
-	translate inputs by $s$, instead of multiplying them by $\omega$.
-\end{proof}
 
@@
 
-\begin{lemma}
-	\label{integer-binomial-polynomial:lem}
+\begin{lemma}[restate=integer-binomial-polynomial:lem,label=integer-binomial-polynomial:lem]
+	\proofref{integer-binomial-polynomial:lem}
 	Let $P \in \Rat[X_1, \dots, X_k]$ be a polynomial.
@@
 \end{lemma}
-\begin{proof}
-	Let us first notice that $\pbinom{X}{k}$ is \kl{integer-valued}
-	when $k \in \Nat$. One way to notice this is that when $X \geq 0$,
-	$\pbinom{X}{k} = \binom{X}{k}$ is the number of ways to choose $k$ elements among $X$,
-	which is an integer. When
-	$X < 0$, then $\pbinom{X}{k} = (-1)^k \pbinom{k - X + 1}{k} = (-1)^k \binom{k - X + 1}{k}$
-	is also an integer. Because \kl{integer-valued} polynomials are closed under products,
-	$\Rel$-linear combination, and 
-    shifting of domain by an integer,
-    we conclude that all \kl{integer binomial polynomials} are
-	\kl{integer-valued}.
-
-	For the converse direction, we follow the original proof of Pólya \cite{POLYA1915,CACHA1996}.
-	The only difference is that we leverage differential operators on multiple indeterminates.
-	To that end, let us consider the operator $\Delta_i$ that maps
-	a polynomial $P$ to the polynomial $\Delta_i(P) \defined P(X_1, \dots, X_i + 1, \dots, X_k) - P$.
-	That is, the discrete partial derivative with respect to the $i$-th indeterminate.
-	Let us notice that $\Delta_i \pbinom{X_i}{\alpha} = \pbinom{X_i}{\alpha - 1}$ when $\alpha > 0$,
-	indeed:
-	\begin{align*}
-		\Delta_i \pbinom{X_i}{\alpha} & = \pbinom{X_i + 1}{\alpha} - \pbinom{X_i}{\alpha}                                                         \\
-		                              & = \frac{(X_i + 1) \cdots (X_i + 1 - \alpha + 1)}{\alpha!} - \frac{X_i \cdots (X_i - \alpha + 1)}{\alpha!} \\
-		                              & = \frac{X_i \cdots (X_i - \alpha + 2)}{\alpha!} \alpha                                                    \\
-		                              & = \pbinom{X_i}{\alpha - 1}
-		\quad .
-	\end{align*}
-	Furthermore, when $\alpha = 0$, $\Delta_i \pbinom{X_i}{0} = 0$.
-
-	Let us now remark that a simple expansion proves $\Delta_i \Delta_j P = \Delta_j \Delta_i P$ for all
-	$1 \leq i,j \leq k$.
-	This justifies the soundness of the notation
-	$\Delta_\alpha$ where $\alpha \colon \set{1, \dots, k} \to \Nat$ is a
-	multi-index, where $\Delta_{\alpha} \defined \Delta_1^{\alpha_1} \cdots
-		\Delta_k^{\alpha_k}$.
-	Let us write $\card{\alpha} \defined \sum_{i = 1}^k \alpha_i$ for the
-	size of a multi-index.
-	We claim that for all $P \in \Rat[X_1, \dots, X_k]$ of degree at most $d$,
-	the following \emph{discrete Taylor expansion} holds:
-	\begin{equation}
-		\label{discrete-taylor:eq}
-		P =
-		\sum_{\card{\alpha} \leq d}
-		\Delta_{\alpha} P (0, \dots, 0)
-		\times
-		\prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
-		\quad .
-	\end{equation}
-
-	To obtain \cref{discrete-taylor:eq}, we proceed by induction on the degree $d$.
-	When $d = 0$, the result is holds because $P = P(0, \dots, 0)$.
-	Assume that the result holds for all polynomials of degree at most $d$.
-	Let $P$ of degree at most $d+1$.
-	Then, there exists coefficients $\seqof{c_{\alpha}}{\card{\alpha} \leq d+1}$
-	in $\Rat$
-	such that
-	\begin{equation*}
-		P = \sum_{\card{\alpha} \leq d+1} c_{\alpha} \prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
-		\quad .
-	\end{equation*}
-	This is because the monomials of the form $\prod_{i = 1}^k \pbinom{X_i}{\alpha_i}$
-	form a basis of $\Rat[X_1, \dots, X_k]$.
-	In particular, when applying one partial derivative, one can leverage the induction hypothesis to obtain:
-	\begin{align*}
-		\Delta_1 P & = \sum_{\card{\alpha} \leq d+1} c_{\alpha} \Delta_1 \prod_{i = 1}^k \pbinom{X_i}{\alpha_i}
-		=
-		\sum_{\card{\alpha} \leq d+1 \wedge \alpha_1 > 0} c_{\alpha}  \pbinom{X_1}{\alpha_1 - 1} \prod_{i = 2}^k \pbinom{X_i}{\alpha_i}
-		\\
-		           & = \sum_{\card{\beta} \leq d} \Delta_{\beta}(\Delta_1 P)(0,\dots,0) \times \pbinom{X_1}{\beta_1} \prod_{i = 2}^k \pbinom{X_i}{\beta_i}
-		\quad .
-	\end{align*}
-
-	Using again the fact that the monomials form a basis, we conclude by
-	identification that for all $\card{\alpha} \leq d+1$ such that $\alpha_1 >
-		0$, we have $c_{\alpha} = \Delta_{\alpha}P (0,\dots,0)$. We can repeat the
-	same process with $\Delta_i$ for $1 \leq i \leq k$ to obtain that
-	$c_{\alpha} = \Delta_{\alpha} P (0,\dots, 0)$ for all $\alpha$ such that $1
-		\leq \card{\alpha} \leq d+1$. Now, for the specific case of $\card{\alpha}
-		= 0$, which is the constant term of $P$, we simply conclude that
-	$P(0,\dots,0) = c_{\alpha}$, and notice that $\Delta_{\alpha} P = P$ in
-	this case.
-
-	Let us now conclude by noticing that if $P$ is \kl{integer-valued},
-	then so is $\Delta_i P$ for all $1 \leq i \leq k$. As a consequence,
-	so are $\Delta_{\alpha} P$, for all multi-indexes $\alpha$, and we conclude
-	that $\Delta_{\alpha}P (0,\dots,0) \in \Rel$ for all $\alpha$.
-	Combined with \cref{discrete-taylor:eq}, we conclude that $P$ is an \kl{integer binomial polynomial}.
-    Remark that $\Delta_{\alpha} P (0, \dots, 0)$ only ever involves evaluation of $P$ on 
-    $\Nat^k$, justifying the weakening of the hypothesis.
-\end{proof}
 
@@
 
-\AP Obtaining an analogue of \cref{integer-binomial-polynomials:cor}
-for \kl{$\Nat$-polyregular functions} requires a bit more work,
-as polynomials in $\Rat[\vec{X}]$ that are \kl{represented} by $\NPoly$ are not
-exactly \kl{natural binomial polynomials}
-(\cref{natural-binomial-polynomial-negative:ex,natural-binomial-polynomial-basis:ex,natural-binomial-polynomial-positive-bad:ex}).
-To address the issues raised by these examples, we introduce the notion of
-\intro{strongly natural binomial polynomials}, as the polynomials $P \in
-\Rat[X]$ such that for all partial valuation $\nu \colon \topartial \Nat$,
-$\restr{P}{\nu}$ is a \kl{natural binomial polynomial}, which 
-characterizes the class of polynomials that are \kl{represented} by $\NPoly$
-(\cref{decide-rat-poly-npoly:cor}).
+\AP Obtaining an analogue of \cref{integer-binomial-polynomials:cor} for
+\kl{$\Nat$-polyregular functions} requires a bit more work, as polynomials in
+$\Rat[\vec{X}]$ that are \kl{represented} by $\NPoly$ are not exactly
+\kl{natural binomial polynomials},
+(\cref{natural-binomial-polynomial-positive-bad:ex}). To address the issues
+raised by the former example, we introduce the notion of \intro{strongly
+natural binomial polynomials}, as the polynomials $P \in \Rat[X]$ such that for
+all partial valuation $\nu \colon \topartial \Nat$, $\restr{P}{\nu}$ is a
+\kl{natural binomial polynomial}, which characterizes the class of polynomials
+that are \kl{represented} by $\NPoly$ (\cref{decide-rat-poly-npoly:cor}).
 
-\begin{example}
-	\label{natural-binomial-polynomial-negative:ex}
-	The polynomial $\pbinom{X - 2}{2}$ takes negative values for $X = 0$.
-\end{example}
-
-\begin{example}
-	\label{natural-binomial-polynomial-basis:ex}
-	The polynomial $\pbinom{X - 2}{2} + 1$ is a \kl{non-negative} \kl{natural binomial polynomial}
-	but in the standard basis, it is written
-	$\pbinom{X}{2} - \pbinom{X}{1} + 2$, which has a negative coefficient.
-\end{example}
-
-\begin{example}
-	\label{natural-binomial-polynomial-positive-bad:ex}
+\begin{example}[restate=natural-binomial-polynomial-positive-bad:ex,label=natural-binomial-polynomial-positive-bad:ex]
+	\proofref{natural-binomial-polynomial-positive-bad:ex}
 	The polynomial
@@
 \begin{proof}
-
     For the implication \cref{rat-npoly-1:item} $\implies$

### 2024-09-05 parts/counter.tex
putting things in appendix

@@
     \label{negative-but-npoly:ex}
-    \proofref{negative-not-nrat:ex}
+    \proofref{negative-but-npoly:ex}
     The polynomials $X$, $X^2 + 3$,

### 2024-09-05 parts/intro.tex
putting things in appendix

@@
 class of functions called (non-commutative) \kl{rational series}
-\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
-\kl{$\Nat$-polyregular functions} correspond to a natural (strict) subclass of
-\kl{$\Nat$-rational series} ($\NRat$), another model of computation that aims
-at generalizing the notion of regular language to functions with outputs in a
-\emph{commutative} semiring \cite{BERE88,BERE10}.\footnote{In particular, the
-    notion of \kl{rational series} with output in $\Gamma^*$ is not well-defined.} It was frequently rediscovered in the literature that
-\kl{$\Nat$-polyregular functions} (although not often called this way) are a
-subclass of \kl{rational series} \cite{SCHU62,KRRC13,CDTL23}.
+\cite{REUT80,BERE88,BERE10}. This other model of computation generalizes the
+notion of regular language to functions with outputs in a \emph{commutative}
+semiring, and it was frequently rediscovered that \kl{$\Nat$-polyregular
+functions} form a (strict) subclass of \kl{rational series} with outputs in
+$\Nat$, called \kl{$\Nat$-rational series} ($\NRat$)
+\cite{SCHU62,KRRC13,CDTL23}.
 
@@
 
-\AP This approach through rational series has several limitations. First,
+\AP This approach through rational series bears numerous limitations. First,
 $\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
@@
 \cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
-is also an open problem (see e.g. \cite{KARH77}), that remains open even in the
-restricted case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It
-was conjectured that, being ``computable using coefficients in $\Nat$'' and being
+is an open problem (see e.g. \cite{KARH77}), that remains open even in the
+simple case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It was
+conjectured that, being ``computable using coefficients in $\Nat$'' and being
 ``star-free'' were two orthogonal properties, and therefore that the following
@@
 study using classical mathematical tools. The simplest kind of functions in
-this class are actual polynomial functions, which were studied by Karhumäki in
-\cite{KARH77}. Our first contribution is to provide a counter-example to the
-main theorem of \cite{KARH77} that claimed to characterize the (multivariate)
-polynomials in $\Rat[\vec{X}]$ that can be computed by some \kl{$\Nat$-rational
-series} (\cref{thm:counter-example}). Our second contribution is to correct
-this error, and provide (effective) characterizations of polynomials in
-$\Rat[\vec{X}]$ that can be computed by \kl{$\Rel$-polyregular functions}
+this class are actual (multivariate) polynomial functions, which were studied
+by Karhumäki in \cite{KARH77}. Our analysis of \kl{commutative}
+\kl{$\Nat$-polyregular functions} is divided in two main parts: the first one
+completely revisits the results of \cite{KARH77} on polynomials, and the second
+one extends the former decidability results to arbitrary \kl{commutative}
+\kl{$\Nat$-polyregular functions}.
+
+Our first contribution is to provide a counter-example to the main theorem of
+\cite{KARH77} that claimed to characterize the (multivariate) polynomials in
+$\Rat[\vec{X}]$ that can be computed by some \kl{$\Nat$-rational series}
+(\cref{thm:counter-example}). Our second contribution is to correct this error,
+and provide (effective) characterizations of polynomials in $\Rat[\vec{X}]$
+that can be computed by \kl{$\Rel$-polyregular functions}
 (\cref{integer-binomial-polynomials:cor}), and by \kl{$\Nat$-polyregular
-functions} (\cref{decide-rat-poly-npoly:cor}). Our third contribution
+functions} (\cref{decide-rat-poly-npoly:cor}). These characterizations also
+show that polynomials computable using $\NRat$ (resp. $\ZRat$) are also
+computable using $\NSF$ (resp. $\ZSF$), that is: polynomials are inherently
+\emph{star-free}. Furthermore, these characterizations provide effective
+descriptions of polynomials that can be expressed in $\ZPoly$ as those obtained
+using integer combinations of products of \emph{binomial coefficients} (called
+\kl{integer binomial polynomials}, defined page \kpageref{integer binomial
+polynomial}) and similarly for $\NPoly$ using the notion of \kl{strongly
+natural binomial polynomials} (defined page \kpageref{strongly natural binomial
+polynomial}).
+
+Our third and last contribution
 of this paper is to answer \cite[Conjecture 7.61]{DOUE23} (stating that $\NSF =

### 2024-09-11 parts/starfree.tex
add fun result

@@
     thanks to \cref{decide-rat-poly-npoly:cor}.
+\end{proof}
+
+It was already known that \kl{$\Rel$-polyregular functions} with unary input
+that are non-negative are \kl{$\Nat$-polyregular} \cite[Proposition 2.1 p
+137]{BERE10}. Let us derive this fact from our \cref{decidable-n-poly:thm}.
+
+\begin{corollary}
+    Let $f \colon \set{a}^* \to \Rel$ be a non-negative \kl{$\Rel$-polyregular function},
+    then $f \in \NPoly$.
+\end{corollary}
+\begin{proof}
+    Since $f$ has unary input, it is \kl{commutative}. Furthermore,
+    $f$ is \kl{$(1,\Nat)$-combinatorial} because for all $q \colon \Nat \to
+    \set{a}$ and all $\omega \geq 1$, $f(q(\omega X))$ is non-negative.
+    When it is a polynomial function, it therefore belongs to $\CorrectPoly$,
+    hence is a \kl{strongly natural binomial polynomial}.
+    We conclude using \cref{decidable-n-poly:thm}.
 \end{proof}

### 2024-09-12 globals/metadata.tex
enormous rewrite

@@
 polynomial growth (the ouptut of the function is bounded by a polynomial in the
-size of the input). This continues the work of Karhumäki, who studied the same
-question under the assumption that the computed function was a polynomial,
-which is more restrictive. Answering this question starts by exhibiting a
-counterexample to the theorem of Karhumäki, proposing a new and valid
-characterization in the case of polynomials, that lifts to the more general
-setting of polynomial growth and commutativity. Surprisingly, we manage to
+size of the input). We
 leverage this effective characterization to decide whether a function computed

### 2024-09-12 parts/appendix.tex
enormous rewrite

@@
 \section{Proofs of section \ref{polynomials:sec}}
+
+\begin{proofof}[n-poly-n-poly:example]
+    The polynomials of $\Nat[\vec{X}]$
+    are obtained from basic functions (constant in $\Nat$,
+    $X_i$ for some $1 \leq i \leq n$)
+    by products and sums (\cref{stability-polyregular:lemma}). Because the basic functions are
+    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions (see
+    \cref{size-of-word-nsf:ex}),
+    that are closed under these operations, we conclude.
+\end{proofof}
 
@@
 	and a partition of
-	$)u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
+	$(u_\ell^\omega)_{[-r:]} u_\ell^{\omega \times (n_\ell - b)}
 		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{n_p} \alpha_{p}$
@@
 	one can define the \kl{polynomial functions}
-	$s_{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
+	$\CutPref{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
 	and
-	$q_{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
+	$\CutSuff{\ell,r} \colon \Nat^{p - \ell + 1} \to \Sigma^*$
 	so that
-	$q(X_1, \dots, X_p) = s_{\ell,r,b}(X_1, \dots, X_{\ell-1})
+	$q(X_1, \dots, X_p) = \CutPref{\ell,r,b}(X_1, \dots, X_{\ell-1})
 		q_{\ell, r}(X_{\ell} - b, \dots, X_p)$
 	under the assumption that $X_\ell \geq b$:
-	\begin{equation*}
-		q_{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
-		\defined
-		u_\ell^\omega\mid_{[-r:]} u_\ell^{\omega \times X_1}
+	\begin{align*}
+        \CutPref{\ell,r,b}(X_1, \dots, X_{\ell - 1})
+        &\defined
+        \alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
+        u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}
+        \quad .
+        \\
+		\CutSuff{\ell,r}(X_{1}, \dots, X_{p - \ell + 1})
+        &\defined
+        (u_\ell^\omega)_{[-r:]} u_\ell^{\omega X_1}
 		\alpha_{\ell+1} \cdots \alpha_{p-1} u_p^{X_{p - \ell + 1}} \alpha_{p}
 		\quad .
-	\end{equation*}
-	Similarly, let us define
-	$s_{\ell, r, b}(X_1, \dots, X_{\ell - 1})$
-	as
-	$\alpha_0 u_1^{\omega X_1} \alpha_2 \dots \alpha_{\ell - 1}
-		u_\ell^{\omega b} (u_{\ell}^\omega)_{[1:r]}$,
-	Remark that $q_{\ell,r}$ has all of its indeterminates
+	\end{align*}
+	Remark that $\CutSuff{\ell,r}$ has all of its indeterminates
 	multiplied by $\omega$, so that the induction hypothesis
@@
 
-	The key remark is that $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
+	The key remark is that $\mu(\CutPref{\ell,r,b}(X_1, \dots, X_{\ell - 1}))$ is a
 	constant function from $\Nat^{\ell-1}$ to $M$, and that its value
@@
 	$m_{\ell,r,1}$ and $m_{\ell,r,0}$ the two elements of the monoid that are
-	obtained by $\mu(s_{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
+	obtained by $\mu(\CutPref{\ell,r,b}(X_1, \dots, X_{\ell-1}))$ respectively when
 	$b = 0$ and when $b > 0$. Finally, let us define $\pi_{\ell,r,0} \colon M^k
@@
 	range for the sum over $(\ell, r, b)$ to save space.
-	Similarly, we will use
-	$q_{\ell,r}(X_1, \dots, X_{p-\ell+1})
-		\defined q_{\ell,r}(X_1+b, X_2, \dots, X_{p - \ell + 1})$
-	at the end of the computation.
 	\begin{align*}
@@
 		 & = \sum_{(\ell, b, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
-		\pi(\mu(s_{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
+		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,r}(n_\ell - b, \dots, n_p)}
+		\pi(\mu(\CutPref{\ell,r,b}(n_1, \dots, n_{\ell-1})), \mu(v_2), \dots, \mu(v_{k+1}))
 		\\
@@
 		\sum_{(\ell, 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
+		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,b,0}(n_\ell - b, \dots, n_p)}
 		\pi(m_{\ell,r,0}, \mu(v_2), \dots, \mu(v_{k+1}))
@@
 		\sum_{(\ell, b > 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell - b, \dots, n_p)}
+		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,r}(n_\ell - b, \dots, n_p)}
 		\pi(m_{\ell,r,1}, \mu(v_2), \dots, \mu(v_{k+1}))
@@
 		\sum_{(\ell, 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,b,0}(n_\ell - b, \dots, n_p)}
+		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,b,0}(n_\ell - b, \dots, n_p)}
 		\pi_{\ell,r,0}(\mu(v_2), \dots, \mu(v_{k+1}))
@@
 		\sum_{(\ell, b > 0, r)}
-		\sum_{v_2 \dots v_{k+1} = q_{\ell,r}(n_\ell-b, \dots, n_p)}
+		\sum_{v_2 \dots v_{k+1} = \CutSuff{\ell,r}(n_\ell-b, \dots, n_p)}
 		\pi_{\ell,r,1}(\mu(v_2), \dots, \mu(v_{k+1}))
@@
 		\sum_{0 \leq r < \omega \times \card{u_\ell}}
-		(\pi_{\ell,r,0})^\dagger (q_{\ell, 0, r}(n_\ell-b, \dots, n_p))
+		(\pi_{\ell,r,0})^\dagger (\CutSuff{\ell, 0, r}(n_\ell-b, \dots, n_p))
 		\\
@@
 		\sum_{0 < b < n_\ell}
-		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
+		(\pi_{\ell,r,1})^{\dagger}(\CutSuff{\ell,r}(n_\ell - b, \dots, n_p))
 	\end{align*}
@@
 	\begin{equation*}
-		(\pi_{\ell,r,1})^{\dagger}(q_{\ell,r}(n_\ell - b, \dots, n_p))
+		(\pi_{\ell,r,1})^{\dagger}(\CutSuff{\ell,r}(n_\ell - b, \dots, n_p))
 		= F_{\ell,r,1}(\omega (n_\ell - b), \dots, \omega n_p)
@@
 	\begin{equation*}
-		(\pi_{\ell,r,0})^{\dagger}(q_{\ell,r}(n_1, \dots, n_{\ell-1})
+		(\pi_{\ell,r,0})^{\dagger}(\CutSuff{\ell,r}(n_1, \dots, n_{\ell-1})
 		= F_{\ell,r,0}(\omega n_1, \dots, \omega n_{\ell-1})
@@
 
-\begin{proofof}[natural-binomial-polynomial-positive-bad:ex]
-    $Q$ belongs to $\Rel[\vec{X}]$, but not to $\CorrectPoly$ because
-    $Q(0,Y,Z)$ has a negative maximal monomial,
-    hence it cannot be \kl{represented} by a \kl{star-free $\Nat$-polyregular function}
-    thanks to \cref{corrected-version:thm}.
-\end{proofof}
-
-\section{Proofs of section \ref{beyond-polynomials:sec}}
-
-\begin{proofof}[decompose-polynomial:lem]
-    Let us first write $f \in \ZPoly$ as a linear combination 
-    $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
-    We 
-    compute $\omega_+$ (resp. $\omega_-$)
-    by
-    applying
-    \cref{n-poly-combinatorics:lem} 
-    to $f_+$ (resp. to $f_-$).
-    Finally, let $\omega$ be the least common multiple
-    of $\omega_+$ and $\omega_-$, i.e. $\omega \defined \ppcm(\omega_+, \omega_-)$.
-
-    Using this $\omega$,
-    for any choice of
-    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
-    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
-    associated to $f_+$ and $f_-$,
-    such that $x_1, \dots, x_p \geq 1$:
+\begin{proofof}[polyrec-integer-strong:lem]
+    Because $n P \in \CorrectPoly$,
+    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ thanks to
+    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
+    to obtain a number $s \geq 1$ (that only depends on $P$) such
+    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
+    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
+    over $\Nat^k$. Let us write the
+    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
+    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
+    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
+    $\card{\alpha} \leq d_2$.
+    Now, because of the equality 
+    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
+    we conclude that we can find a $K \in \Nat$ and coefficients
+    $f_{\alpha} \in \Nat$ such that 
+    $F(X_1, \dots, X_k) =
+    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
+    Now, remark that $P$ is an \kl{integer-valued} polynomial of
+    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
+    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
+    \in \Rel$ for all $\card{\alpha} \leq d$
+    (\cref{integer-binomial-polynomial:lem}). 
+    As a consequence,
+    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
+    and 
+    by identifying coefficients in the previous two expressions, we obtain
+    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
+    Finally,
+    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
+    and therefore
+    $\lambda P (X_1 + s, \dots, X_k + s) =
+    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
+    In particular,
+    we get
     \begin{equation*}
-        f\left(
-        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right) 
-        = F_r^+(x_1, \dots, x_p) - F_r^-(x_1, \dots, x_p) \quad .
+        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
+        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
     \end{equation*}
-    Now, for large enough values of $x_1, \dots, x_p$,
-    this means that 
-    $f\left(
-        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right)$
-    coincides with an \kl{integer binomial polynomial},
-    and we can therefore conclude selecting different values of $r$ and
-    setting some subsets of variables to be assigned to the constant $0$.
+    And we conclude by term identification
+    by term identification
+    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
+    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
+    and $P$
+    is a \kl{natural binomial polynomial}.
+    Remark that this reasoning remains true whenever we partially apply $P$ to some values
+    in $\Nat$, as it amounts to 
+    partially applying the function $f$. We have proven that 
+    $P$ is a \kl{strongly natural binomial polynomial}.
 \end{proofof}

### 2024-09-12 parts/conclusion.tex
enormous rewrite

--- 2024-08-15 parts/conclusion.tex
@@
 
-\begin{lemma}
+\begin{lemma}[restate=pre-compose-growth-commut:lemma,label=pre-compose-growth-commut:lemma]
     \proofref{pre-compose-growth-commut:lemma}
-    \label{pre-compose-growth-commut:lemma}
     Let $f \in \ZRat$, and $k \in \Nat$. Then,
@@
 
-\begin{lemma}
+\begin{lemma}[restate=pre-compose-sf-commut:lemma,label=pre-compose-sf-commut:lemma]
     \proofref{pre-compose-sf-commut:lemma}
-    \label{pre-compose-sf-commut:lemma}
     Let $f \in \ZPoly$. Then, $f \in \ZSF$,
@@
 
-Remark that if those two lemmas were to hold for \kl{$\Nat$-polyregular
-functions}, then the decidability of $\NPoly$ inside $\ZPoly$ would follow
-immediately, and $\NSF$ inside $\NPoly$ would also immediately follow. On the
-one hand, one can guess a candidate function and check for equivalence, on the
-other hand, one can guess a \kl{commutative} \kl{star-free polyregular
-function} and check membership (which is decidable thanks to this paper).
+Remark that if \cref{pre-compose-sf-commut:lemma} were to hold for
+\kl{$\Nat$-polyregular functions}, then the decidability of $\NPoly$ inside
+$\ZPoly$, and the decidability of $\NSF$ inside $\NPoly$ would immediately
+follow. On the one hand, one can guess a candidate function and check for
+equivalence, on the other hand, one can guess a \kl{commutative} \kl{star-free
+polyregular function} and check membership (which is decidable thanks to this
+paper). This is restated in our concluding conjecture.
+
+\begin{conjecture}
+    \label{npoly-zpoly:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \ZPoly[k]$
+    belongs to $\NPoly[k]$ if and only if
+    it is \kl{$(k,\Nat)$-combinatorial}.
+\end{conjecture}

### 2024-09-12 parts/correction.tex
enormous rewrite

@@
 
-\AP We will discuss at the end of \cref{sec:poly-to-n-poly}
-why the
-example of \cref{thm:counter-example} is minimal in the number of
-indeterminates, which first requires us to provide a \emph{correct} analogue of
-\cref{karh:thm}. The
-counterexample provided by \cref{def:bad-polynomial}
-relies on the fact that
-$\CoveredPoly$ is not stable under fixing indeterminates, while
-\kl{$\Nat$-rational polynomials} are. Indeed, the polynomial $\BadPoly$
-introduced in \cref{def:bad-polynomial}
-satisfies $\BadPoly(X,Y,1) = 3X^2 +
-3Y^2 - 2XY$, which has a negative coefficient for a \kl{maximal monomial},
-namely $-2XY$. In this section, we prove that closing $\CoveredPoly$ under
-variable assignments is enough to recover from
-\cref{karh:thm}. We
-use the following notation to fix the value of some indeterminate, if $P(X,Y)$
-is a polynomial in $\Rel[X,Y]$, then $\intro*\restr{P(X,Y)}{X = 1}$ is the
-polynomial $P(1,Y) \in \Rel[Y]$. More generally, if $\nu$ is a partial function
-from $\vec{X}$ to $\Nat$, written $\nu \colon \vec{X} \topartial \Nat$, the
-restriction $\restr{P(\vec{X})}{\nu}$ is the polynomial with indeterminates
-$\vec{Y} \defined \vec{X} - \dom(\nu)$ obtained by fixing the variables of the
-domain of $\nu$.
+\AP We will discuss at the end of \cref{sec:poly-to-n-poly} why
+\cref{thm:counter-example} is minimal in the number of indeterminates, which
+first requires us to provide a \emph{correct} analogue of \cref{karh:thm}. Our
+counterexample relies on the fact that $\CoveredPoly$ is not stable under
+fixing indeterminates, while \kl{$\Nat$-rational polynomials} are. Indeed, the
+polynomial $\BadPoly$ satisfies
+$\BadPoly(X,Y,1) = 3X^2 + 3Y^2 - 2XY$, which has a negative coefficient for a
+\kl{maximal monomial}. Let us now prove that closing
+$\CoveredPoly$ under variable assignments is enough to recover from
+\cref{karh:thm}. We use the following notation to fix the value of some
+indeterminate, if $P(X,Y)$ is a polynomial in $\Rel[X,Y]$, then
+$\intro*\restr{P(X,Y)}{X = 1}$ is the polynomial $P(1,Y) \in \Rel[Y]$. More
+generally, if $\nu$ is a partial function from $\vec{X}$ to $\Nat$, written
+$\nu \colon \vec{X} \topartial \Nat$, the restriction $\restr{P(\vec{X})}{\nu}$
+is the polynomial with indeterminates $\vec{Y} \defined \vec{X} - \dom(\nu)$
+obtained by fixing the variables of the domain of $\nu$.
 
@@
 
-\begin{theorem}[restate=corrected-version:thm,label=corrected-version:thm]
+\begin{theorem}[label=corrected-version:thm,]
 	Let $P \in \Rel[\vec{X}]$.
@@
 
-\begin{remark}
-    \label{undecidable-non-negative:lem}
+\begin{remark}[restate=undecidable-non-negative:lem,label=undecidable-non-negative:lem]
     \proofref{undecidable-non-negative:lem}
@@
 to compute polynomials in $\CorrectPoly$ using $\NPoly$
-(\cref{lem:correct-to-n-rat}). With this in hand, we will next generalize these
-results to polynomials in $\Rat[\vec{X}]$ in \cref{sec:rel-to-rat}.
+(\cref{lem:correct-to-n-rat}). Then, we will next generalize
+\cref{corrected-version:thm} to polynomials in $\Rat[\vec{X}]$ in
+\cref{sec:rel-to-rat}.
 
@@
 \prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
-non-commutative monomial over the monoid $\Sigma^*$. Almost by
-definition, \kl{polynomial functions} are \kl{commutative} \kl{polyregular
-	functions}.
+non-commutative monomial over the monoid $\Sigma^*$. \kl{Polynomial functions}
+are \kl{commutative} \kl{polyregular functions}.
 
@@
 extra assumption that the function is \kl{star-free polyregular}, as described
-in the following \cref{n-sf-combinatorics:lem}.
+in the following \cref{n-sf-combinatorics:lem}. This lemma is not immediately
+of use, but is crucial for the upcoming characterization of $\Nat$-rational
+polynomials in \cref{decide-rat-poly-npoly:cor}, which in turn is a key
+ingredient of our main \cref{zsf-npoly-nsf:thm}.
 
@@
 \begin{proof}
-	Let $f$ be the \kl{commutative}
-	\kl{$\Nat$-polyregular function}
+    Let $f$ be a \kl{commutative} \kl{$\Nat$-rational series}
 	with domain defined as $\Sigma \defined \set{a_1, \dots, a_p}$
 	that \kl{represents} $P$.
+    Because $f$ has \kl{polynomial growth},
+    $f \in \NPoly$ 
+    (\cref{polyregular-polynomial-growth:lemma}).
 	Using \cref{n-poly-combinatorics:lem},
@@
 	\end{equation*}
-	Let us remark that for large enough values of $X$, the \kl{binomial function}
+    For large enough values of $X$, the \kl{binomial function}
 	$\binom{X - p}{k}$ coincides with a polynomial whose dominant coefficient
@@
 
-	Remark that for every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
+	For every partial valuation $\nu \colon \vec{X} \topartial \Nat$,
 	the polynomial $\restr{P}{\nu}$ continues to be represented
@@
 
-\AP We have proven that every \kl{$\Nat$-rational polynomial} is in
-$\CorrectPoly$, and the rest of this section is devoted to proving the converse
-implication. The core of the upcoming \cref{lem:correct-to-n-rat}
-leverages a notion of \intro{discrete derivative} which is built by
-translating the domain of the polynomial. To that end, let us write
-$\intro*\translate{K}$ for the \intro{translation function} that maps a
-polynomial $P \in \Rel[X_1, \dots, X_n]$ to the polynomial $P(X_1 + K, \dots,
-X_n + K)$.
+\AP This section is devoted to proving that polynomials in $\CorrectPoly$ can
+be \kl{represented} by \kl{star-free $\Nat$-polyregular functions}. The key
+lemma of this section is \cref{lem:correct-to-n-rat}, which is proved by
+induction on the number of indeterminates of a given polynomial $P$. In order
+to prove that result, we use the combinatorial
+\cref{derivation-stabilises-correct:lem} that allows us to transform a
+polynomial $P \in \CorrectPoly$ into a polynomial in $\Nat[\vec{X}]$ through a
+well-chosen translation of the indeterminates. This argument is based on the
+notion of \intro{discrete derivative} which is built by translating the domain
+of the polynomial. To that end, let us write $\intro*\translate{K}$ for the
+\intro{translation function} that maps a polynomial $P \in \Rel[X_1, \dots,
+X_n]$ to the polynomial $P(X_1 + K, \dots, X_n + K)$.
 
@@
 \end{definition}
-
-\AP The key lemma of this section is \cref{lem:correct-to-n-rat}, which is
-proved by induction on the number of indeterminates of a given polynomial $P$.
-Notice that in \cref{lem:correct-to-n-rat}, polynomials can be represented
-using \kl{star-free $\Nat$-polyregular functions}, which is a stronger
-statement than the one in \cref{karh:thm}, but is in line with the idea that
-polynomials are not periodic functions (this will be made precise in
-\cref{zsf-nsf:conjecture}). In order to prove that result, we use the
-combinatorial \cref{derivation-stabilises-correct:lem} that allows us to
-transform a polynomial $P \in \CorrectPoly$ into a polynomial in
-$\Nat[\vec{X}]$ through a well-chosen translation of the indeterminates.
 
@@
 decidable, and even show that elements in $\CorrectPoly$ are, up to suitable
-translations, polynomials in $\Nat[\vec{X}]$. Beware that partial applications
-are still needed in this characterization, as \cref{bad-poly-translate:ex}
-illustrates. Effectiveness was the last missing piece of our main result
-\cref{corrected-version:thm}, which corrects and generalizes \cref{karh:thm}.
-\begin{example}
-	\label{bad-poly-translate:ex}
-	The polynomial $\BadPoly$ is not a
-	\kl{$\Nat$-rational polynomial},
-	but is \kl{non-negative} and satisfies
-	$\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
-\end{example}
+translations, polynomials in $\Nat[\vec{X}]$
+(\cref{derivation-translation:lem}). Beware that partial applications are still
+needed in this characterization, as \cref{bad-poly-translate:ex} illustrates.
 
-\begin{lemma}
-	\label{derivation-translation:lem}
+\begin{lemma}[restate=derivation-translation:lem,label=derivation-translation:lem]
 	\proofref{derivation-translation:lem}
@@
 \end{lemma}
+
+\begin{example}
+	\label{bad-poly-translate:ex}
+	The polynomial $\BadPoly$ is not a
+	\kl{$\Nat$-rational polynomial},
+	but is \kl{non-negative} and satisfies
+	$\translate{10}(\BadPoly) \in \Nat[\vec{X}]$.
+\end{example}
+
+We now have all the tools to prove the corrected version of \cref{karh:thm}.
 
@@
 
-\AP Let us complete our analysis of polynomials \kl{represented} by $\NPoly$ or
-$\ZPoly$ by considering polynomials with coefficients in $\Rat$, and justify
-that all the combinatorial work has already happened in $\Rel$ and $\Nat$. A
-first observation is that from \cref{n-sf-combinatorics:lem}, we
-know that the polynomials that can be computed by \kl{star-free
-$\Nat$-polyregular functions} are going to be suitable combinations of
-\kl{natural binomial functions}. This motivates the following definition of the
-``polynomial counterpart" of a binomial coefficient, given two numbers $\ell,k
-\in \Nat$, $\intro*\pbinom{X - \ell}{k}$ defined as $(X - \ell) \cdots (X -
-\ell - k) / k!$,\footnote{ In particular, $\pbinom{X - \ell}{k}$ is defined to
-be $1$ when $k = 0$, and $X - \ell$ when $k = 1$. } that we call a
-\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
-as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
-the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
-- p_{i,j}}{k_{i,j}}$. Similarly, we introduce the class of \intro{integer
+\AP Let us complete our analysis of polynomials \kl{represented} by $\NRat$
+or $\ZRat$ by considering polynomials with coefficients in $\Rat$, and
+justify that all the combinatorial work has already happened in $\Rel$ and
+$\Nat$. A first observation is that from \cref{n-sf-combinatorics:lem}, we know
+that the polynomials that can be computed by \kl{star-free $\Nat$-polyregular
+functions} are going to coincide (on large enough inputs) with \kl{natural binomial
+functions}. For that reason, we introduce the following ``polynomial
+counterpart" of a binomial coefficient: given two numbers $\ell,k \in \Nat$,
+$\intro*\pbinom{X - \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) /
+k!$,\footnote{ In particular, $\pbinom{X - \ell}{k}$ is defined to be $1$ when
+$k = 0$, and $X - \ell$ when $k = 1$. } that we call a \intro{binomial
+monomial}, and we introduce \intro{natural binomial polynomials} as $\Nat$-linear
+combinations of products of \kl{binomial monomials}, i.e., of the shape:
+$P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j -
+p_{i,j}}{k_{i,j}}$. Similarly, we introduce the class of \intro{integer
 binomial polynomials}, which are obtained by $\Rel$-linear combinations of
@@
 \AP
-These definitions are motivated by the classical result of Pólya that
+These definitions are justified by the classical result of Pólya that
 characterizes polynomials $P$ in $\Rat[X]$ that \intro{integer-valued} (i.e., are such
@@
 	\begin{enumerate}
-		\item \label{int-bin-0:item} $P$ is an \kl{integer binomial polynomial},
-		\item \label{int-bin-1:item} $P$ \kl{represented} by a \kl{star-free $\Rel$-polyregular function},
-		\item \label{int-bin-2:item} $P$ \kl{represented} by a \kl{$\Rel$-polyregular function},
-        \item \label{int-bin-3:item} $P$ is \kl{integer-valued}.
+        \item \label{int-bin-3:item} $P$ is \kl{integer-valued},
+        \item \label{int-bin-2s:item} $P$ is \kl{represented} by a \kl{$\Rel$-rational series},
+		\item \label{int-bin-2:item} $P$ is \kl{represented} by a \kl{$\Rel$-polyregular function},
+		\item \label{int-bin-1:item} $P$ is \kl{represented} by a \kl{star-free $\Rel$-polyregular function},
+		\item \label{int-bin-0:item} $P$ is an \kl{integer binomial polynomial}.
 	\end{enumerate}
@@
 \begin{proof}
-	Let us first note that implications \cref{int-bin-1:item}
-	$\implies$ \cref{int-bin-2:item} $\implies$ \cref{int-bin-3:item} are obvious.
+	The implications \cref{int-bin-1:item}
+	$\implies$ \cref{int-bin-2:item} $\implies$ 
+    \cref{int-bin-2s:item}
+    $\implies$ \cref{int-bin-3:item} are obvious.
 	Now, \cref{int-bin-3:item} $\implies$ \cref{int-bin-0:item} is a direct
@@
 $\Rat[\vec{X}]$ that are \kl{represented} by $\NPoly$ are not exactly
-\kl{natural binomial polynomials},
-(\cref{natural-binomial-polynomial-positive-bad:ex}). To address the issues
+\kl{natural binomial polynomials}
+(see \cref{natural-binomial-polynomial-positive-bad:ex}). To address the issues
 raised by the former example, we introduce the notion of \intro{strongly
@@
 \begin{example}[restate=natural-binomial-polynomial-positive-bad:ex,label=natural-binomial-polynomial-positive-bad:ex]
-	\proofref{natural-binomial-polynomial-positive-bad:ex}
+	%\proofref{natural-binomial-polynomial-positive-bad:ex}
 	The polynomial
@@
 	but cannot be computed by a \kl{star-free $\Nat$-polyregular function}.
+    Indeed, 
+    $Q(0,Y,Z)$ has a negative maximal monomial, hence $Q \not \in \CorrectPoly$,
+    and we conclude using 
+    \cref{corrected-version:thm}.
 \end{example}
+
+\begin{lemma}[restate=polyrec-integer-strong:lem,label=polyrec-integer-strong:lem]
+    \proofref{polyrec-integer-strong:lem}
+    Let $P \in \Rat[\vec{X}]$ be an \kl{integer-valued} polynomial, 
+    and $n \in \Nat_{\geq 1}$ be
+    such that $n P \in \Rel[\vec{X}] \cap \CorrectPoly$.
+    Then, $P$ is a \kl{strongly natural binomial polynomial}.
+\end{lemma}
 
@@
 	\begin{enumerate}
+        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$ 
+            and $P$ is \kl{integer-valued},
+        \item \label{rat-npoly-0:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
 		\item \label{rat-npoly-1:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
@@
 		\item \label{rat-npoly-3:item}
-		      $P$ is a \kl{strongly natural binomial polynomial},
-        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$ 
-            and $P$ is \kl{integer-valued}.
+		      $P$ is a \kl{strongly natural binomial polynomial}.
 	\end{enumerate}
@@
 \begin{proof}
+    Let us first remark that $\NPoly \subseteq \NRat$,
+    and that if $P$ is \kl{represented} by a function $f \in \NRat$,
+    then said function has \kl{polynomial growth}, and in particular
+    $f \in \NPoly$ thanks to \cref{polyregular-polynomial-growth:lemma}.
+    As a consequence, \cref{rat-npoly-0:item} $\iff$ \cref{rat-npoly-1:item}.
     For the implication \cref{rat-npoly-1:item} $\implies$
@@
     $\NSF \subseteq \ZPoly$.
-
-    Then, \cref{rat-npoly-4:item} $\implies$
-    \cref{rat-npoly-3:item}, because if $\alpha P \in \CorrectPoly$,
-    it is represented by a \kl{star-free $\Nat$-polyregular function} $f$ by
-    \cref{corrected-version:thm}. Let us now use \cref{n-sf-combinatorics:lem}
-    to obtain a number $s \geq 1$ (that only depends on $P$) such
-    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
-    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
-    over $\Nat^k$. Let us write the
-    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
-    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
-    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
-    $\card{\alpha} \leq d_2$.
-    Now, because of the equality 
-    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
-    we conclude that we can find a $K \in \Nat$ and coefficients
-    $f_{\alpha} \in \Nat$ such that 
-    $F(X_1, \dots, X_k) =
-    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
-    Now, remark that $P$ is an \kl{integer-valued} polynomial of
-    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
-    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
-    \in \Rel$ for all $\card{\alpha} \leq d$
-    (\cref{integer-binomial-polynomial:lem}). 
-    As a consequence,
-    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
-    and 
-    by identifying coefficients in the previous two expressions, we obtain
-    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
-    Finally,
-    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
-    and therefore
-    $\lambda P (X_1 + s, \dots, X_k + s) =
-    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
-    In particular,
-    we get
-    \begin{equation*}
-        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
-        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
-    \end{equation*}
-    And we conclude by term identification
-    by term identification
-    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
-    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
-    and $P$
-    is a \kl{natural binomial polynomial}.
-    Remark that this reasoning remains true whenever we partially apply $P$ to some values
-    in $\Nat$, as it amounts to 
-    partially applying the function $f$. We have proven that 
-    $P$ is a \kl{strongly natural binomial polynomial}.
+    The implication \cref{rat-npoly-4:item} $\implies$
+    \cref{rat-npoly-3:item} is obtained thanks to \cref{polyrec-integer-strong:lem}.
 

### 2024-09-12 parts/counter.tex
enormous rewrite

--- 2024-09-02 parts/counter.tex
@@
 \AP In this section, we will completely characterize which polynomials in
-$\Rat[\vec{X}]$ that are \kl{represented} by \kl{$\Nat$-polyregular functions}
-(resp. \kl{$\Rel$-polyregular functions}). To that end, we start by
-characterizing these classes for polynomials in $\Rel[\vec{X}]$.
-We say that a polynomial $P \in \Rel[X_1, \dots, X_n]$ is an
-\intro{$\Nat$-rational polynomial} if it is \kl{represented} by a
-\kl{$\Nat$-rational series}, in which case it is also \kl{represented} by a
-\kl{$\Nat$-polyregular function}, because it has \kl{polynomial growth}
-(\cref{polyregular-polynomial-growth:lemma}).
-The analogue notion of \intro{$\Rel$-rational polynomials} is not
-of particular interest, since every polynomial $P \in \Rel[X_1, \dots, X_n]$ is
-\kl{represented} by a \kl{$\Rel$-polyregular function}. 
-It is an easy check that
+$\Rat[\vec{X}]$ that are \kl{represented} by \kl{$\Nat$-rational series} (resp.
+\kl{$\Rel$-rational series}). To that end, we start by characterizing these
+classes for polynomials in $\Rel[\vec{X}]$. We say that a polynomial $P \in
+\Rel[X_1, \dots, X_n]$ is an \intro{$\Nat$-rational polynomial} if it is
+\kl{represented} by a \kl{$\Nat$-rational series}. It is an easy check that
 polynomials with coefficients in $\Nat$ are \kl{$\Nat$-rational polynomials}
@@
 a polynomial with negative coefficients that is an \kl{$\Nat$-rational
-polynomial}. 
-The problem of characterizing \kl{$\Nat$-rational polynomials} was claimed to
-be solved in \cite{KARH77}, using the \cref{karh:def}
-to  characterize
-\kl{$\Nat$-rational polynomials}, as restated in \cref{karh:thm}.
+polynomial}. The problem of characterizing \kl{$\Nat$-rational polynomials} was
+claimed to be solved in \cite{KARH77}, using the \cref{karh:def} to
+characterize \kl{$\Nat$-rational polynomials}, as restated in \cref{karh:thm}.
 
-\begin{lemma}
-    \label{n-poly-n-poly:example}
+\begin{lemma}[restate=n-poly-n-poly:example,label=n-poly-n-poly:example]
     \proofref{n-poly-n-poly:example}
@@
 \end{lemma}
-\begin{proof}
-    The polynomials of $\Nat[\vec{X}]$
-    are obtained from basic functions (constant in $\Nat$,
-    $X_i$ for some $1 \leq i \leq n$)
-    by products and sums (\cref{stability-polyregular:lemma}). Because the basic functions are
-    \kl{represented} by \kl{star-free $\Nat$-polyregular} functions (see
-    \cref{size-of-word-nsf:ex}),
-    that are closed under these operations, we conclude.
-\end{proof}
 
-\begin{example}
-    \label{negative-but-npoly:ex}
-    \proofref{negative-not-nrat:ex}
+\begin{example}[restate=negative-but-npoly:ex,label=negative-but-npoly:ex]
+    \proofref{negative-but-npoly:ex}
     The polynomials $X$, $X^2 + 3$,
@@
     are \kl{$\Nat$-rational polynomials},
-    but $- X$ is a \kl{$\Rel$-rational polynomial} that is 
+    but $- X$ is 
     not an \kl{$\Nat$-rational polynomial}.
@@
     but is not an \kl{$\Nat$-rational polynomial}.
-    As a corollary, 
-    the result stated in \cite[Theorem 3.3]{KARH77}, restated
+    As a corollary, \cite[Theorem 3.3]{KARH77}, restated
     in \cref{karh:thm}, is false
@@
     $\BadPoly(\card[a]{w}, \card[b]{w}, \card[c]{w}) = f(w)$.
-
     Remark that for all $x,y,z \geq 0$, $\BadPoly(x,y,z) = 0$
@@
     $z \neq 0$, and $x = y = 0$.
-
     Now, let us consider the language $L \defined \setof{w}{ f(w) = 0}$. By the

### 2024-09-12 parts/intro.tex
enormous rewrite

@@
 characterizations in terms of minimal automaton and/or syntactic monoid.
+
+\AP Let us briefly recall one of the equivalent definitions of a
+\intro{$K$-rational series} (where $K$ is a ring), and refer to \cite{BERE10}
+for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
+to $K$ are the functions computed by \kl{$K$-weighted automata} which are
+non-deterministic finite automata with transitions labelled by elements of $K$,
+and whose output on a given word is the sum over all accepting paths of the
+products of the weights along the path. For instance, the function $w \mapsto
+2^{\card{w}}$ is computed by an \kl{$\Nat$-weighted automaton} with a single
+state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
+series}. We use $\intro*\NRat$ and $\intro*\ZRat$ to denote respectively
+the set of \kl{$\Nat$-rational series} and the set of \kl{$\Rel$-rational
+series}.
+
+\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
+as having outputs in $\Nat$, and this is why they were introduced respectively
+as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
+with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$ for
+\kl{star-free polyregular functions} with unary output) \cite{DOUE21,DOUE22}.
+From this point of view, $\NPoly$ is a subclass of a larger and well-studied
+class of functions called (non-commutative) \kl{rational series}
+\cite{REUT80,BERE88,BERE10}. This other model of computation generalizes the
+notion of regular language to functions with outputs in a \emph{commutative}
+semiring, and it was frequently rediscovered that \kl{$\Nat$-polyregular
+functions} form a (strict) subclass of \kl{rational series} with outputs in
+$\Nat$, called \kl{$\Nat$-rational series} ($\NRat$)
+\cite{SCHU62,KRRC13,CDTL23}.
+
+\AP In order to approach the decision problems on \kl{polyregular functions},
+restricting the output alphabet to a single letter has proven to be a fruitful
+approach \cite{DOUE21,DOUE22}. The key advantage of single-letter output
+alphabets is that the particular order in which letters are produced by the
+machine is irrelevant. This restriction greatly simplifies the definition of
+\kl{polyregular functions}, and opens the door to results that would otherwise
+fail because of this ordering, such as the so-called \emph{pebble minimization
+theorem} of \cite{DOUE21}, which fails in the general case
+\cite{BOJA22,KLEP23}. However, even with the restriction to
+single letter output alphabets, the decidability of \kl{star-free polyregular
+functions} among \kl{polyregular functions} remains an open problem
+\cite[Conjecture 7.61]{DOUE23}.
+
+\AP This approach through rational series bears numerous limitations. First,
+$\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
+because of the extra expressive power introduced by the negative outputs
+\cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
+is an open problem (see e.g. \cite{KARH77}), that remains open even in the
+simple case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It was
+conjectured that, being ``computable using coefficients in $\Nat$'' and being
+``star-free'' were two orthogonal properties, and therefore that the following
+equation would hold: $\NSF = \NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}.
+This would imply the decidability of aperiodicity for single letter output
+polyregular functions since aperiodicity is decidable for $\ZSF$. We refer the
+reader to \cref{previously-known-inclusions:fig} to get an overview of the
+different classes of functions appearing in the paper.
+
+\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
+the membership of $\ZSF$ inside $\ZPoly$,
+relies on the construction of a canonical object called the \emph{residual
+transducer}, the latter being essentially based on differences between
+functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
+proof method does not carry from $\Rel$-output functions to $\Nat$-output
+functions, it was conjectured that the semantic property of being
+\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$ (\cref{zsf-nsf:conjecture}).
+
+\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
+    \label{zsf-nsf:conjecture}
+    Let $k \in \Nat$.
+    A function $f \in \NPoly[k]$
+    belongs to $\NSF[k]$ if and only if
+    it is \kl{ultimately polynomial}.
+    In particular,
+    $\NSF[k] = \ZSF[k] \cap \NPoly$.
+\end{conjecture}
+
+\begin{figure}
+    \centering
+    \includestandalone[width=7cm]{tikz/class-inclusions}
+    \caption{
+        Decidability and inclusions of classes of functions,
+        arranged along two axes. The first one is the complexity
+        of the output alphabet ($\Rel$, $\Nat$, $\Sigma$). The second
+        one is the allowed computational power
+        (\kl{star-free polyregular functions}, \kl{polyregular functions}, 
+        \kl{rational series}).
+        Arrows denote strict inclusions,
+        and effectiveness (both in terms of decidability and of effective
+        representation) is represented by thick double arrows. Inclusions that are
+        suspected to be effective are represented using a dashed arrow together with a
+        question mark.
+    }
+    \label{previously-known-inclusions:fig}
+\end{figure}
+
+In \cite{CDTL23}, the authors introduced the class of \kl{$\Rel$-polyregular
+functions} ($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by
+allowing negative outputs. This remains a subclass of \kl{rational series},
+more precisely, of the class of \kl{$\Rel$-rational series}
+($\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
+been shown to be decidable \cite{CDTL23}. Let us underline that several
+non-equivalent notions of aperiodicity coexist for rational series
+\cite{REUT80,DRGA19,CDTL23}, and that we only refer to the one applying to
+\kl{rational series} that are actually in $\ZPoly$ \cite{CDTL23}.
 
@@
 
-\AP In order to approach the decision problems on \kl{polyregular functions},
-restricting the output alphabet to a single letter has proven to be a fruitful
-approach \cite{DOUE21,DOUE22}. The key advantage of single-letter output
-alphabets is that the particular order in which letters are produced by the
-machine is irrelevant. This restriction greatly simplifies the definition of
-\kl{polyregular functions}, and opens the door to results that would otherwise
-fail because of this ordering, such as the so-called \emph{pebble minimization
-theorem} of \cite{DOUE21}, which fails in the general case
-\cite{BOJA22,KLEP23}. However, even with the restriction to
-single letter output alphabets, the decidability of \kl{star-free polyregular
-functions} among \kl{polyregular functions} remains an open problem
-\cite[Conjecture 7.61]{DOUE23}.
-
-\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
-as having outputs in $\Nat$, and this is why they were introduced respectively
-as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
-with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$ for
-\kl{star-free polyregular functions} with unary output) \cite{DOUE21,DOUE22}.
-From this point of view, $\NPoly$ is a subclass of a larger and well-studied
-class of functions called (non-commutative) \kl{rational series}
-\cite{REUT80,BERE88,BERE10}. This other model of computation generalizes the
-notion of regular language to functions with outputs in a \emph{commutative}
-semiring, and it was frequently rediscovered that \kl{$\Nat$-polyregular
-functions} form a (strict) subclass of \kl{rational series} with outputs in
-$\Nat$, called \kl{$\Nat$-rational series} ($\NRat$)
-\cite{SCHU62,KRRC13,CDTL23}.
-
-In \cite{CDTL23}, the authors introduced the class of \kl{$\Rel$-polyregular
-functions} ($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by
-allowing negative outputs. This remains a subclass of \kl{rational series},
-more precisely, of the class of \kl{$\Rel$-rational series}
-($\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
-been shown to be decidable \cite{CDTL23}. Let us underline that several
-non-equivalent notions of aperiodicity coexist for rational series
-\cite{REUT80,DRGA19,CDTL23}, and that we only refer to the one applying to
-\kl{rational series} that are actually in $\ZPoly$ \cite{CDTL23}.
-
-\AP This approach through rational series bears numerous limitations. First,
-$\ZPoly$ does not correspond to a restriction of \kl{polyregular functions},
-because of the extra expressive power introduced by the negative outputs
-\cite{CDTL23}. Furthermore, deciding the membership of $\NRat$ inside $\ZRat$
-is an open problem (see e.g. \cite{KARH77}), that remains open even in the
-simple case of $\NPoly$ and $\ZPoly$ \cite[Open question 5.55]{DOUE23}. It was
-conjectured that, being ``computable using coefficients in $\Nat$'' and being
-``star-free'' were two orthogonal properties, and therefore that the following
-equation would hold: $\NSF = \NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}.
-This would imply the decidability of aperiodicity for single letter output
-polyregular functions since aperiodicity is decidable for $\ZSF$. We refer the
-reader to \cref{previously-known-inclusions:fig} to get an overview of the
-different classes of functions appearing in the paper.
-
-\begin{figure}
-    \centering
-    \includestandalone[width=7cm]{tikz/class-inclusions}
-    \caption{
-        Decidability and inclusions of classes of functions,
-        arranged along two axes. The first one is the complexity
-        of the output alphabet ($\Rel$, $\Nat$, $\Sigma$). The second
-        one is the allowed computational power
-        (\kl{star-free polyregular functions}, \kl{polyregular functions}, 
-        \kl{rational series}).
-        Arrows denote strict inclusions,
-        and effectiveness (both in terms of decidability and of effective
-        representation) is represented by thick double arrows. Inclusions that are
-        suspected to be effective are represented using a dashed arrow together with a
-        question mark.
-    }
-    \label{previously-known-inclusions:fig}
-\end{figure}
-
 \subparagraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular

### 2024-09-12 parts/prelim.tex
enormous rewrite

@@
 
-\AP In the rest of this paper, we use the symbol $\Rel$ to denote the set of
-integers, $\Rat$ to denote the set of rational numbers, and $\Nat$ to denote
-the set of non-negative integers. The capital letters $\Sigma,\Gamma$ denote
+\AP The capital letters $\Sigma,\Gamma$ denote
 fixed alphabets, i.e. finite set of letters, and $\Sigma^*, \Gamma^*$ (resp.
@@
 monoid $M$ is finite, this $n$ can be uniformly chosen for all elements in $M$.
-
-\AP Let us briefly recall that one of the equivalent definitions of a
-\intro{$K$-rational series} (where $K$ is a ring), and refer to \cite{BERE10}
-for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
-to $K$ are the functions computed by \kl{$K$-weighted automata} which are
-non-deterministic finite automata with transitions labelled by elements of $K$,
-and whose output on a given word is the sum over all accepting paths of the
-products of the weights along the path. For instance, the function $w \mapsto
-2^{\card{w}}$ is computed by an \kl{$\Nat$-weighted automaton} with a single
-state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
-series}. We use $\intro*\NRat$ and $\intro*\ZRat$ to denote respectively
-the set of \kl{$\Nat$-rational series} and the set of \kl{$\Rel$-rational
-series}.
 
@@
 general definition of \kl{polyregular functions}. 
-
-\AP We chose in this paper to provide a very combinatorial description of
+We chose in this paper to provide a very combinatorial description of
 polyregular functions with commutative outputs because it will play nicely with

### 2024-09-12 parts/starfree.tex
enormous rewrite

@@
     \label{decompose-polynomial:lem}
-    \proofref{decompose-polynomial:lem}
     Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
@@
 \end{lemma}
+\begin{proof}
+    Let us first write $f \in \ZPoly$ as a linear combination 
+    $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
+    We 
+    compute $\omega_+$ (resp. $\omega_-$)
+    by
+    applying
+    \cref{n-poly-combinatorics:lem} 
+    to $f_+$ (resp. to $f_-$).
+    Finally, let $\omega$ be the least common multiple
+    of $\omega_+$ and $\omega_-$, i.e. $\omega \defined \ppcm(\omega_+, \omega_-)$.
+    Using this $\omega$,
+    for any choice of
+    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
+    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
+    associated to $f_+$ and $f_-$,
+    such that $x_1, \dots, x_p \geq 1$:
+    \begin{equation*}
+        f\left(
+        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right) 
+        = F_r^+(x_1, \dots, x_p) - F_r^-(x_1, \dots, x_p) \quad .
+    \end{equation*}
+    Now, for large enough values of $x_1, \dots, x_p$,
+    this means that 
+    $f\left(
+        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right)$
+    coincides with an \kl{integer binomial polynomial},
+    and we can therefore conclude selecting different values of $r$ and
+    setting some subsets of variables to be assigned to the constant $0$.
+\end{proof}
 
@@
 \kl{$\Nat$-polyregular functions} can be combined with the recent advances in
-the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} to decide
-aperiodicity. The key ingredient of this study is the use of a semantic
-characterization of \kl{star-free $\Rel$-polyregular functions} among
-\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
-functions by the means of polynomial behaviors (see
-\cref{aperidic-ultimately-polynomial:ex}). Historically, the definition used
-polynomials with coefficients in $\Rat$, which is why we use such polynomials
-here, even though we know that all of those polynomials will actually be
-\kl{integer binomial polynomials} thanks to
-\cref{integer-binomial-polynomials:cor}.
+the study of \kl{$\Rel$-polyregular functions} \cite{CDTL23} allowing to decide
+the membership of $\ZSF$ inside $\ZPoly$. The key ingredient of this study is
+the use of a semantic characterization of \kl{star-free $\Rel$-polyregular
+functions} among \kl{$\Rel$-rational series} that generalizes the notion of
+aperiodicity for languages to functions.
 
@@
 
-Let us recall that a language $L$ is \intro(language){aperiodic} whenever it is
-recognized by an \kl(monoid){aperiodic} finite monoid, which turns out to be
-equivalent to the following characterization among regular languages: for all
-$u,v,w$, there exists $n \in \Nat$, $u v^n w \in L$ if and only if $u v^{n+1} w
-\in L$. As a consequence of this characterization, a language $L$ is aperiodic
-if and only if its indicator function $\ind{L}$ is \kl{ultimately polynomial}.
-This, was already observed in \cite[Claim V.6]{CDTL23}, and restated in
-\cite[Claim 7.45, Lemma 7.53]{DOUE23}. We include a self-contained proof
-hereafter for readability and didactic purposes.
+It was observed in \cite[Claim V.6]{CDTL23}, and in \cite[Claim 7.45, Lemma
+7.53]{DOUE23} that a regular language $L$ is \emph{star-free} if and only if its
+indicator function $\ind{L}$ is \kl{ultimately polynomial}. We can now answer
+\cite[Conjecture 7.61]{DOUE23} positively, by proving that $\NPoly \cap \ZSF =
+\NSF$.
 
-\begin{example}
-    \label{aperidic-ultimately-polynomial:ex}
-    A language $L$ is aperiodic if and only if 
-    $\ind{L}$ is \kl{ultimately polynomial}.
-\end{example}
-\begin{proof}
-    Assume that $\ind{L}$ is \kl{ultimately polynomial}. Let $u,v,w \in
-    \Sigma^*$. There exists a polynomial $P \in \Rat[X]$ and a number $N_0 \in
-    \Nat$, such that $\ind{L}(u v^n w) = P(n)$ for all $n \geq N_0$. Because
-    $\ind{L}$ has co-domain included in $\set{0,1}$, so does $P$, which proves
-    that $P$ is a constant polynomial. Then, $\ind{L}(u v^{n+1} w) = P(n+1) =
-    P(n) = \ind{L}(uv^n w)$ for all $n \geq N_0$. We have proven that $L$
-    is aperiodic.
-
-    Conversely, assume that $L$ is \kl(language){aperiodic}, let $n \in \Nat$ be a number, and
-    let $\alpha_0$, $w_1$, $\alpha_1$, $\ldots$, $\alpha_{n-1}$, $w_n$, $\alpha_n$
-    be words in $\Sigma^*$. Because $L$ is \kl(language){aperiodic}, the language
-    $L$ is recognized by an \kl(monoid){aperiodic} finite monoid $M$, via a
-    morphism $\mu \colon \Sigma^* \to M$ and an accepting part $F \subseteq M$.
-    Because $M$ is a \kl(monoid){aperiodic}, there exists $N_0$
-    such that for all $m \in M$, $m^{N_0+1} = m^{N_0}$. As a consequence,
-    for all $1 \leq i \leq n$,
-    whenever $x_i \geq N_0$, we have $\mu(w_i)^{x_i} = \mu(w_i)^{N_0}$.
-    Therefore, for all $x_1, \dots, x_n \geq N_0$, we have:
-    \begin{align*}
-        \ind{L}\left(\alpha_0 \prod_{i= 1}^n w_i^{x_i} \alpha_i \right) 
-        &= 
-        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{x_i} \cdot \mu(\alpha_i)\right)
-        & \text{ because $M$ recognizes $L$}
-        \\
-        &= 
-        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{N_0} \cdot \mu(\alpha_i)\right)
-        & \text{ by aperiodicity }
-        \\
-        &\defined P(x_1, \dots, x_n) & \text{ constant polynomial }
-    \end{align*}
-    Which proves that $\ind{L}$ is \kl{ultimately polynomial}.
-\end{proof}
-
-\AP The decidability of aperiodicity for \kl{$\Rel$-polyregular functions}, that is
-the membership of $\ZSF$ inside $\ZPoly$,
-relies on the construction of a canonical object called the \emph{residual
-transducer}, the latter being essentially based on differences between
-functions, crucially leveraging negative outputs \cite{CDTL23}. Although the
-proof method does not carry from $\Rel$-output functions to $\Nat$-output,
-functions, it was conjectured that the semantic property of being
-\kl{ultimately polynomial} would also characterize $\NSF$ inside $\NPoly$ (\cref{zsf-nsf:conjecture}).
-
-\begin{conjecture}[{\cite[Conjecture 7.61]{DOUE23}}]
-    \label{zsf-nsf:conjecture}
-    Let $k \in \Nat$.
-    A function $f \in \NPoly[k]$
-    belongs to $\NSF[k]$ if and only if
-    it is \kl{ultimately polynomial}.
-    In particular,
-    $\NSF[k] = \ZSF[k] \cap \NPoly$.
-\end{conjecture}
-
-We answer
-\cref{zsf-nsf:conjecture} 
-positively in the commutative case, by
-leveraging the semantic characterizations respectively of $\ZSF$ inside
-$\ZPoly$ (\kl{ultimately polynomial}) and $\NPoly$ inside $\ZPoly$
-(\kl{$(k,\Nat)$-combinatorial}), which is possible thanks to the decomposition
-obtained in \cref{decompose-polynomial:lem}. Decidability then follows
-immediately from the effective conversions, and the previous decidability
-result for $\ZSF$.
+%\begin{example}[restate=aperidic-ultimately-polynomial:ex,label=aperidic-ultimately-polynomial:ex]
+%    \proofref{aperidic-ultimately-polynomial:ex}
+%    A language $L$ is \kl(language){aperiodic} if and only if 
+%    $\ind{L}$ is \kl{ultimately polynomial}.
+%\end{example}
+%\begin{proof}
+%    Assume that $\ind{L}$ is \kl{ultimately polynomial}. Let $u,v,w \in
+%    \Sigma^*$. There exists a polynomial $P \in \Rat[X]$ and a number $N_0 \in
+%    \Nat$, such that $\ind{L}(u v^n w) = P(n)$ for all $n \geq N_0$. Because
+%    $\ind{L}$ has co-domain included in $\set{0,1}$, so does $P$, which proves
+%    that $P$ is a constant polynomial. Then, $\ind{L}(u v^{n+1} w) = P(n+1) =
+%    P(n) = \ind{L}(uv^n w)$ for all $n \geq N_0$.
+%
+%    Conversely, assume that $L$ is \kl(language){aperiodic}, let $n \in \Nat$ be a number, and
+%    let $\alpha_0$, $w_1$, $\alpha_1$, $\ldots$, $\alpha_{n-1}$, $w_n$, $\alpha_n$
+%    be words in $\Sigma^*$. Because $L$ is \kl(language){aperiodic}, the language
+%    $L$ is recognized by an \kl(monoid){aperiodic} finite monoid $M$, via a
+%    morphism $\mu \colon \Sigma^* \to M$ and an accepting part $F \subseteq M$.
+%    Because $M$ is a \kl(monoid){aperiodic}, there exists $N_0$
+%    such that for all $m \in M$, $m^{N_0+1} = m^{N_0}$. As a consequence,
+%    for all $1 \leq i \leq n$,
+%    whenever $x_i \geq N_0$, we have $\mu(w_i)^{x_i} = \mu(w_i)^{N_0}$.
+%    Therefore, for all $x_1, \dots, x_n \geq N_0$, we have:
+%    \begin{align*}
+%        \ind{L}\left(\alpha_0 \prod_{i= 1}^n w_i^{x_i} \alpha_i \right) 
+%        &= 
+%        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{x_i} \cdot \mu(\alpha_i)\right)
+%        & \text{ because $M$ recognizes $L$}
+%        \\
+%        &= 
+%        \ind{P}\left(\mu(\alpha_0) \cdot \prod_{i= 1}^n \mu(w_i)^{N_0} \cdot \mu(\alpha_i)\right)
+%        & \text{ by aperiodicity }
+%        \\
+%        &\defined P(x_1, \dots, x_n) & \text{ constant polynomial }
+%    \end{align*}
+%    Hence, $\ind{L}$ is \kl{ultimately polynomial}.
+%\end{proof}
 

### 2024-09-13 parts/appendix.tex
Hilbert!

@@
     if $P^2 - 1$ is \kl{non-negative}.
+
+    Let us prove that checking whether $P \in \CoveredPoly$ is undecidable. Let
+    $P \in \Rel[\vec{X}]$ be a polynomial, and let us write $P = P_+ - P_-$
+    where both $P_+$ and $P_-$ belong to $\Nat[\vec{X}]$. Let $Z$ be a fresh
+    indeterminate. We define $Q \defined P_+ (Z - 1) P_-$. By construction, $Q$
+    has \kl{non-negative} \kl{maximal monomials}. Furthermore, $Q$ itself is
+    \kl{non-negative} if and only if $\restr{Q}{Z = 0} = P$ is \kl{non-negative},
+    since for every $n > 0$, $\restr{Q}{Z = n} \in \Nat[\vec{X}]$ is
+    \kl{non-negative}. We conclude that 
+    $P$ is \kl{non-negative} if and only if $Q$ is \kl{non-negative},
+    if and only if $Q \in \CoveredPoly$.
 \end{proofof}

### 2024-09-13 parts/correction.tex
Hilbert!

@@
 fact that is not possible to decide whether a polynomial $P \in \Rel[\vec{X}]$
-is \kl{non-negative} (\cref{undecidable-non-negative:lem}), by reduction to the
-undecidability of Hilbert's Tenth Problem \cite{HILB1902,MATI1970}. That is,
-$\CorrectPoly$ is a decidable class that interpolates between the undecidable
-class of non-negative polynomials and strictly generalize polynomials in
-$\Nat[\vec{X}]$.
+is \kl{non-negative} or if a polynomial $P$ belongs to $\CoveredPoly$
+(\cref{undecidable-non-negative:lem}), by reduction to the undecidability of
+Hilbert's Tenth Problem \cite{HILB1902,MATI1970}. That is, $\CorrectPoly$ is a
+decidable class that strictly contains $\Nat[\vec{X}]$, and is contained in the
+undecidable classes $\CoveredPoly$ and the class of \kl{non-negative}
+polynomials.
 
@@
     $P \in \Rel[\vec{X}]$ is \kl{non-negative} is undecidable.
+    Similarly, checking whether a polynomial $P \in \Rel[\vec{X}]$
+    belongs to $\CoveredPoly$ is undecidable.
 \end{remark}
@@
 
+We conclude from \cref{lem:correct-covered-2} that $\CoveredPoly[X,Y]$ is a
+decidable class of polynomials, hence that Hilbert's Tenth Problem is
+decidable for polynomials in two indeterminates with non-negative maximal
+monomials.
+
 \subsection{From $\Rel$ to $\Rat$}

### 2024-09-16 parts/conclusion.tex
typos

@@
 
-Let us conclude with a more general remark on the status of commutative input
-in the study of unary output polyregular functions. Throughout the paper, we
-have tried to highlight that results on polyregular functions often have the
-following form: a polyregular function $f \colon \Sigma^* \to \Gamma^*$
-satisfies property $X$ if and only if for all \kl{commutative} \kl{star-free
-polyregular function} $g \colon \Delta^* \to \Sigma^*$, $f \circ g$ satisfies
-$X$. Let us formally restate this for two earlier results in the case of
-\kl{$\Rel$-polyregular functions}
-in \cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
+Let us end on a more general discussion regarding the status of commutative
+input functions in the study of unary output polyregular functions. Throughout
+the paper, we have tried to highlight that results on polyregular functions
+often have the following form: a polyregular function $f \colon \Sigma^* \to
+\Gamma^*$ satisfies property $X$ if and only if for all \kl{commutative}
+\kl{star-free polyregular function} $g \colon \Delta^* \to \Sigma^*$, $f \circ
+g$ satisfies $X$. Let us formally restate this for two earlier results in the
+case of \kl{$\Rel$-polyregular functions} in
+\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
 

### 2024-09-16 parts/correction.tex
typos

@@
 
-Let us briefly remark that \cref{corrected-version:thm} is surprising given the
+\Cref{corrected-version:thm} is surprising given the
 fact that is not possible to decide whether a polynomial $P \in \Rel[\vec{X}]$
@@
 to compute polynomials in $\CorrectPoly$ using $\NPoly$
-(\cref{lem:correct-to-n-rat}). Then, we will next generalize
+(\cref{lem:correct-to-n-rat}). Finally, we will next generalize
 \cref{corrected-version:thm} to polynomials in $\Rat[\vec{X}]$ in
@@
 upcoming \cref{k-combinatorial:def,ultimately-polynomial:def}
-in
+of
 \cref{beyond-polynomials:sec}. Our second goal
@@
 
+Conversely, when suitably pumping a \kl{$\Nat$-polyregular function}, one
+always obtains \kl{natural binomial functions}.
+
 \begin{lemma}[restate=n-poly-combinatorics:lem,label=n-poly-combinatorics:lem]
@@
 
-Let us highlight that the multiplicative factor $\omega$ is necessary in
+The multiplicative factor $\omega$ is necessary in
 \cref{n-poly-combinatorics:lem}. Indeed, the function $f \colon \set{a}^* \to
@@
 	we conclude.
-
 	Assume now that $P$ is not a constant polynomial. We will use
@@
 	Therefore,
-	we conclude that
 	$Q \in \Nat[\vec{X}]$.
@@
 		monomials} of $P$. We compute a bound $K$ such that $Q \defined
-		(\Diff{K}{P_1} + \translate{K}(P_2)) \in \Nat[\vec{X}]$ using
-	\cref{derivation-stabilises-correct:lem}. In particular, $Q$ is
+		(\Diff{K}{P_1} + \translate{K}(P_2)) \in \Nat[\vec{X}]$ 
+        (\cref{derivation-stabilises-correct:lem}). In particular, $Q$ is
 	\kl{represented} by a \kl{star-free $\Nat$-polyregular function} using
@@
 
-We conclude from \cref{lem:correct-covered-2} that $\CoveredPoly[X,Y]$ is a
-decidable class of polynomials, hence that Hilbert's Tenth Problem is
-decidable for polynomials in two indeterminates with non-negative maximal
-monomials.
-
 \subsection{From $\Rel$ to $\Rat$}
@@
 justify that all the combinatorial work has already happened in $\Rel$ and
-$\Nat$. A first observation is that from \cref{n-sf-combinatorics:lem}, we know
+$\Nat$. From \cref{n-sf-combinatorics:lem}, we know
 that the polynomials that can be computed by \kl{star-free $\Nat$-polyregular
@@
 		+ 8 \pbinom{Y}{2} + 8 \pbinom{Z}{2} + 4$
-	is a \kl{non-negative} \kl{natural binomial polynomial},
+    is a \kl{non-negative} \kl{natural binomial polynomial} in $\Rel[X,Y,Z]$,
 	but cannot be computed by a \kl{star-free $\Nat$-polyregular function}.

### 2024-09-16 parts/prelim.tex
typos

@@
 
-\subsection{Polynomials} \AP All polynomials considered in this paper have
-coefficients in $\Rel$ unless explicitly stated otherwise. A polynomial $P \in
+\subsection{Polynomials} \AP A polynomial $P \in
 \Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for all non-negative
@@
 
-\AP Let us now give some vocabulary on polynomials with multiple indeterminates
-over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.
+\AP A \intro{monomial} is a product of indeterminates and integers.
 For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
@@
 into a sum of \kl{monomials}.
-
-\AP A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
+A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
 $T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
@@
 our analysis on polynomials. This description is very similar in shape to the
-\emph{finite counting automata} introduced by \cite{SCHU62}. For the sake of
-simplicity, we only consider $\Rel$-valued (resp. $\Nat$-valued) functions, but
-could consider other commutative semigroups.
+\emph{finite counting automata} introduced by \cite{SCHU62}.
 
@@
 \cref{size-of-word-nsf:ex}. Let us also introduce in
-\cref{q-polynomial-n-poly:ex} a function that will appear frequently in this
-paper as an example of \emph{division} computed by a \kl{$\Nat$-polyregular
-function}.
+\cref{q-polynomial-n-poly:ex} a function that serves as an example of
+\emph{division} computed by a \kl{$\Nat$-polyregular function}.
 

### 2024-11-26 reviews/rebuttal-stacs.md
update rebuttal

---
title: "Rebuttal to Reviewer Comments for STACS 2022"
author: "Aliaume LOPEZ"
date: "2024-11-18"
---

\newcommand{\ZPoly}{\mathbb{Z}\mathsf{Poly}}
\newcommand{\NPoly}{\mathbb{N}\mathsf{Poly}}

We would like to thank the reviewers for their insightful comments and feedback.
Due to the space limitation in this rebuttal, we will focus on the critical
points raised by Reviewer 2, and then address the general comments of the other
reviewers.

Theorem 37. Let us first address the main concern of Reviewer 2. It is
decidable whether a commutative function $f$ in $\ZPoly_k$ is in $\NPoly_k$ as
follows: one uses Lemma 36 to obtain a family of polynomials $(P_t)$ that can
be used to compute $f$ on any entry. Then, we check if all these polynomials
are in $\NPoly$, which is possible using Theorem 17. If all of them are, one
can compute $f$ as in the proof of Theorem 37: by computing the type of the
input, and then applying the corresponding polynomial. Conversely, assume that
$f \in \NPoly$, then all of these polynomials are in $\NPoly$, since they are
computed by precomposition of $f$ by a suitable pumping pattern. This also
provides an effective conversion procedure.

As noted by Reviewer 2, this argument requires extra closure properties than
those listed in Lemma 7, namely closure under precomposition by a regular
function to compute the "type of the input." This holds for $\NPoly$ since it
is true for any polyregular function [2, Lemma 8.1].

Example 5. It has been remarked by R1 and R2 that Example 5 contains an
*off-by-one* error, which can be fixed by considering the sum of *non-empty*
prefixes. This is still possible with a very slight modification of the
example, letting $\pi(x,y,z) = x \times y$. Thank you for pointing this out.
This was also a problem in Lemma 19, with the same fix: restricting the sum to
non-empty words.

Lemma 8. The fact that $f \circ c$ is (effectively) a rational series follows
from the more general fact that Weighted Automata are closed under
precomposition with a regular function, but I could not find a clear reference
while writing this rebuttal. For $\mathbb{Z}\mathsf{Poly}$ this is [5,
Proposition II.12]. However, for this specific case, it is relatively simple to
build a weighted automaton computing $f \circ c$ from the automaton
computing $f$. Which will be done in the final version of the paper.

Lemma 33. In the proof, the "polynomial function" $q$ is chosen to be the
simplest one $q(X_1, \dots, X_k) = a_1^{X_1} a_2^{X_2} \dots a_k^{X_k}$. That
way, $P = f \circ q$. Furthermore, in the statement, the intersection with
$\mathbb{Z}[\vec{X}]$ is not useful, and in its proof, $f = g$, and $\lambda =
n$ as guessed by R2.

# Notations

A number of comments of the reviewers pointed out that the notation could be
improved and made more consistent. In particular, the "polynomial functions"
will be renamed "pumping patterns", and "binomial functions" will be renamed
"binomial coefficients". Finally, the class $\mathsf{PolyRec}$ will be renamed
$\mathsf{PolyStrongNNeg}$ to avoid confusion with the class of polynomial
recursive sequences.

# General Comments

*R1. l.65: it could be interested to comment on what else is broken as a
corollary in the article of Karhumaki.*

It turns out that all examples in the article of Karhumaki are correct because
they use at most 2 distinct indeterminates. This is stated in the paragraph
above Lemma 29. We are unaware of results that are broken by this error.

*R3. What do we learn about N/Z-weighted languages overall, and are these
results somehow meaningful in potential application areas of such languages?*

The techniques developed in this paper are tailored to *polynomials* and not
easily generalisable *as is*. However, it designates the problem of deciding
$\mathbb{N}$-definability of $\mathbb{Z}$-rational series in the commutative
case (without polynomial growth assumption) as a natural next step towards
solving Problem 1. Furthermore, this paper demonstrates that understanding the
precise relationship between $\NPoly$ and $\ZPoly$ actually leads to decision
procedures for Problem 2 in the unary case, which was unclear prior to this
work. Finally, the proof techniques being quite different from the usual ones,
it could foster new ideas in the community, for instance regarding *aperiodic*
$\mathbb{N}$/$\mathbb{Z}$-rational series.

*R3. I guess it is undecidable whether a Z-rational series is N-valued?*. Yes,
this is not possible.

The Remark 18 states that deciding if a Z-polynomial $\mathbb{N}$-valued is
undecidable, and a fortiori, it is not possible for $\mathbb{Z}$-rational series.

### 2024-11-26 globals/metadata.tex
update rebuttal

@@
 
-\title{Commutative $\Nat$-polyregular functions}
+\title{Commutative $\Nat$-rational series of polynomial growth}
 
@@
 commutativity (the order of letters in the input does not matter) and
-polynomial growth (the ouptut of the function is bounded by a polynomial in the
+polynomial growth (the output of the function is bounded by a polynomial in the
 size of the input). We

### 2024-11-26 parts/newintro.tex
update rebuttal

%! TeX program = xelatex
%! lang = en-US
% MAIN MATTER
\section{Introduction}
\label{introduction:sec}

\AP Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class of
\kl{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
$\Sigma^*$ to $\mathbb{S}$ that are computed by \kl{$\mathbb{S}$-weighted
automata} \cite{BERE10}. This computational model is a generalization of the
classical notion of non-deterministic finite automata to the weighted setting,
where transitions are labeled with elements of $\mathbb{S}$. The semantics of
\kl{$\mathbb{S}$-weighted automata} on a given word $w$ is defined by the sum
over all accepting runs reading $w$, of the product of the weights of the
transitions taken along this run. In this paper, we are interested in the case
where $\mathbb{S}$ equals $\Nat$ or $\Rel$, hence, in $\Nat$-rational series
($\intro*\NRat$) and $\Rel$-rational series ($\intro*\ZRat$). It is clear that
$\NRat$ is a proper subclass of $\ZRat$, and a longstanding open problem is to
provide an algorithm that decides whether a given $\ZRat$ is in $\NRat$
\cite{KARH77}.

\begin{problem}
    \label{n-in-z-rat:problem}
    Input: A $\ZRat$ $f$. Output: Is $f$ in $\NRat$?
\end{problem}

\Cref{n-in-z-rat:problem} recently received attention in the context of
\kl{polyregular functions}, a computational model that aims to
generalize the theory of regular languages to the setting of string-to-string
functions \cite{BOJA18}. In the case of regular languages, \emph{star-free
languages} form a robust subclass of regular languages described equivalently
in terms of first order logic \cite{MNPA71}, counter-free automata
\cite{MNPA71}, or aperiodic monoids \cite{SCHU65}. Analogously, there exists a
\emph{star-free} fragment of \kl{polyregular functions} called \kl{star-free
polyregular functions} \cite{BOJA18}. One open question in this area is to
decide whether a given \kl{polyregular function} is \kl{star-free}.

\begin{problem}
    \label{sf-polyregular:problem}
    Input: A \kl{polyregular function} $f$. Output: Is $f$ \kl{star-free}?
\end{problem}

\AP In order to approach decision problems on \kl{polyregular functions},
restricting the output alphabet to a single letter has proven to be a fruitful
method \cite{DOUE21,DOUE22}. Because words over a unary alphabet are
canonically identified with natural numbers, unary output \kl{polyregular
functions} are often called \kl{$\Nat$-polyregular functions} ($\NPoly$), and
their \emph{star-free} counterpart \kl{star-free $\Nat$-polyregular functions}
($\NSF$). Coincidentally, \kl{polyregular functions} with unary output form
subclass of \kl{$\Nat$-rational series}, namely the classf of
\kl{$\Nat$-rational series} of \emph{polynomial growth}, i.e. the output of the
function is bounded by a polynomial in the size of the input. In \cite{CDTL23},
the authors introduced the class of \kl{$\Rel$-polyregular functions}
($\ZPoly$) as a subclass of \kl{$\Rel$-rational series} that generalizes
\kl{$\Nat$-polyregular functions} by allowing negative outputs, and showed that
membership in the \emph{star-free} subclass $\ZSF$ inside $\ZPoly$ is decidable
\cite[Theorem V.8]{CDTL23}. Although this could not be immediately leveraged to
decide $\NSF$ inside $\NPoly$, it was conjectured that $\NPoly \cap \ZSF =
\NSF$ \cite[Conjecture 7.61]{DOUE23}. It was believed that understanding the
membership problem of $\NPoly$ inside $\ZPoly$, that is, a restricted version
of \cref{n-in-z-rat:problem}, would be a key step towards proving $\NPoly \cap
\ZSF = \NSF$, which itself would give hope in designing an algorithm for
\cref{sf-polyregular:problem}. We illustrate in
\cref{previously-known-inclusions:fig} the known inclusions and related open
problems between the discussed classes of functions.

\begin{figure}
    \centering
    \includestandalone[height=5cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Rel$, $\Nat$, $\Sigma$). The second
        one is the allowed computational power
        (\kl{star-free polyregular functions}, \kl{polyregular functions}, 
        \kl{rational series}).
        Arrows denote strict inclusions,
        and effectiveness (both in terms of decidability and of effective
        representation) is represented by thick double arrows. Inclusions that are
        suspected to be effective are represented using a dashed arrow together with a
        question mark.
    }
    \label{previously-known-inclusions:fig}
\end{figure}

\subparagraph*{Contributions.} We prove that $\NSF = \NPoly \cap \ZSF$
\cite[Conjecture 7.61]{DOUE23} and design an algorithm that decides whether a
function in $\ZPoly$ is in $\NPoly$ \cite[Open question 5.55]{DOUE23}, both
under the extra assumption of \emph{commutativity}, that is, assuming that the
function is invariant under the permutation of its input. As a consequence, the
upper left square of \cref{previously-known-inclusions:fig} has all of its
arrows decidable and with effective conversion procedures under this extra
assumption. Because \kl{$\Rel$-rational series} with \emph{polynomial growth}
are exactly \kl{$\Rel$-polyregular functions} \cite{CDTL23}, this can be seen
as decision procedure for \cref{n-in-z-rat:problem} under the extra assumption
of \emph{commutativity} and \emph{polynomial growth}. Similarly, our results
provide an algorithm for \cref{sf-polyregular:problem} under the extra
assumption of \emph{commutativity} and \emph{unary output alphabet}.

As an intermediate step, we provide a complete and decidable characterization
of polynomials in $\Rat[\vec{X}]$ that can be computed using $\NRat$ (resp.
$\ZRat)$. These characterizations uncover a fatal flaw in the proof of a former
characterization of such polynomials \cite[Theorem 3.3, page 4]{KARH77}.
Furthermore, these characterizations provide effective descriptions of
polynomials that can be expressed in $\ZRat$ as those obtained using integer
combinations of products of \emph{binomial coefficients} (called \kl{integer
binomial polynomials}, defined page \kpageref{integer binomial polynomial}) and
similarly for $\NRat$ by introducing the notion of \kl{strongly natural
binomial polynomials} (defined page \kpageref{strongly natural binomial
polynomial}), which we believe has its own interest. Finally, these
characterizations demonstrate that polynomials expressible by $\ZRat$ (resp.
$\NRat$) are exactly those expressible by $\ZSF$ (resp. $\NSF$), that is,
polynomials are inherently \emph{star free} functions.

\subparagraph*{Outline of the paper.} In \cref{preliminaries:sec}, we provide a
combinatorial definition of \kl{$\Nat$-polyregular functions} (resp.
\kl{$\Rel$-polyregular functions}), show that one can decide if a function $f
\in \ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}). In
\cref{polynomials:sec}, we provide a counterexample to the flawed result of
\cite[Theorem 3.3, page 4]{KARH77} (\cref{thm:counter-example}), and correct it
by providing effective characterizations of polynomials computed by $\ZRat$
(\cref{integer-binomial-polynomials:cor}) and $\NRat$
(\cref{decide-rat-poly-npoly:cor}). Finally, in \cref{beyond-polynomials:sec},
answer positively to \cite[Open question 5.55]{DOUE23}
(\cref{decidable-n-poly:thm}) and to \cite[Conjecture 7.61]{DOUE23}
(\cref{zsf-npoly-nsf:thm}), both under the extra assumption of
\emph{commutativity}.

### 2024-11-27 parts/correction.tex
typo hunt

@@
 \Cref{corrected-version:thm} is surprising given the
-fact that is not possible to decide whether a polynomial $P \in \Rel[\vec{X}]$
+fact that it is not possible to decide whether a polynomial $P \in \Rel[\vec{X}]$
 is \kl{non-negative} or if a polynomial $P$ belongs to $\CoveredPoly$
@@
 \AP To address our first goal, let us define that a function $q$ is a
-\intro{(noncommutative output) polynomial function} from $\Nat^p$ to $\Sigma^*$
+\intro{pumping pattern} from $\Nat^p$ to $\Sigma^*$
 whenever there exists words $\alpha_0, \dots, \alpha_p \in \Sigma^*$, and words
@@
 polynomial $X(X-1)/2 = \binom{X}{2}$
-of \cref{q-polynomial-n-poly:ex}. A \intro{binomial function} is a
+of \cref{q-polynomial-n-poly:ex}. A \intro{simple binomial function} is a
 function of the form $\binom{X - \ell}{k}$, where $\ell$ and $k$ are natural
 numbers. We extend this to \intro{natural binomial functions} that are obtained
-by considering $\Nat$-linear combinations of products of \kl{binomial
+by considering $\Nat$-linear combinations of products of \kl{simple binomial
 functions}, that is, we consider functions that have the following shape:
@@
 $X \leq \ell$, and is therefore not a polynomial. Let us immediately
-prove that \kl{binomial functions}  can be \kl{represented} in $\NSF$,
+prove that \kl{simple binomial functions}  can be \kl{represented} in $\NSF$,
 generalizing \cref{q-polynomial-n-poly:ex}.
@@
 	can be obtained by translating $f_{r,s}$.
-	Remark that
-	$\binom{x}{s} = \sum_{u_1 \cdots u_{s+1} = a^x} 1$,
-	which is clearly expressible as a \kl{star-free polyregular function}.
+    By definition, $\binom{x}{s} = \card{P_{x,s}}$,
+    where $P_{x,s} \defined \setof{ (x_1, \dots, x_s) \in \Nat^s }{ 1 \leq x_1 < \cdots < x_s \leq x }$.
+    Let us proceed as in \cref{q-polynomial-n-poly:ex} and define 
+    $D_{w,s} \defined \setof{ (u_1, \dots, u_s, u_{s+1}) \in (\Sigma^+)^s \times \Sigma^* }{ w = u_1 u_2 \cdots u_s u_{s+1} }$.
+    It is clear that $D_{a^x,s}$ is in bijection with $P_{x,s}$ for all $x \in \Nat$
+    using the map $(x_1, \dots, x_s) \mapsto (a^{x_1}, \dots, a^{x_s})$.
+    Now, using the monoid $M \defined (\set{0,1}, \max)$
+    and the morphism $\mu(\varepsilon) \defined 0$ and $\mu(a) \defined 1$,
+    one can compute $\card{D_{w,s}}$ 
+    as $\pi^\dagger (w)$ where $\pi \colon M^{s+1} \to \Nat$ is defined by
+    $\pi(m_1, \dots, m_s, m_{s+1}) \defined m_1 \times \cdots \times m_s$.
+    We conclude that $f_{0,s} \in \NSF[s]$ is a \kl{star-free polyregular function}.
 \end{proof}
@@
 	There exists a computable $\omega \in \Nat_{\geq 1}$
-	such that for all \kl{polynomial functions}
+	such that for all \kl{pumping patterns}
 	$q \colon \Nat^p \to \Sigma^*$,
@@
 	There exists a computable $s \in \Nat_{\geq 1}$
-	such that for all \kl{polynomial functions}
+	such that for all \kl{pumping patterns}
 	$q \colon \Nat^p \to \Sigma^*$,
@@
     For large enough values of $X$, the \kl{binomial function}
-	$\binom{X - p}{k}$ coincides with a polynomial whose dominant coefficient
+	$\binom{X - p}{k}$ coincides with a polynomial whose leading coefficient
 	is $1/k!$ which is \kl{non-negative}.
@@
 	by a \kl{$\Nat$-polyregular function}, namely
-	$f$ partially applied to a word $u$ and using a restricted input alphabet. As a consequence,
+	$f_u \colon w \mapsto f(uw)$ 
+    where $w$ belongs to a restricted alphabet.
+    As a consequence,
 	the \kl{maximal monomials} of
@@
 	In this case, $P_1 = P$ and $P_2 = 0$.
-	Furthermore,
-	$\translate{K}(P_1) = P_1$ and $\Diff{K}{P_1} = 0$ for all $K \in \Nat$.
+	Furthermore, $\Diff{K}{P_1} = 0$ for all $K \in \Nat$.
 	We conclude that $\Diff{K}{P_1} + \translate{K}(P_2) = 0$
@@
 	Remark that $\Diff{K}{P_1}$ is already in $\Nat[\vec{X}]$,
-	and the constant term of $\translate{K}(P_2)$ is also
+	and the constant coefficient of $\translate{K}(P_2)$ is also
 	in $\Nat$.
@@
 	it strictly divides some monomial of $P_1$, and
-	equals some monomial of $\Diff{K}{P_1}$. Because every monomial
+	equals some monomial of $\Diff{K}{P_1}$ up to a multiplication by a factor in $\Rat$. Because every monomial
 	of $\Diff{K}{P_1}$ has a coefficient that is a multiple of $K = \alpha \times D$, we can
@@
 	Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial,
-	and therefore must have a positive leading coefficient
+	and therefore must either have a positive leading coefficient
 	(which is the unique \kl{maximal monomial} in this case)
@@
 These definitions are justified by the classical result of Pólya that
-characterizes polynomials $P$ in $\Rat[X]$ that \intro{integer-valued} (i.e., are such
+characterizes polynomials $P$ in $\Rat[X]$ that are \intro{integer-valued} (i.e., are such
 that $P(\Rel) \subseteq \Rel$) as \kl{integer binomial polynomials}
-\cite{POLYA1915,CACHA1996}, which are often called \emph{Newton Polynomials}. Note
+\cite{POLYA1915,CACHA1996}. Note
 that this result straightforwardly extends to multiple indeterminates as we
@@
 	\label{integer-binomial-polynomials:cor}
-	Let $P \in \Rat[\vec{X}]$, and let $\alpha$ be the minimal number in $\Nat$ such that
-	$\alpha P \in \Rel[\vec{X}]$. Then, the following properties are equivalent:
+	Let $P \in \Rat[\vec{X}]$. Then, the following properties are equivalent:
 	\begin{enumerate}
@@
     and $n \in \Nat_{\geq 1}$ be
-    such that $n P \in \Rel[\vec{X}] \cap \CorrectPoly$.
+    such that $n P \in \CorrectPoly$.
     Then, $P$ is a \kl{strongly natural binomial polynomial}.
@@
     \cref{rat-npoly-4:item}, we obtain $\alpha P \in \CorrectPoly$ via
-    \cref{corrected-version:thm} by remarking that the \kl{star-free
-    $\Nat$-polyregular functions} have output in $\Nat$ and are closed under
+    \cref{corrected-version:thm} by remarking that \kl{$\Nat$-polyregular functions} have output in $\Nat$ and are closed under
     multiplication by a constant $\alpha \in \Nat$. The fact that $P$ is

### 2024-11-27 parts/counter.tex
typos
typo hunt

@@
 \AP In this section, we will completely characterize which polynomials in
-$\Rat[\vec{X}]$ that are \kl{represented} by \kl{$\Nat$-rational series} (resp.
+$\Rat[\vec{X}]$ are \kl{represented} by \kl{$\Nat$-rational series} (resp.
 \kl{$\Rel$-rational series}). To that end, we start by characterizing these
@@
 
-Before giving a counterexample to the above statement, let us first provide via
-\cref{non-neg-not-nrat:ex} simple proof that there exists some
+Before giving a counterexample to the above statement, let us first exhibit in
+\cref{non-neg-not-nrat:ex} some
 \kl{non-negative} polynomial that is not an \kl{$\Nat$-rational polynomial}.
@@
 polynomials}. In order to derive this example, we will need the following
-fundamental
-result about the pre-image of regular languages by \kl{polyregular functions}.\footnote{
-    In this particular case, one could have
-    considered more generally \kl{$\Nat$-rational series},
-    and replaced regular languages over a unary alphabet
-    by semi-linear sets.
-}
+fundamental result about the pre-image of regular languages by \kl{polyregular
+functions}.\footnote{ In this particular case, one could have considered more
+generally \kl{$\Nat$-rational series}, and replaced regular languages over a
+unary alphabet by semi-linear sets. } Before that, let us remark that if a
+polynomial $P$ is \kl{represented} by a \kl{$\Nat$-rational series}, then it is
+in fact \kl{represented} by a \kl{$\Nat$-polyregular function} thanks to
+\cref{polyregular-polynomial-growth:lemma}.
 
@@
     Indeed, assume by contradiction that
-    $f \in \NPoly$ \kl{represents} $f$ over the alphabet $\Sigma \defined \set{a,b}$.
+    $f \in \NPoly$ \kl{represents} $P$ over the alphabet $\Sigma \defined \set{a,b}$.
     Then, $f^{-1}(\set{0})$ is a regular language
@@
 \end{example}
+
+Please note that the same argument cannot be leveraged for proving that $P$ is
+not a \kl{$\Rel$-rational polynomial}: \cref{pre-image-regular:fact}
+only holds for
+\emph{string-to-string} functions, and is applied to the specific case where
+the output alphabet is $\set{1}$, i.e., where the output of the function
+belongs to $\set{1}^*$ which is isomorphic to $\Nat$.
 
@@
     if and only if $z(x+y)^2 = -2 (x-y)^2$. Hence,
-    $\BadPoly(x,y,z)=0$ if and only if $z = 0$ and $x = Y$, or 
+    $\BadPoly(x,y,z)=0$ if and only if $z = 0$ and $x = y$, or 
     $z \neq 0$, and $x = y = 0$.
@@
     \card[a]{w} = \card[b]{w} } \cup \set{ c }^*$.
-    Because $L \cap \set{a,b}^*$ not a regular language,
+    Because $L \cap \set{a,b}^*$ is not a regular language,
     we

### 2024-11-27 parts/newintro.tex
typo hunt

@@
 their \emph{star-free} counterpart \kl{star-free $\Nat$-polyregular functions}
-($\NSF$). Coincidentally, \kl{polyregular functions} with unary output form
-subclass of \kl{$\Nat$-rational series}, namely the classf of
+($\NSF$). Coincidentally, \kl{polyregular functions} with unary output forms a
+subclass of \kl{$\Nat$-rational series}, namely the class of
 \kl{$\Nat$-rational series} of \emph{polynomial growth}, i.e. the output of the
@@
 
-\subparagraph*{Contributions.} We prove that $\NSF = \NPoly \cap \ZSF$
+\subparagraph*{Contributions.} In this paper, we work under the extra
+assumption of \emph{commutativity}, that is, assuming that the function is
+invariant under the permutation of its input. In this setting, we 
+prove that $\NPoly \cap \ZSF = \NSF$
 \cite[Conjecture 7.61]{DOUE23} and design an algorithm that decides whether a
-function in $\ZPoly$ is in $\NPoly$ \cite[Open question 5.55]{DOUE23}, both
-under the extra assumption of \emph{commutativity}, that is, assuming that the
-function is invariant under the permutation of its input. As a consequence, the
+function in $\ZPoly$ is in $\NPoly$ \cite[Open question 5.55]{DOUE23}.
+As a consequence, the
 upper left square of \cref{previously-known-inclusions:fig} has all of its
@@
 $\ZRat)$. These characterizations uncover a fatal flaw in the proof of a former
-characterization of such polynomials \cite[Theorem 3.3, page 4]{KARH77}.
-Furthermore, these characterizations provide effective descriptions of
-polynomials that can be expressed in $\ZRat$ as those obtained using integer
-combinations of products of \emph{binomial coefficients} (called \kl{integer
-binomial polynomials}, defined page \kpageref{integer binomial polynomial}) and
-similarly for $\NRat$ by introducing the notion of \kl{strongly natural
-binomial polynomials} (defined page \kpageref{strongly natural binomial
-polynomial}), which we believe has its own interest. Finally, these
-characterizations demonstrate that polynomials expressible by $\ZRat$ (resp.
-$\NRat$) are exactly those expressible by $\ZSF$ (resp. $\NSF$), that is,
-polynomials are inherently \emph{star free} functions.
+characterization of such polynomials \cite[Theorem 3.3, page 4]{KARH77}. We
+also prove that this previous results holds for polynomials with at most two
+indeterminates (\cref{lem:correct-covered-2}), which may explain why
+it was not detected earlier. Furthermore, these characterizations provide
+effective descriptions of polynomials that can be expressed in $\ZRat$ as those
+obtained using integer combinations of products of \emph{binomial coefficients}
+(called \kl{integer binomial polynomials}, defined page \kpageref{integer
+binomial polynomial}) and similarly for $\NRat$ by introducing the notion of
+\kl{strongly natural binomial polynomials} (defined page \kpageref{strongly
+natural binomial polynomial}), which we believe has its own interest. Finally,
+these characterizations demonstrate that polynomials expressible by $\ZRat$
+(resp. $\NRat$) are exactly those expressible by $\ZSF$ (resp. $\NSF$), that
+is, polynomials are inherently \emph{star free} functions.
 
@@
 (\cref{decide-rat-poly-npoly:cor}). Finally, in \cref{beyond-polynomials:sec},
-answer positively to \cite[Open question 5.55]{DOUE23}
+we answer positively to \cite[Open question 5.55]{DOUE23}
 (\cref{decidable-n-poly:thm}) and to \cite[Conjecture 7.61]{DOUE23}

### 2024-11-27 parts/prelim.tex
typo hunt

@@
 
-\AP Let $n \in \Nat$, and let $\Sigma$ be a finite alphabet. Given a function
-$\eta \colon \set{1, \dots, n} \to \Sigma$, we define the $\eta^\dagger \colon
-\Nat^n \to \Sigma^*$ as $\eta^\dagger(\vec{x}) \defined \eta(1)^{x_1} \dots
-\eta(n)^{x_n}$. A function $f \colon \Nat^n \to X$ is \intro{represented} by a
+\AP Let $k \in \Nat$, and let $\Sigma$ be a finite alphabet. Given a function
+$\eta \colon \set{1, \dots, k} \to \Sigma$, we define the $\eta^\dagger \colon
+\Nat^k \to \Sigma^*$ as $\eta^\dagger(\vec{x}) \defined \eta(1)^{x_1} \dots
+\eta(k)^{x_k}$. A function $f \colon \Nat^k \to X$ is \intro{represented} by a
 \kl{commutative} function $g \colon \Sigma^* \to X$ if there exists a map $\eta
-\colon \set{1, \dots, n} \to \Sigma$ such that $g \circ \eta^\dagger = f$. This
+\colon \set{1, \dots, k} \to \Sigma$ such that $g \circ \eta^\dagger = f$. This
 notion will be useful to formally state that a polynomial ``is'' a
@@
 
-\subsection{Polynomials} \AP A polynomial $P \in
-\Rel[X_1, \dots, X_k]$ is \intro{non-negative} when for all non-negative
-integer inputs $n_1, \dots, n_k \geq 0$, the output  $P(n_1, \dots, n_k)$ of
-the polynomial is non-negative. Beware that we do not consider negative values
-as input, as the numbers $n_i$ will ultimately count the number of occurrences
-of a letter in a word. As an example, the polynomial $(X - Y)^2$ is
-\kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
+\subsection{Polynomials} \AP A polynomial $P \in \Rel[X_1, \dots, X_k]$ is
+\intro{non-negative} when for all non-negative integer inputs $n_1, \dots, n_k
+\geq 0$, the output  $P(n_1, \dots, n_k)$ of the polynomial is non-negative. In
+the case of at most three indeterminates, we use variables $X,Y,Z$ instead of
+$X_1, X_2, X_3$ to lighten the notation. Beware that we do not consider
+negative values as input, as the numbers $n_i$ will ultimately count the number
+of occurrences of a letter in a word. As an example, the polynomial $(X - Y)^2$
+is \kl{non-negative}, and so is the polynomial $X^3$, but the polynomial $X^2 -
 2X$ is not.
 
-\AP A \intro{monomial} is a product of indeterminates and integers.
-For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
-\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
-\dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
-Every polynomial $P \in \Rel[X_1, \dots, X_n]$ decomposes uniquely
-into a sum of \kl{monomials}.
-A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
-$T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
+\AP A \intro{monomial} is a product of indeterminates and integers. For
+instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
+\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. Every polynomial $P \in
+\Rel[X_1, \dots, X_n]$ decomposes uniquely into a sum of \kl{monomials}. A
+\kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides $T$
+seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
 \kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
@@
 \intro{maximal monomial} if it is a maximal element for the \kl{divisibility
-ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
-+ X + Y$, the set of \reintro{maximal monomials}, written
-$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  Y^2}$.  In the polynomial
-$P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials} are $X^2$ and
-$Y^2$.
+preordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
++ X + Y$, the set of \reintro{maximal monomials} is $\set{X^2,  -2 XY,  Y^2}$.
+In the polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative}
+\kl{monomials} are $X^2$ and $Y^2$.
 
@@
 general definition of \kl{polyregular functions}. 
-We chose in this paper to provide a very combinatorial description of
+We chose in this paper to provide a combinatorial description of
 polyregular functions with commutative outputs because it will play nicely with
@@
     \label{nat-rel-poly:def}
-    Let $k \in \Nat$. The set $\intro*\ZPoly[k]$ of polyregular
-    \intro{$\Rel$-polyregular functions} of degree at most $k$,
+    Let $d \in \Nat$. The set $\intro*\ZPoly[d]$ of polyregular
+    \intro{$\Rel$-polyregular functions} of degree at most $d$,
     is the set of functions $f \colon \Sigma^* \to \Rel$
@@
     a morphism $\mu \colon \Sigma^* \to M$,
-    and a function $\pi \colon M^{k+1} \to \Rel$
+    and a function $\pi \colon M^{d+1} \to \Rel$
     satisfying for all $w \in \Sigma^*$:
@@
         f (w) = \pi^\dagger (w) \defined
-        \sum_{w = u_1 \cdots u_{k+1}} \pi(\mu(u_1), \dots, \mu(u_{k+1}))
+        \sum_{w = u_1 \cdots u_{d+1}} \pi(\mu(u_1), \dots, \mu(u_{d+1}))
         \quad .
@@
     If the function $\pi$ has codomain $\Nat$,
-    then $f$ is \intro{$\Nat$-polyregular} of degree at most $k$,
-    i.e., $f \in \intro*\NPoly[k]$.
+    then $f$ is \intro{$\Nat$-polyregular} of degree at most $d$,
+    i.e., $f \in \intro*\NPoly[d]$.
     If the monoid $M$ is \kl(monoid){aperiodic}
@@
     the function $f$ is \intro{star-free $\Rel$-polyregular}
-    ($\intro*\ZSF[k]$), resp. \intro{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
+    ($\intro*\ZSF[d]$), resp. \intro{star-free $\Nat$-polyregular} ($\intro*\NSF[d]$).
 \end{definition}
@@
 \AP We complete \cref{nat-rel-poly:def} by letting $\reintro*\NPoly \defined
-\bigcup_{k \in \Nat} \NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
+\bigcup_{d \in \Nat} \NPoly[d]$, and similarly for $\ZPoly$, $\NSF$, and
 $\ZSF$. In order to illustrate these definitions, let us provide an example of
-a \kl{$\Nat$-polyregular function} computed using a finite monoid in
+an \kl{$\Nat$-polyregular function} computed using a finite monoid in
 \cref{size-of-word-nsf:ex}. Let us also introduce in
@@
     Let $f \colon \Sigma^* \to \Nat$ that maps a word $w$
-    to the sum of the length of its prefixes,
-    i.e., $f(w) \defined \sum_{uv = w} \card{u}$.
-    The function $f$ is an \kl{$\Nat$-polyregular function}
-    and 
-    $f(w) = \card{w}( \card{w} - 1) / 2$.
+    to the number of distinct pairs of positions in $w$,
+    i.e., $f(w) = \binom{\card{w}}{2} = \card{w}(\card{w} - 1) / 2$.
+    Then, $f \in \NSF[2]$.
 \end{example}
 \begin{proof}
+    Let us remark that the set $P_w$ of distinct pairs of positions
+    $i < j$ in a word $w$ is in bijection with the set $D_w$ of
+    decompositons of the form $w = xyz$,
+    where $x$ and $y$ are non-empty, 
+    via the map $(i,j) \mapsto (w_{1,i}, w_{i+1,j}, w_{j+1,\card{w}})$.
     Let us write $M \defined (\set{0,1}, \max)$
     which is a finite aperiodic monoid, and $\mu \colon \Sigma^* \to M$
-    that maps $\varepsilon$ to $0$ and the other words to $1$.
+    that maps the empty word $\varepsilon$ to $0$ and the other words to $1$.
     Then, let us define $\pi \colon M^3 \to \Nat$
-    via $\pi(x,y,z) = y$.
-    Remark that 
+    via $\pi(x,y,z) = x \times y$.
+    We conclude because:
     \begin{align*}
@@
         = 
-        \sum_{uz = w} \sum_{xy = u} \pi(\mu(x), \mu(y), \mu(z)) 
+        \sum_{xyz = w \wedge x \neq \varepsilon \wedge y \neq \varepsilon} 1
         = 
-        \sum_{uz = w} \sum_{xy = u} \mu(y) \\
-        &= 
-        \sum_{uz = w} \card{u} 
+        \card{D_w}
+        =
+        \card{P_w}
         = f(w) \quad .
+        \qedhere
     \end{align*}
-    Finally, the fact that 
-    $\sum_{i = 0}^n i = n(n-1) / 2$ justifies the last equality.
 \end{proof}
@@
 based on the notion of \emph{growth rate}. The \intro{growth rate} of a
-function $f \colon \Sigma^* \to \Rel$ is defined as the minimal $k$ such that
-$\card{f(w)} = \bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that
+function $f \colon \Sigma^* \to \Rel$ is defined as the minimal $d$ such that
+$\card{f(w)} = \bigO\left(\card{w}^d\right)$. If such a $d$ exists, we say that
 the function $f$ has \intro{polynomial growth}. It turns out that for all $k
-\in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
-(resp. in $\NRat)$ that have growth rate at most $k$.
+\in \Nat$, $\ZPoly[d]$ (resp. $\NPoly[d]$) are precisely functions in $\ZRat$
+(resp. in $\NRat)$ that have growth rate at most $d$.
 
@@
     \begin{enumerate}
-        \item $f \in \ZPoly[k]$.
-        \item $f$ has \kl{polynomial growth} of degree at most $k$.
+        \item $f \in \ZPoly[d]$.
+        \item $f$ has \kl{polynomial growth} of degree at most $d$.
     \end{enumerate}
@@
     \label{stability-polyregular:lemma}
-    Let $k \geq 1$ and 
-    $f,g \in \NPoly[k]$ (resp. $\ZPoly[k]$, $\NSF[k]$, $\ZSF[k]$),
-    $L$ be a \kl{star-free language} over $\Sigma^*$.
+    Let $d \geq 1$,
+    $f,g \in \NPoly[d]$ (resp. $\ZPoly[d]$, $\NSF[d]$, $\ZSF[d]$),
+    $L$ be a \kl{star-free language} over $\Sigma^*$,
+    and $h \colon \Sigma^* \to \Gamma^*$ be a \kl{polyregular function}
+    (resp. a \kl{star-free polyregular function}).
     Then, the following 
-    are also in $\NPoly[k]$ (resp. $\ZPoly[k]$,
-    $\NSF[k]$, $\ZSF[k]$):
+    are also in $\NPoly[d]$ (resp. $\ZPoly[d]$,
+    $\NSF[d]$, $\ZSF[d]$):
+    $f \circ h$,
     $f + g \defined w \mapsto f(w) + g(w)$,
@@
     $f \circ c$ and $f \circ t$ are both \kl{rational series} that can be
-    effectively computed from $f$, and since equivalence
+    effectively computed from $f$ \textbf{TODO cite}, and since equivalence
     of rational series is decidable 

### 2024-11-27 parts/starfree.tex
typos
typo hunt

@@
 
-\subsection{Deciding $\NPoly$ inside $\ZPoly$}
+%\subsection{Deciding $\NPoly$ inside $\ZPoly$}
 
@@
 \kl{$(k,\Nat)$-combinatorial} function
-\cref{k-combinatorial:def}, following the spirit of previous
+(\cref{k-combinatorial:def}), following the spirit of previous
 characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
@@
 
-\subsection{Deciding $\NSF$ inside $\NPoly$}
+%\subsection{Deciding $\NSF$ inside $\NPoly$}
 

### 2024-11-27 parts/correction.tex
better knowledges

@@
 \prod_{i = 1}^p u_i^{X_i} \alpha_i$. That is, $q$ is syntactically defined by a
-non-commutative monomial over the monoid $\Sigma^*$. \kl{Polynomial functions}
+non-commutative monomial over the monoid $\Sigma^*$. \kl{Pumping patterns}
 are \kl{commutative} \kl{polyregular functions}.
@@
 	\label{binomial-function-star-free:lem}
-	Let $F$ be a \kl{binomial function} from $\Nat^k$ to $\Nat$.
+	Let $F$ be a \kl{simple binomial function} from $\Nat^k$ to $\Nat$.
 	Then it is \kl{represented} by a \kl{star-free polyregular function}.
@@
 	\end{equation*}
-    For large enough values of $X$, the \kl{binomial function}
+    For large enough values of $X$, the \kl{simple binomial function}
 	$\binom{X - p}{k}$ coincides with a polynomial whose leading coefficient

### 2024-11-27 parts/counter.tex
better knowledges

@@
     \label{pre-image-regular:fact}
-    The pre-image of a regular language by a \kl{polyregular function}
+    The pre-image of a regular language by a (string-to-string) \kl{polyregular function}
     is a regular language.
@@
 Please note that the same argument cannot be leveraged for proving that $P$ is
-not a \kl{$\Rel$-rational polynomial}: \cref{pre-image-regular:fact}
+not representd by a \kl{$\Rel$-rational series}: \cref{pre-image-regular:fact}
 only holds for

### 2024-11-27 parts/newintro.tex
better knowledges

@@
 \AP Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class of
-\kl{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
+\intro{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
 $\Sigma^*$ to $\mathbb{S}$ that are computed by \kl{$\mathbb{S}$-weighted
@@
 \Cref{n-in-z-rat:problem} recently received attention in the context of
-\kl{polyregular functions}, a computational model that aims to
+\kl{polyregular functions} ($\intro*\Poly$), a computational model that aims to
 generalize the theory of regular languages to the setting of string-to-string
@@
 \emph{star-free} fragment of \kl{polyregular functions} called \kl{star-free
-polyregular functions} \cite{BOJA18}. One open question in this area is to
+polyregular functions} ($\intro*\SF$) \cite{BOJA18}. One open question in this area is to
 decide whether a given \kl{polyregular function} is \kl{star-free}.

### 2024-11-27 parts/prelim.tex
better knowledges

@@
 in particular the notions of monoid morphisms, idempotents in monoids, monadic
-second-order ($\MSO$) logic and first-order ($\FO$) logic over finite words
+second-order ($\intro*\MSO$) logic and first-order ($\intro*\FO$) logic over finite words
 (see e.g. \cite{THOM97}). As aperiodicity will be a central notion of this
@@
 \AP Because the functions of interest in this paper have output in $\Nat$ or
-$\Rel$, we will only provide the definition of \kl{polyregular functions} for
-these two output semigroups, and we refer the reader to \cite{BOKL19} for the
-general definition of \kl{polyregular functions}. 
-We chose in this paper to provide a combinatorial description of
-polyregular functions with commutative outputs because it will play nicely with
-our analysis on polynomials. This description is very similar in shape to the
-\emph{finite counting automata} introduced by \cite{SCHU62}.
+$\Rel$, we will only provide the definition of \intro{polyregular functions}
+for these two output semigroups, and we refer the reader to \cite{BOKL19} for
+the general definition of \kl{polyregular functions} and their aperiodic
+counterpart, the \intro{star-free polyregular functions}. We chose in this
+paper to provide a combinatorial description of polyregular functions with
+commutative outputs because it will play nicely with our analysis on
+polynomials. This description is very similar in shape to the \emph{finite
+counting automata} introduced by \cite{SCHU62}.
 

### 2024-11-27 parts/starfree.tex
better knowledges

@@
     for all
-    \kl{polynomial functions} $q \colon \Nat^k \to \Sigma^*$,
+    \kl{pumping patterns} $q \colon \Nat^k \to \Sigma^*$,
     there exists a \kl{strongly natural binomial polynomial} $P$
@@
     \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
-    such that for all \kl{polynomial functions} $q$, there exists a
+    such that for all \kl{pumping patterns} $q$, there exists a
     \kl{strongly natural binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
@@
 
-    Let us therefore consider the \kl{polynomial function} $q \colon \Nat^k \to
+    Let us therefore consider the \kl{pumping pattern} $q \colon \Nat^k \to
     \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
@@
     for all $n \in \Nat$,
-    for all \kl{polynomial function} $q \colon \Nat^n \to \Sigma^*$,
+    for all \kl{pumping pattern} $q \colon \Nat^n \to \Sigma^*$,
     there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$

### 2024-11-27 parts/correction.tex
new proof

@@
 \intro{translation function} that maps a polynomial $P \in \Rel[X_1, \dots,
-X_n]$ to the polynomial $P(X_1 + K, \dots, X_n + K)$.
+X_k]$ to the polynomial $P(X_1 + K, \dots, X_k + K)$.
 
@@
 	Let $P = \prod_{i = 1}^k X_i^{\alpha_i}$ be a monomial.
-	Notice that $\translate{K}(P) = \prod_{i = 1}^n (X_i + K)^{\alpha_i}$,
+	Notice that $\translate{K}(P) = \prod_{i = 1}^k (X_i + K)^{\alpha_i}$,
 	and using a binomial expansion
@@
 	Let us assume that the alphabet over which the (\kl{commutative}) functions
-	$f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_n}$, with $a_i$
+	$f_\Delta$ and $f_\nu$ are defined is $\set{a_1, \dots, a_k}$, with $a_i$
 	representing the indeterminate $X_i$ of the polynomials. Now, let us define
@@
 		$\Nat$-polyregular function}, defined on words $w$ of the form $w \defined
-		a_1^{x_1} \cdots a_n^{x_n}$, with $x_1, \dots, x_n \geq 0$.
+		a_1^{x_1} \cdots a_k^{x_k}$, with $x_1, \dots, x_k \geq 0$.
 
@@
 		\begin{cases}
-			f_{[X_i \mapsto x_i]}(w)                     & \text{ if } \exists i \in \set{1, \dots, n}, x_i \leq K \\
-			f_\Delta(a_1^{x_1 - K} \cdots a_n^{x_n - K}) & \text{ otherwise }
+			f_{[X_i \mapsto x_i]}(w)                     & \text{ if } \exists i \in \set{1, \dots, k}, x_i \leq K \\
+			f_\Delta(a_1^{x_1 - K} \cdots a_k^{x_k - K}) & \text{ otherwise }
 		\end{cases}

### 2024-11-27 parts/starfree.tex
new proof

@@
 assumption of \kl{commutativity}. 
-
-%\subsection{Deciding $\NPoly$ inside $\ZPoly$}
-
 To characterize $\NPoly$ inside $\ZPoly$ we introduce the notion of
@@
 
-\AP
-\Cref{k-combinatorial:def} is meaningful under the hypothesis of
-\kl{commutativity}, because the latter reduces $\ZPoly$ to a combination of a
-finite number of polynomials, arranged by counting modulo some numbers. This
-characterization is stated in \cref{decompose-polynomial:lem}, and will
-be the key ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$.
-To actually state the lemma, we will need some notations that will
-formalize the intuitive idea of identifying tuples of integers modulo some
-constant. Let $\omega \in \Nat_{\geq 1}$, we define the set
-$\intro*\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
-\dots, \omega - 1}$ of \intro{types modulo $\omega$}. To a number $n \in \Nat$,
-we associate its \reintro{$\omega$-type} written
-$\intro*\moduloType[\omega](n)$ which is defined as the pair $(t, r)$ where $t
-= \min (n, \omega^2)$ and $r$ is the  remainder of the Euclidean division of
-$n$ by $\omega$. This notion of type is lifted to vectors in $\Nat^p$ for any
-$p \in \Nat$ by pointwise application.
+\AP Given a number $\omega \in \Nat$, let us write
+$\intro*\ModuloTypes[\omega]^k$ for the collection of pairs $(S, \vec{r})$
+where $S \subseteq \set{1, \dots, k}$ and $r \in \set{0, \dots, \omega - 1}^k$.
+To a tuple $\vec{x} \in \Nat^k$, one can associate its \intro{$\omega$-type}
+$\intro*\moduloType[\omega](\vec{x})$ which is the pair $(S, \vec{r})$ where $S
+= \set{i \in \set{1, \dots, k} \mid x_i \geq \omega}$ and $\vec{r} = (x_i \mod
+\omega)_{i \in \set{1, \dots, k}}$.
 
@@
     \kl{$\Rel$-polyregular function},
-    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
+    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_k}$.
     There exists a computable
@@
     and computable 
-    \kl{integer binomial polynomials} $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
-    such that for all $x_1, \dots, x_n \geq 0$:
+    \kl{integer binomial polynomials} 
+    $P_{(S,\vec{r})} \in \Rel[\seqof{X_i}{i \in S}]$ for $(S,\vec{r}) \in \ModuloTypes[\omega]^k$,
+    such that for all $\vec{x} \in \Nat^k$,
     \begin{equation*}
-        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
-        = P_{\moduloType[\omega](x_1, \dots, x_n)}
+        f\left(
+            \prod_{i = 1}^k a_i^{x_i}
+        \right)
+        = P_{(S, \vec{r})}
         \left(
-            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
+            \seqof{\floor{x_i / \omega}}{i \in S}
         \right)
+        \text{ where } (S, \vec{r}) = \moduloType[\omega](\vec{x})
         \quad .
@@
 \begin{proof}
-    Let us first write $f \in \ZPoly$ as a linear combination 
-    $f = f_+ - f_-$ where both $f_+$ and $f_-$ are in $\NPoly$.
-    We 
-    compute $\omega_+$ (resp. $\omega_-$)
-    by
-    applying
-    \cref{n-poly-combinatorics:lem} 
-    to $f_+$ (resp. to $f_-$).
-    Finally, let $\omega$ be the least common multiple
-    of $\omega_+$ and $\omega_-$, i.e. $\omega \defined \ppcm(\omega_+, \omega_-)$.
-    Using this $\omega$,
-    for any choice of
-    $r \defined (r_1, \dots, r_n) \in \set{0, \dots, \omega - 1}$,
-    there exists two \kl{natural binomial functions} $F_r^+$ and $F_r^-$
-    associated to $f_+$ and $f_-$,
-    such that $x_1, \dots, x_p \geq 1$:
-    \begin{equation*}
-        f\left(
-        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right) 
-        = F_r^+(x_1, \dots, x_p) - F_r^-(x_1, \dots, x_p) \quad .
-    \end{equation*}
-    Now, for large enough values of $x_1, \dots, x_p$,
-    this means that 
-    $f\left(
-        a_1^{\omega x_1 + r_1} \cdots a_n ^{\omega x_n + r_n}\right)$
-    coincides with an \kl{integer binomial polynomial},
-    and we can therefore conclude selecting different values of $r$ and
-    setting some subsets of variables to be assigned to the constant $0$.
+    \textbf{TODO}. 
 \end{proof}
@@
     For the converse implication, we use \cref{decompose-polynomial:lem} 
-    effectively compute a polynomial representation of $f$. That is, we obtain
-    an $\omega_1 \in \Nat$ and \kl{integer binomial polynomials}
-    $\seqof{P_{t}}{t \in \ModuloTypes[\omega_1]^k}$. Now, because $f$ is
-    \kl{$(k,\Nat)$-combinatorial}, we also have a number $\omega_2 \in \Nat$
-    such that for all \kl{pumping patterns} $q$, there exists a
-    \kl{strongly natural binomial polynomial} $P$ such that $f \circ q(\omega_2 \vec{X})
-    = P$ over $(\Nat_{\geq 1})^k$. By considering $\omega \defined
-    \omega_1 \times \omega_2$, we can without loss of generality assume that
-    $\omega_1 = \omega_2$ in the rest of this proof.
+    effectively compute an $\omega \in \Nat$ and \kl{integer binomial polynomials}
+    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$.
 
-    Now, consider $t \in \ModuloTypes[\omega]^k$. By
-    definition, this provides us with a set $S \subseteq \set{1, \dots, k}$,
-    and numbers $r_1, \dots, r_k$ between $0$ and
-    $\omega^2-1$ such that for all $x_1, \dots, x_k \geq 1$, the tuple $(x_1
-    \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has type $t$.
-    In particular, for all $x_1, \dots, x_k \geq 1$, the following
+    Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
+    Notice that for all vectors $\vec{x} \in (\Nat_{\geq 1})^k$, 
+    the vector $(x_1 \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has
+    \kl{$\omega$-type} $(S, \vec{r})$.
+    In particular, the following
     equality holds:
@@
         =
-        P_t(\seqof{x_i}{i \in S})
+        P_{(S, \vec{r})}(\seqof{x_i}{i \in S})
+        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
+        \quad .
     \end{equation*}
@@
     \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
-    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. Because $f$ is
-    \kl{$(k,\Nat)$-combinatorial} (with parameter $\omega$), there exists a
-    \kl{strongly natural binomial polynomial} $P$ such that for all $x_1,
-    \dots, x_n \geq 1$, the equality $f \circ q(\omega x_1, \dots, \omega x_k)
-    = P(x_1, \dots, x_k)$ holds. In particular, we conclude that over
-    $(\Nat_{\geq 1})^{\card{S}}$ the polynomial $P_t$ and $P$ coincide. They
-    are therefore equal as multivariate polynomials, and we conclude that $P_t$
-    is a \kl{strongly natural binomial polynomial} \kl{$\Nat$-rational
-    polynomial}.
+    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. 
+    Because $f$ is \kl{$(k,\Nat)$-combinatorial}, there exists an $\omega_2 \in
+    \Nat$ and a \kl{strongly natural binomial polynomial} $P \in \Rel[X_1,
+    \dots, X_k]$ such that $f \circ q = P$ over $(\omega_2 \Nat_{\geq 1})^k$.
+    In particular, the following
+    equality holds:
+    \begin{equation*}
+        P_{(S, \vec{r})}(\seqof{\omega_2 x_i}{i \in S})
+        = 
+        P(\omega_2 x_1, \dots, \omega_2 x_k)
+        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
+        \quad .
+    \end{equation*}
+    Because $\omega_2 \neq 0$, we conclude that $P_{(S, \vec{r})} = P$
+    and that both are polynomials over $\Rel[\seqof{X_i}{i \in S}]$,
+    and furthermore that $P_{(S, \vec{r})}$
+    is a \kl{strongly natural binomial polynomial}.
 
-    Let us now construct a new way to compute $f$, by first
-    computing the \kl{$\omega$-type} of the input (which is possible in
-    $\MSO$), and then computing the suitable \kl{strongly natural binomial polynomials}
+    To conclude, one can compute $f$ by first
+    computing the \kl{$\omega$-type} of the input which is possible using a \kl{polyregular function},
+    and then computing the suitable \kl{strongly natural binomial polynomials}
     which is possible in $\NPoly$
-    thanks to \cref{decide-rat-poly-npoly:cor}.
+    thanks to \cref{decide-rat-poly-npoly:cor}. The resulting composition 
+    belongs to $\NPoly$ thanks to
+    \cref{stability-polyregular:lemma}.
+
+    Let us now argue that the properties are decidable. Indeed, consider a
+    function $f \in \ZPoly[k]$. Again, using
+    \cref{decompose-polynomial:lem}, we can effectively compute the
+    $\omega \in \Nat$ and the \kl{integer binomial polynomials}
+    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$. If all of those polynomials
+    are \kl{strongly natural binomial polynomials} (which is 
+    decidable by \cref{decide-rat-poly-npoly:cor}),
+    then $f \in \NPoly[k]$ as
+    argued above. Conversely, if $f \in \NPoly[k]$, then $f$ is
+    \kl{$(k,\Nat)$-combinatorial} by \cref{n-poly-combinatorics:lem}, and we
+    argued that this implies that $P_{(S, \vec{r})}$ is a \kl{strongly natural
+    binomial polynomial} for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$.
 \end{proof}

### 2024-11-29 parts/conclusion.tex
more typos

@@
 Let us end on a more general discussion regarding the status of commutative
-input functions in the study of unary output polyregular functions. Throughout
-the paper, we have tried to highlight that results on polyregular functions
-often have the following form: a polyregular function $f \colon \Sigma^* \to
-\Gamma^*$ satisfies property $X$ if and only if for all \kl{commutative}
-\kl{star-free polyregular function} $g \colon \Delta^* \to \Sigma^*$, $f \circ
-g$ satisfies $X$. Let us formally restate this for two earlier results in the
-case of \kl{$\Rel$-polyregular functions} in
-\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}.
+input functions in the study of unary output polyregular functions. A
+\emph{quantitative pumping argument} for polyregular function $f \colon
+\Sigma^* \to \Rel$ states that $f$ has property $X$ if and only if for all
+\kl{pumping pattern} $q \colon \Nat^k \to \Sigma^*$, $f \circ q$ has property
+$X$. Let us formalize such a statement for \kl{growth rate} and
+\kl(polyregular){aperiodicity} respectively in \cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma}. Note that we generalized \kl{pumping
+patterns} to \kl{commutative} \kl{star-free polyregular functions} to simplify
+the statements.
 
@@
     \proofref{pre-compose-growth-commut:lemma}
-    Let $f \in \ZRat$, and $k \in \Nat$. Then,
-    $f \in \ZPoly[k]$ if and only if 
+    Let $f \in \ZRat$, and $d \in \Nat$. Then,
+    $f \in \ZPoly[d]$ if and only if 
     for every \kl{commutative} \kl{star-free polyregular function} $h$
             of \kl{growth rate} $l \in \Nat$,
-            $(f \circ h) \in \ZPoly[k\times l]$.
+            $(f \circ h) \in \ZPoly[d \times l]$.
 \end{lemma}
@@
     \label{npoly-zpoly:conjecture}
-    Let $k \in \Nat$.
-    A function $f \in \ZPoly[k]$
-    belongs to $\NPoly[k]$ if and only if
-    it is \kl{$(k,\Nat)$-combinatorial}.
+    Let $f \in \ZPoly$. Then, $f \in \NPoly$ if and only if for every
+    \kl{commutative} \kl{star-free polyregular function} $h$,
+    $(f \circ h) \in \NPoly$.
 \end{conjecture}

### 2024-11-29 parts/starfree.tex
more typos

@@
 \kl{$(k,\Nat)$-combinatorial} function
-(\cref{k-combinatorial:def}), following the spirit of previous
+(\cref{k-combinatorial:def}), following the spirit of revious
 characterizations of subclasses of $\ZPoly$ in terms of \emph{polynomial
@@
     \label{decidable-n-poly:thm}
-    Let $k \in \Nat$, and $f \in \ZPoly[k]$ be \kl{commutative}.
+    Let $k,d \in \Nat$, and $f \in \ZPoly[d]$ be \kl{commutative}
+    over an alphabet of size $k$.
     Then, the following are equivalent:
@@
         \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
-        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
+        \item \label{f-npoly-combi:item} $f \in \NPoly[d]$,
     \end{enumerate}
@@
 
-    To conclude, one can compute $f$ by first
-    computing the \kl{$\omega$-type} of the input which is possible using a \kl{polyregular function},
-    and then computing the suitable \kl{strongly natural binomial polynomials}
-    which is possible in $\NPoly$
-    thanks to \cref{decide-rat-poly-npoly:cor}. The resulting composition 
-    belongs to $\NPoly$ thanks to
-    \cref{stability-polyregular:lemma}.
+    To conclude that $f \in \NPoly$, one can compute $f$ by first computing
+    the \kl{$\omega$-type} of the input which is possible using a
+    \kl{polyregular function}, and then computing
+    the suitable \kl{strongly natural binomial polynomials} which is possible
+    in $\NPoly$ thanks to \cref{decide-rat-poly-npoly:cor}. The
+    resulting composition belongs to $\NPoly$ thanks to
+    \cref{stability-polyregular:lemma},
+    and we conclude that $f \in \NPoly[d]$ because
+    it has \kl{growth rate} at most $d$
+    (\cref{polyregular-polynomial-growth:lemma}).
 
-    Let us now argue that the properties are decidable. Indeed, consider a
-    function $f \in \ZPoly[k]$. Again, using
-    \cref{decompose-polynomial:lem}, we can effectively compute the
-    $\omega \in \Nat$ and the \kl{integer binomial polynomials}
-    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$. If all of those polynomials
-    are \kl{strongly natural binomial polynomials} (which is 
-    decidable by \cref{decide-rat-poly-npoly:cor}),
-    then $f \in \NPoly[k]$ as
-    argued above. Conversely, if $f \in \NPoly[k]$, then $f$ is
-    \kl{$(k,\Nat)$-combinatorial} by \cref{n-poly-combinatorics:lem}, and we
-    argued that this implies that $P_{(S, \vec{r})}$ is a \kl{strongly natural
-    binomial polynomial} for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$.
+    Let us remark that we actually proved that the function $f \in \ZPoly[d]$
+    belongs to $\NPoly[d]$
+    if and only if its corresponding polynomials
+    $P_{(S, \vec{r})}$ are \kl{strongly natural binomial polynomials} for all
+    $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. Because 
+    \cref{decompose-polynomial:lem} is effective, this $\omega \in \Nat$ and polynomials
+    $P_{(S, \vec{r})}$ can be computed from $f$, 
+    and one can decide 
+    whether a polynomial is \kl{strongly natural binomial polynomial} using
+    \cref{decide-rat-poly-npoly:cor}. Hence, one can decide whether
+    $f \in \NPoly[d]$.
 \end{proof}
@@
     when there exists $N_0 \in \Nat$ such that
-    for all $n \in \Nat$,
-    for all \kl{pumping pattern} $q \colon \Nat^n \to \Sigma^*$,
-    there exists a polynomial $P \in \Rat[X_1, \dots, X_n]$
+    for all $k \in \Nat$,
+    for all \kl{pumping pattern} $q \colon \Nat^k \to \Sigma^*$,
+    there exists a polynomial $P \in \Rat[X_1, \dots, X_k]$
     such that:
     \begin{equation*}
-        f \circ p = P
+        f \circ q = P
         \quad 
-        \text{ over } (\Nat_{\geq N_0})^n
+        \text{ over } (\Nat_{\geq N_0})^k
         \quad .
@@
     following previous results for \kl{star-free $\Rel$-polyregular functions}
-    \cite{CDTL23}.
+    \cite[Theorem V.13]{CDTL23}.
 
@@
     $\Sigma$. By definition, there exists $N_0 \in \Nat$, and $P \in
-    \Rat[\seqof{X_a}{a \in \Sigma}]$ such that: \begin{equation*} f\left(
-        \prod_{a \in \Sigma} a^{X_a} \right) = P(\seqof{X_a}{a \in \Sigma})
-        \quad \text{ when } \forall a \in \Sigma, X_a \geq N_0 \quad .
-    \end{equation*} It is clear that $P$ is \kl{represented} by a
-    \kl{$\Nat$-polyregular function}, namely, $f$ partially applied to the word
-    $\prod_{a \in \Sigma} a^{N_0}$, and therefore is in fact represented by a
+    \Rat[\seqof{X_a}{a \in \Sigma}]$ such that:
+
+    \begin{equation*} f\left(
+        \prod_{a \in \Sigma} a^{x_a} \right)
+        = 
+        P(\seqof{x_a}{a \in \Sigma})
+        \quad \forall \vec{x} \in (\Nat_{\geq N_0})^{\Sigma} \quad .
+    \end{equation*}
+
+    It is clear that $\translate{N_0}(P)$ is \kl{represented} by an
+    \kl{$\Nat$-polyregular function}, namely, $f_u \colon w \mapsto f(uw)$
+    where $u \defined \prod_{a \in \Sigma} a^{N_0}$, and is therefore represented by a
     \kl{star-free $\Nat$-polyregular function} thanks to
@@
     For every letter $a \in \Sigma$ and number $0 \leq n \leq N_0$, there exists, by induction hypothesis,
-    a \kl{star-free $\Nat$-polyregular function} $g_a^n$ that represents the
+    a \kl{star-free $\Nat$-polyregular function} $g_{a^n}$ that \kl{represents} the
     function $f_{a^n} \colon (\Sigma \setminus \set{a})^* \to \Rel$
-    that maps $u \in (\Sigma \setminus \set{a})^*$ to $f(a^n u)$.
+    that maps $w \in (\Sigma \setminus \set{a})^*$ to $f(a^n w)$.
 
@@
             g_{a^n} (w) & \text{ if } \card[a]{w} = n \text{ for some } a \in \Sigma \text{ and } n \leq N_0 \\
-            P(\seqof{\card[a]{w}}{a \in \Sigma}) & \text{ otherwise }
+            \translate{N_0}(P)(\seqof{\card[a]{w} - N_0}{a \in \Sigma}) & \text{ otherwise }
         \end{cases}

### 2024-11-29 parts/appendix.tex
big corrections

@@
     Recall that by the undecidability of the Hilbert's Tenth Problem \cite[Problem 10 page 22]{HILB1902}, it is
-    undecidable whether a given polynomial $P \in \Rel[\vec{X}]$ has a solution in
-    $\Rel^n$ \cite{MATI1970,DAVIS1973}.
+    undecidable whether a given polynomial $P \in \Rel[X_1, \dots, X_k]$ has a solution in
+    $\Rel^k$ \cite{MATI1970,DAVIS1973}.
     The result follows by observing that
-    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^n$ if and only 
+    checking whether a polynomial $P \in \Rel[\vec{X}]$ has no solutions in $\Rel^k$ if and only 
     if $P^2 - 1$ is \kl{non-negative}.
@@
     where both $P_+$ and $P_-$ belong to $\Nat[\vec{X}]$. Let $Z$ be a fresh
-    indeterminate. We define $Q \defined P_+ (Z - 1) P_-$. By construction, $Q$
+    indeterminate. We define $Q \defined P_+ + (Z - 1) P_-$. By construction, $Q$
     has \kl{non-negative} \kl{maximal monomials}. Furthermore, $Q$ itself is
@@
 	$\pi \colon M^k \to \Nat$,
-	the lemma holds with $s = 0$ for the function
+	the lemma holds for the function
 	$f = \pi^\dagger$, i.e., such that
@@
 
-	In all cases, we will use a \kl{polynomial function} $q \colon \Nat^p \to
+	In all cases, we will use a \kl{pumping pattern} $q \colon \Nat^p \to
 		\Sigma^*$ defined by $q(X_1, \dots, X_p) \defined \alpha_0 \prod_{i = 1}^p
@@
 	mentioned above,
-	one can define the \kl{polynomial functions}
+	one can define the \kl{pumping patterns}
 	$\CutPref{\ell,r,b} \colon \Nat^{\ell - 1} \to \Sigma^*$
@@
     to obtain a number $s \geq 1$ (that only depends on $P$) such
-    that for all \kl{polynomial function} $q \colon \Nat^k \to \Sigma^*$, $g
-    \circ q(X_1 + s, \dots, X_k + s)$ is a \kl{natural binomial function} $F$
-    over $\Nat^k$. Let us write the
-    \kl{natural binomial function} $F$ as follows $F(X_1, \dots, X_k) =
-    \sum_{\card{\alpha} \leq d_2} e_{\alpha} \prod_{i = 1}^k \binom{X -
-    p_{\alpha,i}}{\alpha(i)}$, with $e_{\alpha} \in \Nat$ for all
-    $\card{\alpha} \leq d_2$.
-    Now, because of the equality 
-    $\binom{X - p - 1}{k} + \binom{X - p - 1}{k - 1} = \binom{X - p}{k}$,
-    we conclude that we can find a $K \in \Nat$ and coefficients
-    $f_{\alpha} \in \Nat$ such that 
-    $F(X_1, \dots, X_k) =
-    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \binom{X_i - K}{\alpha(i)}$.
-    Now, remark that $P$ is an \kl{integer-valued} polynomial of
-    degree $d$, and therefore one can write $P = \sum_{\card{\alpha} \leq d}
-    c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$ with $c_{\alpha}
-    \in \Rel$ for all $\card{\alpha} \leq d$
-    (\cref{integer-binomial-polynomial:lem}). 
-    As a consequence,
-    $\lambda P = \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K - s}{\alpha(i)}$,
-    and 
-    by identifying coefficients in the previous two expressions, we obtain
-    $c_{\alpha} = \lambda f_{\alpha}$ for all $\card{\alpha} \leq d$.
-    Finally,
-    $\lambda P(x_1 + s, \dots, x_k + s) = F(x_1, \dots, x_k)$ for all $\vec{x} \in \Nat^k$,
-    and therefore
-    $\lambda P (X_1 + s, \dots, X_k + s) =
-    \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}$.
-    In particular,
-    we get
+    that for all \kl{pumping patterns} $q \colon \Nat^p \to \Sigma^*$, $f
+    \circ q(X_1 + s, \dots, X_p + s)$ is a \kl{natural binomial function} $F$
+    over $\Nat^p$. We will use this $s$ with $p = k$, where
+    $k = \card{\vec{X}}$ is the number of indeterminates of $P$,
+    and $q(X_1, \dots, X_k) \defined \prod_{i = 1}^k a_i^{X_i}$,
+    assuming the alphabet $\Sigma$ is $\set{a_1, \dots, a_k}$.
+    In particular, $F(x_1, \dots, x_k) = f \circ q (x_1, \dots, x_k) = nP(x_1, \dots, x_k)$
+    for all $\vec{x} \in (\Nat_{\geq s})^k$.
+    Let us write the
+    \kl{natural binomial function} $F$ as follows:
     \begin{equation*}
-        \sum_{\card{\alpha} \leq d} \lambda c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)} =
-        \sum_{\card{\alpha} \leq d_2} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
+        F(X_1, \dots, X_k) =
+        \sum_{\card{\alpha} \leq d} e_{\alpha} \prod_{i = 1}^k \binom{X -
+        p_{\alpha,i}}{\alpha(i)} \quad
+        \text{where } \forall 
+        \card{\alpha} \leq d,
+        e_{\alpha} \in \Nat
+        \quad .
     \end{equation*}
-    And we conclude by term identification
-    by term identification
-    that $d = d_2$, and for all $\card{\alpha} \leq d$, $\lambda c_{\alpha} = f_{\alpha}$.
-    As a consequence, $c_{\alpha} \in \Nat$ for all $\card{\alpha} \leq d$,
-    and $P$
-    is a \kl{natural binomial polynomial}.
-    Remark that this reasoning remains true whenever we partially apply $P$ to some values
-    in $\Nat$, as it amounts to 
-    partially applying the function $f$. We have proven that 
-    $P$ is a \kl{strongly natural binomial polynomial}.
+
+    Because of the equality $\binom{X - p - 1}{k} + \binom{X - p - 1}{k -
+    1} = \binom{X - p}{k}$, we conclude that we can find a $K \in \Nat$ and
+    coefficients $f_{\alpha} \in \Nat$ such that $F(X_1, \dots, X_k) =
+    \sum_{\card{\alpha} \leq d} f_{\alpha} \prod_{i = 1}^k \binom{X_i -
+    K}{\alpha(i)}$, i.e., we can assume that $p_{\alpha,i} = K$ for all
+    $\card{\alpha} \leq d$.
+    Let us remark that $P$ is an \kl{integer-valued} polynomial of
+    degree $d$, and therefore one can use
+    \cref{integer-binomial-polynomial:lem} to rewrite it in the following form:
+    \begin{equation*}
+        P(X_1, \dots, X_k) = \sum_{\card{\alpha} \leq d}
+        c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
+        \quad \text{where } \forall \card{\alpha} \leq d, 
+        c_{\alpha}
+        \in \Rel \quad .
+    \end{equation*}
+
+    Because $F(x_1, \dots, x_k) = P(x_1, \dots, x_k)$ for all
+    $\vec{x} \in (\Nat_{\geq s})^k$, one can write
+    \begin{equation*}
+        n \left(\sum_{\card{\alpha} \leq d} c_{\alpha} \prod_{i = 1}^k \pbinom{x_i - K}{\alpha(i)} \right)
+        = 
+        \sum_{\card{\alpha} \leq d} f_{\alpha} \prod_{i = 1}^k \binom{x_i - K}{\alpha(i)}
+        \quad 
+        \forall \vec{x} \in (\Nat_{\geq s})^k
+    \end{equation*}
+
+    Because this equality holds over $(\Nat_{\geq s})^k$, we conclude that the
+    underlying polynomials are equal in $\Rel[\vec{X}]$, and therefore that:
+    \begin{equation*}
+        \sum_{\card{\alpha} \leq d} n c_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
+        = 
+        \sum_{\card{\alpha} \leq d} f_{\alpha} \prod_{i = 1}^k \pbinom{X_i - K}{\alpha(i)}
+    \end{equation*}
+
+    Identifying the coefficients of the two polynomials (which is possible
+    because the products of binomial coefficients translated by $K$ form a
+    basis of $\Rel[\vec{X}]$), we conclude that for all $\card{\alpha} \leq d$,
+    $n c_{\alpha} = f_{\alpha}$. In particular,
+    for all $\card{\alpha} \leq d$, $c_{\alpha} \in \Nat$,
+    hence $P$ is a \kl{natural binomial polynomial}.
+
+    Remark that this reasoning remains true whenever we partially apply $P$ to
+    some values in $\Nat$, since partial applications of $nP$ continue to be in
+    $\CorrectPoly$ by definition, and remain \kl{integer-valued}. We have
+    proven that $P$ is a \kl{strongly natural binomial polynomial}.
+\end{proofof}
+
+\section{Proofs of section  \ref{beyond-polynomials:sec}}
+
+\begin{proofof}[decompose-polynomial:lem]
+    Let us first remark that one only needs to consider the case where $f$ is a
+    function in $\NPoly$, because every function $f \in \ZPoly$ can be decomposed
+    as $f = f_+ - f_-$ where $f_+$ and $f_-$ 
+    are \kl{$\Nat$-polyregular functions}, and because \kl{integer binomial
+    polynomials} are closed under subtraction.
+    Now, let us prove this result by induction on the number of letters in the alphabet.
+    When the alphabet is empty, the result is trivial because $f$ can only be applied to $\varepsilon$,
+    hence can be represented using a constant polynomial.
+
+    Assume that the result holds for all functions with an alphabet of size at
+    most $k-1$. We first apply \cref{n-poly-combinatorics:lem} to obtain a number
+    $\omega$ such that for all \kl{pumping patterns} $q \colon \Nat^\ell \to
+    \Sigma^*$, there exists a (computable) \kl{natural binomial function} $F_q$
+    such that $f \circ q (\omega X_1, \dots, \omega X_\ell) = F_q$ over
+    $(\Nat_{\geq 1})^\ell$.
+    Given a tuple $\vec{r} \in \set{0, \dots, \omega - 1}^k$, and a 
+    set $S \subseteq \set{1, \dots, k}$,
+    let us define
+    the \kl{pumping pattern} $q_{(S,\vec{r})} \colon \Nat^{\card{S}} \to \Sigma^*$ as
+    $q_{(S, \vec{r})}(\seqof{X_i}{i \in S}) \defined \prod_{i = 1}^k a_i^{r_i} \cdot \prod_{i \in S} a_i^{X_i}$.
+    Let us write $F_{(S,\vec{r})} \colon \Nat^{\card{S}} \to \Nat$ 
+    for the associated
+    \kl{natural binomial function} satisfying:
+    \begin{equation*}
+        f \left(
+            \prod_{i = 1}^k a_i^{r_i}
+            \cdot
+            \prod_{i \in S} a_i^{\omega x_i}
+        \right)
+        = 
+        F_{(S,\vec{r})}(\seqof{x_i}{i \in S})
+        \quad 
+        \forall \vec{x} \in (\Nat_{\geq 1})^{S}
+        \quad .
+    \end{equation*}
+
+    Such a function $F_{(S,\vec{r})}$ coincides with a \kl{natural binomial
+    polynomial} $P_{(S,\vec{r})} \in \Rat[\seqof{X_i}{i \in S}]$ on large
+    enough inputs. Let us write $N_0 \geq 1$ be a number such that
+    $F_{(S,\vec{r})}(\seqof{x_i}{i \in S}) = P_{(S,\vec{r})}(\seqof{x_i}{i \in
+    S})$ for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$, for all $\vec{x}
+    \in (\Nat_{\geq N_0})^{S}$. This $N_0$ is computable because one can
+    collect upper bounds from the syntax of the \kl{natural binomial functions}
+    $F_{(S,\vec{r})}$.
+    We conclude that:
+    \begin{equation*}
+        f\left(
+            \prod_{i = 1}^k a_i^{x_i}
+        \right)
+        = 
+        P_{(S, \vec{r})}(\seqof{\floor{x_i / \omega}}{i \in S})
+        \quad 
+        \text{ where } (S, \vec{r}) = \moduloType[\omega](\vec{x})
+        \text{ and }
+        \vec{x} \in (\Nat_{\geq N_0 \omega})^k
+        \quad .
+    \end{equation*}
+
+    This provides a representation for $f$ on large enough inputs, and we can
+    apply the induction hypothesis on the functions $f_{a^n} \colon (\Sigma
+    \setminus \set{a})^* \to \Nat$ in $\NPoly$ defined by $f_{a^n} (w) \defined
+    f(a^n \cdot w)$, for every $a \in \Sigma$ and $n \leq N_0\omega$. Remark
+    that \kl{integer binomial polynomials} are closed multiplying the input by
+    a constant, hence that we can always move from a decomposition using
+    $\omega$ to a decomposition using a multiple of $\omega$.
+    Let us then consider as our new $\omega'$ the least common multiple of
+    $\omega$, $N_0$, and all the $\omega_{a^n}$'s obtained for each $f_{a^n}$.
+
+    We can now, given an entry $\vec{x} \in \Nat^k$ with
+    $\moduloType[\omega'](x) = (S, \vec{r})$ recognize if a some letter
+    $a$ appears less than $N_0 \omega$ times in the input, and apply the
+    corresponding polynomials from the induction hypothesis if it is the case.
+    Whenever all letters appear at least $N_0 \omega$ times, we can apply the
+    polynomial $P_{(S,\vec{r})}$.
 \end{proofof}

### 2024-11-29 parts/prelim.tex
big corrections

@@
     $f \circ c$ and $f \circ t$ are both \kl{rational series} that can be
-    effectively computed from $f$ \textbf{TODO cite}, and since equivalence
+    effectively computed from $f$,\footnote{
+        This can be done by guessing the second (resp. last) letter of the input word, 
+        remembering the first letter in a state, and 
+        then running the original automaton for $f$ on the modified input, checking at the second position 
+        (resp. the end of the word)
+        if the guess was correct.
+    } and since equivalence
     of rational series is decidable 

### 2024-11-29 parts/starfree.tex
big corrections

@@
 
-\AP Given a number $\omega \in \Nat$, let us write
-$\intro*\ModuloTypes[\omega]^k$ for the collection of pairs $(S, \vec{r})$
-where $S \subseteq \set{1, \dots, k}$ and $r \in \set{0, \dots, \omega - 1}^k$.
-To a tuple $\vec{x} \in \Nat^k$, one can associate its \intro{$\omega$-type}
-$\intro*\moduloType[\omega](\vec{x})$ which is the pair $(S, \vec{r})$ where $S
-= \set{i \in \set{1, \dots, k} \mid x_i \geq \omega}$ and $\vec{r} = (x_i \mod
-\omega)_{i \in \set{1, \dots, k}}$.
+\AP Let us now introduce a decomposition of \kl{commutative}
+\kl{$\Rel$-polyregular functions} into \kl{integer binomial polynomials}. Given
+a number $\omega \in \Nat$, let us write $\intro*\ModuloTypes[\omega]^k$ for
+the collection of pairs $(S, \vec{r})$ where $S \subseteq \set{1, \dots, k}$
+and $r \in \set{0, \dots, \omega - 1}^k$. To a tuple $\vec{x} \in \Nat^k$, one
+can associate its \intro{$\omega$-type}, written
+$\intro*\moduloType[\omega](\vec{x})$, which is the pair $(S, \vec{r})$ where
+$S = \set{i \in \set{1, \dots, k} \mid x_i \geq \omega}$ and $\vec{r} = (x_i
+\mod \omega)_{i \in \set{1, \dots, k}}$.
 
-\begin{lemma}
-    \label{decompose-polynomial:lem}
+\begin{lemma}[restate=decompose-polynomial:lem,label=decompose-polynomial:lem]
+    \proofref{decompose-polynomial:lem}
     Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
@@
     \kl{integer binomial polynomials} 
-    $P_{(S,\vec{r})} \in \Rel[\seqof{X_i}{i \in S}]$ for $(S,\vec{r}) \in \ModuloTypes[\omega]^k$,
+    $P_{(S,\vec{r})} \in \Rat[\seqof{X_i}{i \in S}]$ for $(S,\vec{r}) \in \ModuloTypes[\omega]^k$,
     such that for all $\vec{x} \in \Nat^k$,
@@
 \end{lemma}
-\begin{proof}
-    \textbf{TODO}. 
-\end{proof}
 
@@
 \begin{proof}
-    The implication
-    \cref{f-npoly-combi:item}
-    $\implies$ \cref{f-combinatorial:item} immediately follows
-    from 
-    \cref{n-poly-combinatorics:lem} and \cref{decide-rat-poly-npoly:cor}.
-    For the converse implication, we use \cref{decompose-polynomial:lem} 
-    effectively compute an $\omega \in \Nat$ and \kl{integer binomial polynomials}
-    $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^k}$.
+    Let $f \in \ZPoly[d]$ be \kl{commutative} over an alphabet of size $k$.
+    We apply \cref{decompose-polynomial:lem} to compute an $\omega \in \Nat$ and
+    \kl{integer binomial polynomials} $\seqof{P_{(S,\vec{r})}}{(S,\vec{r}) \in
+    \ModuloTypes[\omega]^k}$ such that for all $\vec{x} \in \Nat^k$,
+    $f\left(\prod_{i = 1}^k a_i^{x_i}\right) = P_{(S, \vec{r})}(\seqof{\floor{x_i / \omega}}{i \in S})$,
+    where $(S, \vec{r}) = \moduloType[\omega](\vec{x})$.
+    We are first going to prove that $f \in \NPoly[d]$ if and only if $P_{(S,
+    \vec{r})}$ is a \kl{strongly natural binomial polynomial} for all $(S,
+    \vec{r}) \in \ModuloTypes[\omega]^k$. This will also provide decidability
+    of \cref{f-npoly-combi:item}, since one can decide whether a
+    polynomial is \kl{strongly natural binomial polynomial} using
+    \cref{decide-rat-poly-npoly:cor}.
 
+    Assume that $f \in \NPoly$, then by definition, the polynomials $P_{(S,
+    \vec{r})}$ are \kl{represented} by an \kl{$\Nat$-polyregular function},
+    hence are \kl{strongly natural binomial polynomials}
+    (\cref{decide-rat-poly-npoly:cor}). Conversely, if
+    $P_{(S, \vec{r})}$ is a \kl{strongly natural binomial polynomial} for all
+    $(S, \vec{r}) \in \ModuloTypes[\omega]^k$, then $f \in \NPoly$ because one
+    can compute the \kl{$\omega$-type} of the input using a \kl{polyregular
+    function}, and then compute the suitable \kl{strongly natural binomial
+    polynomial} $P_{(S, \vec{r})}$ which is possible in $\NPoly$ thanks to
+    \cref{decide-rat-poly-npoly:cor}. The resulting composition
+    belongs to $\NPoly$ thanks to
+    \cref{stability-polyregular:lemma},
+    and we conclude that $f \in \NPoly[d]$ because it has \kl{growth rate} at
+    most $d$
+    (\cref{polyregular-polynomial-growth:lemma}).
+
+    Note that the same proof scheme can be used to conclude that
+    \cref{f-npoly-combi:item} implies
+    \cref{f-combinatorial:item}. For the converse implication, we are
+    going to introduce $\omega_2$ associated to the fact that $f$ is
+    \kl{$(k,\Nat)$-combinatorial}. Because polynomials \kl{represented} by
+    \kl{$\Nat$-polyregular functions} and \kl{integer binomial polynomials} are
+    both closed under multiplication by a constant factor, we can assume that
+    $\omega = \omega_2$ in the decomposition of $f$.
     Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
@@
     \Sigma^*$ that is simply defined as $q(X_1, \dots, X_k) \defined \prod_{i =
-    1}^k a_i^{ X_i \omega \ind{S}(i) + r_i}$. 
-    Because $f$ is \kl{$(k,\Nat)$-combinatorial}, there exists an $\omega_2 \in
-    \Nat$ and a \kl{strongly natural binomial polynomial} $P \in \Rel[X_1,
-    \dots, X_k]$ such that $f \circ q = P$ over $(\omega_2 \Nat_{\geq 1})^k$.
-    In particular, the following
-    equality holds:
-    \begin{equation*}
-        P_{(S, \vec{r})}(\seqof{\omega_2 x_i}{i \in S})
-        = 
-        P(\omega_2 x_1, \dots, \omega_2 x_k)
-        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
-        \quad .
-    \end{equation*}
-    Because $\omega_2 \neq 0$, we conclude that $P_{(S, \vec{r})} = P$
-    and that both are polynomials over $\Rel[\seqof{X_i}{i \in S}]$,
-    and furthermore that $P_{(S, \vec{r})}$
-    is a \kl{strongly natural binomial polynomial}.
-
-    To conclude that $f \in \NPoly$, one can compute $f$ by first computing
-    the \kl{$\omega$-type} of the input which is possible using a
-    \kl{polyregular function}, and then computing
-    the suitable \kl{strongly natural binomial polynomials} which is possible
-    in $\NPoly$ thanks to \cref{decide-rat-poly-npoly:cor}. The
-    resulting composition belongs to $\NPoly$ thanks to
-    \cref{stability-polyregular:lemma},
-    and we conclude that $f \in \NPoly[d]$ because
-    it has \kl{growth rate} at most $d$
-    (\cref{polyregular-polynomial-growth:lemma}).
-
-    Let us remark that we actually proved that the function $f \in \ZPoly[d]$
-    belongs to $\NPoly[d]$
-    if and only if its corresponding polynomials
-    $P_{(S, \vec{r})}$ are \kl{strongly natural binomial polynomials} for all
-    $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. Because 
-    \cref{decompose-polynomial:lem} is effective, this $\omega \in \Nat$ and polynomials
-    $P_{(S, \vec{r})}$ can be computed from $f$, 
-    and one can decide 
-    whether a polynomial is \kl{strongly natural binomial polynomial} using
-    \cref{decide-rat-poly-npoly:cor}. Hence, one can decide whether
+    1}^k a_i^{ X_i \ind{S}(i) + r_i}$. 
+    Because $f$ is \kl{$(k,\Nat)$-combinatorial} with parameter $\omega$,
+    there exists a \kl{strongly natural binomial polynomial} $P \in \Rat[X_1,
+    \dots, X_k]$ such that $f \circ q (\omega X_1, \dots, \omega X_k) = P (X_1, \dots, X_k)$ 
+    over $(\Nat_{\geq 1})^k$.
+    This proves that
+    $P_{(S, \vec{r})}(\seqof{X_i}{i \in S})$
+    equals
+    $P(X_1, \dots, X_k)$ as polynomials,  hence,
+    that $P_{(S, \vec{r})}$ is a \kl{strongly natural binomial polynomial}
+    for all $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. We have proven that
     $f \in \NPoly[d]$.
@@
 
-    Let us now define the function $g \colon \Sigma^* \to \Rel$ as follows:
+    Let us conclude by computing $f$ 
+    using the following \kl{star-free $\Nat$-polyregular function} $g \colon \Sigma^* \to \Rel$:
     \begin{equation*}
@@
         \end{cases}
+        \qedhere
     \end{equation*}
-    We conclude that $g$ is a \kl{star-free $\Nat$-polyregular function}, and by definition
-    $f = g$.
 \end{proof}

### 2024-11-29 parts/prelim.tex
knowledge correction

@@
     $f,g \in \NPoly[d]$ (resp. $\ZPoly[d]$, $\NSF[d]$, $\ZSF[d]$),
-    $L$ be a \kl{star-free language} over $\Sigma^*$,
+    $L$ be a star-free language over $\Sigma^*$,
     and $h \colon \Sigma^* \to \Gamma^*$ be a \kl{polyregular function}

### 2024-12-01 parts/appendix.tex
pushing to arxiv

@@
     because the products of binomial coefficients translated by $K$ form a
-    basis of $\Rel[\vec{X}]$), we conclude that for all $\card{\alpha} \leq d$,
+    basis of $\Rat[\vec{X}]$), we conclude that for all $\card{\alpha} \leq d$,
     $n c_{\alpha} = f_{\alpha}$. In particular,

### 2024-12-01 parts/prelim.tex
pushing to arxiv

@@
 \intro{maximal monomial} if it is a maximal element for the \kl{divisibility
-preordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
-+ X + Y$, the set of \reintro{maximal monomials} is $\set{X^2,  -2 XY,  Y^2}$.
-In the polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative}
-\kl{monomials} are $X^2$ and $Y^2$.
+preordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY +
+Y^2 + X + Y$, the set of \reintro{maximal monomials} is $\set{X^2,  -2 XY,
+Y^2}$. For instance, the \kl{non-negative} \kl{monomials} of $P(X,Y) \defined
+(X - Y)^2$ are $X^2$ and $Y^2$.
 
@@
     \label{q-polynomial-n-poly:ex}
-    Let $f \colon \Sigma^* \to \Nat$ that maps a word $w$
+    Let $f \colon \Sigma^* \to \Nat$ be the function that maps a word $w$
     to the number of distinct pairs of positions in $w$,
@@
     $i < j$ in a word $w$ is in bijection with the set $D_w$ of
-    decompositons of the form $w = xyz$,
+    decompositions of the form $w = xyz$,
     where $x$ and $y$ are non-empty, 

### 2024-12-01 parts/starfree.tex
pushing to arxiv

@@
     \kl{$\Nat$-polyregular functions} and \kl{integer binomial polynomials} are
-    both closed under multiplication by a constant factor, we can assume that
+    both closed under multiplication of their input by a constant factor, we can assume that
     $\omega = \omega_2$ in the decomposition of $f$.

### 2025-09-09 globals/metadata.tex
better knowledges
fix: typos + funding

@@
 
-\title{Commutative $\Nat$-rational series of polynomial growth}
+\title{Commutative $\Nat$-rational Series of Polynomial Growth}
 
@@
        {ad.lopez@uw.edu.pl}%
-       {https://orcid.org/0000-0002-1825-0097}%
-       {}
+       {https://orcid.org/0000-0002-4205-327X}%
+       {Aliaume Lopez was supported by the Polish National Science Centre (NCN) grant “Polynomial
+finite state computation” (2022/46/A/ST6/00072).}
 
@@
 \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-\relatedversiondetails[linktext={arXiv:2404.02232}, cite={lopez2024com:arxiv}]{Full Version}{https://arxiv.org/abs/2404.02232}
+\relatedversiondetails[linktext={arXiv:2404.02232}, cite={LOPEZ24a}]{Full Version}{https://arxiv.org/abs/2404.02232}
 
-\ccsdesc[500]{Theory of Computation~Models of Computation}
-\ccsdesc[500]{Theory of computation~Automata extensions}
+\ccsdesc{Theory of computation~Quantitative automata}
+\ccsdesc{Theory of computation~Transducers}
 
@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{Event Short Title}
-\EventAcronym{Event Acronym}
+\EventEditors{Olaf Beyersdorff, Micha\l{} Pilipczuk, Elaine Pimentel, and Nguyen Kim Thang}
+\EventNoEds{4}
+\EventLongTitle{42nd International Symposium on Theoretical Aspects of Computer Science (STACS 2025)}
+\EventShortTitle{STACS 2025}
+\EventAcronym{STACS}
 \EventYear{2025}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventDate{March 4--7, 2025}
+\EventLocation{Jena, Germany}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{285}
+\SeriesVolume{327}
+\ArticleNo{58}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-09-22 pnrs.tex
rewrote intro
publish to arxiv
moving fact to appendix
knowledges
up to 227
update for CSL«
test
wohooh
update rebuttal
better knowledges
more typos
pushing to arxiv
fix: typos + funding
new: upgrade to paper-meta

--- 2024-01-17 pnrs.tex
@@
-%! TeX program = xelatex
-%! lang = en-US
-\documentclass[sigconf,natbib=false,screen, review,anonymous]{acmart}
+%! TEX program = pdflatex
+% WARNING: this is a generated file.
+%
+% Please do not edit this file directly. 
+% - If you want to update the medatata of the paper (title, authors, abstract), please
+%   edit the `paper-meta.yaml` file in the root of the repository.
+% - If you want to update the content of the paper, please edit the latex files
+%   in the `src` directory.
+% - If you want to update the template itself (e.g., change the layout), please
+%   edit the `templates/plain-article.tex` file instead.
+\documentclass[9pt,a4paper,twosided]{article}
 
-% INPUTS
-\input{globals/packages}
-\input{globals/colorscheme}
+% we setup a custom geometry because the default one is too narrow
+\usepackage{geometry}
+\geometry{margin=3.5cm}
 
-% CONFIG
-\input{globals/macros}
-\input{globals/mathdefs}
+% utf-8 for old systems
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+
+% babel for language settings
+\usepackage[english]{babel}
+
+% microtype for better typography
+\usepackage{microtype}
+
+\usepackage{todonotes}
+\usepackage{lineno}
+
+% math packages
+\usepackage{amsmath,amsthm,amssymb,stmaryrd,thmtools,upgreek}
+
+% configure some theorems
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{conjecture}[theorem]{Conjecture}
+\newtheorem{assumption}[theorem]{Assumption}
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem{example}[theorem]{Example}
+
+% graphics packages
+\usepackage{graphicx}
+\usepackage[obeyclassoptions,mode=tex]{standalone}
+\usepackage{tikz}
+\usetikzlibrary{backgrounds}
+\usetikzlibrary{shapes.geometric}
+\usetikzlibrary{decorations.markings}
+\usetikzlibrary{positioning}
+\usetikzlibrary{automata}
+\usetikzlibrary{tikzmark}
+\usetikzlibrary{patterns}
+\usetikzlibrary{arrows}
+
+\tikzset{every state/.style={minimum size=1pt}}
+\usepackage{tikz-cd}
+
+% ornaments
+\usepackage{pgfornament}
+
+% links inside the document
+\usepackage{hyperref}
+\usepackage[capitalise,noabbrev,nameinlink]{cleveref}
+\Crefname{assumption}{Assumption}{Assumptions}
+\usepackage[composition,hyperref,xcolor,cleveref]{knowledge}
+\knowledgeconfigure{notion}
+
+% Tables 
+\usepackage{booktabs}
+\usepackage{varwidth}
+
+% Algorithms
+\usepackage{algorithm2e}
+\Crefname{algocfline}{Algorithm}{Algorithms}
+\crefname{algocfline}{Algorithm}{Algorithms}
+\crefname{algocf}{Algorithm}{Algorithms}
+\Crefname{algocf}{Algorithm}{Algorithms}
+
+% Packages for macro definitions
+\usepackage{xparse}
+\usepackage{xpatch}
+\usepackage{tokcycle}
+\usepackage{ifthen}
+
+% Proofs
+\usepackage{bussproofs}
+
+% Colors 
+\usepackage{ensps-colorscheme}
+% parametrize knowledge colors and style
+% intro => A4 + emph 
+% kl    => normal color, normal size not emph
+\knowledgestyle{intro notion}{color={A2}, emphasize}
+\knowledgestyle{notion}{color={B1}}
+\hypersetup{
+    colorlinks=true,
+    breaklinks=true,
+    linkcolor=A4,
+    citecolor=A4,
+    urlcolor=A4,
+    filecolor=A4,
+}
+
+% Knowledge logo
+\newcommand{\klogo}{%
+\begin{tikzpicture}[scale=0.2,line/.style={draw, line width=0.2pt, line cap=round, line join=round}]
+\coordinate (A00) at (0,0);
+\coordinate (A01) at (0,1);
+\coordinate (A10) at (1,0);
+\coordinate (B10) at (1,0.2);
+\coordinate (B01) at (0.2,1);
+
+\coordinate (C01) at (0.4,0.7);
+\coordinate (C10) at (0.7,0.4);
+\coordinate (C12) at (0.4,1.2);
+\coordinate (C21) at (1.2, 0.4);
+\coordinate (C22) at (1.2, 1.2);
+
+\coordinate (D00) at (C10);
+\coordinate (D01) at (0.8,0.5);
+\coordinate (D10) at (0.8,0.3);
+
+\coordinate (E01) at (0.3,0.7);
+\coordinate (E10) at (0.5,0.7);
+
+\draw[line] (B01) -- (A01) -- (A00) -- (A10) -- (B10);
+\draw[line] (C01) -- (C12) -- (C22) -- (C21) -- (C10);
+
+\draw[line] (D01) -- (D00) -- (D10);
+\draw[line] (E01) -- (E10);
+
+\end{tikzpicture}%
+}
+
+% we include whatever the user wants to include in the header
+
+% we include libraries (tex files) usually written in the `lib` directory
+\input{globals/mathdefs.tex}
 \input{globals/knowledges.kl}
-\input{globals/metadata}
+\input{globals/macros.tex}
 
-\usepackage{comment}
-%\includecomment{forshort}
-%\excludecomment{proof}
+% We include the title and author information based on the 
+% `paper-meta.yaml` file.
 
-\makeabstract
-% DOCUMENT
+\title{Commutative \(\Nat\)-rational Series of Polynomial Growth}
+
+\author{
+Aliaume Lopez\thanks{University of Warsaw}
+}
+
+% For the date, we first check if the user has provided a date,
+% and otherwise use the git meta inforamtion (if available).
+\date{2025-09-22 10:31:27
++0200\footnote{00d33fe682fe4f73d3af284af9d9a8d732af14d2 -- branch theoretics at git@github.com:AliaumeL/polynomial-n-rational-series.git}}
+
+\newcommand{\repositoryUrl}{\url{https://github.com/AliaumeL/polynomial-n-rational-series}}
+
+\knowledge{notion}
+ | kl-usage
+
+% Now, we create the document itself.
 \begin{document}
+% Generate the title page
+\maketitle
+% Print the abstract
+\begin{abstract}
+    This paper studies which functions computed by \(\Rel\)-weighted
+    automata can be realised by \(\Nat\)-weighted automata, under two
+    extra assumptions: commutativity (the order of letters in the input
+    does not matter) and polynomial growth (the output of the function
+    is bounded by a polynomial in the size of the input). We leverage
+    this effective characterization to decide whether a function
+    computed by a commutative \(\Nat\)-weighted automaton of polynomial
+    growth is star-free, a notion borrowed from the theory of regular
+    languages that has been the subject of many investigations in the
+    context of string-to-string functions
+    \paragraph{Keywords:}
+    Rational series, Weighted automata, Polyregular
+function, Commutative function
+\paragraph{Repository:} \repositoryUrl
+\paragraph{conference} \hyperlink{https://doi.org/10.4230/LIPIcs.STACS.2025.67}{STACS'25} \cite{lopez:LIPIcs.STACS.2025.67}
+\end{abstract}
 
-% FRONT MATTER
-\maketitle
-\acknowledge
+\klogo\ This document uses \href{https://ctan.org/pkg/knowledge}{knowledge}:
+\kl[kl-usage]{notion} points to its \intro[kl-usage]{definition}.
 
-\input{parts/intro.tex}
+% Include the content of the paper
+\lowcotwo~This is a low-co2 research paper:
+\lowcotwourl[\lowcotwoversion]. This research was developed, written,
+submitted and presented without the use of air travel.
+
+\input{parts/newintro.tex}
 \input{parts/prelim.tex}
@@
 
-% BACKMATTER
-\printbibliography
+% Include the bibliography
+\bibliographystyle{plainurl}
+\bibliography{short.bib}
 
-\appendix
+% If there are any appendices, we include them here.
 

### 2025-09-25 parts/newintro.tex
fix: pandoc + new intro

@@
 % MAIN MATTER
-\section{Introduction}
-\label{introduction:sec}
+\section{Introduction}\label{introduction:sec}
 
-\AP Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class of
-\intro{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
-$\Sigma^*$ to $\mathbb{S}$ that are computed by \kl{$\mathbb{S}$-weighted
-automata} \cite{BERE10}. This computational model is a generalization of the
-classical notion of non-deterministic finite automata to the weighted setting,
-where transitions are labeled with elements of $\mathbb{S}$. The semantics of
-\kl{$\mathbb{S}$-weighted automata} on a given word $w$ is defined by the sum
-over all accepting runs reading $w$, of the product of the weights of the
-transitions taken along this run. In this paper, we are interested in the case
-where $\mathbb{S}$ equals $\Nat$ or $\Rel$, hence, in $\Nat$-rational series
-($\intro*\NRat$) and $\Rel$-rational series ($\intro*\ZRat$). It is clear that
-$\NRat$ is a proper subclass of $\ZRat$, and a longstanding open problem is to
-provide an algorithm that decides whether a given $\ZRat$ is in $\NRat$
-\cite{KARH77}.
+\AP\ Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class
+of \intro{(noncommutative) $\mathbb{S}$-rational series} is defined as
+functions from $\Sigma^*$ to $\mathbb{S}$ that are computed by
+\kl{$\mathbb{S}$-weighted automata}~\cite{BERE10}. This computational model is
+a generalization of the classical notion of non-deterministic finite automata
+to the weighted setting, where transitions are labeled with elements of
+$\mathbb{S}$. The semantics of \kl{$\mathbb{S}$-weighted automata} on a given
+word $w$ is defined by the sum over all accepting runs reading $w$, of the
+product of the weights of the transitions taken along this run. When the
+semiring $\mathbb{S}$ is the Boolean semiring $\mathbb{B}$, one recovers the
+classical class of regular languages via non-deterministic finite automata.
 
-\begin{problem}
-    \label{n-in-z-rat:problem}
+\begin{example}\label{ab-ba-power-two:ex}
+  The function $f \colon \set{a,b}^* \to \Nat$ that maps a word $w$ to
+  $2^{|w|_a - |w|_b} + 2^{|w|_b - |w|_a}$ is a
+  $\Rat$-rational series.
+  It can be computed by the following $\Rat$-weighted automaton.
+  \begin{center}
+    \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
+      % initial state q_0 edge from above it
+      \node[state,initial above
+      ] (q_0)   {$q_0$};
+      \node[state,accepting] (q_1) [right=of q_0] {$q_1$};
+      \node[state,accepting] (q_2) [left=of q_0] {$q_2$};
+      \path[->]
+        (q_0) edge [bend left] node {$a \mid 2$} (q_1)
+              edge [bend left] node {$a \mid 0.5$} (q_2)
+              edge [bend right] node [swap] {$b \mid 2$} (q_2)
+              edge [bend right] node [swap] {$b \mid 0.5$} (q_1)
+        (q_1) edge [loop above] node {$a \mid 2$} ()
+              edge [loop below] node {$b \mid 0.5$} (q_0)
+        (q_2) edge [loop above]  node {$b \mid 2$} ()
+              edge [loop below] node {$a \mid 0.5$} ()
+              ;
+        \node[draw=none] (outq1) [right=1cm of q_1] {$1$};
+        \node[draw=none] (outq2) [left=1cm of q_2] {$1$};
+        \node[draw=none] (outq0) [below=1cm of q_0] {$1$};
+        \path[->]
+          (q_1) edge [] (outq1)
+          (q_2) edge [] (outq2)
+          (q_0) edge [] (outq0)
+          ;
+    \end{tikzpicture}
+  \end{center}
+\end{example}
+
+\AP\ When the semiring $\mathbb{S}$ is a commutative field, for instance in the
+case of the rational numbers $\Rat$, the class of \kl{$\mathbb{S}$-rational
+series} is well understood, and enjoys many decidability properties by virtue
+of its correspondence with \emph{linear representations}~\cite{BERE10}.
+From this correspondence, one can for instance decide whether two
+$\Rat$-rational series are equal, or prove that a 
+$\Real$-rational series that has values in $\Rat$ is in fact a 
+$\Rat$-rational series.
+However, the situation is much less clear when $\mathbb{S}$ is merely a
+commutative ring or worse, a commutative semiring: the corresponding theories
+of \emph{modules} or \emph{semimodules} are much more involved than the one for
+vector spaces, and many questions are open. For instance, the function $f$ from
+\cref{ab-ba-power-two:ex} cannot be computed by a $\Nat$-weighted automaton,
+even though it is a $\Rat$-rational series that only takes natural values. Even
+more, we can prove that $f$ cannot be computed by any $\Rel$-weighted
+automaton.
+
+In this paper, we are interested in the case where $\mathbb{S}$ equals $\Nat$
+(semiring) or $\Rel$ (ring), hence, in $\Nat$-rational series ($\intro*\NRat$)
+and $\Rel$-rational series ($\intro*\ZRat$). It is clear that $\NRat$ is a
+proper subclass of $\ZRat$, and a longstanding open problem is to provide an
+algorithm that decides whether a given $\ZRat$ is in $\NRat$~\cite{KARH77}.
+
+\begin{problem}\label{n-in-z-rat:problem}
     Input: A $\ZRat$ $f$. Output: Is $f$ in $\NRat$?
 \end{problem}
+
+A similar open problem has been studied in the context of rational weighted
+automata, where it amounts to deciding whether a given $\Rat$-weighted
+automaton is equivalent to a $\Rat_+$-weighted automaton. Let us briefly
+explain why this problem is equivalent to \cref{n-in-z-rat:problem}. First,
+given a $\Rat$-weighted automaton $A$, one can construct a $\Rel$-weighted
+automaton $A'$ by multiplying all weights by the least common multiple of the
+denominators of the weights of $A$. Semantically, $A'(w) = c^{|w|} A(w)$ for
+some constant $c \in \Nat$. If $A'$ is equivalent to an $\Nat$-weighted
+automaton $B$, then $A$ is $c^{-|w|} B(w)$, hence equivalent to a
+$\Rat_+$-weighted automaton. For the converse implication, if $A$ is a
+$\Rat$-weighted automaton equivalent to a $\Rat_+$-weighted automaton, then
+$A'$ is also equivalent to some $\Rat_+$-weighted automaton $B'$, but all the
+outputs of $B'$ are natural numbers, and because $\Rat_+$ is a Fatou extension
+of $\Nat$,\footnote{rework this} one concludes that $B'$ is equivalent to an
+$\Nat$-weighted automaton $B''$~\cite{BERE10}.
 
@@
 generalize the theory of regular languages to the setting of string-to-string
-functions \cite{BOJA18}. In the case of regular languages, \emph{star-free
+functions~\cite{BOJA18}. In the case of regular languages, \emph{star-free
 languages} form a robust subclass of regular languages described equivalently
@@
 restricting the output alphabet to a single letter has proven to be a fruitful
-method \cite{DOUE21,DOUE22}. Because words over a unary alphabet are
+method~\cite{DOUE21,DOUE22}. Because words over a unary alphabet are
 canonically identified with natural numbers, unary output \kl{polyregular
@@
 \kl{$\Nat$-rational series} of \emph{polynomial growth}, i.e. the output of the
-function is bounded by a polynomial in the size of the input. In \cite{CDTL23},
+function is bounded by a polynomial in the size of the input 
+\cite{SCHU62}.
+In \cite{CDTL23},
 the authors introduced the class of \kl{$\Rel$-polyregular functions}
@@
 also prove that this previous results holds for polynomials with at most two
-indeterminates (\cref{lem:correct-covered-2}), which may explain why
-it was not detected earlier. Furthermore, these characterizations provide
-effective descriptions of polynomials that can be expressed in $\ZRat$ as those
-obtained using integer combinations of products of \emph{binomial coefficients}
-(called \kl{integer binomial polynomials}, defined page \kpageref{integer
-binomial polynomial}) and similarly for $\NRat$ by introducing the notion of
+indeterminates (\cref{lem:correct-covered-2}), which may explain why it was not
+detected earlier. Furthermore, these characterizations provide effective
+descriptions of polynomials that can be expressed in $\ZRat$ as those obtained
+using integer combinations of products of \emph{binomial coefficients} (called
+\kl{integer binomial polynomials}, defined page \kpageref{integer binomial
+polynomial}) and similarly for $\NRat$ by introducing the notion of
 \kl{strongly natural binomial polynomials} (defined page \kpageref{strongly
-natural binomial polynomial}), which we believe has its own interest. Finally,
+natural binomial polynomial}), which we believe has its own interest. In turn,
 these characterizations demonstrate that polynomials expressible by $\ZRat$
@@
 is, polynomials are inherently \emph{star free} functions.
+
+Finally, we show that non-commutative polyregular functions in $\NPoly$ (resp.
+$\NSF$) have a canonical representation in terms of their \emph{residuals}, and
+propose a definition of star-free $\Rel$-rational series and $\Nat$-rational
+series, that we believe is of independent interest.
 

### 2025-09-25 pnrs.tex
fix: pandoc + new intro

@@
 % and otherwise use the git meta inforamtion (if available).
-\date{2025-09-22 10:31:27
-+0200\footnote{00d33fe682fe4f73d3af284af9d9a8d732af14d2 -- branch theoretics at git@github.com:AliaumeL/polynomial-n-rational-series.git}}
+\date{2025-09-22 11:34:34
++0200\footnote{6d8e3bf714a8e6f6efba8be209bb9ae529722a78 -- branch theoretics at git@github.com:AliaumeL/polynomial-n-rational-series.git}}
 
@@
 \paragraph{Repository:} \repositoryUrl
-\paragraph{conference} \hyperlink{https://doi.org/10.4230/LIPIcs.STACS.2025.67}{STACS'25} \cite{lopez:LIPIcs.STACS.2025.67}
+\paragraph{conference} \href{https://doi.org/10.4230/LIPIcs.STACS.2025.67}{STACS'25} \cite{lopez:LIPIcs.STACS.2025.67}
 \end{abstract}
